---
title: "Ch1_SW_Differences"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---


```{r,echo=FALSE}
  knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

```{r,echo=FALSE}
library(tidyverse)
library(ggplot2)
library(dataRetrieval)
library(knitr)
library(ggplot2)
library(zoo) #for moving window
library(stringr)
library(vioplot)
library(purrr)
library(lavaan)
library(blavaan)
library(ggcorrplot)
library(lme4)
library(broom.mixed)
```


The purpose of this script is to:

    - ORGANIZE DATA 
    - Import and arrange .csv files of finalized paired sample data

    - TEST MODELs
    - Model difference in SWA between dammed and undammed pairs as a function of
      - Dam Presence
      - Topographic Openness
      - Riparian Area
      - Contributing Drainage Area
      
    - Model 1: Dam Presence Only
    - Model 2: Dam Presence + Context
      
----

Read in Flow data from USGS LBV Guston Coulee gage:

```{r,echo=FALSE, warning=FALSE}
# LOWER BEAVER CK
# Gage near Guston Coulee
usgsGage_06166000 <- readNWISdv(
  siteNumbers = "06166000",
  parameterCd = "00060",
  startDate = "2019-06-01",
  endDate   = "2023-10-01"
)

###########################################

# Clean USGS
usgs_df <- usgsGage_06166000 %>%
  mutate(
    source = "USGS",
    date   = as.Date(Date),
    flow   = X_00060_00003
    ) %>%
  filter(
    (date >= as.Date("2019-06-01") & date <= as.Date("2019-10-01")) |
    (date >= as.Date("2021-06-01") & date <= as.Date("2021-10-01")) |
    (date >= as.Date("2023-06-01") & date <= as.Date("2023-10-01")) ) %>% 
  mutate(
    zFlow = scale(flow)
  ) %>% 
  select(source,
         date,
         zFlow) %>% 
  mutate(surveyPeriod = case_when(between(date, as.Date("2019-06-01"), as.Date("2019-06-14")) ~ 191,
                                between(date, as.Date("2019-06-15"), as.Date("2019-06-30")) ~ 192,
                                between(date, as.Date("2019-07-01"), as.Date("2019-07-14")) ~ 193,
                                between(date, as.Date("2019-07-15"), as.Date("2019-07-31")) ~ 194,
                                between(date, as.Date("2019-08-01"), as.Date("2019-08-14")) ~ 195,
                                between(date, as.Date("2019-08-15"), as.Date("2019-08-31")) ~ 196,
                                between(date, as.Date("2019-09-01"), as.Date("2019-09-14")) ~ 197,
                                between(date, as.Date("2019-09-15"), as.Date("2019-09-30")) ~ 198,
                                between(date, as.Date("2021-06-01"), as.Date("2021-06-14")) ~ 211,
                                between(date, as.Date("2021-06-15"), as.Date("2021-06-30")) ~ 212,
                                between(date, as.Date("2021-07-01"), as.Date("2021-07-14")) ~ 213,
                                between(date, as.Date("2021-07-15"), as.Date("2021-07-31")) ~ 214,
                                between(date, as.Date("2021-08-01"), as.Date("2021-08-14")) ~ 215,
                                between(date, as.Date("2021-08-15"), as.Date("2021-08-31")) ~ 216,
                                between(date, as.Date("2021-09-01"), as.Date("2021-09-14")) ~ 217,
                                between(date, as.Date("2021-09-15"), as.Date("2021-09-30")) ~ 218,
                                between(date, as.Date("2023-06-01"), as.Date("2023-06-14")) ~ 231,
                                between(date, as.Date("2023-06-15"), as.Date("2023-06-30")) ~ 232,
                                between(date, as.Date("2023-07-01"), as.Date("2023-07-14")) ~ 233,
                                between(date, as.Date("2023-07-15"), as.Date("2023-07-31")) ~ 234,
                                between(date, as.Date("2023-08-01"), as.Date("2023-08-14")) ~ 235,
                                between(date, as.Date("2023-08-15"), as.Date("2023-08-31")) ~ 236,
                                between(date, as.Date("2023-09-01"), as.Date("2023-09-14")) ~ 237,
                                between(date, as.Date("2023-09-15"), as.Date("2023-09-30")) ~ 238,
                                TRUE ~ NA_real_)) %>% 
  filter(surveyPeriod > 0)

usgs_df$surveyPeriod <- as.factor(usgs_df$surveyPeriod)


mean_flow_per_survey_period <- usgs_df %>% 
  group_by(surveyPeriod) %>%
  summarise(zFlow_mean = mean(zFlow),
            .groups = "drop")



lateSummerFlow <- usgs_df %>% 
  filter(surveyPeriod != 191,
         surveyPeriod != 192,
         surveyPeriod != 193,
         surveyPeriod != 194,
         surveyPeriod != 211,
         surveyPeriod != 212,
         surveyPeriod != 213,
         surveyPeriod != 214,
         surveyPeriod != 231,
         surveyPeriod != 232,
         surveyPeriod != 233,
         surveyPeriod != 234,
         )
```


2019:
----

High early/mid-summer flow followed by moderate/low late season. 

```{r,echo=FALSE}

flow19 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2019-06-01") & date <= as.Date("2019-10-01"))
  )


####

ggplot(flow19, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '19 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```

----
  
2021:
----

Low flow conditions throughout the summer season. Part of 3 year sustained drought conditions (2020, 2021, 2022).



```{r,echo=FALSE}

flow21 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2021-06-01") & date <= as.Date("2021-10-01"))
  )


####

ggplot(flow21, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '21 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```  
  
----

2023:
----

High early flow conditions followed by moderate to lower late summer. 



```{r,echo=FALSE}
flow23 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2023-06-01") & date <= as.Date("2023-10-01"))
  )


####

ggplot(flow23, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '23 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```  





```{r, echo=FALSE}

#Read in data frames for each HUC12/Year combo:

# WILLOW:
#2019:
wc_19 <- read.csv("WC_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'WC')
#2021:
wc_21 <- read.csv("WC_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'WC')
#2023:
wc_23 <- read.csv("WC_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'WC')
#AllYears:
WC <- bind_rows(wc_19,wc_21,wc_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LONE TREE:
#2019:
lt_19 <- read.csv("LT_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'LT')
#2021:
lt_21 <- read.csv("LT_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'LT')
#2023:
lt_23 <- read.csv("LT_2023_DF_Final_Long_0427Update_250Spacing.csv")%>% 
  mutate(year = 2023,
         watershed = 'LT')

#AllYears:
LT <- bind_rows(lt_19,lt_21,lt_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))

# BRAZIL:
#2019:
brz_19 <- read.csv("BRZ_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'BRZ')
#2021:
brz_21 <- read.csv("BRZ_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'BRZ')
#2023:
brz_23 <- read.csv("BRZ_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'BRZ')

#AllYears:
BRZ <- bind_rows(brz_19,brz_21,brz_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LARB:
#2019:
lrb_19 <- read.csv("LRB_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'LRB')
#2021:
lrb_21 <- read.csv("LRB_2021_DF_Final_Long_0427Update_250Spacing.csv")%>% 
  mutate(year = 2021,
         watershed = 'LRB')
#2023:
lrb_23 <- read.csv("LRB_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'LRB')

#AllYears:
LRB <- bind_rows(lrb_19,lrb_21,lrb_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LOWER BEAVER:
#2019:
lbv_19 <- read.csv("LBV_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'LBV')
#2021:
lbv_21 <- read.csv("LBV_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'LBV')
#2023:
lbv_23 <- read.csv("LBV_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'LBV')

#AllYears:
LBV <- bind_rows(lbv_19,lbv_21,lbv_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# UPPER BEAVER:
#2019:
ubv_19 <- read.csv("UBV_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'UBV')
#2021:
ubv_21 <- read.csv("UBV_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'UBV')
#2023:
ubv_23 <- read.csv("UBV_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'UBV')

#AllYears:
UBV <- bind_rows(ubv_19,ubv_21,ubv_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
```


----

Currently each incoming data frame is of the following structure:

```{r,echo=FALSE}
str(brz_19)
```

Where:

    - clusterID_seq = dammed/undammed sample pair ID
    - window_type = whether a sample window is 'dammed' or 'undammed'
    - watershed = watershed identifier (e.g., "BRZ" = Brazil Ck)
    - year = year
    - window_length = length (m) of stream flow line within sample window
    - catchment_km2 = contributing catchment area/upstream drainage area (km^2)
    - riparian_area_m2 = riparian area (m^2) of sample window; 2-m vertical
      buffer from DEM low point (stream flow line)
    - TO = mean topographic openness across window
    - p_window: avg dam detection prob for target window; see CP Ch1
    - f_window: avg dam false positive prob for target window; see CP Ch1
    - beav_count_all: all unique detected dams
    - beav_count_conf: unique confidently detected dams
    - anthro_count: all unique detected anthropogenic dams
    - isAnthro: National Wetland Inventory (NWI) "dyked/impounded"
    - Nhat = estimated dam count within sample window
      - ...= beav_count_all/(p_window(1-f_window)  [McCallum 2005]
    - TO_mean = mean topographic openness across window
    - time_period = 2wk interval 06/01 - 09/30
    - SWA = surface water area (m^2) for window during time (t)
    - t = time (t) {same as time_period}

----
----
----
----

WILLOW
----
----

Compute **Riparian Inundation**

For survey window[i] at time[t], the proportion of riparian area that is covered in surface water

$SWA = Surface Water Area (m^2)$

$RIPA = Riparian Area (m^2)$

$$RI[i,t] = SWA[i,t] / RIPA[i] $$


```{r, echo=FALSE}
# COMPUTE RI HERE!!!
WC_RI <- WC %>% 
  mutate(RI = SWA/riparian_area_m2)
```

Visualize riparian inundation **(RI)** $SWA/RIPA$ between dammed and undammed across years.

```{r, echo=FALSE, fig.width=10}
WC_RI %>%
  filter(!is.na(RI)) %>%   # optional but recommended
  ggplot(aes(x = factor(year),
             y = RI,
             fill = window_type)) +
  geom_boxplot(outlier.alpha = 0.4) +
  scale_fill_manual(values = c("dammed" = "brown",
                               "undammed" = "steelblue")) +
  labs(
    x = "Year",
    y = "Riparian Inundation (RI)",
    fill = "Window type",
    title = "RI across dammed vs undammed windows",
    subtitle = "Willow Creek (2019, 2021, 2023)"
  ) +
  theme_bw(base_size = 14)

```

```{r,echo=FALSE}
WC_RI %>%
  filter(year == 2019) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Willow Ck; Summer 2019") +
     theme_bw(base_size = 15)

```

```{r,echo=FALSE}
WC_RI %>%
  filter(year == 2021) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Willow Ck; Summer 2021") +
     theme_bw(base_size = 15)

```

```{r,echo=FALSE}
WC_RI %>%
  filter(year == 2023) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Willow Ck; Summer 2023") +
     theme_bw(base_size = 15)

```

----

Add **relative flow data**. 

NOTE: currently this represents data from the USGS Guston Coulee gage that has been z-standardized over the current data acquisition range (June-Sept; 2019, 2021, 2023). 

Then, for each 2-week survey time period (1:8), the average relative flow $[mean(zFlow)]$ was computed, and added as a data column. IT IS ASSUMED that this flow data is representative of the RELATIVE flow pattern experienced throughout the study area during this time. Consider incorporating relative HHP simulated flow data again (?)

----

Visualize Surface Water Area across Relative Flow and Topographic Openness


```{r,echo=FALSE, fig.width=10}
WC_RelFlow <- WC_RI %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = Nhat/window_length)


WC_RelFlow %>% 
  arrange(SWA) %>%
  ggplot(
       aes(x = zFlow_mean, 
           y = TO, 
           color = SWA,
           shape = window_type)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "mako",begin = 0,end   = 1,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = "Topographic Openness (TO)",
    color = expression("Surface Water Area" (m^2)),
    title = "Surface Water Area Across RelativeFlow and TO space",
    subtitle = "Willow Creek All Years"
  ) +
  theme_bw(base_size = 14)
```


----

----  

----

----



Visualize covariate distributions b/t dammed and undammed windows:

```{r,include=FALSE}
ggplot(WC_RelFlow, 
       aes(x = TO, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)


ggplot(WC_RelFlow, 
       aes(x = riparian_area_m2, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(WC_RelFlow, 
       aes(x = SWA, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(WC_RelFlow, 
       aes(x = RI, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(WC_RelFlow, 
       aes(x = catchment_km2, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)
```



----

Visualize covariate correlations for Willow Creek :

```{r,echo=FALSE, fig.width=10}
#2019
wc19_numCols <- WC_RelFlow %>% 
  filter(year == "2019") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
wc19_corr_matrix <- cor(wc19_numCols)

# Create the plot
wc19_ggcorr_plot <- ggcorrplot(wc19_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Willow Creek 2019") +
  theme(plot.title = element_text(hjust = 0.5))

wc19_ggcorr_plot

#2021
wc21_numCols <- WC_RelFlow %>% 
  filter(year == "2021") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
wc21_corr_matrix <- cor(wc21_numCols)

# Create the plot
wc21_ggcorr_plot <- ggcorrplot(wc21_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Willow Creek 2021") +
  theme(plot.title = element_text(hjust = 0.5))

wc21_ggcorr_plot


#2023
wc23_numCols <- WC_RelFlow %>% 
  filter(year == "2023") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
wc23_corr_matrix <- cor(wc23_numCols)

# Create the plot
wc23_ggcorr_plot <- ggcorrplot(wc23_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Willow Creek 2023") +
  theme(plot.title = element_text(hjust = 0.5))

wc23_ggcorr_plot
```

----

----

----

Compute Differences (Dammed[i] - Undammed):
----

Compute difference in RI $(\Delta RI)$ between dammed and undammed pairs. The hypothesis is that dammed reaches will inundate more of their respective available space (m^2) than their paired undammed reaches. (+)$(\Delta RI)$ means the dammed site in a pair inundated more of its available riparian space than its undammed counterpart; (-) is vice versa.

The degree to which beaver dams influence $\Delta RI$ confounded by pairwise differences in reach characteristics (e.g., $TO$, $UDA$)

----

Proposed model:

$$\Delta RI[i] = \alpha + \beta1 * \Delta UDA[i] + \beta2 * \Delta TO + \beta3 *  Nhat[i_{\text{dammed}}]$$


Wherein, the difference in riparian inundation $(\Delta RI)$ for sample pair $[i]$ is a linear function of the pair's difference in stream network position $(\Delta UDA)$, the pair's difference in local channel confinement $(\Delta TO)$, and the pair's difference in number of dams $(\Delta Nhat)$ (NOTE: delta_Nhat is just Nhat of the dammed site in pair [i]).

Consider, instead of Nhat, using $dam/m = Nhat/windowLength (m)$ to get at dam density within the window, rather than raw dam count.


```{r,echo=FALSE}
wc_sample_pairs <- WC_RelFlow %>%
  select(clusterID_seq,
         year,
         t,
         surveyPeriod,
         window_type, 
         RI, 
         TO, 
         riparian_area_m2, 
         zFlow_mean,
         catchment_km2,
         Nhat,
         dam_m) %>%
  pivot_wider(
    names_from = window_type,
    values_from = c(RI, TO, riparian_area_m2, catchment_km2, Nhat,dam_m),
    names_sep = "_"
  ) %>% 
  mutate(delta_RI = RI_dammed - RI_undammed)

wc_sample_differences <- wc_sample_pairs %>%
  mutate(
    delta_TO   = TO_dammed - TO_undammed,
    delta_RIPA = riparian_area_m2_dammed - riparian_area_m2_undammed,
    delta_UDA  = catchment_km2_dammed - catchment_km2_undammed,
    delta_Nhat = Nhat_dammed,
    delta_dam_m = dam_m_dammed
  ) %>% 
  select(clusterID_seq,
         year,
         t,
         surveyPeriod,
         zFlow_mean,
         delta_RI,
         delta_TO,
         delta_RIPA,
         delta_UDA,
         delta_Nhat,
         delta_dam_m
         )

str(wc_sample_differences)
```


----

Visualize $\Delta RI$ across $\Delta Nhat$

```{r, echo=FALSE, fig.width=10}
wc_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2019') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Willow Creek 2019"
  ) +
  theme_bw(base_size = 14)


#####

wc_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2021') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Willow Creek 2021"
  ) +
  theme_bw(base_size = 14)


#####

wc_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2023') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Willow Creek 2023"
  ) +
  theme_bw(base_size = 14)

```


----


----

Visualize $\Delta RI$ across **damDensity** $(dam/m)$

```{r, echo=FALSE, fig.width=10}
wc_sample_differences %>% 
  filter(year=='2019') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Willow Creek 2019"
  ) +
  theme_bw(base_size = 14)

#####

wc_sample_differences %>% 
  filter(year=='2021') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Willow Creek 2021"
  ) +
  theme_bw(base_size = 14)

#####

wc_sample_differences %>% 
  filter(year=='2023') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Willow Creek 2023"
  ) +
  theme_bw(base_size = 14)

#####
```


----

Dig into where $\Delta RI >0$; these illustrate contexts where dams may be positively influencing SWA.

Look at TO and RIPA and UDA for delta_RI > 0 dam clusters

```{r,echo=FALSE, fig.width=10}
wc_posDamDiff <- wc_sample_pairs %>% 
  filter(delta_RI > 0)

wc_posDamDiff$year <- as.factor(wc_posDamDiff$year)

wc_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = TO_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 5, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = "Topographic Openness (TO)",
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    shape = "Year",
    title = expression("Sites with Positive deltaRI Across Flow and TO"),
    subtitle = "Willow Creek All Years"
  ) +
  theme_bw(base_size = 14)



wc_sample_differences %>%
  filter(delta_RI >0) %>% 
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = delta_TO, 
           color = delta_RI,
           shape = as.factor(year))) +
  geom_point(size = 5, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Difference in Topographic Openness (  "*Delta*" TO)"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    shape = "Year",
    title = expression("Sites with Positive deltaRI Across Flow and " *Delta* "TO"),
    subtitle = "Willow Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
wc_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = riparian_area_m2_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Riparian Area ("*m^2*")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and RIPA"),
    subtitle = "Willow Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
wc_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = catchment_km2_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Upstream Drainage Area ("*km^2*")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and UDA"),
    subtitle = "Willow Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
wc_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = dam_m_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Dam Density (" *DamCount[i]/streamLength(m)[i]* ")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and UDA"),
    subtitle = "Willow Creek All Years"
  ) +
  theme_bw(base_size = 14)
```



```{r,echo=TRUE}
wcRI_Flow <- wc_sample_pairs %>% 
  select(clusterID_seq,
         surveyPeriod,
         zFlow_mean,
         RI_dammed,
         RI_undammed) %>% 
  pivot_longer(
    cols = c(RI_dammed, RI_undammed),
    names_to = "window_type",
    values_to = "RI"
  )

ggplot(wcRI_Flow,
       aes(x = zFlow_mean,
           y = RI,
           color = as.factor(window_type))) +
  scale_color_manual(values = c("RI_dammed" = "brown",
                                "RI_undammed" = "steelblue")) +
  geom_point(alpha = 0.4, size = 2) +
  geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
  labs(
    x = "z-Standardized Flow",
    y = "Riparian Inundation (RI)",
    color = "Window Type",
    title = "Riparian Inundation Across Flow",
    subtitle = "WC All Years"
  ) +
  theme_bw(base_size = 14)
```

----
----
----
----




----
----
----
----

LONE TREE   
----
----

Compute **Riparian Inundation**

For survey window[i] at time[t], the proportion of riparian area that is covered in surface water

$SWA = Surface Water Area (m^2)$

$RIPA = Riparian Area (m^2)$

$$RI[i,t] = SWA[i,t] / RIPA[i] $$


```{r, echo=FALSE}
# COMPUTE RI HERE!!!
LT_RI <- LT %>% 
  mutate(RI = SWA/riparian_area_m2)
```

Visualize riparian inundation **(RI)** $SWA/RIPA$ between dammed and undammed across years.

```{r, echo=FALSE, fig.width=10}
LT_RI %>%
  filter(!is.na(RI)) %>%   # optional but recommended
  ggplot(aes(x = factor(year),
             y = RI,
             fill = window_type)) +
  geom_boxplot(outlier.alpha = 0.4) +
  scale_fill_manual(values = c("dammed" = "brown",
                               "undammed" = "steelblue")) +
  labs(
    x = "Year",
    y = "Riparian Inundation (RI)",
    fill = "Window type",
    title = "RI across dammed vs undammed windows",
    subtitle = "Lone Tree Creek (2019, 2021, 2023)"
  ) +
  theme_bw(base_size = 14)

```


```{r,echo=FALSE}
LT_RI %>%
  filter(year == 2019) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Lone Tree; Summer 2019") +
     theme_bw(base_size = 15)

```

```{r,echo=FALSE}
LT_RI %>%
  filter(year == 2021) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Lone Tree; Summer 2021") +
     theme_bw(base_size = 15)

```

```{r,echo=FALSE}
LT_RI %>%
  filter(year == 2023) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Lone Tree; Summer 2023") +
     theme_bw(base_size = 15)

```


----

Add **relative flow data**. 

----

Visualize Surface Water Area across Relative Flow and Topographic Openness


```{r,echo=FALSE, fig.width=10}
LT_RelFlow <- LT_RI %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = Nhat/window_length)


LT_RelFlow %>% 
  arrange(SWA) %>%
  ggplot(
       aes(x = zFlow_mean, 
           y = TO, 
           color = SWA,
           shape = window_type)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "mako",begin = 0,end   = 1,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = "Topographic Openness (TO)",
    color = expression("Surface Water Area" (m^2)),
    title = "Surface Water Area Across RelativeFlow and TO space",
    subtitle = "Lone Tree Creek All Years"
  ) +
  theme_bw(base_size = 14)
```


----

----  

----

----



Visualize covariate distributions b/t dammed and undammed windows:

```{r,include=FALSE}
ggplot(LT_RelFlow, 
       aes(x = TO, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)


ggplot(LT_RelFlow, 
       aes(x = riparian_area_m2, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(LT_RelFlow, 
       aes(x = SWA, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(LT_RelFlow, 
       aes(x = RI, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(LT_RelFlow, 
       aes(x = catchment_km2, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)
```



----

Visualize covariate correlations for Willow Creek :

```{r,echo=FALSE, fig.width=10}
#2019
lt19_numCols <- LT_RelFlow %>% 
  filter(year == "2019") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
lt19_corr_matrix <- cor(lt19_numCols)

# Create the plot
lt19_ggcorr_plot <- ggcorrplot(lt19_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Willow Creek 2019") +
  theme(plot.title = element_text(hjust = 0.5))

lt19_ggcorr_plot

#2021
lt21_numCols <- LT_RelFlow %>% 
  filter(year == "2021") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
lt21_corr_matrix <- cor(lt21_numCols)

# Create the plot
lt21_ggcorr_plot <- ggcorrplot(lt21_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Willow Creek 2021") +
  theme(plot.title = element_text(hjust = 0.5))

lt21_ggcorr_plot


#2023
lt23_numCols <- LT_RelFlow %>% 
  filter(year == "2023") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
lt23_corr_matrix <- cor(lt23_numCols)

# Create the plot
lt23_ggcorr_plot <- ggcorrplot(lt23_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Willow Creek 2023") +
  theme(plot.title = element_text(hjust = 0.5))

lt23_ggcorr_plot
```

----

----

----

Compute Differences (Dammed[i] - Undammed):
----


```{r,echo=FALSE}
lt_sample_pairs <- LT_RelFlow %>%
  select(clusterID_seq,
         year,
         t,
         surveyPeriod,
         window_type,
         SWA,
         RI, 
         TO, 
         riparian_area_m2, 
         zFlow_mean,
         catchment_km2,
         Nhat,
         dam_m) %>%
  pivot_wider(
    names_from = window_type,
    values_from = c(SWA, RI, TO, riparian_area_m2, catchment_km2, Nhat,dam_m),
    names_sep = "_"
  ) %>% 
  mutate(delta_RI = RI_dammed - RI_undammed,
         delta_SWA = SWA_dammed - SWA_undammed)

lt_sample_differences <- lt_sample_pairs %>%
  mutate(
    delta_TO   = TO_dammed - TO_undammed,
    delta_RIPA = riparian_area_m2_dammed - riparian_area_m2_undammed,
    delta_UDA  = catchment_km2_dammed - catchment_km2_undammed,
    delta_Nhat = Nhat_dammed,
    delta_dam_m = dam_m_dammed
  ) %>% 
  select(clusterID_seq,
         year,
         t,
         surveyPeriod,
         zFlow_mean,
         delta_SWA,
         delta_RI,
         delta_TO,
         delta_RIPA,
         delta_UDA,
         delta_Nhat,
         delta_dam_m
         )

str(lt_sample_differences)
```


----

Visualize $\Delta RI$ across $\Delta Nhat$

```{r, echo=FALSE, fig.width=10}
lt_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2019') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Lone Tree Creek 2019"
  ) +
  theme_bw(base_size = 14)


#####

lt_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2021') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Lone Tree Creek 2021"
  ) +
  theme_bw(base_size = 14)


#####

lt_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2023') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Lone Tree Creek 2023"
  ) +
  theme_bw(base_size = 14)

```


----


----

Visualize $\Delta RI$ across **damDensity** $(dam/m)$

```{r, echo=FALSE, fig.width=10}
lt_sample_differences %>% 
  filter(year=='2019') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Lone Tree Creek 2019"
  ) +
  theme_bw(base_size = 14)

#####

lt_sample_differences %>% 
  filter(year=='2021') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Lone Tree Creek 2021"
  ) +
  theme_bw(base_size = 14)

#####

lt_sample_differences %>% 
  filter(year=='2023') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Lone Tree Creek 2023"
  ) +
  theme_bw(base_size = 14)

#####
```


----

Dig into where $\Delta RI >0$; these illustrate contexts where dams may be positively influencing SWA.

Look at TO and RIPA and UDA for delta_RI > 0 dam clusters

```{r,echo=FALSE, fig.width=10}
lt_posDamDiff <- lt_sample_pairs %>% 
  filter(delta_SWA > 0)

lt_posDamDiff$year <- as.factor(lt_posDamDiff$year)

lt_posDamDiff %>%
  arrange(delta_SWA) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = TO_dammed, 
           color = delta_SWA,
           shape = year)) +
  geom_point(size = 5, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = "Topographic Openness (TO)",
    color = expression("("*Delta*" SWA) (SWA_dammed - SWA_undammed)"),
    title = expression("Sites with Positive deltaSWA Across Flow and TO"),
    subtitle = "Lone Tree Creek All Years"
  ) +
  theme_bw(base_size = 14)



lt_sample_differences %>%
  filter(delta_SWA >0) %>% 
  arrange(delta_SWA) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = delta_TO, 
           color = delta_SWA,
           shape = as.factor(year))) +
  geom_point(size = 5, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Difference in Topographic Openness (  "*Delta*" TO)"),
    color = expression("("*Delta*" SWA) (SWA_dammed - SWA_undammed)"),
    shape = "Year",
    title = expression("Sites with Positive deltaSWA Across Flow and " *Delta* "TO"),
    subtitle = "Lone Tree Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
lt_posDamDiff %>%
  arrange(delta_SWA) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = (riparian_area_m2_dammed-riparian_area_m2_undammed), 
           color = delta_SWA,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Difference in Riparian Area ( " *m^2*")"),
    color = expression("("*Delta*" SWA) (SWA_dammed - SWA_undammed)"),
    title = expression("Sites with Positive deltaSWA Across Flow and RIPA"),
    subtitle = "Lone Tree Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
lt_posDamDiff %>%
  arrange(delta_SWA) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = (catchment_km2_dammed - catchment_km2_undammed), 
           color = delta_SWA,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Upstream Drainage Area ("*km^2*")"),
    color = expression("("*Delta*" SWA) (SWA_dammed - SWA_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and UDA"),
    subtitle = "Lone Tree Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}

lt_sample_pairs <- lt_sample_pairs %>%
  mutate(
    delta_bin = case_when(
      as.numeric(scale(delta_SWA)) < 0        ~ "Negative",
      as.numeric(scale(delta_SWA)) >= 0 & as.numeric(scale(delta_SWA)) <= 0.5   ~ "None",
      as.numeric(scale(delta_SWA)) > 0.5 & as.numeric(scale(delta_SWA)) <  2.5    ~ "Positive_weak",
      as.numeric(scale(delta_SWA)) >= 2.5         ~ "Positive_strong"
    )
  ) %>% 
  filter(!is.na(delta_RI))

lt_sample_pairs$delta_bin <- factor(
  lt_sample_pairs$delta_bin,
  levels = c("Negative", "None", "Positive_weak", "Positive_strong")
)

lt_sample_pairs$year <- as.factor(lt_sample_pairs$year)

lt_sample_pairs <- lt_sample_pairs %>% 
  rename("Year" = "year")

bin_colors <- c(
  Negative = "grey", 
  None   = "grey", 
  Positive_weak   = "lightblue2", 
  Positive_strong = "navy"   
)

lt_sample_pairs %>%
  arrange(delta_SWA) %>% 
  ggplot(
       aes(x = as.numeric(scale(riparian_area_m2_dammed - riparian_area_m2_undammed)), 
           y = as.numeric(scale((TO_dammed - TO_undammed))), 
           color = delta_bin,
           shape = Year)) +
  geom_hline(yintercept = 0, linewidth = 0.5, color = "black") +
  geom_vline(xintercept = 0, linewidth = 0.5, color = "black") +
  geom_point(
  aes(
    x = as.numeric(scale(riparian_area_m2_dammed - riparian_area_m2_undammed)) +
        as.numeric(as.factor(surveyPeriod)) * 0.02
  ),
  size = 5, alpha = 1
) +
  scale_color_manual(values = bin_colors) +
  labs(
  x = expression(Delta*RIPA),
  y = expression(Delta*TO),
  title = expression(Delta * SWA * " Across " * Delta * TO * " and " * Delta * RIPA),
  subtitle = "Lone Tree Creek All Years",
  color = expression(Delta*SWA)
) +
  theme_bw(base_size = 14)

```





```{r,echo=FALSE, fig.width=10}
lt_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = dam_m_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Dam Density (" *DamCount[i]/streamLength(m)[i]* ")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and UDA"),
    subtitle = "Lone Tree Creek All Years"
  ) +
  theme_bw(base_size = 14)
```


```{r,echo=TRUE}
ltRI_Flow <- lt_sample_pairs %>% 
  select(clusterID_seq,
         surveyPeriod,
         zFlow_mean,
         RI_dammed,
         RI_undammed) %>% 
  pivot_longer(
    cols = c(RI_dammed, RI_undammed),
    names_to = "window_type",
    values_to = "RI"
  )

ggplot(ltRI_Flow,
       aes(x = zFlow_mean,
           y = RI,
           color = as.factor(window_type))) +
  scale_color_manual(values = c("RI_dammed" = "brown",
                                "RI_undammed" = "steelblue")) +
  geom_point(alpha = 0.4, size = 2) +
  geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
  labs(
    x = "z-Standardized Flow",
    y = "Riparian Inundation (RI)",
    color = "Window Type",
    title = "Riparian Inundation Across Flow",
    subtitle = "LT All Years"
  ) +
  theme_bw(base_size = 14)
```



----
----
----
----


----
----
----
----


BRAZIL
----
----

Compute **Riparian Inundation**

For survey window[i] at time[t], the proportion of riparian area that is covered in surface water

$SWA = Surface Water Area (m^2)$

$RIPA = Riparian Area (m^2)$

$$RI[i,t] = SWA[i,t] / RIPA[i] $$


```{r, echo=FALSE}
# COMPUTE RI HERE!!!
BRZ_RI <- BRZ %>% 
  mutate(RI = SWA/riparian_area_m2)
```

Visualize riparian inundation **(RI)** $SWA/RIPA$ between dammed and undammed across years.

```{r, echo=FALSE, fig.width=10}
BRZ_RI %>%
  filter(!is.na(RI)) %>%   # optional but recommended
  ggplot(aes(x = factor(year),
             y = RI,
             fill = window_type)) +
  geom_boxplot(outlier.alpha = 0.4) +
  scale_fill_manual(values = c("dammed" = "brown",
                               "undammed" = "steelblue")) +
  labs(
    x = "Year",
    y = "Riparian Inundation (RI)",
    fill = "Window type",
    title = "RI across dammed vs undammed windows",
    subtitle = "Brazil Creek (2019, 2021, 2023)"
  ) +
  theme_bw(base_size = 14)

```

```{r,echo=FALSE}
BRZ_RI %>%
  filter(year == 2019) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Brazil Ck; Summer 2019") +
     theme_bw(base_size = 15)

```

```{r,echo=FALSE}
BRZ_RI %>%
  filter(year == 2021) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Brazil Ck; Summer 2021") +
     theme_bw(base_size = 15)

```

```{r,echo=FALSE}
BRZ_RI %>%
  filter(year == 2023) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Brazil Ck; Summer 2023") +
     theme_bw(base_size = 15)

```


----

Add **relative flow data**. 

NOTE: currently this represents data from the USGS Guston Coulee gage that has been z-standardized over the current data acquisition range (June-Sept; 2019, 2021, 2023). 

Then, for each 2-week survey time period (1:8), the average relative flow $[mean(zFlow)]$ was computed, and added as a data column. IT IS ASSUMED that this flow data is representative of the RELATIVE flow pattern experienced throughout the study area during this time. Consider incorporating relative HHP simulated flow data again (?)

----

Visualize Surface Water Area across Relative Flow and Topographic Openness


```{r,echo=FALSE, fig.width=10}
BRZ_RelFlow <- BRZ_RI %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = Nhat/window_length)


BRZ_RelFlow %>% 
  arrange(SWA) %>%
  ggplot(
       aes(x = zFlow_mean, 
           y = TO, 
           color = SWA,
           shape = window_type)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "mako",begin = 0,end   = 1,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = "Topographic Openness (TO)",
    color = expression("Surface Water Area" (m^2)),
    title = "Surface Water Area Across RelativeFlow and TO space",
    subtitle = "Brazil Creek All Years"
  ) +
  theme_bw(base_size = 14)
```


----

----  

----

----



Visualize covariate distributions b/t dammed and undammed windows:

```{r,include=FALSE}
ggplot(BRZ_RelFlow, 
       aes(x = TO, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)


ggplot(BRZ_RelFlow, 
       aes(x = riparian_area_m2, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(BRZ_RelFlow, 
       aes(x = SWA, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(BRZ_RelFlow, 
       aes(x = RI, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(BRZ_RelFlow, 
       aes(x = catchment_km2, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)
```


----

Visualize covariate correlations for Willow Creek :

```{r,echo=FALSE, fig.width=10}
#2019
brz19_numCols <- BRZ_RelFlow %>% 
  filter(year == "2019") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
brz19_corr_matrix <- cor(brz19_numCols)

# Create the plot
brz19_ggcorr_plot <- ggcorrplot(brz19_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Brazil Creek 2019") +
  theme(plot.title = element_text(hjust = 0.5))

brz19_ggcorr_plot

#2021
brz21_numCols <- BRZ_RelFlow %>% 
  filter(year == "2021") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
brz21_corr_matrix <- cor(brz21_numCols)

# Create the plot
brz21_ggcorr_plot <- ggcorrplot(brz21_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Brazil Creek 2021") +
  theme(plot.title = element_text(hjust = 0.5))

brz21_ggcorr_plot


#2023
brz23_numCols <- BRZ_RelFlow %>% 
  filter(year == "2023") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
brz23_corr_matrix <- cor(brz23_numCols)

# Create the plot
brz23_ggcorr_plot <- ggcorrplot(brz23_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Brazil Creek 2023") +
  theme(plot.title = element_text(hjust = 0.5))

brz23_ggcorr_plot
```

----

----

----

Compute Differences (Dammed[i] - Undammed):
----

Compute difference in RI $(\Delta RI)$ between dammed and undammed pairs. The hypothesis is that dammed reaches will inundate more of their respective available space (m^2) than their paired undammed reaches. (+)$(\Delta RI)$ means the dammed site in a pair inundated more of its available riparian space than its undammed counterpart; (-) is vice versa.

The degree to which beaver dams influence $\Delta RI$ confounded by pairwise differences in reach characteristics (e.g., $TO$, $UDA$)

----

Proposed model:

$$\Delta RI[i] = \alpha + \beta1 * \Delta UDA[i] + \beta2 * \Delta TO + \beta3 *  Nhat[i_{\text{dammed}}]$$


Wherein, the difference in riparian inundation $(\Delta RI)$ for sample pair $[i]$ is a linear function of the pair's difference in stream network position $(\Delta UDA)$, the pair's difference in local channel confinement $(\Delta TO)$, and the pair's difference in number of dams $(\Delta Nhat)$ (NOTE: delta_Nhat is just Nhat of the dammed site in pair [i]).

Consider, instead of Nhat, using $dam/m = Nhat/windowLength (m)$ to get at dam density within the window, rather than raw dam count.


```{r,echo=FALSE}
brz_sample_pairs <- BRZ_RelFlow %>%
  select(clusterID_seq,
         year,
         t,
         surveyPeriod,
         window_type, 
         RI, 
         TO, 
         riparian_area_m2, 
         zFlow_mean,
         catchment_km2,
         Nhat,
         dam_m) %>%
  pivot_wider(
    names_from = window_type,
    values_from = c(RI, TO, riparian_area_m2, catchment_km2, Nhat,dam_m),
    names_sep = "_"
  ) %>% 
  mutate(delta_RI = RI_dammed - RI_undammed)

brz_sample_differences <- brz_sample_pairs %>%
  mutate(
    delta_TO   = TO_dammed - TO_undammed,
    delta_RIPA = riparian_area_m2_dammed - riparian_area_m2_undammed,
    delta_UDA  = catchment_km2_dammed - catchment_km2_undammed,
    delta_Nhat = Nhat_dammed,
    delta_dam_m = dam_m_dammed
  ) %>% 
  select(clusterID_seq,
         year,
         t,
         surveyPeriod,
         zFlow_mean,
         delta_RI,
         delta_TO,
         delta_RIPA,
         delta_UDA,
         delta_Nhat,
         delta_dam_m
         )

str(brz_sample_differences)
```


----

Visualize $\Delta RI$ across $\Delta Nhat$

```{r, echo=FALSE, fig.width=10}
brz_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2019') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Brazil Creek 2019"
  ) +
  theme_bw(base_size = 14)


#####

brz_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2021') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Brazil Creek 2021"
  ) +
  theme_bw(base_size = 14)


#####

brz_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2023') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Brazil Creek 2023"
  ) +
  theme_bw(base_size = 14)

```


----


----

Visualize $\Delta RI$ across **damDensity** $(dam/m)$

```{r, echo=FALSE, fig.width=10}
brz_sample_differences %>% 
  filter(year=='2019') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Brazil Creek 2019"
  ) +
  theme_bw(base_size = 14)

#####

brz_sample_differences %>% 
  filter(year=='2021') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Brazil Creek 2021"
  ) +
  theme_bw(base_size = 14)

#####

brz_sample_differences %>% 
  filter(year=='2023') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Brazil Creek 2023"
  ) +
  theme_bw(base_size = 14)

#####
```


----

Dig into where $\Delta RI >0$; these illustrate contexts where dams may be positively influencing SWA.

Look at TO and RIPA and UDA for delta_RI > 0 dam clusters

```{r,echo=FALSE, fig.width=10}
brz_posDamDiff <- brz_sample_pairs %>% 
  filter(delta_RI > 0)

brz_posDamDiff$year <- as.factor(brz_posDamDiff$year)

brz_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = TO_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 5, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = "Topographic Openness (TO)",
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and TO"),
    subtitle = "Brazil Creek All Years"
  ) +
  theme_bw(base_size = 14)



brz_sample_differences %>%
  filter(delta_RI >0) %>% 
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = delta_TO, 
           color = delta_RI,
           shape = as.factor(year))) +
  geom_point(size = 5, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Difference in Topographic Openness (  "*Delta*" TO)"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    shape = "Year",
    title = expression("Sites with Positive deltaRI Across Flow and " *Delta* "TO"),
    subtitle = "Brazil Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
brz_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = riparian_area_m2_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Riparian Area ("*m^2*")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and RIPA"),
    subtitle = "Brazil Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
brz_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = catchment_km2_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Upstream Drainage Area ("*km^2*")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and UDA"),
    subtitle = "Brazil Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
brz_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = dam_m_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Dam Density (" *DamCount[i]/streamLength(m)[i]* ")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and UDA"),
    subtitle = "Brazil Creek All Years"
  ) +
  theme_bw(base_size = 14)
```


```{r,echo=TRUE}
brzRI_Flow <- brz_sample_pairs %>% 
  select(clusterID_seq,
         surveyPeriod,
         zFlow_mean,
         RI_dammed,
         RI_undammed) %>% 
  pivot_longer(
    cols = c(RI_dammed, RI_undammed),
    names_to = "window_type",
    values_to = "RI"
  )

ggplot(brzRI_Flow,
       aes(x = zFlow_mean,
           y = RI,
           color = as.factor(window_type))) +
  scale_color_manual(values = c("RI_dammed" = "brown",
                                "RI_undammed" = "steelblue")) +
  geom_point(alpha = 0.4, size = 2) +
  geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
  labs(
    x = "z-Standardized Flow",
    y = "Riparian Inundation (RI)",
    color = "Window Type",
    title = "Riparian Inundation Across Flow",
    subtitle = "BRZ All Years"
  ) +
  theme_bw(base_size = 14)
```


----
----
----
----




----
----
----
----


LARB
----
----

Compute **Riparian Inundation**

For survey window[i] at time[t], the proportion of riparian area that is covered in surface water

$SWA = Surface Water Area (m^2)$

$RIPA = Riparian Area (m^2)$

$$RI[i,t] = SWA[i,t] / RIPA[i] $$


```{r, echo=FALSE}
# COMPUTE RI HERE!!!
LRB_RI <- LRB %>% 
  mutate(RI = SWA/riparian_area_m2)
```

Visualize riparian inundation **(RI)** $SWA/RIPA$ between dammed and undammed across years.

```{r, echo=FALSE, fig.width=10}
LRB_RI %>%
  filter(!is.na(RI)) %>%   # optional but recommended
  ggplot(aes(x = factor(year),
             y = RI,
             fill = window_type)) +
  geom_boxplot(outlier.alpha = 0.4) +
  scale_fill_manual(values = c("dammed" = "brown",
                               "undammed" = "steelblue")) +
  labs(
    x = "Year",
    y = "Riparian Inundation (RI)",
    fill = "Window type",
    title = "RI across dammed vs undammed windows",
    subtitle = "Larb Creek (2019, 2021, 2023)"
  ) +
  theme_bw(base_size = 14)

```

```{r,echo=FALSE}
LRB_RI %>%
  filter(year == 2019) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Larb Ck; Summer 2019") +
     theme_bw(base_size = 15)

```

```{r,echo=FALSE}
LRB_RI %>%
  filter(year == 2021) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Larb Ck; Summer 2021") +
     theme_bw(base_size = 15)

```

```{r,echo=FALSE}
LRB_RI %>%
  filter(year == 2023) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Larb Ck; Summer 2023") +
     theme_bw(base_size = 15)

```


----

Add **relative flow data**. 

NOTE: currently this represents data from the USGS Guston Coulee gage that has been z-standardized over the current data acquisition range (June-Sept; 2019, 2021, 2023). 

Then, for each 2-week survey time period (1:8), the average relative flow $[mean(zFlow)]$ was computed, and added as a data column. IT IS ASSUMED that this flow data is representative of the RELATIVE flow pattern experienced throughout the study area during this time. Consider incorporating relative HHP simulated flow data again (?)

----

Visualize Surface Water Area across Relative Flow and Topographic Openness


```{r,echo=FALSE, fig.width=10}
LRB_RelFlow <- LRB_RI %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = Nhat/window_length)


LRB_RelFlow %>% 
  arrange(SWA) %>%
  ggplot(
       aes(x = zFlow_mean, 
           y = TO, 
           color = SWA,
           shape = window_type)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "mako",begin = 0,end   = 1,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = "Topographic Openness (TO)",
    color = expression("Surface Water Area" (m^2)),
    title = "Surface Water Area Across RelativeFlow and TO space",
    subtitle = "Larb Creek All Years"
  ) +
  theme_bw(base_size = 14)
```


----

----  

----

----


Visualize covariate distributions b/t dammed and undammed windows:

```{r,include=FALSE}
ggplot(LRB_RelFlow, 
       aes(x = TO, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)


ggplot(LRB_RelFlow, 
       aes(x = riparian_area_m2, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(LRB_RelFlow, 
       aes(x = SWA, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(LRB_RelFlow, 
       aes(x = RI, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(LRB_RelFlow, 
       aes(x = catchment_km2, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)
```


----

Visualize covariate correlations for Willow Creek :

```{r,echo=FALSE, fig.width=10}
#2019
lrb19_numCols <- LRB_RelFlow %>% 
  filter(year == "2019") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
lrb19_corr_matrix <- cor(lrb19_numCols)

# Create the plot
lrb19_ggcorr_plot <- ggcorrplot(lrb19_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Larb Creek 2019") +
  theme(plot.title = element_text(hjust = 0.5))

lrb19_ggcorr_plot

#2021
lrb21_numCols <- LRB_RelFlow %>% 
  filter(year == "2021") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
lrb21_corr_matrix <- cor(lrb21_numCols)

# Create the plot
lrb21_ggcorr_plot <- ggcorrplot(lrb21_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Larb Creek 2021") +
  theme(plot.title = element_text(hjust = 0.5))

lrb21_ggcorr_plot


#2023
lrb23_numCols <- LRB_RelFlow %>% 
  filter(year == "2023") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
lrb23_corr_matrix <- cor(lrb23_numCols)

# Create the plot
lrb23_ggcorr_plot <- ggcorrplot(lrb23_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Larb Creek 2023") +
  theme(plot.title = element_text(hjust = 0.5))

lrb23_ggcorr_plot
```

----

----

----

Compute Differences (Dammed[i] - Undammed):
----

Compute difference in RI $(\Delta RI)$ between dammed and undammed pairs. The hypothesis is that dammed reaches will inundate more of their respective available space (m^2) than their paired undammed reaches. (+)$(\Delta RI)$ means the dammed site in a pair inundated more of its available riparian space than its undammed counterpart; (-) is vice versa.

The degree to which beaver dams influence $\Delta RI$ confounded by pairwise differences in reach characteristics (e.g., $TO$, $UDA$)

----

Proposed model:

$$\Delta RI[i] = \alpha + \beta1 * \Delta UDA[i] + \beta2 * \Delta TO + \beta3 *  Nhat[i_{\text{dammed}}]$$


Wherein, the difference in riparian inundation $(\Delta RI)$ for sample pair $[i]$ is a linear function of the pair's difference in stream network position $(\Delta UDA)$, the pair's difference in local channel confinement $(\Delta TO)$, and the pair's difference in number of dams $(\Delta Nhat)$ (NOTE: delta_Nhat is just Nhat of the dammed site in pair [i]).

Consider, instead of Nhat, using $dam/m = Nhat/windowLength (m)$ to get at dam density within the window, rather than raw dam count.


```{r,echo=FALSE}
lrb_sample_pairs <- LRB_RelFlow %>%
  select(clusterID_seq,
         year,
         t,
         surveyPeriod,
         window_type, 
         RI, 
         TO, 
         riparian_area_m2, 
         zFlow_mean,
         catchment_km2,
         Nhat,
         dam_m) %>%
  pivot_wider(
    names_from = window_type,
    values_from = c(RI, TO, riparian_area_m2, catchment_km2, Nhat,dam_m),
    names_sep = "_"
  ) %>% 
  mutate(delta_RI = RI_dammed - RI_undammed)

lrb_sample_differences <- lrb_sample_pairs %>%
  mutate(
    delta_TO   = TO_dammed - TO_undammed,
    delta_RIPA = riparian_area_m2_dammed - riparian_area_m2_undammed,
    delta_UDA  = catchment_km2_dammed - catchment_km2_undammed,
    delta_Nhat = Nhat_dammed,
    delta_dam_m = dam_m_dammed
  ) %>% 
  select(clusterID_seq,
         year,
         t,
         surveyPeriod,
         zFlow_mean,
         delta_RI,
         delta_TO,
         delta_RIPA,
         delta_UDA,
         delta_Nhat,
         delta_dam_m
         )

str(lrb_sample_differences)
```


----

Visualize $\Delta RI$ across $\Delta Nhat$

```{r, echo=FALSE, fig.width=10}
lrb_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2019') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Larb Creek 2019"
  ) +
  theme_bw(base_size = 14)


#####

lrb_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2021') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Larb Creek 2021"
  ) +
  theme_bw(base_size = 14)


#####

lrb_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2023') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Larb Creek 2023"
  ) +
  theme_bw(base_size = 14)

```


----


----

Visualize $\Delta RI$ across **damDensity** $(dam/m)$

```{r, echo=FALSE, fig.width=10}
lrb_sample_differences %>% 
  filter(year=='2019') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Larb Creek 2019"
  ) +
  theme_bw(base_size = 14)

#####

lrb_sample_differences %>% 
  filter(year=='2021') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Larb Creek 2021"
  ) +
  theme_bw(base_size = 14)

#####

lrb_sample_differences %>% 
  filter(year=='2023') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Larb Creek 2023"
  ) +
  theme_bw(base_size = 14)

#####
```


----

Dig into where $\Delta RI >0$; these illustrate contexts where dams may be positively influencing SWA.

Look at TO and RIPA and UDA for delta_RI > 0 dam clusters

```{r,echo=FALSE, fig.width=10}
lrb_posDamDiff <- lrb_sample_pairs %>% 
  filter(delta_RI > 0)

lrb_posDamDiff$year <- as.factor(lrb_posDamDiff$year)

lrb_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = TO_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = "Topographic Openness (TO)",
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and TO"),
    subtitle = "Larb Creek All Years"
  ) +
  theme_bw(base_size = 14)


lrb_sample_differences %>%
  filter(delta_RI >0) %>% 
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = delta_TO, 
           color = delta_RI,
           shape = as.factor(year))) +
  geom_point(size = 5, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Difference in Topographic Openness (  "*Delta*" TO)"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    shape = "Year",
    title = expression("Sites with Positive deltaRI Across Flow and " *Delta* "TO"),
    subtitle = "Larb Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
lrb_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = riparian_area_m2_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Riparian Area ("*m^2*")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and RIPA"),
    subtitle = "Larb Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
lrb_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = catchment_km2_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Upstream Drainage Area ("*km^2*")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and UDA"),
    subtitle = "Larb Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
lrb_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = dam_m_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Dam Density (" *DamCount[i]/streamLength(m)[i]* ")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and UDA"),
    subtitle = "Larb Creek All Years"
  ) +
  theme_bw(base_size = 14)
```


```{r,echo=TRUE}
lrbRI_Flow <- lrb_sample_pairs %>% 
  select(clusterID_seq,
         surveyPeriod,
         zFlow_mean,
         RI_dammed,
         RI_undammed) %>% 
  pivot_longer(
    cols = c(RI_dammed, RI_undammed),
    names_to = "window_type",
    values_to = "RI"
  )

ggplot(lrbRI_Flow,
       aes(x = zFlow_mean,
           y = RI,
           color = as.factor(window_type))) +
  scale_color_manual(values = c("RI_dammed" = "brown",
                                "RI_undammed" = "steelblue")) +
  geom_point(alpha = 0.4, size = 2) +
  geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
  labs(
    x = "z-Standardized Flow",
    y = "Riparian Inundation (RI)",
    color = "Window Type",
    title = "Riparian Inundation Across Flow",
    subtitle = "LRB All Years"
  ) +
  theme_bw(base_size = 14)
```

----
----
----
----




----
----
----
----


LOWER BEAVER
----
----

Compute **Riparian Inundation**

For survey window[i] at time[t], the proportion of riparian area that is covered in surface water

$SWA = Surface Water Area (m^2)$

$RIPA = Riparian Area (m^2)$

$$RI[i,t] = SWA[i,t] / RIPA[i] $$


```{r, echo=FALSE}
# COMPUTE RI HERE!!!
LBV_RI <- LBV %>% 
  mutate(RI = SWA/riparian_area_m2)
```

Visualize riparian inundation **(RI)** $SWA/RIPA$ between dammed and undammed across years.

```{r, echo=FALSE, fig.width=10}
LBV_RI %>%
  filter(!is.na(RI)) %>%   # optional but recommended
  ggplot(aes(x = factor(year),
             y = RI,
             fill = window_type)) +
  geom_boxplot(outlier.alpha = 0.4) +
  scale_fill_manual(values = c("dammed" = "brown",
                               "undammed" = "steelblue")) +
  labs(
    x = "Year",
    y = "Riparian Inundation (RI)",
    fill = "Window type",
    title = "RI across dammed vs undammed windows",
    subtitle = "Lower Beaver Creek (2019, 2021, 2023)"
  ) +
  theme_bw(base_size = 14)

```

```{r,echo=FALSE}
LBV_RI %>%
  filter(year == 2019) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Lower Beaver Ck; Summer 2019") +
     theme_bw(base_size = 15)

```

```{r,echo=FALSE}
LBV_RI %>%
  filter(year == 2021) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Lower Beaver Ck; Summer 2021") +
     theme_bw(base_size = 15)

```

```{r,echo=FALSE}
LBV_RI %>%
  filter(year == 2023) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Lower Beaver Ck; Summer 2023") +
     theme_bw(base_size = 15)

```


----

Add **relative flow data**. 

NOTE: currently this represents data from the USGS Guston Coulee gage that has been z-standardized over the current data acquisition range (June-Sept; 2019, 2021, 2023). 

Then, for each 2-week survey time period (1:8), the average relative flow $[mean(zFlow)]$ was computed, and added as a data column. IT IS ASSUMED that this flow data is representative of the RELATIVE flow pattern experienced throughout the study area during this time. Consider incorporating relative HHP simulated flow data again (?)

----

Visualize Surface Water Area across Relative Flow and Topographic Openness


```{r,echo=FALSE, fig.width=10}
LBV_RelFlow <- LBV_RI %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = Nhat/window_length)


LBV_RelFlow %>% 
  arrange(SWA) %>%
  ggplot(
       aes(x = zFlow_mean, 
           y = TO, 
           color = SWA,
           shape = window_type)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "mako",begin = 0,end   = 1,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = "Topographic Openness (TO)",
    color = expression("Surface Water Area" (m^2)),
    title = "Surface Water Area Across RelativeFlow and TO space",
    subtitle = "Lower Beaver Creek All Years"
  ) +
  theme_bw(base_size = 14)
```


----

----  

----

----



Visualize covariate distributions b/t dammed and undammed windows:

```{r,include=FALSE}
ggplot(LBV_RelFlow, 
       aes(x = TO, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)


ggplot(LBV_RelFlow, 
       aes(x = riparian_area_m2, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(LBV_RelFlow, 
       aes(x = SWA, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(LBV_RelFlow, 
       aes(x = RI, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(LBV_RelFlow, 
       aes(x = catchment_km2, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)
```



----

Visualize covariate correlations for Willow Creek :

```{r,echo=FALSE, fig.width=10}
#2019
lbv19_numCols <- LBV_RelFlow %>% 
  filter(year == "2019") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
lbv19_corr_matrix <- cor(lbv19_numCols)

# Create the plot
lbv19_ggcorr_plot <- ggcorrplot(lbv19_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Lower Beaver Creek 2019") +
  theme(plot.title = element_text(hjust = 0.5))

lbv19_ggcorr_plot

#2021
lbv21_numCols <- LBV_RelFlow %>% 
  filter(year == "2021") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
lbv21_corr_matrix <- cor(lbv21_numCols)

# Create the plot
lbv21_ggcorr_plot <- ggcorrplot(lbv21_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Lower Beaver Creek 2021") +
  theme(plot.title = element_text(hjust = 0.5))

lbv21_ggcorr_plot


#2023
lbv23_numCols <- LBV_RelFlow %>% 
  filter(year == "2023") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
lbv23_corr_matrix <- cor(lbv23_numCols)

# Create the plot
lbv23_ggcorr_plot <- ggcorrplot(lbv23_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Lower Beaver Creek 2023") +
  theme(plot.title = element_text(hjust = 0.5))

lbv23_ggcorr_plot
```

----

----

----

Compute Differences (Dammed[i] - Undammed):
----

Compute difference in RI $(\Delta RI)$ between dammed and undammed pairs. The hypothesis is that dammed reaches will inundate more of their respective available space (m^2) than their paired undammed reaches. (+)$(\Delta RI)$ means the dammed site in a pair inundated more of its available riparian space than its undammed counterpart; (-) is vice versa.

The degree to which beaver dams influence $\Delta RI$ confounded by pairwise differences in reach characteristics (e.g., $TO$, $UDA$)

----

Proposed model:

$$\Delta RI[i] = \alpha + \beta1 * \Delta UDA[i] + \beta2 * \Delta TO + \beta3 *  Nhat[i_{\text{dammed}}]$$


Wherein, the difference in riparian inundation $(\Delta RI)$ for sample pair $[i]$ is a linear function of the pair's difference in stream network position $(\Delta UDA)$, the pair's difference in local channel confinement $(\Delta TO)$, and the pair's difference in number of dams $(\Delta Nhat)$ (NOTE: delta_Nhat is just Nhat of the dammed site in pair [i]).

Consider, instead of Nhat, using $dam/m = Nhat/windowLength (m)$ to get at dam density within the window, rather than raw dam count.


```{r,echo=FALSE}
lbv_sample_pairs <- LBV_RelFlow %>%
  select(clusterID_seq,
         year,
         t,
         surveyPeriod,
         window_type, 
         RI, 
         TO, 
         riparian_area_m2, 
         zFlow_mean,
         catchment_km2,
         Nhat,
         dam_m) %>%
  pivot_wider(
    names_from = window_type,
    values_from = c(RI, TO, riparian_area_m2, catchment_km2, Nhat,dam_m),
    names_sep = "_"
  ) %>% 
  mutate(delta_RI = RI_dammed - RI_undammed)

lbv_sample_differences <- lbv_sample_pairs %>%
  mutate(
    delta_TO   = TO_dammed - TO_undammed,
    delta_RIPA = riparian_area_m2_dammed - riparian_area_m2_undammed,
    delta_UDA  = catchment_km2_dammed - catchment_km2_undammed,
    delta_Nhat = Nhat_dammed,
    delta_dam_m = dam_m_dammed
  ) %>% 
  select(clusterID_seq,
         year,
         t,
         surveyPeriod,
         zFlow_mean,
         delta_RI,
         delta_TO,
         delta_RIPA,
         delta_UDA,
         delta_Nhat,
         delta_dam_m
         )

str(lbv_sample_differences)
```


----

Visualize $\Delta RI$ across $\Delta Nhat$

```{r, echo=FALSE, fig.width=10}
lbv_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2019') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Lower Beaver Creek 2019"
  ) +
  theme_bw(base_size = 14)


#####

lbv_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2021') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Lower Beaver Creek 2021"
  ) +
  theme_bw(base_size = 14)


#####

lbv_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2023') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Lower Beaver Creek 2023"
  ) +
  theme_bw(base_size = 14)

```


----


----

Visualize $\Delta RI$ across **damDensity** $(dam/m)$

```{r, echo=FALSE, fig.width=10}
lbv_sample_differences %>% 
  filter(year=='2019') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Lower Beaver Creek 2019"
  ) +
  theme_bw(base_size = 14)

#####

lbv_sample_differences %>% 
  filter(year=='2021') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Lower Beaver Creek 2021"
  ) +
  theme_bw(base_size = 14)

#####

lbv_sample_differences %>% 
  filter(year=='2023') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Lower Beaver Creek 2023"
  ) +
  theme_bw(base_size = 14)

#####
```


----

Dig into where $\Delta RI >0$; these illustrate contexts where dams may be positively influencing SWA.

Look at TO and RIPA and UDA for delta_RI > 0 dam clusters

```{r,echo=FALSE, fig.width=10}
lbv_posDamDiff <- lbv_sample_pairs %>% 
  filter(delta_RI > 0)

lbv_posDamDiff$year <- as.factor(lbv_posDamDiff$year)

lbv_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = TO_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = "Topographic Openness (TO)",
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and TO"),
    subtitle = "Lower Beaver Creek All Years"
  ) +
  theme_bw(base_size = 14)


lbv_sample_differences %>%
  filter(delta_RI >0) %>% 
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = delta_TO, 
           color = delta_RI,
           shape = as.factor(year))) +
  geom_point(size = 5, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Difference in Topographic Openness (  "*Delta*" TO)"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    shape = "Year",
    title = expression("Sites with Positive deltaRI Across Flow and " *Delta* "TO"),
    subtitle = "Lower Beaver Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
lbv_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = riparian_area_m2_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Riparian Area ("*m^2*")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and RIPA"),
    subtitle = "Lower Beaver Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
lbv_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = catchment_km2_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Upstream Drainage Area ("*km^2*")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and UDA"),
    subtitle = "Lower Beaver Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
lbv_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = dam_m_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Dam Density (" *DamCount[i]/streamLength(m)[i]* ")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and UDA"),
    subtitle = "Lower Beaver Creek All Years"
  ) +
  theme_bw(base_size = 14)
```


```{r,echo=TRUE}
lbvRI_Flow <- lbv_sample_pairs %>% 
  select(clusterID_seq,
         surveyPeriod,
         zFlow_mean,
         RI_dammed,
         RI_undammed) %>% 
  pivot_longer(
    cols = c(RI_dammed, RI_undammed),
    names_to = "window_type",
    values_to = "RI"
  )

ggplot(lbvRI_Flow,
       aes(x = zFlow_mean,
           y = RI,
           color = as.factor(window_type))) +
  scale_color_manual(values = c("RI_dammed" = "brown",
                                "RI_undammed" = "steelblue")) +
  geom_point(alpha = 0.4, size = 2) +
  geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
  labs(
    x = "z-Standardized Flow",
    y = "Riparian Inundation (RI)",
    color = "Window Type",
    title = "Riparian Inundation Across Flow",
    subtitle = "LBV All Years"
  ) +
  theme_bw(base_size = 14)
```



----
----
----
----




----
----
----
----


UPPER BEAVER
----
----

Compute **Riparian Inundation**

For survey window[i] at time[t], the proportion of riparian area that is covered in surface water

$SWA = Surface Water Area (m^2)$

$RIPA = Riparian Area (m^2)$

$$RI[i,t] = SWA[i,t] / RIPA[i] $$


```{r, echo=FALSE}
# COMPUTE RI HERE!!!
UBV_RI <- UBV %>% 
  mutate(RI = SWA/riparian_area_m2)
```

Visualize riparian inundation **(RI)** $SWA/RIPA$ between dammed and undammed across years.

```{r, echo=FALSE, fig.width=10}
UBV_RI %>%
  filter(!is.na(RI)) %>%   # optional but recommended
  ggplot(aes(x = factor(year),
             y = RI,
             fill = window_type)) +
  geom_boxplot(outlier.alpha = 0.4) +
  scale_fill_manual(values = c("dammed" = "brown",
                               "undammed" = "steelblue")) +
  labs(
    x = "Year",
    y = "Riparian Inundation (RI)",
    fill = "Window type",
    title = "RI across dammed vs undammed windows",
    subtitle = "Upper Beaver Creek (2019, 2021, 2023)"
  ) +
  theme_bw(base_size = 14)

```


```{r,echo=FALSE}
UBV_RI %>%
  filter(year == 2019) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Upper Beaver Ck; Summer 2019") +
     theme_bw(base_size = 15)

```

```{r,echo=FALSE}
UBV_RI %>%
  filter(year == 2021) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Upper Beaver Ck; Summer 2021") +
     theme_bw(base_size = 15)

```

```{r,echo=FALSE}
UBV_RI %>%
  filter(year == 2023) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Upper Beaver Ck; Summer 2023") +
     theme_bw(base_size = 15)

```


----

Add **relative flow data**. 

NOTE: currently this represents data from the USGS Guston Coulee gage that has been z-standardized over the current data acquisition range (June-Sept; 2019, 2021, 2023). 

Then, for each 2-week survey time period (1:8), the average relative flow $[mean(zFlow)]$ was computed, and added as a data column. IT IS ASSUMED that this flow data is representative of the RELATIVE flow pattern experienced throughout the study area during this time. Consider incorporating relative HHP simulated flow data again (?)

----

Visualize Surface Water Area across Relative Flow and Topographic Openness


```{r,echo=FALSE, fig.width=10}
UBV_RelFlow <- UBV_RI %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = Nhat/window_length)


UBV_RelFlow %>% 
  arrange(SWA) %>%
  ggplot(
       aes(x = zFlow_mean, 
           y = TO, 
           color = SWA,
           shape = window_type)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "mako",begin = 0,end   = 1,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = "Topographic Openness (TO)",
    color = expression("Surface Water Area" (m^2)),
    title = "Surface Water Area Across RelativeFlow and TO space",
    subtitle = "Upper Beaver Creek All Years"
  ) +
  theme_bw(base_size = 14)
```


----

----  

----

----



Visualize covariate distributions b/t dammed and undammed windows:

```{r,include=FALSE}
ggplot(UBV_RelFlow, 
       aes(x = TO, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)


ggplot(UBV_RelFlow, 
       aes(x = riparian_area_m2, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(UBV_RelFlow, 
       aes(x = SWA, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(UBV_RelFlow, 
       aes(x = RI, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)

ggplot(UBV_RelFlow, 
       aes(x = catchment_km2, fill = window_type)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("undammed" = "steelblue", "dammed" = "brown")) +
  theme_bw(base_size = 14)
```



----

Visualize covariate correlations for Willow Creek :

```{r,echo=FALSE, fig.width=10}
#2019
ubv19_numCols <- UBV_RelFlow %>% 
  filter(year == "2019") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
ubv19_corr_matrix <- cor(ubv19_numCols)

# Create the plot
ubv19_ggcorr_plot <- ggcorrplot(ubv19_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Upper Beaver Creek 2019") +
  theme(plot.title = element_text(hjust = 0.5))

ubv19_ggcorr_plot

#2021
ubv21_numCols <- UBV_RelFlow %>% 
  filter(year == "2021") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
ubv21_corr_matrix <- cor(ubv21_numCols)

# Create the plot
ubv21_ggcorr_plot <- ggcorrplot(ubv21_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Upper Beaver Creek 2021") +
  theme(plot.title = element_text(hjust = 0.5))

ubv21_ggcorr_plot


#2023
ubv23_numCols <- UBV_RelFlow %>% 
  filter(year == "2023") %>% 
  select(clusterID_seq,
         riparian_area_m2,
         window_length,
         SWA,
         RI,
         TO,
         Nhat,
         dam_m,
         catchment_km2) %>% 
  filter(!is.na(SWA))
# Correlation matrix
ubv23_corr_matrix <- cor(ubv23_numCols)

# Create the plot
ubv23_ggcorr_plot <- ggcorrplot(ubv23_corr_matrix, lab = TRUE, type = "upper")+
  ggtitle("Correlation matrix for Upper Beaver Creek 2023") +
  theme(plot.title = element_text(hjust = 0.5))

ubv23_ggcorr_plot
```

----

----

----

Compute Differences (Dammed[i] - Undammed):
----

Compute difference in RI $(\Delta RI)$ between dammed and undammed pairs. The hypothesis is that dammed reaches will inundate more of their respective available space (m^2) than their paired undammed reaches. (+)$(\Delta RI)$ means the dammed site in a pair inundated more of its available riparian space than its undammed counterpart; (-) is vice versa.

The degree to which beaver dams influence $\Delta RI$ confounded by pairwise differences in reach characteristics (e.g., $TO$, $UDA$)

----

Proposed model:

$$\Delta RI[i] = \alpha + \beta1 * \Delta UDA[i] + \beta2 * \Delta TO + \beta3 *  Nhat[i_{\text{dammed}}]$$


Wherein, the difference in riparian inundation $(\Delta RI)$ for sample pair $[i]$ is a linear function of the pair's difference in stream network position $(\Delta UDA)$, the pair's difference in local channel confinement $(\Delta TO)$, and the pair's difference in number of dams $(\Delta Nhat)$ (NOTE: delta_Nhat is just Nhat of the dammed site in pair [i]).

Consider, instead of Nhat, using $dam/m = Nhat/windowLength (m)$ to get at dam density within the window, rather than raw dam count.


```{r,echo=FALSE}
ubv_sample_pairs <- UBV_RelFlow %>%
  select(clusterID_seq,
         year,
         t,
         surveyPeriod,
         window_type, 
         RI, 
         TO, 
         riparian_area_m2, 
         zFlow_mean,
         catchment_km2,
         Nhat,
         dam_m) %>%
  pivot_wider(
    names_from = window_type,
    values_from = c(RI, TO, riparian_area_m2, catchment_km2, Nhat,dam_m),
    names_sep = "_"
  ) %>% 
  mutate(delta_RI = RI_dammed - RI_undammed)

ubv_sample_differences <- ubv_sample_pairs %>%
  mutate(
    delta_TO   = TO_dammed - TO_undammed,
    delta_RIPA = riparian_area_m2_dammed - riparian_area_m2_undammed,
    delta_UDA  = catchment_km2_dammed - catchment_km2_undammed,
    delta_Nhat = Nhat_dammed,
    delta_dam_m = dam_m_dammed
  ) %>% 
  select(clusterID_seq,
         year,
         t,
         surveyPeriod,
         zFlow_mean,
         delta_RI,
         delta_TO,
         delta_RIPA,
         delta_UDA,
         delta_Nhat,
         delta_dam_m
         )

str(ubv_sample_differences)
```


----

Visualize $\Delta RI$ across $\Delta Nhat$

```{r, echo=FALSE, fig.width=10}
ubv_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2019') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Upper Beaver Creek 2019"
  ) +
  theme_bw(base_size = 14)


#####

ubv_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2021') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Upper Beaver Creek 2021"
  ) +
  theme_bw(base_size = 14)


#####

ubv_sample_differences %>%
  arrange(delta_RI) %>% 
  filter(year=='2023') %>% 
  ggplot(
    aes(
      x = factor(round(delta_Nhat)),   # <-- THIS is the key
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_violin() +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("(" * Delta * " Dam Count)"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Difference in Dam Count",
    subtitle = "Upper Beaver Creek 2023"
  ) +
  theme_bw(base_size = 14)

```


----


----

Visualize $\Delta RI$ across **damDensity** $(dam/m)$

```{r, echo=FALSE, fig.width=10}
ubv_sample_differences %>% 
  filter(year=='2019') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Upper Beaver Creek 2019"
  ) +
  theme_bw(base_size = 14)

#####

ubv_sample_differences %>% 
  filter(year=='2021') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Upper Beaver Creek 2021"
  ) +
  theme_bw(base_size = 14)

#####

ubv_sample_differences %>% 
  filter(year=='2023') %>% 
  ggplot(
    aes(
      x = delta_dam_m,
      y = delta_RI,
      colour = zFlow_mean
    )
  ) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0,
             color = "black",
             linewidth = 1.2) +
  labs(
    x = expression("Dam Density (" * DamCount[i]/StreamLength[i] * ")"),
    y = expression("(" * Delta * " RI)"),
    title = "Difference in Riparian Inundation by Dam Density (Dam/Stream_m)",
    subtitle = "Upper Beaver Creek 2023"
  ) +
  theme_bw(base_size = 14)

#####
```


----

Dig into where $\Delta RI >0$; these illustrate contexts where dams may be positively influencing SWA.

Look at TO and RIPA and UDA for delta_RI > 0 dam clusters

```{r,echo=FALSE, fig.width=10}
ubv_posDamDiff <- ubv_sample_pairs %>% 
  filter(delta_RI > 0)

ubv_posDamDiff$year <- as.factor(ubv_posDamDiff$year)

ubv_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = TO_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = "Topographic Openness (TO)",
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and TO"),
    subtitle = "Upper Beaver Creek All Years"
  ) +
  theme_bw(base_size = 14)



ubv_sample_differences %>%
  filter(delta_RI >0) %>% 
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = delta_TO, 
           color = delta_RI,
           shape = as.factor(year))) +
  geom_point(size = 5, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Difference in Topographic Openness (  "*Delta*" TO)"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    shape = "Year",
    title = expression("Sites with Positive deltaRI Across Flow and " *Delta* "TO"),
    subtitle = "Upper Beaver Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
ubv_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = riparian_area_m2_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Riparian Area ("*m^2*")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and RIPA"),
    subtitle = "Upper Beaver Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
ubv_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = catchment_km2_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Upstream Drainage Area ("*km^2*")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and UDA"),
    subtitle = "Upper Beaver Creek All Years"
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE, fig.width=10}
ubv_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = dam_m_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Dam Density (" *DamCount[i]/streamLength(m)[i]* ")"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and UDA"),
    subtitle = "Upper Beaver Creek All Years"
  ) +
  theme_bw(base_size = 14)
```


```{r,echo=FALSE}
ubvRI_Flow <- ubv_sample_pairs %>% 
  select(clusterID_seq,
         surveyPeriod,
         zFlow_mean,
         RI_dammed,
         RI_undammed) %>% 
  pivot_longer(
    cols = c(RI_dammed, RI_undammed),
    names_to = "window_type",
    values_to = "RI"
  )

ggplot(ubvRI_Flow,
       aes(x = zFlow_mean,
           y = RI,
           color = as.factor(window_type))) +
  scale_color_manual(values = c("RI_dammed" = "brown",
                                "RI_undammed" = "steelblue")) +
  geom_point(alpha = 0.4, size = 2) +
  geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
  labs(
    x = "z-Standardized Flow",
    y = "Riparian Inundation (RI)",
    color = "Window Type",
    title = "Riparian Inundation Across Flow",
    subtitle = "UBV All Years"
  ) +
  theme_bw(base_size = 14)
```



----
----
----
----

----
----
----
----
----
----
----
----


$\Delta RI > 0$ means that the dammed sample in sample pair[i] ponded a greater proportion of its riparian zone than the un-dammed sample in sample pair[i].

----

    Contexts in which $\Delta RI > 0$: 
        
      - Brazil Creek
        - High dam density within the watershed
          - 0.04 dam/km for the whole study area
          - Brazil:
            - 2019:
            - 2021:
          
            
      - Upper Beaver Creek:
        - Relatively high dam density coupled with high channel openness/Low 
          channel confinemnet
        - avg(TO) = 60
        - Most positive effects seen at TO >= 50 (ish)
        
        
      - Lone Tree 2019;
        - Lone Tree is one of the more intermittent streams in our study area. 
          This data suggests that when water is plentiful on the landscape (2019),
          beaver dams on Lone Tree may increase the proportion of riparian inundation,
          and maintain that inundation longer than un-dammed stream reaches. In 
          more intermittent streams like Lone Tree, dam influence may have a 
          "hydrologic floor" - a threshold of relative flow, above which influence
          tends to be positive; below which, influence tends to be negligible 
        - IN LT_2019; Where RI_dammed > RI_undammed; typically TO_dammed > TO_undammed; 
          and TO_dammed > 50
        
        
The following are meant to further explore these relationships^


NOTE: I checked persistence, colonization, and occupancy and they were all about the same for each watershed. 



Contexts in which beaver dams seem to have a positive influence on surface water/riparian inundation.


----

----


BRAZIL CREEK:
----


    - High dam density maintains riparian inundation even in low water (2021)
    
    - Positive effect seen across TO, AND despite relatively low TO watershed-wide
      - Though typically postive delta_RI is associated with postive delta_TO
    

```{r,echo=FALSE}
# Import dam density estimates for target watersheds:
# From Piper et al., 202_

cp_watershed_damDensity <- read.csv("ObservedDamDensity_dams_per_km_byStream_byYear.csv") %>% 
  mutate(zDamDensity = scale(dams_per_km))

cp_watershed_col_ext <- read.csv("site_year_phi_gamma_means.csv") %>% 
  mutate(Stream = case_when(str_detect(unique_id, "BRZ") ~ "BRZ",
                            str_detect(unique_id, "WC") ~ "WC",
                            str_detect(unique_id, "LT") ~ "LT",
                            str_detect(unique_id, "LRB") ~ "LRB",
                            str_detect(unique_id, "LBV") ~ "LBV",
                            str_detect(unique_id, "UBV") ~ "UBV",
                            TRUE ~ NA)
         ) %>% 
  group_by(Stream, year) %>% 
  summarise (phi_mean = mean(phi_mean),
             gamma_mean = mean(gamma_mean))
  
  
  
cp_watershed_psi <- read.csv("site_year_psi_probs.csv") %>% 
  mutate(Stream = case_when(str_detect(site_id, "BRZ") ~ "BRZ",
                            str_detect(site_id, "WC") ~ "WC",
                            str_detect(site_id, "LT") ~ "LT",
                            str_detect(site_id, "LRB") ~ "LRB",
                            str_detect(site_id, "LBV") ~ "LBV",
                            str_detect(site_id, "UBV") ~ "UBV",
                            TRUE ~ NA) 
         ) %>% 
  group_by(Stream, year) %>% 
  summarise (psi = mean(psi))
```



```{r,echo=FALSE}
brz_damDens <- tribble(
  ~Stream, ~Year,
  "BRZ", 2019,
  "BRZ", 2021
)

brz_damDens_plot <- cp_watershed_damDensity %>%
  mutate(
    StreamYear = paste(Stream, Year),
    highlight = if_else(
      paste(Stream, Year) %in% paste(brz_damDens$Stream, brz_damDens$Year),
      "highlight", "other"
    )
  )

ggplot(brz_damDens_plot, aes(x = zDamDensity, y = fct_reorder(StreamYear, zDamDensity))) +
  geom_segment(aes(xend = 0, yend = StreamYear),
               color = "grey80", linewidth = 1) +
  geom_point(aes(color = highlight), size = 4) +
  scale_color_manual(values = c("highlight" = "violetred4", "other" = "grey70")) +
  labs(
    x = "Watershed-level Dam Density (z-scored)",
    y = "Watershed-Year",
    color = "",
    title = "Contexts Where Beaver Dams Increase Riparian Inundation",
    subtitle = "Brazil 2019 & Brazil 2021"
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")

```

```{r,echo=FALSE}
brz_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = TO_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 5, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = "Topographic Openness (TO)",
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and TO"),
    subtitle = "Brazil Creek All Years"
  ) +
  theme_bw(base_size = 14)



brz_sample_differences %>%
  filter(delta_RI >0) %>% 
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = delta_TO, 
           color = delta_RI,
           shape = as.factor(year))) +
  geom_point(size = 5, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Difference in Topographic Openness (  "*Delta*" TO)"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    shape = "Year",
    title = expression("Sites with Positive deltaRI Across Flow and " *Delta* "TO"),
    subtitle = "Brazil Creek All Years"
  ) +
  theme_bw(base_size = 14)
```


----


UPPER BEAVER CREEK:
----

    - potential difference is minimal, though tends to be positive
      - mostly in 2019; though maybe a bit in 2021
    
    - Average to above average dam density for study area, though not crazy high
    
    - High topographic openness for the study area; low channel confinement
    

```{r,echo=FALSE}

ubv_damDens <- tribble(
  ~Stream, ~Year,
  "UBV", 2019,
  "UBV", 2021
)

ubv_damDens_plot <- cp_watershed_damDensity %>%
  mutate(
    StreamYear = paste(Stream, Year),
    highlight = if_else(
      paste(Stream, Year) %in% paste(ubv_damDens$Stream, ubv_damDens$Year),
      "highlight", "other"
    )
  )

ggplot(ubv_damDens_plot, aes(x = zDamDensity, y = fct_reorder(StreamYear, zDamDensity))) +
  geom_segment(aes(xend = 0, yend = StreamYear),
               color = "grey80", linewidth = 1) +
  geom_point(aes(color = highlight), size = 4) +
  scale_color_manual(values = c("highlight" = "violetred4", "other" = "grey70")) +
  labs(
    x = "Watershed-level Dam Density (z-scored)",
    y = "Watershed-Year",
    color = "",
    title = "Contexts Where Beaver Dams Increase Riparian Inundation",
    subtitle = "Upper Beaver 2019 & Upper Beaver 2021"
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")

```

```{r,echo=FALSE}
ubv_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = TO_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = "Topographic Openness (TO)",
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and TO"),
    subtitle = "Upper Beaver Creek All Years"
  ) +
  theme_bw(base_size = 14)



ubv_sample_differences %>%
  filter(delta_RI >0) %>% 
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = delta_TO, 
           color = delta_RI,
           shape = as.factor(year))) +
  geom_point(size = 5, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Difference in Topographic Openness (  "*Delta*" TO)"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    shape = "Year",
    title = expression("Sites with Positive deltaRI Across Flow and " *Delta* "TO"),
    subtitle = "Upper Beaver Creek All Years"
  ) +
  theme_bw(base_size = 14)

```



----


LONE TREE CREEK:
----

    - Potentially notable positive effect of dams when water availability is "high enough"
    
    - Note above average dam density (all years)
    
    - Note that even when dam density was highest (2021), lack of water did not support dam capacity to fill riparian space (e.g., little effect in 2021). However, with more water in the system (2019) dams are able to hold water and fill available riparian space moreso than un-dammed counterparts. 
    
    - Also, note that positive dam effect was seen where TO was relatively high, and difference in TO of sample pair was relatively high.
    
    - Consider implications for reservoir management; BEAVER DAMS NEED WATER TO WORK
    
    
```{r,echo=FALSE}
lt_damDens <- tribble(
  ~Stream, ~Year,
  "LT", 2019,
  "LT", 2021
)

lt_damDens_plot <- cp_watershed_damDensity %>%
  mutate(
    StreamYear = paste(Stream, Year),
    highlight = if_else(
      paste(Stream, Year) %in% paste(lt_damDens$Stream, lt_damDens$Year),
      "highlight", "other"
    )
  )

ggplot(lt_damDens_plot, aes(x = zDamDensity, y = fct_reorder(StreamYear, zDamDensity))) +
  geom_segment(aes(xend = 0, yend = StreamYear),
               color = "grey80", linewidth = 1) +
  geom_point(aes(color = highlight), size = 4) +
  scale_color_manual(values = c("highlight" = "violetred4", "other" = "grey70")) +
  labs(
    x = "Watershed-level Dam Density (z-scored)",
    y = "Watershed-Year",
    color = "",
    title = "Contexts Where Beaver Dams Increase Riparian Inundation",
    subtitle = "Lone Tree 2019, 2021"
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")

```

```{r,echo=FALSE}
lt_posDamDiff %>%
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = TO_dammed, 
           color = delta_RI,
           shape = year)) +
  geom_point(size = 5, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = "Topographic Openness (TO)",
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    title = expression("Sites with Positive deltaRI Across Flow and TO"),
    subtitle = "Lone Tree Creek All Years"
  ) +
  theme_bw(base_size = 14)



lt_sample_differences %>%
  filter(delta_RI >0) %>% 
  arrange(delta_RI) %>% 
  ggplot(
       aes(x = zFlow_mean, 
           y = delta_TO, 
           color = delta_RI,
           shape = as.factor(year))) +
  geom_point(size = 5, alpha = 0.5) +
  scale_color_viridis_c(option = "inferno",begin = 0,end   = 0.875,direction = -1) +
  labs(
    x = "z-Standardized Flow",
    y = expression("Difference in Topographic Openness (  "*Delta*" TO)"),
    color = expression("("*Delta*" RI) (RI_dammed - RI_undammed)"),
    shape = "Year",
    title = expression("Sites with Positive deltaRI Across Flow and " *Delta* "TO"),
    subtitle = "Lone Tree Creek All Years"
  ) +
  theme_bw(base_size = 14)

```

```{r,echo=FALSE}
LT_RI %>%
  filter(year == 2019) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Lone Tree; Summer 2019") +
     theme_bw(base_size = 15)

```

```{r,echo=FALSE}
LT_RI %>%
  filter(year == 2021) %>% 
     ggplot(aes(x = t,
                y = RI,
                color = window_type)) +
     geom_point(alpha = 0.4, size = 2) +
     geom_smooth(method = "loess",
                 se = TRUE,
                 linewidth = 1.2) +
     scale_color_manual(values = c("dammed" = "brown",
                                   "undammed" = "steelblue")) +
     scale_x_continuous(breaks = 1:8,
                        labels = paste0("t", 1:8)) +
     labs(x = "Time Period (t)",
          y = "Riparian Inundation (SWA/RIPA)",
          color = "Window Type",
          title = "Riparian proportion inundated over time",
          subtitle = "Lone Tree; Summer 2021") +
     theme_bw(base_size = 15)

```



Next week you will show us...
Major questions for SFS talk

Q1: How well can we map surface water in prairie streams?

Q2a: To what extent do beaver dams influence riparian inundation?
  - Bayesian GLM
  
Q2b: Under what conditions do beaver dams increase the extent of riparian inundation?
  - Geomorphic (TO; confinement)
  - Hydrologic (relative flow from Guston Coulee gage)
  - Riparian area (m^2)
  - Watershed-scale dam density
  

Final results to present at SFS 
Take homes 
Please work on slides for the SFS talk (which will be the theme for writing chapter 1)... 




----
----
----
----
----
----
----
----


Compare Hierarchical GLMs
----

Proposed Models:
----


Both: 

$$  SWA \sim N(\mu, \sigma);$$

M1:

$$\mu = alpha + \eta_{stream} + \eta_{survey} + (\beta_1 + \beta_{stream} + \beta_{year}) * DamPresence$$


    This models the effect of dam presence on surface water area ("B1"), with:          
      - Random intercepts for:
        - Stream/watershed to account for unmeasureable differences in watersheds
        such as human development intensity (e.g., reservoirs, check dams), and 
        presence of springs, and so on.
        
        - Unique survey period (Year.t[1:8]) to account for unmeasureable seasonal hydrologic 
        change
      
      - Random slopes for:
        - Stream/watershed to account for dam*watershed specific effects
        - Year to account for massive hydrologic differences among sample years



M2:

$$\mu = alpha + \eta_{stream} + \eta_{survey} + (\beta_1 + \beta_{stream} + \beta_{year}) * DamPresence + \beta_2*TO + \beta_3*RiparianArea + \beta_4 * ContributingDrainageArea$$


    This models the effect of dam presence on surface water area ("B1"), with:
    
      - All of the same things as Model 1 and:
      
      - Effect of topographic openness on SWA
      - Effect of riparian area on SWA
      - Effect of contributing drainage area on SWA
      


----

Example of Input Data Structure:
---- 

```{r,echo=FALSE}
# Organize data for the proposed HGLM

# Add dam Density per watershed
ws_damDensity <- rename(cp_watershed_damDensity,
                        watershed = Stream) %>% 
  select(watershed,
         Year,
         dams_per_km,
         zDamDensity) %>% 
  group_by(watershed,
           Year)

colnames(ws_damDensity) <- c("watershed",
                             "year",
                             "ws_damDensity",
                             "z_ws_damDensity")

# Create GLM data frame:

aoi_GLM_Data <- WC_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         RI,
         dam_m) %>% 
  bind_rows(LT_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         RI,
         dam_m)) %>% 
  bind_rows(BRZ_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         RI,
         dam_m)) %>% 
  bind_rows(LRB_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         RI,
         dam_m)) %>% 
  bind_rows(LBV_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         RI,
         dam_m)) %>% 
  bind_rows(UBV_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         RI,
         dam_m))


aoi_GLM_Data$clusterID_unique <- interaction(
  aoi_GLM_Data$year,
  aoi_GLM_Data$watershed,
  aoi_GLM_Data$clusterID_seq,
  drop = TRUE
)

aoi_GLM_Data$clusterID_num <- as.numeric(aoi_GLM_Data$clusterID_unique)

#aoi_GLM_Data$eventID <- interaction(
#  aoi_GLM_Data$year,
#  aoi_GLM_Data$surveyPeriod,
#  drop = TRUE
#)

#aoi_GLM_Data$eventID_num <- as.numeric(aoi_GLM_Data$eventID)

aoi_GLM_Data$stream <- as.numeric(as.factor(aoi_GLM_Data$watershed))

aoi_GLM_Data$sampleEvent <- as.numeric(aoi_GLM_Data$surveyPeriod)

aoi_GLM_Data$year_id <- as.numeric(as.factor(aoi_GLM_Data$year))

aoi_GLM_Data <- aoi_GLM_Data %>% 
  mutate(damPresence = case_when(Nhat > 0 ~ 1,
                                 Nhat == 0 ~ 0))

str(aoi_GLM_Data)



aoi_GLM_Data_mod <- aoi_GLM_Data %>%
  filter(!is.na(SWA))   # or !is.na(RI) if that’s the response






# Build explicit lookups from the data actually used in the model
stream_lookup <- aoi_GLM_Data_mod %>%
  distinct(stream, watershed) %>%
  arrange(stream)

year_lookup <- aoi_GLM_Data_mod %>%
  distinct(year_id, year) %>%
  arrange(year_id)

uda_lookup <- aoi_GLM_Data_mod %>%
  group_by(watershed) %>%
  summarise(max_catchment = max(catchment_km2, na.rm = TRUE)) %>%
  arrange(max_catchment)



# Start from the full modeled data and remove early summer time periods
# This analysis will look exclusively at late season mesic resource availability
# 08/01-09/30
# t5:t8
aoi_GLM_Data_mod_3_4 <- aoi_GLM_Data_mod %>% 
  filter(!(sampleEvent %in% 1:4) &      # drop 2019 t1-t4
           !(sampleEvent %in% 9:12) &   # drop 2021 t1-t4
           !(sampleEvent %in% 17:20))   # drop 2023 t1–t4

event_lookup <- data.frame(
  old = sort(unique(aoi_GLM_Data_mod_3_4$sampleEvent)),
  new = 1:length(unique(aoi_GLM_Data_mod_3_4$sampleEvent))
)

aoi_GLM_Data_mod_3_4 <- aoi_GLM_Data_mod_3_4 %>%
  left_join(event_lookup, by = c("sampleEvent" = "old")) %>%
  mutate(sampleEvent_new = new) %>%
  select(-new)

event_year_3_4 <- tapply(aoi_GLM_Data_mod_3_4$year_id,
                         aoi_GLM_Data_mod_3_4$sampleEvent_new,
                         unique)
event_year_3_4 <- as.numeric(event_year_3_4)

```

----

Implement HGLMMs:
----


M1: Dam Presence Only
----


```{r, echo=FALSE}
library(jagsUI)

Sys.time()

sink("m1_HGLM_SWA.jags")

cat("
model {

  for (i in 1:N) {
    SWA[i] ~ dnorm(mu[i], tau)

  mu[i] <- alpha
             + (beta1 + b_stream[stream[i]] + b_year[year_id[i]]) * damPresence[i]
             + eta_stream[stream[i]]
             + eta_event[sampleEvent[i]]
             + eta_cluster[clusterID[i]]
  }

  # Fixed effects
  alpha ~ dnorm(0, 0.001)
  beta1 ~ dnorm(0, 0.001)

  # Random Effects
  # Stream/Watershed
  for (j in 1:n_stream) {
    eta_stream[j] ~ dnorm(0, tau_stream)     # random intercept for stream
    b_stream[j]   ~ dnorm(0, tau_b_stream)   # random slope for stream
  }
  
  # Year
  for(yr in 1:n_year) {
    b_year[yr] ~ dnorm(0, tau_b_year)        # random slope for year
  }
  
  # Sampling Event
  for (k in 1:n_event) {
    eta_event[k] ~ dnorm(0, tau_event)       # random intercept for year/survey 'sampling event'
  }
  
  # Cluster ID
  for (c in 1:n_cluster) {
    eta_cluster[c] ~ dnorm(0, tau_cluster)
  }

  # Hyperpriors
  sigma ~ dnorm(0, 1 / pow(sd_SWA, 2)) T(0,)   # half-normal with sd ≈ sd_SWA
  tau <- pow(sigma, -2)

  sigma_stream ~ dunif(0, 10)
  tau_stream <- pow(sigma_stream, -2)

  sigma_b_stream ~ dunif(0, 10)
  tau_b_stream <- pow(sigma_b_stream, -2)
  
  sigma_b_year ~ dunif(0,10)
  tau_b_year <- pow(sigma_b_year, -2)

  sigma_event ~ dunif(0, 10)
  tau_event <- pow(sigma_event, -2)
  
  sigma_cluster ~ dunif(0,10)
  tau_cluster <- pow(sigma_cluster, -2)
}
", fill = TRUE)

sink()


m1_data_list <- list(
  N = nrow(aoi_GLM_Data_mod),
  SWA = aoi_GLM_Data_mod$SWA,
  damPresence = aoi_GLM_Data_mod$damPresence,
  stream = aoi_GLM_Data_mod$stream,
  year_id = aoi_GLM_Data_mod$year_id,
  sampleEvent = aoi_GLM_Data_mod$sampleEvent,
  clusterID = aoi_GLM_Data_mod$clusterID_num,
  n_stream = length(unique(aoi_GLM_Data_mod$stream)),
  n_event = length(unique(aoi_GLM_Data_mod$sampleEvent)),
  n_year = length(unique(aoi_GLM_Data_mod$year_id)),
  n_cluster = length(unique(aoi_GLM_Data_mod$clusterID_num)),
  sd_SWA = sd(aoi_GLM_Data_mod$SWA, na.rm = T)
)

inits <- function(){list(
    alpha = rnorm(1, 0, 1),
    beta1 = rnorm(1, 0, 1)
    )}

parameters_m1 <- c(
  "alpha",
  "beta1",
  "b_stream",
  "b_year",
  "sigma",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma_stream",
  "sigma_b_stream",
  "sigma_b_year",
  "sigma_event",
  "sigma_cluster"
)

nchains <- 3
nthin <- 10
niter <- 20000
nburn <- 5000

m1_HGLM_SWA <- jags(data = m1_data_list,
                   inits = inits,
                   parameters.to.save = parameters_m1,
                   model.file = "m1_HGLM_SWA.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

print(m1_HGLM_SWA) # model with local and watershed dam density as fixed effects

```
difference in sys.time = elapsed time for model to run

```{r, echo=FALSE}
# Extract posteriors for parameters of interest

#library(dplyr)
#library(tidyr)
#library(ggplot2)
#library(purrr)

# Extract posterior draws
M1post <- m1_HGLM_SWA$sims.list

# Select only scalar parameters (convert random effects matrices)
# Scalars
M1df_scalar <- tibble(
  parameter = c("alpha", "beta1"),
  mean = c(mean(M1post$alpha), 
           mean(M1post$beta1)),
  lower = c(quantile(M1post$alpha, 0.025), 
            quantile(M1post$beta1, 0.025)),
  upper = c(quantile(M1post$alpha, 0.975), 
            quantile(M1post$beta1, 0.975))
)

# Stream-level slopes
M1df_stream <- map_df(1:ncol(M1post$b_stream), function(j){
  tibble(
    parameter = paste0("b_stream[", j, "]"),
    mean = mean(M1post$b_stream[, j]),
    lower = quantile(M1post$b_stream[, j], 0.025),
    upper = quantile(M1post$b_stream[, j], 0.975)
  )
})

# Year-level slopes
M1df_year <- map_df(1:ncol(M1post$b_year), function(y){
  tibble(
    parameter = paste0("b_year[", y, "]"),
    mean = mean(M1post$b_year[, y]),
    lower = quantile(M1post$b_year[, y], 0.025),
    upper = quantile(M1post$b_year[, y], 0.975)
  )
})

M1summary_df <- bind_rows(M1df_scalar, M1df_stream, M1df_year)

# Give parameters meaningful names
pretty_names <- c(
  alpha           = "Intercept",
  beta1           = "Dam Presence",
  "b_stream[1]"   = "Brazil Ck",
  "b_stream[2]"   = "Lower Beaver Ck",
  "b_stream[3]"   = "Larb Ck",
  "b_stream[4]"   = "Lone Tree Ck",
  "b_stream[5]"   = "Upper Beaver Ck",
  "b_stream[6]"   = "Willow Ck",
  "b_year[1]"     = "2019",
  "b_year[2]"     = "2021",
  "b_year[3]"     = "2023"
)


# Plot
ggplot(M1summary_df %>% filter(parameter == "beta1"), 
       aes(x = parameter, 
           y = mean)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower, 
                    ymax = upper), width = 0.1) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  coord_flip() +
  scale_x_discrete(labels = pretty_names) +
  scale_y_continuous(limits = c(-100, 800)) +
  theme_minimal(base_size = 14) +
  labs(
    title = expression("Model 1: " * SWA ~ Dam_Presence ),
    subtitle = "Posterior Means and 95% Credible Intervals",
    x = "Parameter",
    y = "Posterior Mean"
  ) +
  theme(axis.text.y = element_text(size = 12))

print(mean(m1_HGLM_SWA$sims.list$beta1))
```


^^^ Average dam effect across the study area is 283.9. This translates to:

"On average, dammed reaches pond about 284 m^2 more surface water than undammed 
reaches."


----


```{r,echo=FALSE}
# HEAT MAP IT

M1n_sims   <- length(M1post$beta1)
M1n_stream <- ncol(M1post$b_stream)
M1n_year   <- ncol(M1post$b_year)

#range(M1post$b_stream[,3] + M1post$b_year[,2] + M1post$beta1)


M1DamEffect <- array(NA, dim = c(M1n_sims, 
                                 M1n_stream, 
                                 M1n_year))

for(j in 1:M1n_stream){
  for(y in 1:M1n_year){
    M1DamEffect[, j, y] <- M1post$beta1 +
                           M1post$b_stream[, j] +
                           M1post$b_year[, y]
  }
}

```


```{r,echo=FALSE}
#Summarize posteriors for each 

M1summary_list <- list()

for(j in 1:M1n_stream){
  for(y in 1:M1n_year){
    
    eff <- as.vector(M1DamEffect[, j, y, drop = FALSE])
    
    M1summary_list[[paste(j, y, sep = "_")]] <- tibble(
      stream = j,
      year = y,
      mean = mean(eff),
      lower = quantile(eff, 0.025),
      upper = quantile(eff, 0.975),
      positive = quantile(eff, 0.025) > 0
    )
  }
}

# Build explicit lookups from the data actually used in the model
stream_lookup <- aoi_GLM_Data_mod %>%
  distinct(stream, watershed) %>%
  arrange(stream)

year_lookup <- aoi_GLM_Data_mod %>%
  distinct(year_id, year) %>%
  arrange(year_id)

M1effect_df <- bind_rows(M1summary_list) %>%
  left_join(stream_lookup, by = c("stream")) %>%
  left_join(year_lookup,  by = c("year" = "year_id")) %>%
  rename(year_id = year)

# order watersheds by factor levels of ascending max CDA
M1effect_df <- M1effect_df %>%
  mutate(watershed = factor(watershed, levels = uda_lookup$watershed))



# Plot the heatmap

ggplot(M1effect_df, aes(x = as.factor(year.y), 
                        y = watershed, 
                        fill = mean)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(
    low = "goldenrod",
    mid = "white",
    high = "navy",
    midpoint = 0,
    name = "Dam Effect"
  ) +
  labs(
    x = "Year",
    y = "Watershed",
    title = "M1: Magnitude of Beaver Dam Effect on Surface Water Area"
  ) +
  theme_minimal(base_size = 14)
```


^^^^ With SUBSTANTIAL variation among watersheds and years

----

----

----

M2: Dam Presence + Context
----


```{r, echo=FALSE}
library(jagsUI)

Sys.time()

sink("m2_HGLM_SWA.jags")

cat("
model {

  for (i in 1:N) {
    SWA[i] ~ dnorm(mu[i], tau)

    mu[i] <- alpha
             + (beta1 + b_stream[stream[i]] + b_year[year_id[i]]) * damPresence[i]
             + beta2 * TO[i] 
             + beta3 * RIPA[i]
             + beta4 * UDA[i]
             + eta_stream[stream[i]]
             + eta_event[sampleEvent[i]]
             + eta_cluster[clusterID[i]]
  }

  # Fixed effects
  alpha ~ dnorm(0, 0.001)
  beta1 ~ dnorm(0, 0.001)
  beta2 ~ dnorm(0, 0.001)
  beta3 ~ dnorm(0, 0.001)
  beta4 ~ dnorm(0, 0.001)

  # Random Effects
  # Stream/Watershed
  for (j in 1:n_stream) {
    eta_stream[j] ~ dnorm(0, tau_stream)     # random intercept for stream
    b_stream[j]   ~ dnorm(0, tau_b_stream)   # random slope for stream
  }
  
  # Year
  for(yr in 1:n_year) {
    b_year[yr] ~ dnorm(0, tau_b_year)        # random slope for year
  }
  
  # Sampling Event
  for (k in 1:n_event) {
    eta_event[k] ~ dnorm(0, tau_event)       # random intercept for year/survey 'sampling event'
  }
  
  # Cluster ID
  for (c in 1:n_cluster) {
    eta_cluster[c] ~ dnorm(0, tau_cluster)
  }

  # Hyperpriors
  sigma ~ dnorm(0, 1 / pow(sd_SWA, 2)) T(0,)   # half-normal with sd ≈ sd_SWA
  tau <- pow(sigma, -2)

  sigma_stream ~ dunif(0, 10)
  tau_stream <- pow(sigma_stream, -2)

  sigma_b_stream ~ dunif(0, 10)
  tau_b_stream <- pow(sigma_b_stream, -2)
  
  sigma_b_year ~ dunif(0,10)
  tau_b_year <- pow(sigma_b_year, -2)

  sigma_event ~ dunif(0, 10)
  tau_event <- pow(sigma_event, -2)
  
  sigma_cluster ~ dunif(0,10)
  tau_cluster <- pow(sigma_cluster, -2)
}
", fill = TRUE)

sink()


m2_data_list <- list(
  N = nrow(aoi_GLM_Data_mod),
  SWA = aoi_GLM_Data_mod$SWA,
  damPresence = aoi_GLM_Data_mod$damPresence,
  TO = as.numeric(scale(1-(aoi_GLM_Data_mod$TO/90))),
  RIPA = as.numeric(scale(aoi_GLM_Data_mod$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_GLM_Data_mod$catchment_km2)),
  stream = aoi_GLM_Data_mod$stream,
  year_id = aoi_GLM_Data_mod$year_id,
  sampleEvent = aoi_GLM_Data_mod$sampleEvent,
  clusterID = aoi_GLM_Data_mod$clusterID_num,
  n_stream = length(unique(aoi_GLM_Data_mod$stream)),
  n_event = length(unique(aoi_GLM_Data_mod$sampleEvent)),
  n_year = length(unique(aoi_GLM_Data_mod$year_id)),
  n_cluster = length(unique(aoi_GLM_Data_mod$clusterID_num)),
  sd_SWA = sd(aoi_GLM_Data_mod$SWA, na.rm = T)
)

parameters_m2 <- c(
  "alpha",
  "beta1",
  "b_stream",
  "b_year",
  "beta2",
  "beta3",
  "beta4",
  "sigma",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma_stream",
  "sigma_b_stream",
  "sigma_b_year",
  "sigma_event",
  "sigma_cluster"
)

m2_HGLM_SWA <- jags(data = m2_data_list,
                   inits = inits,
                   parameters.to.save = parameters_m2,
                   model.file = "m2_HGLM_SWA.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

print(m2_HGLM_SWA)

```
difference in sys.time = elapsed time for model to run


**NOTE:**
----

In M2, TO was transformed to "channel incision" by taking the inverse of proportion openness:

$$ChannelIncision = 1 - (TO/90) $$

Where $(TO/90)$ is the proportion of max openness (90 degrees = "flat"/"fully open")




```{r, echo=FALSE}
# Extract posteriors for parameters of interest

#library(dplyr)
#library(tidyr)
#library(ggplot2)
#library(purrr)

# Extract posterior draws
M2post <- m2_HGLM_SWA$sims.list

# Select only scalar parameters (convert random effects matrices)
# Scalars
M2df_scalar <- tibble(
  parameter = c("alpha", "beta1", "beta2", "beta3", "beta4"),
  mean = c(mean(M2post$alpha), 
           mean(M2post$beta1), 
           mean(M2post$beta2), 
           mean(M2post$beta3),
           mean(M2post$beta4)),
  lower = c(quantile(M2post$alpha, 0.025), 
            quantile(M2post$beta1, 0.025),
            quantile(M2post$beta2, 0.025),
            quantile(M2post$beta3, 0.025),
            quantile(M2post$beta4, 0.025)),
  upper = c(quantile(M2post$alpha, 0.975), 
            quantile(M2post$beta1, 0.975),
            quantile(M2post$beta2, 0.975),
            quantile(M2post$beta3, 0.975),
            quantile(M2post$beta4, 0.975))
)

# Stream-level slopes
M2df_stream <- map_df(1:ncol(M2post$b_stream), function(j){
  tibble(
    parameter = paste0("b_stream[", j, "]"),
    mean = mean(M2post$b_stream[, j]),
    lower = quantile(M2post$b_stream[, j], 0.025),
    upper = quantile(M2post$b_stream[, j], 0.975)
  )
})

# Year-level slopes
M2df_year <- map_df(1:ncol(M2post$b_year), function(y){
  tibble(
    parameter = paste0("b_year[", y, "]"),
    mean = mean(M2post$b_year[, y]),
    lower = quantile(M2post$b_year[, y], 0.025),
    upper = quantile(M2post$b_year[, y], 0.975)
  )
})

M2summary_df <- bind_rows(M2df_scalar, M2df_stream, M2df_year)

# Give parameters meaningful names
pretty_names <- c(
  alpha           = "Intercept",
  beta1           = "Dam Presence",
  beta2           = "Topographic Openness",
  beta3           = "Riparian Area (m^2)",
  beta4           = "Contributing Drainage Area",
  "b_stream[1]"   = "Brazil Ck",
  "b_stream[2]"   = "Lower Beaver Ck",
  "b_stream[3]"   = "Larb Ck",
  "b_stream[4]"   = "Lone Tree Ck",
  "b_stream[5]"   = "Upper Beaver Ck",
  "b_stream[6]"   = "Willow Ck",
  "b_year[1]"     = "2019",
  "b_year[2]"     = "2021",
  "b_year[3]"     = "2023"
)


# Plot
ggplot(M2summary_df %>% filter(parameter == "beta1" |
                               parameter == "beta2" |
                               parameter == "beta3" |
                               parameter == "beta4"), 
       aes(x = parameter, 
           y = mean)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower, 
                    ymax = upper), width = 0.1) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  coord_flip() +
  scale_x_discrete(labels = pretty_names) +
  scale_y_continuous(limits = c(-100, 800)) +
  theme_minimal(base_size = 14) +
  labs(
    title = expression("Model 2: (" * SWA * " ~ " * Dam_Presence * ") Considering Geomorphic Context" ),
    subtitle = "Posterior Means and 95% Credible Intervals",
    x = "Parameter",
    y = "Posterior Mean"
  ) +
  theme(axis.text.y = element_text(size = 12))

```

M2 Mean Dam Effect:

Same interpretation as M1

```{r,echo=FALSE}
print(mean(m2_HGLM_SWA$sims.list$beta1))
```

M2 Mean TO Effect

Interpretation: SWA(m^2) change per 1 standard deviation increase in TO

```{r,echo=FALSE}
print(mean(m2_HGLM_SWA$sims.list$beta2))
```

M2 Mean RIPA Effect:

Interpretation: SWA(m^2) change per 1 standard deviation increase in RIPA


```{r,echo=FALSE}
print(mean(m2_HGLM_SWA$sims.list$beta3))
```

M2 Mean CDA Effect

Interpretation: SWA(m^2) change per 1 standard deviation increase in CDA


```{r,echo=FALSE}
print(mean(m2_HGLM_SWA$sims.list$beta4))
```



```{r,echo=FALSE}
# HEAT MAP IT

M2n_sims   <- length(M2post$beta1)
M2n_stream <- ncol(M2post$b_stream)
M2n_year   <- ncol(M2post$b_year)

#range(M2post$b_stream[,3] + M2post$b_year[,2] + M2post$beta1)


M2DamEffect <- array(NA, dim = c(M2n_sims, 
                                 M2n_stream, 
                                 M2n_year))

for(j in 1:M2n_stream){
  for(y in 1:M2n_year){
    M2DamEffect[, j, y] <- M2post$beta1 +
                           M2post$b_stream[, j] +
                           M2post$b_year[, y]
  }
}

```


```{r,echo=FALSE}
#Summarize posteriors for each 

M2summary_list <- list()

for(j in 1:M2n_stream){
  for(y in 1:M2n_year){
    
    eff <- as.vector(M2DamEffect[, j, y, drop = FALSE])
    
    M2summary_list[[paste(j, y, sep = "_")]] <- tibble(
      stream = j,
      year = y,
      mean = mean(eff),
      lower = quantile(eff, 0.025),
      upper = quantile(eff, 0.975),
      positive = quantile(eff, 0.025) > 0
    )
  }
}


M2effect_df <- bind_rows(M2summary_list) %>%
  left_join(stream_lookup, by = c("stream")) %>%
  left_join(year_lookup,  by = c("year" = "year_id")) %>%
  rename(year_id = year) %>% 
  left_join(uda_lookup, by = "watershed")

# order watersheds by factor levels of ascending max UDA
M2effect_df <- M2effect_df %>%
  mutate(watershed = factor(
    watershed,
    levels = c("WC", "LT", "BRZ", "LRB", "LBV", "UBV")
  ))


# Plot the heatmap

ggplot(M2effect_df, aes(x = as.factor(year.y), 
                        y = watershed, 
                        fill = mean)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(
    low = "grey90",
    mid = "grey100",
    high = "dodgerblue4",
    midpoint = 0,
    name = "Model 2 
Dam Effect"
  ) +
  labs(
    x = "Year",
    y = "Watershed 
Arranged By Drainage Area",
    title = "Magnitude of Dam Effect by Watershed and Year"
  ) +
  theme_minimal(base_size = 14)
```

^^^ Less dam effect overall, still lots of variation among watersheds and years

Also, streams are ordered here ^^ by contributing drainage area. Higher on the y axis = greater drainage area





Visualize proportional effect of beaver dams on riparian inundation
----

'average dam inundated ~2% of available riparian space'

```{r, echo=FALSE}
# Posterior draws of dam effect (population-level)
dam_post <- M2post$beta1   # vector of length ~4500

# Median riparian area (real units)
RIPA_med <- median(aoi_GLM_Data_mod$riparian_area_m2)

# Posterior distribution of proportional inundation
prop_post <- dam_post / RIPA_med

# Summaries
quantile(prop_post, c(0.025, 0.5, 0.975))


prop_df1 <- tibble(prop = prop_post)

ggplot(prop_df1, aes(x = prop)) +
  geom_density(fill = "skyblue", alpha = 0.4, linewidth = 1.2) +
  geom_vline(xintercept = median(prop_post), color = "blue", linewidth = 1) +
  geom_vline(xintercept = quantile(prop_post, 0.025), linetype = "dashed") +
  geom_vline(xintercept = quantile(prop_post, 0.975), linetype = "dashed") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    x = "Proportion of Riparian Area Inundated by Dams",
    y = "Posterior Density",
    title = "Posterior Distribution of Proportional Riparian Inundation",
    subtitle = expression("Based on population-level dam effect ( " * beta * 1 * " ) and median riparian area")
  ) +
  theme_minimal(base_size = 14)



```

Check proportional change from undammed to dammed
----

```{r, echo=TRUE}
# Compute predicted baseline for 'typical' reach for each stream/year

TO_z_med <- median(as.numeric(scale(aoi_GLM_Data_mod$TO)))

RIPA_z_med <- median(as.numeric(scale(aoi_GLM_Data_mod$riparian_area_m2)))

UDA_z_med <- median(as.numeric(scale(aoi_GLM_Data_mod$catchment_km2)))

post <- M2post

SWA0_global <- post$alpha +
  post$beta2 * TO_z_med +
  post$beta3 * RIPA_z_med +
  post$beta4 * UDA_z_med

SWA0_global <- pmax(SWA0_global, 0)  # enforce SWA ≥ 0


# global effect

prop_global <- post$beta1 / SWA0_global
quantile(prop_global, c(0.025, 0.5, 0.975))


prop_df <- tibble(prop = prop_global)

ggplot(prop_df, aes(x = prop)) +
  geom_density(fill = "skyblue", alpha = 0.4, linewidth = 1.2) +
  geom_vline(xintercept = median(prop_global), color = "blue", linewidth = 1) +
  geom_vline(xintercept = quantile(prop_global, 0.025), linetype = "dashed") +
  geom_vline(xintercept = quantile(prop_global, 0.975), linetype = "dashed") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    x = "Proportional Change in SWA (Dammed vs Undammed)",
    y = "Posterior Density",
    title = "Global Proportional Effect of Beaver Dams",
    subtitle = "Population-level dam effect relative to global baseline SWA"
  ) +
  theme_minimal(base_size = 14)


```





----

----

----

----

----

----

----


This next section intends to explore seasonal variation in beaver dam effect. 

Uses nested random slopes for beaver dam effect to allow for stream, year, and 
sampleEvent to cause deviation from the global mean beaver effect.


Implement HGLMMs:
----


M3: Dam Presence Only; seasonal random slope (sampleEvent)
----


```{r, echo=FALSE}
library(jagsUI)

Sys.time()

sink("m3_HGLM_SWA.jags")

cat("
model {

  for (i in 1:N) {
    SWA[i] ~ dnorm(mu[i], tau)

  mu[i] <- alpha
             + (beta1
                + b_stream_event[stream[i],sampleEvent[i]]) * damPresence[i]
             + eta_stream[stream[i]]
             + eta_event[sampleEvent[i]]
             + eta_cluster[clusterID[i]]
  }

  # Fixed effects
  alpha ~ dnorm(0, 0.001)
  beta1 ~ dnorm(0, 0.001)

  # Random Effects
  # Stream/Watershed
  for (j in 1:n_stream) {
    eta_stream[j] ~ dnorm(0, tau_stream)     # random intercept for stream
    b_stream[j]   ~ dnorm(0, tau_b_stream)   # random slope for stream
  }
  
  # Year-level dam effect
  for(yr in 1:n_year) {
    b_year[yr] ~ dnorm(0, tau_b_year)        # random slope for year
  }
  
  # Event-level dam effect nested within year
  for (k in 1:n_event) {
    b_event[k] ~ dnorm(b_year[event_year[k]],tau_b_event)       # random slope for year/survey 'sampling event'
    eta_event[k] ~ dnorm(0, tau_event)
  }
  
  # Stream/Event Interaction
  for(j in 1:n_stream){
    for(k in 1:n_event){
      b_stream_event[j,k] ~ dnorm(b_event[k] + b_stream[j], tau_b_stream_event)
    }
  }
  
  # Cluster ID
  for (c in 1:n_cluster) {
    eta_cluster[c] ~ dnorm(0, tau_cluster)
  }

  # Hyperpriors
  sigma ~ dnorm(0, 1 / pow(sd_SWA, 2)) T(0,)   # half-normal with sd ≈ sd_SWA
  tau <- pow(sigma, -2)

  sigma_stream ~ dunif(0, 10)
  tau_stream <- pow(sigma_stream, -2)

  sigma_b_stream ~ dunif(0, 10)
  tau_b_stream <- pow(sigma_b_stream, -2)
  
  sigma_b_year ~ dunif(0,10)
  tau_b_year <- pow(sigma_b_year, -2)

  sigma_b_event ~ dunif(0, 10)
  tau_b_event <- pow(sigma_b_event, -2)
  
  sigma_event ~ dunif(0,10)
  tau_event <- pow(sigma_event, -2)
  
  sigma_b_stream_event ~ dunif(0,10)
  tau_b_stream_event <- pow(sigma_b_stream_event, -2)
  
  sigma_cluster ~ dunif(0,10)
  tau_cluster <- pow(sigma_cluster, -2)
}
", fill = TRUE)

sink()


m3_data_list <- list(
  N = nrow(aoi_GLM_Data_mod_3_4),
  SWA = aoi_GLM_Data_mod_3_4$SWA,
  damPresence = aoi_GLM_Data_mod_3_4$damPresence,
  stream = aoi_GLM_Data_mod_3_4$stream,
  clusterID = aoi_GLM_Data_mod_3_4$clusterID_num,
  n_stream = length(unique(aoi_GLM_Data_mod_3_4$stream)),
  n_year = length(unique(aoi_GLM_Data_mod_3_4$year_id)),
  n_cluster = length(unique(aoi_GLM_Data_mod_3_4$clusterID_num)),
  sampleEvent = aoi_GLM_Data_mod_3_4$sampleEvent_new,
  n_event = length(unique(aoi_GLM_Data_mod_3_4$sampleEvent_new)),
  event_year = event_year_3_4,
  sd_SWA = sd(aoi_GLM_Data_mod_3_4$SWA, na.rm = T)
)

m3inits <- function(){list(
    alpha = rnorm(1, 0, 1),
    beta1 = rnorm(1, 0, 1)
    )}

parameters_m3 <- c(
  "alpha",
  "beta1",
  "b_stream",
  "b_year",
  "b_event",
  "b_stream_event",
  "sigma",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma_stream",
  "sigma_b_stream",
  "sigma_b_year",
  "sigma_b_event",
  "sigma_b_stream_event",
  "sigma_event",
  "sigma_cluster"
)

nchains <- 3
nthin <- 10
niter <- 20000
nburn <- 5000

m3_HGLM_SWA <- jags(data = m3_data_list,
                   inits = m3inits,
                   parameters.to.save = parameters_m3,
                   model.file = "m3_HGLM_SWA.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

print(m3_HGLM_SWA) # model with local and watershed dam density as fixed effects

```
difference in sys.time = elapsed time for model to run

```{r, echo=FALSE}
# Extract posteriors for parameters of interest

#library(dplyr)
#library(tidyr)
#library(ggplot2)
#library(purrr)

# Extract posterior draws
M3post <- m3_HGLM_SWA$sims.list

# Select only scalar parameters (convert random effects matrices)
# Scalars
M3df_scalar <- tibble(
  parameter = c("alpha", "beta1"),
  mean = c(mean(M3post$alpha), 
           mean(M3post$beta1)),
  lower = c(quantile(M3post$alpha, 0.025), 
            quantile(M3post$beta1, 0.025)),
  upper = c(quantile(M3post$alpha, 0.975), 
            quantile(M3post$beta1, 0.975))
)

# Stream-level slopes
M3df_stream <- map_df(1:ncol(M3post$b_stream), function(j){
  tibble(
    parameter = paste0("b_stream[", j, "]"),
    mean = mean(M3post$b_stream[, j]),
    lower = quantile(M3post$b_stream[, j], 0.025),
    upper = quantile(M3post$b_stream[, j], 0.975)
  )
})

# Season-level slopes
M3df_event <- map_df(1:ncol(M3post$b_event), function(y){
  tibble(
    parameter = paste0("b_event[", y, "]"),
    mean = mean(M3post$b_event[, y]),
    lower = quantile(M3post$b_event[, y], 0.025),
    upper = quantile(M3post$b_event[, y], 0.975)
  )
})

M3summary_df <- bind_rows(M3df_scalar, M3df_stream, M3df_event)

# Give parameters meaningful names
pretty_names <- c(
  alpha           = "Intercept",
  beta1           = "Dam Presence",
  "b_stream[1]"   = "Brazil Ck",
  "b_stream[2]"   = "Lower Beaver Ck",
  "b_stream[3]"   = "Larb Ck",
  "b_stream[4]"   = "Lone Tree Ck",
  "b_stream[5]"   = "Upper Beaver Ck",
  "b_stream[6]"   = "Willow Ck",
  "b_event[1]"     = "19_t1",
  "b_event[2]"     = "19_t2",
  "b_event[3]"     = "19_t3",
  "b_event[4]"     = "19_t4",
  "b_event[5]"     = "19_t5",
  "b_event[6]"     = "19_t6",
  "b_event[7]"     = "19_t7",
  "b_event[8]"     = "19_t8",
  "b_event[9]"     = "21_t1",
  "b_event[10]"     = "21_t2",
  "b_event[11]"     = "21_t3",
  "b_event[12]"     = "21_t4",
  "b_event[13]"     = "21_t5",
  "b_event[14]"     = "21_t6",
  "b_event[15]"     = "21_t7",
  "b_event[16]"     = "21_t8",
  "b_event[17]"     = "23_t5",
  "b_event[18]"     = "23_t6",
  "b_event[19]"     = "23_t7",
  "b_event[20]"     = "23_t8"
)


# Plot
ggplot(M3summary_df, 
       aes(x = parameter, 
           y = mean)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower, 
                    ymax = upper), width = 0.1) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  coord_flip() +
  scale_x_discrete(labels = pretty_names) +
  theme_minimal(base_size = 14) +
  labs(
    title = expression("Model 3: " * SWA ~ Dam_Presence ),
    subtitle = "Posterior Means and 95% Credible Intervals",
    x = "Parameter",
    y = "Posterior Mean"
  ) +
  theme(axis.text.y = element_text(size = 12))

print(mean(m3_HGLM_SWA$sims.list$beta1))
```


^^^ Average dam effect across the study area is 283.9. This translates to:

"On average, dammed reaches pond about 284 m^2 more surface water than undammed 
reaches."


----


```{r,echo=FALSE}
# HEAT MAP IT

M3n_sims   <- length(M3post$beta1)
M3n_stream <- dim(M3post$b_stream_event)[2]
M3n_event   <- dim(M3post$b_stream_event)[3]

#range(M3post$b_stream[,3] + M3post$b_year[,2] + M3post$beta1)


M3DamEffect <- array(NA, dim = c(M3n_sims, 
                                 M3n_stream, 
                                 M3n_event))

for(j in 1:M3n_stream){
  for(e in 1:M3n_event){
    M3DamEffect[, j, e] <- M3post$beta1 +
                           M3post$b_stream_event[,j,e]
  }
}

```


```{r,echo=FALSE}
#Summarize posteriors for each 

M3summary_list <- list()

for(j in 1:M3n_stream){
  for(e in 1:M3n_event){
    
    eff <- as.vector(M3DamEffect[, j, e, drop = FALSE])
    
    M3summary_list[[paste(j, e, sep = "_")]] <- tibble(
      stream = j,
      event = e,
      mean = mean(eff),
      lower = quantile(eff, 0.025),
      upper = quantile(eff, 0.975),
      positive = quantile(eff, 0.025) > 0
    )
  }
}

# Build explicit lookups from the data actually used in the model
stream_lookup <- aoi_GLM_Data_mod %>%
  distinct(stream, watershed) %>%
  arrange(stream)

# Event lookup created above

M3effect_df <- bind_rows(M3summary_list) %>%
  left_join(stream_lookup, by = c("stream")) %>%
  left_join(event_lookup,  by = c("event" = "new"))

# order watersheds by factor levels of ascending max CDA
M3effect_df <- M3effect_df %>%
  mutate(watershed = factor(watershed, levels = uda_lookup$watershed))



# Plot the heatmap

ggplot(M3effect_df, aes(x = as.factor(event), 
                        y = watershed, 
                        fill = mean)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(
    low = "white",
    mid = "cornflowerblue",
    high = "navy",
    midpoint = 186,
    name = "Dam Effect"
  ) +
  labs(
    x = "Sampling Event",
    y = "Watershed",
    title = "M3: Magnitude of Beaver Dam Effect on Surface Water Area"
  ) +
  theme_minimal(base_size = 14)
```


^^^^ With SUBSTANTIAL variation among watersheds and years



Compare b1 trajectory across event, by stream:

```{r, echo=FALSE}
M3line_df <- M3effect_df %>%
  arrange(watershed, event) %>%
  mutate(event = as.numeric(event)) %>%
  group_by(watershed) %>%
  do({
    df <- .

    gap_rows <- tibble(
      stream = df$stream[1],
      event = c(4.5, 8.5),   # ← add as many gaps as you want
      mean = NA,
      lower = NA,
      upper = NA,
      positive = NA,
      watershed = df$watershed[1]
    )

    bind_rows(df, gap_rows)
  }) %>%
  arrange(watershed, event)



ggplot(M3line_df, aes(x = event, y = mean, color = watershed, group = watershed)) +
  geom_line(linewidth = 1.2, na.rm = TRUE) +
  geom_point(size = 2, na.rm = TRUE) +
  scale_x_continuous(
    breaks = unique(M3effect_df$event),
    labels = unique(event_lookup$label)   # if you want "19_t1", "19_t2", etc.
  ) +
  labs(
    title = "Seasonal Trajectory of Beaver Dam Effect on SWA, by Watershed",
    subtitle = "Gaps represent breaks by Year; t1:t4 excluded for 2023",
    x = "Sampling Events (t5 : t8) \n (August : September) \n (2019, 2021, 2023)",
    y = "Dam Effect (SWA m²)",
    color = "Watershed"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

```



----

----

----

M4: Dam Presence + Context with updated random effects structure
----


```{r, echo=FALSE}
library(jagsUI)

Sys.time()

sink("m4_HGLM_SWA.jags")

cat("
    model {

  for (i in 1:N) {
    SWA[i] ~ dnorm(mu[i], tau)

  mu[i] <- alpha
             + (beta1
                + b_stream_event[stream[i],sampleEvent[i]]) * damPresence[i]
             + beta2 * TO[i] 
             + beta3 * RIPA[i]
             + beta4 * UDA[i]
             + eta_stream[stream[i]]
             + eta_event[sampleEvent[i]]
             + eta_cluster[clusterID[i]]
  }

  # Fixed effects
  alpha ~ dnorm(0, 0.001)
  beta1 ~ dnorm(0, 0.001)
  beta2 ~ dnorm(0, 0.001)
  beta3 ~ dnorm(0, 0.001)
  beta4 ~ dnorm(0, 0.001)

  # Random Effects
  # Stream/Watershed
  for (j in 1:n_stream) {
    eta_stream[j] ~ dnorm(0, tau_stream)     # random intercept for stream
    b_stream[j]   ~ dnorm(0, tau_b_stream)   # random slope for stream
  }
  
  # Year-level dam effect
  for(yr in 1:n_year) {
    b_year[yr] ~ dnorm(0, tau_b_year)        # random slope for year
  }
  
  # Event-level dam effect nested within year
  for (k in 1:n_event) {
    b_event[k] ~ dnorm(b_year[event_year[k]],tau_b_event)       # random slope for year/survey 'sampling event'
    eta_event[k] ~ dnorm(0, tau_event)
  }
  
  # Stream/Event Interaction
  for(j in 1:n_stream){
    for(k in 1:n_event){
      b_stream_event[j,k] ~ dnorm(b_event[k] + b_stream[j], tau_b_stream_event)
    }
  }
  
  # Cluster ID
  for (c in 1:n_cluster) {
    eta_cluster[c] ~ dnorm(0, tau_cluster)
  }

  # Hyperpriors
  sigma ~ dnorm(0, 1 / pow(sd_SWA, 2)) T(0,)   # half-normal with sd ≈ sd_SWA
  tau <- pow(sigma, -2)

  sigma_stream ~ dunif(0, 10)
  tau_stream <- pow(sigma_stream, -2)

  sigma_b_stream ~ dunif(0, 10)
  tau_b_stream <- pow(sigma_b_stream, -2)
  
  sigma_b_year ~ dunif(0,10)
  tau_b_year <- pow(sigma_b_year, -2)

  sigma_b_event ~ dunif(0, 10)
  tau_b_event <- pow(sigma_b_event, -2)
  
  sigma_event ~ dunif(0,10)
  tau_event <- pow(sigma_event, -2)
  
  sigma_b_stream_event ~ dunif(0,10)
  tau_b_stream_event <- pow(sigma_b_stream_event, -2)
  
  sigma_cluster ~ dunif(0,10)
  tau_cluster <- pow(sigma_cluster, -2)
}
", fill = TRUE)

sink()


m4_data_list <- list(
  N = nrow(aoi_GLM_Data_mod_3_4),
  SWA = aoi_GLM_Data_mod_3_4$SWA,
  damPresence = aoi_GLM_Data_mod_3_4$damPresence,
  TO = as.numeric(scale(1-(aoi_GLM_Data_mod_3_4$TO/90))),
  RIPA = as.numeric(scale(aoi_GLM_Data_mod_3_4$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_GLM_Data_mod_3_4$catchment_km2)),
  stream = aoi_GLM_Data_mod_3_4$stream,
  clusterID = aoi_GLM_Data_mod_3_4$clusterID_num,
  n_stream = length(unique(aoi_GLM_Data_mod_3_4$stream)),
  n_year = length(unique(aoi_GLM_Data_mod_3_4$year_id)),
  n_cluster = length(unique(aoi_GLM_Data_mod_3_4$clusterID_num)),
  sampleEvent = aoi_GLM_Data_mod_3_4$sampleEvent_new,
  n_event = length(unique(aoi_GLM_Data_mod_3_4$sampleEvent_new)),
  event_year = event_year_3_4,
  sd_SWA = sd(aoi_GLM_Data_mod_3_4$SWA, na.rm = TRUE)
)

m4inits <- function(){list(
    alpha = rnorm(1, 0, 1),
    beta1 = rnorm(1, 0, 1)
    )}

parameters_m4 <- c(
  "alpha",
  "beta1",
  "beta2",
  "beta3",
  "beta4",
  "b_stream",
  "b_year",
  "b_event",
  "b_stream_event",
  "sigma",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma_stream",
  "sigma_b_stream",
  "sigma_b_year",
  "sigma_b_event",
  "sigma_b_stream_event",
  "sigma_event",
  "sigma_cluster"
)

nchains <- 3
nthin <- 10
niter <- 20000
nburn <- 5000

m4_HGLM_SWA <- jags(data = m4_data_list,
                   inits = m4inits,
                   parameters.to.save = parameters_m4,
                   model.file = "m4_HGLM_SWA.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

print(m4_HGLM_SWA) # model with local and watershed dam density as fixed effects
```
difference in sys.time = elapsed time for model to run


**NOTE:**
----

In M4, TO was transformed to "channel incision" by taking the inverse of proportion openness:

$$ChannelIncision = 1 - (TO/90) $$

Where $(TO/90)$ is the proportion of max openness (90 degrees = "flat"/"fully open")




```{r, echo=FALSE}
# Extract posteriors for parameters of interest

#library(dplyr)
#library(tidyr)
#library(ggplot2)
#library(purrr)

# Extract posterior draws
m4post <- m4_HGLM_SWA$sims.list

# Select only scalar parameters (convert random effects matrices)
# Scalars
m4df_scalar <- tibble(
  parameter = c("alpha", "beta1", "beta2", "beta3", "beta4"),
  mean = c(mean(m4post$alpha), 
           mean(m4post$beta1), 
           mean(m4post$beta2), 
           mean(m4post$beta3),
           mean(m4post$beta4)),
  lower = c(quantile(m4post$alpha, 0.025), 
            quantile(m4post$beta1, 0.025),
            quantile(m4post$beta2, 0.025),
            quantile(m4post$beta3, 0.025),
            quantile(m4post$beta4, 0.025)),
  upper = c(quantile(m4post$alpha, 0.975), 
            quantile(m4post$beta1, 0.975),
            quantile(m4post$beta2, 0.975),
            quantile(m4post$beta3, 0.975),
            quantile(m4post$beta4, 0.975))
)


m4summary_df <- m4df_scalar

# Give parameters meaningful names
pretty_names <- c(
  alpha           = "Intercept",
  beta1           = "Dam Presence",
  beta2           = "Channel Incision",
  beta3           = "Riparian Area (m^2)",
  beta4           = "Contributing Drainage Area"
)


# Plot
ggplot(
  m4summary_df %>% 
    filter(parameter %in% c("beta1", "beta2", "beta3", "beta4")),
  aes(x = parameter, y = mean)
) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  coord_flip() +
  scale_x_discrete(labels = pretty_names) +
  theme_minimal(base_size = 14) +
  labs(
    title = expression("Model 4: (" * SWA * " ~ " * Dam_Presence * ") + Geomorphic Context"),
    subtitle = "Posterior Means and 95% Credible Intervals",
    x = "Parameter",
    y = "Posterior Mean"
  ) +
  theme(axis.text.y = element_text(size = 12))
```

m4 Mean Dam Effect:

Same interpretation as M1

```{r,echo=FALSE}
print(mean(m4_HGLM_SWA$sims.list$beta1))
```

m4 Mean TO Effect

Interpretation: SWA(m^2) change per 1 standard deviation increase in TO

```{r,echo=FALSE}
print(mean(m4_HGLM_SWA$sims.list$beta2))
```

m4 Mean RIPA Effect:

Interpretation: SWA(m^2) change per 1 standard deviation increase in RIPA


```{r,echo=FALSE}
print(mean(m4_HGLM_SWA$sims.list$beta3))
```

m4 Mean CDA Effect

Interpretation: SWA(m^2) change per 1 standard deviation increase in CDA


```{r,echo=FALSE}
print(mean(m4_HGLM_SWA$sims.list$beta4))
```



```{r,echo=FALSE}
# HEAT MAP IT

m4n_sims   <- length(m4post$beta1)
m4n_stream <- ncol(m4post$b_stream)
m4n_event   <- ncol(m4post$b_event)

#range(m4post$b_stream[,3] + m4post$b_year[,2] + m4post$beta1)


m4DamEffect <- array(NA, dim = c(m4n_sims, 
                                 m4n_stream, 
                                 m4n_event))

for(j in 1:m4n_stream){
  for(k in 1:m4n_event){
    m4DamEffect[, j, k] <-
      m4post$beta1 +
      m4post$b_stream[, j] +
      m4post$b_event[, k]
  }
}

```


```{r,echo=FALSE}
#Summarize posteriors for each 

m4summary_list <- list()

for(j in 1:m4n_stream){
  for(k in 1:m4n_event){
    
    eff <- as.vector(m4DamEffect[, j, k, drop = FALSE])
    
    m4summary_list[[paste(j, k, sep = "_")]] <- tibble(
      stream = j,
      event = k,
      mean = mean(eff),
      lower = quantile(eff, 0.025),
      upper = quantile(eff, 0.975),
      positive = quantile(eff, 0.025) > 0
    )
  }
}


m4effect_df <- bind_rows(m4summary_list) %>%
  left_join(stream_lookup, by = c("stream")) %>% 
  left_join(event_lookup, by = c("event" = "new"))


# Plot the heatmap

ggplot(m4effect_df, aes(x = as.factor(old), 
                        y = watershed, 
                        fill = mean)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(
    low = "white",
    mid = "cornflowerblue",
    high = "black",
    midpoint = 193,
    name = "Model 2 
Dam Effect"
  ) +
  labs(
    x = "Sample Event",
    y = "Watershed",
    title = "Magnitude of Dam Effect by Watershed and SampleEvent",
    subtitle = "Aug:Sep; 5:8 = 2019; 13:16 = 2021; 21:24 = 2023"
  ) +
  theme_minimal(base_size = 14)
```

^^^ Less dam effect overall, still lots of variation among watersheds and years

Also, streams are ordered here ^^ by contributing drainage area. Higher on the y axis = greater drainage area





Visualize proportional change from undammed to dammed
----

```{r, echo=TRUE}
# Compute predicted baseline for 'typical' reach for each stream/year

post <- m4post

# --- Recreate standardized covariates exactly as in m4_data_list ---

# Channel incision = 1 - (TO/90)
TO_ci <- 1 - (aoi_GLM_Data_mod_3_4$TO / 90)
TO_z  <- as.numeric(scale(TO_ci))

RIPA_z <- as.numeric(scale(aoi_GLM_Data_mod_3_4$riparian_area_m2))
UDA_z  <- as.numeric(scale(aoi_GLM_Data_mod_3_4$catchment_km2))

# Medians of standardized covariates
TO_z_med   <- median(TO_z, na.rm = TRUE)
RIPA_z_med <- median(RIPA_z, na.rm = TRUE)
UDA_z_med  <- median(UDA_z, na.rm = TRUE)

# Global basline SWA
SWA0_global <- post$alpha +
  post$beta2 * TO_z_med +
  post$beta3 * RIPA_z_med +
  post$beta4 * UDA_z_med

SWA0_global <- pmax(SWA0_global, 0)  # enforce SWA ≥ 0


# global effect

prop_global <- post$beta1 / SWA0_global
quantile(prop_global, c(0.025, 0.5, 0.975))


prop_df <- tibble(prop = prop_global)




hist(SWA0_global,
     main = "Posterior Distribution of Global Baseline SWA",
     xlab = "Baseline SWA (m²)")



ggplot(prop_df, aes(x = prop)) +
  geom_density(fill = "skyblue", alpha = 0.4, linewidth = 1.2) +
  geom_vline(xintercept = median(prop_global), color = "blue", linewidth = 1) +
  geom_vline(xintercept = quantile(prop_global, 0.025), linetype = "dashed") +
  geom_vline(xintercept = quantile(prop_global, 0.975), linetype = "dashed") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    x = "Proportional Change in SWA (Dammed vs Undammed)",
    y = "Posterior Density",
    title = "Global Proportional Effect of Beaver Dams (Model 4)",
    subtitle = "Population-level dam effect relative to global baseline SWA"
  ) +
  theme_minimal(base_size = 14)
```

