---
title: "Ch1_HGLMM_Late season influence of beaver dams on surface water area"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---


The purpose of this script is to explore the effect of beaver dams on surface water area, relative to local geomorphology, and contributing drainage area. This modeling focuses solely on late summer months; August : September. 

Two models are presented herein. Both models have the same random effects structure for stream, year, and sample event. Model 1 examines the effect of beaver dam presence on surface water area, with no consideration for local context. Model 2 includes additional fixed effects of channel incision, riparian area, and drainage area. 

----

```{r,echo=FALSE}
  knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

```{r,echo=FALSE}
library(tidyverse)
library(ggplot2)
library(dataRetrieval)
library(knitr)
library(ggplot2)
library(zoo) #for moving window
library(stringr)
library(vioplot)
library(purrr)
library(lavaan)
library(blavaan)
library(ggcorrplot)
library(lme4)
library(broom.mixed)
```
      
----

Data from USGS LBV Guston Coulee gage, near Saco, MT:
----

```{r,echo=FALSE, warning=FALSE}
# LOWER BEAVER CK
# Gage near Guston Coulee
usgsGage_06166000 <- readNWISdv(
  siteNumbers = "06166000",
  parameterCd = "00060",
  startDate = "2019-06-01",
  endDate   = "2023-10-01"
)

###########################################

# Clean USGS
usgs_df <- usgsGage_06166000 %>%
  mutate(
    source = "USGS",
    date   = as.Date(Date),
    flow   = X_00060_00003
    ) %>%
  filter(
    (date >= as.Date("2019-06-01") & date <= as.Date("2019-10-01")) |
    (date >= as.Date("2021-06-01") & date <= as.Date("2021-10-01")) |
    (date >= as.Date("2023-06-01") & date <= as.Date("2023-10-01")) ) %>% 
  mutate(
    zFlow = scale(flow)
  ) %>% 
  select(source,
         date,
         zFlow) %>% 
  mutate(surveyPeriod = case_when(between(date, as.Date("2019-06-01"), as.Date("2019-06-14")) ~ 191,
                                between(date, as.Date("2019-06-15"), as.Date("2019-06-30")) ~ 192,
                                between(date, as.Date("2019-07-01"), as.Date("2019-07-14")) ~ 193,
                                between(date, as.Date("2019-07-15"), as.Date("2019-07-31")) ~ 194,
                                between(date, as.Date("2019-08-01"), as.Date("2019-08-14")) ~ 195,
                                between(date, as.Date("2019-08-15"), as.Date("2019-08-31")) ~ 196,
                                between(date, as.Date("2019-09-01"), as.Date("2019-09-14")) ~ 197,
                                between(date, as.Date("2019-09-15"), as.Date("2019-09-30")) ~ 198,
                                between(date, as.Date("2021-06-01"), as.Date("2021-06-14")) ~ 211,
                                between(date, as.Date("2021-06-15"), as.Date("2021-06-30")) ~ 212,
                                between(date, as.Date("2021-07-01"), as.Date("2021-07-14")) ~ 213,
                                between(date, as.Date("2021-07-15"), as.Date("2021-07-31")) ~ 214,
                                between(date, as.Date("2021-08-01"), as.Date("2021-08-14")) ~ 215,
                                between(date, as.Date("2021-08-15"), as.Date("2021-08-31")) ~ 216,
                                between(date, as.Date("2021-09-01"), as.Date("2021-09-14")) ~ 217,
                                between(date, as.Date("2021-09-15"), as.Date("2021-09-30")) ~ 218,
                                between(date, as.Date("2023-06-01"), as.Date("2023-06-14")) ~ 231,
                                between(date, as.Date("2023-06-15"), as.Date("2023-06-30")) ~ 232,
                                between(date, as.Date("2023-07-01"), as.Date("2023-07-14")) ~ 233,
                                between(date, as.Date("2023-07-15"), as.Date("2023-07-31")) ~ 234,
                                between(date, as.Date("2023-08-01"), as.Date("2023-08-14")) ~ 235,
                                between(date, as.Date("2023-08-15"), as.Date("2023-08-31")) ~ 236,
                                between(date, as.Date("2023-09-01"), as.Date("2023-09-14")) ~ 237,
                                between(date, as.Date("2023-09-15"), as.Date("2023-09-30")) ~ 238,
                                TRUE ~ NA_real_)) %>% 
  filter(surveyPeriod > 0)

usgs_df$surveyPeriod <- as.factor(usgs_df$surveyPeriod)


mean_flow_per_survey_period <- usgs_df %>% 
  group_by(surveyPeriod) %>%
  summarise(zFlow_mean = mean(zFlow),
            .groups = "drop")



lateSummerFlow <- usgs_df %>% 
  filter(surveyPeriod != 191,
         surveyPeriod != 192,
         surveyPeriod != 193,
         surveyPeriod != 194,
         surveyPeriod != 211,
         surveyPeriod != 212,
         surveyPeriod != 213,
         surveyPeriod != 214,
         surveyPeriod != 231,
         surveyPeriod != 232,
         surveyPeriod != 233,
         surveyPeriod != 234,
         )
```

Late Summer Flow Conditions
----

2019:
----

Moderate August with peak in September 

```{r,echo=FALSE}

flow19 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2019-06-01") & date <= as.Date("2019-10-01"))
  )


####

ggplot(flow19, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '19 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```

----
  
2021:
----

Low flow conditions throughout the summer season. Part of 3 year sustained drought conditions (2020, 2021, 2022).


Gage recorded no flow during August & September



```{r,echo=FALSE}

flow21 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2021-06-01") & date <= as.Date("2021-10-01"))
  )


####

ggplot(flow21, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '21 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```  
  
----

2023:
----

Moderate to low August & September flows; after early season extreme flooding. 



```{r,echo=FALSE}
flow23 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2023-06-01") & date <= as.Date("2023-10-01"))
  )


####

ggplot(flow23, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '23 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```  



GEE-derived data for each stream:
----

```{r, echo=FALSE}

#Read in data frames for each HUC12/Year combo:

# WILLOW:
#2019:
wc_19 <- read.csv("WC_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'WC')
#2021:
wc_21 <- read.csv("WC_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'WC')
#2023:
wc_23 <- read.csv("WC_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'WC')
#AllYears:
WC <- bind_rows(wc_19,wc_21,wc_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LONE TREE:
#2019:
lt_19 <- read.csv("LT_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'LT')
#2021:
lt_21 <- read.csv("LT_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'LT')
#2023:
lt_23 <- read.csv("LT_2023_DF_Final_Long_0427Update_250Spacing.csv")%>% 
  mutate(year = 2023,
         watershed = 'LT')

#AllYears:
LT <- bind_rows(lt_19,lt_21,lt_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))

# BRAZIL:
#2019:
brz_19 <- read.csv("BRZ_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'BRZ')
#2021:
brz_21 <- read.csv("BRZ_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'BRZ')
#2023:
brz_23 <- read.csv("BRZ_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'BRZ')

#AllYears:
BRZ <- bind_rows(brz_19,brz_21,brz_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LARB:
#2019:
lrb_19 <- read.csv("LRB_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'LRB')
#2021:
lrb_21 <- read.csv("LRB_2021_DF_Final_Long_0427Update_250Spacing.csv")%>% 
  mutate(year = 2021,
         watershed = 'LRB')
#2023:
lrb_23 <- read.csv("LRB_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'LRB')

#AllYears:
LRB <- bind_rows(lrb_19,lrb_21,lrb_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LOWER BEAVER:
#2019:
lbv_19 <- read.csv("LBV_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'LBV')
#2021:
lbv_21 <- read.csv("LBV_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'LBV')
#2023:
lbv_23 <- read.csv("LBV_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'LBV')

#AllYears:
LBV <- bind_rows(lbv_19,lbv_21,lbv_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# UPPER BEAVER:
#2019:
ubv_19 <- read.csv("UBV_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'UBV')
#2021:
ubv_21 <- read.csv("UBV_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'UBV')
#2023:
ubv_23 <- read.csv("UBV_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'UBV')

#AllYears:
UBV <- bind_rows(ubv_19,ubv_21,ubv_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
```


----

Currently each incoming data frame is of the following structure:

```{r,include=FALSE}
str(brz_19)
```

Where:

    - clusterID_seq = dammed/undammed sample pair ID
    - window_type = whether a sample window is 'dammed' or 'undammed'
    - watershed = watershed identifier (e.g., "BRZ" = Brazil Ck)
    - year = year
    - window_length = length (m) of stream flow line within sample window
    - catchment_km2 = contributing catchment area/upstream drainage area (km^2)
    - riparian_area_m2 = riparian area (m^2) of sample window; 2-m vertical
      buffer from DEM low point (stream flow line)
    - TO = mean topographic openness across window
    - p_window: avg dam detection prob for target window; see CP Ch1
    - f_window: avg dam false positive prob for target window; see CP Ch1
    - beav_count_all: all unique detected dams
    - beav_count_conf: unique confidently detected dams
    - anthro_count: all unique detected anthropogenic dams
    - isAnthro: National Wetland Inventory (NWI) "dyked/impounded"
    - Nhat = estimated dam count within sample window
      - ...= beav_count_all/(p_window(1-f_window)  [McCallum 2005]
    - TO_mean = mean topographic openness across window
    - time_period = 2wk interval 06/01 - 09/30
    - SWA = surface water area (m^2) for window during time (t)
    - t = time (t) {same as time_period}

----
----
----
----


**Relative flow data.**

NOTE: currently this represents data from the USGS Guston Coulee gage that has been z-standardized over the current data acquisition range (June-Sept; 2019, 2021, 2023). 

Then, for each 2-week survey time period (1:8), the average relative flow $[mean(zFlow)]$ was computed, and added as a data column. IT IS ASSUMED that this flow data is representative of the RELATIVE flow pattern experienced throughout the study area during this time.

----

Add relFlow data:

```{r,include=FALSE, fig.width=10}
WC_RelFlow <- WC %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = Nhat/window_length)
```

```{r,include=FALSE, fig.width=10}
LT_RelFlow <- LT %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = Nhat/window_length)

```

```{r,include=FALSE, fig.width=10}
BRZ_RelFlow <- BRZ %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = Nhat/window_length)
```

```{r,include=FALSE, fig.width=10}
LRB_RelFlow <- LRB %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = Nhat/window_length)

```

```{r,include=FALSE, fig.width=10}
LBV_RelFlow <- LBV %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = Nhat/window_length)
```

```{r,include=FALSE, fig.width=10}
UBV_RelFlow <- UBV %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = Nhat/window_length)
```





Hierarchical Generalized Linear Mixed Effects Models
----

Proposed Models:
----


**Both models:** 

$$  SWA \sim N(\mu, \sigma);$$



**M1:**


$$\mu = \alpha + \eta_{stream} + \eta_{event} + \eta_{cluster} + (\beta1+\beta_{streamEvent})*damPresence $$

----
----

**M2:**

$$\mu = \alpha + \eta_{stream} + \eta_{event} + \eta_{cluster} + (\beta_1+\beta_{streamEvent})*damPresence + \beta_2 * Incision + \beta_3 * riparianArea + \beta_4 * drainageArea$$

----
----
---- 

```{r,include=FALSE}
# Create GLMM data frame:

aoi_GLM_Data <- WC_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_m) %>% 
  bind_rows(LT_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_m)) %>% 
  bind_rows(BRZ_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_m)) %>% 
  bind_rows(LRB_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_m)) %>% 
  bind_rows(LBV_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_m)) %>% 
  bind_rows(UBV_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_m))


aoi_GLM_Data$clusterID_unique <- interaction(
  aoi_GLM_Data$year,
  aoi_GLM_Data$watershed,
  aoi_GLM_Data$clusterID_seq,
  drop = TRUE
)

aoi_GLM_Data$clusterID_num <- as.numeric(aoi_GLM_Data$clusterID_unique)

#aoi_GLM_Data$eventID <- interaction(
#  aoi_GLM_Data$year,
#  aoi_GLM_Data$surveyPeriod,
#  drop = TRUE
#)

#aoi_GLM_Data$eventID_num <- as.numeric(aoi_GLM_Data$eventID)

aoi_GLM_Data$stream <- as.numeric(as.factor(aoi_GLM_Data$watershed))

aoi_GLM_Data$sampleEvent <- as.numeric(aoi_GLM_Data$surveyPeriod)

aoi_GLM_Data$year_id <- as.numeric(as.factor(aoi_GLM_Data$year))

aoi_GLM_Data <- aoi_GLM_Data %>% 
  mutate(damPresence = case_when(Nhat > 0 ~ 1,
                                 Nhat == 0 ~ 0))

str(aoi_GLM_Data)



aoi_GLM_Data_mod <- aoi_GLM_Data %>%
  filter(!is.na(SWA))   # or !is.na(RI) if that’s the response


# Build explicit lookups from the data actually used in the model
stream_lookup <- aoi_GLM_Data_mod %>%
  distinct(stream, watershed) %>%
  arrange(stream)

year_lookup <- aoi_GLM_Data_mod %>%
  distinct(year_id, year) %>%
  arrange(year_id)

uda_lookup <- aoi_GLM_Data_mod %>%
  group_by(watershed) %>%
  summarise(max_catchment = max(catchment_km2, na.rm = TRUE)) %>%
  arrange(max_catchment)



# Start from the full modeled data and remove early summer time periods
# This analysis will look exclusively at late season mesic resource availability
# 08/01-09/30
# t5:t8
aoi_GLM_Data_mod_3_4 <- aoi_GLM_Data_mod %>% 
  filter(!(sampleEvent %in% 1:4) &      # drop 2019 t1-t4
           !(sampleEvent %in% 9:12) &   # drop 2021 t1-t4
           !(sampleEvent %in% 17:20))   # drop 2023 t1–t4

event_lookup <- data.frame(
  old = sort(unique(aoi_GLM_Data_mod_3_4$sampleEvent)),
  new = 1:length(unique(aoi_GLM_Data_mod_3_4$sampleEvent))
)

aoi_GLM_Data_mod_3_4 <- aoi_GLM_Data_mod_3_4 %>%
  left_join(event_lookup, by = c("sampleEvent" = "old")) %>%
  mutate(sampleEvent_new = new) %>%
  select(-new)

event_year_3_4 <- tapply(aoi_GLM_Data_mod_3_4$year_id,
                         aoi_GLM_Data_mod_3_4$sampleEvent_new,
                         unique)
event_year_3_4 <- as.numeric(event_year_3_4)




######
###### Z-STANDARDIZE LOG(SWA) FOR THE SAKE OF PRIORS
######

SWA_log <- log(aoi_GLM_Data_mod_3_4$SWA +1)

SWA_log_mean <- mean(SWA_log, na.rm = T)

SWA_log_sd <- sd(SWA_log, na.rm = T)

SWA_response <- (SWA_log - SWA_log_mean)/SWA_log_sd
```



This section explores late season SWA in relation to beaver dam presence.

Uses nested random slopes for beaver dam effect to allow for stream, year, and sampleEvent to cause deviation from the global mean beaver effect.


Implement HGLMMs:
----


M3: Dam Presence Only; seasonal random slope (sampleEvent)
----

```{r,echo=FALSE}
    library(jagsUI)



    Sys.time()
    
    sink("HGLMM_damPres.jags")
    
    cat("
    model {
    
      for (i in 1:N) {
        SWA[i] ~ dnorm(mu[i], tau)
    
      mu[i] <- alpha
                 + (beta1
                    + b_stream[stream[i]] 
                    + b_year[year[i]]
                    + b_event[sampleEvent[i]]) * damPresence[i]
                 + eta_stream[stream[i]]
                 + eta_event[sampleEvent[i]]
                 + eta_cluster[clusterID[i]]
      }
    
      # Fixed effects (standardized log-SWA scale)
      alpha ~ dnorm(0, 1)  #mean SWA (no dam) ~ 0 +- 1 sd
      beta1 ~ dnorm(0, 1)  #dam effect ~ 0 +- 1 sd
    
      # Random Effects
      # Stream/Watershed
      for (j in 1:n_stream) {
        eta_stream[j] ~ dnorm(0, tau_stream)     # random intercept for stream
        b_stream[j]   ~ dnorm(0, tau_b_stream)   # random slope for stream
      }
      
      # Year-level dam effect
      for(yr in 1:n_year) {
        b_year[yr] ~ dnorm(0, tau_b_year)        # random slope for year
      }
      
      # Event-level dam effect nested within year
      for (k in 1:n_event) {
        b_event[k] ~ dnorm(b_year[event_year[k]],tau_b_event)       # nested random slope
        eta_event[k] ~ dnorm(0, tau_event)       # random intercept for event
      }
      
      # Cluster ID
      for (c in 1:n_cluster) {
        eta_cluster[c] ~ dnorm(0, tau_cluster)    # random intercept for sampleCluster
      }
    
      # Hyperpriors (SDs on standardized log-SWA scale)
      sigma ~ dnorm(0, 1) T(0,)   # residual SD; half normal
      tau <- pow(sigma, -2)
    
      sigma_stream ~ dnorm(0,1) T(0,)
      tau_stream <- pow(sigma_stream, -2)
    
      sigma_b_stream ~ dnorm(0,0.5) T(0,)
      tau_b_stream <- pow(sigma_b_stream, -2)
      
      sigma_b_year ~ dnorm(0,0.5) T(0,)
      tau_b_year <- pow(sigma_b_year, -2)
    
      sigma_b_event ~ dnorm(0,0.5) T(0,)
      tau_b_event <- pow(sigma_b_event, -2)
      
      sigma_event ~ dnorm(0,1) T(0,)
      tau_event <- pow(sigma_event, -2)
      
      sigma_cluster ~ dnorm(0,1) T(0,)
      tau_cluster <- pow(sigma_cluster, -2)
    }
    ", fill = TRUE)
    
    sink()
```

```{r,echo=FALSE}
m3_data_list <- list(
  N = nrow(aoi_GLM_Data_mod_3_4),
  SWA = SWA_response,
  damPresence = aoi_GLM_Data_mod_3_4$damPresence,
  stream = aoi_GLM_Data_mod_3_4$stream,
  year = aoi_GLM_Data_mod_3_4$year_id,
  clusterID = aoi_GLM_Data_mod_3_4$clusterID_num,
  n_stream = length(unique(aoi_GLM_Data_mod_3_4$stream)),
  n_year = length(unique(aoi_GLM_Data_mod_3_4$year_id)),
  n_cluster = length(unique(aoi_GLM_Data_mod_3_4$clusterID_num)),
  sampleEvent = aoi_GLM_Data_mod_3_4$sampleEvent_new,
  n_event = length(unique(aoi_GLM_Data_mod_3_4$sampleEvent_new)),
  event_year = event_year_3_4,
  SWA_log_mean = SWA_log_mean,
  SWA_log_sd = SWA_log_sd
)

m3inits <- function(){list(
    alpha = rnorm(1, 0, 1),
    beta1 = rnorm(1, 0, 1)
    )}

parameters_m3 <- c(
  "alpha",
  "beta1",
  "b_stream",
  "b_year",
  "b_event",
  "sigma",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma_stream",
  "sigma_b_stream",
  "sigma_b_year",
  "sigma_b_event",
  "sigma_event",
  "sigma_cluster"
)

nchains <- 3
nthin <- 10
niter <- 20000
nburn <- 5000

m3_HGLM_SWA <- jags(data = m3_data_list,
                   inits = m3inits,
                   parameters.to.save = parameters_m3,
                   model.file = "HGLMM_damPres.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

print(m3_HGLM_SWA) # model with local and watershed dam density as fixed effects

```

difference in sys.time = elapsed time for model to run

```{r, echo=FALSE}
# Extract posteriors for parameters of interest

#library(dplyr)
#library(tidyr)
#library(ggplot2)
#library(purrr)

# Extract posterior draws
M3post <- m3_HGLM_SWA$sims.list

# Summaries for alpha and beta 1
M3df_scalar <- tibble(
  parameter = c("alpha", "beta1"),
  mean = c(mean(M3post$alpha), 
           mean(M3post$beta1)),
  lower = c(quantile(M3post$alpha, 0.025), 
            quantile(M3post$beta1, 0.025)),
  upper = c(quantile(M3post$alpha, 0.975), 
            quantile(M3post$beta1, 0.975))
)

# Stream-level slopes
M3df_stream <- map_df(1:ncol(M3post$b_stream), function(j){
  tibble(
    parameter = paste0("b_stream[", j, "]"),
    mean = mean(M3post$b_stream[, j]),
    lower = quantile(M3post$b_stream[, j], 0.025),
    upper = quantile(M3post$b_stream[, j], 0.975)
  )
})

# Sample Event-level slopes
M3df_event <- map_df(1:ncol(M3post$b_event), function(y){
  tibble(
    parameter = paste0("b_event[", y, "]"),
    mean = mean(M3post$b_event[, y]),
    lower = quantile(M3post$b_event[, y], 0.025),
    upper = quantile(M3post$b_event[, y], 0.975)
  )
})

M3summary_df <- bind_rows(M3df_scalar, M3df_stream, M3df_event)

# Give parameters meaningful names
pretty_names <- c(
  alpha           = "Intercept",
  beta1           = "B1: DamPresence",
  "b_stream[1]"   = "Random Slope: Brazil",
  "b_stream[2]"   = "Random Slope: LowerBeaver",
  "b_stream[3]"   = "Random Slope: Larb",
  "b_stream[4]"   = "Random Slope: LoneTree",
  "b_stream[5]"   = "Random Slope: UpperBeaver",
  "b_stream[6]"   = "Random Slope: Willow",
  "b_event[1]"     = "Random Slope: 2019 Early Aug",
  "b_event[2]"     = "Random Slope: 2019 Late Aug",
  "b_event[3]"     = "Random Slope: 2019 Early Sep",
  "b_event[4]"     = "Random Slope: 2019 Late Sep",
  "b_event[5]"     = "Random Slope: 2021 Early Aug",
  "b_event[6]"     = "Random Slope: 2021 Late Aug",
  "b_event[7]"     = "Random Slope: 2021 Early Sep",
  "b_event[8]"     = "Random Slope: 2021 Late Sep",
  "b_event[9]"     = "Random Slope: 2023 Early Aug",
  "b_event[10]"     = "Random Slope: 2023 Late Aug",
  "b_event[11]"     = "Random Slope: 2023 Early Sep",
  "b_event[12]"     = "Random Slope: 2023 Late Sep"
)


# Plot
ggplot(M3summary_df, 
       aes(x = parameter, 
           y = mean)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower, 
                    ymax = upper), width = 0.1) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  coord_flip() +
  scale_x_discrete(labels = pretty_names) +
  theme_minimal(base_size = 14) +
  labs(
    title = expression("Model 3: " * SWA ~ Dam_Presence ),
    subtitle = "Posterior Means and 95% Credible Intervals",
    x = "Parameter",
    y = "Posterior Mean"
  ) +
  theme(axis.text.y = element_text(size = 12))

print(mean(m3_HGLM_SWA$sims.list$beta1))
```


This is on a z-standardized log(SWA) scale ^^^

----

Compute predicted dam effects:

```{r,echo=FALSE}
# Data dimensions:
M3n_sims   <- length(M3post$beta1)
M3n_stream <- ncol(M3post$b_stream)
M3n_event   <- ncol(M3post$b_event)

#range(M3post$b_stream[,3] + M3post$b_year[,2] + M3post$beta1)

# Compute dam effect for each streamxevent:

M3DamEffect <- array(NA, dim = c(M3n_sims, 
                                 M3n_stream, 
                                 M3n_event))

for(j in 1:M3n_stream){
  for(e in 1:M3n_event){
    
    yr <- event_year_3_4[e] #maps event -> year index
    
    M3DamEffect[,j,e] <- M3post$beta1 +
                         M3post$b_stream[,j] +
                         M3post$b_year[,yr] +
                         M3post$b_event[,e]
  }
}

# --- Compute baseline (no-dam) standardized log-SWA ---
M3Baseline_std <- array(NA, dim = c(M3n_sims, M3n_stream, M3n_event))

for(j in 1:M3n_stream){
  for(e in 1:M3n_event){
    M3Baseline_std[, j, e] <-
      M3post$alpha +
      M3post$eta_stream[, j] +
      M3post$eta_event[, e]
  }
}

# --- Compute dammed standardized log-SWA ---
M3Dammed_std <- M3Baseline_std + M3DamEffect

# --- Back-transform both to raw SWA (m²) ---
M3Baseline_raw <- exp(M3Baseline_std * SWA_log_sd + SWA_log_mean)-1
M3Dammed_raw   <- exp(M3Dammed_std   * SWA_log_sd + SWA_log_mean)-1

# --- Absolute increase in SWA (m²) ---
M3Delta_raw <- M3Dammed_raw - M3Baseline_raw

# --- Proportional increase in SWA ---
M3Ratio_raw <- (M3Dammed_raw + 1) / (M3Baseline_raw + 1)
M3Prop_raw  <- M3Ratio_raw - 1
```


Compute "typical dam effect across watersheds and late summer sampling events"

```{r,echo = F}
# All stream × event combinations
M3_delta_vec <- as.vector(M3Delta_raw)

M3_typical_effect_mean  <- mean(M3_delta_vec, na.rm = TRUE)
M3_typical_effect_lower <- quantile(M3_delta_vec, 0.025, na.rm = TRUE)
M3_typical_effect_upper <- quantile(M3_delta_vec, 0.975, na.rm = TRUE)

M3_typical_effect_mean
M3_typical_effect_lower
M3_typical_effect_upper

```


Compute global proportional change in SWA caused by dam presence:

```{r,echo=FALSE}
# Flatten all sims × streams × events
M3_prop_vec <- as.vector(M3Prop_raw)

M3_global_prop_mean  <- mean(M3_prop_vec, na.rm = TRUE)
M3_global_prop_lower <- quantile(M3_prop_vec, 0.025, na.rm = TRUE)
M3_global_prop_upper <- quantile(M3_prop_vec, 0.975, na.rm = TRUE)

M3_global_prop_mean
M3_global_prop_lower
M3_global_prop_upper




# visualize it!
M3_globalPropChange_df <- tibble(prop = M3_prop_vec)

ggplot(M3_globalPropChange_df, aes(x = prop)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  geom_vline(xintercept = M3_global_prop_mean, color = "black", linewidth = 1) +
  geom_vline(xintercept = M3_global_prop_lower, linetype = "dashed") +
  geom_vline(xintercept = M3_global_prop_upper, linetype = "dashed") +
  labs(
    title = "Model 3: Global Proportional Change in SWA Due to Beaver Dam Presence",
    x = "Proportional Increase",
    y = "Posterior Density"
  ) +
  xlim(-2,20) +
  theme_minimal(base_size = 14)
```




Compute proportional change in SWA across watersheds and sampling events

```{r,echo=FALSE}
#Summarize posteriors for each 

M3summary_list <- list()

for(j in 1:M3n_stream){
  for(e in 1:M3n_event){
    
    eff <- as.vector(M3Prop_raw[, j, e, drop = FALSE])
    
    M3summary_list[[paste(j, e, sep = "_")]] <- tibble(
      stream = j,
      event = e,
      mean = mean(eff),
      lower = quantile(eff, 0.025),
      upper = quantile(eff, 0.975),
      positive = quantile(eff, 0.025) > 0
    )
  }
}

# Build explicit lookups from the data actually used in the model
stream_lookup <- aoi_GLM_Data_mod %>%
  distinct(stream, watershed) %>%
  arrange(stream)

# Event lookup created above

M3effect_df <- bind_rows(M3summary_list) %>%
  left_join(stream_lookup, by = c("stream")) %>%
  left_join(event_lookup,  by = c("event" = "new"))

# order watersheds by factor levels of ascending max CDA
M3effect_df <- M3effect_df %>%
  mutate(watershed = factor(watershed, levels = uda_lookup$watershed))



# Plot the heatmap

ggplot(M3effect_df, aes(x = as.factor(old), 
                        y = watershed, 
                        fill = mean)) +
  geom_tile(color = "black") +
  scale_fill_viridis_c(
    option = "C",
    name = "Proportional Increase\n(SWA_dam / SWA_base - 1)"
  )    +
  labs(
    x = "Sampling Event",
    y = "Watershed",
    title = "M3: Predicted proportional increase in SWA due to beaver dam presence"
  ) +
  theme_minimal(base_size = 14)
```




Compare b1 trajectory across event, by stream:

```{r, echo=FALSE}
M3line_df <- M3effect_df %>%
  arrange(watershed, event) %>%
  mutate(event = as.numeric(event)) %>%
  group_by(watershed) %>%
  do({
    df <- .

    gap_rows <- tibble(
      stream = df$stream[1],
      event = c(4.5, 8.5),   # ← add as many gaps as you want
      mean = NA,
      lower = NA,
      upper = NA,
      positive = NA,
      watershed = df$watershed[1]
    )

    bind_rows(df, gap_rows)
  }) %>%
  arrange(watershed, event)



ggplot(M3line_df, aes(x = event, y = mean, color = watershed, group = watershed)) +
  geom_line(linewidth = 1.2, na.rm = TRUE) +
  geom_point(size = 2, na.rm = TRUE) +
  scale_x_continuous(
    breaks = unique(M3effect_df$event),
    labels = unique(event_lookup$label)   # if you want "19_t1", "19_t2", etc.
  ) +
  labs(
    title = "M3: Proportional change in SWA given beaver dam presence",
    subtitle = "Gaps represent breaks by Year",
    x = "Sampling Events (t5 : t8) \n (August : September) \n (2019, 2021, 2023)",
    y = "Proportional Difference from Baseline SWA",
    color = "Watershed"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

```



----

----

----

M4: Dam Presence + Context with updated random effects structure
----



    library(jagsUI)
    
    Sys.time()
    
    sink("HGLMM_allCovar.jags")
    
    cat("
        model {
    
      for (i in 1:N) {
        SWA[i] ~ dnorm(mu[i], tau)
    
      mu[i] <- alpha
                 + (beta1
                    + b_stream_event[stream[i],sampleEvent[i]]) * damPresence[i]
                 + beta2 * TO[i] 
                 + beta3 * RIPA[i]
                 + beta4 * UDA[i]
                 + eta_stream[stream[i]]
                 + eta_event[sampleEvent[i]]
                 + eta_cluster[clusterID[i]]
      }
    
      # Fixed effects
      alpha ~ dnorm(0, 0.001)
      beta1 ~ dnorm(0, 0.001)
      beta2 ~ dnorm(0, 0.001)
      beta3 ~ dnorm(0, 0.001)
      beta4 ~ dnorm(0, 0.001)
    
      # Random Effects
      # Stream/Watershed
      for (j in 1:n_stream) {
        eta_stream[j] ~ dnorm(0, tau_stream)     # random intercept for stream
        b_stream[j]   ~ dnorm(0, tau_b_stream)   # random slope for stream
      }
      
      # Year-level dam effect
      for(yr in 1:n_year) {
        b_year[yr] ~ dnorm(0, tau_b_year)        # random slope for year
      }
      
      # Event-level dam effect nested within year
      for (k in 1:n_event) {
        b_event[k] ~ dnorm(b_year[event_year[k]],tau_b_event)       # random slope for year/survey 'sampling event'
        eta_event[k] ~ dnorm(0, tau_event)
      }
      
      # Stream/Event Interaction
      for(j in 1:n_stream){
        for(k in 1:n_event){
          b_stream_event[j,k] ~ dnorm(b_event[k] + b_stream[j], tau_b_stream_event)
        }
      }
      
      # Cluster ID
      for (c in 1:n_cluster) {
        eta_cluster[c] ~ dnorm(0, tau_cluster)
      }
    
      # Hyperpriors
      sigma ~ dnorm(0, 1 / pow(sd_SWA, 2)) T(0,)   # half-normal with sd ≈ sd_SWA
      tau <- pow(sigma, -2)
    
      sigma_stream ~ dunif(0, 10)
      tau_stream <- pow(sigma_stream, -2)
    
      sigma_b_stream ~ dunif(0, 10)
      tau_b_stream <- pow(sigma_b_stream, -2)
      
      sigma_b_year ~ dunif(0,10)
      tau_b_year <- pow(sigma_b_year, -2)
    
      sigma_b_event ~ dunif(0, 10)
      tau_b_event <- pow(sigma_b_event, -2)
      
      sigma_event ~ dunif(0,10)
      tau_event <- pow(sigma_event, -2)
      
      sigma_b_stream_event ~ dunif(0,10)
      tau_b_stream_event <- pow(sigma_b_stream_event, -2)
      
      sigma_cluster ~ dunif(0,10)
      tau_cluster <- pow(sigma_cluster, -2)
    }
    ", fill = TRUE)
    
    sink()



```{r,echo=FALSE}
m4_data_list <- list(
  N = nrow(aoi_GLM_Data_mod_3_4),
  SWA = aoi_GLM_Data_mod_3_4$SWA,
  damPresence = aoi_GLM_Data_mod_3_4$damPresence,
  TO = as.numeric(scale(1-(aoi_GLM_Data_mod_3_4$TO/90))),
  RIPA = as.numeric(scale(aoi_GLM_Data_mod_3_4$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_GLM_Data_mod_3_4$catchment_km2)),
  stream = aoi_GLM_Data_mod_3_4$stream,
  clusterID = aoi_GLM_Data_mod_3_4$clusterID_num,
  n_stream = length(unique(aoi_GLM_Data_mod_3_4$stream)),
  n_year = length(unique(aoi_GLM_Data_mod_3_4$year_id)),
  n_cluster = length(unique(aoi_GLM_Data_mod_3_4$clusterID_num)),
  sampleEvent = aoi_GLM_Data_mod_3_4$sampleEvent_new,
  n_event = length(unique(aoi_GLM_Data_mod_3_4$sampleEvent_new)),
  event_year = event_year_3_4,
  sd_SWA = sd(aoi_GLM_Data_mod_3_4$SWA, na.rm = TRUE)
)

m4inits <- function(){list(
    alpha = rnorm(1, 0, 1),
    beta1 = rnorm(1, 0, 1)
    )}

parameters_m4 <- c(
  "alpha",
  "beta1",
  "beta2",
  "beta3",
  "beta4",
  "b_stream",
  "b_year",
  "b_event",
  "b_stream_event",
  "sigma",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma_stream",
  "sigma_b_stream",
  "sigma_b_year",
  "sigma_b_event",
  "sigma_b_stream_event",
  "sigma_event",
  "sigma_cluster"
)

nchains <- 3
nthin <- 10
niter <- 20000
nburn <- 5000

m4_HGLM_SWA <- jags(data = m4_data_list,
                   inits = m4inits,
                   parameters.to.save = parameters_m4,
                   model.file = "HGLMM_allCovar.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(m4_HGLM_SWA) # model with local and watershed dam density as fixed effects
```
difference in sys.time = elapsed time for model to run


----
----

**NOTE:**

TO was transformed to "channel incision" by taking the inverse of proportion openness:


$$ChannelIncision = 1 - (TO/90) $$


Where $(TO/90)$ is the proportion of max openness (90 degrees = "flat"/"fully open")


----
----


```{r, echo=FALSE}
# Extract posteriors for parameters of interest

#library(dplyr)
#library(tidyr)
#library(ggplot2)
#library(purrr)

# Extract posterior draws
m4post <- m4_HGLM_SWA$sims.list

# Select only scalar parameters (convert random effects matrices)
# Scalars
m4df_scalar <- tibble(
  parameter = c("alpha", "beta1", "beta2", "beta3", "beta4"),
  mean = c(mean(m4post$alpha), 
           mean(m4post$beta1), 
           mean(m4post$beta2), 
           mean(m4post$beta3),
           mean(m4post$beta4)),
  lower = c(quantile(m4post$alpha, 0.025), 
            quantile(m4post$beta1, 0.025),
            quantile(m4post$beta2, 0.025),
            quantile(m4post$beta3, 0.025),
            quantile(m4post$beta4, 0.025)),
  upper = c(quantile(m4post$alpha, 0.975), 
            quantile(m4post$beta1, 0.975),
            quantile(m4post$beta2, 0.975),
            quantile(m4post$beta3, 0.975),
            quantile(m4post$beta4, 0.975))
)

# Stream-level slopes
m4df_stream <- map_df(1:ncol(m4post$b_stream), function(j){
  tibble(
    parameter = paste0("b_stream[", j, "]"),
    mean = mean(m4post$b_stream[, j]),
    lower = quantile(m4post$b_stream[, j], 0.025),
    upper = quantile(m4post$b_stream[, j], 0.975)
  )
})

# Season-level slopes
m4df_event <- map_df(1:ncol(m4post$b_event), function(y){
  tibble(
    parameter = paste0("b_event[", y, "]"),
    mean = mean(m4post$b_event[, y]),
    lower = quantile(m4post$b_event[, y], 0.025),
    upper = quantile(m4post$b_event[, y], 0.975)
  )
})

m4summary_df <- bind_rows(m4df_scalar, m4df_stream, m4df_event)

#m4summary_df <- m4df_scalar

# Give parameters meaningful names
pretty_names2 <- c(
  alpha           = "Intercept",
  beta1           = "Dam Presence",
  beta2           = "Channel Incision",
  beta3           = "Riparian Area (m^2)",
  beta4           = "Contributing Drainage Area",
  "b_stream[1]"   = "Brazil Ck",
  "b_stream[2]"   = "Lower Beaver Ck",
  "b_stream[3]"   = "Larb Ck",
  "b_stream[4]"   = "Lone Tree Ck",
  "b_stream[5]"   = "Upper Beaver Ck",
  "b_stream[6]"   = "Willow Ck",
  "b_event[1]"     = "19_t5",
  "b_event[2]"     = "19_t6",
  "b_event[3]"     = "19_t7",
  "b_event[4]"     = "19_t8",
  "b_event[5]"     = "21_t5",
  "b_event[6]"     = "21_t6",
  "b_event[7]"     = "21_t7",
  "b_event[8]"     = "21_t8",
  "b_event[9]"     = "23_t5",
  "b_event[10]"     = "23_t6",
  "b_event[11]"     = "23_t7",
  "b_event[12]"     = "23_t8"
)


# Plot
ggplot(
  m4summary_df, 
  aes(x = parameter, y = mean)
) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  coord_flip() +
  scale_x_discrete(labels = pretty_names2) +
  theme_minimal(base_size = 14) +
  labs(
    title = expression("Model 4: (" * SWA * " ~ " * Dam_Presence * ") + Geomorphic Context"),
    subtitle = "Posterior Means and 95% Credible Intervals",
    x = "Parameter",
    y = "Posterior Mean"
  ) +
  theme(axis.text.y = element_text(size = 12))
```

m4 Mean Dam Effect:

```{r,echo=FALSE}
print(mean(m4_HGLM_SWA$sims.list$beta1))
```

----

m4 Mean TO Effect

Interpretation: SWA(m^2) change per 1 standard deviation increase in channel incision

```{r,echo=FALSE}
print(mean(m4_HGLM_SWA$sims.list$beta2))
```

----

m4 Mean RIPA Effect:

Interpretation: SWA(m^2) change per 1 standard deviation increase in riparian area


```{r,echo=FALSE}
print(mean(m4_HGLM_SWA$sims.list$beta3))
```

----

m4 Mean CDA Effect

Interpretation: SWA(m^2) change per 1 standard deviation increase in contributing drainage area


```{r,echo=FALSE}
print(mean(m4_HGLM_SWA$sims.list$beta4))
```



```{r,echo=FALSE}
# HEAT MAP IT

m4n_sims   <- length(m4post$beta1)
m4n_stream <- ncol(m4post$b_stream)
m4n_event   <- ncol(m4post$b_event)

#range(m4post$b_stream[,3] + m4post$b_year[,2] + m4post$beta1)


m4DamEffect <- array(NA, dim = c(m4n_sims, 
                                 m4n_stream, 
                                 m4n_event))

for(j in 1:m4n_stream){
  for(k in 1:m4n_event){
    m4DamEffect[, j, k] <-
      m4post$beta1 +
      m4post$b_stream[, j] +
      m4post$b_event[, k]
  }
}

```


```{r,echo=FALSE}
#Summarize posteriors for each 

m4summary_list <- list()

for(j in 1:m4n_stream){
  for(k in 1:m4n_event){
    
    eff <- as.vector(m4DamEffect[, j, k, drop = FALSE])
    
    m4summary_list[[paste(j, k, sep = "_")]] <- tibble(
      stream = j,
      event = k,
      mean = mean(eff),
      lower = quantile(eff, 0.025),
      upper = quantile(eff, 0.975),
      positive = quantile(eff, 0.025) > 0
    )
  }
}


m4effect_df <- bind_rows(m4summary_list) %>%
  left_join(stream_lookup, by = c("stream")) %>% 
  left_join(event_lookup, by = c("event" = "new"))


# Plot the heatmap

ggplot(m4effect_df, aes(x = as.factor(old), 
                        y = watershed, 
                        fill = mean)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(
    low = "white",
    mid = "cornflowerblue",
    high = "black",
    midpoint = 193,
    name = "Model 2 
Dam Effect"
  ) +
  labs(
    x = "Sample Event",
    y = "Watershed",
    title = "M4: Magnitude of Dam Effect by Watershed and SampleEvent",
    subtitle = "Aug:Sep; 5:8 = 2019; 13:16 = 2021; 21:24 = 2023"
  ) +
  theme_minimal(base_size = 14)
```






Visualize proportional change from undammed to dammed
----

```{r, echo=FALSE}
# Compute predicted baseline for 'typical' reach for each stream/year

post <- m4post

# --- Recreate standardized covariates exactly as in m4_data_list ---

# Channel incision = 1 - (TO/90)
TO_ci <- 1 - (aoi_GLM_Data_mod_3_4$TO / 90)
TO_z  <- as.numeric(scale(TO_ci))

RIPA_z <- as.numeric(scale(aoi_GLM_Data_mod_3_4$riparian_area_m2))
UDA_z  <- as.numeric(scale(aoi_GLM_Data_mod_3_4$catchment_km2))

# Medians of standardized covariates
TO_z_med   <- median(TO_z, na.rm = TRUE)
RIPA_z_med <- median(RIPA_z, na.rm = TRUE)
UDA_z_med  <- median(UDA_z, na.rm = TRUE)

# Global basline SWA
SWA0_global <- post$alpha +
  post$beta2 * TO_z_med +
  post$beta3 * RIPA_z_med +
  post$beta4 * UDA_z_med

SWA0_global <- pmax(SWA0_global, 0)  # enforce SWA ≥ 0


# global effect

prop_global <- post$beta1 / SWA0_global
quantile(prop_global, c(0.025, 0.5, 0.975))


prop_df <- tibble(prop = prop_global)




hist(SWA0_global,
     main = "Posterior Distribution of Global Baseline SWA",
     xlab = "Baseline SWA (m²)")

```
```{r,echo=TRUE}
mean(SWA0_global)
sd(SWA0_global)
```

```{r,echo=FALSE}
ggplot(prop_df, aes(x = prop)) +
  geom_density(fill = "skyblue", alpha = 0.4, linewidth = 1.2) +
  geom_vline(xintercept = mean(prop_global), color = "blue", linewidth = 1) +
  geom_vline(xintercept = quantile(prop_global, 0.025), linetype = "dashed") +
  geom_vline(xintercept = quantile(prop_global, 0.975), linetype = "dashed") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1),
                     limits = c(0, NA)) +
  labs(
    x = "Proportional Change in SWA (Dammed vs Undammed)",
    y = "Posterior Density",
    title = "Global Proportional Effect of Beaver Dams (Model 4)",
    subtitle = "Population-level dam effect relative to global baseline SWA"
  ) +
  theme_minimal(base_size = 14)
```

```{r,echo=TRUE}
mean(prop_df$prop)*100
```

    NOTE: Blue line = mean() of posterior estimate for proportional change
          Dashed lines = 95% CI


----
----

**^^^^typical dam effect = (+)~185m^2 SWA compared to undammed. Typical baseline SWA condition = ~134m^2. Hence, near doubling when we look at proportional change.**

----
----

Compare b1 trajectory across event, by stream:
----

```{r, echo=FALSE}
M4line_df <- m4effect_df %>%
  arrange(watershed, event) %>%
  mutate(event = as.numeric(event)) %>%
  group_by(watershed) %>%
  do({
    df <- .

    gap_rows <- tibble(
      stream = df$stream[1],
      event = c(4.5, 8.5),   # ← add as many gaps as you want
      mean = NA,
      lower = NA,
      upper = NA,
      positive = NA,
      watershed = df$watershed[1]
    )

    bind_rows(df, gap_rows)
  }) %>%
  arrange(watershed, event)



ggplot(M4line_df, aes(x = event, y = mean, color = watershed, group = watershed)) +
  geom_line(linewidth = 1.2, na.rm = TRUE) +
  geom_point(size = 2, na.rm = TRUE) +
  scale_x_continuous(
    breaks = unique(m4effect_df$event),
    labels = unique(event_lookup$label)   # if you want "19_t1", "19_t2", etc.
  ) +
  labs(
    title = "M4: Seasonal Trajectory of Beaver Dam Effect on SWA, by Watershed",
    subtitle = "Gaps represent breaks by Year",
    x = "Sampling Events (t5 : t8) \n (August : September) \n (2019, 2021, 2023)",
    y = "Dam Effect (SWA m^2)",
    color = "Watershed"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

```


