---
title: "Exploring damEffect on SWA with RF model"
author: "Matt Webster"
date: "2026-07-09"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---




**NOTE**
----

**RF EXPLORATION ON HOLD UNTIL FURTHER NOTICE**
----

07.14.2026
----

Pursue hurdle model with hierarchical linear mixed effects framework
----

----
----

----
----

----
----

----
----



The purpose of this script is to explore the effect of beaver dams on surface water area, relative to local geomorphology, and contributing drainage area. This modeling focuses solely on late summer months; August & September. 

This particular iteration explores the use of random forest models to assess importance of dam_m and damDensity as predictors of surface water area. 


----
----

```{r,echo=FALSE}
  knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

```{r,echo=FALSE}
library(tidyverse)
library(ggplot2)
library(dataRetrieval)
library(knitr)
library(ggplot2)
library(zoo) #for moving window
library(stringr)
library(vioplot)
library(purrr)
library(lavaan)
library(blavaan)
library(ggcorrplot)
library(lme4)
library(broom.mixed)
library(randomForest)
library(ranger)
```
      
----

Data from USGS LBV Guston Coulee gage, near Saco, MT:
----

```{r,echo=FALSE, warning=FALSE}
# LOWER BEAVER CK
# Gage near Guston Coulee
usgsGage_06166000 <- read_waterdata_daily(
  monitoring_location_id = "USGS-06166000",
  parameter_code = "00060",
  time = c("2019-06-01","2023-10-01")
)

###########################################

# Clean USGS
usgs_df <- usgsGage_06166000 %>%
  mutate(
    source = "USGS",
    date   = as.Date(time),
    flow   = value
    ) %>%
  filter(
    (date >= as.Date("2019-06-01") & date <= as.Date("2019-10-01")) |
    (date >= as.Date("2021-06-01") & date <= as.Date("2021-10-01")) |
    (date >= as.Date("2023-06-01") & date <= as.Date("2023-10-01")) ) %>% 
  mutate(
    zFlow = scale(flow)
  ) %>% 
  select(source,
         date,
         zFlow) %>% 
  mutate(surveyPeriod = case_when(between(date, as.Date("2019-06-01"), as.Date("2019-06-14")) ~ 191,
                                between(date, as.Date("2019-06-15"), as.Date("2019-06-30")) ~ 192,
                                between(date, as.Date("2019-07-01"), as.Date("2019-07-14")) ~ 193,
                                between(date, as.Date("2019-07-15"), as.Date("2019-07-31")) ~ 194,
                                between(date, as.Date("2019-08-01"), as.Date("2019-08-14")) ~ 195,
                                between(date, as.Date("2019-08-15"), as.Date("2019-08-31")) ~ 196,
                                between(date, as.Date("2019-09-01"), as.Date("2019-09-14")) ~ 197,
                                between(date, as.Date("2019-09-15"), as.Date("2019-09-30")) ~ 198,
                                between(date, as.Date("2021-06-01"), as.Date("2021-06-14")) ~ 211,
                                between(date, as.Date("2021-06-15"), as.Date("2021-06-30")) ~ 212,
                                between(date, as.Date("2021-07-01"), as.Date("2021-07-14")) ~ 213,
                                between(date, as.Date("2021-07-15"), as.Date("2021-07-31")) ~ 214,
                                between(date, as.Date("2021-08-01"), as.Date("2021-08-14")) ~ 215,
                                between(date, as.Date("2021-08-15"), as.Date("2021-08-31")) ~ 216,
                                between(date, as.Date("2021-09-01"), as.Date("2021-09-14")) ~ 217,
                                between(date, as.Date("2021-09-15"), as.Date("2021-09-30")) ~ 218,
                                between(date, as.Date("2023-06-01"), as.Date("2023-06-14")) ~ 231,
                                between(date, as.Date("2023-06-15"), as.Date("2023-06-30")) ~ 232,
                                between(date, as.Date("2023-07-01"), as.Date("2023-07-14")) ~ 233,
                                between(date, as.Date("2023-07-15"), as.Date("2023-07-31")) ~ 234,
                                between(date, as.Date("2023-08-01"), as.Date("2023-08-14")) ~ 235,
                                between(date, as.Date("2023-08-15"), as.Date("2023-08-31")) ~ 236,
                                between(date, as.Date("2023-09-01"), as.Date("2023-09-14")) ~ 237,
                                between(date, as.Date("2023-09-15"), as.Date("2023-09-30")) ~ 238,
                                TRUE ~ NA_real_)) %>% 
  filter(surveyPeriod > 0)

usgs_df$surveyPeriod <- as.factor(usgs_df$surveyPeriod)


mean_flow_per_survey_period <- usgs_df %>% 
  group_by(surveyPeriod) %>%
  summarise(zFlow_mean = mean(zFlow),
            .groups = "drop")



lateSummerFlow <- usgs_df %>% 
  filter(surveyPeriod != 191,
         surveyPeriod != 192,
         surveyPeriod != 193,
         surveyPeriod != 194,
         surveyPeriod != 211,
         surveyPeriod != 212,
         surveyPeriod != 213,
         surveyPeriod != 214,
         surveyPeriod != 231,
         surveyPeriod != 232,
         surveyPeriod != 233,
         surveyPeriod != 234,
         )
```

Late Summer Flow Conditions
----

2019:
----

Moderate August with peak in September 

```{r,echo=FALSE}

flow19 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2019-06-01") & date <= as.Date("2019-10-01"))
  )


####

ggplot(flow19, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '19 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```

----
  
2021:
----

Low flow conditions throughout the summer season. Part of 3 year sustained drought conditions (2020, 2021, 2022).


Gage recorded no flow during August & September



```{r,echo=FALSE}

flow21 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2021-06-01") & date <= as.Date("2021-10-01"))
  )


####

ggplot(flow21, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '21 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```  
  
----

2023:
----

Moderate to low August & September flows; after early season extreme flooding. 



```{r,echo=FALSE}
flow23 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2023-06-01") & date <= as.Date("2023-10-01"))
  )


####

ggplot(flow23, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '23 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```  



GEE-derived data for each stream:
----

```{r, echo=FALSE}

#Read in data frames for each HUC12/Year combo:

# WILLOW:
#2019:
wc_19 <- read.csv("WC_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'WC')
#2021:
wc_21 <- read.csv("WC_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'WC')
#2023:
wc_23 <- read.csv("WC_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'WC')
#AllYears:
WC <- bind_rows(wc_19,wc_21,wc_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LONE TREE:
#2019:
lt_19 <- read.csv("LT_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'LT')
#2021:
lt_21 <- read.csv("LT_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'LT')
#2023:
lt_23 <- read.csv("LT_2023_DF_Final_Long_0427Update_250Spacing.csv")%>% 
  mutate(year = 2023,
         watershed = 'LT')

#AllYears:
LT <- bind_rows(lt_19,lt_21,lt_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))

# BRAZIL:
#2019:
brz_19 <- read.csv("BRZ_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'BRZ')
#2021:
brz_21 <- read.csv("BRZ_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'BRZ')
#2023:
brz_23 <- read.csv("BRZ_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'BRZ')

#AllYears:
BRZ <- bind_rows(brz_19,brz_21,brz_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LARB:
#2019:
lrb_19 <- read.csv("LRB_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'LRB')
#2021:
lrb_21 <- read.csv("LRB_2021_DF_Final_Long_0427Update_250Spacing.csv")%>% 
  mutate(year = 2021,
         watershed = 'LRB')
#2023:
lrb_23 <- read.csv("LRB_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'LRB')

#AllYears:
LRB <- bind_rows(lrb_19,lrb_21,lrb_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LOWER BEAVER:
#2019:
lbv_19 <- read.csv("LBV_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'LBV')
#2021:
lbv_21 <- read.csv("LBV_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'LBV')
#2023:
lbv_23 <- read.csv("LBV_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'LBV')

#AllYears:
LBV <- bind_rows(lbv_19,lbv_21,lbv_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# UPPER BEAVER:
#2019:
ubv_19 <- read.csv("UBV_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'UBV')
#2021:
ubv_21 <- read.csv("UBV_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'UBV')
#2023:
ubv_23 <- read.csv("UBV_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'UBV')

#AllYears:
UBV <- bind_rows(ubv_19,ubv_21,ubv_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
```


----

Currently each incoming data frame is of the following structure:

```{r,include=FALSE}
str(brz_19)
```

Where:

    - clusterID_seq = dammed/undammed sample pair ID
    - window_type = whether a sample window is 'dammed' or 'undammed'
    - watershed = watershed identifier (e.g., "BRZ" = Brazil Ck)
    - year = year
    - window_length = length (m) of stream flow line within sample window
    - catchment_km2 = contributing catchment area/upstream drainage area (km^2)
    - riparian_area_m2 = riparian area (m^2) of sample window; 2-m vertical
      buffer from DEM low point (stream flow line)
    - TO = mean topographic openness across window
    - p_window: avg dam detection prob for target window; see CP Ch1
    - f_window: avg dam false positive prob for target window; see CP Ch1
    - beav_count_all: all unique detected dams
    - beav_count_conf: unique confidently detected dams
    - anthro_count: all unique detected anthropogenic dams
    - isAnthro: National Wetland Inventory (NWI) "dyked/impounded"
    - Nhat = estimated dam count within sample window
      - ...= beav_count_all/(p_window(1-f_window)  [McCallum 2005]
    - TO_mean = mean topographic openness across window
    - time_period = 2wk interval 06/01 - 09/30
    - SWA = surface water area (m^2) for window during time (t)
    - t = time (t) {same as time_period}

----
----
----
----


**Relative flow data.**

NOTE: currently this represents data from the USGS Guston Coulee gage that has been z-standardized over the current data acquisition range (June-Sept; 2019, 2021, 2023). 

Then, for each 2-week survey time period (1:8), the average relative flow $[mean(zFlow)]$ was computed, and added as a data column. IT IS ASSUMED that this flow data is representative of the RELATIVE flow pattern experienced throughout the study area during this time.

----

Add relFlow data:

```{r,include=FALSE, fig.width=10}
WC_RelFlow <- WC %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = (Nhat/window_length)*1000)
```

```{r,include=FALSE, fig.width=10}
LT_RelFlow <- LT %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = (Nhat/window_length)*1000)

```

```{r,include=FALSE, fig.width=10}
BRZ_RelFlow <- BRZ %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = (Nhat/window_length)*1000)
```

```{r,include=FALSE, fig.width=10}
LRB_RelFlow <- LRB %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = (Nhat/window_length)*1000)

```

```{r,include=FALSE, fig.width=10}
LBV_RelFlow <- LBV %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = (Nhat/window_length)*1000)
```

```{r,include=FALSE, fig.width=10}
UBV_RelFlow <- UBV %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_m = (Nhat/window_length)*1000)
```


----
----

```{r,include=FALSE}
# Create GLMM data frame:

aoi_GLM_Data <- WC_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_m) %>% 
  bind_rows(LT_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_m)) %>% 
  bind_rows(BRZ_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_m)) %>% 
  bind_rows(LRB_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_m)) %>% 
  bind_rows(LBV_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_m)) %>% 
  bind_rows(UBV_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_m))


aoi_GLM_Data$clusterID_unique <- interaction(
  aoi_GLM_Data$year,
  aoi_GLM_Data$watershed,
  aoi_GLM_Data$clusterID_seq,
  drop = TRUE
)

aoi_GLM_Data$clusterID_num <- as.numeric(aoi_GLM_Data$clusterID_unique)

#aoi_GLM_Data$eventID <- interaction(
#  aoi_GLM_Data$year,
#  aoi_GLM_Data$surveyPeriod,
#  drop = TRUE
#)

#aoi_GLM_Data$eventID_num <- as.numeric(aoi_GLM_Data$eventID)

aoi_GLM_Data$stream <- as.numeric(as.factor(aoi_GLM_Data$watershed))

aoi_GLM_Data$sampleEvent <- as.numeric(aoi_GLM_Data$surveyPeriod)

aoi_GLM_Data$year_id <- as.numeric(as.factor(aoi_GLM_Data$year))

aoi_GLM_Data <- aoi_GLM_Data %>% 
  mutate(dam_m = case_when(Nhat > 0 ~ 1,
                                 Nhat == 0 ~ 0))

str(aoi_GLM_Data)



aoi_GLM_Data_mod <- aoi_GLM_Data %>%
  filter(!is.na(SWA))   # or !is.na(RI) if that’s the response


# Build explicit lookups from the data actually used in the model
stream_lookup <- aoi_GLM_Data_mod %>%
  distinct(stream, watershed) %>%
  arrange(stream)
stream_lookup$stream <- as.factor(stream_lookup$stream)

year_lookup <- aoi_GLM_Data_mod %>%
  distinct(year_id, year) %>%
  arrange(year_id)

uda_lookup <- aoi_GLM_Data_mod %>%
  group_by(watershed) %>%
  summarise(max_catchment = max(catchment_km2, na.rm = TRUE)) %>%
  arrange(max_catchment)



# Start from the full modeled data and remove early summer time periods
# This analysis will look exclusively at late season mesic resource availability
# 08/01-09/30
# t5:t8
aoi_GLM_Data_mod_3_4 <- aoi_GLM_Data_mod %>% 
  filter(!(sampleEvent %in% 1:4) &      # drop 2019 t1-t4
           !(sampleEvent %in% 9:12) &   # drop 2021 t1-t4
           !(sampleEvent %in% 17:20))   # drop 2023 t1–t4

event_lookup <- data.frame(
  old = sort(unique(aoi_GLM_Data_mod_3_4$sampleEvent)),
  new = 1:length(unique(aoi_GLM_Data_mod_3_4$sampleEvent))
)

aoi_GLM_Data_mod_3_4 <- aoi_GLM_Data_mod_3_4 %>%
  left_join(event_lookup, by = c("sampleEvent" = "old")) %>%
  mutate(sampleEvent_new = new) %>%
  select(-new)

event_year_3_4 <- tapply(aoi_GLM_Data_mod_3_4$year_id,
                         aoi_GLM_Data_mod_3_4$sampleEvent_new,
                         unique)
event_year_3_4 <- as.numeric(event_year_3_4)



aoi_GLM_Data_mod_3_4$SWA <- as.numeric(aoi_GLM_Data_mod_3_4$SWA)
```


----

----

Whittle data to just what you want to fit the model with:

    clusterID_seq
    SWA
    dam_m
    TO
    riparian_area_m2
    catchment_km2
    stream
    year_id
    sampleEvent_new

```{r,echo=FALSE}
RF_dat <- aoi_GLM_Data_mod_3_4 |> 
  select(SWA,
         dam_m,
         TO,
         riparian_area_m2,
         catchment_km2,
         stream,
         year_id,
         sampleEvent_new,
         clusterID_seq) 

RF_dat$stream           <- as.factor(RF_dat$stream)
RF_dat$year_id          <- as.factor(RF_dat$year_id)
RF_dat$sampleEvent_new  <- as.factor(RF_dat$sampleEvent_new)
RF_dat$clusterID_seq    <- as.factor(RF_dat$clusterID_seq)


str(RF_dat)
```
----

----


RF MODELING
----

Find any 'NA' data, and impute values using median/proximity 
```{r, echo=FALSE}
sum(!complete.cases(RF_dat))
```

Great! No NA values means no need for data imputation prior to modeling.

Fit RF model to data:
```{r,echo=FALSE}


set.seed(12345)

# Fit RF model using package "ranger"
RF_model <- ranger(
  formula = SWA ~ .,
  data = RF_dat,
  importance = "permutation",
  num.trees = 500
)

# Display model summary
RF_model

# RMSE from OOB error
sqrt(RF_model$prediction.error)

# Variable importance
barplot(RF_model$variable.importance,
        las = 2,
        main = "Permutation Importance")

# Partial dependence plots
# Partial dependence plots
TO_pdp <- pdp::partial(RF_model, pred.var = "TO", train = RF_dat)
ggplot(TO_pdp, aes(TO, yhat)) + geom_line() + theme_bw()

RIPA_pdp <- pdp::partial(RF_model, pred.var = "riparian_area_m2", train = RF_dat)
ggplot(RIPA_pdp, aes(riparian_area_m2, yhat)) + geom_line() + theme_bw()

UDA_pdp <- pdp::partial(RF_model, pred.var = "catchment_km2", train = RF_dat)
ggplot(UDA_pdp, aes(catchment_km2, yhat)) + geom_line() + theme_bw()

damDens_pdp <- pdp::partial(RF_model, pred.var = "dam_m", train = RF_dat)
ggplot(damDens_pdp, aes(dam_m, yhat)) + geom_line() + theme_bw()

```




```{r,echo=FALSE}
# 2D interaction PDP
pdp_inter_TO_dens <- pdp::partial(
  RF_model,
  pred.var = c("TO", "dam_m"),
  train = RF_dat,
  grid.resolution = 30
)

# Plot interaction surface
ggplot(pdp_inter_TO_dens, aes(TO, dam_m, fill = yhat)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    x = "TO",
    y = "Dam Density (dams/km)",
    fill = "Predicted SWA",
    title = "Interaction Surface: SWA ~ TO × DamDensity"
  ) +
  theme_bw() +
  theme(text = element_text(size = 14))

```


```{r,echo=FALSE}
# 2D interaction PDP
pdp_inter_RIPA_dens <- pdp::partial(
  RF_model,
  pred.var = c("riparian_area_m2", "dam_m"),
  train = RF_dat,
  grid.resolution = 30
)

# Plot interaction surface
ggplot(pdp_inter_RIPA_dens, aes(riparian_area_m2, dam_m, fill = yhat)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    x = "Riprarian Area (m^2)",
    y = "Dam Density (dams/km)",
    fill = "Predicted SWA",
    title = "Interaction Surface: SWA ~ RiparianArea × DamDensity"
  ) +
  theme_bw() +
  theme(text = element_text(size = 14))

```

```{r,echo=FALSE}
# 2D interaction PDP
pdp_inter_UDA_dens <- pdp::partial(
  RF_model,
  pred.var = c("catchment_km2", "dam_m"),
  train = RF_dat,
  grid.resolution = 30
)

# Plot interaction surface
ggplot(pdp_inter_UDA_dens, aes(catchment_km2, dam_m, fill = yhat)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    x = "Upstream Drainage Area (km^2)",
    y = "Dam Density (dams/km)",
    fill = "Predicted SWA",
    title = "Interaction Surface: SWA ~ DrainageArea × DamDensity"
  ) +
  theme_bw() +
  theme(text = element_text(size = 14))
```


```{r,echo=FALSE}
# 2D interaction PDP
pdp_inter_stream_dens <- pdp::partial(
  RF_model,
  pred.var = c("stream", "dam_m"),
  train = RF_dat,
  grid.resolution = 30
)

pdp_inter_stream_dens2 <- pdp_inter_stream_dens %>%
  left_join(stream_lookup, by = "stream")


# Plot interaction surface
ggplot(pdp_inter_stream_dens2, aes(watershed, dam_m, fill = yhat)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    x = "Watershed",
    y = "Dam Density (dams/km)",
    fill = "Predicted SWA",
    title = "Interaction Surface: SWA ~ Watershed × DamDensity"
  ) +
  theme_bw() +
  theme(text = element_text(size = 14))
```



```{r,echo=FALSE}
RF_model <- ranger(
  SWA ~ .,
  data = RF_dat,
  importance = "impurity",
  num.trees = 500
)

barplot(RF_model$variable.importance,
        las = 2,
        main = "Impurity Importance")

```