---
title: "SW_Hurdle_Models"
author: "Matt Webster"
date: "2026-07-28"
output: html_document
editor_options: 
  chunk_output_type: console
---


The purpose of this RMarkdown is to implement hurdle models for SW ~ damPresence, & SW ~ damDensity (with and without geomorphic covariates)

4 models total.

    WAIC suggests that models without geomorphic covariates (m1 & m3) perform better than those
    with geomorphic covariates (m2 & m4), and that dam presence models perform better than their
    dam density model counterparts. The WAIC metric is concerned with out-of-bag predictive 
    capabilities of a model set. Since we are not aiming for this to predict SW~beaver relationships
    outside of our study, and the exclusion of geomorhpic covariates leads to the inflation of 
    estimated dam effects, I think it's best to gain inference from model 4, as long as parameters
    are converging relatively well and making good sense.
    



```{r,echo=FALSE,}
suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(ggplot2)
  library(dataRetrieval)
  library(knitr)
  library(ggplot2)
  library(zoo) #for moving window
  library(stringr)
  library(vioplot)
  library(purrr)
  library(lavaan)
  library(blavaan)
  library(ggcorrplot)
  library(lme4)
  library(broom.mixed)
  library(coda)
  library(scales)
  library(loo)
})
```


----

**Hurdle HGLMM**


    This analysis explores late summner surface water presence and area in relation to beaver dam presence,
    & density.
    
    All 4 models presented herein are constructed in a hurdle model framework with nested random slopes for 
    beaver dam effect to allow for stream, year, and sampleEvent to cause deviation from the
    global mean beaver effect. Random intercepts are also included for stream, event, and
    cluster ID.
    
    
    Pt1: Surface water presence/absence; logistic regression
    
    Pt2: Surface water amount, conditional on presence; log(Normal) regression
    
    Uses log(SWA) for for 6 watersheds, 3 years, 4 sample periods (Aug-Sep) per year.


----

**Final Data:**

```{r,echo=FALSE}
aoi_HurdleData <- read.csv("aoi_HurdleData_091026Update.csv") %>% 
  select(!"X")


# add year/event index
event_year_3_4 <- tapply(aoi_HurdleData$year_id,
                         aoi_HurdleData$sampleEvent_new,
                         unique)
event_year_3_4 <- as.numeric(event_year_3_4)


# Full Hurdle data
N <- nrow(aoi_HurdleData)
y <- aoi_HurdleData$y
y2 <- aoi_HurdleData$logSWA
stream <- aoi_HurdleData$stream
year <- aoi_HurdleData$year_id
event <- aoi_HurdleData$sampleEvent_new
cluster <- as.numeric(as.factor(aoi_HurdleData$clusterID_unique))
damPres <- aoi_HurdleData$damPresence
damDen <- as.numeric(aoi_HurdleData$dam_km_std)
damDen_raw <- aoi_HurdleData$dam_km

n_stream <- length(unique(aoi_HurdleData$stream))
n_year <- length(unique(aoi_HurdleData$year_id))
n_event <- length(unique(aoi_HurdleData$sampleEvent_new))
n_cluster <- length(unique(as.numeric(as.factor(aoi_HurdleData$clusterID_unique))))

str(aoi_HurdleData)
```

----

For surface water models, data of interest include:

    - y: surface water presence (0,1); response variable for logistic portion of hurdle model
    
    - y2: log-transformed surface water area log(SWA[i]+1); response variable for HGLMM portion of hurdle model
    
    - stream: index for random intercept on SWA and random slope on dam effect
    
    - year: index for random slope on dam effect
    
    - sampleEvent: index for random intercept on SWA and random slope on dam effect 
      (nested within year; represents each 2 week sample period for SW layer )
    
    - clusterID_seq: index for random intercept on SWA; represents sample pairing
    
    - damPres: dam presence (0,1); fixed effect on SWA
    
    - damDen: dams/km; z-standardized ("...$dam_km_std"); fixed effect on SWA
    
    Only used in models 2 and 4:
    
    - incis: channel incision: 1-(TO/90) (inverse of topographic openness); z-standardized; fixed effect on SWA
    
    - RIPA: riparian area (m^2); z-standardized; fixed effect on SWA 
    
    - UDA: upstream drainage area (km ^2); z-standardized; fixed effect on SWA


----

**Overall Takeaways**

    - Estimates may be notably inflated when geomorphic conditions are not taken into consideration (see m1 vs m2). The following regards models 2 and 4, which estimate the effect of damPresence or damDensity on surface water, conditional on geomorphic covariates:
    
    - Dam presence has very little influence on surface water presence/absence. Surface water presence is instead
      driven by local geomorphic characteristics and random effects of stream/year/season.
      
    - Likewise, 1 standard deviation increase in damDensity corresponds with a negligible increase in the odds of surface water presence.
    
    - When surface water is present, presence of beaver dam(s) is associated with a 25% increase in surface water area from baseline conditions, although there is considerable variation among watersheds and time periods.
    
    - Where surface water is present, 1 standard deviation increase in damDensity corresponds with a modest 
      increase in surface water area (15%). Higher dam densities (3-4 SD increase) do notably 
      increase surface water area relative to baseline conditions (upwards of ~50% increase in SWA). 

    - Of course there is A LOT of variation in these patterns by stream, year, and sampling event.


----

----

----

----









SW ~ Dam Presence
----

----

----

**Model 1: SWA ~ DamPresence**

Effect of damPresence without geomorphic covariates


```{r,include=FALSE} 
suppressPackageStartupMessages({
library(jagsUI)
})
```
    
    sink("HGLMM_Hurdle_m1_SW_damPres.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR SW PRESENCE (1/2 Normal on SDs):
    
      sigma_zeta_stream  ~ dnorm(0,1) T(0,)
      tau_zeta_stream    <- pow(sigma_zeta_stream, -2)
    
      sigma_zeta_year    ~ dnorm(0,1) T(0,)
      tau_zeta_year      <- pow(sigma_zeta_year, -2)
    
      sigma_zeta_event   ~ dnorm(0,1) T(0,)
      tau_zeta_event     <- pow(sigma_zeta_event, -2)
    
      sigma_iota_stream  ~ dnorm(0,1) T(0,)
      tau_iota_stream    <- pow(sigma_iota_stream, -2)
      
      sigma_iota_year    ~ dnorm(0,1) T(0,)
      tau_iota_year      <- pow(sigma_iota_year, -2)
    
      sigma_iota_event   ~ dnorm(0,1) T(0,)
      tau_iota_event     <- pow(sigma_iota_event, -2)
    
      sigma_iota_cluster ~ dnorm(0,1) T(0,)
      tau_iota_cluster   <- pow(sigma_iota_cluster, -2)
    
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 1) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 1) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 1) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
      
      sigma_eta_year     ~ dnorm(0, 1) T(0,)
      tau_eta_year       <- pow(sigma_eta_year, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Presence/absence:
      epsilon ~ dnorm(0,0.04)
      zeta_damPres ~ dnorm(0,0.04)
    
      # Amount:
      alpha ~ dnorm(3.57,0.04)    # SWA intercept within 5 SD of empirical mean
      beta_damPres ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW PRESENCE:
    
      for(j in 1:n_stream){
        zeta_stream[j] ~ dnorm(0, tau_zeta_stream)
        iota_stream[j] ~ dnorm(0, tau_iota_stream)
      }
    
      for(yr in 1:n_year){
        zeta_year[yr] ~ dnorm(0, tau_zeta_year)
        iota_year[yr] ~ dnorm(0, tau_iota_year)
      }
    
      for(k in 1:n_event){
        zeta_event[k] ~ dnorm(zeta_year[event_year[k]], tau_zeta_event)
        iota_event[k] ~ dnorm(iota_year[event_year[k]], tau_iota_event)
      }
    
      for(c in 1:n_cluster){
        iota_cluster[c] ~ dnorm(0, tau_iota_cluster)
      }
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
        eta_year[yr] ~ dnorm(0, tau_eta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(eta_year[event_year[k]], tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # DETECTION PRIORS:
      
      for(k in 1:n_event){
      p_TP[k] ~ dnorm(0.76, pow(0.16,-2)) T(0,1)   # empirically derived recall
      p_FP[k] ~ dnorm(0.01, pow(0.005,-2)) T(0,1)  # empirically derived FPR
      }
      
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Presence/absence:
        z[i] ~ dbern(psi[i])
    
        logit(psi[i]) <- epsilon
                         + (zeta_damPres
                            + zeta_stream[stream[i]]
                            + zeta_year[year[i]]
                            + zeta_event[event[i]]) * damPres[i]
                         + iota_stream[stream[i]]
                         + iota_event[event[i]]
                         + iota_cluster[cluster[i]]
    
        # Detection:
        p_eff[i] <- z[i] * p_TP[event[i]] + (1-z[i]) * p_FP[event[i]]
        y[i]     ~ dbern(p_eff[i])
    
        # Amount:
        y2[i] ~ dnorm(mu[i], tau_eff[i])
    
        mu[i] <- z[i] *
                  (alpha
                    + (beta_damPres
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damPres[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]])
    
        tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
      
      ##########################################################################
                    
      # Derived parameter for WAIC computation:

      loglik[i] <- logdensity.norm(y[i], mu[i], tau)
      }
    
    }
    ", fill = TRUE)
    
    sink()
   
    

```{r,echo=FALSE}
hurdle_damPres_m1_data_list <- list(
  N = N,
  y = y,
  y2 = y2,
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damPres = damPres,
  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)),
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hurdle_damPres_m1_inits <- function(){ # let jags figure it out.
  list()
}

hurdle_damPres_m1_params <- c(
  "psi",
  "mu",
  "epsilon",
  "alpha",
  "zeta_damPres",
  "zeta_incis",
  "zeta_RIPA",
  "zeta_UDA",
  "zeta_stream",
  "zeta_year",
  "zeta_event",
  "iota_stream",
  "iota_event",
  "iota_year",
  "iota_cluster",
  "beta_damPres",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_year",
  "eta_cluster",
  "sigma",
  "sigma_zeta_stream",
  "sigma_zeta_year",
  "sigma_zeta_event",
  "sigma_iota_stream",
  "sigma_iota_event",
  "sigma_iota_cluster",
  "sigma_iota_year",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_year",
  "sigma_eta_event",
  "sigma_eta_cluster",
  "p_TP",
  "p_FP",
  "z",
  "tau_eff",
  "loglik"
)

nchains <- 3
nthin <- 5
niter <- 100000
nburn <- 25000

set.seed(12345)

Sys.time()

hurdle_damPres_m1_SW <- jags(data = hurdle_damPres_m1_data_list,
                   inits = hurdle_damPres_m1_inits,
                   parameters.to.save = hurdle_damPres_m1_params,
                   model.file = "HGLMM_Hurdle_m1_SW_damPres.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()
```

<!---
----

DIAGNOSTICS M1:
----

```{r,include=FALSE}
#Extract posteriors
damPres_m1_post <- hurdle_damPres_m1_SW$sims.list
```

**Assess "MCMC chain health":**

**Rubin-Gelman Rhat** -- assess convergence

**Effective Sample Size** -- assess posterior independence

```{r,include=FALSE}
# Choose only the parameters you care about
params_m1 <- c(
  "epsilon",
  "alpha",
  "zeta_damPres",
  "beta_damPres",
  "sigma",
  "sigma_zeta_stream",
  "sigma_beta_stream",
  "sigma_eta_cluster",
  "sigma_iota_cluster",
  "p_TP[1]",
  "p_FP[1]")

# Gelman–Rubin convergence diagnostic
gelman.diag(hurdle_damPres_m1_SW$samples[, params_m1], multivariate = FALSE)

# Effective sample size
effectiveSize(hurdle_damPres_m1_SW$samples[, params_m1])
```



    Good Rhat values, all around 1.00.
    
    Mostly high effective sample sizes.
    
    Important to anchor alpha_prior around the empirical mean, this helps things stabilize.
     


----

**Check traceplots for critical parameters**

```{r,include=FALSE}
par(mfrow = c(3, 4),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


tplot_m1_params <- c("epsilon",
  "alpha",
  "zeta_damPres",
  "beta_damPres",
  "sigma",
  "sigma_zeta_stream",
  "sigma_beta_stream",
  "sigma_eta_cluster",
  "sigma_iota_cluster",
  "p_TP[1]",
  "p_FP[1]")


m1_mcmc_sub <- hurdle_damPres_m1_SW$samples[, tplot_m1_params]

coda::traceplot(m1_mcmc_sub)
```


    Traceplots are all grassy. Note, sigma_eta_cluster is slow to converge. However, chains are sufficiently overlapping, and Rhat/ESS looks acceptable.

----

**Posterior predictive plots:**

Predicted Psi (m1_psi_post) vs observed SW Presence (y)

```{r,include=FALSE}

par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

# Posterior mean psi
m1_psi_post <- apply(hurdle_damPres_m1_SW$sims.list$psi, 2, mean)

# Compare to observed y
plot(m1_psi_post, y, pch=16, col=rgb(0,0,0,0.2))


```


Predicted log(SWA) (m1_mu_post) vs detected log(SWA) (y2)

Goal is 1:1 (ish)

```{r,include=FALSE}
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

# Posterior mean of predicted log(VA + 1)
m1_mu_SW_post <- apply(hurdle_damPres_m1_SW$sims.list$mu, 2, mean)

# Observed log vegetation area
m1_obs_logSW <- y2

# Fit regression on the log scale
m1_fit_log <- lm(m1_obs_logSW ~ m1_mu_SW_post)

m1_slope_log <- coef(m1_fit_log)[2]
m1_R2_log    <- summary(m1_fit_log)$r.squared

# Plot
plot(m1_mu_SW_post,
     m1_obs_logSW,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated log(SWA + 1)",
     ylab = "Observed log(SWA + 1)")

abline(0, 1, col = "red", lwd = 2)

```

Slope:
```{r,include=FALSE}
m1_slope_log
```

    SLOPE = 1 is ideal. 1.008 is pretty dang close. Shows that the model is well fit to the data.

----

**Random effects shrinkage: **

Goal is near 0(ish)

1. zeta_stream; random effect of stream on sw presence
2. beta_stream; random effect of stream on sw amount
3. zeta_event; random effect of sample event on sw presence
4. beta_event; random effect of sample event on sw amount

```{r, include=FALSE, }
par(mfrow = c(3,5),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

boxplot(hurdle_damPres_m1_SW$sims.list$iota_stream,
        main = "iota_stream")
boxplot(hurdle_damPres_m1_SW$sims.list$eta_stream,
        main = "eta_stream")
boxplot(hurdle_damPres_m1_SW$sims.list$zeta_stream,
        main = "zeta_stream")
boxplot(hurdle_damPres_m1_SW$sims.list$beta_stream,
        main = "beta_stream")
boxplot(hurdle_damPres_m1_SW$sims.list$zeta_year,
        main = "zeta_year")
boxplot(hurdle_damPres_m1_SW$sims.list$beta_year,
        main = "beta_year")
boxplot(hurdle_damPres_m1_SW$sims.list$iota_event,
        main = "iota_event")

boxplot(hurdle_damPres_m1_SW$sims.list$eta_event,
        main = "eta_event")

boxplot(hurdle_damPres_m1_SW$sims.list$zeta_event,
        main = "zeta_event")
boxplot(hurdle_damPres_m1_SW$sims.list$beta_event,
        main = "beta_event")


```

----


**Posterior mean dam effects:**

damPres -> swPres (logistic regression)

```{r,include=FALSE}
m1_coeffparams <- c(#"alpha",
            "zeta_damPres",
            "beta_damPres",
            "sigma")

m1_coef_df <- hurdle_damPres_m1_SW$summary[m1_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m1_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.2,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```

```{r,include=FALSE}
summary(hurdle_damPres_m1_SW$sims.list$zeta_damPres)

exp(mean(hurdle_damPres_m1_SW$sims.list$zeta_damPres))
```

```{r,include=FALSE}
#library(tidyverse)
#library(scales)
par(mfrow = c(1, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

#Posterior distribution of damEffect on vegArea:

hist(hurdle_damPres_m1_SW$sims.list$zeta_damPres,
     breaks = 50,
     col = "steelblue",
     main = "Posterior of damPresence effect on logit(Psi)",
     xlab = "beta_damPres")
abline(v = mean(hurdle_damPres_m1_SW$sims.list$zeta_damPres), col = "red", lwd = 2)
```

    When you don't account for geomorphic conditions, dam presence corresponds to an 
    estimated 40-50% increase in the odds of surface‑water presence.
    
----


damPres -> swAmount (hglmm)

```{r,include=FALSE}
summary(hurdle_damPres_m1_SW$sims.list$beta_damPres)

exp(mean(hurdle_damPres_m1_SW$sims.list$beta_damPres))
```

```{r,include=FALSE}
#library(tidyverse)
#library(scales)
par(mfrow = c(1, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

#Posterior distribution of damEffect on vegArea:

hist(hurdle_damPres_m1_SW$sims.list$beta_damPres,
     breaks = 50,
     col = "steelblue",
     main = "Posterior of damPresence effect on log(SWA)",
     xlab = "beta_damPres")
abline(v = mean(hurdle_damPres_m1_SW$sims.list$beta_damPres), col = "red", lwd = 2)
```
    
    When you don't account for geomorphic conditions, dam presence corresponds to an 
    estimated 30-35% increase in surface water area.


----



**Prior-posterior overlap:**

```{r,include=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 


damPres_m1_alpha_post <- hurdle_damPres_m1_SW$sims.list$alpha
dampres_m1_epsilon_post <- hurdle_damPres_m1_SW$sims.list$epsilon
damPres_m1_beta_post  <- hurdle_damPres_m1_SW$sims.list$beta_damPres
damPres_m1_zeta_post  <- hurdle_damPres_m1_SW$sims.list$zeta_damPres
damPres_m1_sigBetaStream <- hurdle_damPres_m1_SW$sims.list$sigma_beta_stream

damPres_m1Mod_n <- length(damPres_m1_alpha_post)

damPres_m1_alpha_prior <- rnorm(damPres_m1Mod_n, 3.57, sqrt(1/0.04))
dampres_m1_epsilon_prior <- rnorm(damPres_m1Mod_n, 0, sqrt(1/0.04))
damPres_m1_beta_prior  <- rnorm(damPres_m1Mod_n, 0, sqrt(1/0.04))
damPres_m1_zeta_prior  <- rnorm(damPres_m1Mod_n, 0, sqrt(1/0.04))
damPres_m1_sigBetaStream_prior <- pmax(rnorm(damPres_m1Mod_n, 0, 1), 0)


# Plot overlap:

plot_prior_post <- function(prior, post, name) {
  
  df <- tibble(
    value = c(prior, post),
    dist  = rep(c("Prior", "Posterior"), each = length(prior))
  )
  
  ggplot(df, aes(x = value, fill = dist)) +
    geom_density(alpha = 0.4) +
    scale_fill_manual(values = c("gray60", "steelblue")) +
    labs(
      title = paste("Prior vs Posterior for", name),
      x = name,
      y = "Density"
    ) +
    theme_minimal(base_size = 14)
}

plot_prior_post(damPres_m1_alpha_prior, damPres_m1_alpha_post, "alpha_m1")
plot_prior_post(dampres_m1_epsilon_prior, dampres_m1_epsilon_post, "epsilon_m1")
plot_prior_post(damPres_m1_beta_prior, damPres_m1_beta_post, "beta_damPres_m1")
plot_prior_post(damPres_m1_zeta_prior, damPres_m1_zeta_post, "zeta_damPres_m1")
plot_prior_post(damPres_m1_sigBetaStream_prior, damPres_m1_sigBetaStream, "sigma_beta_stream_m1")

```

    These PPO plots indicate that the priors are reasonable, but are not driving the model -- 
    the data are sufficiently informing the model, and shifting the posterior distributions 
    away from those prior expectations


----

----

----

Now, include geomorphic covariates:

----

----

----
--->

**Model 2: SWA ~ DamPresence + Geom.Covar.**


```{r,include=FALSE}
#library(jagsUI)
```
    sink("HGLMM_Hurdle_m2_SW_damPres.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR SW PRESENCE (1/2 Normal on SDs):
    
      sigma_zeta_stream  ~ dnorm(0,1) T(0,)
      tau_zeta_stream    <- pow(sigma_zeta_stream, -2)
    
      sigma_zeta_year    ~ dnorm(0,1) T(0,)
      tau_zeta_year      <- pow(sigma_zeta_year, -2)
    
      sigma_zeta_event   ~ dnorm(0,1) T(0,)
      tau_zeta_event     <- pow(sigma_zeta_event, -2)
    
      sigma_iota_stream  ~ dnorm(0,1) T(0,)
      tau_iota_stream    <- pow(sigma_iota_stream, -2)
      
      sigma_iota_year    ~ dnorm(0,1) T(0,)
      tau_iota_year      <- pow(sigma_iota_year, -2)
    
      sigma_iota_event   ~ dnorm(0,1) T(0,)
      tau_iota_event     <- pow(sigma_iota_event, -2)
    
      sigma_iota_cluster ~ dnorm(0,1) T(0,)
      tau_iota_cluster   <- pow(sigma_iota_cluster, -2)
    
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 1) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 1) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 1) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
      
      sigma_eta_year     ~ dnorm(0,1) T(0,)
      tau_eta_year     <- pow(sigma_eta_year, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Presence/absence:
      epsilon ~ dnorm(0,0.04)
      zeta_damPres ~ dnorm(0,0.04)
      zeta_incis ~ dnorm(0,0.04)
      zeta_RIPA ~ dnorm(0,0.04)
      zeta_UDA ~ dnorm(0,0.04)
    
      # Amount:
      alpha ~ dnorm(3.57,0.04)   #anchor alpha within 5 SD of empirical mean logSWA
      beta_damPres ~ dnorm(0,0.04)
      beta_incis ~ dnorm(0,0.04)
      beta_RIPA ~ dnorm(0,0.04)
      beta_UDA ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW PRESENCE:
    
      for(j in 1:n_stream){
        zeta_stream[j] ~ dnorm(0, tau_zeta_stream)
        iota_stream[j] ~ dnorm(0, tau_iota_stream)
      }
    
      for(yr in 1:n_year){
        zeta_year[yr] ~ dnorm(0, tau_zeta_year)
        iota_year[yr] ~ dnorm(0, tau_iota_year)
      }
    
      for(k in 1:n_event){
        zeta_event[k] ~ dnorm(zeta_year[event_year[k]], tau_zeta_event)
        iota_event[k] ~ dnorm(iota_year[event_year[k]], tau_iota_event)
      }
    
      for(c in 1:n_cluster){
        iota_cluster[c] ~ dnorm(0, tau_iota_cluster)
      }
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
        eta_year[yr] ~ dnorm(0, tau_eta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(eta_year[event_year[k]], tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # DETECTION PRIORS:
      
      for(k in 1:n_event){
      p_TP[k] ~ dnorm(0.76, pow(0.16,-2)) T(0,1)   # empirically derived recall
      p_FP[k] ~ dnorm(0.01, pow(0.005,-2)) T(0,1)  # empirically derived FPR
      }
      
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Presence/absence:
        z[i] ~ dbern(psi[i])
    
        logit(psi[i]) <- epsilon
                         + (zeta_damPres
                            + zeta_stream[stream[i]]
                            + zeta_year[year[i]]
                            + zeta_event[event[i]]) * damPres[i]
                         + zeta_incis * incis[i]
                         + zeta_RIPA * RIPA[i]
                         + zeta_UDA * UDA[i]
                         + iota_stream[stream[i]]
                         + iota_event[event[i]]
                         + iota_cluster[cluster[i]]
    
        # Detection:
        p_eff[i] <- z[i] * p_TP[event[i]] + (1-z[i]) * p_FP[event[i]]
        y[i]     ~ dbern(p_eff[i])
    
        # Amount:
        y2[i] ~ dnorm(mu[i], tau_eff[i])
    
        mu[i] <- z[i] *
                  (alpha
                    + (beta_damPres
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damPres[i]
                    + beta_incis * incis[i] 
                    + beta_RIPA * RIPA[i]
                    + beta_UDA * UDA[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]])
    
        tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
        
      ##########################################################################
                    
      # Derived parameter for WAIC computation:

      loglik[i] <- logdensity.norm(y[i], mu[i], tau)
      }
    
    }
    ", fill = TRUE)
    
    sink()
  
    

```{r,include=FALSE}
hurdle_damPres_m2_data_list <- list(
  N = N,
  y = y,
  y2 = y2,
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damPres = damPres,
  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)),
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hurdle_damPres_m2_inits <- function(){
  list(
    alpha        = rnorm(1, 0, 1),
    epsilon      = rnorm(1, 0, 1),

    sigma        = runif(1, 0.1, 2),
    sigma_zeta_stream  = runif(1, 0.1, 2),
    sigma_zeta_year    = runif(1, 0.1, 2),
    sigma_zeta_event   = runif(1, 0.1, 2),
    sigma_iota_stream  = runif(1, 0.1, 2),
    sigma_iota_event   = runif(1, 0.1, 2),
    sigma_iota_cluster = runif(1, 0.1, 2),

    sigma_beta_stream  = runif(1, 0.1, 2),
    sigma_beta_year    = runif(1, 0.1, 2),
    sigma_beta_event   = runif(1, 0.1, 2),
    sigma_eta_stream   = runif(1, 0.1, 2),
    sigma_eta_event    = runif(1, 0.1, 2),
    sigma_eta_cluster  = runif(1, 0.1, 2)
  )
}

hurdle_damPres_m2_params <- c(
  "psi",
  "mu",
  "epsilon",
  "alpha",
  "zeta_damPres",
  "zeta_incis",
  "zeta_RIPA",
  "zeta_UDA",
  "zeta_stream",
  "zeta_year",
  "zeta_event",
  "iota_stream",
  "iota_event",
  "iota_cluster",
  "beta_damPres",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_zeta_stream",
  "sigma_zeta_year",
  "sigma_zeta_event",
  "sigma_iota_stream",
  "sigma_iota_event",
  "sigma_iota_year",
  "sigma_iota_cluster",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_event",
  "sigma_eta_year",
  "sigma_eta_cluster",
  "p_TP",
  "p_FP",
  "z",
  "tau_eff",
  "loglik"
)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000

set.seed(12345)

Sys.time()

hurdle_damPres_m2_SW <- jags(data = hurdle_damPres_m2_data_list,
                   inits = hurdle_damPres_m2_inits,
                   parameters.to.save = hurdle_damPres_m2_params,
                   model.file = "HGLMM_Hurdle_m2_SW_damPres.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hurdle_damPres_m2_SW) # model with local and watershed dam density as fixed effects
#print(hurdle_damPres_m2_SW$summary[grep("beta_", rownames(hurdle_damPres_m2_SW$summary)), ])
#print(hurdle_damPres_m2_SW$summary[grep("zeta_", rownames(hurdle_damPres_m2_SW$summary)), ])

```

<!---
----

DIAGNOSTICS M2:
----


```{r,include=FALSE}
#Extract posteriors
damPres_m2_post <- hurdle_damPres_m2_SW$sims.list
```

**Assess "MCMC chain health":**

**Rubin-Gelman Rhat** -- assess convergence

**Effective Sample Size** -- assess posterior independence

```{r,include=FALSE}
# Choose only the parameters you care about
params <- c(
  "epsilon",
  "alpha",
  "zeta_damPres",
  "beta_damPres",
  "zeta_incis",
  "beta_incis",
  "zeta_RIPA",
  "beta_RIPA",
  "zeta_UDA",
  "beta_UDA",
  "sigma",
  "sigma_zeta_stream",
  "sigma_iota_cluster",
  "p_TP[1]",
  "p_FP[1]")

# Gelman–Rubin convergence diagnostic
gelman.diag(hurdle_damPres_m2_SW$samples[, params], multivariate = FALSE)

# Effective sample size
effectiveSize(hurdle_damPres_m2_SW$samples[, params])

```



    Decent Rhat values, all around 1.00.
    
    Mostly high effective sample sizes. 
    
    Slow convergence for beta_damPres (means it's harder to estimate)


----

**Check traceplots for critical parameters**

```{r,include=FALSE}

par(mfrow = c(3, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


m2_tplot_params <- c("alpha","beta_damPres","zeta_damPres", "sigma","sigma_zeta_stream","sigma_beta_stream")


m2_mcmc_sub <- hurdle_damPres_m2_SW$samples[, m2_tplot_params]

coda::traceplot(m2_mcmc_sub)
```


    Traceplots are all mostly grassy. Again, see zeta/beta_damPres. REAL slow to 
    converge; expected, given model structure.


----

**Posterior predictive plots:**

Predicted Psi (psi_post) vs observed SW Presence (y)

```{r,include=FALSE}

par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

# Posterior mean psi
psi_post <- apply(hurdle_damPres_m2_SW$sims.list$psi, 2, mean)

# Compare to observed y
plot(psi_post, y, pch=16, col=rgb(0,0,0,0.2))


```


Predicted log(SWA) (mu_post[y==1]) vs detected SWA (y2[y==1]) conditional on SW presence

Goal is 1:1 (ish)

```{r,include=FALSE}
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

# Posterior mean of predicted log(VA + 1)
m2_mu_SW_post <- apply(hurdle_damPres_m2_SW$sims.list$mu, 2, mean)

# Observed log vegetation area
m2_obs_logSW <- y2

# Fit regression on the log scale
m2_fit_log <- lm(m2_obs_logSW ~ m2_mu_SW_post)

m2_slope_log <- coef(m2_fit_log)[2]
m2_R2_log    <- summary(m2_fit_log)$r.squared

# Plot
plot(m2_mu_SW_post,
     m2_obs_logSW,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated log(SWA + 1)",
     ylab = "Observed log(SWA + 1)")

abline(0, 1, col = "red", lwd = 2)

```

Slope:
```{r,include=FALSE}
m2_slope_log
```


    1.007 indicates that the model-predicted log(SWA+1) space is very similar to the observed data.
----

**Random effects shrinkage: **

Goal is near 0(ish)

1. zeta_stream; random effect of stream on sw presence
2. beta_stream; random effect of stream on sw amount
3. zeta_event; random effect of sample event on sw presence
4. beta_event; random effect of sample event on sw amount

```{r, include=FALSE, }
par(mfrow = c(3,5),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

boxplot(hurdle_damPres_m2_SW$sims.list$iota_stream,
        main = "iota_stream")
boxplot(hurdle_damPres_m2_SW$sims.list$eta_stream,
        main = "eta_stream")
boxplot(hurdle_damPres_m2_SW$sims.list$zeta_stream,
        main = "zeta_stream")
boxplot(hurdle_damPres_m2_SW$sims.list$beta_stream,
        main = "beta_stream")
boxplot(hurdle_damPres_m2_SW$sims.list$zeta_year,
        main = "zeta_year")
boxplot(hurdle_damPres_m2_SW$sims.list$beta_year,
        main = "beta_year")
boxplot(hurdle_damPres_m2_SW$sims.list$iota_event,
        main = "iota_event")

boxplot(hurdle_damPres_m2_SW$sims.list$eta_event,
        main = "eta_event")

boxplot(hurdle_damPres_m2_SW$sims.list$zeta_event,
        main = "zeta_event")
boxplot(hurdle_damPres_m2_SW$sims.list$beta_event,
        main = "beta_event")


```


----


**Posterior mean dam effects:**


damPres -> swPres (logistic regression)

```{r,include=FALSE}
m2_coeffparams <- c(#"alpha",
            "zeta_damPres",
            "zeta_incis",
            "zeta_RIPA",
            "zeta_UDA",
            "beta_damPres",
            "beta_incis",
            "beta_RIPA",
            "beta_UDA",
            "sigma")

m2_coef_df <- hurdle_damPres_m2_SW$summary[m2_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m2_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.2,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```

```{r,include=FALSE}
summary(hurdle_damPres_m2_SW$sims.list$zeta_damPres)

exp(mean(hurdle_damPres_m2_SW$sims.list$zeta_damPres))
```

```{r,include=FALSE}
#library(tidyverse)
#library(scales)
par(mfrow = c(1, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

#Posterior distribution of damEffect on vegArea:

hist(hurdle_damPres_m2_SW$sims.list$zeta_damPres,
     breaks = 50,
     col = "steelblue",
     main = "Posterior of damPresence effect on logit(Psi)",
     xlab = "beta_damPres")
abline(v = mean(hurdle_damPres_m2_SW$sims.list$zeta_damPres), col = "red", lwd = 2)
```

    When you consider local geomorphic conditions, dam presence increases the odds of 
    surface‑water presence by about 6% (compared to 40-50% w/o geom.covar.). This 
    indicates that these geomorphic conditions are driving surface water presence, 
    moreso than dam presence.
    
----


damPres -> swAmount (hglmm)

```{r,include=FALSE}
summary(hurdle_damPres_m2_SW$sims.list$beta_damPres)

exp(mean(hurdle_damPres_m2_SW$sims.list$beta_damPres))
```

```{r,include=FALSE}
#library(tidyverse)
#library(scales)
par(mfrow = c(1, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

#Posterior distribution of damEffect on vegArea:

hist(hurdle_damPres_m2_SW$sims.list$beta_damPres,
     breaks = 50,
     col = "steelblue",
     main = "Posterior of damPresence effect on log(SWA)",
     xlab = "beta_damPres")
abline(v = mean(hurdle_damPres_m2_SW$sims.list$beta_damPres), col = "red", lwd = 2)
```
    
    When you account for geomorphic conditions, dam presence corresponds to an 
    estimated 20-25%% increase in surface water area. LOTS of variation.


----

**Prior-posterior overlap:**


```{r,include=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)


damPres_m2_alpha_post <- hurdle_damPres_m2_SW$sims.list$alpha
damPres_m2_beta_post  <- hurdle_damPres_m2_SW$sims.list$beta_damPres
damPres_m2_zeta_post  <- hurdle_damPres_m2_SW$sims.list$zeta_damPres
damPres_m2_sigBetaStream <- hurdle_damPres_m2_SW$sims.list$sigma_beta_stream

damPres_m2Mod_n <- length(damPres_m2_alpha_post)

damPres_m2_alpha_prior <- rnorm(damPres_m2Mod_n, 3.57, sqrt(1/0.04))
damPres_m2_beta_prior  <- rnorm(damPres_m2Mod_n, 0, sqrt(1/0.04))
damPres_m2_zeta_prior  <- rnorm(damPres_m2Mod_n, 0, sqrt(1/0.04))
damPres_m2_sigBetaStream_prior <- pmax(rnorm(damPres_m2Mod_n, 0, 1), 0)


# Plot overlap:

#plot_prior_post <- function(prior, post, name) {
  
#  df <- tibble(
#    value = c(prior, post),
#    dist  = rep(c("Prior", "Posterior"), each = length(prior))
#  )
  
#  ggplot(df, aes(x = value, fill = dist)) +
#    geom_density(alpha = 0.4) +
#    scale_fill_manual(values = c("gray60", "steelblue")) +
#    labs(
#      title = paste("Prior vs Posterior for", name),
#      x = name,
#      y = "Density"
#    ) +
#    theme_minimal(base_size = 14)
#}

plot_prior_post(damPres_m2_alpha_prior, damPres_m2_alpha_post, "alpha_m2")
plot_prior_post(damPres_m2_beta_prior, damPres_m2_beta_post, "beta_damPres_m2")
plot_prior_post(damPres_m2_zeta_prior, damPres_m2_zeta_post, "zeta_damPres_m2")
plot_prior_post(damPres_m2_sigBetaStream_prior, damPres_m2_sigBetaStream, "sigma_beta_stream_m2")

```

    PPOs still suggest data driven model. 


----


**Marginal effect** of dam presence on surface water presence, given geomorphic context.

**Note** there's a  TINY positive effect of dam presence on surface water presence. Not really perceptible from these plots. 

```{r,include=FALSE, warning=FALSE}
# Build a grid of raw geomorphic covariate values
damPres_m2_geom_grid <- tibble(
  TO_raw = seq(min(aoi_HurdleData$TO),
               max(aoi_HurdleData$TO), length.out = 50),

  incis_raw = 1 - (TO_raw / 90),   # raw incision scale

  RIPA_raw = seq(min(aoi_HurdleData$riparian_area_m2),
                 max(aoi_HurdleData$riparian_area_m2), length.out = 50),

  UDA_raw  = seq(min(aoi_HurdleData$catchment_km2),
                 max(aoi_HurdleData$catchment_km2), length.out = 50)
) %>%
  # Convert raw → standardized (to match model scale)
  mutate(
    incis = as.numeric(scale(incis_raw)),
    RIPA  = as.numeric(scale(RIPA_raw)),
    UDA   = as.numeric(scale(UDA_raw))
  )

damPres_m2_compute_psi <- function(dam_flag, geom_grid, damPres_m2_post) {

  eps <- mean(damPres_m2_post$epsilon)
  zd  <- mean(damPres_m2_post$zeta_damPres)
  zi  <- mean(damPres_m2_post$zeta_incis)
  zr  <- mean(damPres_m2_post$zeta_RIPA)
  zu  <- mean(damPres_m2_post$zeta_UDA)

  geom_grid %>%
    mutate(
      psi = plogis(
        eps +
        zd * dam_flag +
        zi * incis +      # standardized incision
        zr * RIPA +       # standardized RIPA
        zu * UDA          # standardized UDA
      ),
      dam = dam_flag
    )
}


```

```{r,include=FALSE}
damPres_m2_psi_nodam <- damPres_m2_compute_psi(0, damPres_m2_geom_grid, damPres_m2_post)
damPres_m2_psi_dam   <- damPres_m2_compute_psi(1, damPres_m2_geom_grid, damPres_m2_post)

damPres_m2_psi_df <- bind_rows(damPres_m2_psi_nodam, damPres_m2_psi_dam)
```

```{r,include=FALSE}
ggplot(damPres_m2_psi_df, aes(x = incis_raw, y = psi, color = factor(dam))) +
  geom_line(size = 1.2) +
  labs(
    title = "Marginal Effect of Dam Presence on Surface Water Presence",
    subtitle = "Predicted probability of water (psi) across raw incision values",
    x = "Channel Incision (raw scale: 1 - TO/90)",
    y = "Probability of Surface Water",
    color = "Dam Presence"
  ) +
  scale_color_manual(values = c("brown4","steelblue"),
                     labels = c("No Dam", "Dam Present")) +
  theme_minimal(base_size = 14)

```

```{r,include=FALSE}
ggplot(damPres_m2_psi_df, aes(x = RIPA_raw, y = psi, color = factor(dam))) +
  geom_line(size = 1.2) +
  labs(
    title = "Marginal Effect of Dam Presence on Surface Water Presence",
    subtitle = "Predicted psi across raw riparian area (m^2)",
    x = "Riparian Area (m^2)",
    y = "Probability of Surface Water",
    color = "Dam Presence"
  ) +
  scale_color_manual(values = c("brown4","steelblue"),
                     labels = c("No Dam", "Dam Present")) +
  theme_minimal(base_size = 14)


```

```{r,include=FALSE}
ggplot(damPres_m2_psi_df, aes(x = UDA_raw, y = psi, color = factor(dam))) +
  geom_line(size = 1.2) +
  labs(
    title = "Marginal Effect of Dam Presence on Surface Water Presence",
    subtitle = "Predicted psi across raw contributing drainage area (km^2)",
    x = "Contributing Drainage Area (km^2)",
    y = "Probability of Surface Water",
    color = "Dam Presence"
  ) +
  scale_color_manual(values = c("brown4","steelblue"),
                     labels = c("No Dam", "Dam Present")) +
  theme_minimal(base_size = 14)


```


----


**Partial dependence plots**

Constrain PDP to empirical ranges for SWA and target covariates

```{r,include=FALSE}

damPres_m2_post <- hurdle_damPres_m2_SW$sims.list

# Raw covariates
incis_raw <- 1 - aoi_HurdleData$TO/90
RIPA_raw  <- aoi_HurdleData$riparian_area_m2
UDA_raw   <- aoi_HurdleData$catchment_km2

logSWA_mean <- mean(aoi_HurdleData$logSWA)
logSWA_sd   <- sd(aoi_HurdleData$logSWA)

# Standardization stats
incis_mean_raw <- mean(incis_raw)
incis_sd_raw   <- sd(incis_raw)

RIPA_mean_raw  <- mean(RIPA_raw)
RIPA_sd_raw    <- sd(RIPA_raw)

UDA_mean_raw   <- mean(UDA_raw)
UDA_sd_raw     <- sd(UDA_raw)


##############################3

# Helper Functions to Compute SWA from posterior draws
compute_swa_draws <- function(incis, RIPA, UDA, post, dam_flag) {
  n_draws <- length(damPres_m2_post$alpha)
  swa_mat <- matrix(NA_real_, nrow = length(incis), ncol = n_draws)

  for (j in seq_len(n_draws)) {
    lp <- damPres_m2_post$alpha[j] +
          damPres_m2_post$beta_incis[j] * incis +
          damPres_m2_post$beta_RIPA[j]  * RIPA +
          damPres_m2_post$beta_UDA[j]   * UDA +
          damPres_m2_post$beta_damPres[j] * dam_flag

    swa_mat[, j] <- exp(lp) - 1
  }

  swa_mat
}

summarize_swa <- function(swa_mat) {
  tibble(
    SWA_med = apply(swa_mat, 1, median),
    SWA_lo  = apply(swa_mat, 1, quantile, probs = 0.025),
    SWA_hi  = apply(swa_mat, 1, quantile, probs = 0.975)
  )
}




#########################################

# PDP Builder:

build_pdp <- function(focal_raw, focal_name,
                      incis_raw, RIPA_raw, UDA_raw,
                      post,
                      incis_mean_raw, incis_sd_raw,
                      RIPA_mean_raw, RIPA_sd_raw,
                      UDA_mean_raw, UDA_sd_raw) {

  # Sequence only across empirical range
  focal_seq <- seq(min(focal_raw), max(focal_raw), length.out = 200)

  # Conditional medians of other covariates
  incis_med <- median(incis_raw)
  RIPA_med  <- median(RIPA_raw)
  UDA_med   <- median(UDA_raw)

  # Build grid (200 rows)
  grid <- tibble(
    focal_raw = focal_seq,
    incis_raw = if (focal_name == "incis") focal_seq else incis_med,
    RIPA_raw  = if (focal_name == "RIPA")  focal_seq else RIPA_med,
    UDA_raw   = if (focal_name == "UDA")   focal_seq else UDA_med
  ) %>%
    mutate(
      incis = (incis_raw - incis_mean_raw) / incis_sd_raw,
      RIPA  = (RIPA_raw  - RIPA_mean_raw)  / RIPA_sd_raw,
      UDA   = (UDA_raw   - UDA_mean_raw)   / UDA_sd_raw
    )

  # Compute SWA draws
  swa_nodam <- compute_swa_draws(grid$incis, grid$RIPA, grid$UDA, post, dam_flag = 0)
  swa_dam   <- compute_swa_draws(grid$incis, grid$RIPA, grid$UDA, post, dam_flag = 1)

  # Summaries (400 rows)
  pdp <- bind_rows(
    summarize_swa(swa_nodam) %>% mutate(dam = "No Dam"),
    summarize_swa(swa_dam)   %>% mutate(dam = "Dam Present")
  )

  # Repeat grid twice (400 rows)
  grid2 <- bind_rows(grid, grid)

  # Bind columns safely
  bind_cols(grid2, pdp)
}



#####

pdp_incis <- build_pdp(
  focal_raw = incis_raw,
  focal_name = "incis",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_m2_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)

pdp_RIPA <- build_pdp(
  focal_raw = RIPA_raw,
  focal_name = "RIPA",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_m2_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)

pdp_UDA <- build_pdp(
  focal_raw = UDA_raw,
  focal_name = "UDA",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_m2_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)



##########################################

# Plot Function:

plot_pdp <- function(pdp, xvar, xtitle) {
  ggplot(pdp,
         aes(x = .data[[xvar]], y = SWA_med, color = dam, fill = dam)) +
    geom_ribbon(aes(ymin = SWA_lo, ymax = SWA_hi), alpha = 0.2, color = NA) +
    geom_line(size = 1.2) +
    scale_color_manual(values = c("steelblue", "brown4")) +
    scale_fill_manual(values = c("steelblue", "brown4")) +
    scale_y_continuous(labels = comma) +
    labs(
      title = paste("Partial Dependence of SWA on", xtitle),
      subtitle = "Predicted SWA with 95% credible intervals\nConstrained to empirical covariate range",
      x = xtitle,
      y = "Predicted SWA (m²)",
      color = "Dam Presence",
      fill  = "Dam Presence"
    ) +
    theme_minimal(base_size = 14)
}


###########################################3


plot_pdp(pdp_incis, "incis_raw", "Channel Incision (raw)")
plot_pdp(pdp_RIPA,  "RIPA_raw",  "Riparian Area (m²)")
plot_pdp(pdp_UDA,   "UDA_raw",   "Contributing Drainage Area (km²)")


```

--->

----

----

----

----

----

----

----

----

----


SW ~ Dam Density
----


**Hurdle HGLMM**


**MODEL 3: SW ~ damDensity (no Geom.Covar.)**

```{r,include=FALSE}
#library(jagsUI)
```
    sink("HGLMM_Hurdle_m3_SW_damDen.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR SW PRESENCE (1/2 Normal on SDs):
    
      sigma_zeta_stream  ~ dnorm(0,1) T(0,)
      tau_zeta_stream    <- pow(sigma_zeta_stream, -2)
    
      sigma_zeta_year    ~ dnorm(0,1) T(0,)
      tau_zeta_year      <- pow(sigma_zeta_year, -2)
    
      sigma_zeta_event   ~ dnorm(0,1) T(0,)
      tau_zeta_event     <- pow(sigma_zeta_event, -2)
    
      sigma_iota_stream  ~ dnorm(0,1) T(0,)
      tau_iota_stream    <- pow(sigma_iota_stream, -2)
      
      sigma_iota_year    ~ dnorm(0,1) T(0,)
      tau_iota_year      <- pow(sigma_iota_year, -2)
    
      sigma_iota_event   ~ dnorm(0,1) T(0,)
      tau_iota_event     <- pow(sigma_iota_event, -2)
    
      sigma_iota_cluster ~ dnorm(0,1) T(0,)
      tau_iota_cluster   <- pow(sigma_iota_cluster, -2)
    
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 4) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 4) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 4) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
      
      sigma_eta_year     ~ dnorm(0, 1) T(0,)
      tau_eta_year       <- pow(sigma_eta_year, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Presence/absence:
      epsilon ~ dnorm(0,0.04)
      zeta_damDen ~ dnorm(0,0.04)
    
      # Amount:
      alpha ~ dnorm(3.57,0.04)  # within 5 SD of empirical mean logSWA
      beta_damDen ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW PRESENCE:
    
      for(j in 1:n_stream){
        zeta_stream[j] ~ dnorm(0, tau_zeta_stream)
        iota_stream[j] ~ dnorm(0, tau_iota_stream)
      }
    
      for(yr in 1:n_year){
        zeta_year[yr] ~ dnorm(0, tau_zeta_year)
        iota_year[yr] ~ dnorm(0, tau_iota_year)
      }
    
      for(k in 1:n_event){
        zeta_event[k] ~ dnorm(zeta_year[event_year[k]], tau_zeta_event)
        iota_event[k] ~ dnorm(iota_year[event_year[k]], tau_iota_event)
      }
    
      for(c in 1:n_cluster){
        iota_cluster[c] ~ dnorm(0, tau_iota_cluster)
      }
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
        eta_year[yr] ~ dnorm(0, tau_eta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(eta_year[event_year[k]], tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # DETECTION PRIORS:
      
      for(k in 1:n_event){
      p_TP[k] ~ dnorm(0.76, pow(0.16,-2)) T(0,1)   # empirically derived recall
      p_FP[k] ~ dnorm(0.01, pow(0.005,-2)) T(0,1)  # empirically derived FPR
      }
      
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Presence/absence:
        z[i] ~ dbern(psi[i])
    
        logit(psi[i]) <- epsilon
                         + (zeta_damDen
                            + zeta_stream[stream[i]]
                            + zeta_year[year[i]]
                            + zeta_event[event[i]]) * damDen[i]
                         + iota_stream[stream[i]]
                         + iota_event[event[i]]
                         + iota_cluster[cluster[i]]
    
        # Detection:
        p_eff[i] <- z[i] * p_TP[event[i]] + (1-z[i]) * p_FP[event[i]]
        y[i]     ~ dbern(p_eff[i])
    
        # Amount:
        y2[i] ~ dnorm(mu[i], tau_eff[i])
    
        mu[i] <- z[i] *
                  (alpha
                    + (beta_damDen
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damDen[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]])
    
        tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
        
      ##########################################################################
                    
      # Derived parameter for WAIC computation:

      loglik[i] <- logdensity.norm(y[i], mu[i], tau)
      }
    
    }
    ", fill = TRUE)
    
    sink()
  
    

```{r,include=FALSE}
hurdle_damDen_m3_data_list <- list(
  N = N,
  y = y,
  y2 = y2,
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damDen = damDen,
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hurdle_damDen_m3_inits <- function(){
  list()
}

hurdle_damDen_m3_params <- c(
  "psi",
  "mu",
  "epsilon",
  "alpha",
  "zeta_damDen",
  "zeta_stream",
  "zeta_year",
  "zeta_event",
  "iota_stream",
  "iota_year",
  "iota_event",
  "iota_cluster",
  "beta_damDen",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_year",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_zeta_stream",
  "sigma_zeta_year",
  "sigma_zeta_event",
  "sigma_iota_stream",
  "sigma_iota_year",
  "sigma_iota_event",
  "sigma_iota_cluster",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_year",
  "sigma_eta_event",
  "sigma_eta_cluster",
  "p_TP",
  "p_FP",
  "z",
  "tau_eff",
  "loglik"
)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000


set.seed(12345)

Sys.time()

hurdle_damDen_m3_SW <- jags(data = hurdle_damDen_m3_data_list,
                   inits = hurdle_damDen_m3_inits,
                   parameters.to.save = hurdle_damDen_m3_params,
                   model.file = "HGLMM_Hurdle_m3_SW_damDen.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

```


----

<!---
DIAGNOSTICS:
----


```{r,include=FALSE}
# Extract posterior draws

damDen_m3_post <- hurdle_damDen_m3_SW$sims.list 
```

**Assess "Chain health" Rhat + ESS:**

```{r,include=FALSE} 
# Choose only the parameters you care about
params_m3 <- c(
  "epsilon",
  "alpha",
  "zeta_damDen",
  "beta_damDen",
  "sigma",
  "sigma_zeta_stream",
  "sigma_iota_cluster",
  "p_TP[1]",
  "p_FP[1]")

# Gelman–Rubin convergence diagnostic
gelman.diag(hurdle_damDen_m3_SW$samples[, params_m3], multivariate = FALSE)

# Effective sample size
effectiveSize(hurdle_damDen_m3_SW$samples[, params_m3])

```

    Rhat & Effective Sample Size:
    
    Rhat values look pretty good. All near 1.00. Zeta_damDen is slower to converge, but
    still does so eventually. 
    
    Effective sample sizes look good for all parameters (note beta_damDen). 

----

**Check traceplots** for critical parameters:

```{r,include=FALSE} 

par(mfrow = c(2, 3),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


tplot_params_m3 <- c("alpha","beta_damDen","zeta_damDen", "sigma","sigma_zeta_stream","sigma_beta_stream")


mcmc_sub_m3 <- hurdle_damDen_m3_SW$samples[, tplot_params_m3]

coda::traceplot(mcmc_sub_m3)
```


    Traceplots:
    
    Traceplots all look sufficiently "grassy," indicating sufficient chain convergence>
    NOTE: zeta_damDen has trouble converging. This is likely because 1 SD change in dam
    density has a very small, yet highly variable estimated effect on surface water presence.
    This is no surprise given the results of model 2. 


----

**Posterior predictive plots:**

Predicted Probability of SW presence (Psi) vs observed SW presence (y)

```{r,include=FALSE} 
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

# Posterior mean psi
psi_post_m3 <- apply(hurdle_damDen_m3_SW$sims.list$psi, 2, mean)

# Compare to observed y
plot(psi_post_m3, y, pch=16, col=rgb(0,0,0,0.2))

```

----

Predicted log(SWA) vs observed log(SWA), conditional on SW presence

Goal is 1:1 (ish)

```{r,include=FALSE}
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

# Posterior mean of predicted log(VA + 1)
m3_mu_SW_post <- apply(hurdle_damDen_m3_SW$sims.list$mu, 2, mean)

# Observed log vegetation area
m3_obs_logSW <- y2

# Fit regression on the log scale
m3_fit_log <- lm(m3_obs_logSW ~ m3_mu_SW_post)

m3_slope_log <- coef(m3_fit_log)[2]
m3_R2_log    <- summary(m3_fit_log)$r.squared

# Plot
plot(m3_mu_SW_post,
     m3_obs_logSW,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated log(SWA + 1)",
     ylab = "Observed log(SWA + 1)")

abline(0, 1, col = "red", lwd = 2)

```

Slope:
```{r,include=FALSE}
m3_slope_log
```

    1.008 suggests the model predicts the variation in logSWA data well.

----

**Random effects shrinkage: **

Goal is centered around 0(ish)


```{r, include=FALSE, }
par(mfrow = c(3,5),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

boxplot(hurdle_damDen_m3_SW$sims.list$iota_stream,
        main = "iota_stream")
boxplot(hurdle_damDen_m3_SW$sims.list$eta_stream,
        main = "eta_stream")
boxplot(hurdle_damDen_m3_SW$sims.list$zeta_stream,
        main = "zeta_stream")
boxplot(hurdle_damDen_m3_SW$sims.list$beta_stream,
        main = "beta_stream")
boxplot(hurdle_damDen_m3_SW$sims.list$zeta_year,
        main = "zeta_year")
boxplot(hurdle_damDen_m3_SW$sims.list$beta_year,
        main = "beta_year")
boxplot(hurdle_damDen_m3_SW$sims.list$iota_event,
        main = "iota_event")

boxplot(hurdle_damDen_m3_SW$sims.list$iota_year,
        main = "iota_year")

boxplot(hurdle_damDen_m3_SW$sims.list$eta_event,
        main = "eta_event")

boxplot(hurdle_damDen_m3_SW$sims.list$eta_year,
        main = "eta_year")

boxplot(hurdle_damDen_m3_SW$sims.list$zeta_event,
        main = "zeta_event")
boxplot(hurdle_damDen_m3_SW$sims.list$beta_event,
        main = "beta_event")


```


----

----


**Posterior mean dam effects:**

damDen -> swPres

```{r,include=FALSE}
m3_coeffparams <- c(#"alpha",
            "zeta_damDen",
            "beta_damDen",
            "sigma")

m3_coef_df <- hurdle_damDen_m3_SW$summary[m3_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m3_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.2,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```

```{r,include=FALSE}
summary(hurdle_damDen_m3_SW$sims.list$zeta_damDen)

exp(mean(hurdle_damDen_m3_SW$sims.list$zeta_damDen))
```

```{r,include=FALSE}
#library(tidyverse)
#library(scales)
par(mfrow = c(1, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

#Posterior distribution of damEffect on vegArea:

hist(hurdle_damDen_m3_SW$sims.list$zeta_damDen,
     breaks = 50,
     col = "steelblue",
     main = "Posterior of damDensity effect on logit(Psi)",
     xlab = "beta_damPres")
abline(v = mean(hurdle_damDen_m3_SW$sims.list$zeta_damDen), col = "red", lwd = 2)
```

    Without considering local geomorphic conditions, a 1 SD increase dam density is associated
    with basically no change in the odds of surface water presence.
    
----


damPres -> swAmount (hglmm)

```{r,include=FALSE}
summary(hurdle_damDen_m3_SW$sims.list$beta_damDen)

exp(mean(hurdle_damDen_m3_SW$sims.list$beta_damDen))
```

```{r,include=FALSE}
#library(tidyverse)
#library(scales)
par(mfrow = c(1, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

#Posterior distribution of damEffect on vegArea:

hist(hurdle_damDen_m3_SW$sims.list$beta_damDen,
     breaks = 50,
     col = "steelblue",
     main = "Posterior of damDensity effect on log(SWA)",
     xlab = "beta_damPres")
abline(v = mean(hurdle_damDen_m3_SW$sims.list$beta_damDen), col = "red", lwd = 2)
```
    
    Without accounting for geomorphic conditions, a 1 SD increase in dam density corresponds 
    with a 20-26% increase in logSWA.



----

**Prior-posterior overlap:**

```{r,include=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 

damDen_alpha_post_m3 <- hurdle_damDen_m3_SW$sims.list$alpha
damDen_beta_post_m3  <- hurdle_damDen_m3_SW$sims.list$beta_damDen
damDen_zeta_post_m3  <- hurdle_damDen_m3_SW$sims.list$zeta_damDen
damDen_sigBetaStream_post_m3 <- hurdle_damDen_m3_SW$sims.list$sigma_beta_stream

damDenMod_m3_n <- length(damDen_alpha_post_m3)

damDen_alpha_prior_m3 <- rnorm(damDenMod_m3_n, 3.57, sqrt(1/0.04))
damDen_beta_prior_m3  <- rnorm(damDenMod_m3_n, 0, sqrt(1/0.04))
damDen_zeta_prior_m3  <- rnorm(damDenMod_m3_n, 0, sqrt(1/0.04))
damDen_sigBetaStream_prior_m3 <- pmax(rnorm(damDenMod_m3_n, 0, 1), 0)


# Plot overlap:

#plot_prior_post <- function(prior, post, name) {
  
#  df <- tibble(
#    value = c(prior, post),
#    dist  = rep(c("Prior", "Posterior"), each = length(prior))
#  )
  
#  ggplot(df, aes(x = value, fill = dist)) +
#    geom_density(alpha = 0.4) +
#    scale_fill_manual(values = c("gray60", "steelblue")) +
#    labs(
#      title = paste("Prior vs Posterior for", name),
#      x = name,
#      y = "Density"
#    ) +
#    theme_minimal(base_size = 14)
#}

plot_prior_post(damDen_alpha_prior_m3, damDen_alpha_post_m3, "alpha_m3")
plot_prior_post(damDen_beta_prior_m3, damDen_beta_post_m3, "beta_damDen_m3")
plot_prior_post(damDen_zeta_prior_m3, damDen_zeta_post_m3, "zeta_damDen_m3")
plot_prior_post(damDen_sigBetaStream_prior_m3, damDen_sigBetaStream_post_m3, "sigma_beta_stream_m3")
```


    PPOs still look pretty good.

----

**Change in SWA across damDensity by stream and sample event**

```{r,echo=FALSE}

m3_dam_seq <- seq(min(aoi_HurdleData$dam_km),
               max(aoi_HurdleData$dam_km),
               length.out = 50)

# build PDP grid
m3_pdp_stream <- map_dfr(1:n_stream, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m3_post$alpha
  beta_j  <- damDen_m3_post$beta_damDen 
              + damDen_m3_post$beta_stream[, j] 
              + rowMeans(damDen_m3_post$beta_year) 
              + rowMeans(damDen_m3_post$beta_event)
  eta_j   <- damDen_m3_post$eta_stream[, j] 
              + rowMeans(damDen_m3_post$eta_event)

  map_dfr(m3_dam_seq, function(d) {

    mu_draws <- alpha_j + beta_j * d + eta_j
    swa_draws <- exp(mu_draws) - 1

    tibble(
      stream = j,
      damDen = d,
      SWA_med = median(swa_draws),
      SWA_lo  = quantile(swa_draws, 0.025),
      SWA_hi  = quantile(swa_draws, 0.975)
    )
  })
})

stream_lookup <- aoi_HurdleData %>%
  distinct(stream, watershed) %>%
  arrange(stream)

m3_pdp_stream <- m3_pdp_stream %>%
  left_join(stream_lookup, by = "stream")



# plot
ggplot(m3_pdp_stream, aes(x = watershed, y = SWA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Partial Dependence of SWA on Stream and Dam Density",
    x = "Stream",
    y = "Predicted SWA (m²)",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


###############################################################
################################################################

# build PDP grid
m3_pdp_event <- map_dfr(1:n_event, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m3_post$alpha
  beta_j <- damDen_m3_post$beta_damDen                 # dam effect
             + damDen_m3_post$beta_event[, j]           # plus event effect
             + damDen_m3_post$beta_year[, event_year_3_4[j]] # plus nested year effect
             + rowMeans(damDen_m3_post$beta_stream)      # for a typical stream

  eta_j <- damDen_m3_post$eta_event[, j] 
            + damDen_m3_post$eta_year[, event_year_3_4[j]]


  map_dfr(m3_dam_seq, function(d) {

    mu_draws <- alpha_j + beta_j * d + eta_j
    swa_draws <- exp(mu_draws) - 1

    tibble(
      event = j,
      damDen = d,
      SWA_med = median(swa_draws),
      SWA_lo  = quantile(swa_draws, 0.025),
      SWA_hi  = quantile(swa_draws, 0.975)
    )
  })
})

event_year_lookup <- tibble(
  event = 1:12,
  event_label = c(
    "2019_Aug_1","2019_Sep_1","2019_Aug_2","2019_Sep_2",
    "2021_Aug_1","2021_Sep_1","2021_Aug_2","2021_Sep_2",
    "2023_Aug_1","2023_Sep_1","2023_Aug_2","2023_Sep_2"
  )
)


m3_pdp_event_named <- m3_pdp_event %>%
  left_join(event_year_lookup, by = "event")



# plot
ggplot(m3_pdp_event_named, aes(x = event_label, y = SWA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Partial Dependence of SWA on Sample Event and Dam Density",
    x = "Sample Event",
    y = "Predicted SWA (m²)",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```

--->
----

----

----

----

----



**MODEL 4: SW ~  damDensity + Geom.Covar.**

```{r,echo=FALSE}
#library(jagsUI)
```
    sink("HGLMM_Hurdle_m4_SW_DamDen.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR SW PRESENCE (1/2 Normal on SDs):
    
      sigma_zeta_stream  ~ dnorm(0,1) T(0,)
      tau_zeta_stream    <- pow(sigma_zeta_stream, -2)
    
      sigma_zeta_year    ~ dnorm(0,1) T(0,)
      tau_zeta_year      <- pow(sigma_zeta_year, -2)
    
      sigma_zeta_event   ~ dnorm(0,1) T(0,)
      tau_zeta_event     <- pow(sigma_zeta_event, -2)
    
      sigma_iota_stream  ~ dnorm(0,1) T(0,)
      tau_iota_stream    <- pow(sigma_iota_stream, -2)
      
      sigma_iota_year    ~ dnorm(0,1) T(0,)
      tau_iota_year      <- pow(sigma_iota_year, -2)
    
      sigma_iota_event   ~ dnorm(0,1) T(0,)
      tau_iota_event     <- pow(sigma_iota_event, -2)
    
      sigma_iota_cluster ~ dnorm(0,1) T(0,)
      tau_iota_cluster   <- pow(sigma_iota_cluster, -2)
    
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 4) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 4) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 4) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
      
      sigma_eta_year     ~ dnorm(0, 1) T(0,)
      tau_eta_year       <- pow(sigma_eta_year, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Presence/absence:
      epsilon ~ dnorm(0,0.04)
      zeta_damDen ~ dnorm(0,0.04)
      zeta_incis ~ dnorm(0,0.04)
      zeta_RIPA ~ dnorm(0,0.04)
      zeta_UDA ~ dnorm(0,0.04)
    
      # Amount:
      alpha ~ dnorm(3.57,0.04)     # within 5 SD of empirical mean logSWA
      beta_damDen ~ dnorm(0,0.04)
      beta_incis ~ dnorm(0,0.04)
      beta_RIPA ~ dnorm(0,0.04)
      beta_UDA ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW PRESENCE:
    
      for(j in 1:n_stream){
        zeta_stream[j] ~ dnorm(0, tau_zeta_stream)
        iota_stream[j] ~ dnorm(0, tau_iota_stream)
      }
    
      for(yr in 1:n_year){
        zeta_year[yr] ~ dnorm(0, tau_zeta_year)
        iota_year[yr] ~ dnorm(0, tau_iota_year)
      }
    
      for(k in 1:n_event){
        zeta_event[k] ~ dnorm(zeta_year[event_year[k]], tau_zeta_event)
        iota_event[k] ~ dnorm(iota_year[event_year[k]], tau_iota_event)
      }
    
      for(c in 1:n_cluster){
        iota_cluster[c] ~ dnorm(0, tau_iota_cluster)
      }
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
        eta_year[yr] ~ dnorm(0, tau_eta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(eta_year[event_year[k]], tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # DETECTION PRIORS:
      
      for(k in 1:n_event){
      p_TP[k] ~ dnorm(0.76, pow(0.16,-2)) T(0,1)   # empirically derived recall
      p_FP[k] ~ dnorm(0.01, pow(0.005,-2)) T(0,1)  # empirically derived FPR
      }
      
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Presence/absence:
        z[i] ~ dbern(psi[i])
    
        logit(psi[i]) <- epsilon
                         + (zeta_damDen
                            + zeta_stream[stream[i]]
                            + zeta_year[year[i]]
                            + zeta_event[event[i]]) * damDen[i]
                         + zeta_incis * incis[i]
                         + zeta_RIPA * RIPA[i]
                         + zeta_UDA * UDA[i]
                         + iota_stream[stream[i]]
                         + iota_event[event[i]]
                         + iota_cluster[cluster[i]]
    
        # Detection:
        p_eff[i] <- z[i] * p_TP[event[i]] + (1-z[i]) * p_FP[event[i]]
        y[i]     ~ dbern(p_eff[i])
    
        # Amount:
        y2[i] ~ dnorm(mu[i], tau_eff[i])
    
        mu[i] <- z[i] *
                  (alpha
                    + (beta_damDen
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damDen[i]
                    + beta_incis * incis[i] 
                    + beta_RIPA * RIPA[i]
                    + beta_UDA * UDA[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]])
    
        tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
        
        
      ##########################################################################
                    
      # Derived parameter for WAIC computation:

      loglik[i] <- logdensity.norm(y[i], mu[i], tau)
      }
    
    }
    ", fill = TRUE)
    
    sink()
  
    

```{r,echo=FALSE}
# hurdle_damDen_m4_data_list <- list(
#   N = N,
#   y = y,
#   y2 = y2,
#   stream = stream,
#   year = year,
#   event = event,
#   cluster = cluster,
#   damDen = damDen,
#   incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
#   RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2)),
#   UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)),
#   n_stream = n_stream,
#   n_year = n_year,
#   n_event = n_event,
#   n_cluster = n_cluster,
#   event_year = event_year_3_4
# )

# hurdle_damDen_m4_inits <- function(){
#   list()
# }

# hurdle_damDen_m4_params <- c(
#   "psi",
#   "mu",
#   "epsilon",
#   "alpha",
#   "zeta_damDen",
#   "zeta_incis",
#   "zeta_RIPA",
#   "zeta_UDA",
#   "zeta_stream",
#   "zeta_year",
#   "zeta_event",
#   "iota_stream",
#   "iota_year",
#   "iota_event",
#   "iota_cluster",
#   "beta_damDen",
#   "beta_incis",
#   "beta_RIPA",
#   "beta_UDA",
#   "beta_stream",
#   "beta_year",
#   "beta_event",
#   "eta_stream",
#   "eta_year",
#   "eta_event",
#   "eta_cluster",
#   "sigma",
#   "sigma_zeta_stream",
#   "sigma_zeta_year",
#   "sigma_zeta_event",
#   "sigma_iota_stream",
#   "sigma_iota_year",
#   "sigma_iota_event",
#   "sigma_iota_cluster",
#   "sigma_beta_stream",
#   "sigma_beta_year",
#   "sigma_beta_event",
#   "sigma_eta_stream",
#   "sigma_eta_year",
#   "sigma_eta_event",
#   "sigma_eta_cluster",
#   "p_TP",
#   "p_FP",
#   "z",
#   "tau_eff",
#   "loglik"
# )

# nchains <- 3
# nthin <- 5
# niter <- 100000
# nburn <- 25000

# set.seed(12345)

# Sys.time()

# hurdle_damDen_m4_SW <- jags(data = hurdle_damDen_m4_data_list,
  #                  inits = hurdle_damDen_m4_inits,
  #                  parameters.to.save = hurdle_damDen_m4_params,
  #                  model.file = "HGLMM_Hurdle_m4_SW_DamDen.jags",
  #                  parallel = T,
  #                  n.chains = nchains,
  #                  n.thin = nthin,
  #                  n.iter = niter,
  #                  n.burnin = nburn)


# Sys.time()
```



```{r,echo=FALSE}
# SAVE THE MODEL OUTPUT SO YOU DON'T HAVE TO RUN IT EVERY TIME!!

#saveRDS(hurdle_damDen_m4_SW, "C:\Users\mattw\Desktop\Ch1_ModelOutputs\hurdle_damDen_m4_SW_model.rds")


# hurdle_damDen_m4_SW <- readRDS("C:\Users\mattw\Desktop\Ch1_ModelOutputs\hurdle_damDen_m4_SW_model.rds")
```





----


DIAGNOSTICS:
----

```{r,echo=FALSE}
# Extract posterior draws

damDen_m4_post <- hurdle_damDen_m4_SW$sims.list 
```

**Assess "Chain health" Rhat + ESS:**

```{r,echo=FALSE} 
# Choose only the parameters you care about
params_m4 <- c(
  "epsilon",
  "alpha",
  "zeta_damDen",
  "beta_damDen",
  "zeta_incis",
  "beta_incis",
  "zeta_RIPA",
  "beta_RIPA",
  "zeta_UDA",
  "beta_UDA",
  "sigma",
  "sigma_zeta_stream",
  "sigma_iota_cluster",
  "p_TP[1]",
  "p_FP[1]")

# Gelman–Rubin convergence diagnostic
gelman.diag(hurdle_damDen_m4_SW$samples[, params_m4], multivariate = FALSE)

# Effective sample size
effectiveSize(hurdle_damDen_m4_SW$samples[, params_m4])

```

    Rhat & Effective Sample Size:
    
    Acceptable. Most Rhats = 1. zeta_damDen is slow to converge, and having trouble
    being estimated. It is a REALLY small effect, with LOTS of variation across grouping
    random variables, so it's hard for the MCMC chains to converge on a single estimate.


----

**Check traceplots** for critical parameters:

```{r,echo=FALSE} 

par(mfrow = c(3, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


tplot_params_m4 <- c("alpha","beta_damDen","zeta_damDen", "sigma","sigma_zeta_stream","sigma_beta_stream")


mcmc_sub_m4 <- hurdle_damDen_m4_SW$samples[, tplot_params_m4]

coda::traceplot(mcmc_sub_m4)
```


    Traceplots all look okay, again see zeta_damDen. Could increase iterations, or tighten prior...?
    
    Matt: Closely read Ogle and Barber (2020) Ensuring identifiability in hierarchical mixed effects Bayesian models
    
    for potential ways to improve/considerations re: identifiability...


----

**Posterior predictive plots:**

Predicted Psi vs observed SW presence

```{r,echo=FALSE} 
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

# Posterior mean psi
psi_post_m4 <- apply(hurdle_damDen_m4_SW$sims.list$psi, 2, mean)

# Compare to observed y
plot(psi_post_m4, y, pch=16, col=rgb(0,0,0,0.2))


```

----

Predicted SW vs observed SW

Goal is 1:1 (ish)

```{r,echo=FALSE}
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

# Posterior mean of predicted log(VA + 1)
m4_mu_SW_post <- apply(hurdle_damDen_m4_SW$sims.list$mu, 2, mean)

# Observed log vegetation area
m4_obs_logSW <- y2

# Fit regression on the log scale
m4_fit_log <- lm(m4_obs_logSW ~ m4_mu_SW_post)

m4_slope_log <- coef(m4_fit_log)[2]
m4_R2_log    <- summary(m4_fit_log)$r.squared

# Plot
plot(m4_mu_SW_post,
     m4_obs_logSW,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated log(SWA + 1)",
     ylab = "Observed log(SWA + 1)")

abline(0, 1, col = "red", lwd = 2)

```

Slope:
```{r,echo=FALSE}
m4_slope_log
```

    1.008 suggests model fits the data well, though slightly underestimates variance in
    logSWA at lower values.

----

**Random effects shrinkage: **

Goal is centered around 0(ish)

. iota_stream; stream random intercept for SW presence
. eta_stream; stream random intercept for SWA
. zeta_stream; stream random slope for damDen effect on SW presence
. beta_stream; stream random slope for damDen effect on SWA
. zeta_event; event random slope for damDen effect on SW presence
. beta_event; event random slope for damDen effect on SWA

```{r, echo=FALSE, }
par(mfrow = c(3,5),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

boxplot(hurdle_damDen_m4_SW$sims.list$iota_stream,
        main = "iota_stream")
boxplot(hurdle_damDen_m4_SW$sims.list$eta_stream,
        main = "eta_stream")
boxplot(hurdle_damDen_m4_SW$sims.list$zeta_stream,
        main = "zeta_stream")
boxplot(hurdle_damDen_m4_SW$sims.list$beta_stream,
        main = "beta_stream")
boxplot(hurdle_damDen_m4_SW$sims.list$zeta_year,
        main = "zeta_year")
boxplot(hurdle_damDen_m4_SW$sims.list$beta_year,
        main = "beta_year")
boxplot(hurdle_damDen_m4_SW$sims.list$iota_event,
        main = "iota_event")

boxplot(hurdle_damDen_m4_SW$sims.list$iota_year,
        main = "iota_year")

boxplot(hurdle_damDen_m4_SW$sims.list$eta_event,
        main = "eta_event")

boxplot(hurdle_damDen_m4_SW$sims.list$eta_year,
        main = "eta_year")

boxplot(hurdle_damDen_m4_SW$sims.list$zeta_event,
        main = "zeta_event")
boxplot(hurdle_damDen_m4_SW$sims.list$beta_event,
        main = "beta_event")


```

    These all seem reasonable.

----

**Prior-posterior overlap:**

```{r,echo=FALSE}
damDen_alpha_m4_post <- hurdle_damDen_m4_SW$sims.list$alpha
damDen_beta_m4_post  <- hurdle_damDen_m4_SW$sims.list$beta_damDen
damDen_zeta_m4_post  <- hurdle_damDen_m4_SW$sims.list$zeta_damDen
damDen_sigBetaStream_m4_post <- hurdle_damDen_m4_SW$sims.list$sigma_beta_stream

damDenMod_m4_n <- length(damDen_alpha_m4_post)

damDen_alpha_m4_prior <- rnorm(damDenMod_m4_n, 0, sqrt(3.57/0.04))
damDen_beta_m4_prior  <- rnorm(damDenMod_m4_n, 0, sqrt(1/0.04))
damDen_zeta_m4_prior  <- rnorm(damDenMod_m4_n, 0, sqrt(1/0.04))
damDen_sigBetaStream_m4_prior <- pmax(rnorm(damDenMod_m4_n, 0, 1), 0)


# Plot overlap:

plot_prior_post <- function(prior, post, name) {
  
  df <- tibble(
    value = c(prior, post),
    dist  = rep(c("Prior", "Posterior"), each = length(prior))
  )
  
  ggplot(df, aes(x = value, fill = dist)) +
    geom_density(alpha = 0.4) +
    scale_fill_manual(values = c("gray60", "steelblue")) +
    labs(
      title = paste("Prior vs Posterior for", name),
      x = name,
      y = "Density"
    ) +
    theme_minimal(base_size = 14)
}

plot_prior_post(damDen_alpha_m4_prior, damDen_alpha_m4_post, "alpha_m4")
plot_prior_post(damDen_beta_m4_prior, damDen_beta_m4_post, "beta_damDen_m4")
plot_prior_post(damDen_zeta_m4_prior, damDen_zeta_m4_post, "zeta_damDen_m4")
plot_prior_post(damDen_sigBetaStream_m4_prior, damDen_sigBetaStream_m4_post, "sigma_beta_stream_m4")
```

    PPOs look good pretty good. Data seems to be driving posterior estimates.


----

----


**Posterior Estimates**

```{r,echo=FALSE}
m4_epsilon <- mean(hurdle_damDen_m4_SW$sims.list$epsilon)

m4_zeta_dd <- mean(hurdle_damDen_m4_SW$sims.list$zeta_damDen)

m4_zeta_in <- mean(hurdle_damDen_m4_SW$sims.list$zeta_incis)

m4_zeta_rip <- mean(hurdle_damDen_m4_SW$sims.list$zeta_RIPA)

m4_zeta_uda <- mean(hurdle_damDen_m4_SW$sims.list$zeta_UDA)

m4_alpha <- mean(hurdle_damDen_m4_SW$sims.list$alpha)

m4_beta_dd <- mean(hurdle_damDen_m4_SW$sims.list$beta_damDen)

m4_beta_in <- mean(hurdle_damDen_m4_SW$sims.list$beta_incis)

m4_beta_rip <- mean(hurdle_damDen_m4_SW$sims.list$beta_RIPA)

m4_beta_uda <- mean(hurdle_damDen_m4_SW$sims.list$beta_UDA)
```

```{r,echo=FALSE}
m4_coeffparams <- c(#"alpha",
            "zeta_damDen",
            "zeta_incis",
            "zeta_RIPA",
            "zeta_UDA",
            "beta_damDen",
            "beta_incis",
            "beta_RIPA",
            "beta_UDA",
            "sigma")

m4_coef_df <- hurdle_damDen_m4_SW$summary[m4_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m4_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.2,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    x = "Effect Size",
    y = ""
  ) +
  theme_bw(base_size = 14)
```

----

**Estimated effects of dam density**

----

**Effect of 1 SD increase in dam density on surface water (SW) presence:**

```{r,echo=FALSE}
summary(hurdle_damDen_m4_SW$sims.list$zeta_damDen)

exp(mean(hurdle_damDen_m4_SW$sims.list$zeta_damDen))


p_SW_damDen_0 <- plogis(m4_epsilon)
#p_SW_damDen_0

p_SW_damDen_1 <- plogis(m4_epsilon + 0.017)
#p_SW_damDen_1

print("Change in Psi(SW) with 1 SD increase in dam density")
(p_SW_damDen_1 - p_SW_damDen_0)
```


    For a typical stream at mean covariate conditions, a 1 SD increase dam density is associated
    with basically no change in the probability of surface water presence (0.017 change on 
    log-odds scale corresponds to an increase of (0.2%) on the probability scale).
    
    NOTE: HEAVILY OVERLAPS 0. This indicates that there is no consistent influence of damDensity
    across grouping variables (stream,year,event) in the data. Some settings see strong positive
    influence of dams, some settings see strong negative influence of dams, most see a weak or 
    negligible influence of dams.


**Effect of 1 SD increase in dam density on SW amount:**

```{r,echo=FALSE}
summary(hurdle_damDen_m4_SW$sims.list$beta_damDen)

exp(mean(hurdle_damDen_m4_SW$sims.list$beta_damDen))


SWA_damDen_0 <- exp(m4_alpha)
#SWA_damDen_0

SWA_damDen_1 <- exp(m4_alpha + 0.12)
#SWA_damDen_1

print("SWA change with 1 SD increase in dam density")
(SWA_damDen_1 - SWA_damDen_0)

```

    A 1 SD increase dam density is associated with ~ 12% increase in log(SWA+1),
    which, for a "typical" stream under mean geomorphic covariate conditions, corresponds
    to an increase of ~ 42 m^2 of SW.
    
    Note: at the high end of dam density, a typical stream at mean covariate conditions
    would see an estimated increase of ~150-300 m^2 of SW.
    
    NOTE: see above note re: 0 overlap and consistency of damDensity effect across the data
    
    
What is 300-m^2 compared to the baseline SW?
What is 300-m^2 compared to riparian_area_m2_0?

----

**Estimated effects of geomorphic covariates**

----

**Effect of 1 SD increase in incision on surface water (SW) presence:**

```{r,echo=FALSE}
summary(hurdle_damDen_m4_SW$sims.list$zeta_incis)

exp(mean(hurdle_damDen_m4_SW$sims.list$zeta_incis))


p_SW_incis_0 <- plogis(m4_epsilon)
#p_SW_incis_0

p_SW_incis_1 <- plogis(m4_epsilon - 1.10)
#p_SW_incis_1

print("Change in Psi(SW) with 1 SD increase in incision")
(p_SW_incis_1 - p_SW_incis_0)
```

    A 1 SD increase in incision (incis) (which corresponds to ~0.1 increase on the raw scale) is associated with
    a ~77% decrease in odds of SW presence, which is ~21% drop in probability of SW (Psi), for a 
    typical stream, at mean covariate conditions. 
    
    Convincing negative effect, not overlappping 0.
    
    
  

    
**Effect of 1 SD increase in incision on SW amount:**

```{r,echo=FALSE}
summary(hurdle_damDen_m4_SW$sims.list$beta_incis)

exp(mean(hurdle_damDen_m4_SW$sims.list$beta_incis))


SWA_incis_0 <- exp(m4_alpha)
#SWA_incis_0

SWA_incis_1 <- exp(m4_alpha - 0.68)
#SWA_incis_1

print("SWA change with 1 SD increase in incision")
(SWA_incis_1 - SWA_incis_0)
```

    A 1 SD increase in incision is associated with a ~51% decrease in surface water area (log(SWA)),
    conditional on surface water presence (z=1). For a typical stream at mean covariate conditions, 
    this corresponds to a decrease in ~162 m^2. 
    
    Strong negative effect, not overlapping 0.

----

**Effect of 1 SD increase in riparian area on surface water (SW) presence:**

```{r,echo=FALSE}
summary(hurdle_damDen_m4_SW$sims.list$zeta_RIPA)

exp(mean(hurdle_damDen_m4_SW$sims.list$zeta_RIPA))



p_SW_RIPA_0 <- plogis(m4_epsilon)
#p_SW_RIPA_0

p_SW_RIPA_1 <- plogis(m4_epsilon + 0.48)
#p_SW_RIPA_1

print("Change in Psi(SW) with 1 SD increase in riparian area")
(p_SW_RIPA_1 - p_SW_RIPA_0)
```
    
    A 1 SD increase in riparian area (corresponds with an increase of ~1000 m^2
    on the raw RIPA scale), is associated with an increase of 0.48 in ODDS, which corresponds
    with a ~5.9% increase in the PROBABILITY of SW presence.


    
**Effect of 1 SD increase in riparian area on SW amount:**

```{r,echo=FALSE}
summary(hurdle_damDen_m4_SW$sims.list$beta_RIPA)

exp(mean(hurdle_damDen_m4_SW$sims.list$beta_RIPA))




SWA_RIPA_0 <- exp(m4_alpha)
#SWA_UDA_0

SWA_RIPA_1 <- exp(m4_alpha - 0.07)
#SWA_UDA_1

print("SWA change with 1 SD increase in riparian area")
(SWA_RIPA_1 - SWA_RIPA_0)
```


    A 1 SD increase in RIPA is associated with a ~ 7% decrease in log(SWA). Meager negative
    effect, slightly overlapping 0. This corresponds with a decrease of ~22 m^2
    (for typical stream under mean conditions).
    

----

**Effect of 1 SD increase in upstream drainage area (UDA) on SW presence:**

```{r,echo=FALSE}
summary(hurdle_damDen_m4_SW$sims.list$zeta_UDA)

exp(mean(hurdle_damDen_m4_SW$sims.list$zeta_UDA))



p_SW_UDA_0 <- plogis(m4_epsilon)
#p_SW_UDA_0

p_SW_UDA_1 <- plogis(m4_epsilon + 2.06)
#p_SW_UDA_1

print("Change in Psi(SW) with 1 SD increase in drainage area")
(p_SW_UDA_1 - p_SW_UDA_0)

```
  
    A 1 SD increase in Upstream Drainage Area (UDA) (corresponds with an increase of ~1320 km^2
    on the raw UDA scale), is associated with a ~786% increase in ODDS, which corresponds to a
    ~15% increase in the PROBABILITY of SW presence (for typical stream under mean conditions)

----  
    
**Effect of 1 SD increase in Upstream Drainage Area on SW amount:**

```{r,echo=FALSE}
summary(hurdle_damDen_m4_SW$sims.list$beta_UDA)

exp(mean(hurdle_damDen_m4_SW$sims.list$beta_UDA))


SWA_UDA_0 <- exp(m4_alpha)
#SWA_UDA_0

SWA_UDA_1 <- exp(m4_alpha + 0.56)
#SWA_UDA_1

print("SWA change with 1 SD increase in drainage area")
(SWA_UDA_1 - SWA_UDA_0)

```

    A 1 SD increase in Upstream Drainage Area (UDA) is associated with a 56% increase in log(SWA),
    conditional on surface water presence. This corresponds to an increase of ~246.43 m^2 in SWA (m^2)
    for a "typical" stream at mean covariate conditions.

----

----

----

**Marginal effect** of dam Density on surface water presence, given geomorphic context

```{r,echo=FALSE}
# Build a grid of raw geomorphic covariate values
damDen_m4_geomCovar_grid <- tibble(
  TO_raw = seq(min(aoi_HurdleData$TO),
               max(aoi_HurdleData$TO), length.out = 50),

  incis_raw = 1 - (TO_raw / 90),   # raw incision scale

  RIPA_raw = seq(quantile(aoi_HurdleData$riparian_area_m2, 0.05),
                 quantile(aoi_HurdleData$riparian_area_m2, 0.95),
                 length.out = 50),

  UDA_raw = seq(quantile(aoi_HurdleData$catchment_km2, 0.05),
              quantile(aoi_HurdleData$catchment_km2, 0.95),
              length.out = 50)

) %>%
  # Convert raw → standardized (to match model scale)
  mutate(
    incis = as.numeric(scale(incis_raw)),
    RIPA  = as.numeric(scale(RIPA_raw)),
    UDA   = as.numeric(scale(UDA_raw))
  )

# Make dam density sequence
damDen_m4_seq_raw <- seq(min(aoi_HurdleData$dam_km), max(aoi_HurdleData$dam_km), length.out=50)
damDen_m4_seq_scaled <- seq(min(aoi_HurdleData$dam_km_std),max(aoi_HurdleData$dam_km_std),length.out=50)

# Compute surface water across dam density sequence
damDen_m4_compute_psi <- function(damDen, damDen_scaled, geom_grid, damDen_post) {

  eps <- mean(damDen_post$epsilon)
  zd  <- mean(damDen_post$zeta_damDen)
  zi  <- mean(damDen_post$zeta_incis)
  zr  <- mean(damDen_post$zeta_RIPA)
  zu  <- mean(damDen_post$zeta_UDA)

  geom_grid %>%
    mutate(
      psi = plogis(
        eps +
        zd * damDen_scaled +
        zi * incis +      # standardized incision
        zr * RIPA +       # standardized RIPA
        zu * UDA          # standardized UDA
      ),
      damDen = damDen,
      damDen_scaled = damDen_scaled
    )
}


```

```{r,echo=FALSE}
damDen_m4_psi_df <- map2_df(damDen_m4_seq_raw,
                           damDen_m4_seq_scaled,
                           ~ damDen_m4_compute_psi(.x,
                                                   .y,
                                                   damDen_m4_geomCovar_grid,
                                                   damDen_m4_post))
```

```{r,echo=FALSE}
ggplot(damDen_m4_psi_df, aes(x = incis_raw, y = psi, color = damDen)) +
  geom_point() +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "Marginal Effects of Dam Density and Channel Incision on the Probability of Surface Water Presence",
    x = "Channel Incision (raw scale: 1 - TO/90)",
    y = expression("Probability of Surface Water (" * Psi * ")"),
    color = "Dam Density (dams/km)"
  ) +
  theme_minimal(base_size = 14)

```

```{r,echo=FALSE}
ggplot(damDen_m4_psi_df, aes(x = RIPA_raw, y = psi, color = damDen)) +
  geom_point() +
  scale_color_viridis_c(option="plasma") +
  labs(
    title = "Marginal Effects of Dam Density and Riparian Area on the Probability of Surface Water Presence",
    x = expression("Riparian Area" * (m^2)),
    y = expression("Probability of Surface Water Presence (" * Psi * ")"),
    color = "Dam Density (dams/km)"
  ) +
  theme_minimal(base_size = 14)
```

```{r,echo=FALSE}
ggplot(damDen_m4_psi_df, aes(x = UDA_raw, y = psi, color = damDen)) +
  geom_point() +
  scale_color_viridis_c(option="plasma") +
  labs(
    title = "Marginal Effects of Dam Density and Upstream Drainage Area on the Probability of Surface Water Presence",
    x = expression("Upstream Drainage Area" * (km^2)),
    y = expression("Probability of Surface Water Presence (" * Psi * ")"),
    color = "Dam Density (dams/km)"
    ) +
  theme_minimal(base_size = 14)
```

----

**Partial dependence plots**

Constrain PDP to empirical ranges for SWA and target covariates

Dam density effect across target covariate range; all other covariates held at median conditions

```{r,echo=FALSE}
library(tidyverse)
library(scales)

# Raw covariates
incis_m4_raw <- 1 - aoi_HurdleData$TO/90
RIPA_m4_raw  <- aoi_HurdleData$riparian_area_m2
UDA_m4_raw   <- aoi_HurdleData$catchment_km2
#damDen <- aoi_HurdleData$dam_km_std
damDen_m4_raw <- aoi_HurdleData$dam_km
damDen_m4_mean_raw <- mean(damDen_m4_raw)
damDen_m4_sd_raw   <- sd(damDen_m4_raw)

logSWA_m4_mean <- mean(aoi_HurdleData$logSWA)
logSWA_m4_sd   <- sd(aoi_HurdleData$logSWA)

# Standardization stats
incis_m4_mean_raw <- mean(incis_m4_raw)
incis_m4_sd_raw   <- sd(incis_m4_raw)

RIPA_m4_mean_raw  <- mean(RIPA_m4_raw)
RIPA_m4_sd_raw    <- sd(RIPA_m4_raw)

UDA_m4_mean_raw   <- mean(UDA_m4_raw)
UDA_m4_sd_raw     <- sd(UDA_m4_raw)


##############################3

# Helper Functions to Compute SWA from posterior draws
compute_swa_draws_damDen <- function(incis, RIPA, UDA, post, damDen_val) {
  n_draws <- length(post$alpha)
  swa_mat <- matrix(NA_real_, nrow = length(incis), ncol = n_draws)

  for (j in seq_len(n_draws)) {
    lp <- post$alpha[j] +
          post$beta_incis[j] * incis +
          post$beta_RIPA[j]  * RIPA +
          post$beta_UDA[j]   * UDA +
          post$beta_damDen[j] * damDen_val

    swa_mat[, j] <- exp(lp) - 1
  }

  swa_mat
}

summarize_swa <- function(swa_mat) {
  tibble(
    SWA_med = apply(swa_mat, 1, median),
    SWA_lo  = apply(swa_mat, 1, quantile, probs = 0.025),
    SWA_hi  = apply(swa_mat, 1, quantile, probs = 0.975)
  )
}




#########################################

# PDP Builder:

build_pdp_damDen <- function(focal_raw, focal_name,
                      incis_raw, RIPA_raw, UDA_raw,
                      post,
                      incis_mean_raw, incis_sd_raw,
                      RIPA_mean_raw, RIPA_sd_raw,
                      UDA_mean_raw, UDA_sd_raw,
                      damDen_raw) {

  # Sequence only across empirical range
  focal_seq <- seq(min(focal_raw), max(focal_raw), length.out = 200)

  # Conditional medians of other covariates
  incis_med <- median(incis_raw)
  RIPA_med  <- median(RIPA_raw)
  UDA_med   <- median(UDA_raw)
  
  # Empirical dam density range (standardized)
  damDen_seq <- seq(min(damDen_raw), max(damDen_raw), length.out = 10)

  # Build grid (200 rows)
  grid <- tibble(
    focal_raw = focal_seq,
    incis_raw = if (focal_name == "incis") focal_seq else incis_med,
    RIPA_raw  = if (focal_name == "RIPA")  focal_seq else RIPA_med,
    UDA_raw   = if (focal_name == "UDA")   focal_seq else UDA_med
  ) %>%
    mutate(
      incis = (incis_raw - incis_mean_raw) / incis_sd_raw,
      RIPA  = (RIPA_raw  - RIPA_mean_raw)  / RIPA_sd_raw,
      UDA   = (UDA_raw   - UDA_mean_raw)   / UDA_sd_raw
    )
  # Compute PDP curves for each dam density value
  pdp_list <- map(damDen_seq, function(dd) {
    swa_mat <- compute_swa_draws_damDen(grid$incis, grid$RIPA, grid$UDA, post,
                                        (dd - damDen_m4_mean_raw) / damDen_m4_sd_raw)
    summarize_swa(swa_mat) %>% mutate(damDen = dd)
  })

  pdp <- bind_rows(pdp_list)

  # Repeat grid for each damDen value
  grid2 <- bind_rows(replicate(length(damDen_seq), grid, simplify = FALSE))

  bind_cols(grid2, pdp)
}



#####

pdp_incis_m4_damDen <- build_pdp_damDen(
  focal_raw = incis_m4_raw,
  focal_name = "incis",
  incis_m4_raw, RIPA_m4_raw, UDA_m4_raw,
  damDen_m4_post,
  incis_m4_mean_raw, incis_m4_sd_raw,
  RIPA_m4_mean_raw, RIPA_m4_sd_raw,
  UDA_m4_mean_raw, UDA_m4_sd_raw, damDen_m4_raw
)

pdp_RIPA_m4_damDen <- build_pdp_damDen(
  focal_raw = RIPA_m4_raw,
  focal_name = "RIPA",
  incis_m4_raw, RIPA_m4_raw, UDA_m4_raw,
  damDen_m4_post,
  incis_m4_mean_raw, incis_m4_sd_raw,
  RIPA_m4_mean_raw, RIPA_m4_sd_raw,
  UDA_m4_mean_raw, UDA_m4_sd_raw, damDen_m4_raw
)

pdp_UDA_m4_damDen <- build_pdp_damDen(
  focal_raw = UDA_m4_raw,
  focal_name = "UDA",
  incis_m4_raw, RIPA_m4_raw, UDA_m4_raw,
  damDen_m4_post,
  incis_m4_mean_raw, incis_m4_sd_raw,
  RIPA_m4_mean_raw, RIPA_m4_sd_raw,
  UDA_m4_mean_raw, UDA_m4_sd_raw, damDen_m4_raw
)



##########################################3

# Plot Function:

plot_pdp_damDen <- function(pdp, xvar, xtitle) {
  ggplot(pdp,
         aes(x = .data[[xvar]], y = SWA_med,
             color = damDen, fill = damDen)) +
    #geom_ribbon(aes(ymin = SWA_lo, ymax = SWA_hi), alpha = 0.15, color = NA) +
    geom_point() +
    scale_color_viridis_c(option = "plasma") +
    scale_fill_viridis_c(option = "plasma") +
    scale_y_continuous(labels = comma) +
    labs(
      title = paste("Partial Dependence of SWA on dam density and", xtitle),
      #subtitle = "Across empirical dam density range",
      x = xtitle,
      y = "Predicted SWA (m^2)",
      color = "Dam Density (dams/km)",
      fill  = "Dam Density (dams/km)"
    ) +
    theme_minimal(base_size = 14)
}



###########################################


plot_pdp_damDen(pdp_incis_m4_damDen, "incis_raw", "Channel Incision (raw)")
plot_pdp_damDen(pdp_RIPA_m4_damDen,  "RIPA_raw",  "Riparian Area (m²)")
plot_pdp_damDen(pdp_UDA_m4_damDen,   "UDA_raw",   "Contributing Drainage Area (km²)")


```

----

**Change in SWA across damDensity by stream and sampleEvent**

```{r,echo=FALSE}
m4_dam_seq_raw <- seq(min(aoi_HurdleData$dam_km),
                      max(aoi_HurdleData$dam_km),
                      length.out = 50)

m4_dam_seq_scaled <- (m4_dam_seq_raw - damDen_m4_mean_raw) / damDen_m4_sd_raw

m4_pdp_stream <- map_dfr(1:n_stream, function(j) {

  alpha_j <- damDen_m4_post$alpha

  beta_j <- damDen_m4_post$beta_damDen +
            damDen_m4_post$beta_stream[, j] +
            rowMeans(damDen_m4_post$beta_event) +
            rowMeans(damDen_m4_post$beta_year)

  eta_j <- damDen_m4_post$eta_stream[, j] +
           rowMeans(damDen_m4_post$eta_event) +
           rowMeans(damDen_m4_post$eta_year)

  map2_dfr(m4_dam_seq_raw, m4_dam_seq_scaled, function(d_raw, d_scaled) {

    mu_draws <- alpha_j + beta_j * d_scaled + eta_j
    swa_draws <- exp(mu_draws) - 1

    tibble(
      stream = j,
      damDen = d_raw,
      SWA_med = median(swa_draws),
      SWA_lo  = quantile(swa_draws, 0.025),
      SWA_hi  = quantile(swa_draws, 0.975)
    )
  })
})

stream_lookup <- aoi_HurdleData %>%
  distinct(stream, watershed) %>%
  arrange(stream)

m4_pdp_stream <- m4_pdp_stream %>%
  left_join(stream_lookup, by = "stream")

ggplot(m4_pdp_stream, aes(x = watershed, y = SWA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "SWA Conditional on Stream Identity and Dam Density",
    subtitle = "Predicted SWA across dam density, conditioning on stream random effects and holding geomorphic covariates and event random effectrs at mean values",
    x = "Stream",
    y = "Predicted SWA (m²)",
    color = "Dam Density (dams/km)"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

###############################################################
################################################################
m4_pdp_event <- map_dfr(1:n_event, function(j) {

  alpha_j <- damDen_m4_post$alpha

  beta_j <- damDen_m4_post$beta_damDen +
            damDen_m4_post$beta_event[, j] +
            damDen_m4_post$beta_year[, event_year_3_4[j]] +
            rowMeans(damDen_m4_post$beta_stream)

  eta_j <- damDen_m4_post$eta_event[, j] +
           damDen_m4_post$eta_year[, event_year_3_4[j]] +
           rowMeans(damDen_m4_post$eta_stream)

  map2_dfr(m4_dam_seq_raw, m4_dam_seq_scaled, function(d_raw, d_scaled) {

    mu_draws <- alpha_j + beta_j * d_scaled + eta_j
    swa_draws <- exp(mu_draws) - 1

    tibble(
      event = j,
      damDen = d_raw,
      SWA_med = median(swa_draws),
      SWA_lo  = quantile(swa_draws, 0.025),
      SWA_hi  = quantile(swa_draws, 0.975)
    )
  })
})

event_year_lookup <- tibble(
  event = 1:12,
  event_label = c(
    "2019_Aug_1","2019_Sep_1","2019_Aug_2","2019_Sep_2",
    "2021_Aug_1","2021_Sep_1","2021_Aug_2","2021_Sep_2",
    "2023_Aug_1","2023_Sep_1","2023_Aug_2","2023_Sep_2"
  )
)

m4_pdp_event_named <- m4_pdp_event %>%
  left_join(event_year_lookup, by = "event")

ggplot(m4_pdp_event_named, aes(x = event_label, y = SWA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "SWA Conditional on Sample Event and Dam Density",
    subtitle = "Predicted SWA across dam density, conditioning on sample event random effects and holding geomorphic covariates and stream random effects at mean values",
    x = "Sample Event",
    y = "Predicted SWA (m²)",
    color = "Dam Density (dams/km)"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```





----

----

**Model Comparison**
```{r,echo=FALSE,warning=FALSE}

# Extract log-likelihood matrices
ll_m1 <- hurdle_damPres_m1_SW$sims.list$loglik
ll_m2 <- hurdle_damPres_m2_SW$sims.list$loglik
ll_m3 <- hurdle_damDen_m3_SW$sims.list$loglik
ll_m4 <- hurdle_damDen_m4_SW$sims.list$loglik

# Compute WAIC
waic_m1 <- waic(ll_m1)
waic_m2 <- waic(ll_m2)
waic_m3 <- waic(ll_m3)
waic_m4 <- waic(ll_m4)

# Combine into comparison table
waic_table <- tibble(
  model = c("m1", "m2", "m3", "m4"),
  WAIC  = c(waic_m1$estimates["waic", "Estimate"],
            waic_m2$estimates["waic", "Estimate"],
            waic_m3$estimates["waic", "Estimate"],
            waic_m4$estimates["waic", "Estimate"])
) %>%
  arrange(WAIC)

waic_table

```
