---
title: "VegAreaModels"
author: "Matt Webster"
date: "2026-07-28"
output: html_document
editor_options: 
  chunk_output_type: console
---



The purpose of this RMarkdown is to implement hierarchical generalized linear mixed effects models for VegArea ~ damPresence, & VegArea ~ damDensity (with and without geomorphic covariates).

4 models total.
    
    WAIC indicates that all four models perform comparably well. Given this, I only include code and results for m4, which models the effect of dam density on vegetation area, conditional on geomorphic covariates. 
    
    
    
Apply "Model 4" to veg data at each buffer size; 0,10,20,30-m. 1 model for each buffer size. Then compare model outputs for each. How do posterior estimates change across buffers?    
    
    
----

**HGLMM**


    This analysis explores late August vegetation area in relation to beaver dam presence,
    & density.
    
    2 models for dam presence (1 without geomorphic covariates; 1 with geomorphic covariates)
    2 models for dam density (...^^^...)
    
    All 4 models presented herein are constructed in a hierarchical generalized linear mixed
    effects model framework. Stream, year, and sample clusterID are included as random intercepts
    for baseline vegetation area. Random slopes were removed from this analysis, due to lack of 
    group-level variation to support model complexity. See "Ch1_Model_Data_Setup.Rmd". For
    the same reason, (lack of variation), sampleEvent was removed as a random effect - no
    notable difference in VA among sampleEvents.
    
    Models herein:
    
      - VA ~ damPresence (no geom.covar.)
      - VA ~ damPresence (geom.covar.)
      - VA ~ damDensity (no geom.covar.)
      - VA ~ damDensity (geom.covar.)
    
    Model structure is simplified from part 2 from the surface water hurdle models. Model 
    should be structurally parallel with NDVI model.
    
    Uses log(VegArea) for for 6 watersheds, & 3 years.
    
    
----
    

```{r,echo=FALSE,}
suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(ggplot2)
  library(dataRetrieval)
  library(knitr)
  library(ggplot2)
  library(zoo) #for moving window
  library(stringr)
  library(vioplot)
  library(purrr)
  library(lavaan)
  library(blavaan)
  library(ggcorrplot)
  library(lme4)
  library(broom.mixed)
  library(coda)
  library(scales)
  library(loo)
  library(jagsUI)
})
```


----


**Final Data:**

    NOTE: This ONLY uses data from 08/01-08/15. This is to avoid issues with pseudo-replication after the omission of random effects for sample event.
    
    Modeling reflects late August effect of beaver dams on vegetation area. Riparian vegetation = mesic resource that is critical for grouse recruitment.

```{r,echo=FALSE}
aoi_HurdleData <- read.csv("aoi_HurdleData_091026Update.csv") %>% 
  filter(t == 6) # Late August



# Full Hurdle data
N <- nrow(aoi_HurdleData)

#y <- aoi_HurdleData$y
#y2 <- aoi_HurdleData$logSWA
#y_veg <- aoi_HurdleData$y_veg

y_vegA_0 <- aoi_HurdleData$VA_0

y_vegA_10 <- aoi_HurdleData$VA_10

y_vegA_20 <- aoi_HurdleData$VA_20

y_vegA_30 <- aoi_HurdleData$VA_30

stream <- aoi_HurdleData$stream
year <- aoi_HurdleData$year_id
event <- aoi_HurdleData$sampleEvent_new
cluster <- as.numeric(as.factor(aoi_HurdleData$clusterID_unique))
damPres <- aoi_HurdleData$damPresence
damDen <- as.numeric(aoi_HurdleData$dam_km_std)
damDen_raw <- aoi_HurdleData$dam_km

n_stream <- length(unique(aoi_HurdleData$stream))
n_year <- length(unique(aoi_HurdleData$year_id))
n_event <- length(unique(aoi_HurdleData$sampleEvent_new))
n_cluster <- length(unique(as.numeric(as.factor(aoi_HurdleData$clusterID_unique))))

```

----

    Right skew of vegetation area data warrants log-transformation prior to HGLMM.
    
    Because vegetation is almost always present in our sample reaches, a
    hurdle model is not necessary here -- no need to model the presence/absence 
    process if it's almost always present. 
      
      

```{r,echo=FALSE}
par(mfrow = c(2,1))
hist(aoi_HurdleData$VA_0)

hist(log(aoi_HurdleData$VA_0+1))
```


----


For vegetation models, data of interest include:

     - y_vegA_...: log-transformed vegetation area log(VegArea[i]+1) for a given buffer; response variable
    
    - stream: index for random intercept on VegArea
    
    - year: index for random intercept on VegArea
    
    - clusterID_seq: index for random intercept on VegArea; represents sample pairing
    
    - damPres: dam presence (0,1); fixed effect on VegArea
    
    - damDen: dams/km; z-standardized ("...$dam_km_std"); fixed effect on VegArea
    
    Only used in models 2 and 4:
    
    - incis: channel incision: 1-(TO/90) (inverse of topographic openness); z-standardized; fixed effect on 
    VegArea
    
    - RIPA: riparian area (m^2); z-standardized; fixed effect on VegArea 
    
    - UDA: upstream drainage area (km ^2); z-standardized; fixed effect on VegArea


----


**Overall Takeaways**

    Dam presence is associated with ~25% increase in log() vegetation area
    
    1 SD increase in dam density is associated with ~13% increase in log() vegetation area.
    
    1 SD increase in riparian area is associated with a 25-30% increase in log() vegetation area.
    
    Incision and upstream drainage area have negligible influence on vegetation area.
    
    Still lots of variation among streams and years. Note the annual differences and the lag effect of early summer 2023 flooding.



**(HOW) DOES THIS PATTERN CHANGE ACROSS BUFFER SIZES?!!!**
----

    RUN m4 FOR EACH BUFFER TREATMENT, TO FIND OUT

----

----

----

----

----

Vegetation Area ~ Dam Presence
----

----

----

**Model 1: VegArea ~ DamPresence**

Effect of damPresence without geomorphic covariates

----

```{r,echo=FALSE}
library(jagsUI)
```
    sink("HGLMM_m1_Veg_damPresence.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR VEG AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
    
      sigma_eta_year    ~ dnorm(0, 1) T(0,)
      tau_eta_year      <- pow(sigma_eta_year, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
      
      alpha ~ dnorm(9.07, 0.04)   # within 5SD of empirical mean
      beta_damPres ~ dnorm(0,0.04)
    
      ##########################################################################
      
      # RANDOM EFFECT PRIORS FOR VEG AREA:
    
      for(j in 1:n_stream){
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        eta_year[yr] ~ dnorm(0, tau_eta_year)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
     
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Amount:
        y[i] ~ dnorm(mu[i], tau)
    
        mu[i] <- alpha
                    + beta_damPres * damPres[i]
                    + eta_stream[stream[i]]
                    + eta_year[year[i]]
                    + eta_cluster[cluster[i]]
      
      ##########################################################################
                    
      # Derived parameter for WAIC computation:

      loglik[i] <- logdensity.norm(y[i], mu[i], tau)

      }
    
    }
    ", fill = TRUE)
    
    sink()

    

```{r,echo=FALSE}
hglmm_m1_veg_damPres_data_list <- list(
  N = N,
  y = log(y_vegA + 1),
  stream = stream,
  year = year,
  cluster = cluster,
  damPres = damPres,
  n_stream = n_stream,
  n_year = n_year,
  n_cluster = n_cluster
)

hglmm_m1_veg_damPres_inits <- function(){
  list(  )
}

hglmm_m1_veg_damPres_params <- c(
  "mu",
  "alpha",
  "beta_damPres",
  "eta_stream",
  "eta_year",
  "eta_cluster",
  "sigma",
  "sigma_eta_stream",
  "sigma_eta_year",
  "sigma_eta_cluster",
  "loglik"
)

nchains <- 3
nthin <- 5
niter <- 100000
nburn <- 25000

Sys.time()

set.seed(12345)

hglmm_m1_veg_damPres <- jags(data = hglmm_m1_veg_damPres_data_list,
                   inits = hglmm_m1_veg_damPres_inits,
                   parameters.to.save = hglmm_m1_veg_damPres_params,
                   model.file = "HGLMM_m1_Veg_damPresence.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

```

<!---
----

DIAGNOSTICS:

Extract posteriors:
```{r,echo=FALSE}
damPres_m1_veg_post <- hglmm_m1_veg_damPres$sims.list
```

"Chain health" Rhat + ESS:
```{r,echo=FALSE}
# Choose only the parameters you care about
params_m1_veg <- c(
  "alpha",
  "beta_damPres",
  "sigma")

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m1_veg_damPres$samples[, params_m1_veg], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m1_veg_damPres$samples[, params_m1_veg])

```

----

    Rhat values of 1 are ideal.
    
    ESS looks good.


----

**Check traceplots** for critical parameters:

```{r,echo=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

tplot_m1_veg_params <- c("alpha","beta_damPres","sigma", "sigma_eta_stream")


m1_mcmc_veg_sub <- hglmm_m1_veg_damPres$samples[, tplot_m1_veg_params]

coda::traceplot(m1_mcmc_veg_sub)
```

----

    Similarly, traceplots indicate solid convergence.

----

**Posterior predictive plots:**

Model-estimated vegArea vs. observed vegArea

Should be about 1

```{r,echo=FALSE}
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

# Posterior mean of predicted log(VA + 1)
m1_mu_veg_post <- apply(hglmm_m1_veg_damPres$sims.list$mu, 2, mean)

# Observed log vegetation area
m1_obs_logVA <- log(y_vegA + 1)

# Fit regression on the log scale
m1_fit_log <- lm(m1_obs_logVA ~ m1_mu_veg_post)

m1_slope_log <- coef(m1_fit_log)[2]
m1_R2_log    <- summary(m1_fit_log)$r.squared

# Plot
plot(m1_mu_veg_post,
     m1_obs_logVA,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated log(VA + 1)",
     ylab = "Observed log(VA + 1)")

abline(0, 1, col = "red", lwd = 2)

```

Slope:
```{r,echo=FALSE}
m1_slope_log
```

----
    
    Comparison of model-estimated log-transformed vegetation area with observed log-
    transformed vegetation area. "Perfect" model fit would have a slope = 1. A slope
    of 1.1 is reasonable, given model complexity. This indicates that our model
    slightly underestimates the observed variation in log(vegArea+1) [especially at 
    low VA values], but overall it performs well.

----

**Random effects shrinkage: **

Should be centered around 0(ish)

1. eta_stream; random effect of stream on vegArea intercept (n=6)
2. eta_year; random effect of year on vegArea intercept (n=3)

```{r, echo=FALSE}

par(mfrow = c(2, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hglmm_m1_veg_damPres$sims.list$eta_stream)

boxplot(hglmm_m1_veg_damPres$sims.list$eta_year)
```

----

**Posterior mean dam effects:**

damPres -> vegArea (no geom.covar.)

```{r,echo=FALSE}
m1_coeffparams <- c(#"alpha",
            "beta_damPres",
            "sigma")

m1_coef_df <- hglmm_m1_veg_damPres$summary[m1_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m1_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.2,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE}
summary(hglmm_m1_veg_damPres$sims.list$beta_damPres)

exp(mean(hglmm_m1_veg_damPres$sims.list$beta_damPres))
```

```{r,echo=FALSE}
#library(tidyverse)
#library(scales)
par(mfrow = c(1, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

#Posterior distribution of damEffect on vegArea:

hist(hglmm_m1_veg_damPres$sims.list$beta_damPres,
     breaks = 50,
     col = "steelblue",
     main = "Posterior of damPresence effect on log(VA + 1)",
     xlab = "beta_damPres")
abline(v = median(hglmm_m1_veg_damPres$sims.list$beta_damPres), col = "red", lwd = 2)
```
    
    Without consideration of geomorphic covariates, dam presence is associated with a
    ~28-32% increase in log-transformed vegetation area.


----

**Prior-posterior overlap:**

```{r,echo=FALSE}
damPres_m1_veg_alpha_post <- hglmm_m1_veg_damPres$sims.list$alpha
damPres_m1_veg_beta_post  <- hglmm_m1_veg_damPres$sims.list$beta_damPres

damPres_m1_veg_Mod_n <- length(damPres_m1_veg_alpha_post)

damPres_m1_veg_alpha_prior <- rnorm(damPres_m1_veg_Mod_n, 9.07, sqrt(1/0.04))
damPres_m1_veg_beta_prior  <- rnorm(damPres_m1_veg_Mod_n, 0, sqrt(1/0.04))


# Plot overlap:


plot_prior_post <- function(prior, post, name) {
  
  df <- tibble(
    value = c(prior, post),
    dist  = rep(c("Prior", "Posterior"), each = length(prior))
  )
  
  ggplot(df, aes(x = value, fill = dist)) +
    geom_density(alpha = 0.4) +
    scale_fill_manual(values = c("gray60", "steelblue")) +
    labs(
      title = paste("Prior vs Posterior for", name),
      x = name,
      y = "Density"
    ) +
    theme_minimal(base_size = 14)
}

plot_prior_post(damPres_m1_veg_alpha_prior, damPres_m1_veg_alpha_post, "alpha")
plot_prior_post(damPres_m1_veg_beta_prior, damPres_m1_veg_beta_post, "beta_damPres")

```

----

    PPO indicates that the model is not being driven by prior specifications

----




**Predicted VegArea with and without Dam Presence**


```{r,echo=FALSE}

m1_n_draws <- length(damPres_m1_veg_post$alpha)

# Random effects averaged out
m1_mu_no_dam <- damPres_m1_veg_post$alpha +
             damPres_m1_veg_post$eta_stream[ , stream] +
             damPres_m1_veg_post$eta_year[ , year] +
             damPres_m1_veg_post$eta_cluster[ , cluster]

m1_mu_yes_dam <- damPres_m1_veg_post$alpha +
              damPres_m1_veg_post$beta_damPres +
              damPres_m1_veg_post$eta_stream[ , stream] +
              damPres_m1_veg_post$eta_year[ , year] +
              damPres_m1_veg_post$eta_cluster[ , cluster]


m1_VA_no_dam  <- exp(m1_mu_no_dam) - 1
m1_VA_yes_dam <- exp(m1_mu_yes_dam) - 1


m1_VA_no_dam_vec  <- as.numeric(m1_VA_no_dam)
m1_VA_yes_dam_vec <- as.numeric(m1_VA_yes_dam)


m1_VA_no_dam_med  <- apply(m1_VA_no_dam, 2, median)
m1_VA_yes_dam_med <- apply(m1_VA_yes_dam, 2, median)

m1_VA_df <- tibble(
  VA = c(m1_VA_no_dam_med, m1_VA_yes_dam_med),
  Dam = factor(rep(c("No Dam", "Dam Present"),
                   each = length(m1_VA_no_dam_med)))
)


ggplot(m1_VA_df, aes(x = Dam, y = VA, fill = Dam)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8) +
  scale_fill_manual(values = c("steelblue", "brown4")) +
  labs(
    title = "Predicted Vegetation Area by Dam Presence",
    y = "Vegetation Area (m^2)",
    x = ""
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")
```

--->
----

----

----

----

----

----

----

**Model 2: VegArea ~ DamPresence + geom.covar.**

Effect of damPresence WITH geomorphic covariates

----

```{r,echo=FALSE}
#library(jagsUI)
```
    sink("HGLMM_m2_Veg_damPresence.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR VEG AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
      
      sigma_eta_year   ~ dnorm(0, 1) T(0,)
      tau_eta_year     <- pow(sigma_eta_year, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
       
      alpha ~ dnorm(9.07,0.04)  # anchored within 5 SD of empirical mean
      beta_damPres ~ dnorm(0,0.04)
      beta_incis ~ dnorm(0,0.04)
      beta_RIPA ~ dnorm(0,0.04)
      beta_UDA ~ dnorm(0,0.04)
    
      ##########################################################################
      
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        eta_year[yr] ~ dnorm(0, tau_eta_year)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
     
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Amount:
        y[i] ~ dnorm(mu[i], tau)
    
        mu[i] <- alpha
                    + beta_damPres * damPres[i]
                    + beta_incis * incis[i] 
                    + beta_RIPA * RIPA[i]
                    + beta_UDA * UDA[i]
                    + eta_stream[stream[i]]
                    + eta_year[year[i]]
                    + eta_cluster[cluster[i]]
                    
      ##########################################################################
                    
      # Derived parameter for WAIC computation:

      loglik[i] <- logdensity.norm(y[i], mu[i], tau)
      }
    
    }
    ", fill = TRUE)
    
    sink()

    

```{r,echo=FALSE}
hglmm_m2_veg_damPres_data_list <- list(
  N = N,
  y = log(y_vegA + 1),
  stream = stream,
  year = year,
  cluster = cluster,
  damPres = damPres,
  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)),
  n_stream = n_stream,
  n_year = n_year,
  n_cluster = n_cluster
)

hglmm_m2_veg_damPres_inits <- function(){
  list()
}

hglmm_m2_veg_damPres_params <- c(
  "mu",
  "alpha",
  "beta_damPres",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "eta_stream",
  "eta_year",
  "eta_cluster",
  "sigma",
  "sigma_eta_stream",
  "sigma_eta_year",
  "sigma_eta_cluster",
  "loglik"
)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000

Sys.time()

set.seed(12345)

hglmm_m2_veg_damPres <- jags(data = hglmm_m2_veg_damPres_data_list,
                   inits = hglmm_m2_veg_damPres_inits,
                   parameters.to.save = hglmm_m2_veg_damPres_params,
                   model.file = "HGLMM_m2_Veg_damPresence.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hglmm_m1_veg_damPres) # model with local and watershed dam density as fixed effects
#print(hglmm_m1_veg_damPres$summary[grep("beta_", rownames(hurdle_damPres_SW$summary)), ])

```

<!---

----

DIAGNOSTICS:

Extract posteriors:
```{r,echo=FALSE}
damPres_m2_veg_post <- hglmm_m2_veg_damPres$sims.list
```

"Chain health" Rhat + ESS:
```{r,echo=FALSE}
# Choose only the parameters you care about
params_m2_veg <- c(
  "alpha",
  "beta_damPres",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "sigma")

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m2_veg_damPres$samples[, params_m2_veg], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m2_veg_damPres$samples[, params_m2_veg])

```

----

    Rhat = 1
    ESS = high

----

**Check traceplots** for critical parameters:

```{r,echo=FALSE}
par(mfrow = c(2, 4),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

tplot_m2_veg_params <- c("alpha","beta_damPres",
                         "beta_incis","beta_RIPA",
                         "beta_UDA", "sigma", "sigma_eta_stream")


m2_mcmc_veg_sub <- hglmm_m2_veg_damPres$samples[, tplot_m2_veg_params]

coda::traceplot(m2_mcmc_veg_sub)
```

----

    Grassy.
    

----

**Posterior predictive plots:**

Model-estimated vegArea vs. observed vegArea

Should be about 1

```{r,echo=FALSE}
par(mfrow = c(1,1),        # 1 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

# Posterior mean of predicted log(VA + 1)
m2_mu_veg_post <- apply(hglmm_m2_veg_damPres$sims.list$mu, 2, mean)

# Observed log vegetation area
m2_obs_logVA <- log(y_vegA + 1)

# Fit regression on the log scale
m2_fit_log <- lm(m2_obs_logVA ~ m2_mu_veg_post)

m2_slope_log <- coef(m2_fit_log)[2]
m2_R2_log    <- summary(m2_fit_log)$r.squared

# Plot
plot(m2_mu_veg_post,
     m2_obs_logVA,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated log(VA + 1)",
     ylab = "Observed log(VA + 1)")

abline(0, 1, col = "red", lwd = 2)
```

Slope:
```{r,echo=FALSE}
m2_slope_log
```


----

    Comparison of model-estimated log-transformed vegetation area with observed log-
    transformed vegetation area. "Perfect" model fit would have a slope = 1. A slope
    of 1.1 is good, given model complexity. This indicates that our model
    underestimates the observed variation in log(vegArea+1), but overall
    performs well. Highest disagreement at lower observed veg areas. Slight improvement
    from model 1

----

**Random effects shrinkage: **

Should be centered around 0(ish)

1. eta_stream; random effect of stream on vegArea
2. eta_year; random effect of year on vegArea

```{r, echo=FALSE}

par(mfrow = c(2, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hglmm_m2_veg_damPres$sims.list$eta_stream)

boxplot(hglmm_m2_veg_damPres$sims.list$eta_year)
```

----

**Posterior mean dam effects:**

damPres -> vegArea (w/geom.covar.)


```{r,echo=FALSE}
m2_coeffparams <- c(#"alpha",
            "beta_damPres",
            "beta_incis",
            "beta_RIPA",
            "beta_UDA",
            "sigma")

m2_coef_df <- hglmm_m2_veg_damPres$summary[m2_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m2_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.2,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```


```{r,echo=FALSE}
summary(hglmm_m2_veg_damPres$sims.list$beta_damPres)

exp(mean(hglmm_m2_veg_damPres$sims.list$beta_damPres))
```


```{r,echo=FALSE}
#library(tidyverse)
#library(scales)

par(mfrow = c(1, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  
#Posterior distribution of damEffect on vegArea:

hist(hglmm_m2_veg_damPres$sims.list$beta_damPres,
     breaks = 50,
     col = "steelblue",
     main = "Posterior distribution of damPres effect on log(VA + 1)",
     xlab = "beta_damPres",
     xlim = c(-0.2,0.75))
# Add median line
abline(v = mean(hglmm_m2_veg_damPres$sims.list$beta_damPres), col = "red", lwd = 2)

```
    
    When you also consider site-level geomorphic conditions, dam presence is associated 
    with an estimated ~22-25% increase in log-transformed vegetation area. More conservative than M1.
    Still nicely identified, and not overlapping 0. 

----
----

Prior-posterior overlap:

```{r,echo=FALSE}
damPres_m2_veg_alpha_post <- hglmm_m2_veg_damPres$sims.list$alpha
damPres_m2_veg_beta_post  <- hglmm_m2_veg_damPres$sims.list$beta_damPres
damPres_m2_veg_betaIncis_post <- hglmm_m2_veg_damPres$sims.list$beta_incis

damPres_m2_veg_Mod_n <- length(damPres_m2_veg_alpha_post)

damPres_m2_veg_alpha_prior <- rnorm(damPres_m2_veg_Mod_n, 9.07, sqrt(1/0.04))
damPres_m2_veg_beta_prior  <- rnorm(damPres_m2_veg_Mod_n, 0, sqrt(1/0.04))
damPres_m2_veg_betaIncis_prior <- rnorm(damPres_m2_veg_Mod_n, 0, sqrt(1/0.04))


# Plot overlap:

#plot_prior_post <- function(prior, post, name) {
  
#  df <- tibble(
#    value = c(prior, post),
#    dist  = rep(c("Prior", "Posterior"), each = length(prior))
#  )
  
#  ggplot(df, aes(x = value, fill = dist)) +
#    geom_density(alpha = 0.4) +
#    scale_fill_manual(values = c("gray60", "steelblue")) +
#    labs(
#      title = paste("Prior vs Posterior for", name),
#      x = name,
#      y = "Density"
#    ) +
#    theme_minimal(base_size = 14)
#}

plot_prior_post(damPres_m2_veg_alpha_prior, damPres_m2_veg_alpha_post, "alpha")
plot_prior_post(damPres_m2_veg_beta_prior, damPres_m2_veg_beta_post, "beta_damPres")
plot_prior_post(damPres_m2_veg_betaIncis_prior, damPres_m2_veg_betaIncis_post,"beta_incis")

```

----

    PPOs still look good. 


**Predicted VegArea with and without Dam Presence**

```{r,echo=FALSE}

m2_n_draws <- length(damPres_m2_veg_post$alpha)

# Random effects averaged out
m2_mu_no_dam <- damPres_m2_veg_post$alpha +
             damPres_m2_veg_post$eta_stream[ , stream] +
             damPres_m2_veg_post$eta_year[ , year] +
             damPres_m2_veg_post$eta_cluster[ , cluster] +
             damPres_m2_veg_post$beta_incis +
             damPres_m2_veg_post$beta_RIPA +
             damPres_m2_veg_post$beta_UDA

m2_mu_yes_dam <- damPres_m2_veg_post$alpha +
              damPres_m2_veg_post$beta_damPres +
              damPres_m2_veg_post$eta_stream[ , stream] +
              damPres_m2_veg_post$eta_year[ , year] +
              damPres_m2_veg_post$eta_cluster[ , cluster] +
              damPres_m2_veg_post$beta_incis +
              damPres_m2_veg_post$beta_RIPA +
              damPres_m2_veg_post$beta_UDA


m2_VA_no_dam  <- exp(m2_mu_no_dam) - 1
m2_VA_yes_dam <- exp(m2_mu_yes_dam) - 1


m2_VA_no_dam_vec  <- as.numeric(m2_VA_no_dam)
m2_VA_yes_dam_vec <- as.numeric(m2_VA_yes_dam)


m2_VA_no_dam_med  <- apply(m2_VA_no_dam, 2, median)
m2_VA_yes_dam_med <- apply(m2_VA_yes_dam, 2, median)

m2_VA_df <- tibble(
  VA = c(m2_VA_no_dam_med, m2_VA_yes_dam_med),
  Dam = factor(rep(c("No Dam", "Dam Present"),
                   each = length(m2_VA_no_dam_med)))
)


ggplot(m2_VA_df, aes(x = Dam, y = VA, fill = Dam)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, alpha = 0.8) +
  scale_fill_manual(values = c("steelblue", "brown4")) +
  labs(
    title = "Predicted Vegetation Area by Dam Presence",
    y = "Vegetation Area (m^2)",
    x = ""
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")
```

----

----

----

----


**Partial dependence plots**

Constrain PDP to empirical ranges for VA and target covariates

```{r,echo=FALSE}

names(damPres_m2_veg_post)

# Raw covariates
incis_raw <- 1 - aoi_HurdleData$TO/90
RIPA_raw  <- aoi_HurdleData$riparian_area_m2
UDA_raw   <- aoi_HurdleData$catchment_km2

logVA_mean <- mean(aoi_HurdleData$logVA)
logVA_sd   <- sd(aoi_HurdleData$logVA)

# Standardization stats
incis_mean_raw <- mean(incis_raw)
incis_sd_raw   <- sd(incis_raw)

RIPA_mean_raw  <- mean(RIPA_raw)
RIPA_sd_raw    <- sd(RIPA_raw)

UDA_mean_raw   <- mean(UDA_raw)
UDA_sd_raw     <- sd(UDA_raw)


##############################3

# Helper Functions to Compute VA from posterior draws
compute_va_draws <- function(incis, RIPA, UDA, post, dam_flag) {
  n_draws <- length(damPres_m2_veg_post$alpha)
  va_mat <- matrix(NA_real_, nrow = length(incis), ncol = n_draws)

  for (j in seq_len(n_draws)) {
    lp <- post$alpha[j] +
          post$beta_incis[j] * incis +
          post$beta_RIPA[j]  * RIPA +
          post$beta_UDA[j]   * UDA +
          post$beta_damPres[j] * dam_flag

    va_mat[, j] <- exp(lp) - 1
  }

  va_mat
}

summarize_va <- function(va_mat) {
  tibble(
    VA_med = apply(va_mat, 1, median),
    VA_lo  = apply(va_mat, 1, quantile, probs = 0.025),
    VA_hi  = apply(va_mat, 1, quantile, probs = 0.975)
  )
}




#########################################

# PDP Builder:

build_pdp <- function(focal_raw, focal_name,
                      incis_raw, RIPA_raw, UDA_raw,
                      post,
                      incis_mean_raw, incis_sd_raw,
                      RIPA_mean_raw, RIPA_sd_raw,
                      UDA_mean_raw, UDA_sd_raw) {

  # Sequence only across empirical range
  focal_seq <- seq(min(focal_raw), max(focal_raw), length.out = 200)

  # Conditional medians of other covariates
  incis_med <- median(incis_raw)
  RIPA_med  <- median(RIPA_raw)
  UDA_med   <- median(UDA_raw)

  # Build grid (200 rows)
  grid <- tibble(
    focal_raw = focal_seq,
    incis_raw = if (focal_name == "incis") focal_seq else incis_med,
    RIPA_raw  = if (focal_name == "RIPA")  focal_seq else RIPA_med,
    UDA_raw   = if (focal_name == "UDA")   focal_seq else UDA_med
  ) %>%
    mutate(
      incis = (incis_raw - incis_mean_raw) / incis_sd_raw,
      RIPA  = (RIPA_raw  - RIPA_mean_raw)  / RIPA_sd_raw,
      UDA   = (UDA_raw   - UDA_mean_raw)   / UDA_sd_raw
    )

  # Compute VA draws
  va_nodam <- compute_va_draws(grid$incis, grid$RIPA, grid$UDA, post, dam_flag = 0)
  va_dam   <- compute_va_draws(grid$incis, grid$RIPA, grid$UDA, post, dam_flag = 1)

  # Summaries (400 rows)
  pdp <- bind_rows(
    summarize_va(va_nodam) %>% mutate(dam = "No Dam"),
    summarize_va(va_dam)   %>% mutate(dam = "Dam Present")
  )

  # Repeat grid twice (400 rows)
  grid2 <- bind_rows(grid, grid)

  # Bind columns safely
  bind_cols(grid2, pdp)
}



#####

m2_pdp_incis <- build_pdp(
  focal_raw = incis_raw,
  focal_name = "incis",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_m2_veg_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)

m2_pdp_RIPA <- build_pdp(
  focal_raw = RIPA_raw,
  focal_name = "RIPA",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_m2_veg_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)

m2_pdp_UDA <- build_pdp(
  focal_raw = UDA_raw,
  focal_name = "UDA",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_m2_veg_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)



##########################################

# Plot Function:

plot_pdp <- function(pdp, xvar, xtitle) {
  ggplot(pdp,
         aes(x = .data[[xvar]], y = VA_med, color = dam, fill = dam)) +
    geom_ribbon(aes(ymin = VA_lo, ymax = VA_hi), alpha = 0.2, color = NA) +
    geom_line(linewidth = 1.2) +   # updated from size = 1.2
    scale_color_manual(values = c("steelblue", "brown4")) +
    scale_fill_manual(values = c("steelblue", "brown4")) +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = paste("Partial Dependence of VA on", xtitle),
      subtitle = "Predicted VegArea with 95% credible intervals\nConstrained to empirical covariate range",
      x = xtitle,
      y = "Predicted Vegetation Area (m^2)",
      color = "Dam Presence",
      fill  = "Dam Presence"
    ) +
    theme_minimal(base_size = 14)
}



###########################################3

plot_pdp(m2_pdp_incis, "incis_raw", "Channel Incision (raw)")
plot_pdp(m2_pdp_RIPA,  "RIPA_raw",  "Riparian Area (m^2)")
plot_pdp(m2_pdp_UDA,   "UDA_raw",   "Contributing Drainage Area (km^2)")


```


----

----

----

----
--->

----

----

----

----

----



VegArea ~ Dam Density
----


**Hurdle HGLMM**


**MODEL 3: VegArea ~ damDensity (no Geom.Covar.)**

```{r,echo=FALSE}
#library(jagsUI)
```
    sink("HGLMM_m3_Veg_damDensity.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
      
      sigma_eta_year   ~ dnorm(0, 1) T(0,)
      tau_eta_year     <- pow(sigma_eta_year, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      alpha ~ dnorm(9.07,0.04)  # within 5 SD of empirical mean log(VA+1)
      beta_damDen ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR Veg AMOUNT:
    
      for(j in 1:n_stream){
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        eta_year[yr] ~ dnorm(0, tau_eta_year)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
      
      for(i in 1:N){
        
        y[i] ~ dnorm(mu[i], tau)
      
        mu[i] <- alpha
                    + beta_damDen * damDen[i]
                    + eta_stream[stream[i]]
                    + eta_year[year[i]]
                    + eta_cluster[cluster[i]]
                    
      ##########################################################################
                    
      # Derived parameter for WAIC computation:
      
      loglik[i] <- logdensity.norm(y[i], mu[i], tau)
      }
    }
    ", fill = TRUE)
    
    sink()

    

```{r,echo=FALSE}
hglmm_damDen_m3_data_list <- list(
  N = N,
  y = log(y_vegA + 1),
  stream = stream,
  year = year,
  cluster = cluster,
  damDen = damDen,
  n_stream = n_stream,
  n_year = n_year,
  n_cluster = n_cluster
)

hglmm_damDen_m3_inits <- function(){
  list()
}

hglmm_damDen_m3_params <- c(
  "alpha",
  "mu",
  "beta_damDen",
  "eta_stream",
  "eta_year",
  "eta_cluster",
  "sigma",
  "sigma_eta_stream",
  "sigma_eta_year",
  "sigma_eta_cluster",
  "loglik"
)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000


Sys.time()

set.seed(12345)

hglmm_m3_damDen_Veg <- jags(data = hglmm_damDen_m3_data_list,
                   inits = hglmm_damDen_m3_inits,
                   parameters.to.save = hglmm_damDen_m3_params,
                   model.file = "HGLMM_m3_Veg_damDensity.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()


```

<!---
----


DIAGNOSTICS:
----


```{r,echo=FALSE}
# Extract posterior draws

damDen_m3_post <- hglmm_m3_damDen_Veg$sims.list 
```

**Assess "Chain health" Rhat + ESS:**

```{r,echo=FALSE} 
# Choose only the parameters you care about
params_m3 <- c(
  "alpha",
  "beta_damDen",
  "sigma")

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m3_damDen_Veg$samples[, params_m3], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m3_damDen_Veg$samples[, params_m3])

```

    Rhat & Effective Sample Size:
    
    Rhat values look good. All 1.00.
    
    Effective sample sizes look ok for all parameters.

----

**Check traceplots** for critical parameters:

```{r,echo=FALSE} 

par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


tplot_params_m3 <- c("alpha","beta_damDen","sigma","sigma_eta_stream")


mcmc_sub_m3 <- hglmm_m3_damDen_Veg$samples[, tplot_params_m3]

coda::traceplot(mcmc_sub_m3)
```


    Traceplots:
    
    Traceplots all look sufficiently "grassy," indicating sufficient chain convergence.


----

**Predicted log(VA) vs observed log(VA)**

Goal is 1:1 (ish)

```{r,echo=FALSE} 
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 

# Posterior mean of predicted log(VA + 1)
m3_mu_veg_post <- apply(hglmm_m3_damDen_Veg$sims.list$mu, 2, mean)

# Observed log vegetation area
m3_obs_logVA <- log(y_vegA + 1)

# Fit regression on the log scale
m3_fit_log <- lm(m3_obs_logVA ~ m3_mu_veg_post)

m3_slope_log <- coef(m3_fit_log)[2]
m3_R2_log    <- summary(m3_fit_log)$r.squared

# Plot
plot(m3_mu_veg_post,
     m3_obs_logVA,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated log(VA + 1)",
     ylab = "Observed log(VA + 1)")

abline(0, 1, col = "red", lwd = 2)
```

Slope:
```{r,echo=FALSE}
m3_slope_log
```

    1.1 again. See above comment re: underestimation of variation at lower log(VA+1)
    
----

**Random effects shrinkage: **

Goal is centered around 0(ish)

1. eta_stream; random effect of stream on veg area; n=6
2. eta_year; random effect of year on veg area; n=3

```{r, echo=FALSE} 
par(mfrow = c(2, 1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hglmm_m3_damDen_Veg$sims.list$eta_stream)

boxplot(hglmm_m3_damDen_Veg$sims.list$eta_year)
```


----

----

**Posterior mean dam density effect**

damDen -> vegArea (no geom.covar.)

```{r,echo=FALSE}
m3_coeffparams <- c(#"alpha",
            "beta_damDen",
            "sigma")

m3_coef_df <- hglmm_m3_damDen_Veg$summary[m3_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m3_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.2,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```


```{r,echo=FALSE}
summary(hglmm_m3_damDen_Veg$sims.list$beta_damDen)

exp(mean(hglmm_m3_damDen_Veg$sims.list$beta_damDen))
```

```{r,echo=FALSE}
#library(tidyverse)
#library(scales)

par(mfrow = c(1, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  
#Posterior distribution of damEffect on vegArea:

hist(hglmm_m3_damDen_Veg$sims.list$beta_damDen,
     breaks = 50,
     col = "steelblue",
     main = "Posterior distribution of damDensity effect on log(VA + 1)",
     xlab = "beta_damDen",
     xlim = c(0,0.4))
# Add median line
abline(v = mean(hglmm_m3_damDen_Veg$sims.list$beta_damDen), col = "red", lwd = 2)

```

    Without geomorphic covariates, a 1 SD increase in dam density corresponds with a ~ 15% increase in log() vegetation area.


----

**Prior-posterior overlap:**

```{r,echo=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 

damDen_alpha_post_m3 <- hglmm_m3_damDen_Veg$sims.list$alpha
damDen_beta_post_m3  <- hglmm_m3_damDen_Veg$sims.list$beta_damDen

damDenMod_m3_n <- length(damDen_alpha_post_m3)

damDen_alpha_prior_m3 <- rnorm(damDenMod_m3_n, 9.07, sqrt(1/0.04))
damDen_beta_prior_m3  <- rnorm(damDenMod_m3_n, 0, sqrt(1/0.04))


# Plot overlap:

#plot_prior_post <- function(prior, post, name) {
  
#  df <- tibble(
#    value = c(prior, post),
#    dist  = rep(c("Prior", "Posterior"), each = length(prior))
#  )
  
#  ggplot(df, aes(x = value, fill = dist)) +
#    geom_density(alpha = 0.4) +
#    scale_fill_manual(values = c("gray60", "steelblue")) +
#    labs(
#      title = paste("Prior vs Posterior for", name),
#      x = name,
#      y = "Density"
#    ) +
#    theme_minimal(base_size = 14)
#}

plot_prior_post(damDen_alpha_prior_m3, damDen_alpha_post_m3, "alpha_m3")
plot_prior_post(damDen_beta_prior_m3, damDen_beta_post_m3, "beta_damDen_m3")
```


    PPOs still look pretty good.

----

**Change in VA across damDensity by stream and Year**

```{r,echo=FALSE}

m3_dam_seq <- seq(min(aoi_HurdleData$dam_km),
               max(aoi_HurdleData$dam_km),
               length.out = 50)

# build PDP grid
m3_pdp_stream <- map_dfr(1:n_stream, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m3_post$alpha
  beta_j  <- damDen_m3_post$beta_damDen
  eta_j   <- damDen_m3_post$eta_stream[, j] + rowMeans(damDen_m3_post$eta_year)

  map_dfr(m3_dam_seq, function(d) {

    mu_draws <- alpha_j + beta_j * d + eta_j
    va_draws <- exp(mu_draws) - 1

    tibble(
      stream = j,
      damDen = d,
      VA_med = median(va_draws),
      VA_lo  = quantile(va_draws, 0.025),
      VA_hi  = quantile(va_draws, 0.975)
    )
  })
})

stream_lookup <- aoi_HurdleData %>%
  distinct(stream, watershed) %>%
  arrange(stream)

m3_pdp_stream <- m3_pdp_stream %>%
  left_join(stream_lookup, by = "stream")



# plot
ggplot(m3_pdp_stream, aes(x = watershed, y = VA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Partial Dependence of VA on Stream and Dam Density",
    x = "Stream",
    y = "Predicted VA (m^2)",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


###############################################################
################################################################

# build PDP grid
m3_pdp_year <- map_dfr(1:n_year, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m3_post$alpha
  beta_j  <- damDen_m3_post$beta_damDen
  eta_j   <- damDen_m3_post$eta_year[, j] + rowMeans(damDen_m3_post$eta_stream)

  map_dfr(m3_dam_seq, function(d) {

    mu_draws <- alpha_j + beta_j * d + eta_j
    va_draws <- exp(mu_draws) - 1

    tibble(
      year = j,
      damDen = d,
      VA_med = median(va_draws),
      VA_lo  = quantile(va_draws, 0.025),
      VA_hi  = quantile(va_draws, 0.975)
    )
  })
})

year_lookup <- tibble(
  year = 1:3,
  year_label = c(
    "2019",
    "2021",
    "2023"
  )
)


m3_pdp_year_named <- m3_pdp_year %>%
  left_join(year_lookup, by = "year")



# plot
ggplot(m3_pdp_year_named, aes(x = year_label, y = VA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Partial Dependence of VA on Year and Dam Density",
    x = "Year",
    y = "Predicted VA (m^2)",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```

--->
----

----

----

----

----



**MODEL 4: SW ~  damDensity + Geom.Covar.**

```{r,echo=FALSE}
#library(jagsUI)
```
    sink("HGLMM_m4_Veg_damDensity.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      ##########################################################################
      # HYPERPRIORS FOR Vegetation Area (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
    
      sigma_eta_year    ~ dnorm(0, 1) T(0,)
      tau_eta_year      <- pow(sigma_eta_year, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Amount:
      alpha ~ dnorm(9.07,0.04)  # within 5 SD of empirical mean vegArea
      beta_damDen ~ dnorm(0,0.04)
      beta_incis ~ dnorm(0,0.04)
      beta_RIPA ~ dnorm(0,0.04)
      beta_UDA ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR VegArea:
    
      for(j in 1:n_stream){
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        eta_year[yr] ~ dnorm(0, tau_eta_year)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
      
        y[i] ~ dnorm(mu[i], tau) 
        
        mu[i] <- alpha
                    + beta_damDen * damDen[i]
                    + beta_incis * incis[i] 
                    + beta_RIPA * RIPA[i]
                    + beta_UDA * UDA[i]
                    + eta_stream[stream[i]]
                    + eta_year[year[i]]
                    + eta_cluster[cluster[i]]
                    
      ##########################################################################
                    
      # Derived parameter for WAIC computation:

      loglik[i] <- logdensity.norm(y[i], mu[i], tau)
      }
    
    }
    ", fill = TRUE)
    
    sink()
 



M4_0-m Buffer:
----

```{r,echo=FALSE}
#hglmm_damDen_m4_0_data_list <- list(
 # N = N,
#  y = log(y_vegA_0 + 1),
 # stream = stream,
#  year = year,
 # cluster = cluster,
#  damDen = damDen,
 # incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
#  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2_0)), #RIPA of DEM-derived pondableZone
 # UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)), # Upstream Drainage Area
#  n_stream = n_stream,
 # n_year = n_year,
#  n_cluster = n_cluster
#)

#hglmm_damDen_m4_inits <- function(){
#  list()
#}

#hglmm_damDen_m4_params <- c(
#  "alpha",
#  "mu",
#  "beta_damDen",
#  "beta_incis",
#  "beta_RIPA",
#  "beta_UDA",
#  "eta_stream",
#  "eta_year",
#  "eta_cluster",
#  "sigma",
#  "sigma_eta_stream",
#  "sigma_eta_event",
#  "sigma_eta_cluster",
#  "loglik"
#)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000


#Sys.time()

#set.seed(12345)

#hglmm_m4_0_damDen_Veg <- jags(data = hglmm_damDen_m4_0_data_list,
#                   inits = hglmm_damDen_m4_inits,
#                   parameters.to.save = hglmm_damDen_m4_params,
#                   model.file = "HGLMM_m4_Veg_damDensity.jags",
#                   parallel = T,
#                   n.chains = nchains,
#                   n.thin = nthin,
#                   n.iter = niter,
#                   n.burnin = nburn)


#Sys.time()

#print(hglmm_m4_0_damDen_Veg) # model with local and watershed dam density as fixed effects
#print(hglmm_m4_0_damDen_Veg$summary[grep("beta_", rownames(hglmm_m4_0_damDen_Veg$summary)), ])
#print(hglmm_m4_0_damDen_Veg$summary[grep("zeta_", rownames(hglmm_m4_0_damDen_Veg$summary)), ])

```


----

M4_10-m Buffer:
----

```{r,echo=FALSE}
#hglmm_damDen_m4_10_data_list <- list(
#  N = N,
# y = log(y_vegA_10 + 1),
#  stream = stream,
#  year = year,
#  cluster = cluster,
#  damDen = damDen,
#  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
##  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2_0)), #RIPA of DEM-derived pondableZone
#  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)), # Upstream Drainage Area
#  n_stream = n_stream,
#  n_year = n_year,
#  n_cluster = n_cluster
#)

#hglmm_damDen_m4_inits <- function(){
#  list()
#}

#hglmm_damDen_m4_params <- c(
#  "alpha",
#  "mu",
#  "beta_damDen",
#  "beta_incis",
#  "beta_RIPA",
#  "beta_UDA",
#  "eta_stream",
#  "eta_year",
#  "eta_cluster",
#  "sigma",
#  "sigma_eta_stream",
#  "sigma_eta_event",
#  "sigma_eta_cluster",
#  "loglik"
#)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000


#Sys.time()

#set.seed(12345)

#hglmm_m4_10_damDen_Veg <- jags(data = hglmm_damDen_m4_10_data_list,
 #                  inits = hglmm_damDen_m4_inits,
#                   parameters.to.save = hglmm_damDen_m4_params,
 #                  model.file = "HGLMM_m4_Veg_damDensity.jags",
#                   parallel = T,
 #                  n.chains = nchains,
#                   n.thin = nthin,
 #                  n.iter = niter,
#                   n.burnin = nburn)


#Sys.time()

#print(hglmm_m4_0_damDen_Veg) # model with local and watershed dam density as fixed effects
#print(hglmm_m4_10_damDen_Veg$summary[grep("beta_", rownames(hglmm_m4_10_damDen_Veg$summary)), ])
#print(hglmm_m4_10_damDen_Veg$summary[grep("zeta_", rownames(hglmm_m4_10_damDen_Veg$summary)), ])

```

----

M4_20-m Buffer:
----

```{r,echo=FALSE}
#hglmm_damDen_m4_20_data_list <- list(
#  N = N,
#  y = log(y_vegA_20 + 1),
#  stream = stream,
#  year = year,
#  cluster = cluster,
#  damDen = damDen,
#  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
#  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2_0)), #RIPA of DEM-derived pondableZone
#  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)), # Upstream Drainage Area
#  n_stream = n_stream,
#  n_year = n_year,
#  n_cluster = n_cluster
#)

#hglmm_damDen_m4_inits <- function(){
#  list()
#}

#hglmm_damDen_m4_params <- c(
#  "alpha",
#  "mu",
#  "beta_damDen",
#  "beta_incis",
#  "beta_RIPA",
#  "beta_UDA",
#  "eta_stream",
#  "eta_year",
#  "eta_cluster",
#  "sigma",
#  "sigma_eta_stream",
#  "sigma_eta_event",
#  "sigma_eta_cluster",
#  "loglik"
#)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000


#Sys.time()

#set.seed(12345)

#hglmm_m4_20_damDen_Veg <- jags(data = hglmm_damDen_m4_20_data_list,
#                   inits = hglmm_damDen_m4_inits,
#                   parameters.to.save = hglmm_damDen_m4_params,
#                   model.file = "HGLMM_m4_Veg_damDensity.jags",
#                   parallel = T,
#                   n.chains = nchains,
##                   n.thin = nthin,
#                   n.iter = niter,
#                   n.burnin = nburn)


#Sys.time()

#print(hglmm_m4_20_damDen_Veg) # model with local and watershed dam density as fixed effects
#print(hglmm_m4_20_damDen_Veg$summary[grep("beta_", rownames(hglmm_m4_20_damDen_Veg$summary)), ])
#print(hglmm_m4_20_damDen_Veg$summary[grep("zeta_", rownames(hglmm_m4_20_damDen_Veg$summary)), ])

```

----

M4_30-m Buffer:
----

```{r,echo=FALSE}
#hglmm_damDen_m4_30_data_list <- list(
#  N = N,
#  y = log(y_vegA_30 + 1),
#  stream = stream,
#  year = year,
#  cluster = cluster,
#  damDen = damDen,
#  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
#  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2_0)), #RIPA of DEM-derived pondableZone
#  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)), # Upstream Drainage Area
#  n_stream = n_stream,
#  n_year = n_year,
#  n_cluster = n_cluster
#)

#hglmm_damDen_m4_inits <- function(){
#  list()
#}

#hglmm_damDen_m4_params <- c(
#  "alpha",
#  "mu",
#  "beta_damDen",
#  "beta_incis",
#  "beta_RIPA",
#  "beta_UDA",
#  "eta_stream",
#  "eta_year",
#  "eta_cluster",
#  "sigma",
#  "sigma_eta_stream",
#  "sigma_eta_event",
#  "sigma_eta_cluster",
#  "loglik"
#)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000


#Sys.time()

#set.seed(12345)

#hglmm_m4_30_damDen_Veg <- jags(data = hglmm_damDen_m4_30_data_list,
#                   inits = hglmm_damDen_m4_inits,
#                   parameters.to.save = hglmm_damDen_m4_params,
#                   model.file = "HGLMM_m4_Veg_damDensity.jags",
#                   parallel = T,
#                   n.chains = nchains,
#                   n.thin = nthin,
#                   n.iter = niter,
#                   n.burnin = nburn)


#Sys.time()

#print(hglmm_m4_30_damDen_Veg) # model with local and watershed dam density as fixed effects
#print(hglmm_m4_30_damDen_Veg$summary[grep("beta_", rownames(hglmm_m4_30_damDen_Veg$summary)), ])
#print(hglmm_m4_30_damDen_Veg$summary[grep("zeta_", rownames(hglmm_m4_30_damDen_Veg$summary)), ])

```



```{r,echo=FALSE}
# SAVE THE MODEL OUTPUTS SO YOU DON'T HAVE TO RUN THEM EVERY TIME!!

#saveRDS(hglmm_m4_0_damDen_Veg,
#        "C:\Users\mattw\Desktop\Ch1_ModelOutputs\hglmm_m4_0_damDen_Veg.rds")

#saveRDS(hglmm_m4_10_damDen_Veg,
#        "C:\Users\mattw\Desktop\Ch1_ModelOutputs\hglmm_m4_10_damDen_Veg.rds")

#saveRDS(hglmm_m4_20_damDen_Veg,
#        "C:\Users\mattw\Desktop\Ch1_ModelOutputs\hglmm_m4_20_damDen_Veg.rds")

#saveRDS(hglmm_m4_30_damDen_Veg,
#        "C:\Users\mattw\Desktop\Ch1_ModelOutputs\hglmm_m4_30_damDen_Veg.rds")


#hglmm_m4_0_damDen_Veg <- readRDS("C:\Users\mattw\Desktop\Ch1_ModelOutputs\hglmm_m4_0_damDen_Veg.rds")

#hglmm_m4_10_damDen_Veg <- readRDS("C:\Users\mattw\Desktop\Ch1_ModelOutputs\hglmm_m4_10_damDen_Veg.rds")

#hglmm_m4_20_damDen_Veg <- readRDS("C:\Users\mattw\Desktop\Ch1_ModelOutputs\hglmm_m4_20_damDen_Veg.rds")

#hglmm_m4_30_damDen_Veg <- readRDS("C:\Users\mattw\Desktop\Ch1_ModelOutputs\hglmm_m4_30_damDen_Veg.rds")
```



----









DIAGNOSTICS:
----


```{r,echo=FALSE}
# Extract posterior draws

damDen_m4_0_post <- hglmm_m4_0_damDen_Veg$sims.list 

damDen_m4_10_post <- hglmm_m4_10_damDen_Veg$sims.list 

damDen_m4_20_post <- hglmm_m4_20_damDen_Veg$sims.list 

damDen_m4_30_post <- hglmm_m4_30_damDen_Veg$sims.list 

```

**Assess "Chain health" Rhat + ESS:**

**0-m Buff:**

```{r,echo=FALSE} 
# Choose only the parameters you care about
params_m4 <- c(
  "alpha",
  "beta_damDen",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "sigma")

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m4_0_damDen_Veg$samples[, params_m4], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m4_0_damDen_Veg$samples[, params_m4])

```

    Rhat & Effective Sample Size:
    
    Solid. Rhats near ____; ESS ____
    
    
    


**10-m Buff:**

```{r,echo=FALSE} 

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m4_10_damDen_Veg$samples[, params_m4], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m4_10_damDen_Veg$samples[, params_m4])

```

    Rhat & Effective Sample Size:
    
    Solid. Rhats near ____; ESS ____
    
    
    
**20-m Buff:**

```{r,echo=FALSE} 

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m4_20_damDen_Veg$samples[, params_m4], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m4_20_damDen_Veg$samples[, params_m4])

```

    Rhat & Effective Sample Size:
    
    Solid. Rhats near ____; ESS ____




**30-m Buff:**

```{r,echo=FALSE} 

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m4_30_damDen_Veg$samples[, params_m4], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m4_30_damDen_Veg$samples[, params_m4])

```

    Rhat & Effective Sample Size:
    
    Solid. Rhats near ____; ESS ____
    
    
    
<!---
----

**Check traceplots** for critical parameters:

```{r,echo=FALSE} 

par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


tplot_params_m4 <- c("alpha","beta_damDen","beta_incis","beta_RIPA","beta_UDA",
                     "sigma","sigma_eta_stream")


mcmc_sub_m4_0 <- hglmm_m4_0_damDen_Veg$samples[, tplot_params_m4]

coda::traceplot(mcmc_sub_m4_0)
```


    Traceplots all look good.


----


**Predicted log(VA) vs observed log(VA)**

Goal is 1:1 (ish)

```{r,echo=FALSE} 
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 

# Posterior mean of predicted log(VA + 1)
m4_0_mu_veg_post <- apply(hglmm_m4_0_damDen_Veg$sims.list$mu, 2, mean)

# Observed log vegetation area
m4_0_obs_logVA <- log(y_vegA_0 + 1)

# Fit regression on the log scale
m4_0_fit_log <- lm(m4_0_obs_logVA ~ m4_0_mu_veg_post)

m4_0_slope_log <- coef(m4_0_fit_log)[2]
m4_0_R2_log    <- summary(m4_0_fit_log)$r.squared

# Plot
plot(m4_0_mu_veg_post,
     m4_0_obs_logVA,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated log(VA_0 + 1)",
     ylab = "Observed log(VA_0 + 1)")

abline(0, 1, col = "red", lwd = 2)

```

Slope:
```{r,echo=FALSE}
m4_0_slope_log
```

    Near 1 = good. Noticeable underestimation of variance at low VA_0

----




**Random effects shrinkage: **

Goal is centered around 0(ish)

1. eta_year; random effect of year log(VA+1) intercept
2. eta_stream; random effect of stream on log(VA+1) intercept

```{r, echo=FALSE} 
par(mfrow = c(1, 2),        # rows, columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hglmm_m4_0_damDen_Veg$sims.list$eta_year)
boxplot(hglmm_m4_0_damDen_Veg$sims.list$eta_stream)
```


----

**Prior-posterior overlap:**

```{r,echo=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 

damDen_alpha_m4_0_post <- hglmm_m4_0_damDen_Veg$sims.list$alpha
damDen_beta_m4_0_post  <- hglmm_m4_0_damDen_Veg$sims.list$beta_damDen
damDen_beta_incis_m4_0_post <- hglmm_m4_0_damDen_Veg$sims.list$beta_incis
damDen_sigmaEtaStream_m4_0_post <- hglmm_m4_0_damDen_Veg$sims.list$sigma_eta_stream

damDenMod_m4_0_n <- length(damDen_alpha_m4_0_post)

damDen_alpha_m4_0_prior <- rnorm(damDenMod_m4_0_n, 0, sqrt(1/0.04))
damDen_beta_m4_0_prior  <- rnorm(damDenMod_m4_0_n, 0, sqrt(1/0.04))
damDen_beta_incis_m4_0_prior  <- rnorm(damDenMod_m4_0_n, 0, sqrt(1/0.04))
damDen_sigmaEtaStream_m4_0_prior <- pmax(rnorm(damDenMod_m4_0_n, 0, 1), 0)


# Plot overlap:

plot_prior_post <- function(prior, post, name) {
  
  df <- tibble(
    value = c(prior, post),
    dist  = rep(c("Prior", "Posterior"), each = length(prior))
  )
  
  ggplot(df, aes(x = value, fill = dist)) +
    geom_density(alpha = 0.4) +
    scale_fill_manual(values = c("gray60", "steelblue")) +
    labs(
      title = paste("Prior vs Posterior for", name),
      x = name,
      y = "Density"
    ) +
    theme_minimal(base_size = 14)
}

plot_prior_post(damDen_alpha_m4_0_prior, damDen_alpha_m4_0_post, "alpha_m4")
plot_prior_post(damDen_beta_m4_0_prior, damDen_beta_m4_0_post, "beta_damDen_m4")
plot_prior_post(damDen_beta_incis_m4_0_prior, damDen_beta_incis_m4_0_post, "beta_incis_m4")
plot_prior_post(damDen_sigmaEtaStream_m4_0_prior, damDen_sigmaEtaStream_m4_0_post, "sigma_eta_stream_m4")
```

    PPOs look good pretty good.

----
--->


**Posterior Estimates**


**0-m Buffer; i.e., DEM-derived 'pondable zone'**

```{r,echo=FALSE}

m4_0_alpha <- mean(hglmm_m4_0_damDen_Veg$sims.list$alpha)

m4_0_beta_dd <- mean(hglmm_m4_0_damDen_Veg$sims.list$beta_damDen)

m4_0_beta_in <- mean(hglmm_m4_0_damDen_Veg$sims.list$beta_incis)

m4_0_beta_rip <- mean(hglmm_m4_0_damDen_Veg$sims.list$beta_RIPA)

m4_0_beta_uda <- mean(hglmm_m4_0_damDen_Veg$sims.list$beta_UDA)

```

```{r,echo=FALSE}
m4_coeffparams <- c(#"alpha",
            "beta_damDen",
            "beta_incis",
            "beta_RIPA",
            "beta_UDA",
            "sigma")

m4_0_coef_df <- hglmm_m4_0_damDen_Veg$summary[m4_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m4_0_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.5,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    subtitle = "0-m Buffer; DEM-derived 'pondable zone'",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```

----

**Effect of 1 SD increase in dam density on vegetation area:**

```{r,echo=FALSE}
summary(hglmm_m4_0_damDen_Veg$sims.list$beta_damDen)

exp(mean(hglmm_m4_0_damDen_Veg$sims.list$beta_damDen))

VA_0_damDen_0 <- exp(m4_0_alpha)  # with geom covar. at z-scaled mean (0)
#VA_damDen_0

VA_0_damDen_1 <- exp(m4_0_alpha + 0.067)  #make sure this value is correct
#VA_damDen_1

print("VA_0 change with 1 SD increase in dam density")
(VA_0_damDen_1 - VA_0_damDen_0)

```

    A 1 SD increase dam density is associated with ~ 6-7% increase in log() vegetation area WITHIN 0-m BUFFER around DEM-delineated riparian/pondable zone.
    For a "typical" stream under mean geomorphic covariate conditions, this  5% corresponds
    to an increase of ~ 250 m^2 of riparian vegetation.
    
    Note: at the upper end of the observed dam density range, a typical stream at mean covariate 
    conditions could see an estimated increase of ~1500 m^2 of riparian vegetation
    
    
    Note: If we consider the mean surface area across dammed and un-dammed sample windows in the 0-m buffered riparian zone ~ 10200 m^2, ..., a 1-5 SD increase in dams km^-1 (~3-15 dams km-1) is associated with a ~ 2-15 % increase in the proportion of vegetation cover across the near-stream riparian zone
    
----

**Estimated effects of geomorphic covariates**

----

**Effect of 1 SD increase in incision on vegetation area:**

```{r,echo=FALSE}
summary(hglmm_m4_0_damDen_Veg$sims.list$beta_incis)

exp(mean(hglmm_m4_0_damDen_Veg$sims.list$beta_incis))


VA_0_incis_0 <- exp(m4_0_alpha)
#VA_incis_0

VA_0_incis_1 <- exp(m4_0_alpha - 0.056) # make sure this is correct
#VA_incis_1

print("VA change with 1 SD increase in incision")
(VA_0_incis_1 - VA_0_incis_0)
```

    A 1 SD increase in incision (~0.1) is associated with a 5.6% decrease in log() vegetation area. 
    For a typical stream at mean covariate conditions, this corresponds to an decrease in ~ 200 m^2. 
    
    Effect of incision on vegetation area is modestly negative, but heavily overlaps 0.
    

----
 
**Effect of 1 SD increase in pond-able space on vegetation area:**

```{r,echo=FALSE}
summary(hglmm_m4_0_damDen_Veg$sims.list$beta_RIPA)

exp(mean(hglmm_m4_0_damDen_Veg$sims.list$beta_RIPA))


VA_0_RIPA_0 <- exp(m4_0_alpha)
#VA_0_RIPA_0

VA_0_RIPA_1 <- exp(m4_0_alpha + 0.52)  # make sure this is correct
#VA_0_RIPA_1

print("VA change with 1 SD increase in riparian area")
(VA_0_RIPA_1 - VA_0_RIPA_0)
```


    A 1 SD increase in RIPA (11000 m^2) is associated with a ~ 52% increase in log() vegetation area WITHIN 0-m RIP BUFFER. 
    
    This corresponds with an increase of ~2500 m^2 (for typical stream under mean conditions).
    
    Considering that mean riparian_area of the 0-buff zone is ~10200 m^2, an increase of 2500 corresponds to ~25% increase.

----
    
**Effect of 1 SD increase in Upstream Drainage Area on vegetation area:**

```{r,echo=FALSE}
summary(hglmm_m4_0_damDen_Veg$sims.list$beta_UDA)

exp(mean(hglmm_m4_0_damDen_Veg$sims.list$beta_UDA))


VA_0_UDA_0 <- exp(m4_0_alpha)
#VA_UDA_0

VA_0_UDA_1 <- exp(m4_0_alpha - 0.45)   # make sure this is correct
#VA_UDA_1

print("VA change with 1 SD increase in drainage area")
(VA_0_UDA_1 - VA_0_UDA_0)

```

    A 1 SD increase in Upstream Drainage Area (UDA) (~1300km^2) is associated with a 45% decrease in 
    log() vegetation area. This corresponds to an decrease of ~ 1311 m^2 in VA_0 (m^2) for a 
    "typical" stream at mean covariate conditions. 
    
    Corresponds to a ~13% reduction in proportion of RIP zone covered (if median 0-Buff RIPzone is 5495).
    
    IMPORTANT NOTE: THIS 0-BUFF ZONE IS WHERE WATER IS PRESENT. UDA IS STRONGLY POSITIVELY ASSOCIATED WITH SWA IN THIS ZONE, SO NATURALLY IT WILL IMPACT VEG AREA (CAN'T HAVE VEG WHERE THERE'S WATER).
  

----

----

----

----

**10-m Planar Buffer around DEM-derived 'pondable zone'**

```{r,echo=FALSE}

m4_10_alpha <- mean(hglmm_m4_10_damDen_Veg$sims.list$alpha)

m4_10_beta_dd <- mean(hglmm_m4_10_damDen_Veg$sims.list$beta_damDen)

m4_10_beta_in <- mean(hglmm_m4_10_damDen_Veg$sims.list$beta_incis)

m4_10_beta_rip <- mean(hglmm_m4_10_damDen_Veg$sims.list$beta_RIPA)

m4_10_beta_uda <- mean(hglmm_m4_10_damDen_Veg$sims.list$beta_UDA)

```

```{r,echo=FALSE}
m4_10_coef_df <- hglmm_m4_10_damDen_Veg$summary[m4_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m4_10_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.5,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    subtitle = "10-m Planar Buffer",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```

----

**Effect of 1 SD increase in dam density on vegetation area:**

```{r,echo=FALSE}
summary(hglmm_m4_10_damDen_Veg$sims.list$beta_damDen)

exp(mean(hglmm_m4_10_damDen_Veg$sims.list$beta_damDen))

VA_10_damDen_0 <- exp(m4_10_alpha)  # with geom covar. at z-scaled mean (0)
#VA_10_damDen_0

VA_10_damDen_1 <- exp(m4_10_alpha + 0.08)  #make sure this value is correct
#VA_10_damDen_1

print("VA_10 change with 1 SD increase in dam density")
(VA_10_damDen_1 - VA_10_damDen_0)

```

    A 1 SD increase dam density is associated with ~ 8% increase in log() vegetation area WITHIN 10-m BUFFER around DEM-delineated riparian/pondable zone.
    
    For a "typical" stream under mean geomorphic covariate conditions, this  8% corresponds
    to an increase of ~ 560 m^2 of riparian vegetation.
    
    Note: at the upper end of the observed dam density range, a typical stream at mean covariate 
    conditions could see an estimated increase of ~3300 m^2 of riparian vegetation
    
    
    Note: Considering that the mean surface area of sample windows in the 10-m buffered riparian zone ~ 17300, ..., a 1-5 SD increase in dams km^-1 (~3-15 dams km-1) is associated with a ~ 3-19 % increase in the proportion of vegetation cover across that buffer space.
    
----

**Estimated effects of geomorphic covariates**

----

**Effect of 1 SD increase in incision on vegetation area:**

```{r,echo=FALSE}
summary(hglmm_m4_10_damDen_Veg$sims.list$beta_incis)

exp(mean(hglmm_m4_10_damDen_Veg$sims.list$beta_incis))


VA_10_incis_0 <- exp(m4_10_alpha)
#VA_incis_10

VA_10_incis_1 <- exp(m4_10_alpha - 0.019) # make sure this is correct
#VA_incis_1

print("VA change with 1 SD increase in incision")
(VA_10_incis_1 - VA_10_incis_0)
```

    A 1 SD increase in incision (~0.1) is associated with a ~2% decrease in log() vegetation area, within the 10-m RIP buffer. 
    For a typical stream at mean covariate conditions, this corresponds to a decrease in ~ 127 m^2. 
    
    Realistically, incision doesn't seem to influence vegetation area.
    
    <Weak negative effect, heavily overlapping 0.>

----
 
**Effect of 1 SD increase in riparian area on vegetation area:**

```{r,echo=FALSE}
summary(hglmm_m4_10_damDen_Veg$sims.list$beta_RIPA)

exp(mean(hglmm_m4_10_damDen_Veg$sims.list$beta_RIPA))


VA_10_RIPA_0 <- exp(m4_10_alpha)
#SWA_UDA_10

VA_10_RIPA_1 <- exp(m4_10_alpha + 0.37)  # make sure this is correct
#SWA_UDA_1

print("VA change with 1 SD increase in riparian area")
(VA_10_RIPA_1 - VA_10_RIPA_0)
```


    A 1 SD increase in RIPA (11000 m^2) is associated with a ~ 37% increase in log() vegetation area WITHIN 10-m RIP buffer 
    
    This corresponds with an increase of ~3020 m^2 (for typical stream under mean conditions).
    
    Considering that mean surface area of the 10-m buffered zone is ~17300 m^2, an increase of 3020 corresponds to ~17% increase in log() vegetation cover within that buffered space.

----
    
**Effect of 1 SD increase in Upstream Drainage Area on SW amount:**

```{r,echo=FALSE}
summary(hglmm_m4_10_damDen_Veg$sims.list$beta_UDA)

exp(mean(hglmm_m4_10_damDen_Veg$sims.list$beta_UDA))


VA_10_UDA_0 <- exp(m4_10_alpha)
#SWA_UDA_10

VA_10_UDA_1 <- exp(m4_10_alpha - 0.18)   # make sure this is correct
#SWA_UDA_1

print("VA change with 1 SD increase in drainage area")
(VA_10_UDA_1 - VA_10_UDA_0)

```

    A 1 SD increase in Upstream Drainage Area (UDA) (~1300km^2) is associated with an 18% decrease in 
    log() vegetation area. This corresponds to an decrease of ~ 1100 m^2 in VA_10 (m^2) for a 
    "typical" stream at mean covariate conditions. 
    
    Corresponds to a ~6% reduction in proportion of surface area covered (if at mean 10-Buff RIPzone area = 17300).
   

----

----

----

----

**20-m Planar Buffer around DEM-derived 'pondable zone'**

```{r,echo=FALSE}

m4_20_alpha <- mean(hglmm_m4_20_damDen_Veg$sims.list$alpha)

m4_20_beta_dd <- mean(hglmm_m4_20_damDen_Veg$sims.list$beta_damDen)

m4_20_beta_in <- mean(hglmm_m4_20_damDen_Veg$sims.list$beta_incis)

m4_20_beta_rip <- mean(hglmm_m4_20_damDen_Veg$sims.list$beta_RIPA)

m4_20_beta_uda <- mean(hglmm_m4_20_damDen_Veg$sims.list$beta_UDA)

```

```{r,echo=FALSE}
m4_20_coef_df <- hglmm_m4_20_damDen_Veg$summary[m4_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m4_20_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.5,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    subtitle = "20-m Planar Buffer",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```

----

**Effect of 1 SD increase in dam density on vegetation area:**

```{r,echo=FALSE}
summary(hglmm_m4_20_damDen_Veg$sims.list$beta_damDen)

exp(mean(hglmm_m4_20_damDen_Veg$sims.list$beta_damDen))

VA_20_damDen_0 <- exp(m4_20_alpha)  # with geom covar. at z-scaled mean (0)
#VA_damDen_20

VA_20_damDen_1 <- exp(m4_20_alpha + 0.10)  #make sure this value is correct
#VA_damDen_1

print("VA_20 change with 1 SD increase in dam density")
(VA_20_damDen_1 - VA_20_damDen_0)

```

    A 1 SD increase dam density is associated with ~ 10% increase in log() vegetation area WITHIN 20-m BUFFER around DEM-delineated riparian/pondable zone.
    
    For a "typical" stream under mean geomorphic covariate conditions, this  10% corresponds
    to an increase of ~ 930 m^2 of riparian vegetation.
    
    Note: at the upper end of the observed dam density range, a typical stream at mean covariate 
    conditions could see an estimated increase of ~ 5800 m^2 of riparian vegetation
    
    
    Note: Considering that the mean surface area of sample windows in the 20-m buffered riparian zone ~ 23600 m^2, ..., a 1-5 SD increase in dams km^-1 (~3-15 dams km-1) is associated with a ~ 4-25 % increase in the proportion of vegetation cover across that buffer space.
    
----

**Estimated effects of geomorphic covariates**

----

**Effect of 1 SD increase in incision on vegetation area:**

```{r,echo=FALSE}
summary(hglmm_m4_20_damDen_Veg$sims.list$beta_incis)

exp(mean(hglmm_m4_20_damDen_Veg$sims.list$beta_incis))


VA_20_incis_0 <- exp(m4_20_alpha)
#VA_incis_20

VA_20_incis_1 <- exp(m4_20_alpha - 0.016) # make sure this is correct
#VA_incis_1

print("VA change with 1 SD increase in incision")
(VA_20_incis_1 - VA_20_incis_0)
```

    A 1 SD increase in incision (~0.1) is associated with a ~2% decrease in log() vegetation area, within the 20-m RIP buffer. 
    For a typical stream at mean covariate conditions, this corresponds to a decrease in ~ 140 m^2. 
    
    Realistically, incision doesn't seem to influence vegetation area.
    
    <Weak negative effect, heavily overlapping 0.>

----
 
**Effect of 1 SD increase in riparian area on vegetation area:**

```{r,echo=FALSE}
summary(hglmm_m4_20_damDen_Veg$sims.list$beta_RIPA)

exp(mean(hglmm_m4_20_damDen_Veg$sims.list$beta_RIPA))


VA_20_RIPA_0 <- exp(m4_20_alpha)
#SWA_UDA_20

VA_20_RIPA_1 <- exp(m4_20_alpha + 0.30)  # make sure this is correct
#SWA_UDA_1

print("VA change with 1 SD increase in riparian area")
(VA_20_RIPA_1 - VA_20_RIPA_0)
```


    A 1 SD increase in RIPA (11000 m^2) is associated with a ~ 30% increase in log() vegetation area WITHIN 20-m RIP buffer 
    
    This corresponds with an increase of ~3100 m^2 (for typical stream under mean conditions).
    
    Considering that mean surface area of the 20-m buffered zone is ~23600 m^2, an increase of 3100 m^2 corresponds to ~13% increase in the proportion 20-m buffered space with veg cover.

----
    
**Effect of 1 SD increase in Upstream Drainage Area on SW amount:**

```{r,echo=FALSE}
summary(hglmm_m4_20_damDen_Veg$sims.list$beta_UDA)

exp(mean(hglmm_m4_20_damDen_Veg$sims.list$beta_UDA))


VA_20_UDA_0 <- exp(m4_20_alpha)
#SWA_UDA_20

VA_20_UDA_1 <- exp(m4_20_alpha - 0.077)   # make sure this is correct
#SWA_UDA_1

print("VA change with 1 SD increase in drainage area")
(VA_20_UDA_1 - VA_20_UDA_0)

```

    A 1 SD increase in Upstream Drainage Area (UDA) (~1300km^2) is associated with a 7-8% decrease in 
    log() vegetation area. This corresponds to an decrease of ~ 660 m^2 in VA_20 (m^2) for a 
    "typical" stream at mean covariate conditions. 
    
    Corresponds to a ~3% reduction in proportion of surface area covered (if mean 20-Buff RIPzone is 23600).




----

----

----

----

**30-m Planar Buffer around DEM-derived 'pondable zone'**

```{r,echo=FALSE}

m4_30_alpha <- mean(hglmm_m4_30_damDen_Veg$sims.list$alpha)

m4_30_beta_dd <- mean(hglmm_m4_30_damDen_Veg$sims.list$beta_damDen)

m4_30_beta_in <- mean(hglmm_m4_30_damDen_Veg$sims.list$beta_incis)

m4_30_beta_rip <- mean(hglmm_m4_30_damDen_Veg$sims.list$beta_RIPA)

m4_30_beta_uda <- mean(hglmm_m4_30_damDen_Veg$sims.list$beta_UDA)

```

```{r,echo=FALSE}
m4_30_coef_df <- hglmm_m4_30_damDen_Veg$summary[m4_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m4_30_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.5,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    subtitle = "30-m Planar Buffer",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```

----

**Effect of 1 SD increase in dam density on vegetation area:**

```{r,echo=FALSE}
summary(hglmm_m4_30_damDen_Veg$sims.list$beta_damDen)

exp(mean(hglmm_m4_30_damDen_Veg$sims.list$beta_damDen))

VA_30_damDen_0 <- exp(m4_30_alpha)  # with geom covar. at z-scaled mean (0)
#VA_damDen_30

VA_30_damDen_1 <- exp(m4_30_alpha + 0.11)  #make sure this value is correct
#VA_damDen_1

print("VA_30 change with 1 SD increase in dam density")
(VA_30_damDen_1 - VA_30_damDen_0)

```

    A 1 SD increase dam density is associated with ~ 11% increase in log() vegetation area WITHIN 30-m BUFFER around DEM-delineated riparian/pondable zone.
    
    For a "typical" stream under mean geomorphic covariate conditions, this  11% corresponds
    to an increase of ~ 1200 m^2 of riparian vegetation.
    
    Note: at the upper end of the observed dam density range, a typical stream at mean covariate 
    conditions could see an estimated increase of ~ 7500 m^2 of riparian vegetation
    
    
    Note: Considering that the mean surface area of sample windows in the 30-m buffered riparian zone ~ 29400, ..., a 1-5 SD increase in dams km^-1 (~3-15 dams km-1) is associated with a ~ 4-25% increase in the proportion of vegetation cover across that buffer space [same proportional change as with 20-m buffer].
    
----

**Estimated effects of geomorphic covariates**

----

**Effect of 1 SD increase in incision on vegetation area:**

```{r,echo=FALSE}
summary(hglmm_m4_30_damDen_Veg$sims.list$beta_incis)

exp(mean(hglmm_m4_30_damDen_Veg$sims.list$beta_incis))


VA_30_incis_0 <- exp(m4_30_alpha)
#VA_incis_30

VA_30_incis_1 <- exp(m4_30_alpha + 0.011) # make sure this is correct
#VA_incis_1

print("VA change with 1 SD increase in incision")
(VA_30_incis_1 - VA_30_incis_0)
```

    A 1 SD increase in incision (~0.1) is associated with a ~1% decrease in log() vegetation area, within the 30-m RIP buffer. 
    For a typical stream at mean covariate conditions, this corresponds to a decrease in ~ 114 m^2. 
    
    Realistically, incision doesn't seem to influence vegetation area.
    
    <Weak negative effect, heavily overlapping 0.>

----
 
**Effect of 1 SD increase in riparian area on vegetation area:**

```{r,echo=FALSE}
summary(hglmm_m4_30_damDen_Veg$sims.list$beta_RIPA)

exp(mean(hglmm_m4_30_damDen_Veg$sims.list$beta_RIPA))


VA_30_RIPA_0 <- exp(m4_30_alpha)
#SWA_UDA_30

VA_30_RIPA_1 <- exp(m4_30_alpha + 0.26)  # make sure this is correct
#SWA_UDA_1

print("VA change with 1 SD increase in riparian area")
(VA_30_RIPA_1 - VA_30_RIPA_0)
```


    A 1 SD increase in RIPA (11000 m^2) is associated with a ~ 27% increase in log() vegetation area WITHIN 30-m RIP buffer 
    
    This corresponds with an increase of ~3000 m^2 (for typical stream under mean conditions).
    
    Considering that mean surface area of the 30-m buffered zone is ~29400 m^2, an increase of ~3000 m^2 corresponds to ~10% increase.

----
    
**Effect of 1 SD increase in Upstream Drainage Area on SW amount:**

```{r,echo=FALSE}
summary(hglmm_m4_30_damDen_Veg$sims.list$beta_UDA)

exp(mean(hglmm_m4_30_damDen_Veg$sims.list$beta_UDA))


VA_30_UDA_0 <- exp(m4_30_alpha)
#SWA_UDA_30

VA_30_UDA_1 <- exp(m4_30_alpha - 0.035)   # make sure this is correct
#SWA_UDA_1

print("VA change with 1 SD increase in drainage area")
(VA_30_UDA_1 - VA_30_UDA_0)

```

    A 1 SD increase in Upstream Drainage Area (UDA) (~1300km^2) is associated with a 4% decrease in 
    vegetation area. This corresponds to an decrease of ~ 355 m^2 in VA_30 (m^2) for a 
    "typical" stream at mean covariate conditions. 
    
    Corresponds to a ~1% reduction in proportion of surface area covered (if mean 30-Buff RIPzone is ___).
    
    HEAVILY overlaps 0.



----

**Partial dependence plots**

Constrain PDP to empirical ranges for SWA and target covariates

```{r,echo=FALSE}
# Raw covariates
incis_m4_0_raw <- 1 - aoi_HurdleData$TO/90
RIPA_m4_0_raw  <- aoi_HurdleData$riparian_area_m2
UDA_m4_0_raw   <- aoi_HurdleData$catchment_km2
#damDen <- aoi_HurdleData$dam_km_std
damDen_m4_0_raw <- aoi_HurdleData$dam_km
damDen_m4_0_mean_raw <- mean(damDen_m4_raw)
damDen_m4_0_sd_raw   <- sd(damDen_m4_raw)

logVA_m4_0_mean <- mean(log(aoi_HurdleData$VA_0+1))
logVA_m4_0_sd   <- sd(log(aoi_HurdleData$VA_0+1))

# Standardization stats
incis_m4_0_mean_raw <- mean(incis_m4_0_raw)
incis_m4_0_sd_raw   <- sd(incis_m4_0_raw)

RIPA_m4_0_mean_raw  <- mean(RIPA_m4_0_raw)
RIPA_m4_0_sd_raw    <- sd(RIPA_m4_0_raw)

UDA_m4_0_mean_raw   <- mean(UDA_m4_0_raw)
UDA_m4_0_sd_raw     <- sd(UDA_m4_0_raw)


##############################3

# Helper Functions to Compute VA from posterior draws
compute_va_draws_damDen <- function(incis, RIPA, UDA, post, damDen_val) {
  n_draws <- length(post$alpha)
  va_mat <- matrix(NA_real_, nrow = length(incis), ncol = n_draws)

  for (j in seq_len(n_draws)) {
    lp <- post$alpha[j] +
          post$beta_incis[j] * incis +
          post$beta_RIPA[j]  * RIPA +
          post$beta_UDA[j]   * UDA +
          post$beta_damDen[j] * damDen_val

    va_mat[, j] <- exp(lp) - 1
  }

  va_mat
}

summarize_va <- function(va_mat) {
  tibble(
    VA_med = apply(va_mat, 1, median),
    VA_lo  = apply(va_mat, 1, quantile, probs = 0.025),
    VA_hi  = apply(va_mat, 1, quantile, probs = 0.975)
  )
}




#########################################

# PDP Builder:

build_pdp_damDen <- function(focal_raw, focal_name,
                      incis_raw, RIPA_raw, UDA_raw,
                      post,
                      incis_mean_raw, incis_sd_raw,
                      RIPA_mean_raw, RIPA_sd_raw,
                      UDA_mean_raw, UDA_sd_raw,
                      damDen_raw) {

  # Sequence only across empirical range
  focal_seq <- seq(min(focal_raw), max(focal_raw), length.out = 200)

  # Conditional medians of other covariates
  incis_med <- median(incis_raw)
  RIPA_med  <- median(RIPA_raw)
  UDA_med   <- median(UDA_raw)
  
  # Empirical dam density range (standardized)
  damDen_seq <- seq(min(damDen_raw), max(damDen_raw), length.out = 10)

  # Build grid (200 rows)
  grid <- tibble(
    focal_raw = focal_seq,
    incis_raw = if (focal_name == "incis") focal_seq else incis_med,
    RIPA_raw  = if (focal_name == "RIPA")  focal_seq else RIPA_med,
    UDA_raw   = if (focal_name == "UDA")   focal_seq else UDA_med
  ) %>%
    mutate(
      incis = (incis_raw - incis_mean_raw) / incis_sd_raw,
      RIPA  = (RIPA_raw  - RIPA_mean_raw)  / RIPA_sd_raw,
      UDA   = (UDA_raw   - UDA_mean_raw)   / UDA_sd_raw
    )
  # Compute PDP curves for each dam density value
  pdp_list <- map(damDen_seq, function(dd) {
    va_mat <- compute_va_draws_damDen(grid$incis, grid$RIPA, grid$UDA, post,
                                        (dd - damDen_m4_mean_raw) / damDen_m4_sd_raw)
    summarize_va(va_mat) %>% mutate(damDen = dd)
  })

  pdp <- bind_rows(pdp_list)

  # Repeat grid for each damDen value
  grid2 <- bind_rows(replicate(length(damDen_seq), grid, simplify = FALSE))

  bind_cols(grid2, pdp)
}



#####

pdp_incis_m4_0_damDen <- build_pdp_damDen(
  focal_raw = incis_m4_0_raw,
  focal_name = "incis",
  incis_m4_0_raw, RIPA_m4_0_raw, UDA_m4_0_raw,
  damDen_m4_0_post,
  incis_m4_0_mean_raw, incis_m4_0_sd_raw,
  RIPA_m4_0_mean_raw, RIPA_m4_0_sd_raw,
  UDA_m4_0_mean_raw, UDA_m4_0_sd_raw, damDen_m4_0_raw
)

pdp_RIPA_m4_0_damDen <- build_pdp_damDen(
  focal_raw = RIPA_m4_0_raw,
  focal_name = "RIPA",
  incis_m4_0_raw, RIPA_m4_0_raw, UDA_m4_0_raw,
  damDen_m4_0_post,
  incis_m4_0_mean_raw, incis_m4_0_sd_raw,
  RIPA_m4_0_mean_raw, RIPA_m4_0_sd_raw,
  UDA_m4_0_mean_raw, UDA_m4_0_sd_raw, damDen_m4_0_raw
)

pdp_UDA_m4_0_damDen <- build_pdp_damDen(
  focal_raw = UDA_m4_0_raw,
  focal_name = "UDA",
  incis_m4_0_raw, RIPA_m4_0_raw, UDA_m4_0_raw,
  damDen_m4_0_post,
  incis_m4_0_mean_raw, incis_m4_0_sd_raw,
  RIPA_m4_0_mean_raw, RIPA_m4_0_sd_raw,
  UDA_m4_0_mean_raw, UDA_m4_0_sd_raw, damDen_m4_0_raw
)



##########################################3

# Plot Function:

plot_pdp_damDen <- function(pdp, xvar, xtitle) {
  ggplot(pdp,
         aes(x = .data[[xvar]], y = VA_med,
             color = damDen, fill = damDen)) +
    geom_point() +
    scale_color_viridis_c(option = "plasma") +
    scale_fill_viridis_c(option = "plasma") +
    scale_y_continuous(labels = comma) +
    labs(
      title = paste("Partial Dependence of VA_0 on dam density and", xtitle),
      subtitle = "Across empirical dam density range",
      x = xtitle,
      y = "Predicted VA (m^2)",
      color = "Dam Density (dams/km)",
      fill  = "Dam Density (dams/km)"
    ) +
    theme_minimal(base_size = 14)
}



###########################################3


plot_pdp_damDen(pdp_incis_m4_0_damDen, "incis_raw", "Channel Incision (raw)")
plot_pdp_damDen(pdp_RIPA_m4_0_damDen,  "RIPA_raw",  "Riparian Area (m^2)")
plot_pdp_damDen(pdp_UDA_m4_0_damDen,   "UDA_raw",   "Contributing Drainage Area (km^2)")


```

----

**Change in VA across damDensity by stream and sampleEvent**


```{r,echo=FALSE}
m4_dam_seq_raw <- seq(min(aoi_HurdleData$dam_km),
                      max(aoi_HurdleData$dam_km),
                      length.out = 50)

m4_dam_seq_scaled <- (m4_dam_seq_raw - damDen_m4_mean_raw) / damDen_m4_sd_raw

m4_0_pdp_stream <- map_dfr(1:n_stream, function(j) {

  alpha_j <- damDen_m4_0_post$alpha

  beta_j <- damDen_m4_0_post$beta_damDen

  eta_j <- damDen_m4_0_post$eta_stream[, j] +
           rowMeans(damDen_m4_0_post$eta_year)

  map2_dfr(m4_dam_seq_raw, m4_dam_seq_scaled, function(d_raw, d_scaled) {

    mu_draws <- alpha_j + beta_j * d_scaled + eta_j
    va_draws <- exp(mu_draws) - 1

    tibble(
      stream = j,
      damDen = d_raw,
      VA_med = median(va_draws),
      VA_lo  = quantile(va_draws, 0.025),
      VA_hi  = quantile(va_draws, 0.975)
    )
  })
})

stream_lookup <- aoi_HurdleData %>%
  distinct(stream, watershed) %>%
  arrange(stream)

m4_0_pdp_stream <- m4_0_pdp_stream %>%
  left_join(stream_lookup, by = "stream")

ggplot(m4_0_pdp_stream, aes(x = watershed, y = VA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "Partial Dependence of VA_0 on Stream and Dam Density",
    x = "Stream",
    y = "Predicted VA_0 (m^2)",
    color = "Dam Density (dams/km)"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

###############################################################
################################################################
m4_0_pdp_year <- map_dfr(1:n_year, function(j) {

  alpha_j <- damDen_m4_0_post$alpha

  beta_j <- damDen_m4_0_post$beta_damDen

  eta_j <- damDen_m4_0_post$eta_year[,j] +
           rowMeans(damDen_m4_post$eta_stream)

  map2_dfr(m4_dam_seq_raw, m4_dam_seq_scaled, function(d_raw, d_scaled) {

    mu_draws <- alpha_j + beta_j * d_scaled + eta_j
    va_draws <- exp(mu_draws) - 1

    tibble(
      year = j,
      damDen = d_raw,
      VA_med = median(va_draws),
      VA_lo  = quantile(va_draws, 0.025),
      VA_hi  = quantile(va_draws, 0.975)
    )
  })
})

year_lookup <- tibble(
  year = 1:3,
  year_label = c(
    "2019",
    "2021",
    "2023"
  )
)

m4_0_pdp_year_named <- m4_0_pdp_year %>%
  left_join(year_lookup, by = "year")

ggplot(m4_0_pdp_year_named, aes(x = year_label, y = VA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "Partial Dependence of VA_0 on Year and Dam Density",
    x = "Year",
    y = "Predicted VA_0 (m²)",
    color = "Dam Density (dams/km)"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```



**CONTINUE FROM HERE 09/10/2026 1051; ADD 10, 20, 30-m buffer**



----








**Model Comparison**
```{r,echo=FALSE,warning=FALSE}

# Extract log-likelihood matrices
ll_m1 <- hglmm_m1_veg_damPres$sims.list$loglik
ll_m2 <- hglmm_m2_veg_damPres$sims.list$loglik
ll_m3 <- hglmm_m3_damDen_Veg$sims.list$loglik
ll_m4 <- hglmm_m4_damDen_Veg$sims.list$loglik

# Compute WAIC
waic_m1 <- waic(ll_m1)
waic_m2 <- waic(ll_m2)
waic_m3 <- waic(ll_m3)
waic_m4 <- waic(ll_m4)

# Combine into comparison table
waic_table <- tibble(
  model = c("m1", "m2", "m3", "m4"),
  WAIC  = c(waic_m1$estimates["waic", "Estimate"],
            waic_m2$estimates["waic", "Estimate"],
            waic_m3$estimates["waic", "Estimate"],
            waic_m4$estimates["waic", "Estimate"])
) %>%
  arrange(WAIC)

waic_table

```

    According to WAIC comparison, all 4 models perform comparably. M4 provides the
    most complete story, as it covers dam presence/absence via the damDensity gradient
    as well as geomorphic covariates.
    
    Recommendation: use M4 (VA ~ damDensity + geomorphicCovariates) for manuscript
    results
    
    