---
title: "Ch1_Hurdle_SWA~Beaver"
author: "Matt Webster"
date: "2026-07-14"
output: html_document
editor_options: 
  chunk_output_type: console
---




The purpose of this script is to explore the effect of beaver dams on surface water area, relative to local geomorphology, and contributing drainage area. This modeling focuses solely on late summer months; August & September. 

Two models are presented herein. Both models have the same random effects structure for stream, year, and sample event. Model 1 examines the effect of beaver dam presence on surface water area, with no consideration for local context. Model 2 includes additional fixed effects of channel incision, riparian area, and drainage area.

Both models are constructed as **hurdle** models. 1) Estimate the relative influence of dam presence on surface water presence. 2) For cases where water is present, estimate the relative influence of dam presence on surface water area. 

----
----

```{r,echo=FALSE}
  knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

```{r,echo=FALSE}
library(tidyverse)
library(ggplot2)
library(dataRetrieval)
library(knitr)
library(ggplot2)
library(zoo) #for moving window
library(stringr)
library(vioplot)
library(purrr)
library(lavaan)
library(blavaan)
library(ggcorrplot)
library(lme4)
library(broom.mixed)
library(jagsUI)
```
      
----

Data from USGS LBV Guston Coulee gage, near Saco, MT:
----

```{r,echo=FALSE, warning=FALSE}
# LOWER BEAVER CK
# Gage near Guston Coulee
usgsGage_06166000 <- read_waterdata_daily(
  monitoring_location_id = "USGS-06166000",
  parameter_code = "00060",
  time = c("2019-06-01","2023-10-01")
)

###########################################

# Clean USGS
usgs_df <- usgsGage_06166000 %>%
  mutate(
    source = "USGS",
    date   = as.Date(time),
    flow   = value
    ) %>%
  filter(
    (date >= as.Date("2019-06-01") & date <= as.Date("2019-10-01")) |
    (date >= as.Date("2021-06-01") & date <= as.Date("2021-10-01")) |
    (date >= as.Date("2023-06-01") & date <= as.Date("2023-10-01")) ) %>% 
  mutate(
    zFlow = scale(flow)
  ) %>% 
  select(source,
         date,
         zFlow) %>% 
  mutate(surveyPeriod = case_when(between(date, as.Date("2019-06-01"), as.Date("2019-06-14")) ~ 191,
                                between(date, as.Date("2019-06-15"), as.Date("2019-06-30")) ~ 192,
                                between(date, as.Date("2019-07-01"), as.Date("2019-07-14")) ~ 193,
                                between(date, as.Date("2019-07-15"), as.Date("2019-07-31")) ~ 194,
                                between(date, as.Date("2019-08-01"), as.Date("2019-08-14")) ~ 195,
                                between(date, as.Date("2019-08-15"), as.Date("2019-08-31")) ~ 196,
                                between(date, as.Date("2019-09-01"), as.Date("2019-09-14")) ~ 197,
                                between(date, as.Date("2019-09-15"), as.Date("2019-09-30")) ~ 198,
                                between(date, as.Date("2021-06-01"), as.Date("2021-06-14")) ~ 211,
                                between(date, as.Date("2021-06-15"), as.Date("2021-06-30")) ~ 212,
                                between(date, as.Date("2021-07-01"), as.Date("2021-07-14")) ~ 213,
                                between(date, as.Date("2021-07-15"), as.Date("2021-07-31")) ~ 214,
                                between(date, as.Date("2021-08-01"), as.Date("2021-08-14")) ~ 215,
                                between(date, as.Date("2021-08-15"), as.Date("2021-08-31")) ~ 216,
                                between(date, as.Date("2021-09-01"), as.Date("2021-09-14")) ~ 217,
                                between(date, as.Date("2021-09-15"), as.Date("2021-09-30")) ~ 218,
                                between(date, as.Date("2023-06-01"), as.Date("2023-06-14")) ~ 231,
                                between(date, as.Date("2023-06-15"), as.Date("2023-06-30")) ~ 232,
                                between(date, as.Date("2023-07-01"), as.Date("2023-07-14")) ~ 233,
                                between(date, as.Date("2023-07-15"), as.Date("2023-07-31")) ~ 234,
                                between(date, as.Date("2023-08-01"), as.Date("2023-08-14")) ~ 235,
                                between(date, as.Date("2023-08-15"), as.Date("2023-08-31")) ~ 236,
                                between(date, as.Date("2023-09-01"), as.Date("2023-09-14")) ~ 237,
                                between(date, as.Date("2023-09-15"), as.Date("2023-09-30")) ~ 238,
                                TRUE ~ NA_real_)) %>% 
  filter(surveyPeriod > 0)

usgs_df$surveyPeriod <- as.factor(usgs_df$surveyPeriod)


mean_flow_per_survey_period <- usgs_df %>% 
  group_by(surveyPeriod) %>%
  summarise(zFlow_mean = mean(zFlow),
            .groups = "drop")



lateSummerFlow <- usgs_df %>% 
  filter(surveyPeriod != 191,
         surveyPeriod != 192,
         surveyPeriod != 193,
         surveyPeriod != 194,
         surveyPeriod != 211,
         surveyPeriod != 212,
         surveyPeriod != 213,
         surveyPeriod != 214,
         surveyPeriod != 231,
         surveyPeriod != 232,
         surveyPeriod != 233,
         surveyPeriod != 234,
         )
```

Late Summer Flow Conditions
----

2019:
----

Moderate August with peak in September 

```{r,echo=FALSE}

flow19 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2019-06-01") & date <= as.Date("2019-10-01"))
  )


####

ggplot(flow19, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '19 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```

----
  
2021:
----

Low flow conditions throughout the summer season. Part of 3 year sustained drought conditions (2020, 2021, 2022).


Gage recorded no flow during August & September



```{r,echo=FALSE}

flow21 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2021-06-01") & date <= as.Date("2021-10-01"))
  )


####

ggplot(flow21, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '21 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```  
  
----

2023:
----

Moderate to low August & September flows; after early season extreme flooding. 



```{r,echo=FALSE}
flow23 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2023-06-01") & date <= as.Date("2023-10-01"))
  )


####

ggplot(flow23, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '23 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```  



GEE-derived data for each stream:
----

```{r, echo=FALSE}

#Read in data frames for each HUC12/Year combo:

# WILLOW:
#2019:
wc_19 <- read.csv("WC_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'WC')
#2021:
wc_21 <- read.csv("WC_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'WC')
#2023:
wc_23 <- read.csv("WC_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'WC')
#AllYears:
WC <- bind_rows(wc_19,wc_21,wc_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LONE TREE:
#2019:
lt_19 <- read.csv("LT_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'LT')
#2021:
lt_21 <- read.csv("LT_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'LT')
#2023:
lt_23 <- read.csv("LT_2023_DF_Final_Long_0427Update_250Spacing.csv")%>% 
  mutate(year = 2023,
         watershed = 'LT')

#AllYears:
LT <- bind_rows(lt_19,lt_21,lt_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))

# BRAZIL:
#2019:
brz_19 <- read.csv("BRZ_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'BRZ')
#2021:
brz_21 <- read.csv("BRZ_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'BRZ')
#2023:
brz_23 <- read.csv("BRZ_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'BRZ')

#AllYears:
BRZ <- bind_rows(brz_19,brz_21,brz_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LARB:
#2019:
lrb_19 <- read.csv("LRB_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'LRB')
#2021:
lrb_21 <- read.csv("LRB_2021_DF_Final_Long_0427Update_250Spacing.csv")%>% 
  mutate(year = 2021,
         watershed = 'LRB')
#2023:
lrb_23 <- read.csv("LRB_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'LRB')

#AllYears:
LRB <- bind_rows(lrb_19,lrb_21,lrb_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LOWER BEAVER:
#2019:
lbv_19 <- read.csv("LBV_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'LBV')
#2021:
lbv_21 <- read.csv("LBV_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'LBV')
#2023:
lbv_23 <- read.csv("LBV_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'LBV')

#AllYears:
LBV <- bind_rows(lbv_19,lbv_21,lbv_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# UPPER BEAVER:
#2019:
ubv_19 <- read.csv("UBV_2019_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2019,
         watershed = 'UBV')
#2021:
ubv_21 <- read.csv("UBV_2021_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2021,
         watershed = 'UBV')
#2023:
ubv_23 <- read.csv("UBV_2023_DF_Final_Long_0427Update_250Spacing.csv") %>% 
  mutate(year = 2023,
         watershed = 'UBV')

#AllYears:
UBV <- bind_rows(ubv_19,ubv_21,ubv_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
```


----

Currently each incoming data frame is of the following structure:

```{r,include=FALSE}
str(brz_19)
```

Where:

    - clusterID_seq = dammed/undammed sample pair ID
    - window_type = whether a sample window is 'dammed' or 'undammed'
    - watershed = watershed identifier (e.g., "BRZ" = Brazil Ck)
    - year = year
    - window_length = length (m) of stream flow line within sample window
    - catchment_km2 = contributing catchment area/upstream drainage area (km^2)
    - riparian_area_m2 = riparian area (m^2) of sample window; 2-m vertical
      buffer from DEM low point (stream flow line)
    - TO = mean topographic openness across window
    - p_window: avg dam detection prob for target window; see CP Ch1
    - f_window: avg dam false positive prob for target window; see CP Ch1
    - beav_count_all: all unique detected dams
    - beav_count_conf: unique confidently detected dams
    - anthro_count: all unique detected anthropogenic dams
    - isAnthro: National Wetland Inventory (NWI) "dyked/impounded"
    - Nhat = estimated dam count within sample window
      - ...= beav_count_all/(p_window(1-f_window)  [McCallum 2005]
    - TO_mean = mean topographic openness across window
    - time_period = 2wk interval 06/01 - 09/30
    - SWA = surface water area (m^2) for window during time (t)
    - t = time (t) {same as time_period}

----
----
----
----


**Relative flow data.**

NOTE: currently this represents data from the USGS Guston Coulee gage that has been z-standardized over the current data acquisition range (June-Sept; 2019, 2021, 2023). 

Then, for each 2-week survey time period (1:8), the average relative flow $[mean(zFlow)]$ was computed, and added as a data column. IT IS ASSUMED that this flow data is representative of the RELATIVE flow pattern experienced throughout the study area during this time.

----

Add relFlow data:

```{r,include=FALSE, fig.width=10}
WC_RelFlow <- WC %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)
```

```{r,include=FALSE, fig.width=10}
LT_RelFlow <- LT %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)

```

```{r,include=FALSE, fig.width=10}
BRZ_RelFlow <- BRZ %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)
```

```{r,include=FALSE, fig.width=10}
LRB_RelFlow <- LRB %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)

```

```{r,include=FALSE, fig.width=10}
LBV_RelFlow <- LBV %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)
```

```{r,include=FALSE, fig.width=10}
UBV_RelFlow <- UBV %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)
```





Hierarchical Generalized Linear Mixed Effects Models
----

Proposed Models:
----


**Both models:** 

$$  SWA \sim N(\mu, \sigma);$$

----
----

**M1:**


$$\mu = \alpha + \eta_{stream} + \eta_{event} + \eta_{cluster} + (\beta1+\beta_{stream} + \beta_{year} + \beta_{event})*damPresence $$

----
----

**M2:**

$$\mu = \alpha + \eta_{stream} + \eta_{event} + \eta_{cluster} + (\beta1+\beta_{stream} + \beta_{year} + \beta_{event})*damPresence + \beta_2 * Incision + \beta_3 * riparianArea + \beta_4 * drainageArea$$

----
----

```{r,include=FALSE}
# Create GLMM data frame:

aoi_GLM_Data <- WC_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_km) %>% 
  bind_rows(LT_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_km)) %>% 
  bind_rows(BRZ_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_km)) %>% 
  bind_rows(LRB_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_km)) %>% 
  bind_rows(LBV_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_km)) %>% 
  bind_rows(UBV_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_km))


aoi_GLM_Data$clusterID_unique <- interaction(
  aoi_GLM_Data$year,
  aoi_GLM_Data$watershed,
  aoi_GLM_Data$clusterID_seq,
  drop = TRUE
)

aoi_GLM_Data$clusterID_num <- as.numeric(aoi_GLM_Data$clusterID_unique)

#aoi_GLM_Data$eventID <- interaction(
#  aoi_GLM_Data$year,
#  aoi_GLM_Data$surveyPeriod,
#  drop = TRUE
#)

#aoi_GLM_Data$eventID_num <- as.numeric(aoi_GLM_Data$eventID)

aoi_GLM_Data$stream <- as.numeric(as.factor(aoi_GLM_Data$watershed))

aoi_GLM_Data$sampleEvent <- as.numeric(aoi_GLM_Data$surveyPeriod)

aoi_GLM_Data$year_id <- as.numeric(as.factor(aoi_GLM_Data$year))

aoi_GLM_Data <- aoi_GLM_Data %>% 
  mutate(damPresence = case_when(Nhat > 0 ~ 1,
                                 Nhat == 0 ~ 0))

str(aoi_GLM_Data)



aoi_GLM_Data_mod <- aoi_GLM_Data %>%
  filter(!is.na(SWA))   # or !is.na(RI) if that’s the response


# Build explicit lookups from the data actually used in the model
stream_lookup <- aoi_GLM_Data_mod %>%
  distinct(stream, watershed) %>%
  arrange(stream)

year_lookup <- aoi_GLM_Data_mod %>%
  distinct(year_id, year) %>%
  arrange(year_id)

uda_lookup <- aoi_GLM_Data_mod %>%
  group_by(watershed) %>%
  summarise(max_catchment = max(catchment_km2, na.rm = TRUE)) %>%
  arrange(max_catchment)



# Start from the full modeled data and remove early summer time periods
# This analysis will look exclusively at late season mesic resource availability
# 08/01-09/30
# t5:t8
aoi_GLM_Data_mod_3_4 <- aoi_GLM_Data_mod %>% 
  filter(!(sampleEvent %in% 1:4) &      # drop 2019 t1-t4
           !(sampleEvent %in% 9:12) &   # drop 2021 t1-t4
           !(sampleEvent %in% 17:20))   # drop 2023 t1–t4

event_lookup <- data.frame(
  old = sort(unique(aoi_GLM_Data_mod_3_4$sampleEvent)),
  new = 1:length(unique(aoi_GLM_Data_mod_3_4$sampleEvent))
)

aoi_GLM_Data_mod_3_4 <- aoi_GLM_Data_mod_3_4 %>%
  left_join(event_lookup, by = c("sampleEvent" = "old")) %>%
  mutate(sampleEvent_new = new) %>%
  select(-new)

event_year_3_4 <- tapply(aoi_GLM_Data_mod_3_4$year_id,
                         aoi_GLM_Data_mod_3_4$sampleEvent_new,
                         unique)
event_year_3_4 <- as.numeric(event_year_3_4)



aoi_GLM_Data_mod_3_4$SWA <- ifelse(aoi_GLM_Data_mod_3_4$SWA < 1,
                                   0,
                                   aoi_GLM_Data_mod_3_4$SWA)
```

Log-transform SWA to 'normalize' the right skew:

```{r,echo=FALSE}
aoi_GLM_Data_mod_3_4$logSWA <- log(aoi_GLM_Data_mod_3_4$SWA+1)

```

Format data for hurdle:

```{r,echo=TRUE}
# Create binary indicator of surface water presence, for logistic part of hurdle model:
aoi_HurdleData <- aoi_GLM_Data_mod_3_4
aoi_HurdleData$y <- ifelse(aoi_GLM_Data_mod_3_4$SWA>0,1,0)
aoi_HurdleData$dam_km_std <- scale(aoi_HurdleData$dam_km)

# Identify positive-SWA subset for the continuous/magnitude submodel:
#pos_idx <- which(aoi_HurdleData$SWA > 0)

#N_pos <- length(pos_idx) 
#y_pos <- aoi_HurdleData$logSWA[pos_idx] #response subset for magnitude sub-model

#stream_pos <- aoi_HurdleData$stream[pos_idx]
#year_pos <- aoi_HurdleData$year_id[pos_idx]
#event_pos <- aoi_HurdleData$sampleEvent_new[pos_idx]
#cluster_pos <- aoi_HurdleData$clusterID_num[pos_idx]
#damPres_pos <- aoi_HurdleData$damPresence[pos_idx]
#damDen_pos <- aoi_HurdleData$dam_km[pos_idx] #only for densit model

# Full data indices for Bernoulli sub-model

N <- nrow(aoi_HurdleData)
y <- aoi_HurdleData$y
y2 <- aoi_HurdleData$SWA
stream <- aoi_HurdleData$stream
year <- aoi_HurdleData$year_id
event <- aoi_HurdleData$sampleEvent_new
cluster <- aoi_HurdleData$clusterID_num
damPres <- aoi_HurdleData$damPresence
damDen <- aoi_HurdleData$dam_km

n_stream <- length(unique(aoi_HurdleData$stream))
n_year <- length(unique(aoi_HurdleData$year_id))
n_event <- length(unique(aoi_HurdleData$sampleEvent_new))
n_cluster <- length(unique(aoi_HurdleData$clusterID_num))
```


**Visualize SWA spread:**

**Raw SWA (m^2):**

```{r,echo=FALSE}
hist(aoi_HurdleData$SWA,
     main = "Distribution of SWA",
     xlab = "log(SWA)")
```

**log(SWA):**

```{r,echo=FALSE}
hist(aoi_HurdleData$logSWA,
     main = "Distribution of log(SWA)",
     xlab = "log(SWA)")
```

----
----

----
----

----
----

    This section explores late summner SWA in relation to beaver dam presence.
    
    Uses hurdle model with nested random slopes for beaver dam effect to allow for stream, year, 
    and sampleEvent to cause deviation from the global mean beaver effect.
    
    Pt1: Surface water presence/absence; logistic regression
    Pt2: Surface water amount; log(Normal) regression
    
    Uses log(SWA) for AOI all years. Adjust priors accordingly...

----
----

**Hurdle HGLMM**
----

```{r,echo=FALSE}

sink("HGLMM_hurdle_damPresence.jags")

cat("
    model{
    
    # Surface water presence/absence sub-model
    
    for(i in 1:N){
    
    z[i] ~ dbern(psi[i])
    
    logit(psi[i]) <- epsilon
                     + (zeta_damPres
                        + zeta_stream[stream[i]]
                        + zeta_year[year[i]]
                        + zeta_event[event[i]]) * damPres[i]
                     + iota_stream[stream[i]]
                     + iota_event[event[i]]
                     + iota_cluster[cluster[i]]
                     
    ######################################
    # Surface water detection sub-model:
    
    p_eff[i] <- p_detect[event[i]] * z[i]
    
    y[i] ~ dbern(p_eff[i])    # presence/absence data (0,1)
    
    #######################################
    # Surface water amount sub-model:
    
    y2[i] ~ dnorm(mu[i], tau_eff[i])
      
    mu[i] <- z[i] *
              (alpha
                + (beta_damPres
                  + beta_stream[stream[i]]
                  + beta_year[year[i]]
                  + beta_event[event[i]]) * damPres[i]
                + eta_stream[stream[i]]
                + eta_event[event[i]]
                + eta_cluster[cluster[i]])
    
    tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100000
    }
    
    ##########################################################################
    # PRIORS FOR SW PRESENCE:
    # Fixed effects priors for SW presence/absence sub-model:
    
    epsilon ~ dnorm(0,0.04)      # intercept; weakly informative
    zeta_damPres ~ dnorm(0,0.04) # weak global dam effect on SW presence
    
    # Random effects priors for SW presence/absence sub-model:
    
    for(j in 1:n_stream){
      zeta_stream[j] ~ dnorm(0, tau_zeta_stream)
      iota_stream[j] ~ dnorm(0, tau_iota_stream)
    }
    
    for(yr in 1:n_year){
      zeta_year[yr] ~ dnorm(0, tau_zeta_year)
    }
    
    for(k in 1:n_event){
      zeta_event[k] ~ dnorm(0, tau_zeta_event)
      iota_event[k] ~ dnorm(0, tau_iota_event)
    }
    
    for(c in 1:n_cluster){
      iota_cluster[c] ~ dnorm(0, tau_iota_cluster)
    }
    
    # Hyperpriors for SW presence/absence submodel (1/2 Normal on SDs):
    
    sigma_zeta_stream ~ dnorm(0,1) T(0,)
    tau_zeta_stream  <- pow(sigma_zeta_stream, -2)  #precision = 1/sd^2
    
    sigma_zeta_year ~ dnorm(0,1) T(0,)
    tau_zeta_year <- pow(sigma_zeta_year, -2)
    
    sigma_zeta_event ~ dnorm(0,1) T(0,)
    tau_zeta_event <- pow(sigma_zeta_event, -2)
    
    sigma_iota_stream ~ dnorm(0,1) T(0,)
    tau_iota_stream <- pow(sigma_iota_stream, -2)
    
    sigma_iota_event ~ dnorm(0,1) T(0,)
    tau_iota_event <- pow(sigma_iota_event, -2)
    
    sigma_iota_cluster ~ dnorm(0,1) T(0,)
    tau_iota_cluster <- pow(sigma_iota_cluster, -2)
    
    ######################################################################
    
    # PRIORS FOR SW DETECTION:
    
    for(k in 1:n_event){
      p_detect[k] ~ dnorm(0.76, pow(0.16,-2)) T(0,1)  #truncated normal; empirically centered
    }
    
    ######################################################################
    
    # PRIORS FOR SW AMOUNT:
    # Fixed effects priors for SW amount sub-model; log(SWA+1) scale:
    
    alpha ~ dnorm(0,0.04)        # weakly info. intercept prior
    beta_damPres ~ dnorm(0,0.25) # moderately info. global dam effect prior; more info than alpha
    
    # Random effects priors for SW amount sub-model; log(SWA+1) scale:
    
    for(j in 1:n_stream){
      beta_stream[j] ~ dnorm(0, tau_beta_stream)
      eta_stream[j] ~ dnorm(0, tau_eta_stream)
    }
    
    for(yr in 1:n_year){
      beta_year[yr] ~ dnorm(0, tau_beta_year)
    }
    
    for(k in 1:n_event){
      beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
      eta_event[k] ~ dnorm(0, tau_eta_event)
    }
  
    for(c in 1:n_cluster){
      eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
    }
    
    # Hyperpriors for SW amount sub-model; still 1/2 Normal distributions
    
    sigma ~ dnorm(0,1) T(0,)
    tau <- pow(sigma, -2)
    
    sigma_beta_stream ~ dnorm(0, 0.7) T(0,)
    tau_beta_stream <- pow(sigma_beta_stream, -2)
    
    sigma_beta_year ~ dnorm(0, 0.7) T(0,)
    tau_beta_year <- pow(sigma_beta_year, -2)
    
    sigma_beta_event ~ dnorm(0, 0.7) T(0,)
    tau_beta_event <- pow(sigma_beta_event, -2)
    
    sigma_eta_stream ~ dnorm(0, 1) T(0,)
    tau_eta_stream <- pow(sigma_eta_stream, -2)
    
    sigma_eta_event ~ dnorm(0, 1) T(0,)
    tau_eta_event <- pow(sigma_eta_event, -2)
    
    sigma_eta_cluster ~ dnorm(0, 1) T(0,)
    tau_eta_cluster <- pow(sigma_eta_cluster, -2)
    
    }
    
    ", fill = TRUE)

sink()
```

```{r,echo=FALSE}
hurdle_damPres_data_list <- list(
  N = N,
  y = y,
  y2 = y2,
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damPres = damPres,
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hurdle_damPres_inits <- function(){
  list(
    alpha        = rnorm(1, 0, 1),
    epsilon      = rnorm(1, 0, 1),

    sigma        = runif(1, 0.1, 2),
    sigma_zeta_stream  = runif(1, 0.1, 2),
    sigma_zeta_year    = runif(1, 0.1, 2),
    sigma_zeta_event   = runif(1, 0.1, 2),
    sigma_iota_stream  = runif(1, 0.1, 2),
    sigma_iota_event   = runif(1, 0.1, 2),
    sigma_iota_cluster = runif(1, 0.1, 2),

    sigma_beta_stream  = runif(1, 0.1, 2),
    sigma_beta_year    = runif(1, 0.1, 2),
    sigma_beta_event   = runif(1, 0.1, 2),
    sigma_eta_stream   = runif(1, 0.1, 2),
    sigma_eta_event    = runif(1, 0.1, 2),
    sigma_eta_cluster  = runif(1, 0.1, 2)
  )
}

hurdle_damPres_params <- c(
  "alpha",
  "zeta_damPres",
  "zeta_stream",
  "zeta_year",
  "zeta_event",
  "iota_stream",
  "iota_event",
  "iota_cluster",
  "beta_damPres",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_zeta_stream",
  "sigma_zeta_year",
  "sigma_zeta_event",
  "sigma_iota_stream",
  "sigma_iota_event",
  "sigma_iota_cluster",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_event",
  "sigma_eta_cluster",
  "p_detect"
)

nchains <- 3
nthin <- 10
niter <- 20000
nburn <- 5000

hurdle_damPres_SW <- jags(data = hurdle_damPres_data_list,
                   inits = hurdle_damPres_inits,
                   parameters.to.save = hurdle_damPres_params,
                   model.file = "HGLMM_hurdle_damPresence.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hurdle_damPres_SW) # model with local and watershed dam density as fixed effects
print(hurdle_damPres_SW$summary[grep("beta_", rownames(hurdle_damPres_SW$summary)), ])
print(hurdle_damPres_SW$summary[grep("zeta_", rownames(hurdle_damPres_SW$summary)), ])

```





----
----

Dam Density
----

**COPY PASTE AND EDIT DAM PRESENCE MODEL FOR USE AS DAM DENSITY MODEL**