---
title: "Hurdle_Data_Prep"
author: "Matt Webster"
date: "2026-07-28"
output: html_document
editor_options: 
  chunk_output_type: console
---

The purpose of this RMarkdown document is to set up the dataframe for use in Ch1 models. Data herein is prepared for the folowing models:


    Surface Water Hurdle Models:
      
      SW ~ damPresence
      SW ~ damPresence + geomorphic covar.
      SW ~ damDensity
      SW ~ damDensity + geomorphic covar.
      
    Veg Area HGLMMs:
    
      VegArea ~ damPresence
      VegArea ~ damPresence + geomorphic covar.
      VegArea ~ damDensity
      VegArea ~ damDensity + geomorphic covar.
      
    NDVI HGLMMs:
    
      NDVI ~ damPresence
      NDVI ~ damPresence + geomorphic covar.
      NDVI ~ damDensity
      NDVI ~ damDensity + geomorphic covar.
      
      
----

----

----

----



```{r,echo=FALSE}
  knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

```{r,echo=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dataRetrieval)
library(knitr)
library(ggplot2)
library(zoo) #for moving window
library(stringr)
library(vioplot)
library(purrr)
library(lavaan)
library(blavaan)
library(ggcorrplot)
library(lme4)
library(broom.mixed)
library(coda)
library(scales)
```
      



USGS Flow Data
----

**Data from USGS LBV Guston Coulee gage, near Saco, MT:**

```{r,echo=FALSE, warning=FALSE}
# LOWER BEAVER CK
# Gage near Guston Coulee
usgsGage_06166000 <- read_waterdata_daily(
  monitoring_location_id = "USGS-06166000",
  parameter_code = "00060",
  time = c("2019-06-01","2023-10-01")
)

###########################################

# Clean USGS
usgs_df <- usgsGage_06166000 %>%
  mutate(
    source = "USGS",
    date   = as.Date(time),
    flow   = value
    ) %>%
  filter(
    (date >= as.Date("2019-06-01") & date <= as.Date("2019-10-01")) |
    (date >= as.Date("2021-06-01") & date <= as.Date("2021-10-01")) |
    (date >= as.Date("2023-06-01") & date <= as.Date("2023-10-01")) ) %>% 
  mutate(
    zFlow = scale(flow)
  ) %>% 
  select(source,
         date,
         zFlow) %>% 
  mutate(surveyPeriod = case_when(between(date, as.Date("2019-06-01"), as.Date("2019-06-14")) ~ 191,
                                between(date, as.Date("2019-06-15"), as.Date("2019-06-30")) ~ 192,
                                between(date, as.Date("2019-07-01"), as.Date("2019-07-14")) ~ 193,
                                between(date, as.Date("2019-07-15"), as.Date("2019-07-31")) ~ 194,
                                between(date, as.Date("2019-08-01"), as.Date("2019-08-14")) ~ 195,
                                between(date, as.Date("2019-08-15"), as.Date("2019-08-31")) ~ 196,
                                between(date, as.Date("2019-09-01"), as.Date("2019-09-14")) ~ 197,
                                between(date, as.Date("2019-09-15"), as.Date("2019-09-30")) ~ 198,
                                between(date, as.Date("2021-06-01"), as.Date("2021-06-14")) ~ 211,
                                between(date, as.Date("2021-06-15"), as.Date("2021-06-30")) ~ 212,
                                between(date, as.Date("2021-07-01"), as.Date("2021-07-14")) ~ 213,
                                between(date, as.Date("2021-07-15"), as.Date("2021-07-31")) ~ 214,
                                between(date, as.Date("2021-08-01"), as.Date("2021-08-14")) ~ 215,
                                between(date, as.Date("2021-08-15"), as.Date("2021-08-31")) ~ 216,
                                between(date, as.Date("2021-09-01"), as.Date("2021-09-14")) ~ 217,
                                between(date, as.Date("2021-09-15"), as.Date("2021-09-30")) ~ 218,
                                between(date, as.Date("2023-06-01"), as.Date("2023-06-14")) ~ 231,
                                between(date, as.Date("2023-06-15"), as.Date("2023-06-30")) ~ 232,
                                between(date, as.Date("2023-07-01"), as.Date("2023-07-14")) ~ 233,
                                between(date, as.Date("2023-07-15"), as.Date("2023-07-31")) ~ 234,
                                between(date, as.Date("2023-08-01"), as.Date("2023-08-14")) ~ 235,
                                between(date, as.Date("2023-08-15"), as.Date("2023-08-31")) ~ 236,
                                between(date, as.Date("2023-09-01"), as.Date("2023-09-14")) ~ 237,
                                between(date, as.Date("2023-09-15"), as.Date("2023-09-30")) ~ 238,
                                TRUE ~ NA_real_)) %>% 
  filter(surveyPeriod > 0)

usgs_df$surveyPeriod <- as.factor(usgs_df$surveyPeriod)


mean_flow_per_survey_period <- usgs_df %>% 
  group_by(surveyPeriod) %>%
  summarise(zFlow_mean = mean(zFlow),
            .groups = "drop")



lateSummerFlow <- usgs_df %>% 
  filter(surveyPeriod != 191,
         surveyPeriod != 192,
         surveyPeriod != 193,
         surveyPeriod != 194,
         surveyPeriod != 211,
         surveyPeriod != 212,
         surveyPeriod != 213,
         surveyPeriod != 214,
         surveyPeriod != 231,
         surveyPeriod != 232,
         surveyPeriod != 233,
         surveyPeriod != 234,
         )
```

**Late Summer Flow Conditions**


**2019:**


Moderate August with small peak in September


```{r,echo=FALSE}

flow19 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2019-06-01") & date <= as.Date("2019-10-01"))
  )


####

ggplot(flow19, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '19 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```


----


**2021:**


Low flow conditions throughout the summer season. Part of 3 year sustained drought conditions (2020, 2021, 2022).


Gage recorded no flow during August & September



```{r,echo=FALSE}

flow21 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2021-06-01") & date <= as.Date("2021-10-01"))
  )


####

ggplot(flow21, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '21 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```  
  
  
----


**2023:**


Moderate to low August & September flows; after early season extreme flooding. 



```{r,echo=FALSE}
flow23 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2023-06-01") & date <= as.Date("2023-10-01"))
  )


####

ggplot(flow23, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '23 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```  

----

GEE-derived data for each stream:
----

```{r, echo=FALSE}

#Read in data frames for each HUC12/Year combo:

# WILLOW:
#2019:
wc_19 <- read.csv("WC_2019_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2019,
         watershed = 'WC')
#2021:
wc_21 <- read.csv("WC_2021_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2021,
         watershed = 'WC')
#2023:
wc_23 <- read.csv("WC_2023_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2023,
         watershed = 'WC')
#AllYears:
WC <- bind_rows(wc_19,wc_21,wc_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LONE TREE:
#2019:
lt_19 <- read.csv("LT_2019_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2019,
         watershed = 'LT')
#2021:
lt_21 <- read.csv("LT_2021_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2021,
         watershed = 'LT')
#2023:
lt_23 <- read.csv("LT_2023_DF_Final_Long_0630Update.csv")%>% 
  mutate(year = 2023,
         watershed = 'LT')

#AllYears:
LT <- bind_rows(lt_19,lt_21,lt_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))

# BRAZIL:
#2019:
brz_19 <- read.csv("BRZ_2019_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2019,
         watershed = 'BRZ')
#2021:
brz_21 <- read.csv("BRZ_2021_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2021,
         watershed = 'BRZ')
#2023:
brz_23 <- read.csv("BRZ_2023_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2023,
         watershed = 'BRZ')

#AllYears:
BRZ <- bind_rows(brz_19,brz_21,brz_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LARB:
#2019:
lrb_19 <- read.csv("LRB_2019_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2019,
         watershed = 'LRB')
#2021:
lrb_21 <- read.csv("LRB_2021_DF_Final_Long_0630Update.csv")%>% 
  mutate(year = 2021,
         watershed = 'LRB')
#2023:
lrb_23 <- read.csv("LRB_2023_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2023,
         watershed = 'LRB')

#AllYears:
LRB <- bind_rows(lrb_19,lrb_21,lrb_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LOWER BEAVER:
#2019:
lbv_19 <- read.csv("LBV_2019_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2019,
         watershed = 'LBV')
#2021:
lbv_21 <- read.csv("LBV_2021_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2021,
         watershed = 'LBV')
#2023:
lbv_23 <- read.csv("LBV_2023_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2023,
         watershed = 'LBV')

#AllYears:
LBV <- bind_rows(lbv_19,lbv_21,lbv_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# UPPER BEAVER:
#2019:
ubv_19 <- read.csv("UBV_2019_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2019,
         watershed = 'UBV')
#2021:
ubv_21 <- read.csv("UBV_2021_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2021,
         watershed = 'UBV')
#2023:
ubv_23 <- read.csv("UBV_2023_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2023,
         watershed = 'UBV')

#AllYears:
UBV <- bind_rows(ubv_19,ubv_21,ubv_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
```


----

Currently each incoming data frame is of the following structure:

```{r,include=FALSE}
str(brz_19)
```

Where:

    - clusterID_seq = dammed/undammed sample pair ID
    - window_type = whether a sample window is 'dammed' or 'undammed'
    - watershed = watershed identifier (e.g., "BRZ" = Brazil Ck)
    - year = year
    - window_length = length (m) of stream flow line within sample window
    - catchment_km2 = contributing catchment area/upstream drainage area (km^2)
    - riparian_area_m2 = riparian area (m^2) of sample window; 2-m vertical
      buffer from DEM low point (stream flow line)
    - TO = mean topographic openness across window
    - p_window: avg dam detection prob for target window; see CP Ch1
    - f_window: avg dam false positive prob for target window; see CP Ch1
    - beav_count_all: all unique detected dams
    - beav_count_conf: unique confidently detected dams
    - anthro_count: all unique detected anthropogenic dams
    - isAnthro: National Wetland Inventory (NWI) "dyked/impounded"
    - Nhat = estimated dam count within sample window
      - ...= beav_count_all/(p_window(1-f_window)  [McCallum 2005]
    - TO_mean = mean topographic openness across window
    - time_period = 2wk interval 06/01 - 09/30
    - SWA = surface water area (m^2) for window during time (t)
    - t = time (t) {same as time_period}



----

----

----

----



```{r,include=FALSE, fig.width=10}
WC_RelFlow <- WC %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)
```

```{r,include=FALSE, fig.width=10}
LT_RelFlow <- LT %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)

```

```{r,include=FALSE, fig.width=10}
BRZ_RelFlow <- BRZ %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)
```

```{r,include=FALSE, fig.width=10}
LRB_RelFlow <- LRB %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)

```

```{r,include=FALSE, fig.width=10}
LBV_RelFlow <- LBV %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)
```

```{r,include=FALSE, fig.width=10}
UBV_RelFlow <- UBV %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)
```


----

----

----

----


```{r,include=FALSE}
# Create GLMM data frame:

aoi_GLM_Data <- WC_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         VA,
         NDVI,
         riparian_area_m2,
         dam_km) %>% 
  bind_rows(LT_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         VA,
         NDVI,
         riparian_area_m2,
         dam_km)) %>% 
  bind_rows(BRZ_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         VA,
         NDVI,
         riparian_area_m2,
         dam_km)) %>% 
  bind_rows(LRB_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         VA,
         NDVI,
         riparian_area_m2,
         dam_km)) %>% 
  bind_rows(LBV_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         VA,
         NDVI,
         riparian_area_m2,
         dam_km)) %>% 
  bind_rows(UBV_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         VA,
         NDVI,
         riparian_area_m2,
         dam_km))


aoi_GLM_Data$clusterID_unique <- interaction(
  aoi_GLM_Data$year,
  aoi_GLM_Data$watershed,
  aoi_GLM_Data$clusterID_seq,
  drop = TRUE
)

aoi_GLM_Data$clusterID_num <- as.numeric(aoi_GLM_Data$clusterID_unique)

#aoi_GLM_Data$eventID <- interaction(
#  aoi_GLM_Data$year,
#  aoi_GLM_Data$surveyPeriod,
#  drop = TRUE
#)

#aoi_GLM_Data$eventID_num <- as.numeric(aoi_GLM_Data$eventID)

aoi_GLM_Data$stream <- as.numeric(as.factor(aoi_GLM_Data$watershed))

aoi_GLM_Data$sampleEvent <- as.numeric(aoi_GLM_Data$surveyPeriod)

aoi_GLM_Data$year_id <- as.numeric(as.factor(aoi_GLM_Data$year))

aoi_GLM_Data <- aoi_GLM_Data %>% 
  mutate(damPresence = case_when(Nhat > 0 ~ 1,
                                 Nhat == 0 ~ 0))

str(aoi_GLM_Data)



aoi_GLM_Data_mod <- aoi_GLM_Data %>%
  filter(!is.na(SWA))   # or !is.na(RI) if that’s the response


# Build explicit lookups from the data actually used in the model
stream_lookup <- aoi_GLM_Data_mod %>%
  distinct(stream, watershed) %>%
  arrange(stream)

year_lookup <- aoi_GLM_Data_mod %>%
  distinct(year_id, year) %>%
  arrange(year_id)

uda_lookup <- aoi_GLM_Data_mod %>%
  group_by(watershed) %>%
  summarise(max_catchment = max(catchment_km2, na.rm = TRUE)) %>%
  arrange(max_catchment)



# Start from the full modeled data and remove early summer time periods
# This analysis will look exclusively at late season mesic resource availability.
# Imported data has already been filtered to the target time frame. 
# 08/01-09/30
# t5:t8

aoi_glm_data_mod_refined <- aoi_GLM_Data_mod

# If this is not the case for whatever reason, use the code block below

#aoi_glm_data_mod_refined <- aoi_GLM_Data_mod %>% 
#  filter(!(sampleEvent %in% 1:4) &      # drop 2019 t1-t4
#           !(sampleEvent %in% 9:12) &   # drop 2021 t1-t4
#           !(sampleEvent %in% 17:20))   # drop 2023 t1–t4

event_lookup <- data.frame(
  old = sort(unique(aoi_glm_data_mod_refined$sampleEvent)),
  new = 1:length(unique(aoi_glm_data_mod_refined$sampleEvent))
)

aoi_glm_data_mod_refined <- aoi_glm_data_mod_refined %>%
  left_join(event_lookup, by = c("sampleEvent" = "old")) %>%
  mutate(sampleEvent_new = new) %>%
  select(-new)

event_year_3_4 <- tapply(aoi_glm_data_mod_refined$year_id,
                         aoi_glm_data_mod_refined$sampleEvent_new,
                         unique)
event_year_3_4 <- as.numeric(event_year_3_4)



aoi_glm_data_mod_refined$SWA <- ifelse(aoi_glm_data_mod_refined$SWA < 1,
                                   0,
                                   aoi_glm_data_mod_refined$SWA)
```


```{r,echo=FALSE}

# LOG TRANSFORM SWA TO ADDRESSS RIGHT SKEW

aoi_glm_data_mod_refined$logSWA <- log(aoi_glm_data_mod_refined$SWA+1)

```


Format data for hurdle:
----

```{r,echo=TRUE}
# Create binary indicator of surface water presence, for logistic part of hurdle model:
aoi_HurdleData <- aoi_glm_data_mod_refined

aoi_HurdleData$y <- ifelse(aoi_glm_data_mod_refined$SWA>0,1,0)
aoi_HurdleData$y_veg <- ifelse(aoi_glm_data_mod_refined$VA>0,1,0)
aoi_HurdleData$dam_km_std <- scale(aoi_HurdleData$dam_km)

str(aoi_HurdleData)

# write.csv(aoi_HurdleData,"aoi_HurdleData.csv")

```

----

**Visualize SWA distribution:**

----

**Raw SWA (m^2):**

```{r,echo=FALSE}
hist(aoi_HurdleData$SWA,
     main = "Distribution of SWA",
     xlab = "log(SWA)")
```

----

**Log-transform SWA to 'normalize' the right skew:**

**log(SWA):**

```{r,echo=FALSE}
hist(aoi_HurdleData$logSWA,
     main = "Distribution of log(SWA)",
     xlab = "log(SWA)")
```

This persistent spike at 0 is why we're using the hurdle approach for SWA. First model SW(0,1); then for SW = 1 model SWA.


----

**Visualize veg area (VA) distribution**

**Raw VA (m^2):**

```{r,echo=FALSE}
hist(aoi_HurdleData$VA,
     main = "Distribution of Vegetation area (VA)",
     xlab = "Vegetation Area (VA) (m^2)")
```


----

**Log-transform VA to normalize the right skew:**

**log(VA):**

```{r,echo=FALSE}
hist(log(aoi_HurdleData$VA + 1),
     main = "Distribution of log(VegArea)",
     xlab = "log(VA)")
```


Long tail, but no spike at 0. Will model log(vegArea) ~ damPres + ...

----

**Visualize NDVI distribution**

```{r,echo=FALSE}
hist(aoi_HurdleData$NDVI)
```

Pretty much normal. No transformation needed (?)