---
title: "SW_Hurdle_Models"
author: "Matt Webster"
date: "2026-07-28"
output: html_document
editor_options: 
  chunk_output_type: console
---


The purpose of this RMarkdown is to implement hurdle models for SW ~ damPresence, & SW ~ damDensity (with and without geomorphic covariates)

4 models total.
    



```{r,echo=FALSE,}
suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(ggplot2)
  library(dataRetrieval)
  library(knitr)
  library(ggplot2)
  library(zoo) #for moving window
  library(stringr)
  library(vioplot)
  library(purrr)
  library(lavaan)
  library(blavaan)
  library(ggcorrplot)
  library(lme4)
  library(broom.mixed)
  library(coda)
  library(scales)
})
```


----


**Final Data:**

```{r,echo=FALSE}
aoi_HurdleData <- read.csv("aoi_HurdleData.csv")


# add year/event index
event_year_3_4 <- tapply(aoi_HurdleData$year_id,
                         aoi_HurdleData$sampleEvent_new,
                         unique)
event_year_3_4 <- as.numeric(event_year_3_4)


# Full Hurdle data
N <- nrow(aoi_HurdleData)
y <- aoi_HurdleData$y
y2 <- aoi_HurdleData$logSWA
#y_veg <- aoi_HurdleData$y_veg
#y_vegA <- aoi_HurdleData$VA
stream <- aoi_HurdleData$stream
year <- aoi_HurdleData$year_id
event <- aoi_HurdleData$sampleEvent_new
cluster <- aoi_HurdleData$clusterID_num
damPres <- aoi_HurdleData$damPresence
damDen <- as.numeric(aoi_HurdleData$dam_km_std)
damDen_raw <- aoi_HurdleData$dam_km

n_stream <- length(unique(aoi_HurdleData$stream))
n_year <- length(unique(aoi_HurdleData$year_id))
n_event <- length(unique(aoi_HurdleData$sampleEvent_new))
n_cluster <- length(unique(aoi_HurdleData$clusterID_num))

str(aoi_HurdleData)
```


For surface water models, data of interest include:

    - y: surface water presence (0,1); response variable for logistic portion of hurdle model
    
    - y2: log-transformed surface water area log(SWA[i]+1); response variable for HGLMM portion of hurdle model
    
    - stream: index for random intercept on SWA and random slope on dam effect
    
    - year: index for random slope on dam effect
    
    - sampleEvent: index for random intercept on SWA and random slope on dam effect 
      (nested within year; represents each 2 week sample period for SW layer )
    
    - clusterID_seq: index for random intercept on SWA; represents sample pairing
    
    - damPres: dam presence (0,1); fixed effect on SWA
    
    - damDen: dams/km; z-standardized ("...$dam_km_std"); fixed effect on SWA
    
    Only used in models 2 and 4:
    
    - incis: channel incision: 1-(TO/90) (inverse of topographic openness); z-standardized; fixed effect on SWA
    
    - RIPA: riparian area (m^2); z-standardized; fixed effect on SWA 
    
    - UDA: upstream drainage area (km ^2); z-standardized; fixed effect on SWA


----

**Hurdle HGLMM**


    This analysis explores late summner surface water presence and area in relation to beaver dam presence,
    & density.
    
    All 4 models presented herein are constructed in a hurdle model framework with nested random slopes for 
    beaver dam effect to allow for stream, year, and sampleEvent to cause deviation 
    from the global mean beaver effect.
    
    
    Pt1: Surface water presence/absence; logistic regression
    
    Pt2: Surface water amount, conditional on presence; log(Normal) regression
    
    Uses log(SWA) for for 6 watersheds, 3 years, 4 sample periods (Aug-Sep) per year.


----

**Overall Takeaways**

    - Estimates may be notably inflated when geomorphic conditions are not taken into consideration (see m1 vs m2). The following regards models 2 and 4, which estimate the effect of damPresence or damDensity on surface water, conditional on geomorphic covariates:
    
    - Dam presence has very little influence on surface water presence/absence. Surface water presence is instead
      driven by local geomorphic characteristics and random effects of stream/year/season.
      
    - Likewise, 1 standard deviation increase in damDensity corresponds with a negligible increase in the odds of surface water presence.
    
    - When surface water is present, presence of beaver dam(s) is associated with a 25% increase in surface water area from baseline conditions, although there is considerable variation among watersheds and time periods.
    
    - Where surface water is present, 1 standard deviation increase in damDensity corresponds with a modest 
      increase in surface water area (15%). Higher dam densities (3-4 SD increase) do 
      notably increase surface water area relative to baseline conditionS (upwards of ~50% increase in SWA). 

    - Of course there is A LOT of variation in these patterns by stream, year, and sampling event.

----

----


```{r,echo=FALSE}
aoi_HurdleData %>%
  ggplot(aes(x = logSWA)) +
  geom_density(fill = "steelblue", alpha = 0.4) +
  facet_wrap(~ stream, scales = "free_y") +
  labs(
    title = "log-SWA distributions by stream",
    x = "log-SWA",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)

aoi_HurdleData %>%
  ggplot(aes(x = logSWA)) +
  geom_density(fill = "steelblue", alpha = 0.4) +
  facet_wrap(~ year_id, scales = "free_y") +
  labs(
    title = "log-SWA distributions by Year",
    x = "log-SWA",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)

aoi_HurdleData %>%
  ggplot(aes(x = logSWA)) +
  geom_density(fill = "steelblue", alpha = 0.4) +
  facet_wrap(~ sampleEvent_new, scales = "free_y") +
  labs(
    title = "log-SWA distributions by sampleEvent",
    x = "log-SWA",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)


```


----

----

----

----

SW ~ Dam Presence
----

----

----

**Model 1: SWA ~ DamPresence**

Effect of damPresence without geomorphic covariates

```{r,echo=FALSE}
suppressPackageStartupMessages({
library(jagsUI)
})

```
    
    sink("HGLMM_hurdle_damPresence_m1.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR SW PRESENCE (1/2 Normal on SDs):
    
      sigma_zeta_stream  ~ dnorm(0,1) T(0,)
      tau_zeta_stream    <- pow(sigma_zeta_stream, -2)
    
      sigma_zeta_year    ~ dnorm(0,1) T(0,)
      tau_zeta_year      <- pow(sigma_zeta_year, -2)
    
      sigma_zeta_event   ~ dnorm(0,1) T(0,)
      tau_zeta_event     <- pow(sigma_zeta_event, -2)
    
      sigma_iota_stream  ~ dnorm(0,1) T(0,)
      tau_iota_stream    <- pow(sigma_iota_stream, -2)
    
      sigma_iota_event   ~ dnorm(0,1) T(0,)
      tau_iota_event     <- pow(sigma_iota_event, -2)
    
      sigma_iota_cluster ~ dnorm(0,1) T(0,)
      tau_iota_cluster   <- pow(sigma_iota_cluster, -2)
    
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 1) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 1) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 1) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Presence/absence:
      epsilon ~ dnorm(0,0.04)
      zeta_damPres ~ dnorm(0,0.04)
    
      # Amount:
      alpha ~ dnorm(0,0.04)
      beta_damPres ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW PRESENCE:
    
      for(j in 1:n_stream){
        zeta_stream[j] ~ dnorm(0, tau_zeta_stream)
        iota_stream[j] ~ dnorm(0, tau_iota_stream)
      }
    
      for(yr in 1:n_year){
        zeta_year[yr] ~ dnorm(0, tau_zeta_year)
      }
    
      for(k in 1:n_event){
        zeta_event[k] ~ dnorm(0, tau_zeta_event)
        iota_event[k] ~ dnorm(0, tau_iota_event)
      }
    
      for(c in 1:n_cluster){
        iota_cluster[c] ~ dnorm(0, tau_iota_cluster)
      }
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(0, tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # DETECTION PRIORS:
      
      for(k in 1:n_event){
      p_TP[k] ~ dnorm(0.76, pow(0.16,-2)) T(0,1)   # empirically derived recall
      p_FP[k] ~ dnorm(0.01, pow(0.005,-2)) T(0,1)  # empirically derived FPR
      }
      
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Presence/absence:
        z[i] ~ dbern(psi[i])
    
        logit(psi[i]) <- epsilon
                         + (zeta_damPres
                            + zeta_stream[stream[i]]
                            + zeta_year[year[i]]
                            + zeta_event[event[i]]) * damPres[i]
                         + iota_stream[stream[i]]
                         + iota_event[event[i]]
                         + iota_cluster[cluster[i]]
    
        # Detection:
        p_eff[i] <- z[i] * p_TP[event[i]] + (1-z[i]) * p_FP[event[i]]
        y[i]     ~ dbern(p_eff[i])
    
        # Amount:
        y2[i] ~ dnorm(mu[i], tau_eff[i])
    
        mu[i] <- z[i] *
                  (alpha
                    + (beta_damPres
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damPres[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]])
    
        tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
      }
    
    }
    ", fill = TRUE)
    
    sink()
   
    

```{r,echo=FALSE}
hurdle_damPres_m1_data_list <- list(
  N = N,
  y = y,
  y2 = y2,
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damPres = damPres,
  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)),
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hurdle_damPres_m1_inits <- function(){
  list(
    alpha        = rnorm(1, 0, 1),
    epsilon      = rnorm(1, 0, 1),

    sigma        = runif(1, 0.1, 2),
    sigma_zeta_stream  = runif(1, 0.1, 2),
    sigma_zeta_year    = runif(1, 0.1, 2),
    sigma_zeta_event   = runif(1, 0.1, 2),
    sigma_iota_stream  = runif(1, 0.1, 2),
    sigma_iota_event   = runif(1, 0.1, 2),
    sigma_iota_cluster = runif(1, 0.1, 2),

    sigma_beta_stream  = runif(1, 0.1, 2),
    sigma_beta_year    = runif(1, 0.1, 2),
    sigma_beta_event   = runif(1, 0.1, 2),
    sigma_eta_stream   = runif(1, 0.1, 2),
    sigma_eta_event    = runif(1, 0.1, 2),
    sigma_eta_cluster  = runif(1, 0.1, 2)
  )
}

hurdle_damPres_m1_params <- c(
  "psi",
  "mu",
  "epsilon",
  "alpha",
  "zeta_damPres",
  "zeta_incis",
  "zeta_RIPA",
  "zeta_UDA",
  "zeta_stream",
  "zeta_year",
  "zeta_event",
  "iota_stream",
  "iota_event",
  "iota_cluster",
  "beta_damPres",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_zeta_stream",
  "sigma_zeta_year",
  "sigma_zeta_event",
  "sigma_iota_stream",
  "sigma_iota_event",
  "sigma_iota_cluster",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_event",
  "sigma_eta_cluster",
  "p_TP",
  "p_FP",
  "z",
  "tau_eff"
)

nchains <- 3
nthin <- 5
niter <- 100000
nburn <- 25000

Sys.time()

hurdle_damPres_m1_SW <- jags(data = hurdle_damPres_m1_data_list,
                   inits = hurdle_damPres_m1_inits,
                   parameters.to.save = hurdle_damPres_m1_params,
                   model.file = "HGLMM_hurdle_damPresence_m1.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hurdle_damPres_m1_SW) # model with local and watershed dam density as fixed effects
#print(hurdle_damPres_m1_SW$summary[grep("beta_", rownames(hurdle_damPres_m1_SW$summary)), ])
#print(hurdle_damPres_m1_SW$summary[grep("zeta_", rownames(hurdle_damPres_m1_SW$summary)), ])

```


----

DIAGNOSTICS M1:
----

```{r,echo=FALSE}
# Plotting specifications from Rmarkdown.

par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

```


```{r,echo=FALSE}
#Extract posteriors
damPres_m1_post <- hurdle_damPres_m1_SW$sims.list
```

**Assess "MCMC chain health":**

**Rubin-Gelman Rhat** -- assess convergence

**Effective Sample Size** -- assess posterior independence

```{r,echo=FALSE}
# Choose only the parameters you care about
params_m1 <- c(
  "epsilon",
  "alpha",
  "zeta_damPres",
  "beta_damPres",
  "sigma",
  "sigma_zeta_stream",
  "sigma_iota_cluster",
  "p_TP[1]",
  "p_FP[1]")

# Gelman–Rubin convergence diagnostic
gelman.diag(hurdle_damPres_m1_SW$samples[, params_m1], multivariate = FALSE)

# Effective sample size
effectiveSize(hurdle_damPres_m1_SW$samples[, params_m1])
```



    Good Rhat values, all around 1.00.
    
    Mostly high effective sample sizes. 
    
    Note: beta_damPres has the most trouble converging; however, Rhat near 1, 
    and moderate ESS for beta_damPres suggest that the parameter is still 
    reliably usable. 


----

**Check traceplots for critical parameters**

```{r,echo=FALSE}
tplot_m1_params <- c("alpha","beta_damPres","zeta_damPres", "sigma","sigma_zeta_stream","sigma_beta_stream")


m1_mcmc_sub <- hurdle_damPres_m1_SW$samples[, tplot_m1_params]

coda::traceplot(m1_mcmc_sub)
```


    Traceplots are all mostly grassy. Again, see beta_damPres. Slow to 
    converge; expected, given Rhat and ESS.


----

**Posterior predictive plots:**

Predicted Psi (m1_psi_post) vs observed SW Presence (y)

```{r,echo=FALSE}

par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

# Posterior mean psi
m1_psi_post <- apply(hurdle_damPres_m1_SW$sims.list$psi, 2, mean)

# Compare to observed y
plot(m1_psi_post, y, pch=16, col=rgb(0,0,0,0.2))


```


Predicted log(SWA) (m1_mu_post) vs detected log(SWA) (y2)

Goal is 1:1 (ish)

```{r,echo=FALSE}
m1_mu_post <- apply(hurdle_damPres_m1_SW$sims.list$mu, 2, mean)

plot(m1_mu_post[y==1], y2[y==1], pch=16, col=rgb(0,0,1,0.3))
abline(0,1,col="red")
```

----

**Random effects shrinkage: **

Goal is near 0(ish)

1. zeta_stream; random effect of stream on sw presence
2. beta_stream; random effect of stream on sw amount
3. zeta_event; random effect of sample event on sw presence
4. beta_event; random effect of sample event on sw amount

```{r, echo=FALSE, }
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

boxplot(hurdle_damPres_m1_SW$sims.list$zeta_stream,
        main = "zeta_stream")
boxplot(hurdle_damPres_m1_SW$sims.list$beta_stream,
        main = "beta_stream")
boxplot(hurdle_damPres_m1_SW$sims.list$zeta_event,
        main = "zeta_event")
boxplot(hurdle_damPres_m1_SW$sims.list$beta_event,
        main = "beta_event")
```

<!---
Hyperpriors on stream effect: goal sd 0.1:2

```{r,echo=FALSE}

summary(hurdle_damPres_m1_SW$sims.list$sigma_zeta_stream)
summary(hurdle_damPres_m1_SW$sims.list$sigma_beta_stream)

```

Check detection posteriors: goal is near empirical mean recall

```{r,echo=FALSE}
summary(hurdle_damPres_m1_SW$sims.list$p_TP)

summary(hurdle_damPres_m1_SW$sims.list$p_FP)
```

    TP is high and precise; FP is low and precise. This makes sense because these are based on empirical mean() and sd() recall and flase positive rates from the RF classifier (see multiprobability threshod validation).


Check z latent states

```{r,echo=FALSE}
z_post <- apply(hurdle_damPres_m1_SW$sims.list$z, 2, mean)

hist(z_post, main = "Split of posterior z")
```

    Latent z's split mostly evenly

Check tau_eff behavior:

```{r,echo=FALSE}
summary(hurdle_damPres_m1_SW$sims.list$tau_eff)

```

    This shows that the tau_eff trick worked: 
    
    tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
    
    so when z is 1, tau_eff is reasonable. when z is 0, tau_eff is 100 and the prior distribution is basically flat

Check dam effects:
--->


----


**Posterior mean dam effects:**

damPres --> swPres (logistic regression)

```{r,echo=FALSE}
summary(hurdle_damPres_m1_SW$sims.list$zeta_damPres)

exp(mean(hurdle_damPres_m1_SW$sims.list$zeta_damPres))
```

    When you don't account for geomorphic conditions, dam presence corresponds to an estimated 40-50% increase in 
    the odds of surface‑water presence.
    
----


damPres --> swAmount (hglmm)

```{r,echo=FALSE}
summary(hurdle_damPres_m1_SW$sims.list$beta_damPres)

exp(mean(hurdle_damPres_m1_SW$sims.list$beta_damPres))
```

    When you don't account for geomorphic conditions, dam presence corresponds to an estimated 30-40% increase 
    in surface water area.


----



**Prior-posterior overlap:**

```{r,echo=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 


damPres_m1_alpha_post <- hurdle_damPres_m1_SW$sims.list$alpha
damPres_m1_beta_post  <- hurdle_damPres_m1_SW$sims.list$beta_damPres
damPres_m1_zeta_post  <- hurdle_damPres_m1_SW$sims.list$zeta_damPres
damPres_m1_sigBetaStream <- hurdle_damPres_m1_SW$sims.list$sigma_beta_stream

damPres_m1Mod_n <- length(damPres_m1_alpha_post)

damPres_m1_alpha_prior <- rnorm(damPres_m1Mod_n, 0, sqrt(1/0.04))
damPres_m1_beta_prior  <- rnorm(damPres_m1Mod_n, 0, sqrt(1/0.04))
damPres_m1_zeta_prior  <- rnorm(damPres_m1Mod_n, 0, sqrt(1/0.04))
damPres_m1_sigBetaStream_prior <- pmax(rnorm(damPres_m1Mod_n, 0, 1), 0)


# Plot overlap:

plot_prior_post <- function(prior, post, name) {
  
  df <- tibble(
    value = c(prior, post),
    dist  = rep(c("Prior", "Posterior"), each = length(prior))
  )
  
  ggplot(df, aes(x = value, fill = dist)) +
    geom_density(alpha = 0.4) +
    scale_fill_manual(values = c("gray60", "steelblue")) +
    labs(
      title = paste("Prior vs Posterior for", name),
      x = name,
      y = "Density"
    ) +
    theme_minimal(base_size = 14)
}

plot_prior_post(damPres_m1_alpha_prior, damPres_m1_alpha_post, "alpha_m1")
plot_prior_post(damPres_m1_beta_prior, damPres_m1_beta_post, "beta_damPres_m1")
plot_prior_post(damPres_m1_zeta_prior, damPres_m1_zeta_post, "zeta_damPres_m1")
plot_prior_post(damPres_m1_sigBetaStream_prior, damPres_m1_sigBetaStream, "sigma_beta_stream_m1")

```

    These PPO plots indicate that the priors are reasonable, but are not driving the model -- 
    the data are sufficiently informing the model, and shifting the posterior distributions 
    away from those prior expectations


----

----

----

Now, include geomorphic covariates:

----

----

----


**Model 2: SWA ~ DamPresence + Geom.Covar.**


```{r,echo=FALSE}
#library(jagsUI)
```
    sink("HGLMM_hurdle_damPresence_m2.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR SW PRESENCE (1/2 Normal on SDs):
    
      sigma_zeta_stream  ~ dnorm(0,1) T(0,)
      tau_zeta_stream    <- pow(sigma_zeta_stream, -2)
    
      sigma_zeta_year    ~ dnorm(0,1) T(0,)
      tau_zeta_year      <- pow(sigma_zeta_year, -2)
    
      sigma_zeta_event   ~ dnorm(0,1) T(0,)
      tau_zeta_event     <- pow(sigma_zeta_event, -2)
    
      sigma_iota_stream  ~ dnorm(0,1) T(0,)
      tau_iota_stream    <- pow(sigma_iota_stream, -2)
    
      sigma_iota_event   ~ dnorm(0,1) T(0,)
      tau_iota_event     <- pow(sigma_iota_event, -2)
    
      sigma_iota_cluster ~ dnorm(0,1) T(0,)
      tau_iota_cluster   <- pow(sigma_iota_cluster, -2)
    
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 1) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 1) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 1) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Presence/absence:
      epsilon ~ dnorm(0,0.04)
      zeta_damPres ~ dnorm(0,0.04)
      zeta_incis ~ dnorm(0,0.04)
      zeta_RIPA ~ dnorm(0,0.04)
      zeta_UDA ~ dnorm(0,0.04)
    
      # Amount:
      alpha ~ dnorm(0,0.04)
      beta_damPres ~ dnorm(0,0.04)
      beta_incis ~ dnorm(0,0.04)
      beta_RIPA ~ dnorm(0,0.04)
      beta_UDA ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW PRESENCE:
    
      for(j in 1:n_stream){
        zeta_stream[j] ~ dnorm(0, tau_zeta_stream)
        iota_stream[j] ~ dnorm(0, tau_iota_stream)
      }
    
      for(yr in 1:n_year){
        zeta_year[yr] ~ dnorm(0, tau_zeta_year)
      }
    
      for(k in 1:n_event){
        zeta_event[k] ~ dnorm(0, tau_zeta_event)
        iota_event[k] ~ dnorm(0, tau_iota_event)
      }
    
      for(c in 1:n_cluster){
        iota_cluster[c] ~ dnorm(0, tau_iota_cluster)
      }
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(0, tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # DETECTION PRIORS:
      
      for(k in 1:n_event){
      p_TP[k] ~ dnorm(0.76, pow(0.16,-2)) T(0,1)   # empirically derived recall
      p_FP[k] ~ dnorm(0.01, pow(0.005,-2)) T(0,1)  # empirically derived FPR
      }
      
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Presence/absence:
        z[i] ~ dbern(psi[i])
    
        logit(psi[i]) <- epsilon
                         + (zeta_damPres
                            + zeta_stream[stream[i]]
                            + zeta_year[year[i]]
                            + zeta_event[event[i]]) * damPres[i]
                         + zeta_incis * incis[i]
                         + zeta_RIPA * RIPA[i]
                         + zeta_UDA * UDA[i]
                         + iota_stream[stream[i]]
                         + iota_event[event[i]]
                         + iota_cluster[cluster[i]]
    
        # Detection:
        p_eff[i] <- z[i] * p_TP[event[i]] + (1-z[i]) * p_FP[event[i]]
        y[i]     ~ dbern(p_eff[i])
    
        # Amount:
        y2[i] ~ dnorm(mu[i], tau_eff[i])
    
        mu[i] <- z[i] *
                  (alpha
                    + (beta_damPres
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damPres[i]
                    + beta_incis * incis[i] 
                    + beta_RIPA * RIPA[i]
                    + beta_UDA * UDA[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]])
    
        tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
      }
    
    }
    ", fill = TRUE)
    
    sink()
   
    

```{r,echo=FALSE}
hurdle_damPres_m2_data_list <- list(
  N = N,
  y = y,
  y2 = y2,
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damPres = damPres,
  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)),
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hurdle_damPres_m2_inits <- function(){
  list(
    alpha        = rnorm(1, 0, 1),
    epsilon      = rnorm(1, 0, 1),

    sigma        = runif(1, 0.1, 2),
    sigma_zeta_stream  = runif(1, 0.1, 2),
    sigma_zeta_year    = runif(1, 0.1, 2),
    sigma_zeta_event   = runif(1, 0.1, 2),
    sigma_iota_stream  = runif(1, 0.1, 2),
    sigma_iota_event   = runif(1, 0.1, 2),
    sigma_iota_cluster = runif(1, 0.1, 2),

    sigma_beta_stream  = runif(1, 0.1, 2),
    sigma_beta_year    = runif(1, 0.1, 2),
    sigma_beta_event   = runif(1, 0.1, 2),
    sigma_eta_stream   = runif(1, 0.1, 2),
    sigma_eta_event    = runif(1, 0.1, 2),
    sigma_eta_cluster  = runif(1, 0.1, 2)
  )
}

hurdle_damPres_m2_params <- c(
  "psi",
  "mu",
  "epsilon",
  "alpha",
  "zeta_damPres",
  "zeta_incis",
  "zeta_RIPA",
  "zeta_UDA",
  "zeta_stream",
  "zeta_year",
  "zeta_event",
  "iota_stream",
  "iota_event",
  "iota_cluster",
  "beta_damPres",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_zeta_stream",
  "sigma_zeta_year",
  "sigma_zeta_event",
  "sigma_iota_stream",
  "sigma_iota_event",
  "sigma_iota_cluster",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_event",
  "sigma_eta_cluster",
  "p_TP",
  "p_FP",
  "z",
  "tau_eff"
)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000

Sys.time()

hurdle_damPres_m2_SW <- jags(data = hurdle_damPres_m2_data_list,
                   inits = hurdle_damPres_m2_inits,
                   parameters.to.save = hurdle_damPres_m2_params,
                   model.file = "HGLMM_hurdle_damPresence_m2.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hurdle_damPres_m2_SW) # model with local and watershed dam density as fixed effects
#print(hurdle_damPres_m2_SW$summary[grep("beta_", rownames(hurdle_damPres_m2_SW$summary)), ])
#print(hurdle_damPres_m2_SW$summary[grep("zeta_", rownames(hurdle_damPres_m2_SW$summary)), ])

```


----

DIAGNOSTICS M2:
----

```{r,echo=FALSE}
# Plotting specifications from Rmarkdown.

par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

```


```{r,echo=FALSE}
#Extract posteriors
damPres_m2_post <- hurdle_damPres_m2_SW$sims.list
```

**Assess "MCMC chain health":**

**Rubin-Gelman Rhat** -- assess convergence

**Effective Sample Size** -- assess posterior independence

```{r,echo=FALSE}
# Choose only the parameters you care about
params <- c(
  "epsilon",
  "alpha",
  "zeta_damPres",
  "beta_damPres",
  "zeta_incis",
  "beta_incis",
  "zeta_RIPA",
  "beta_RIPA",
  "zeta_UDA",
  "beta_UDA",
  "sigma",
  "sigma_zeta_stream",
  "sigma_iota_cluster",
  "p_TP[1]",
  "p_FP[1]")

# Gelman–Rubin convergence diagnostic
gelman.diag(hurdle_damPres_m2_SW$samples[, params], multivariate = FALSE)

# Effective sample size
effectiveSize(hurdle_damPres_m2_SW$samples[, params])

```



    Decent Rhat values, all around 1.00.
    
    Mostly high effective sample sizes. 
    
    Same slow convergence for beta_damPres (hard to estimate)


----

**Check traceplots for critical parameters**

```{r,echo=FALSE}
m2_tplot_params <- c("alpha","beta_damPres","zeta_damPres", "sigma","sigma_zeta_stream","sigma_beta_stream")


m2_mcmc_sub <- hurdle_damPres_m2_SW$samples[, m2_tplot_params]

coda::traceplot(m2_mcmc_sub)
```


    Traceplots are all mostly grassy. Again, see beta_damPres. REAL slow to 
    converge; expected, given Rhat and ESS.


----

**Posterior predictive plots:**

Predicted Psi (psi_post) vs observed SW Presence (y)

```{r,echo=FALSE}

par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

# Posterior mean psi
psi_post <- apply(hurdle_damPres_m2_SW$sims.list$psi, 2, mean)

# Compare to observed y
plot(psi_post, y, pch=16, col=rgb(0,0,0,0.2))


```


Predicted log(SWA) (mu_post[y==1]) vs detected SWA (y2[y==1]) conditional on SW presence

Goal is 1:1 (ish)

```{r,echo=FALSE}
mu_post <- apply(hurdle_damPres_m2_SW$sims.list$mu, 2, mean)

plot(mu_post[y==1], y2[y==1], pch=16, col=rgb(0,0,1,0.3))
abline(0,1,col="red")


```

----

**Random effects shrinkage: **

Goal is near 0(ish)

1. zeta_stream; random effect of stream on sw presence
2. beta_stream; random effect of stream on sw amount
3. zeta_event; random effect of sample event on sw presence
4. beta_event; random effect of sample event on sw amount

```{r, echo=FALSE, }
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

boxplot(hurdle_damPres_m2_SW$sims.list$zeta_stream,
        main = "zeta_stream")
boxplot(hurdle_damPres_m2_SW$sims.list$beta_stream,
        main = "beta_stream")
boxplot(hurdle_damPres_m2_SW$sims.list$zeta_event,
        main = "zeta_event")
boxplot(hurdle_damPres_m2_SW$sims.list$beta_event,
        main = "beta_event")
```

<!---
Hyperpriors on stream effect: goal sd 0.1:2

```{r,echo=FALSE}

summary(hurdle_damPres_m2_SW$sims.list$sigma_zeta_stream)
summary(hurdle_damPres_m2_SW$sims.list$sigma_beta_stream)

```

Check detection posteriors: goal is near empirical mean recall

```{r,echo=FALSE}
summary(hurdle_damPres_m2_SW$sims.list$p_TP)

summary(hurdle_damPres_m2_SW$sims.list$p_FP)
```

    TP is high and precise; FP is low and precise. This makes sense because these are based on empirical mean() and sd() recall and flase positive rates from the RF classifier (see multiprobability threshod validation).


Check z latent states

```{r,echo=FALSE}
z_post <- apply(hurdle_damPres_m2_SW$sims.list$z, 2, mean)

hist(z_post, main = "Split of posterior z")
```

    Latent z's split mostly evenly

Check tau_eff behavior:

```{r,echo=FALSE}
summary(hurdle_damPres_m2_SW$sims.list$tau_eff)

```

    This shows that the tau_eff trick worked: 
    
    tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
    
    so when z is 1, tau_eff is reasonable. when z is 0, tau_eff is 100 and the prior distribution is basically flat

Check dam effects:
--->


----


**Posterior mean dam effects:**

damPres --> swPres (logistic regression)

```{r,echo=FALSE}
summary(hurdle_damPres_m2_SW$sims.list$zeta_damPres)

exp(mean(hurdle_damPres_m2_SW$sims.list$zeta_damPres))
```

    When you consider local geomorphic conditions, dam presence increases the odds of surface‑water presence by
    only about 6% (compared to 40-50% w/o geom.covar.). This indicates that dam presence has a negligible
    effect on whether water is present at all.
    


----


damPres --> swAmount (hglmm)

```{r,echo=FALSE}
summary(hurdle_damPres_m2_SW$sims.list$beta_damPres)

exp(mean(hurdle_damPres_m2_SW$sims.list$beta_damPres))
```

    Once geom. covar. are accounted for, dam presence corresponds with an estimated
    ~20-25% increase in surface water area


----



**Prior-posterior overlap:**


```{r,echo=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)


damPres_m2_alpha_post <- hurdle_damPres_m2_SW$sims.list$alpha
damPres_m2_beta_post  <- hurdle_damPres_m2_SW$sims.list$beta_damPres
damPres_m2_zeta_post  <- hurdle_damPres_m2_SW$sims.list$zeta_damPres
damPres_m2_sigBetaStream <- hurdle_damPres_m2_SW$sims.list$sigma_beta_stream

damPres_m2Mod_n <- length(damPres_m2_alpha_post)

damPres_m2_alpha_prior <- rnorm(damPres_m2Mod_n, 0, sqrt(1/0.04))
damPres_m2_beta_prior  <- rnorm(damPres_m2Mod_n, 0, sqrt(1/0.04))
damPres_m2_zeta_prior  <- rnorm(damPres_m2Mod_n, 0, sqrt(1/0.04))
damPres_m2_sigBetaStream_prior <- pmax(rnorm(damPres_m2Mod_n, 0, 1), 0)


# Plot overlap:

#plot_prior_post <- function(prior, post, name) {
  
#  df <- tibble(
#    value = c(prior, post),
#    dist  = rep(c("Prior", "Posterior"), each = length(prior))
#  )
  
#  ggplot(df, aes(x = value, fill = dist)) +
#    geom_density(alpha = 0.4) +
#    scale_fill_manual(values = c("gray60", "steelblue")) +
#    labs(
#      title = paste("Prior vs Posterior for", name),
#      x = name,
#      y = "Density"
#    ) +
#    theme_minimal(base_size = 14)
#}

plot_prior_post(damPres_m2_alpha_prior, damPres_m2_alpha_post, "alpha_m2")
plot_prior_post(damPres_m2_beta_prior, damPres_m2_beta_post, "beta_damPres_m2")
plot_prior_post(damPres_m2_zeta_prior, damPres_m2_zeta_post, "zeta_damPres_m2")
plot_prior_post(damPres_m2_sigBetaStream_prior, damPres_m2_sigBetaStream, "sigma_beta_stream_m2")

```

    PPOs still suggest data driven model. 


----


**Marginal effect** of dam presence on surface water presence, given geomorphic context.

**Note** there's a  TINY positive effect of dam presence on surface water presence. Not really perceptible from these plots. 

```{r,echo=FALSE, warning=FALSE}
# Build a grid of raw geomorphic covariate values
damPres_m2_geom_grid <- tibble(
  TO_raw = seq(min(aoi_HurdleData$TO),
               max(aoi_HurdleData$TO), length.out = 50),

  incis_raw = 1 - (TO_raw / 90),   # raw incision scale

  RIPA_raw = seq(min(aoi_HurdleData$riparian_area_m2),
                 max(aoi_HurdleData$riparian_area_m2), length.out = 50),

  UDA_raw  = seq(min(aoi_HurdleData$catchment_km2),
                 max(aoi_HurdleData$catchment_km2), length.out = 50)
) %>%
  # Convert raw → standardized (to match model scale)
  mutate(
    incis = as.numeric(scale(incis_raw)),
    RIPA  = as.numeric(scale(RIPA_raw)),
    UDA   = as.numeric(scale(UDA_raw))
  )

damPres_m2_compute_psi <- function(dam_flag, geom_grid, damPres_m2_post) {

  eps <- mean(damPres_m2_post$epsilon)
  zd  <- mean(damPres_m2_post$zeta_damPres)
  zi  <- mean(damPres_m2_post$zeta_incis)
  zr  <- mean(damPres_m2_post$zeta_RIPA)
  zu  <- mean(damPres_m2_post$zeta_UDA)

  geom_grid %>%
    mutate(
      psi = plogis(
        eps +
        zd * dam_flag +
        zi * incis +      # standardized incision
        zr * RIPA +       # standardized RIPA
        zu * UDA          # standardized UDA
      ),
      dam = dam_flag
    )
}


```

```{r,echo=FALSE}
damPres_m2_psi_nodam <- damPres_m2_compute_psi(0, damPres_m2_geom_grid, damPres_m2_post)
damPres_m2_psi_dam   <- damPres_m2_compute_psi(1, damPres_m2_geom_grid, damPres_m2_post)

damPres_m2_psi_df <- bind_rows(damPres_m2_psi_nodam, damPres_m2_psi_dam)
```

```{r,echo=FALSE}
ggplot(damPres_m2_psi_df, aes(x = incis_raw, y = psi, color = factor(dam))) +
  geom_line(size = 1.2) +
  labs(
    title = "Marginal Effect of Dam Presence on Surface Water Presence",
    subtitle = "Predicted probability of water (psi) across raw incision values",
    x = "Channel Incision (raw scale: 1 - TO/90)",
    y = "Probability of Surface Water",
    color = "Dam Presence"
  ) +
  scale_color_manual(values = c("brown4","steelblue"),
                     labels = c("No Dam", "Dam Present")) +
  theme_minimal(base_size = 14)

```

```{r,echo=FALSE}
ggplot(damPres_m2_psi_df, aes(x = RIPA_raw, y = psi, color = factor(dam))) +
  geom_line(size = 1.2) +
  labs(
    title = "Marginal Effect of Dam Presence on Surface Water Presence",
    subtitle = "Predicted psi across raw riparian area (m^2)",
    x = "Riparian Area (m^2)",
    y = "Probability of Surface Water",
    color = "Dam Presence"
  ) +
  scale_color_manual(values = c("brown4","steelblue"),
                     labels = c("No Dam", "Dam Present")) +
  theme_minimal(base_size = 14)


```

```{r,echo=FALSE}
ggplot(damPres_m2_psi_df, aes(x = UDA_raw, y = psi, color = factor(dam))) +
  geom_line(size = 1.2) +
  labs(
    title = "Marginal Effect of Dam Presence on Surface Water Presence",
    subtitle = "Predicted psi across raw contributing drainage area (km^2)",
    x = "Contributing Drainage Area (km^2)",
    y = "Probability of Surface Water",
    color = "Dam Presence"
  ) +
  scale_color_manual(values = c("brown4","steelblue"),
                     labels = c("No Dam", "Dam Present")) +
  theme_minimal(base_size = 14)


```


----


**Partial dependence plots**

Constrain PDP to empirical ranges for SWA and target covariates

```{r,echo=FALSE}

damPres_m2_post <- hurdle_damPres_m2_SW$sims.list

# Raw covariates
incis_raw <- 1 - aoi_HurdleData$TO/90
RIPA_raw  <- aoi_HurdleData$riparian_area_m2
UDA_raw   <- aoi_HurdleData$catchment_km2

logSWA_mean <- mean(aoi_HurdleData$logSWA)
logSWA_sd   <- sd(aoi_HurdleData$logSWA)

# Standardization stats
incis_mean_raw <- mean(incis_raw)
incis_sd_raw   <- sd(incis_raw)

RIPA_mean_raw  <- mean(RIPA_raw)
RIPA_sd_raw    <- sd(RIPA_raw)

UDA_mean_raw   <- mean(UDA_raw)
UDA_sd_raw     <- sd(UDA_raw)


##############################3

# Helper Functions to Compute SWA from posterior draws
compute_swa_draws <- function(incis, RIPA, UDA, post, dam_flag) {
  n_draws <- length(damPres_m2_post$alpha)
  swa_mat <- matrix(NA_real_, nrow = length(incis), ncol = n_draws)

  for (j in seq_len(n_draws)) {
    lp <- damPres_m2_post$alpha[j] +
          damPres_m2_post$beta_incis[j] * incis +
          damPres_m2_post$beta_RIPA[j]  * RIPA +
          damPres_m2_post$beta_UDA[j]   * UDA +
          damPres_m2_post$beta_damPres[j] * dam_flag

    swa_mat[, j] <- exp(lp) - 1
  }

  swa_mat
}

summarize_swa <- function(swa_mat) {
  tibble(
    SWA_med = apply(swa_mat, 1, median),
    SWA_lo  = apply(swa_mat, 1, quantile, probs = 0.025),
    SWA_hi  = apply(swa_mat, 1, quantile, probs = 0.975)
  )
}




#########################################

# PDP Builder:

build_pdp <- function(focal_raw, focal_name,
                      incis_raw, RIPA_raw, UDA_raw,
                      post,
                      incis_mean_raw, incis_sd_raw,
                      RIPA_mean_raw, RIPA_sd_raw,
                      UDA_mean_raw, UDA_sd_raw) {

  # Sequence only across empirical range
  focal_seq <- seq(min(focal_raw), max(focal_raw), length.out = 200)

  # Conditional medians of other covariates
  incis_med <- median(incis_raw)
  RIPA_med  <- median(RIPA_raw)
  UDA_med   <- median(UDA_raw)

  # Build grid (200 rows)
  grid <- tibble(
    focal_raw = focal_seq,
    incis_raw = if (focal_name == "incis") focal_seq else incis_med,
    RIPA_raw  = if (focal_name == "RIPA")  focal_seq else RIPA_med,
    UDA_raw   = if (focal_name == "UDA")   focal_seq else UDA_med
  ) %>%
    mutate(
      incis = (incis_raw - incis_mean_raw) / incis_sd_raw,
      RIPA  = (RIPA_raw  - RIPA_mean_raw)  / RIPA_sd_raw,
      UDA   = (UDA_raw   - UDA_mean_raw)   / UDA_sd_raw
    )

  # Compute SWA draws
  swa_nodam <- compute_swa_draws(grid$incis, grid$RIPA, grid$UDA, post, dam_flag = 0)
  swa_dam   <- compute_swa_draws(grid$incis, grid$RIPA, grid$UDA, post, dam_flag = 1)

  # Summaries (400 rows)
  pdp <- bind_rows(
    summarize_swa(swa_nodam) %>% mutate(dam = "No Dam"),
    summarize_swa(swa_dam)   %>% mutate(dam = "Dam Present")
  )

  # Repeat grid twice (400 rows)
  grid2 <- bind_rows(grid, grid)

  # Bind columns safely
  bind_cols(grid2, pdp)
}



#####

pdp_incis <- build_pdp(
  focal_raw = incis_raw,
  focal_name = "incis",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_m2_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)

pdp_RIPA <- build_pdp(
  focal_raw = RIPA_raw,
  focal_name = "RIPA",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_m2_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)

pdp_UDA <- build_pdp(
  focal_raw = UDA_raw,
  focal_name = "UDA",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_m2_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)



##########################################

# Plot Function:

plot_pdp <- function(pdp, xvar, xtitle) {
  ggplot(pdp,
         aes(x = .data[[xvar]], y = SWA_med, color = dam, fill = dam)) +
    geom_ribbon(aes(ymin = SWA_lo, ymax = SWA_hi), alpha = 0.2, color = NA) +
    geom_line(size = 1.2) +
    scale_color_manual(values = c("steelblue", "brown4")) +
    scale_fill_manual(values = c("steelblue", "brown4")) +
    scale_y_continuous(labels = comma) +
    labs(
      title = paste("Partial Dependence of SWA on", xtitle),
      subtitle = "Predicted SWA with 95% credible intervals\nConstrained to empirical covariate range",
      x = xtitle,
      y = "Predicted SWA (m²)",
      color = "Dam Presence",
      fill  = "Dam Presence"
    ) +
    theme_minimal(base_size = 14)
}


###########################################3


plot_pdp(pdp_incis, "incis_raw", "Channel Incision (raw)")
plot_pdp(pdp_RIPA,  "RIPA_raw",  "Riparian Area (m²)")
plot_pdp(pdp_UDA,   "UDA_raw",   "Contributing Drainage Area (km²)")


```


----

----

----

----

----

----

----

----

----


SW ~ Dam Density
----


**Hurdle HGLMM**


**MODEL 3: SW ~ damDensity (no Geom.Covar.)**

```{r,echo=FALSE}
#library(jagsUI)
```
    sink("HGLMM_hurdle_damDensity_m3.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR SW PRESENCE (1/2 Normal on SDs):
    
      sigma_zeta_stream  ~ dnorm(0,1) T(0,)
      tau_zeta_stream    <- pow(sigma_zeta_stream, -2)
    
      sigma_zeta_year    ~ dnorm(0,1) T(0,)
      tau_zeta_year      <- pow(sigma_zeta_year, -2)
    
      sigma_zeta_event   ~ dnorm(0,1) T(0,)
      tau_zeta_event     <- pow(sigma_zeta_event, -2)
    
      sigma_iota_stream  ~ dnorm(0,1) T(0,)
      tau_iota_stream    <- pow(sigma_iota_stream, -2)
    
      sigma_iota_event   ~ dnorm(0,1) T(0,)
      tau_iota_event     <- pow(sigma_iota_event, -2)
    
      sigma_iota_cluster ~ dnorm(0,1) T(0,)
      tau_iota_cluster   <- pow(sigma_iota_cluster, -2)
    
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 4) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 4) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 4) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Presence/absence:
      epsilon ~ dnorm(0,0.04)
      zeta_damDen ~ dnorm(0,0.04)
    
      # Amount:
      alpha ~ dnorm(0,0.04)
      beta_damDen ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW PRESENCE:
    
      for(j in 1:n_stream){
        zeta_stream[j] ~ dnorm(0, tau_zeta_stream)
        iota_stream[j] ~ dnorm(0, tau_iota_stream)
      }
    
      for(yr in 1:n_year){
        zeta_year[yr] ~ dnorm(0, tau_zeta_year)
      }
    
      for(k in 1:n_event){
        zeta_event[k] ~ dnorm(0, tau_zeta_event)
        iota_event[k] ~ dnorm(0, tau_iota_event)
      }
    
      for(c in 1:n_cluster){
        iota_cluster[c] ~ dnorm(0, tau_iota_cluster)
      }
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(0, tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # DETECTION PRIORS:
      
      for(k in 1:n_event){
      p_TP[k] ~ dnorm(0.76, pow(0.16,-2)) T(0,1)   # empirically derived recall
      p_FP[k] ~ dnorm(0.01, pow(0.005,-2)) T(0,1)  # empirically derived FPR
      }
      
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Presence/absence:
        z[i] ~ dbern(psi[i])
    
        logit(psi[i]) <- epsilon
                         + (zeta_damDen
                            + zeta_stream[stream[i]]
                            + zeta_year[year[i]]
                            + zeta_event[event[i]]) * damDen[i]
                         + iota_stream[stream[i]]
                         + iota_event[event[i]]
                         + iota_cluster[cluster[i]]
    
        # Detection:
        p_eff[i] <- z[i] * p_TP[event[i]] + (1-z[i]) * p_FP[event[i]]
        y[i]     ~ dbern(p_eff[i])
    
        # Amount:
        y2[i] ~ dnorm(mu[i], tau_eff[i])
    
        mu[i] <- z[i] *
                  (alpha
                    + (beta_damDen
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damDen[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]])
    
        tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
      }
    
    }
    ", fill = TRUE)
    
    sink()
  
    

```{r,echo=FALSE}
hurdle_damDen_m3_data_list <- list(
  N = N,
  y = y,
  y2 = y2,
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damDen = damDen,
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hurdle_damDen_m3_inits <- function(){
  list(
    alpha        = rnorm(1, 0, 1),
    epsilon      = rnorm(1, 0, 1),

    sigma        = runif(1, 0.1, 2),
    sigma_zeta_stream  = runif(1, 0.1, 2),
    sigma_zeta_year    = runif(1, 0.1, 2),
    sigma_zeta_event   = runif(1, 0.1, 2),
    sigma_iota_stream  = runif(1, 0.1, 2),
    sigma_iota_event   = runif(1, 0.1, 2),
    sigma_iota_cluster = runif(1, 0.1, 2),

    sigma_beta_stream  = runif(1, 0.1, 2),
    sigma_beta_year    = runif(1, 0.1, 2),
    sigma_beta_event   = runif(1, 0.1, 2),
    sigma_eta_stream   = runif(1, 0.1, 2),
    sigma_eta_event    = runif(1, 0.1, 2),
    sigma_eta_cluster  = runif(1, 0.1, 2)
  )
}

hurdle_damDen_m3_params <- c(
  "psi",
  "mu",
  "epsilon",
  "alpha",
  "zeta_damDen",
  "zeta_stream",
  "zeta_year",
  "zeta_event",
  "iota_stream",
  "iota_event",
  "iota_cluster",
  "beta_damDen",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_zeta_stream",
  "sigma_zeta_year",
  "sigma_zeta_event",
  "sigma_iota_stream",
  "sigma_iota_event",
  "sigma_iota_cluster",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_event",
  "sigma_eta_cluster",
  "p_TP",
  "p_FP",
  "z",
  "tau_eff"
)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000


Sys.time()

hurdle_damDen_m3_SW <- jags(data = hurdle_damDen_m3_data_list,
                   inits = hurdle_damDen_m3_inits,
                   parameters.to.save = hurdle_damDen_m3_params,
                   model.file = "HGLMM_hurdle_damDensity_m3.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hurdle_damDen_m3_SW) # model with local and watershed dam density as fixed effects
#print(hurdle_damDen_m3_SW$summary[grep("beta_", rownames(hurdle_damDen_m3_SW$summary)), ])
#print(hurdle_damDen_m3_SW$summary[grep("zeta_", rownames(hurdle_damDen_m3_SW$summary)), ])

```


----


DIAGNOSTICS:
----


```{r,echo=FALSE}
# Extract posterior draws

damDen_m3_post <- hurdle_damDen_m3_SW$sims.list 
```

**Assess "Chain health" Rhat + ESS:**

```{r,echo=FALSE} 
# Choose only the parameters you care about
params_m3 <- c(
  "epsilon",
  "alpha",
  "zeta_damDen",
  "beta_damDen",
  "sigma",
  "sigma_zeta_stream",
  "sigma_iota_cluster",
  "p_TP[1]",
  "p_FP[1]")

# Gelman–Rubin convergence diagnostic
gelman.diag(hurdle_damDen_m3_SW$samples[, params_m3], multivariate = FALSE)

# Effective sample size
effectiveSize(hurdle_damDen_m3_SW$samples[, params_m3])

```

    Rhat & Effective Sample Size:
    
    Rhat values look good. All near 1.00. Beta_damDen is slower to converge, but
    still does so nicely.
    
    Effective sample sizes look good for all parameters (note beta_damDen). 

----

**Check traceplots** for critical parameters:

```{r,echo=FALSE} 

par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


tplot_params_m3 <- c("alpha","beta_damDen","zeta_damDen", "sigma","sigma_zeta_stream","sigma_beta_stream")


mcmc_sub_m3 <- hurdle_damDen_m3_SW$samples[, tplot_params_m3]

coda::traceplot(mcmc_sub_m3)
```


    Traceplots:
    
    Traceplots all look sufficiently "grassy," indicating sufficient chain convergence


----

**Posterior predictive plots:**

Predicted Probability of SW presence (Psi) vs observed SW presence (y)

```{r,echo=FALSE} 
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

# Posterior mean psi
psi_post_m3 <- apply(hurdle_damDen_m3_SW$sims.list$psi, 2, mean)

# Compare to observed y
plot(psi_post_m3, y, pch=16, col=rgb(0,0,0,0.2))

```

----

Predicted log(SWA) vs observed log(SWA), conditional on SW presence

Goal is 1:1 (ish)

```{r,echo=FALSE} 
mu_post_m3 <- apply(hurdle_damDen_m3_SW$sims.list$mu, 2, mean)

plot(mu_post_m3[y==1], y2[y==1], pch=16, col=rgb(0,0,1,0.3))
abline(0,1,col="red")


```

----

**Random effects shrinkage: **

Goal is centered around 0(ish)

1. zeta_stream; random effect of stream on sw presence
2. beta_stream; random effect of stream on sw amount
3. zeta_event; random effect of sample event on sw presence
4. beta_event; random effect of sample event on sw amount

```{r, echo=FALSE} 
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hurdle_damDen_m3_SW$sims.list$zeta_stream)
boxplot(hurdle_damDen_m3_SW$sims.list$beta_stream)
boxplot(hurdle_damDen_m3_SW$sims.list$zeta_event)
boxplot(hurdle_damDen_m3_SW$sims.list$beta_event)
```


----

----


**Posterior mean dam effects:**

damDen --> swPres

```{r,echo=FALSE}
summary(hurdle_damDen_m3_SW$sims.list$zeta_damDen)

exp(mean(hurdle_damDen_m3_SW$sims.list$zeta_damDen))
```

    When geomorphic context is not considered, 1 std dev. increase in dam density
    corresponds to an estimated ~17% increase in the odds of surface water presence
    
    1 std dev. increase in damDensity is about +3.14 dams
    
    Massive overlap of 0. 
    
    Though slightly positive, there's no consistent global effect of a 1 std dev increase in dam density on surface water presence.

----

damDen --> swAmount

```{r,echo=FALSE}
summary(hurdle_damDen_m3_SW$sims.list$beta_damDen)

exp(mean(hurdle_damDen_m3_SW$sims.list$beta_damDen))
```

    Without geomorphic covariates, dam presence corresponds with a ~ 19% increase in surface water area. Again, lots of variation, and notable overlap of 0.


----

**Prior-posterior overlap:**

```{r,echo=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 

damDen_alpha_post_m3 <- hurdle_damDen_m3_SW$sims.list$alpha
damDen_beta_post_m3  <- hurdle_damDen_m3_SW$sims.list$beta_damDen
damDen_zeta_post_m3  <- hurdle_damDen_m3_SW$sims.list$zeta_damDen
damDen_sigBetaStream_post_m3 <- hurdle_damDen_m3_SW$sims.list$sigma_beta_stream

damDenMod_m3_n <- length(damDen_alpha_post_m3)

damDen_alpha_prior_m3 <- rnorm(damDenMod_m3_n, 0, sqrt(1/0.04))
damDen_beta_prior_m3  <- rnorm(damDenMod_m3_n, 0, sqrt(1/0.04))
damDen_zeta_prior_m3  <- rnorm(damDenMod_m3_n, 0, sqrt(1/0.04))
damDen_sigBetaStream_prior_m3 <- pmax(rnorm(damDenMod_m3_n, 0, 1), 0)


# Plot overlap:

#plot_prior_post <- function(prior, post, name) {
  
#  df <- tibble(
#    value = c(prior, post),
#    dist  = rep(c("Prior", "Posterior"), each = length(prior))
#  )
  
#  ggplot(df, aes(x = value, fill = dist)) +
#    geom_density(alpha = 0.4) +
#    scale_fill_manual(values = c("gray60", "steelblue")) +
#    labs(
#      title = paste("Prior vs Posterior for", name),
#      x = name,
#      y = "Density"
#    ) +
#    theme_minimal(base_size = 14)
#}

plot_prior_post(damDen_alpha_prior_m3, damDen_alpha_post_m3, "alpha_m3")
plot_prior_post(damDen_beta_prior_m3, damDen_beta_post_m3, "beta_damDen_m3")
plot_prior_post(damDen_zeta_prior_m3, damDen_zeta_post_m3, "zeta_damDen_m3")
plot_prior_post(damDen_sigBetaStream_prior_m3, damDen_sigBetaStream_post_m3, "sigma_beta_stream_m3")
```


    PPOs still look pretty good.

----

**Change in SWA across damDensity by stream and sampleEvent**

```{r,echo=FALSE}

m3_dam_seq <- seq(min(aoi_HurdleData$dam_km),
               max(aoi_HurdleData$dam_km),
               length.out = 50)

# build PDP grid
m3_pdp_stream <- map_dfr(1:n_stream, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m3_post$alpha
  beta_j  <- damDen_m3_post$beta_damDen + damDen_m3_post$beta_stream[, j]
  eta_j   <- damDen_m3_post$eta_stream[, j]

  map_dfr(m3_dam_seq, function(d) {

    mu_draws <- alpha_j + beta_j * d + eta_j
    swa_draws <- exp(mu_draws) - 1

    tibble(
      stream = j,
      damDen = d,
      SWA_med = median(swa_draws),
      SWA_lo  = quantile(swa_draws, 0.025),
      SWA_hi  = quantile(swa_draws, 0.975)
    )
  })
})

stream_lookup <- aoi_HurdleData %>%
  distinct(stream, watershed) %>%
  arrange(stream)

m3_pdp_stream <- m3_pdp_stream %>%
  left_join(stream_lookup, by = "stream")



# plot
ggplot(m3_pdp_stream, aes(x = watershed, y = SWA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Partial Dependence of SWA on Stream and Dam Density",
    x = "Stream",
    y = "Predicted SWA (m²)",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


###############################################################
################################################################

# build PDP grid
m3_pdp_event <- map_dfr(1:n_event, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m3_post$alpha
  beta_j  <- damDen_m3_post$beta_damDen + damDen_m3_post$beta_event[, j]
  eta_j   <- damDen_m3_post$eta_event[, j]

  map_dfr(m3_dam_seq, function(d) {

    mu_draws <- alpha_j + beta_j * d + eta_j
    swa_draws <- exp(mu_draws) - 1

    tibble(
      event = j,
      damDen = d,
      SWA_med = median(swa_draws),
      SWA_lo  = quantile(swa_draws, 0.025),
      SWA_hi  = quantile(swa_draws, 0.975)
    )
  })
})

event_year_lookup <- tibble(
  event = 1:12,
  event_label = c(
    "2019_Aug_1","2019_Sep_1","2019_Aug_2","2019_Sep_2",
    "2021_Aug_1","2021_Sep_1","2021_Aug_2","2021_Sep_2",
    "2023_Aug_1","2023_Sep_1","2023_Aug_2","2023_Sep_2"
  )
)


m3_pdp_event_named <- m3_pdp_event %>%
  left_join(event_year_lookup, by = "event")



# plot
ggplot(m3_pdp_event_named, aes(x = event_label, y = SWA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Partial Dependence of SWA on Sample Event and Dam Density",
    x = "Sample Event",
    y = "Predicted SWA (m²)",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```


----

----

----

----

----



**MODEL 4: SW ~  damDensity + Geom.Covar.**

```{r,echo=FALSE}
#library(jagsUI)
```
    sink("HGLMM_hurdle_damDensity_m4.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR SW PRESENCE (1/2 Normal on SDs):
    
      sigma_zeta_stream  ~ dnorm(0,1) T(0,)
      tau_zeta_stream    <- pow(sigma_zeta_stream, -2)
    
      sigma_zeta_year    ~ dnorm(0,1) T(0,)
      tau_zeta_year      <- pow(sigma_zeta_year, -2)
    
      sigma_zeta_event   ~ dnorm(0,1) T(0,)
      tau_zeta_event     <- pow(sigma_zeta_event, -2)
    
      sigma_iota_stream  ~ dnorm(0,1) T(0,)
      tau_iota_stream    <- pow(sigma_iota_stream, -2)
    
      sigma_iota_event   ~ dnorm(0,1) T(0,)
      tau_iota_event     <- pow(sigma_iota_event, -2)
    
      sigma_iota_cluster ~ dnorm(0,1) T(0,)
      tau_iota_cluster   <- pow(sigma_iota_cluster, -2)
    
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 4) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 4) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 4) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Presence/absence:
      epsilon ~ dnorm(0,0.04)
      zeta_damDen ~ dnorm(0,0.04)
      zeta_incis ~ dnorm(0,0.04)
      zeta_RIPA ~ dnorm(0,0.04)
      zeta_UDA ~ dnorm(0,0.04)
    
      # Amount:
      alpha ~ dnorm(0,0.04)
      beta_damDen ~ dnorm(0,0.04)
      beta_incis ~ dnorm(0,0.04)
      beta_RIPA ~ dnorm(0,0.04)
      beta_UDA ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW PRESENCE:
    
      for(j in 1:n_stream){
        zeta_stream[j] ~ dnorm(0, tau_zeta_stream)
        iota_stream[j] ~ dnorm(0, tau_iota_stream)
      }
    
      for(yr in 1:n_year){
        zeta_year[yr] ~ dnorm(0, tau_zeta_year)
      }
    
      for(k in 1:n_event){
        zeta_event[k] ~ dnorm(0, tau_zeta_event)
        iota_event[k] ~ dnorm(0, tau_iota_event)
      }
    
      for(c in 1:n_cluster){
        iota_cluster[c] ~ dnorm(0, tau_iota_cluster)
      }
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(0, tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # DETECTION PRIORS:
      
      for(k in 1:n_event){
      p_TP[k] ~ dnorm(0.76, pow(0.16,-2)) T(0,1)   # empirically derived recall
      p_FP[k] ~ dnorm(0.01, pow(0.005,-2)) T(0,1)  # empirically derived FPR
      }
      
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Presence/absence:
        z[i] ~ dbern(psi[i])
    
        logit(psi[i]) <- epsilon
                         + (zeta_damDen
                            + zeta_stream[stream[i]]
                            + zeta_year[year[i]]
                            + zeta_event[event[i]]) * damDen[i]
                         + zeta_incis * incis[i]
                         + zeta_RIPA * RIPA[i]
                         + zeta_UDA * UDA[i]
                         + iota_stream[stream[i]]
                         + iota_event[event[i]]
                         + iota_cluster[cluster[i]]
    
        # Detection:
        p_eff[i] <- z[i] * p_TP[event[i]] + (1-z[i]) * p_FP[event[i]]
        y[i]     ~ dbern(p_eff[i])
    
        # Amount:
        y2[i] ~ dnorm(mu[i], tau_eff[i])
    
        mu[i] <- z[i] *
                  (alpha
                    + (beta_damDen
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damDen[i]
                    + beta_incis * incis[i] 
                    + beta_RIPA * RIPA[i]
                    + beta_UDA * UDA[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]])
    
        tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
      }
    
    }
    ", fill = TRUE)
    
    sink()
   
    

```{r,echo=FALSE}
hurdle_damDen_m4_data_list <- list(
  N = N,
  y = y,
  y2 = y2,
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damDen = damDen,
  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)),
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hurdle_damDen_m4_inits <- function(){
  list(
    alpha        = rnorm(1, 0, 1),
    epsilon      = rnorm(1, 0, 1),

    sigma        = runif(1, 0.1, 2),
    sigma_zeta_stream  = runif(1, 0.1, 2),
    sigma_zeta_year    = runif(1, 0.1, 2),
    sigma_zeta_event   = runif(1, 0.1, 2),
    sigma_iota_stream  = runif(1, 0.1, 2),
    sigma_iota_event   = runif(1, 0.1, 2),
    sigma_iota_cluster = runif(1, 0.1, 2),

    sigma_beta_stream  = runif(1, 0.1, 2),
    sigma_beta_year    = runif(1, 0.1, 2),
    sigma_beta_event   = runif(1, 0.1, 2),
    sigma_eta_stream   = runif(1, 0.1, 2),
    sigma_eta_event    = runif(1, 0.1, 2),
    sigma_eta_cluster  = runif(1, 0.1, 2)
  )
}

hurdle_damDen_m4_params <- c(
  "psi",
  "mu",
  "epsilon",
  "alpha",
  "zeta_damDen",
  "zeta_incis",
  "zeta_RIPA",
  "zeta_UDA",
  "zeta_stream",
  "zeta_year",
  "zeta_event",
  "iota_stream",
  "iota_event",
  "iota_cluster",
  "beta_damDen",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_zeta_stream",
  "sigma_zeta_year",
  "sigma_zeta_event",
  "sigma_iota_stream",
  "sigma_iota_event",
  "sigma_iota_cluster",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_event",
  "sigma_eta_cluster",
  "p_TP",
  "p_FP",
  "z",
  "tau_eff"
)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000


Sys.time()

hurdle_damDen_m4_SW <- jags(data = hurdle_damDen_m4_data_list,
                   inits = hurdle_damDen_m4_inits,
                   parameters.to.save = hurdle_damDen_m4_params,
                   model.file = "HGLMM_hurdle_damDensity_m4.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hurdle_damPres_SW) # model with local and watershed dam density as fixed effects
#print(hurdle_damPres_SW$summary[grep("beta_", rownames(hurdle_damPres_SW$summary)), ])
#print(hurdle_damPres_SW$summary[grep("zeta_", rownames(hurdle_damPres_SW$summary)), ])

```


----


DIAGNOSTICS:
----


```{r,echo=FALSE}
# Extract posterior draws

damDen_m4_post <- hurdle_damDen_m4_SW$sims.list 
```

**Assess "Chain health" Rhat + ESS:**

```{r,echo=FALSE} 
# Choose only the parameters you care about
params_m4 <- c(
  "epsilon",
  "alpha",
  "zeta_damDen",
  "beta_damDen",
  "zeta_incis",
  "beta_incis",
  "zeta_RIPA",
  "beta_RIPA",
  "zeta_UDA",
  "beta_UDA",
  "sigma",
  "sigma_zeta_stream",
  "sigma_iota_cluster",
  "p_TP[1]",
  "p_FP[1]")

# Gelman–Rubin convergence diagnostic
gelman.diag(hurdle_damDen_m4_SW$samples[, params_m4], multivariate = FALSE)

# Effective sample size
effectiveSize(hurdle_damDen_m4_SW$samples[, params_m4])

```

    Rhat & Effective Sample Size:
    
    Solid. Rhats near 1.0 (1.03-1.05 is fine); ESS all in the thousands. 
    beta_damDen still the hardest to estimate


----

**Check traceplots** for critical parameters:

```{r,echo=FALSE} 

par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


tplot_params_m4 <- c("alpha","beta_damDen","zeta_damDen", "sigma","sigma_zeta_stream","sigma_beta_stream")


mcmc_sub_m4 <- hurdle_damDen_m4_SW$samples[, tplot_params_m4]

coda::traceplot(mcmc_sub_m4)
```


    Traceplots all look good, again see beta_damDen. 


----

**Posterior predictive plots:**

Predicted Psi vs observed SW presence

```{r,echo=FALSE} 
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

# Posterior mean psi
psi_post_m4 <- apply(hurdle_damDen_m4_SW$sims.list$psi, 2, mean)

# Compare to observed y
plot(psi_post_m4, y, pch=16, col=rgb(0,0,0,0.2))


```

----

Predicted SW vs observed SW

Goal is 1:1 (ish)

```{r,echo=FALSE} 
mu_post_m4 <- apply(hurdle_damDen_m4_SW$sims.list$mu, 2, mean)

plot(mu_post_m4[y==1], y2[y==1], pch=16, col=rgb(0,0,1,0.3))
abline(0,1,col="red")


```

----

**Random effects shrinkage: **

Goal is centered around 0(ish)

1. zeta_stream; random effect of stream on sw presence
2. beta_stream; random effect of stream on sw amount
3. zeta_event; random effect of sample event on sw presence
4. beta_event; random effect of sample event on sw amount

```{r, echo=FALSE} 
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hurdle_damDen_m4_SW$sims.list$zeta_stream)
boxplot(hurdle_damDen_m4_SW$sims.list$beta_stream)
boxplot(hurdle_damDen_m4_SW$sims.list$zeta_event)
boxplot(hurdle_damDen_m4_SW$sims.list$beta_event)
```


----

----


**Posterior mean dam effects:**

damDen --> swPres

```{r,echo=FALSE}
summary(hurdle_damDen_m4_SW$sims.list$zeta_damDen)

exp(mean(hurdle_damDen_m4_SW$sims.list$zeta_damDen))
```

    When we account for geomorphic context, a 1 std dev. increase in dam density corresponds with an estimated 3% increase in the odds of surface water presence
    
    + 1 sd of damDensity ~= + 3.14 dams/km
    
    Effect = NEGLIGIBLE

----

damDen --> swAmount

```{r,echo=FALSE}
summary(hurdle_damDen_m4_SW$sims.list$beta_damDen)

exp(mean(hurdle_damDen_m4_SW$sims.list$beta_damDen))
```

    When we account for geomorhpic context, a 1 std dev. increase in dam density corresponds with a ~ 15% increase in surface water area (conditional on SW presence).
    
    Again - notable overlap of 0 for ALL global parameter estimates, due to massive variation across landscape/time.


----

**Prior-posterior overlap:**

```{r,echo=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 

damDen_alpha_m4_post <- hurdle_damDen_m4_SW$sims.list$alpha
damDen_beta_m4_post  <- hurdle_damDen_m4_SW$sims.list$beta_damDen
damDen_zeta_m4_post  <- hurdle_damDen_m4_SW$sims.list$zeta_damDen
damDen_sigBetaStream_m4_post <- hurdle_damDen_m4_SW$sims.list$sigma_beta_stream

damDenMod_m4_n <- length(damDen_alpha_m4_post)

damDen_alpha_m4_prior <- rnorm(damDenMod_m4_n, 0, sqrt(1/0.04))
damDen_beta_m4_prior  <- rnorm(damDenMod_m4_n, 0, sqrt(1/0.04))
damDen_zeta_m4_prior  <- rnorm(damDenMod_m4_n, 0, sqrt(1/0.04))
damDen_sigBetaStream_m4_prior <- pmax(rnorm(damDenMod_m4_n, 0, 1), 0)


# Plot overlap:

#plot_prior_post <- function(prior, post, name) {
  
#  df <- tibble(
#    value = c(prior, post),
#    dist  = rep(c("Prior", "Posterior"), each = length(prior))
#  )
  
#  ggplot(df, aes(x = value, fill = dist)) +
#    geom_density(alpha = 0.4) +
#    scale_fill_manual(values = c("gray60", "steelblue")) +
#    labs(
#      title = paste("Prior vs Posterior for", name),
#      x = name,
#      y = "Density"
#    ) +
#    theme_minimal(base_size = 14)
#}

plot_prior_post(damDen_alpha_m4_prior, damDen_alpha_m4_post, "alpha_m4")
plot_prior_post(damDen_beta_m4_prior, damDen_beta_m4_post, "beta_damDen_m4")
plot_prior_post(damDen_zeta_m4_prior, damDen_zeta_m4_post, "zeta_damDen_m4")
plot_prior_post(damDen_sigBetaStream_m4_prior, damDen_sigBetaStream_m4_post, "sigma_beta_stream_m4")
```

    PPOs still look good pretty good.

----

**Marginal effect** of dam Density on surface water presence, given geomorphic context

**Note** there's a small but negligble positive effect of a 1 SD increase in DamDensity on surface water presence. However, higher dam densities do increase probability of SW presence more notably (empirical range of dam/km is up to +4 standard dev. from mean; 4*0.13 = 0.52 upwards of 52% increase in SWA at highest damDensity) 

```{r,echo=FALSE}
# Build a grid of raw geomorphic covariate values
damDen_m4_geom_grid <- tibble(
  TO_raw = seq(min(aoi_HurdleData$TO),
               max(aoi_HurdleData$TO), length.out = 50),

  incis_raw = 1 - (TO_raw / 90),   # raw incision scale

  RIPA_raw = seq(min(aoi_HurdleData$riparian_area_m2),
                 max(aoi_HurdleData$riparian_area_m2), length.out = 50),

  UDA_raw  = seq(min(aoi_HurdleData$catchment_km2),
                 max(aoi_HurdleData$catchment_km2), length.out = 50)
) %>%
  # Convert raw → standardized (to match model scale)
  mutate(
    incis = as.numeric(scale(incis_raw)),
    RIPA  = as.numeric(scale(RIPA_raw)),
    UDA   = as.numeric(scale(UDA_raw))
  )

# Make dam density sequence
damDen_m4_seq <- seq(min(aoi_HurdleData$dam_km),max(aoi_HurdleData$dam_km),length.out=10)

# Compute surface water across dam density sequence
damDen_m4_compute_psi <- function(damDen_val, geom_grid, damDen_post) {

  eps <- mean(damDen_post$epsilon)
  zd  <- mean(damDen_post$zeta_damDen)
  zi  <- mean(damDen_post$zeta_incis)
  zr  <- mean(damDen_post$zeta_RIPA)
  zu  <- mean(damDen_post$zeta_UDA)

  geom_grid %>%
    mutate(
      psi = plogis(
        eps +
        zd * damDen_val +
        zi * incis +      # standardized incision
        zr * RIPA +       # standardized RIPA
        zu * UDA          # standardized UDA
      ),
      damDen = damDen_val
    )
}


```

```{r,echo=FALSE}
damDen_m4_psi_df <- map_df(damDen_m4_seq,
                        ~ damDen_m4_compute_psi(.x,
                                             damDen_m4_geom_grid,
                                             damDen_m4_post))
```

```{r,echo=FALSE}
ggplot(damDen_m4_psi_df, aes(x = incis_raw, y = psi, color = damDen)) +
  geom_point() +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "Marginal Effect of Dam Density on Surface Water Presence",
    subtitle = "Predicted probability of water presence (psi) across raw incision values",
    x = "Channel Incision (raw scale: 1 - TO/90)",
    y = "Probability of Surface Water",
    color = "Dam Density (dams/km)"
  ) +
  theme_minimal(base_size = 14)

```

```{r,echo=FALSE}
ggplot(damDen_m4_psi_df, aes(x = RIPA_raw, y = psi, color = damDen)) +
  geom_point() +
  scale_color_viridis_c(option="plasma") +
  labs(
    title = "Marginal Effect of Dam Density on Surface Water Presence",
    subtitle = "Predicted probability of water (psi) across raw riparian area (m^2)",
    x = "Riparian Area (m^2)",
    y = "Probability of Surface Water",
    color = "Dam Density (dams/km)"
  ) +
  theme_minimal(base_size = 14)
```

```{r,echo=FALSE}
ggplot(damDen_m4_psi_df, aes(x = UDA_raw, y = psi, color = damDen)) +
  geom_point() +
  scale_color_viridis_c(option="plasma") +
  labs(
    title = "Marginal Effect of Dam Density on Surface Water Presence",
    subtitle = "Predicted probability of water (psi) across raw upstream drainage area (km^2)",
    x = "Upstream Drainage Area (km^2)",
    y = "Probability of Surface Water",
    color = "Dam Density (dams/km)"
  ) +
  theme_minimal(base_size = 14)


```

----

**Partial dependence plots**

Constrain PDP to empirical ranges for SWA and target covariates

```{r,echo=FALSE}
library(tidyverse)
library(scales)

# Raw covariates
incis_m4_raw <- 1 - aoi_HurdleData$TO/90
RIPA_m4_raw  <- aoi_HurdleData$riparian_area_m2
UDA_m4_raw   <- aoi_HurdleData$catchment_km2
#damDen <- aoi_HurdleData$dam_km_std
damDen_m4_raw <- aoi_HurdleData$dam_km
damDen_m4_mean_raw <- mean(damDen_m4_raw)
damDen_m4_sd_raw   <- sd(damDen_m4_raw)

logSWA_m4_mean <- mean(aoi_HurdleData$logSWA)
logSWA_m4_sd   <- sd(aoi_HurdleData$logSWA)

# Standardization stats
incis_m4_mean_raw <- mean(incis_m4_raw)
incis_m4_sd_raw   <- sd(incis_m4_raw)

RIPA_m4_mean_raw  <- mean(RIPA_m4_raw)
RIPA_m4_sd_raw    <- sd(RIPA_m4_raw)

UDA_m4_mean_raw   <- mean(UDA_m4_raw)
UDA_m4_sd_raw     <- sd(UDA_m4_raw)


##############################3

# Helper Functions to Compute SWA from posterior draws
compute_swa_draws_damDen <- function(incis, RIPA, UDA, post, damDen_val) {
  n_draws <- length(post$alpha)
  swa_mat <- matrix(NA_real_, nrow = length(incis), ncol = n_draws)

  for (j in seq_len(n_draws)) {
    lp <- post$alpha[j] +
          post$beta_incis[j] * incis +
          post$beta_RIPA[j]  * RIPA +
          post$beta_UDA[j]   * UDA +
          post$beta_damDen[j] * damDen_val

    swa_mat[, j] <- exp(lp) - 1
  }

  swa_mat
}

summarize_swa <- function(swa_mat) {
  tibble(
    SWA_med = apply(swa_mat, 1, median),
    SWA_lo  = apply(swa_mat, 1, quantile, probs = 0.025),
    SWA_hi  = apply(swa_mat, 1, quantile, probs = 0.975)
  )
}




#########################################

# PDP Builder:

build_pdp_damDen <- function(focal_raw, focal_name,
                      incis_raw, RIPA_raw, UDA_raw,
                      post,
                      incis_mean_raw, incis_sd_raw,
                      RIPA_mean_raw, RIPA_sd_raw,
                      UDA_mean_raw, UDA_sd_raw,
                      damDen_raw) {

  # Sequence only across empirical range
  focal_seq <- seq(min(focal_raw), max(focal_raw), length.out = 200)

  # Conditional medians of other covariates
  incis_med <- median(incis_raw)
  RIPA_med  <- median(RIPA_raw)
  UDA_med   <- median(UDA_raw)
  
  # Empirical dam density range (standardized)
  damDen_seq <- seq(min(damDen_raw), max(damDen_raw), length.out = 10)

  # Build grid (200 rows)
  grid <- tibble(
    focal_raw = focal_seq,
    incis_raw = if (focal_name == "incis") focal_seq else incis_med,
    RIPA_raw  = if (focal_name == "RIPA")  focal_seq else RIPA_med,
    UDA_raw   = if (focal_name == "UDA")   focal_seq else UDA_med
  ) %>%
    mutate(
      incis = (incis_raw - incis_mean_raw) / incis_sd_raw,
      RIPA  = (RIPA_raw  - RIPA_mean_raw)  / RIPA_sd_raw,
      UDA   = (UDA_raw   - UDA_mean_raw)   / UDA_sd_raw
    )
  # Compute PDP curves for each dam density value
  pdp_list <- map(damDen_seq, function(dd) {
    swa_mat <- compute_swa_draws_damDen(grid$incis, grid$RIPA, grid$UDA, post,
                                        (dd - damDen_m4_mean_raw) / damDen_m4_sd_raw)
    summarize_swa(swa_mat) %>% mutate(damDen = dd)
  })

  pdp <- bind_rows(pdp_list)

  # Repeat grid for each damDen value
  grid2 <- bind_rows(replicate(length(damDen_seq), grid, simplify = FALSE))

  bind_cols(grid2, pdp)
}



#####

pdp_incis_m4_damDen <- build_pdp_damDen(
  focal_raw = incis_m4_raw,
  focal_name = "incis",
  incis_m4_raw, RIPA_m4_raw, UDA_m4_raw,
  damDen_m4_post,
  incis_m4_mean_raw, incis_m4_sd_raw,
  RIPA_m4_mean_raw, RIPA_m4_sd_raw,
  UDA_m4_mean_raw, UDA_m4_sd_raw, damDen_m4_raw
)

pdp_RIPA_m4_damDen <- build_pdp_damDen(
  focal_raw = RIPA_m4_raw,
  focal_name = "RIPA",
  incis_m4_raw, RIPA_m4_raw, UDA_m4_raw,
  damDen_m4_post,
  incis_m4_mean_raw, incis_m4_sd_raw,
  RIPA_m4_mean_raw, RIPA_m4_sd_raw,
  UDA_m4_mean_raw, UDA_m4_sd_raw, damDen_m4_raw
)

pdp_UDA_m4_damDen <- build_pdp_damDen(
  focal_raw = UDA_m4_raw,
  focal_name = "UDA",
  incis_raw, RIPA_m4_raw, UDA_m4_raw,
  damDen_m4_post,
  incis_m4_mean_raw, incis_m4_sd_raw,
  RIPA_m4_mean_raw, RIPA_m4_sd_raw,
  UDA_m4_mean_raw, UDA_m4_sd_raw, damDen_m4_raw
)



##########################################3

# Plot Function:

plot_pdp_damDen <- function(pdp, xvar, xtitle) {
  ggplot(pdp,
         aes(x = .data[[xvar]], y = SWA_med,
             color = damDen, fill = damDen)) +
    #geom_ribbon(aes(ymin = SWA_lo, ymax = SWA_hi), alpha = 0.15, color = NA) +
    geom_point() +
    scale_color_viridis_c(option = "plasma") +
    scale_fill_viridis_c(option = "plasma") +
    scale_y_continuous(labels = comma) +
    labs(
      title = paste("Partial Dependence of SWA on dam density and", xtitle),
      subtitle = "Across empirical dam density range",
      x = xtitle,
      y = "Predicted SWA (m^2)",
      color = "Dam Density (dams/km)",
      fill  = "Dam Density (dams/km)"
    ) +
    theme_minimal(base_size = 14)
}



###########################################3


plot_pdp_damDen(pdp_incis_m4_damDen, "incis_raw", "Channel Incision (raw)")
plot_pdp_damDen(pdp_RIPA_m4_damDen,  "RIPA_raw",  "Riparian Area (m²)")
plot_pdp_damDen(pdp_UDA_m4_damDen,   "UDA_raw",   "Contributing Drainage Area (km²)")


```

----

**Change in SWA across damDensity by stream and sampleEvent**

```{r,echo=FALSE}

m4_dam_seq <- seq(min(aoi_HurdleData$dam_km),
               max(aoi_HurdleData$dam_km),
               length.out = 50)

# build PDP grid
m4_pdp_stream <- map_dfr(1:n_stream, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m4_post$alpha
  beta_j  <- damDen_m4_post$beta_damDen + damDen_m4_post$beta_stream[, j]
  eta_j   <- damDen_m4_post$eta_stream[, j]

  map_dfr(m4_dam_seq, function(d) {

    mu_draws <- alpha_j + beta_j * d + eta_j
    swa_draws <- exp(mu_draws) - 1

    tibble(
      stream = j,
      damDen = d,
      SWA_med = median(swa_draws),
      SWA_lo  = quantile(swa_draws, 0.025),
      SWA_hi  = quantile(swa_draws, 0.975)
    )
  })
})

stream_lookup <- aoi_HurdleData %>%
  distinct(stream, watershed) %>%
  arrange(stream)

m4_pdp_stream <- m4_pdp_stream %>%
  left_join(stream_lookup, by = "stream")



# plot
ggplot(m4_pdp_stream, aes(x = watershed, y = SWA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "Partial Dependence of SWA on Stream and Dam Density",
    x = "Stream",
    y = "Predicted SWA (m²)",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


###############################################################
################################################################

# build PDP grid
m4_pdp_event <- map_dfr(1:n_event, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m4_post$alpha
  beta_j  <- damDen_m4_post$beta_damDen + damDen_m4_post$beta_event[, j]
  eta_j   <- damDen_m4_post$eta_event[, j]

  map_dfr(m4_dam_seq, function(d) {

    mu_draws <- alpha_j + beta_j * d + eta_j
    swa_draws <- exp(mu_draws) - 1

    tibble(
      event = j,
      damDen = d,
      SWA_med = median(swa_draws),
      SWA_lo  = quantile(swa_draws, 0.025),
      SWA_hi  = quantile(swa_draws, 0.975)
    )
  })
})

event_year_lookup <- tibble(
  event = 1:12,
  event_label = c(
    "2019_Aug_1","2019_Sep_1","2019_Aug_2","2019_Sep_2",
    "2021_Aug_1","2021_Sep_1","2021_Aug_2","2021_Sep_2",
    "2023_Aug_1","2023_Sep_1","2023_Aug_2","2023_Sep_2"
  )
)


m4_pdp_event_named <- m4_pdp_event %>%
  left_join(event_year_lookup, by = "event")



# plot
ggplot(m4_pdp_event_named, aes(x = event_label, y = SWA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "Partial Dependence of SWA on Sample Event and Dam Density",
    x = "Sample Event",
    y = "Predicted SWA (m²)",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```