---
title: "NDVI_Models"
author: "Matt Webster"
date: "2026-07-28"
output: html_document
editor_options: 
  chunk_output_type: console
---


The purpose of this RMarkdown is to implement hierarchical generalized linear mixed effects models for NDVI ~ damPresence, & NDVI ~ damDensity (with and without geomorphic covariates). Structurally parallel to VegArea ~ dam models.

4 models total.
    
    
    WAIC indicates that all four models perform comparably well. Given this, I only include code and results for m4, which models the effect of dam density on NDVI, conditional on geomorphic covariates.

Apply "Model 4" to veg data at each buffer size; 0,10,20,30-m. 1 model for each buffer size. Then compare model outputs for each. How do posterior estimates change across buffers? 


----

**HGLMM**


    This analysis explores late August vegetation greenness (NDVI) in relation to 
    beaver dam presence, & density.
    
    2 models for dam presence (1 without geomorphic covariates; 1 with geomorphic covariates)
    2 models for dam density (...^^^...)
    
    All 4 models presented herein are constructed in a hierarchical generalized linear 
    mixed effects model framework with stream, year, and cluster_ID as random intercepts
    for NDVI. 
    
      - NDVI ~ damPresence (no geom.covar.)
      - NDVI ~ damPresence (geom.covar.)
      - NDVI ~ damDensity (no geom.covar.)
      - NDVI ~ damDensity (no geom.covar.)
    
    Model structure is paralleled with HGLMMs for veg area.
    
    Uses mean(NDVI) for each sample window for 6 watersheds, 3 years, 08/15-08/31
    
    
----

```{r,echo=FALSE,}
suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(ggplot2)
  library(dataRetrieval)
  library(knitr)
  library(ggplot2)
  library(zoo) #for moving window
  library(stringr)
  library(vioplot)
  library(purrr)
  library(lavaan)
  library(blavaan)
  library(ggcorrplot)
  library(lme4)
  library(broom.mixed)
  library(coda)
  library(scales)
  library(loo)
  library(jagsUI)
})
```




**Final Data:**

NOTE: This ONLY uses data from 08/15-08/31. This is to avoid issues with pseudo-replication after the omission of random effects for sample event.

Modeling reflects LATE August effect of beaver dams on NDVI. Green riparian 
vegetation = mesic resource that is critical for grouse recruitment.


```{r,echo=FALSE}
aoi_HurdleData <- read.csv("aoi_HurdleData_091026Update.csv") %>% 
  filter(t == 6) # LATE August


# add year/event index
event_year_3_4 <- tapply(aoi_HurdleData$year_id,
                         aoi_HurdleData$sampleEvent_new,
                         unique)
event_year_3_4 <- as.numeric(event_year_3_4)


# Full Hurdle data
N <- nrow(aoi_HurdleData)
#y <- aoi_HurdleData$y
#y2 <- aoi_HurdleData$logSWA
#y_veg <- aoi_HurdleData$y_veg
#y_vegA <- aoi_HurdleData$VA
y_NDVI_0 <- aoi_HurdleData$NDVI_0

y_NDVI_10 <- aoi_HurdleData$NDVI_10

y_NDVI_20 <- aoi_HurdleData$NDVI_20

y_NDVI_30 <- aoi_HurdleData$NDVI_30

stream <- aoi_HurdleData$stream
year <- aoi_HurdleData$year_id
event <- aoi_HurdleData$sampleEvent_new
cluster <- as.numeric(as.factor(aoi_HurdleData$clusterID_unique))
damPres <- aoi_HurdleData$damPresence
damDen <- as.numeric(aoi_HurdleData$dam_km_std)
damDen_raw <- aoi_HurdleData$dam_km

n_stream <- length(unique(aoi_HurdleData$stream))
n_year <- length(unique(aoi_HurdleData$year_id))
n_event <- length(unique(aoi_HurdleData$sampleEvent_new))
n_cluster <- length(unique(as.numeric(as.factor(aoi_HurdleData$clusterID_unique))))

str(aoi_HurdleData)
```

----

    Rough normality of NDVI distribution does not require response variable transformation prior
    to modeling. 
      
      

```{r,echo=FALSE}
par(mfrow = c(1,4),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 

hist(aoi_HurdleData$NDVI_0)
hist(aoi_HurdleData$NDVI_10)
hist(aoi_HurdleData$NDVI_20)
hist(aoi_HurdleData$NDVI_30)

```


----


For NDVI models, data of interest include:

     - NDVI_BUFF: average NDVI for sample window[i] within x-m planar buffer (0,10,20,30); response variable
    
    - stream: index for random intercept on NDVI
    
    - year: index for random intercept on NDVI
    
    - clusterID_seq: index for random intercept on NDVI; represents sample pairing
    
    - damPres: dam presence (0,1); fixed effect on NDVI
    
    - damDen: dams/km; z-standardized ("...$dam_km_std"); fixed effect on NDVI
    
    Only used in models 2 and 4:
    
    - incis: channel incision: 1-(TO/90) (inverse of topographic openness); z-standardized;
      fixed effect on NDVI
    
    - RIPA: riparian area (m^2); z-standardized; fixed effect on NDVI 
    
    - UDA: upstream drainage area (km ^2); z-standardized; fixed effect on NDVI
    
----

**Overall Takeaways**
    
    When we simplify the model, we see that:
    
    Dams have a slightly positive, though MINISCULE effect on NDVI. This makes good sense, as
    dam presence/density is unlikely to drive massive changes in riparian vegetation structure.
    It may make yellow willows slightly greener, but it's not changing scrub/grass into lush
    willow/cottonwood forests. 
    
    Dams increase NDVI by ~ 0.01
    
    Drainage area is the strongest positive driver of NDVI.
    
    Riparian area has a slight negative effect on NDVI (more space includes more non-green veg).
    
    Incision has a slight (significant) negative effect on NDVI.

----

----

----

NDVI ~ Dam Presence
----

----

----

**Model 1: NDVI ~ DamPresence**

Effect of damPresence without geomorphic covariates

----

```{r,echo=FALSE}
#library(jagsUI)
```

    sink("HGLMM_m1_NDVI_damPres.jags")
    
    cat("
    model{
      
      # HYPERPRIORS FOR NDVI (1/2 Normal on SDs):
    
      sigma ~ dnorm(0,1) T(0,)
      tau <- pow(sigma, -2)
      
      sigma_eta_stream ~ dnorm(0,1) T(0,)
      tau_eta_stream <- pow(sigma_eta_stream,-2)
      
      sigma_eta_year ~ dnorm(0,1) T(0,)
      tau_eta_year <- pow(sigma_eta_year,-2)
      
      sigma_eta_cluster ~ dnorm(0,1) T(0,)
      tau_eta_cluster <- pow(sigma_eta_cluster,-2)
      
      ##########################################################################
      # RANDOM EFFECT PRIORS:
    
      for(j in 1:n_stream){
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
      
      for(j in 1:n_year){
        eta_year[j] ~ dnorm(0, tau_eta_year)
      }
      
      for(j in 1:n_cluster){
        eta_cluster[j] ~ dnorm(0, tau_eta_cluster)
      }
      
      ##########################################################################
      # FIXED EFFECT PRIORS:
      
      alpha ~ dnorm(0.32,0.04)          # within 5 SD of empirical mean NDVI
      beta_damPres ~ dnorm(0,0.04)
    
      ##########################################################################
      
      # MAIN LOOP: NDVI
    
      for(i in 1:N){
    
        # Amount:
        y[i] ~ dnorm(mu[i], tau)
    
        mu[i] <- alpha
                    + eta_stream[stream[i]]
                    + eta_year[year[i]]
                    + eta_cluster[cluster[i]]
                    + beta_damPres * damPres[i]
                    
      ##########################################################################
                    
      # Derived parameter for WAIC computation:

      loglik[i] <- logdensity.norm(y[i], mu[i], tau)
      
      }
    
    }
    ", fill = TRUE)
    
    sink()

    

```{r,echo=FALSE}
hglmm_m1_NDVI_damPres_data_list <- list(
  N = N,
  y = y_NDVI,
  damPres = damPres,
  stream = stream,
  n_stream = n_stream,
  year = year,
  n_year = n_year,
  cluster = cluster,
  n_cluster = n_cluster
  )

hglmm_m1_NDVI_damPres_inits <- function(){
  list()
}

hglmm_m1_NDVI_damPres_params <- c(
  "mu",
  "alpha",
  "beta_damPres",
  "eta_stream",
  "eta_year",
  "eta_cluster",
  "sigma",
  "sigma_eta_stream",
  "sigma_eta_year",
  "sigma_eta_cluster",
  "loglik"
)

nchains <- 3
nthin <- 5
niter <- 100000
nburn <- 25000

Sys.time()

set.seed(12345)

hglmm_m1_NDVI_damPres <- jags(data = hglmm_m1_NDVI_damPres_data_list,
                   inits = hglmm_m1_NDVI_damPres_inits,
                   parameters.to.save = hglmm_m1_NDVI_damPres_params,
                   model.file = "HGLMM_m1_NDVI_damPres.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hglmm_m1_NDVI_damPres) # model with local and watershed dam density as fixed effects
#print(hglmm_m1_NDVI_damPres$summary[grep("beta_", rownames(hglmm_m1_NDVI_damPres$summary)), ])

```

<!---
----

DIAGNOSTICS:

Extract posteriors:
```{r,echo=FALSE}
damPres_m1_NDVI_post <- hglmm_m1_NDVI_damPres$sims.list
```

"Chain health" Rhat + ESS:
```{r,echo=FALSE}
# Choose only the parameters you care about
params_m1_NDVI <- c(
  "alpha",
  "beta_damPres",
  "sigma",
  "sigma_eta_stream")

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m1_NDVI_damPres$samples[, params_m1_NDVI], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m1_NDVI_damPres$samples[, params_m1_NDVI])

```

----

    Rhat = 1 & large ESS; looks good for all parameters

----

**Check traceplots** for critical parameters:

```{r,echo=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

tplot_m1_NDVI_params <- c("alpha","beta_damPres","sigma", "sigma_eta_stream")


m1_mcmc_NDVI_sub <- hglmm_m1_NDVI_damPres$samples[, tplot_m1_NDVI_params]

coda::traceplot(m1_mcmc_NDVI_sub)
```

----

    Traceplots all look sufficiently grassy, indicating chain convergence.

----

**Posterior predictive plots:**

Model-estimated NDVI vs. observed NDVI

Should be about 1

```{r,echo=FALSE}
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

# Posterior mean of predicted log(NDVI + 1)
m1_mu_NDVI_post <- apply(hglmm_m1_NDVI_damPres$sims.list$mu, 2, mean)

# Observed log vegetation area
m1_obs_NDVI <- y_NDVI

# Fit regression on the log scale
m1_fit <- lm(m1_obs_NDVI ~ m1_mu_NDVI_post)

m1_slope <- coef(m1_fit)[2]
m1_R2   <- summary(m1_fit)$r.squared

# Plot
plot(m1_mu_NDVI_post,
     m1_obs_NDVI,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated NDVI",
     ylab = "Observed NDVI")

abline(0, 1, col = "red", lwd = 2)
```

slope =

```{r,echo=FALSE}
m1_slope
```

----

    Comparison of model-estimated NDVI with observed NDVI. "Perfect" model fit would
    have a slope = 1. A slope of 1.08 is reasonable, given the model. This indicates
    that our model for NDVI is performing fairly well, though slightly underestimating
    the variation in NDVI compared to the raw data.

----

**Random effects shrinkage: **

Should be centered around 0(ish)

1. eta_stream; random effect of stream on NDVI intercept (1:6)
2. eta_year; random effect of year on NDVI intercept (1:3)

```{r, echo=FALSE}

par(mfrow = c(2, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hglmm_m1_NDVI_damPres$sims.list$eta_stream)

boxplot(hglmm_m1_NDVI_damPres$sims.list$eta_year)
```

----

**Posterior mean dam effects:**

NDVI ~ damPres (no geom.covar.)

```{r,echo=FALSE}
m1_coeffparams <- c(#"alpha",
            "beta_damPres",
            "sigma")

m1_coef_df <- hglmm_m1_NDVI_damPres$summary[m1_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m1_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.2,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE}
summary(hglmm_m1_NDVI_damPres$sims.list$beta_damPres)

exp(mean(hglmm_m1_NDVI_damPres$sims.list$beta_damPres))
```


```{r,echo=FALSE}
#library(tidyverse)
#library(scales)


#Posterior distribution of damEffect on vegArea:

hist(hglmm_m1_NDVI_damPres$sims.list$beta_damPres,
     breaks = 50,
     col = "steelblue",
     main = "Posterior of damPresence effect on NDVI ",
     xlab = "beta_damPres")
abline(v = mean(hglmm_m1_NDVI_damPres$sims.list$beta_damPres), col = "red", lwd = 2)
```
    
    Without consideration of geomorphic covariates, dam presence is associated with a
    ~0.011 increase in NDVI. No overlap of zero. Confidently, a very meager effect.

----

**Prior-posterior overlap:**

```{r,echo=FALSE}
damPres_m1_NDVI_alpha_post <- hglmm_m1_NDVI_damPres$sims.list$alpha
damPres_m1_NDVI_beta_post  <- hglmm_m1_NDVI_damPres$sims.list$beta_damPres

damPres_m1_NDVI_Mod_n <- length(damPres_m1_NDVI_alpha_post)

damPres_m1_NDVI_alpha_prior <- rnorm(damPres_m1_NDVI_Mod_n, 0.32, sqrt(1/0.04))
damPres_m1_NDVI_beta_prior  <- rnorm(damPres_m1_NDVI_Mod_n, 0, sqrt(1/0.04))


# Plot overlap:

plot_prior_post <- function(prior, post, name) {
  
  df <- tibble(
    value = c(prior, post),
    dist  = rep(c("Prior", "Posterior"), each = length(prior))
  )
  
  ggplot(df, aes(x = value, fill = dist)) +
    geom_density(alpha = 0.4) +
    scale_fill_manual(values = c("gray60", "steelblue")) +
    labs(
      title = paste("Prior vs Posterior for", name),
      x = name,
      y = "Density"
    ) +
    theme_minimal(base_size = 14) +
    xlim(-1,1)
}

plot_prior_post(damPres_m1_NDVI_alpha_prior, damPres_m1_NDVI_alpha_post, "alpha")
plot_prior_post(damPres_m1_NDVI_beta_prior, damPres_m1_NDVI_beta_post, "beta_damPres")

```

----

    PPO indicates that the model is not being driven by prior specifications

----


**Predicted NDVI with and without Dam Presence**


```{r,echo=FALSE}

m1_n_draws <- length(damPres_m1_NDVI_post$alpha)

# Random effects averaged out
m1_mu_no_dam <- damPres_m1_NDVI_post$alpha +
             damPres_m1_NDVI_post$eta_stream[ , stream] +
             damPres_m1_NDVI_post$eta_year[, year] +
             damPres_m1_NDVI_post$eta_cluster[ , cluster]

m1_mu_yes_dam <- damPres_m1_NDVI_post$alpha +
              damPres_m1_NDVI_post$beta_damPres +
              damPres_m1_NDVI_post$eta_stream[ , stream] +
              damPres_m1_NDVI_post$eta_year[, year] +
              damPres_m1_NDVI_post$eta_cluster[ , cluster]


m1_NDVI_no_dam_vec  <- as.numeric(m1_mu_no_dam)
m1_NDVI_yes_dam_vec <- as.numeric(m1_mu_yes_dam)


m1_NDVI_no_dam_med  <- apply(m1_mu_no_dam, 2, median)
m1_NDVI_yes_dam_med <- apply(m1_mu_yes_dam, 2, median)

m1_NDVI_df <- tibble(
  NDVI = c(m1_NDVI_no_dam_med, m1_NDVI_yes_dam_med),
  Dam = factor(rep(c("No Dam", "Dam Present"),
                   each = length(m1_NDVI_no_dam_med)))
)


ggplot(m1_NDVI_df, aes(x = Dam, y = NDVI, fill = Dam)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8) +
  scale_fill_manual(values = c("steelblue", "brown4")) +
  labs(
    title = "Predicted NDVI by Dam Presence",
    y = "NDVI",
    x = ""
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")
```


--->

----

----

----

**M2: NDVI ~ damPres + Geom.Covar**

Effect of damPresence WITH geomorphic covariates

----

```{r,echo=FALSE}
#library(jagsUI)
```
    sink("HGLMM_m2_NDVI_damPres.jags")
    
    cat("
    model{
      
      # HYPERPRIORS FOR NDVI (1/2 Normal on SDs):
    
      sigma ~ dnorm(0,1) T(0,)
      tau <- pow(sigma, -2)
      
      sigma_eta_stream ~ dnorm(0,1) T(0,)
      tau_eta_stream <- pow(sigma_eta_stream,-2)
      
      sigma_eta_year ~ dnorm(0,1) T(0,)
      tau_eta_year <- pow(sigma_eta_year,-2)
      
      sigma_eta_cluster ~ dnorm(0,1) T(0,)
      tau_eta_cluster <- pow(sigma_eta_cluster,-2)
      
      ##########################################################################
      # RANDOM EFFECT PRIORS:
    
      for(j in 1:n_stream){
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
      
      for(j in 1:n_year){
        eta_year[j] ~ dnorm(0, tau_eta_year)
      }
      
      for(j in 1:n_cluster){
        eta_cluster[j] ~ dnorm(0, tau_eta_cluster)
      }
      
      ##########################################################################
      # FIXED EFFECT PRIORS:
      
      alpha ~ dnorm(0.32, 0.04)          # within 5 SD of empirical mean NDVI
      beta_damPres ~ dnorm(0,0.04)
      beta_incis ~ dnorm(0,0.04)
      beta_RIPA ~ dnorm(0,0.04)
      beta_UDA ~ dnorm(0,0.04)
    
      ##########################################################################
      
      # MAIN LOOP: NDVI
    
      for(i in 1:N){
    
        # Amount:
        y[i] ~ dnorm(mu[i], tau)
    
        mu[i] <- alpha
                    + eta_stream[stream[i]]
                    + eta_year[year[i]]
                    + eta_cluster[cluster[i]]
                    + beta_damPres * damPres[i]
                    + beta_incis * incis[i]
                    + beta_RIPA * RIPA[i]
                    + beta_UDA * UDA[i]
                    
      ##########################################################################
                    
      # Derived parameter for WAIC computation:

      loglik[i] <- logdensity.norm(y[i], mu[i], tau)
      }
    
    }
    ", fill = TRUE)
    
    sink()

    

```{r,echo=FALSE}
hglmm_m2_NDVI_damPres_data_list <- list(
  N = N,
  y = y_NDVI,
  stream = stream,
  year = year,
  cluster = cluster,
  damPres = damPres,
  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)),
  n_stream = n_stream,
  n_year = n_year,
  n_cluster = n_cluster
)


hglmm_m2_NDVI_damPres_inits <- function(){
  list()
}

hglmm_m2_NDVI_damPres_params <- c(
  "mu",
  "alpha",
  "beta_damPres",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "eta_stream",
  "eta_year",
  "eta_cluster",
  "sigma",
  "sigma_eta_stream",
  "sigma_eta_year",
  "sigma_eta_cluster",
  "loglik"
)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000

Sys.time()

set.seed(12345)

hglmm_m2_NDVI_damPres <- jags(data = hglmm_m2_NDVI_damPres_data_list,
                   inits = hglmm_m2_NDVI_damPres_inits,
                   parameters.to.save = hglmm_m2_NDVI_damPres_params,
                   model.file = "HGLMM_m2_NDVI_damPres.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

```

<!---

----

DIAGNOSTICS:

Extract posteriors:
```{r,echo=FALSE}
damPres_m2_NDVI_post <- hglmm_m2_NDVI_damPres$sims.list
```

"Chain health" Rhat + ESS:
```{r,echo=FALSE}
# Choose only the parameters you care about
params_m2_NDVI <- c(
  "alpha",
  "beta_damPres",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "sigma")

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m2_NDVI_damPres$samples[, params_m2_NDVI], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m2_NDVI_damPres$samples[, params_m2_NDVI])

```

----

    Rhats = 1
    
    ESS is high (good) for all parameters
    
----

**Check traceplots** for critical parameters:

```{r,echo=FALSE}
par(mfrow = c(2, 3),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

tplot_m2_NDVI_params <- c("alpha","beta_damPres",
                         "beta_incis","beta_RIPA",
                         "beta_UDA", "sigma")


m2_mcmc_NDVI_sub <- hglmm_m2_NDVI_damPres$samples[, tplot_m2_NDVI_params]

coda::traceplot(m2_mcmc_NDVI_sub)
```

----

    Traceplots are sufficiently "grassy"
    
----

**Posterior predictive plots:**

Model-estimated vegArea vs. observed vegArea

Should be about 1

```{r,echo=FALSE}
par(mfrow = c(1,1),        # 1 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

# Posterior mean of predicted NDVI 
m2_mu_NDVI_post <- apply(hglmm_m2_NDVI_damPres$sims.list$mu, 2, mean)

# Observed log vegetation area
m2_obs_NDVI <- y_NDVI

# Fit regression on the log scale
m2_fit <- lm(m2_obs_NDVI ~ m2_mu_NDVI_post)

m2_slope <- coef(m2_fit)[2]
m2_R2    <- summary(m2_fit)$r.squared

# Plot
plot(m2_mu_NDVI_post,
     m2_obs_NDVI,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated NDVI",
     ylab = "Observed NDVI")

abline(0, 1, col = "red", lwd = 2)

```

Slope:
```{r,echo=FALSE}
m2_slope
```
----

    Comparison of model-estimated log-transformed vegetation area with observed log-
    transformed vegetation area. "Perfect" model fit would have a slope = 1. A slope
    of 1.07 is not bad.

----

**Random effects shrinkage: **

Should be centered around 0(ish)

1. eta_stream; random effect of stream on NDVI intercept (1:6)
2. eta_year; random effect of year on NDVI intercept (1:3)

```{r, echo=FALSE}

par(mfrow = c(2, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hglmm_m2_NDVI_damPres$sims.list$eta_stream)

boxplot(hglmm_m2_NDVI_damPres$sims.list$eta_year)
```

----

**Posterior mean dam effects:**

damPres -> NDVI (w/geom.covar.)


```{r,echo=FALSE}
m2_coeffparams <- c(#"alpha",
            "beta_damPres",
            "beta_incis",
            "beta_RIPA",
            "beta_UDA",
            "sigma")

m2_coef_df <- hglmm_m2_NDVI_damPres$summary[m2_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m2_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.2,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    x = "Effect Size",
    y = ""
  ) +
  theme_bw(base_size = 14)
```


```{r,echo=FALSE}
summary(hglmm_m2_NDVI_damPres$sims.list$beta_damPres)

exp(mean(hglmm_m2_NDVI_damPres$sims.list$beta_damPres))
```


```{r,echo=FALSE}
#library(tidyverse)
#library(scales)

par(mfrow = c(1, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  
#Posterior distribution of damEffect on vegArea:

hist(hglmm_m2_NDVI_damPres$sims.list$beta_damPres,
     breaks = 50,
     col = "steelblue",
     main = "Posterior distribution of damPres effect on NDVI",
     xlab = "beta_damPres",
     xlim = c(0,0.025))
# Add median line
abline(v = median(hglmm_m2_NDVI_damPres$sims.list$beta_damPres), col = "red", lwd = 2)

```


    When you also consider site-level geomorphic conditions, dam presence is STILL associated 
    with an estimated 0.01 increase in NDVI.


----

Prior-posterior overlap:

```{r,echo=FALSE}
damPres_m2_NDVI_alpha_post <- hglmm_m2_NDVI_damPres$sims.list$alpha
damPres_m2_NDVI_betaDP_post  <- hglmm_m2_NDVI_damPres$sims.list$beta_damPres

damPres_m2_NDVI_Mod_n <- length(damPres_m2_NDVI_alpha_post)

damPres_m2_NDVI_alpha_prior <- rnorm(damPres_m2_NDVI_Mod_n, 0.32, sqrt(1/0.04))
damPres_m2_NDVI_betaDP_prior  <- rnorm(damPres_m2_NDVI_Mod_n, 0, sqrt(1/0.04))


# Plot overlap:

#plot_prior_post <- function(prior, post, name) {
  
#  df <- tibble(
#    value = c(prior, post),
#    dist  = rep(c("Prior", "Posterior"), each = length(prior))
#  )
  
#  ggplot(df, aes(x = value, fill = dist)) +
#    geom_density(alpha = 0.4) +
#    scale_fill_manual(values = c("gray60", "steelblue")) +
#    labs(
#      title = paste("Prior vs Posterior for", name),
#      x = name,
#      y = "Density"
#    ) +
#    theme_minimal(base_size = 14)
#}

plot_prior_post(damPres_m2_NDVI_alpha_prior, damPres_m2_NDVI_alpha_post, "alpha")
plot_prior_post(damPres_m2_NDVI_betaDP_prior, damPres_m2_NDVI_betaDP_post, "beta_damPres")

```

----

    PPOs still look pretty good. 

----

**Predicted NDVI with and without Dam Presence**


```{r,echo=FALSE}

m2_n_draws <- length(damPres_m2_NDVI_post$alpha)

# Random effects averaged out
m2_mu_no_dam <- damPres_m2_NDVI_post$alpha +
             damPres_m2_NDVI_post$eta_stream[ , stream] +
             damPres_m2_NDVI_post$eta_year[ , year] +
             damPres_m2_NDVI_post$eta_cluster[ , cluster] +
             damPres_m2_NDVI_post$beta_incis +
             damPres_m2_NDVI_post$beta_RIPA +
             damPres_m2_NDVI_post$beta_UDA

m2_mu_yes_dam <- damPres_m2_NDVI_post$alpha +
              damPres_m2_NDVI_post$beta_damPres +
              damPres_m2_NDVI_post$eta_stream[ , stream] +
              damPres_m2_NDVI_post$eta_year[ , year] +
              damPres_m2_NDVI_post$eta_cluster[ , cluster] +
              damPres_m2_NDVI_post$beta_incis +
              damPres_m2_NDVI_post$beta_RIPA +
              damPres_m2_NDVI_post$beta_UDA

m2_NDVI_no_dam_vec  <- as.numeric(m2_mu_no_dam)
m2_NDVI_yes_dam_vec <- as.numeric(m2_mu_yes_dam)


m2_NDVI_no_dam_med  <- apply(m2_mu_no_dam, 2, median)
m2_NDVI_yes_dam_med <- apply(m2_mu_yes_dam, 2, median)

m2_NDVI_df <- tibble(
  NDVI = c(m2_NDVI_no_dam_med, m2_NDVI_yes_dam_med),
  Dam = factor(rep(c("No Dam", "Dam Present"),
                   each = length(m2_NDVI_no_dam_med)))
)


ggplot(m2_NDVI_df, aes(x = Dam, y = NDVI, fill = Dam)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, alpha = 0.8) +
  scale_fill_manual(values = c("steelblue", "brown4")) +
  labs(
    title = "Predicted NDVI by Dam Presence",
    y = "NDVI",
    x = ""
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")
```

    Pretty much no difference...

----

----

----

----


**Partial dependence plots**

Constrain PDP to empirical ranges for NDVI and target covariates

```{r,echo=FALSE}

names(damPres_m2_NDVI_post)

# Raw covariates
incis_raw <- 1 - aoi_HurdleData$TO/90
RIPA_raw  <- aoi_HurdleData$riparian_area_m2
UDA_raw   <- aoi_HurdleData$catchment_km2

NDVI_mean <- mean(aoi_HurdleData$NDVI)
NDVI_sd   <- sd(aoi_HurdleData$NDVI)

# Standardization stats
incis_mean_raw <- mean(incis_raw)
incis_sd_raw   <- sd(incis_raw)

RIPA_mean_raw  <- mean(RIPA_raw)
RIPA_sd_raw    <- sd(RIPA_raw)

UDA_mean_raw   <- mean(UDA_raw)
UDA_sd_raw     <- sd(UDA_raw)


##############################3

# Helper Functions to Compute NDVI from posterior draws
compute_NDVI_draws <- function(incis, RIPA, UDA, post, dam_flag) {
  n_draws <- length(damPres_m2_NDVI_post$alpha)
  NDVI_mat <- matrix(NA_real_, nrow = length(incis), ncol = n_draws)

  for (j in seq_len(n_draws)) {
    lp <- post$alpha[j] +
          post$beta_incis[j] * incis +
          post$beta_RIPA[j]  * RIPA +
          post$beta_UDA[j]   * UDA +
          post$beta_damPres[j] * dam_flag

    NDVI_mat[, j] <- lp
  }

  NDVI_mat
}

summarize_NDVI <- function(NDVI_mat) {
  tibble(
    NDVI_med = apply(NDVI_mat, 1, median),
    NDVI_lo  = apply(NDVI_mat, 1, quantile, probs = 0.025),
    NDVI_hi  = apply(NDVI_mat, 1, quantile, probs = 0.975)
  )
}




#########################################

# PDP Builder:

build_pdp <- function(focal_raw, focal_name,
                      incis_raw, RIPA_raw, UDA_raw,
                      post,
                      incis_mean_raw, incis_sd_raw,
                      RIPA_mean_raw, RIPA_sd_raw,
                      UDA_mean_raw, UDA_sd_raw) {

  # Sequence only across empirical range
  focal_seq <- seq(min(focal_raw), max(focal_raw), length.out = 200)

  # Conditional medians of other covariates
  incis_med <- median(incis_raw)
  RIPA_med  <- median(RIPA_raw)
  UDA_med   <- median(UDA_raw)

  # Build grid (200 rows)
  grid <- tibble(
    focal_raw = focal_seq,
    incis_raw = if (focal_name == "incis") focal_seq else incis_med,
    RIPA_raw  = if (focal_name == "RIPA")  focal_seq else RIPA_med,
    UDA_raw   = if (focal_name == "UDA")   focal_seq else UDA_med
  ) %>%
    mutate(
      incis = (incis_raw - incis_mean_raw) / incis_sd_raw,
      RIPA  = (RIPA_raw  - RIPA_mean_raw)  / RIPA_sd_raw,
      UDA   = (UDA_raw   - UDA_mean_raw)   / UDA_sd_raw
    )

  # Compute NDVI draws
  NDVI_nodam <- compute_NDVI_draws(grid$incis, grid$RIPA, grid$UDA, post, dam_flag = 0)
  NDVI_dam   <- compute_NDVI_draws(grid$incis, grid$RIPA, grid$UDA, post, dam_flag = 1)

  # Summaries (400 rows)
  pdp <- bind_rows(
    summarize_NDVI(NDVI_nodam) %>% mutate(dam = "No Dam"),
    summarize_NDVI(NDVI_dam)   %>% mutate(dam = "Dam Present")
  )

  # Repeat grid twice (400 rows)
  grid2 <- bind_rows(grid, grid)

  # Bind columns safely
  bind_cols(grid2, pdp)
}



#####

m2_pdp_incis <- build_pdp(
  focal_raw = incis_raw,
  focal_name = "incis",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_m2_NDVI_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)

m2_pdp_RIPA <- build_pdp(
  focal_raw = RIPA_raw,
  focal_name = "RIPA",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_m2_NDVI_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)

m2_pdp_UDA <- build_pdp(
  focal_raw = UDA_raw,
  focal_name = "UDA",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_m2_NDVI_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)



##########################################

# Plot Function:

plot_pdp <- function(pdp, xvar, xtitle) {
  ggplot(pdp,
         aes(x = .data[[xvar]], y = NDVI_med, color = dam, fill = dam)) +
    geom_ribbon(aes(ymin = NDVI_lo, ymax = NDVI_hi), alpha = 0.2, color = NA) +
    geom_line(linewidth = 1.2) +   # updated from size = 1.2
    scale_color_manual(values = c("steelblue", "brown4")) +
    scale_fill_manual(values = c("steelblue", "brown4")) +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = paste("Partial Dependence of NDVI on", xtitle),
      subtitle = "Predicted NDVI with 95% credible intervals\nConstrained to empirical covariate range",
      x = xtitle,
      y = "Predicted NDVI",
      color = "Dam Presence",
      fill  = "Dam Presence"
    ) +
    theme_minimal(base_size = 14)
}



###########################################3

plot_pdp(m2_pdp_incis, "incis_raw", "Channel Incision (raw)")
plot_pdp(m2_pdp_RIPA,  "RIPA_raw",  "Riparian Area (m^2)")
plot_pdp(m2_pdp_UDA,   "UDA_raw",   "Contributing Drainage Area (km^2)")

```


    NO DIFFERENCE IN NDVI BETWEEN DAMMED AND UNDAMMED REACHES

----

----

----

----

----

--->


----

----

----

----



NDVI ~ Dam Density
----

**MODEL 3: NDVI ~ damDensity (no Geom.Covar.)**

```{r,echo=FALSE}
#library(jagsUI)
```
    sink("HGLMM_m3_NDVI_damDensity.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
      
      sigma_eta_year ~ dnorm(0,1) T(0,)
      tau_eta_year <- pow(sigma_eta_year, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Amount:
      alpha ~ dnorm(0.32, 0.04)  # within 5 SD of empirical mean NDVI
      beta_damDen ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR Veg AMOUNT:
    
      for(j in 1:n_stream){
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        eta_year[yr] ~ dnorm(0, tau_eta_year)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # MAIN LOOP:
    
      for(i in 1:N){
        
        y[i] ~ dnorm(mu[i], tau)
    
        mu[i] <- alpha
                    + beta_damDen * damDen[i]
                    + eta_stream[stream[i]]
                    + eta_year[year[i]]
                    + eta_cluster[cluster[i]]    
                    
      ##########################################################################
                    
      # Derived parameter for WAIC computation:

      loglik[i] <- logdensity.norm(y[i], mu[i], tau)
      }
    }
    ", fill = TRUE)
    
    sink()

    

```{r,echo=FALSE}
hglmm_m3_NDVI_damDen_data_list <- list(
  N = N,
  y = y_NDVI,
  stream = stream,
  year = year,
  cluster = cluster,
  damDen = damDen,
  n_stream = n_stream,
  n_year = n_year,
  n_cluster = n_cluster
)

hglmm_m3_NDVI_damDen_inits <- function(){
  list()
}

hglmm_m3_NDVI_damDen_params <- c(
  "alpha",
  "mu",
  "beta_damDen",
  "eta_stream",
  "eta_year",
  "eta_cluster",
  "sigma",
  "sigma_eta_stream",
  "sigma_eta_year",
  "sigma_eta_cluster",
  "loglik"
)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000


Sys.time()

set.seed(12345)

hglmm_m3_NDVI_damDen <- jags(data = hglmm_m3_NDVI_damDen_data_list,
                   inits = hglmm_m3_NDVI_damDen_inits,
                   parameters.to.save = hglmm_m3_NDVI_damDen_params,
                   model.file = "HGLMM_m3_NDVI_damDensity.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

```

<!---



----


DIAGNOSTICS:
----


```{r,echo=FALSE}
# Extract posterior draws

damDen_m3_post <- hglmm_m3_NDVI_damDen$sims.list 
```

**Assess "Chain health" Rhat + ESS:**

```{r,echo=FALSE} 
# Choose only the parameters you care about
params_m3 <- c(
  "alpha",
  "beta_damDen",
  "sigma")

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m3_NDVI_damDen$samples[, params_m3], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m3_NDVI_damDen$samples[, params_m3])

```

    Rhat & Effective Sample Size:
    
    Rhat values look good.
    
    Effective sample sizes look good 

----

**Check traceplots** for critical parameters:

```{r,echo=FALSE} 

par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


tplot_params_m3 <- c("alpha","beta_damDen","sigma","sigma_eta_stream")


mcmc_sub_m3 <- hglmm_m3_NDVI_damDen$samples[, tplot_params_m3]

coda::traceplot(mcmc_sub_m3)
```


    Traceplots:
    
    Traceplots all look sufficiently "grassy," indicating chain convergence.


----

**Predicted NDVI vs observed NDVI**

Goal is 1:1 (ish)

```{r,echo=FALSE} 
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 

# Posterior mean of predicted NDVI 
m3_mu_NDVI_post <- apply(hglmm_m3_NDVI_damDen$sims.list$mu, 2, mean)

# Observed log vegetation area
m3_obs_NDVI <- y_NDVI

# Fit regression on the log scale
m3_fit <- lm(m3_obs_NDVI ~ m3_mu_NDVI_post)

m3_slope <- coef(m3_fit)[2]
m3_R2   <- summary(m3_fit)$r.squared

# Plot
plot(m3_mu_NDVI_post,
     m3_obs_NDVI,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated NDVI",
     ylab = "Observed NDVI")

abline(0, 1, col = "red", lwd = 2)
```

Slope:
```{r,echo=FALSE}
m3_slope
```

    1.07 is okay.

----

**Random effects shrinkage: **

Goal is centered around 0(ish)

1. eta_stream; random effect of stream on NDVI intercept; n=6
2. eta_year; random effect of sample event on NDVI intercept; n=3

```{r, echo=FALSE} 
par(mfrow = c(2, 1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hglmm_m3_NDVI_damDen$sims.list$eta_stream)

boxplot(hglmm_m3_NDVI_damDen$sims.list$eta_year)
```


----

----

**Posterior mean dam density effect**

damDen -> NDVI (no geom.covar.)

```{r,echo=FALSE}
m3_coeffparams <- c(#"alpha",
            "beta_damDen",
            "sigma")

m3_coef_df <- hglmm_m3_NDVI_damDen$summary[m3_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m3_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.2,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```


```{r,echo=FALSE}
summary(hglmm_m3_NDVI_damDen$sims.list$beta_damDen)

exp(mean(hglmm_m3_NDVI_damDen$sims.list$beta_damDen))
```

```{r,echo=FALSE}
#library(tidyverse)
#library(scales)

par(mfrow = c(1, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  
#Posterior distribution of damEffect on vegArea:

hist(hglmm_m3_NDVI_damDen$sims.list$beta_damDen,
     breaks = 50,
     col = "steelblue",
     main = "Posterior distribution of damDen effect on NDVI",
     xlab = "beta_damDen",
     xlim = c(0,0.025))
# Add median line
abline(v = median(hglmm_m3_NDVI_damDen$sims.list$beta_damDen), col = "red", lwd = 2)

```
    
    Without geomorphic covariates, 1 SD increase in dam density is associated with
    0.007 increase in NDVI.... so precisely nothing. Higher dam densities may have
    more influence, but we're only talking 0.1 increase or so.


----

**Prior-posterior overlap:**

```{r,echo=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 

damDen_alpha_post_m3 <- hglmm_m3_NDVI_damDen$sims.list$alpha
damDen_beta_post_m3  <- hglmm_m3_NDVI_damDen$sims.list$beta_damDen

damDenMod_m3_n <- length(damDen_alpha_post_m3)

damDen_alpha_prior_m3 <- rnorm(damDenMod_m3_n, 0.32, sqrt(1/0.04))
damDen_beta_prior_m3  <- rnorm(damDenMod_m3_n, 0, sqrt(1/0.04))


# Plot overlap:

#plot_prior_post <- function(prior, post, name) {
  
#  df <- tibble(
#    value = c(prior, post),
#    dist  = rep(c("Prior", "Posterior"), each = length(prior))
#  )
  
#  ggplot(df, aes(x = value, fill = dist)) +
#    geom_density(alpha = 0.4) +
#    scale_fill_manual(values = c("gray60", "steelblue")) +
#    labs(
#      title = paste("Prior vs Posterior for", name),
#      x = name,
#      y = "Density"
#    ) +
#    theme_minimal(base_size = 14)
#}

plot_prior_post(damDen_alpha_prior_m3, damDen_alpha_post_m3, "alpha_m3")
plot_prior_post(damDen_beta_prior_m3, damDen_beta_post_m3, "beta_damDen_m3")
```


    PPOs still look pretty good.

----

**Change in NDVI across damDensity by stream and year**

```{r,echo=FALSE}

m3_dam_seq <- seq(min(aoi_HurdleData$dam_km),
               max(aoi_HurdleData$dam_km),
               length.out = 50)

# build PDP grid
m3_pdp_stream <- map_dfr(1:n_stream, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m3_post$alpha
  beta_j  <- damDen_m3_post$beta_damDen
  eta_j   <- damDen_m3_post$eta_stream[, j] + rowMeans(damDen_m3_post$eta_year)

  map_dfr(m3_dam_seq, function(d) {

    mu_draws <- alpha_j + beta_j * d + eta_j

    tibble(
      stream = j,
      damDen = d,
      NDVI_med = median(mu_draws),
      NDVI_lo  = quantile(mu_draws, 0.025),
      NDVI_hi  = quantile(mu_draws, 0.975)
    )
  })
})

stream_lookup <- aoi_HurdleData %>%
  distinct(stream, watershed) %>%
  arrange(stream)

m3_pdp_stream <- m3_pdp_stream %>%
  left_join(stream_lookup, by = "stream")



# plot
ggplot(m3_pdp_stream, aes(x = watershed, y = NDVI_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Partial Dependence of Vegetation Area on Stream and Dam Density",
    x = "Stream",
    y = "NDVI",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


###############################################################
################################################################

# build PDP grid
m3_pdp_year <- map_dfr(1:n_year, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m3_post$alpha
  beta_j  <- damDen_m3_post$beta_damDen
  eta_j   <- damDen_m3_post$eta_year[, j] + rowMeans(damDen_m3_post$eta_stream)

  map_dfr(m3_dam_seq, function(d) {

    mu_draws <- alpha_j + beta_j * d + eta_j

    tibble(
      year = j,
      damDen = d,
      NDVI_med = median(mu_draws),
      NDVI_lo  = quantile(mu_draws, 0.025),
      NDVI_hi  = quantile(mu_draws, 0.975)
    )
  })
})

year_lookup <- tibble(
  year = 1:3,
  year_label = c(
    "2019",
    "2021",
    "2023"
  )
)


m3_pdp_year_named <- m3_pdp_year %>%
  left_join(year_lookup, by = "year")



# plot
ggplot(m3_pdp_year_named, aes(x = year_label, y = NDVI_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Partial Dependence of NDVI on Year and Dam Density",
    x = "Year",
    y = "Predicted NDVI",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```


----

----

--->


----

----

----



**MODEL 4: SW ~  damDensity + Geom.Covar.**

```{r,echo=FALSE}
#library(jagsUI)
```

    sink("HGLMM_m4_NDVI_damDensity.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
      
      sigma_eta_year ~ dnorm(0,1) T(0,)
      tau_eta_year <- pow(sigma_eta_year, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Amount:
      alpha ~ dnorm(0.32, 0.04)    # within 5 SD of empirical mean NDVI
      beta_damDen ~ dnorm(0,0.04)
      beta_incis ~ dnorm(0,0.04)
      beta_RIPA ~ dnorm(0,0.04)
      beta_UDA ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR Veg AMOUNT:
    
      for(j in 1:n_stream){
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        eta_year[yr] ~ dnorm(0, tau_eta_year)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # MAIN LOOP:
    
      for(i in 1:N){
        
        y[i] ~ dnorm(mu[i], tau)
    
        mu[i] <- alpha
                    + beta_damDen * damDen[i]
                    + beta_incis * incis[i]
                    + beta_RIPA * RIPA[i]
                    + beta_UDA * UDA[i]
                    + eta_stream[stream[i]]
                    + eta_year[year[i]]
                    + eta_cluster[cluster[i]]
                    
      ##########################################################################
                    
      # Derived parameter for WAIC computation:

      loglik[i] <- logdensity.norm(y[i], mu[i], tau)
      
      }
    }
    ", fill = TRUE)
    
    sink()

    


M4_0-m Buffer:
----

```{r,echo=FALSE}
# hglmm_damDen_m4_0_data_list <- list(
#   N = N,
#   y = log(y_NDVI_0 + 1),
#   stream = stream,
#   year = year,
#   cluster = cluster,
#   damDen = damDen,
#   incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
#   RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2_0)), #RIPA of DEM-derived pondableZone
#   UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)), # Upstream Drainage Area
#   n_stream = n_stream,
#   n_year = n_year,
#   n_cluster = n_cluster
# )

# hglmm_damDen_m4_inits <- function(){
#   list()
# }

# hglmm_damDen_m4_params <- c(
#   "alpha",
#   "mu",
#   "beta_damDen",
#   "beta_incis",
#   "beta_RIPA",
#   "beta_UDA",
#   "eta_stream",
#   "eta_year",
#   "eta_cluster",
#   "sigma",
#   "sigma_eta_stream",
#   "sigma_eta_event",
#   "sigma_eta_cluster",
#   "loglik"
# )

# nchains <- 3
# nthin <- 5
# niter <- 100000
# nburn <- 25000


# Sys.time()

# set.seed(12345)

# hglmm_m4_0_damDen_NDVI <- jags(data = hglmm_damDen_m4_0_data_list,
#                    inits = hglmm_damDen_m4_inits,
#                    parameters.to.save = hglmm_damDen_m4_params,
#                    model.file = "HGLMM_m4_NDVI_damDensity.jags",
#                    parallel = T,
#                    n.chains = nchains,
#                    n.thin = nthin,
#                    n.iter = niter,
#                    n.burnin = nburn)


# Sys.time()

#print(hglmm_m4_0_damDen_NDVI) # model with local and watershed dam density as fixed effects
#print(hglmm_m4_0_damDen_NDVI$summary[grep("beta_", rownames(hglmm_m4_0_damDen_NDVI$summary)), ])
#print(hglmm_m4_0_damDen_NDVI$summary[grep("zeta_", rownames(hglmm_m4_0_damDen_NDVI$summary)), ])

```


----

M4_10-m Buffer:
----

```{r,echo=FALSE}
# hglmm_damDen_m4_10_data_list <- list(
#   N = N,
#   y = log(y_NDVI_10 + 1),
#   stream = stream,
#   year = year,
#   cluster = cluster,
#   damDen = damDen,
#   incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
#   RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2_0)), #RIPA of DEM-derived pondableZone
#   UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)), # Upstream Drainage Area
#   n_stream = n_stream,
#   n_year = n_year,
#   n_cluster = n_cluster
# )

# hglmm_damDen_m4_inits <- function(){
#   list()
# }

# hglmm_damDen_m4_params <- c(
#   "alpha",
#   "mu",
#   "beta_damDen",
#   "beta_incis",
#   "beta_RIPA",
#   "beta_UDA",
#   "eta_stream",
#   "eta_year",
#   "eta_cluster",
#   "sigma",
#   "sigma_eta_stream",
#   "sigma_eta_event",
#   "sigma_eta_cluster",
#   "loglik"
# )

# nchains <- 3
# nthin <- 5
# niter <- 100000
# nburn <- 25000


# Sys.time()

# set.seed(12345)

# hglmm_m4_10_damDen_NDVI <- jags(data = hglmm_damDen_m4_10_data_list,
#                    inits = hglmm_damDen_m4_inits,
#                    parameters.to.save = hglmm_damDen_m4_params,
#                    model.file = "HGLMM_m4_NDVI_damDensity.jags",
#                    parallel = T,
#                    n.chains = nchains,
#                    n.thin = nthin,
#                    n.iter = niter,
#                    n.burnin = nburn)


# Sys.time()

#print(hglmm_m4_0_damDen_NDVI) # model with local and watershed dam density as fixed effects
#print(hglmm_m4_10_damDen_NDVI$summary[grep("beta_", rownames(hglmm_m4_10_damDen_NDVI$summary)), ])
#print(hglmm_m4_10_damDen_NDVI$summary[grep("zeta_", rownames(hglmm_m4_10_damDen_NDVI$summary)), ])

```

----

M4_20-m Buffer:
----

```{r,echo=FALSE}
# hglmm_damDen_m4_20_data_list <- list(
#   N = N,
#   y = log(y_NDVI_20 + 1),
#   stream = stream,
#   year = year,
#   cluster = cluster,
#   damDen = damDen,
#   incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
#   RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2_0)), #RIPA of DEM-derived pondableZone
#   UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)), # Upstream Drainage Area
#   n_stream = n_stream,
#   n_year = n_year,
#   n_cluster = n_cluster
# )

# hglmm_damDen_m4_inits <- function(){
#   list()
# }

# hglmm_damDen_m4_params <- c(
#   "alpha",
#   "mu",
#   "beta_damDen",
#   "beta_incis",
#   "beta_RIPA",
#   "beta_UDA",
#   "eta_stream",
#   "eta_year",
#   "eta_cluster",
#   "sigma",
#   "sigma_eta_stream",
#   "sigma_eta_event",
#   "sigma_eta_cluster",
#   "loglik"
# )

# nchains <- 3
# nthin <- 5
# niter <- 100000
# nburn <- 25000


# Sys.time()

# set.seed(12345)

# hglmm_m4_20_damDen_NDVI <- jags(data = hglmm_damDen_m4_20_data_list,
  #                  inits = hglmm_damDen_m4_inits,
  #                  parameters.to.save = hglmm_damDen_m4_params,
  #                  model.file = "HGLMM_m4_NDVI_damDensity.jags",
  #                  parallel = T,
  #                  n.chains = nchains,
  #                  n.thin = nthin,
  #                  n.iter = niter,
  #                  n.burnin = nburn)


# Sys.time()

#print(hglmm_m4_20_damDen_NDVI) # model with local and watershed dam density as fixed effects
#print(hglmm_m4_20_damDen_NDVI$summary[grep("beta_", rownames(hglmm_m4_20_damDen_NDVI$summary)), ])
#print(hglmm_m4_20_damDen_NDVI$summary[grep("zeta_", rownames(hglmm_m4_20_damDen_NDVI$summary)), ])

```

----

M4_30-m Buffer:
----

```{r,echo=FALSE}
# hglmm_damDen_m4_30_data_list <- list(
#   N = N,
#   y = log(y_NDVI_30 + 1),
#   stream = stream,
#   year = year,
#   cluster = cluster,
#   damDen = damDen,
#   incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
#   RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2_0)), #RIPA of DEM-derived pondableZone
#   UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)), # Upstream Drainage Area
#   n_stream = n_stream,
#   n_year = n_year,
#   n_cluster = n_cluster
# )

# hglmm_damDen_m4_inits <- function(){
#   list()
# }

# hglmm_damDen_m4_params <- c(
#   "alpha",
#   "mu",
#   "beta_damDen",
#   "beta_incis",
#   "beta_RIPA",
#   "beta_UDA",
#   "eta_stream",
#   "eta_year",
#   "eta_cluster",
#   "sigma",
#   "sigma_eta_stream",
#   "sigma_eta_event",
#   "sigma_eta_cluster",
#   "loglik"
# )

# nchains <- 3
# nthin <- 5
# niter <- 100000
# nburn <- 25000


# Sys.time()

# set.seed(12345)

# hglmm_m4_30_damDen_NDVI <- jags(data = hglmm_damDen_m4_30_data_list,
#                    inits = hglmm_damDen_m4_inits,
#                    parameters.to.save = hglmm_damDen_m4_params,
#                    model.file = "HGLMM_m4_NDVI_damDensity.jags",
#                    parallel = T,
#                    n.chains = nchains,
#                    n.thin = nthin,
#                    n.iter = niter,
#                    n.burnin = nburn)


# Sys.time()

#print(hglmm_m4_30_damDen_NDVI) # model with local and watershed dam density as fixed effects
#print(hglmm_m4_30_damDen_NDVI$summary[grep("beta_", rownames(hglmm_m4_30_damDen_NDVI$summary)), ])
#print(hglmm_m4_30_damDen_NDVI$summary[grep("zeta_", rownames(hglmm_m4_30_damDen_NDVI$summary)), ])

```





```{r,echo=FALSE}
# SAVE THE MODEL OUTPUTS SO YOU DON'T HAVE TO RUN THEM EVERY TIME!!

# saveRDS(hglmm_m4_0_damDen_NDVI ,
#         "C:\\Users\\mattw\\Desktop\\Ch1_ModelOutputs\\hglmm_m4_0_damDen_NDVI .rds")

# saveRDS(hglmm_m4_10_damDen_NDVI ,
#         "C:\\Users\\mattw\\Desktop\\Ch1_ModelOutputs\\hglmm_m4_10_damDen_NDVI .rds")

# saveRDS(hglmm_m4_20_damDen_NDVI ,
#         "C:\\Users\\mattw\\Desktop\\Ch1_ModelOutputs\\hglmm_m4_20_damDen_NDVI .rds")

# saveRDS(hglmm_m4_30_damDen_NDVI ,
#         "C:\\Users\\mattw\\Desktop\\Ch1_ModelOutputs\\hglmm_m4_30_damDen_NDVI .rds")


hglmm_m4_0_damDen_NDVI  <- readRDS("C:\Users\mattw\Desktop\Ch1_ModelOutputs\hglmm_m4_0_damDen_NDVI .rds")

hglmm_m4_10_damDen_NDVI  <- readRDS("C:\Users\mattw\Desktop\Ch1_ModelOutputs\hglmm_m4_10_damDen_NDVI .rds")

hglmm_m4_20_damDen_NDVI  <- readRDS("C:\Users\mattw\Desktop\Ch1_ModelOutputs\hglmm_m4_20_damDen_NDVI .rds")

hglmm_m4_30_damDen_NDVI  <- readRDS("C:\Users\mattw\Desktop\Ch1_ModelOutputs\hglmm_m4_30_damDen_NDVI .rds")
```











----









DIAGNOSTICS:
----


```{r,echo=FALSE}
# Extract posterior draws

damDen_m4_0_post <- hglmm_m4_0_damDen_NDVI$sims.list 

damDen_m4_10_post <- hglmm_m4_10_damDen_NDVI$sims.list 

damDen_m4_20_post <- hglmm_m4_20_damDen_NDVI$sims.list 

damDen_m4_30_post <- hglmm_m4_30_damDen_NDVI$sims.list 

```

**Assess "Chain health" Rhat + ESS:**

**0-m Buff:**

```{r,echo=FALSE} 
# Choose only the parameters you care about
params_m4 <- c(
  "alpha",
  "beta_damDen",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "sigma")

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m4_0_damDen_NDVI$samples[, params_m4], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m4_0_damDen_NDVI$samples[, params_m4])

```

    Rhat & Effective Sample Size:
    
    Solid. Rhats near 1; ESS consistently high
    
    
    


**10-m Buff:**

```{r,echo=FALSE} 

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m4_10_damDen_NDVI$samples[, params_m4], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m4_10_damDen_NDVI$samples[, params_m4])

```

    Rhat & Effective Sample Size:
    
    Solid. Rhats near 1; ESS consistently high
    
    
    
**20-m Buff:**

```{r,echo=FALSE} 

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m4_20_damDen_NDVI$samples[, params_m4], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m4_20_damDen_NDVI$samples[, params_m4])

```

    Rhat & Effective Sample Size:
    
    Solid. Rhats near 1; ESS consistently high




**30-m Buff:**

```{r,echo=FALSE} 

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m4_30_damDen_NDVI$samples[, params_m4], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m4_30_damDen_NDVI$samples[, params_m4])

```

    Rhat & Effective Sample Size:
    
    Solid. Rhats near 1; ESS consistently high
    
    
    
<!---
----

**Check traceplots** for critical parameters:

```{r,echo=FALSE} 

par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


tplot_params_m4 <- c("alpha","beta_damDen","beta_incis","beta_RIPA","beta_UDA",
                     "sigma","sigma_eta_stream")


mcmc_sub_m4_0 <- hglmm_m4_0_damDen_NDVI$samples[, tplot_params_m4]

coda::traceplot(mcmc_sub_m4_0)
```


    Traceplots all look good.


----


**Predicted NDVI vs observed NDVI**

Goal is 1:1 (ish)

```{r,echo=FALSE} 
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 

# Posterior mean of predicted NDVI 
m4_0_mu_NDVI_post <- apply(hglmm_m4_0_damDen_NDVI$sims.list$mu, 2, mean)

# Observed NDVI
m4_0_obs_NDVI <- y_NDVI_0

# Fit regression on the log scale
m4_0_fit <- lm(m4_0_obs_NDVI ~ m4_0_mu_NDVI_post)

m4_0_slope <- coef(m4_0_fit)[2]
m4_0_R2    <- summary(m4_0_fit)$r.squared

# Plot
plot(m4_0_mu_NDVI_post,
     m4_0_obs_NDVI,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated NDVI",
     ylab = "Observed NDVI")

abline(0, 1, col = "red", lwd = 2)

```

Slope:
```{r,echo=FALSE}
m4_0_slope
```

    Near 1 = good. Noticeable underestimation of variance at low NDVI_0

----




**Random effects shrinkage: **

Goal is centered around 0(ish)

1. eta_year; random effect of year log(VA+1) intercept
2. eta_stream; random effect of stream on log(VA+1) intercept

```{r, echo=FALSE} 
par(mfrow = c(1, 2),        # rows, columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hglmm_m4_0_damDen_NDVI$sims.list$eta_year)
boxplot(hglmm_m4_0_damDen_NDVI$sims.list$eta_stream)
```


----

**Prior-posterior overlap:**

```{r,echo=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 

damDen_alpha_m4_0_post <- hglmm_m4_0_damDen_NDVI$sims.list$alpha
damDen_beta_m4_0_post  <- hglmm_m4_0_damDen_NDVI$sims.list$beta_damDen
damDen_beta_incis_m4_0_post <- hglmm_m4_0_damDen_NDVI$sims.list$beta_incis
damDen_sigmaEtaStream_m4_0_post <- hglmm_m4_0_damDen_NDVI$sims.list$sigma_eta_stream

damDenMod_m4_0_n <- length(damDen_alpha_m4_0_post)

damDen_alpha_m4_0_prior <- rnorm(damDenMod_m4_0_n, 0, sqrt(1/0.04))
damDen_beta_m4_0_prior  <- rnorm(damDenMod_m4_0_n, 0, sqrt(1/0.04))
damDen_beta_incis_m4_0_prior  <- rnorm(damDenMod_m4_0_n, 0, sqrt(1/0.04))
damDen_sigmaEtaStream_m4_0_prior <- pmax(rnorm(damDenMod_m4_0_n, 0, 1), 0)


# Plot overlap:

plot_prior_post <- function(prior, post, name) {
  
  df <- tibble(
    value = c(prior, post),
    dist  = rep(c("Prior", "Posterior"), each = length(prior))
  )
  
  ggplot(df, aes(x = value, fill = dist)) +
    geom_density(alpha = 0.4) +
    scale_fill_manual(values = c("gray60", "steelblue")) +
    labs(
      title = paste("Prior vs Posterior for", name),
      x = name,
      y = "Density"
    ) +
    theme_minimal(base_size = 14)
}

plot_prior_post(damDen_alpha_m4_0_prior, damDen_alpha_m4_0_post, "alpha_m4")
plot_prior_post(damDen_beta_m4_0_prior, damDen_beta_m4_0_post, "beta_damDen_m4")
plot_prior_post(damDen_beta_incis_m4_0_prior, damDen_beta_incis_m4_0_post, "beta_incis_m4")
plot_prior_post(damDen_sigmaEtaStream_m4_0_prior, damDen_sigmaEtaStream_m4_0_post, "sigma_eta_stream_m4")
```

    PPOs look good pretty good.

----
--->


**Posterior Estimates**


**0-m Buffer; i.e., DEM-derived 'pondable zone'**

```{r,echo=FALSE}

m4_0_alpha <- mean(hglmm_m4_0_damDen_NDVI$sims.list$alpha)

m4_0_beta_dd <- mean(hglmm_m4_0_damDen_NDVI$sims.list$beta_damDen)

m4_0_beta_in <- mean(hglmm_m4_0_damDen_NDVI$sims.list$beta_incis)

m4_0_beta_rip <- mean(hglmm_m4_0_damDen_NDVI$sims.list$beta_RIPA)

m4_0_beta_uda <- mean(hglmm_m4_0_damDen_NDVI$sims.list$beta_UDA)

```

```{r,echo=FALSE}
m4_coeffparams <- c(#"alpha",
            "beta_damDen",
            "beta_incis",
            "beta_RIPA",
            "beta_UDA",
            "sigma")

m4_0_coef_df <- hglmm_m4_0_damDen_NDVI$summary[m4_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m4_0_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.5,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    subtitle = "0-m Buffer; DEM-derived 'pondable zone'",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```

----

**Effect of 1 SD increase in dam density on NDVI:**

```{r,echo=FALSE}
summary(hglmm_m4_0_damDen_NDVI$sims.list$beta_damDen)

#exp(mean(hglmm_m4_0_damDen_NDVI$sims.list$beta_damDen))

#NDVI_0_damDen_0 <- m4_0_alpha  # with geom covar. at z-scaled mean (0)
#NDVI_damDen_0

#NDVI_0_damDen_1 <- m4_0_alpha + 0.005  #make sure this value is correct
#NDVI_damDen_1

print("NDVI_0 change with 1 SD increase in dam density")
(NDVI_0_damDen_1 - NDVI_0_damDen_0)

```

    A 1 SD increase dam density is associated with INCREASE of 0.005 in NDVI WITHIN 0-m BUFFER around DEM-delineated riparian/pondable zone, for a "typical" stream under mean geomorphic covariate conditions.
    
    Note: at the upper end of the observed dam density range, a typical stream at mean covariate 
    conditions could see an estimated 0.025 increase in riparian NDVI
    
    NOTE: THIS IS CONFOUNDED BY PRESENCE/AMMOUNT OF SURFACE WATER IN THE 0-m BUFFER ZONE:
    
    
----

**Estimated effects of geomorphic covariates**

----

**Effect of 1 SD increase in incision on NDVI:**

```{r,echo=FALSE}
summary(hglmm_m4_0_damDen_NDVI$sims.list$beta_incis)

#exp(mean(hglmm_m4_0_damDen_NDVI$sims.list$beta_incis))


#NDVI_0_incis_0 <- m4_0_alpha
#NDVI_incis_0

#NDVI_0_incis_1 <- m4_0_alpha - 0.05 # make sure this is correct
#NDVI_incis_1

#print("NDVI change with 1 SD increase in incision")
#(NDVI_0_incis_1 - NDVI_0_incis_0)
```

    A 1 SD increase in incision (~0.1) is associated with a decrease of 0.005 in NDVI, for a typical stream at mean covariate conditions. 
    
    Realistically, incision doesn't seem to influence NDVI. Weakly positive, with 
    heavy overlap of 0.
    

----
 
**Effect of 1 SD increase in pond-able space on NDVI:**

```{r,echo=FALSE}
summary(hglmm_m4_0_damDen_NDVI$sims.list$beta_RIPA)
```


    A 1 SD increase in RIPA (11000 m^2) is associated with a decrease of 0.006 in ndvi WITHIN 0-m RIP BUFFER
    (for typical stream under mean conditions).
    
----
    
**Effect of 1 SD increase in Upstream Drainage Area on SW amount:**

```{r,echo=FALSE}
summary(hglmm_m4_0_damDen_NDVI$sims.list$beta_UDA)


```

    A 1 SD increase in Upstream Drainage Area (UDA) (~1300km^2) is associated with an increase of 0.016 in 
    NDVI, for a "typical" stream at mean covariate conditions. 
    






NOTE: CHANGE OF NDVI < 0.1 is NEGLIGIBLE. BEAVER DAMS LIKELY DON'T CHANGE VEG STRUCTURE FROM UNIMPOUNDED CONDITIONS
  

----

----

----

----

**10-m Planar Buffer around DEM-derived 'pondable zone'**

```{r,echo=FALSE}

m4_10_alpha <- mean(hglmm_m4_10_damDen_NDVI$sims.list$alpha)

m4_10_beta_dd <- mean(hglmm_m4_10_damDen_NDVI$sims.list$beta_damDen)

m4_10_beta_in <- mean(hglmm_m4_10_damDen_NDVI$sims.list$beta_incis)

m4_10_beta_rip <- mean(hglmm_m4_10_damDen_NDVI$sims.list$beta_RIPA)

m4_10_beta_uda <- mean(hglmm_m4_10_damDen_NDVI$sims.list$beta_UDA)

```

```{r,echo=FALSE}
m4_10_coef_df <- hglmm_m4_10_damDen_NDVI$summary[m4_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m4_10_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.5,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    subtitle = "10-m Planar Buffer",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```

----

**Effect of 1 SD increase in dam density on NDVI:**

```{r,echo=FALSE}
summary(hglmm_m4_10_damDen_NDVI$sims.list$beta_damDen)
```

    A 1 SD increase dam density is associated with an increase of 0.005 in NDVI WITHIN 10-m BUFFER around DEM-delineated riparian/pondable zone, for a "typical" stream under mean geomorphic covariate conditions.
    
    Note: at the upper end of the observed dam density range, a typical stream at mean covariate 
    conditions could see an estimated 0.025 in riparian NDVI
    
    
    NO DIFFERENCE FROM 0-m BUFFER ZONE
    
----

**Estimated effects of geomorphic covariates**

----

**Effect of 1 SD increase in incision on NDVI:**

```{r,echo=FALSE}
summary(hglmm_m4_10_damDen_NDVI$sims.list$beta_incis)

```

    A 1 SD increase in incision (~0.1) is associated with a decrease of 0.005 in NDVI, within the 10-m RIP buffer,
    for a typical stream at mean covariate conditions.

----
 
**Effect of 1 SD increase in riparian area on NDVI:**

```{r,echo=FALSE}
summary(hglmm_m4_10_damDen_NDVI$sims.list$beta_RIPA)

```


    A 1 SD increase in RIPA (11000 m^2) is associated with decrease of 0.006 in NDVI WITHIN 10-m RIP buffer (for typical stream under mean conditions).

----
    
**Effect of 1 SD increase in Upstream Drainage Area on SW amount:**

```{r,echo=FALSE}
summary(hglmm_m4_10_damDen_NDVI$sims.list$beta_UDA)
```

    A 1 SD increase in Upstream Drainage Area (UDA) (~1300km^2) is associated with an increase of 0.022 in 
    NDVI, for a "typical" stream at mean covariate conditions. 
   

----

----

----

----

**20-m Planar Buffer around DEM-derived 'pondable zone'**

```{r,echo=FALSE}

m4_20_alpha <- mean(hglmm_m4_20_damDen_NDVI$sims.list$alpha)

m4_20_beta_dd <- mean(hglmm_m4_20_damDen_NDVI$sims.list$beta_damDen)

m4_20_beta_in <- mean(hglmm_m4_20_damDen_NDVI$sims.list$beta_incis)

m4_20_beta_rip <- mean(hglmm_m4_20_damDen_NDVI$sims.list$beta_RIPA)

m4_20_beta_uda <- mean(hglmm_m4_20_damDen_NDVI$sims.list$beta_UDA)

```

```{r,echo=FALSE}
m4_20_coef_df <- hglmm_m4_20_damDen_NDVI$summary[m4_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m4_20_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.5,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    subtitle = "20-m Planar Buffer",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```

----

**Effect of 1 SD increase in dam density on NDVI:**

```{r,echo=FALSE}
summary(hglmm_m4_20_damDen_NDVI$sims.list$beta_damDen)
```

    A 1 SD increase dam density is associated with an increase of ~0.005 in NDVI WITHIN 20-m BUFFER around DEM-delineated riparian/pondable zone, for a "typical" stream under mean geomorphic covariate conditions.
    
    Note: at the upper end of the observed dam density range, a typical stream at mean covariate 
    conditions could see an estimated 0.025 in riparian NDVI 
    
    
----

**Estimated effects of geomorphic covariates**

----

**Effect of 1 SD increase in incision on NDVI:**

```{r,echo=FALSE}
summary(hglmm_m4_20_damDen_NDVI$sims.list$beta_incis)

```

    A 1 SD increase in incision (~0.1) is associated with a decrease of ~0.005 in NDVI, within the 20-m RIP buffer, 
    for a typical stream at mean covariate conditions. 

----
 
**Effect of 1 SD increase in riparian area on NDVI:**

```{r,echo=FALSE}
summary(hglmm_m4_20_damDen_NDVI$sims.list$beta_RIPA)
```


    A 1 SD increase in RIPA (11000 m^2) is associated with a decrease of 0.004 in NDVI WITHIN 20-m RIP buffer 
    (for typical stream under mean conditions).

----
    
**Effect of 1 SD increase in Upstream Drainage Area on SW amount:**

```{r,echo=FALSE}
summary(hglmm_m4_20_damDen_NDVI$sims.list$beta_UDA)

```

    A 1 SD increase in Upstream Drainage Area (UDA) (~1300km^2) is associated with an increase of 0.021 in 
    NDVI, for a "typical" stream at mean covariate conditions. 




----

----

----

----

**30-m Planar Buffer around DEM-derived 'pondable zone'**

```{r,echo=FALSE}

m4_30_alpha <- mean(hglmm_m4_30_damDen_NDVI$sims.list$alpha)

m4_30_beta_dd <- mean(hglmm_m4_30_damDen_NDVI$sims.list$beta_damDen)

m4_30_beta_in <- mean(hglmm_m4_30_damDen_NDVI$sims.list$beta_incis)

m4_30_beta_rip <- mean(hglmm_m4_30_damDen_NDVI$sims.list$beta_RIPA)

m4_30_beta_uda <- mean(hglmm_m4_30_damDen_NDVI$sims.list$beta_UDA)

```

```{r,echo=FALSE}
m4_30_coef_df <- hglmm_m4_30_damDen_NDVI$summary[m4_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m4_30_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.5,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    subtitle = "30-m Planar Buffer",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```

----

**Effect of 1 SD increase in dam density on NDVI:**

```{r,echo=FALSE}
summary(hglmm_m4_30_damDen_NDVI$sims.list$beta_damDen)

```

    A 1 SD increase dam density is associated with an increase of 0.005 in NDVI WITHIN 30-m BUFFER around DEM-delineated riparian/pondable zone, for a "typical" stream under mean geomorphic covariate conditions.
    
    Note: at the upper end of the observed dam density range, a typical stream at mean covariate 
    conditions could see an estimated 0.25 in riparian NDVI 
    
----

**Estimated effects of geomorphic covariates**

----

**Effect of 1 SD increase in incision on NDVI:**

```{r,echo=FALSE}
summary(hglmm_m4_30_damDen_NDVI$sims.list$beta_incis)

```

    A 1 SD increase in incision (~0.1) is associated with a ____ in NDVI, within the 30-m RIP buffer,
    for a typical stream at mean covariate conditions. 
    
    Realistically, incision doesn't seem to influence NDVI.
    
    <Weak negative effect, heavily overlapping 0.>

----
 
**Effect of 1 SD increase in riparian area on NDVI:**

```{r,echo=FALSE}
summary(hglmm_m4_30_damDen_NDVI$sims.list$beta_RIPA)

```


    A 1 SD increase in RIPA (11000 m^2) is associated with a decrease of 0.004% increase in NDVI WITHIN 30-m RIP buffer (for typical stream under mean conditions).
  

----
    
**Effect of 1 SD increase in Upstream Drainage Area on NDVI:**

```{r,echo=FALSE}
summary(hglmm_m4_30_damDen_NDVI$sims.list$beta_UDA)

```

    A 1 SD increase in Upstream Drainage Area (UDA) (~1300km^2) is associated with an increase of 0.020 in 
    NDVI for a "typical" stream at mean covariate conditions. 
    





**TAKE-AWAY**
----

ALL POSTERIOR ESTIMATES FOR COVARIATE EFFECTS (though some are 'significantly' positive or negative) ARE MINISCULE ON THE RAW NDVI SCALE. A CHANGE OF <0.1 IS TRIVIAL. Beaver dams aren't notably changing vegetation strucuture within 30-m of the stream. None of our included covariates really move the NDVI needle. 

For perspective, the estimated intercept for NDVI is around 0.3... max, 1 SD change in any of these covariates is affecting NDVI by 0.004 - 0.02....practically nothing.














----


----

**Partial dependence plots**

Constrain PDP to empirical ranges for SWA and target covariates

```{r,echo=FALSE}
# Raw covariates
incis_m4_raw <- 1 - aoi_HurdleData$TO/90
RIPA_m4_raw  <- aoi_HurdleData$riparian_area_m2
UDA_m4_raw   <- aoi_HurdleData$catchment_km2
#damDen <- aoi_HurdleData$dam_km_std
damDen_m4_raw <- aoi_HurdleData$dam_km
damDen_m4_mean_raw <- mean(damDen_m4_raw)
damDen_m4_sd_raw   <- sd(damDen_m4_raw)

NDVI_m4_mean <- mean(y_NDVI)
NDVI_m4_sd   <- sd(y_NDVI)

# Standardization stats
incis_m4_mean_raw <- mean(incis_m4_raw)
incis_m4_sd_raw   <- sd(incis_m4_raw)

RIPA_m4_mean_raw  <- mean(RIPA_m4_raw)
RIPA_m4_sd_raw    <- sd(RIPA_m4_raw)

UDA_m4_mean_raw   <- mean(UDA_m4_raw)
UDA_m4_sd_raw     <- sd(UDA_m4_raw)


##############################3

# Helper Functions to Compute SWA from posterior draws
compute_NDVI_draws_damDen <- function(incis, RIPA, UDA, post, damDen_val) {
  n_draws <- length(post$alpha)
  NDVI_mat <- matrix(NA_real_, nrow = length(incis), ncol = n_draws)

  for (j in seq_len(n_draws)) {
    lp <- post$alpha[j] +
          post$beta_incis[j] * incis +
          post$beta_RIPA[j]  * RIPA +
          post$beta_UDA[j]   * UDA +
          post$beta_damDen[j] * damDen_val

    NDVI_mat[, j] <- lp
  }

  NDVI_mat
}

summarize_NDVI <- function(NDVI_mat) {
  tibble(
    NDVI_med = apply(NDVI_mat, 1, median),
    NDVI_lo  = apply(NDVI_mat, 1, quantile, probs = 0.025),
    NDVI_hi  = apply(NDVI_mat, 1, quantile, probs = 0.975)
  )
}




#########################################

# PDP Builder:

build_pdp_damDen <- function(focal_raw, focal_name,
                      incis_raw, RIPA_raw, UDA_raw,
                      post,
                      incis_mean_raw, incis_sd_raw,
                      RIPA_mean_raw, RIPA_sd_raw,
                      UDA_mean_raw, UDA_sd_raw,
                      damDen_raw) {

  # Sequence only across empirical range
  focal_seq <- seq(min(focal_raw), max(focal_raw), length.out = 200)

  # Conditional medians of other covariates
  incis_med <- median(incis_raw)
  RIPA_med  <- median(RIPA_raw)
  UDA_med   <- median(UDA_raw)
  
  # Empirical dam density range (standardized)
  damDen_seq <- seq(min(damDen_raw), max(damDen_raw), length.out = 10)

  # Build grid (200 rows)
  grid <- tibble(
    focal_raw = focal_seq,
    incis_raw = if (focal_name == "incis") focal_seq else incis_med,
    RIPA_raw  = if (focal_name == "RIPA")  focal_seq else RIPA_med,
    UDA_raw   = if (focal_name == "UDA")   focal_seq else UDA_med
  ) %>%
    mutate(
      incis = (incis_raw - incis_mean_raw) / incis_sd_raw,
      RIPA  = (RIPA_raw  - RIPA_mean_raw)  / RIPA_sd_raw,
      UDA   = (UDA_raw   - UDA_mean_raw)   / UDA_sd_raw
    )
  # Compute PDP curves for each dam density value
  pdp_list <- map(damDen_seq, function(dd) {
    NDVI_mat <- compute_NDVI_draws_damDen(grid$incis, grid$RIPA, grid$UDA, post,
                                        (dd - damDen_m4_mean_raw) / damDen_m4_sd_raw)
    summarize_NDVI(NDVI_mat) %>% mutate(damDen = dd)
  })

  pdp <- bind_rows(pdp_list)

  # Repeat grid for each damDen value
  grid2 <- bind_rows(replicate(length(damDen_seq), grid, simplify = FALSE))

  bind_cols(grid2, pdp)
}



#####

pdp_incis_m4_damDen <- build_pdp_damDen(
  focal_raw = incis_m4_raw,
  focal_name = "incis",
  incis_m4_raw, RIPA_m4_raw, UDA_m4_raw,
  damDen_m4_post,
  incis_m4_mean_raw, incis_m4_sd_raw,
  RIPA_m4_mean_raw, RIPA_m4_sd_raw,
  UDA_m4_mean_raw, UDA_m4_sd_raw, damDen_m4_raw
)

pdp_RIPA_m4_damDen <- build_pdp_damDen(
  focal_raw = RIPA_m4_raw,
  focal_name = "RIPA",
  incis_m4_raw, RIPA_m4_raw, UDA_m4_raw,
  damDen_m4_post,
  incis_m4_mean_raw, incis_m4_sd_raw,
  RIPA_m4_mean_raw, RIPA_m4_sd_raw,
  UDA_m4_mean_raw, UDA_m4_sd_raw, damDen_m4_raw
)

pdp_UDA_m4_damDen <- build_pdp_damDen(
  focal_raw = UDA_m4_raw,
  focal_name = "UDA",
  incis_m4_raw, RIPA_m4_raw, UDA_m4_raw,
  damDen_m4_post,
  incis_m4_mean_raw, incis_m4_sd_raw,
  RIPA_m4_mean_raw, RIPA_m4_sd_raw,
  UDA_m4_mean_raw, UDA_m4_sd_raw, damDen_m4_raw
)



##########################################3

# Plot Function:

plot_pdp_damDen <- function(pdp, xvar, xtitle) {
  ggplot(pdp,
         aes(x = .data[[xvar]], y = NDVI_med,
             color = damDen, fill = damDen)) +
    geom_point() +
    scale_color_viridis_c(option = "plasma") +
    scale_fill_viridis_c(option = "plasma") +
    scale_y_continuous(labels = comma) +
    labs(
      title = paste("Partial Dependence of NDVI on Dam Density and", xtitle),
      subtitle = "Across empirical dam density range",
      x = xtitle,
      y = "Predicted NDVI",
      color = "Dam Density (dams/km)",
      fill  = "Dam Density (dams/km)"
    ) +
    theme_minimal(base_size = 14)
}



###########################################3


plot_pdp_damDen(pdp_incis_m4_damDen, "incis_raw", "Channel Incision (raw)")
plot_pdp_damDen(pdp_RIPA_m4_damDen,  "RIPA_raw",  "Riparian Area (m^2)")
plot_pdp_damDen(pdp_UDA_m4_damDen,   "UDA_raw",   "Contributing Drainage Area (km^2)")


```

----

**Change in SWA across damDensity by stream and year**

```{r,echo=FALSE}

m4_dam_seq <- seq(min(aoi_HurdleData$dam_km),
               max(aoi_HurdleData$dam_km),
               length.out = 50)
m4_scaledDamDen_seq <- (m4_dam_seq - damDen_m4_mean_raw) / damDen_m4_sd_raw

# build PDP grid
m4_pdp_stream <- map_dfr(1:n_stream, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m4_post$alpha
  
  beta_j  <- damDen_m4_post$beta_damDen 
  
  eta_j   <- damDen_m4_post$eta_stream[, j] + rowMeans(damDen_m4_post$eta_year)

  map2_dfr(m4_dam_seq, m4_scaledDamDen_seq, function(d_raw, d_scaled) {

    mu_draws <- alpha_j + beta_j * d_scaled + eta_j

    tibble(
      stream = j,
      damDen = d_raw,
      NDVI_med = median(mu_draws),
      NDVI_lo  = quantile(mu_draws, 0.025),
      NDVI_hi  = quantile(mu_draws, 0.975)
    )
  })
})


stream_lookup <- aoi_HurdleData %>%
  distinct(stream, watershed) %>%
  arrange(stream)

m4_pdp_stream <- m4_pdp_stream %>%
  left_join(stream_lookup, by = "stream")



# plot
ggplot(m4_pdp_stream, aes(x = watershed, y = NDVI_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "Partial Dependence of NDVI on Stream and Dam Density",
    x = "Stream",
    y = "Predicted NDVI",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


###############################################################
################################################################

# build PDP grid
m4_pdp_year <- map_dfr(1:n_year, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m4_post$alpha
  
  beta_j  <- damDen_m4_post$beta_damDen
  
  eta_j   <- damDen_m4_post$eta_year[, j] + rowMeans(damDen_m4_post$eta_stream)

  map2_dfr(m4_dam_seq, m4_scaledDamDen_seq, function(d_raw, d_scaled) {

    mu_draws <- alpha_j + beta_j * d_scaled + eta_j

    tibble(
      year = j,
      damDen = d_raw,
      NDVI_med = median(mu_draws),
      NDVI_lo  = quantile(mu_draws, 0.025),
      NDVI_hi  = quantile(mu_draws, 0.975)
    )
  })
})

year_lookup <- tibble(
  year = 1:3,
  year_label = c(
    "2019",
    "2021",
    "2023"
  )
)


m4_pdp_year_named <- m4_pdp_year %>%
  left_join(year_lookup, by = "year")



# plot
ggplot(m4_pdp_year_named, aes(x = year_label, y = NDVI_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "Partial Dependence of NDVI on Year and Dam Density",
    x = "Year",
    y = "Predicted NDVI",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```




----

----








**Model Comparison**
```{r,echo=FALSE,warning=FALSE}

# Extract log-likelihood matrices
ll_m1 <- hglmm_m1_NDVI_damPres$sims.list$loglik
ll_m2 <- hglmm_m2_NDVI_damPres$sims.list$loglik
ll_m3 <- hglmm_m3_NDVI_damDen$sims.list$loglik
ll_m4 <- hglmm_m4_NDVI_damDen$sims.list$loglik

# Compute WAIC
waic_m1 <- waic(ll_m1)
waic_m2 <- waic(ll_m2)
waic_m3 <- waic(ll_m3)
waic_m4 <- waic(ll_m4)

# Combine into comparison table
waic_table <- tibble(
  model = c("m1", "m2", "m3", "m4"),
  WAIC  = c(waic_m1$estimates["waic", "Estimate"],
            waic_m2$estimates["waic", "Estimate"],
            waic_m3$estimates["waic", "Estimate"],
            waic_m4$estimates["waic", "Estimate"])
) %>%
  arrange(WAIC)

waic_table

```
