---
title: "VegAreaModels"
author: "Matt Webster"
date: "2026-07-28"
output: html_document
editor_options: 
  chunk_output_type: console
---



The purpose of this RMarkdown is to implement hierarchical generalized linear mixed effects models for VegArea ~ damPresence, & VegArea ~ damDensity (with and without geomorphic covariates).

4 models total.
    



```{r,echo=FALSE,}
suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(ggplot2)
  library(dataRetrieval)
  library(knitr)
  library(ggplot2)
  library(zoo) #for moving window
  library(stringr)
  library(vioplot)
  library(purrr)
  library(lavaan)
  library(blavaan)
  library(ggcorrplot)
  library(lme4)
  library(broom.mixed)
  library(coda)
  library(scales)
})
```


----


**Final Data:**

```{r,echo=FALSE}
aoi_HurdleData <- read.csv("aoi_HurdleData.csv")


# add year/event index
event_year_3_4 <- tapply(aoi_HurdleData$year_id,
                         aoi_HurdleData$sampleEvent_new,
                         unique)
event_year_3_4 <- as.numeric(event_year_3_4)


# Full Hurdle data
N <- nrow(aoi_HurdleData)
#y <- aoi_HurdleData$y
#y2 <- aoi_HurdleData$logSWA
y_veg <- aoi_HurdleData$y_veg
y_vegA <- aoi_HurdleData$VA
stream <- aoi_HurdleData$stream
year <- aoi_HurdleData$year_id
event <- aoi_HurdleData$sampleEvent_new
cluster <- aoi_HurdleData$clusterID_num
damPres <- aoi_HurdleData$damPresence
damDen <- as.numeric(aoi_HurdleData$dam_km_std)
damDen_raw <- aoi_HurdleData$dam_km

n_stream <- length(unique(aoi_HurdleData$stream))
n_year <- length(unique(aoi_HurdleData$year_id))
n_event <- length(unique(aoi_HurdleData$sampleEvent_new))
n_cluster <- length(unique(aoi_HurdleData$clusterID_num))

str(aoi_HurdleData)
```

----

    Right skew of vegetation area data warrants log-transformation prior to HGLMM.
    
    Because vegetation is almost always present in our sample reaches, a
    hurdle model is not necessary here -- no need to model the presence/absence 
    process if it's almost always present. 
      
      

```{r,echo=FALSE}
hist(aoi_HurdleData$VA)

hist(log(aoi_HurdleData$VA+1))
```


----


For surface water models, data of interest include:

     - y_vegA: log-transformed vegetation area log(VegArea[i]+1); response variable
    
    - stream: index for random intercept on VegArea and random slope on dam effect
    
    - year: index for random slope on dam effect
    
    - sampleEvent: index for random intercept on VegArea and random slope on dam effect 
      (nested within year; represents each 2 week sample period for Veg layer )
    
    - clusterID_seq: index for random intercept on VegArea; represents sample pairing
    
    - damPres: dam presence (0,1); fixed effect on VegArea
    
    - damDen: dams/km; z-standardized ("...$dam_km_std"); fixed effect on VegArea
    
    Only used in models 2 and 4:
    
    - incis: channel incision: 1-(TO/90) (inverse of topographic openness); z-standardized; fixed effect on VegArea
    
    - RIPA: riparian area (m^2); z-standardized; fixed effect on VegArea 
    
    - UDA: upstream drainage area (km ^2); z-standardized; fixed effect on VegArea


----

**HGLMM**


    This analysis explores late summer vegetation area in relation to beaver dam presence,
    & density.
    
    2 models for dam presence (1 without geomorphic covariates; 1 with geomorphic covariates)
    2 models for dam density (...^^^...)
    
    All 4 models presented herein are constructed in a hierarchical generalized linear mixed
    effects model framework with nested random slopes for beaver dam effect to allow
    for stream, year, and sampleEvent to cause deviation from the global mean beaver
    effect. Stream, sampleEvent, and sample clusterID are also included as random intercepts
    for baseline vegetation area. 
    
      - VA ~ damPresence (no geom.covar.)
      - VA ~ damPresence (geom.covar.)
      - VA ~ damDensity (no geom.covar.)
      - VA ~ damDensity (no geom.covar.)
    
    Model structure is paralleled with part 2 from the surface water hurdle models.
    
    Uses log(VegArea) for for 6 watersheds, 3 years, 4 sample periods (Aug-Sep) per year.
    
    
----

**Overall Takeaways**

    ...
    ...
    ...

----

----

----

----

----

----

Vegetation Area ~ Dam Presence
----

----

----

**Model 1: VegArea ~ DamPresence**

Effect of damPresence without geomorphic covariates

----

```{r,echo=FALSE}
#library(jagsUI)

    sink("HGLMM_m1_Veg_damPresence.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR VEG AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 0.7) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 0.7) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 0.7) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
      
      alpha ~ dnorm(0,0.04)
      beta_damPres ~ dnorm(0,0.04)
    
      ##########################################################################
      
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(0, tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
     
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Amount:
        y[i] ~ dnorm(mu[i], tau)
    
        mu[i] <- alpha
                    + (beta_damPres
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damPres[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]]
      }
    
    }
    ", fill = TRUE)
    
    sink()
```
    

```{r,echo=FALSE}
hglmm_m1_veg_damPres_data_list <- list(
  N = N,
  y = log(y_vegA + 1),
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damPres = damPres,
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hglmm_m1_veg_damPres_inits <- function(){
  list(
    alpha        = rnorm(1, 0, 1),
    sigma        = runif(1, 0.1, 2),

    sigma_beta_stream  = runif(1, 0.1, 2),
    sigma_beta_year    = runif(1, 0.1, 2),
    sigma_beta_event   = runif(1, 0.1, 2),
    sigma_eta_stream   = runif(1, 0.1, 2),
    sigma_eta_event    = runif(1, 0.1, 2),
    sigma_eta_cluster  = runif(1, 0.1, 2)
  )
}

hglmm_m1_veg_damPres_params <- c(
  "mu",
  "alpha",
  "beta_damPres",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_event",
  "sigma_eta_cluster"
)

nchains <- 3
nthin <- 5
niter <- 100000
nburn <- 25000

Sys.time()

hglmm_m1_veg_damPres <- jags(data = hglmm_m1_veg_damPres_data_list,
                   inits = hglmm_m1_veg_damPres_inits,
                   parameters.to.save = hglmm_m1_veg_damPres_params,
                   model.file = "HGLMM_m1_Veg_damPresence.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hglmm_m1_veg_damPres) # model with local and watershed dam density as fixed effects
#print(hglmm_m1_veg_damPres$summary[grep("beta_", rownames(hurdle_damPres_SW$summary)), ])

```


----
----

DIAGNOSTICS:

Extract posteriors:
```{r,echo=FALSE}
damPres_m1_veg_post <- hglmm_m1_veg_damPres$sims.list
```

"Chain health" Rhat + ESS:
```{r,echo=FALSE}
# Choose only the parameters you care about
params_m1_veg <- c(
  "alpha",
  "beta_damPres",
  "sigma")

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m1_veg_damPres$samples[, params_m1_veg], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m1_veg_damPres$samples[, params_m1_veg])

```

----

    Acceptable Rhat values, all around 1.00. Though beta_damPres is slightly eleveated, 
    1.02-1.05 is reasonable given the complexity of the nested random effects structure.
    
    Same for ESS. Not ideal for beta_damPres to be that low, but it is not a fatal issue.


----

**Check traceplots** for critical parameters:

```{r,echo=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

tplot_m1_veg_params <- c("alpha","beta_damPres","sigma", "sigma_beta_stream")


m1_mcmc_veg_sub <- hglmm_m1_veg_damPres$samples[, tplot_m1_veg_params]

coda::traceplot(m1_mcmc_veg_sub)
```

----

    Similarly, traceplots indicate solid convergence in all but beta_damPres which
    is slow to converge. Still should be useable

----

**Posterior predictive plots:**

Model-estimated vegArea vs. observed vegArea

Should be about 1

```{r,echo=FALSE}
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

# Posterior mean of predicted log(VA + 1)
m1_mu_veg_post <- apply(hglmm_m1_veg_damPres$sims.list$mu, 2, mean)

# Observed log vegetation area
m1_obs_logVA <- log(y_vegA + 1)

# Fit regression on the log scale
m1_fit_log <- lm(m1_obs_logVA ~ m1_mu_veg_post)

m1_slope_log <- coef(m1_fit_log)[2]
m1_R2_log    <- summary(m1_fit_log)$r.squared

# Plot
plot(m1_mu_veg_post,
     m1_obs_logVA,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated log(VA + 1)",
     ylab = "Observed log(VA + 1)")

abline(0, 1, col = "red", lwd = 2)

# Add slope and R² text
text(x = min(m1_mu_veg_post),
     y = max(m1_obs_logVA),
     labels = paste0("Slope = ", round(m1_slope_log, 3)),
     pos = 4)

#text(x = min(m1_mu_veg_post),
 #    y = max(m1_obs_logVA) - 0.3,
  #   labels = paste0("R² = ", round(m1_R2_log, 3)),
   #  pos = 4)

```

----

    Comparison of model-estimated log-transformed vegetation area with observed log-
    transformed vegetation area. "Perfect" model fit would have a slope = 1. A slope
    of 1.095 is reasonable, given model complexity. This indicates that our model
    slightly underestimates the observed variation in log(vegArea+1), but overall
    performs well.

----

**Random effects shrinkage: **

Should be centered around 0(ish)

1. beta_stream; random effect of stream on sw amount (1:6)
2. beta_event; random effect of sample event on sw amount (1:12; 4 sampleEvents per year)

```{r, echo=FALSE}

par(mfrow = c(2, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hglmm_m1_veg_damPres$sims.list$beta_stream)

boxplot(hglmm_m1_veg_damPres$sims.list$beta_event)
```

----

**Posterior mean dam effects:**

damPres -> vegArea 

```{r,echo=FALSE}
summary(hglmm_m1_veg_damPres$sims.list$beta_damPres)

exp(mean(hglmm_m1_veg_damPres$sims.list$beta_damPres))
```

    Dam presence increases log(vegetaion area+1) by ~23-27%. LOTS of variation. Strong
    overlap of 0.

----
----

Prior-posterior overlap:

```{r,echo=FALSE}
damPres_m1_veg_alpha_post <- hglmm_m1_veg_damPres$sims.list$alpha
damPres_m1_veg_beta_post  <- hglmm_m1_veg_damPres$sims.list$beta_damPres

damPres_m1_veg_Mod_n <- length(damPres_m1_veg_alpha_post)

damPres_m1_veg_alpha_prior <- rnorm(damPres_m1_veg_Mod_n, 0, sqrt(1/0.04))
damPres_m1_veg_beta_prior  <- rnorm(damPres_m1_veg_Mod_n, 0, sqrt(1/0.04))


# Plot overlap:

plot_prior_post <- function(prior, post, name) {
  
  df <- tibble(
    value = c(prior, post),
    dist  = rep(c("Prior", "Posterior"), each = length(prior))
  )
  
  ggplot(df, aes(x = value, fill = dist)) +
    geom_density(alpha = 0.4) +
    scale_fill_manual(values = c("gray60", "steelblue")) +
    labs(
      title = paste("Prior vs Posterior for", name),
      x = name,
      y = "Density"
    ) +
    theme_minimal(base_size = 14)
}

plot_prior_post(damPres_m1_veg_alpha_prior, damPres_m1_veg_alpha_post, "alpha")
plot_prior_post(damPres_m1_veg_beta_prior, damPres_m1_veg_beta_post, "beta_damPres")

```

----

    PPO indicates that the model is not being driven by posterior specifications

----
----

**Assess effect of dam presence on vegetation area**

```{r,echo=FALSE}
#library(tidyverse)
#library(scales)


#Posterior distribution of damEffect on vegArea:

hist(damPres_m1_veg_beta_post,
     breaks = 50,
     col = "steelblue",
     main = "Posterior of damPresence effect on log(VA + 1)",
     xlab = "beta_damPres")
abline(v = median(damPres_m1_veg_beta_post), col = "red", lwd = 2)
```

**Predicted VegArea with and without Dam Presence**

```{r,echo=FALSE}

m1_n_draws <- length(damPres_m1_veg_post$alpha)

# Random effects averaged out
m1_mu_no_dam <- damPres_m1_veg_post$alpha +
             damPres_m1_veg_post$eta_stream[ , stream] +
             damPres_m1_veg_post$eta_event[ , event] +
             damPres_m1_veg_post$eta_cluster[ , cluster]

m1_mu_yes_dam <- damPres_m1_veg_post$alpha +
              damPres_m1_veg_post$beta_damPres +
              damPres_m1_veg_post$eta_stream[ , stream] +
              damPres_m1_veg_post$eta_event[ , event] +
              damPres_m1_veg_post$eta_cluster[ , cluster]


m1_VA_no_dam  <- exp(m1_mu_no_dam) - 1
m1_VA_yes_dam <- exp(m1_mu_yes_dam) - 1


m1_VA_no_dam_vec  <- as.numeric(m1_VA_no_dam)
m1_VA_yes_dam_vec <- as.numeric(m1_VA_yes_dam)

boxplot(
  list(
    `No Dam` = m1_VA_no_dam_vec,
    `Dam Present` = m1_VA_yes_dam_vec
  ),
  col = c("steelblue", "brown4"),
  main = "Predicted Vegetation Area by Dam Presence",
  ylab = "Vegetation Area (m²)"
)

```

**Marginal Effect:**

```{r,echo=FALSE}
delta <- VA_yes_dam - VA_no_dam

hist(delta,
     breaks = 50,
     col = "darkgreen",
     main = "Posterior of Dam Effect on Vegetation Area",
     xlab = "Difference in veg area (m²)")
abline(v = median(delta), col = "red", lwd = 2)


```
