---
title: "VegAreaModels"
author: "Matt Webster"
date: "2026-07-28"
output: html_document
editor_options: 
  chunk_output_type: console
---



The purpose of this RMarkdown is to implement hierarchical generalized linear mixed effects models for VegArea ~ damPresence, & VegArea ~ damDensity (with and without geomorphic covariates).

4 models total.
    



```{r,echo=FALSE,}
suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(ggplot2)
  library(dataRetrieval)
  library(knitr)
  library(ggplot2)
  library(zoo) #for moving window
  library(stringr)
  library(vioplot)
  library(purrr)
  library(lavaan)
  library(blavaan)
  library(ggcorrplot)
  library(lme4)
  library(broom.mixed)
  library(coda)
  library(scales)
})
```


----


**Final Data:**

```{r,echo=FALSE}
aoi_HurdleData <- read.csv("aoi_HurdleData.csv")


# add year/event index
event_year_3_4 <- tapply(aoi_HurdleData$year_id,
                         aoi_HurdleData$sampleEvent_new,
                         unique)
event_year_3_4 <- as.numeric(event_year_3_4)


# Full Hurdle data
N <- nrow(aoi_HurdleData)
#y <- aoi_HurdleData$y
#y2 <- aoi_HurdleData$logSWA
y_veg <- aoi_HurdleData$y_veg
y_vegA <- aoi_HurdleData$VA
stream <- aoi_HurdleData$stream
year <- aoi_HurdleData$year_id
event <- aoi_HurdleData$sampleEvent_new
cluster <- aoi_HurdleData$clusterID_num
damPres <- aoi_HurdleData$damPresence
damDen <- as.numeric(aoi_HurdleData$dam_km_std)
damDen_raw <- aoi_HurdleData$dam_km

n_stream <- length(unique(aoi_HurdleData$stream))
n_year <- length(unique(aoi_HurdleData$year_id))
n_event <- length(unique(aoi_HurdleData$sampleEvent_new))
n_cluster <- length(unique(aoi_HurdleData$clusterID_num))

str(aoi_HurdleData)
```

----

    Right skew of vegetation area data warrants log-transformation prior to HGLMM.
    
    Because vegetation is almost always present in our sample reaches, a
    hurdle model is not necessary here -- no need to model the presence/absence 
    process if it's almost always present. 
      
      

```{r,echo=FALSE}
hist(aoi_HurdleData$VA)

hist(log(aoi_HurdleData$VA+1))
```


----


For surface water models, data of interest include:

     - y_vegA: log-transformed vegetation area log(VegArea[i]+1); response variable
    
    - stream: index for random intercept on VegArea and random slope on dam effect
    
    - year: index for random slope on dam effect
    
    - sampleEvent: index for random intercept on VegArea and random slope on dam effect 
      (nested within year; represents each 2 week sample period for Veg layer )
    
    - clusterID_seq: index for random intercept on VegArea; represents sample pairing
    
    - damPres: dam presence (0,1); fixed effect on VegArea
    
    - damDen: dams/km; z-standardized ("...$dam_km_std"); fixed effect on VegArea
    
    Only used in models 2 and 4:
    
    - incis: channel incision: 1-(TO/90) (inverse of topographic openness); z-standardized; fixed effect on VegArea
    
    - RIPA: riparian area (m^2); z-standardized; fixed effect on VegArea 
    
    - UDA: upstream drainage area (km ^2); z-standardized; fixed effect on VegArea


----

**HGLMM**


    This analysis explores late summer vegetation area in relation to beaver dam presence,
    & density.
    
    2 models for dam presence (1 without geomorphic covariates; 1 with geomorphic covariates)
    2 models for dam density (...^^^...)
    
    All 4 models presented herein are constructed in a hierarchical generalized linear mixed
    effects model framework with nested random slopes for beaver dam effect to allow
    for stream, year, and sampleEvent to cause deviation from the global mean beaver
    effect. Stream, sampleEvent, and sample clusterID are also included as random intercepts
    for baseline vegetation area. 
    
      - VA ~ damPresence (no geom.covar.)
      - VA ~ damPresence (geom.covar.)
      - VA ~ damDensity (no geom.covar.)
      - VA ~ damDensity (no geom.covar.)
    
    Model structure is paralleled with part 2 from the surface water hurdle models.
    
    Uses log(VegArea) for for 6 watersheds, 3 years, 4 sample periods (Aug-Sep) per year.
    
    
----

**Overall Takeaways**

    ...
    ...
    ...

----

----

----

----

----

----

Vegetation Area ~ Dam Presence
----

----

----

**Model 1: VegArea ~ DamPresence**

Effect of damPresence without geomorphic covariates

----

```{r,echo=FALSE}
library(jagsUI)

    sink("HGLMM_m1_Veg_damPresence.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR VEG AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 0.7) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 0.7) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 0.7) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
      
      alpha ~ dnorm(0,0.04)
      beta_damPres ~ dnorm(0,0.04)
    
      ##########################################################################
      
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(0, tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
     
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Amount:
        y[i] ~ dnorm(mu[i], tau)
    
        mu[i] <- alpha
                    + (beta_damPres
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damPres[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]]
      }
    
    }
    ", fill = TRUE)
    
    sink()
```
    

```{r,echo=FALSE}
hglmm_m1_veg_damPres_data_list <- list(
  N = N,
  y = log(y_vegA + 1),
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damPres = damPres,
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hglmm_m1_veg_damPres_inits <- function(){
  list(
    alpha        = rnorm(1, 0, 1),
    sigma        = runif(1, 0.1, 2),

    sigma_beta_stream  = runif(1, 0.1, 2),
    sigma_beta_year    = runif(1, 0.1, 2),
    sigma_beta_event   = runif(1, 0.1, 2),
    sigma_eta_stream   = runif(1, 0.1, 2),
    sigma_eta_event    = runif(1, 0.1, 2),
    sigma_eta_cluster  = runif(1, 0.1, 2)
  )
}

hglmm_m1_veg_damPres_params <- c(
  "mu",
  "alpha",
  "beta_damPres",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_event",
  "sigma_eta_cluster"
)

nchains <- 3
nthin <- 5
niter <- 100000
nburn <- 25000

Sys.time()

set.seed(092323)

hglmm_m1_veg_damPres <- jags(data = hglmm_m1_veg_damPres_data_list,
                   inits = hglmm_m1_veg_damPres_inits,
                   parameters.to.save = hglmm_m1_veg_damPres_params,
                   model.file = "HGLMM_m1_Veg_damPresence.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hglmm_m1_veg_damPres) # model with local and watershed dam density as fixed effects
#print(hglmm_m1_veg_damPres$summary[grep("beta_", rownames(hurdle_damPres_SW$summary)), ])

```


----
----

DIAGNOSTICS:

Extract posteriors:
```{r,echo=FALSE}
damPres_m1_veg_post <- hglmm_m1_veg_damPres$sims.list
```

"Chain health" Rhat + ESS:
```{r,echo=FALSE}
# Choose only the parameters you care about
params_m1_veg <- c(
  "alpha",
  "beta_damPres",
  "sigma")

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m1_veg_damPres$samples[, params_m1_veg], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m1_veg_damPres$samples[, params_m1_veg])

```

----

    Acceptable Rhat values, all around 1.00. Though beta_damPres is slightly eleveated, 
    up to 1.05 or so is reasonable given the complexity of the nested random effects structure.
    
    Same for ESS. Not ideal for beta_damPres to be that low, but it is not a fatal issue.


----

**Check traceplots** for critical parameters:

```{r,echo=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

tplot_m1_veg_params <- c("alpha","beta_damPres","sigma", "sigma_beta_stream")


m1_mcmc_veg_sub <- hglmm_m1_veg_damPres$samples[, tplot_m1_veg_params]

coda::traceplot(m1_mcmc_veg_sub)
```

----

    Similarly, traceplots indicate solid convergence in all but beta_damPres which
    is slow to converge. Still should be useable

----

**Posterior predictive plots:**

Model-estimated vegArea vs. observed vegArea

Should be about 1

```{r,echo=FALSE}
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

# Posterior mean of predicted log(VA + 1)
m1_mu_veg_post <- apply(hglmm_m1_veg_damPres$sims.list$mu, 2, mean)

# Observed log vegetation area
m1_obs_logVA <- log(y_vegA + 1)

# Fit regression on the log scale
m1_fit_log <- lm(m1_obs_logVA ~ m1_mu_veg_post)

m1_slope_log <- coef(m1_fit_log)[2]
m1_R2_log    <- summary(m1_fit_log)$r.squared

# Plot
plot(m1_mu_veg_post,
     m1_obs_logVA,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated log(VA + 1)",
     ylab = "Observed log(VA + 1)")

abline(0, 1, col = "red", lwd = 2)

# Add slope and R² text
text(x = min(m1_mu_veg_post),
     y = max(m1_obs_logVA),
     labels = paste0("Slope = ", round(m1_slope_log, 3)),
     pos = 4)

#text(x = min(m1_mu_veg_post),
 #    y = max(m1_obs_logVA) - 0.3,
  #   labels = paste0("R² = ", round(m1_R2_log, 3)),
   #  pos = 4)

```

----

    Comparison of model-estimated log-transformed vegetation area with observed log-
    transformed vegetation area. "Perfect" model fit would have a slope = 1. A slope
    of 1.09 is reasonable, given model complexity. This indicates that our model
    slightly underestimates the observed variation in log(vegArea+1), but overall
    performs well.

----

**Random effects shrinkage: **

Should be centered around 0(ish)

1. beta_stream; random effect of stream on sw amount (1:6)
2. beta_event; random effect of sample event on sw amount (1:12; 4 sampleEvents per year)

```{r, echo=FALSE}

par(mfrow = c(2, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hglmm_m1_veg_damPres$sims.list$beta_stream)

boxplot(hglmm_m1_veg_damPres$sims.list$beta_event)
```

----

**Posterior mean dam effects:**

damPres -> vegArea (no geom.covar.)

```{r,echo=FALSE}
m1_coeffparams <- c(#"alpha",
            "beta_damPres",
            "sigma")

m1_coef_df <- hglmm_m1_veg_damPres$summary[m1_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m1_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.2,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```

```{r,echo=FALSE}
summary(hglmm_m1_veg_damPres$sims.list$beta_damPres)

exp(mean(hglmm_m1_veg_damPres$sims.list$beta_damPres))
```

    Without consideration of geomorphic covariates, dam presence is associated with a
    ~23-27% increase in vegetation area. LOTS of variation. Strong overlap of 0. MARGINAL
    influence at best.

----
----

Prior-posterior overlap:

```{r,echo=FALSE}
damPres_m1_veg_alpha_post <- hglmm_m1_veg_damPres$sims.list$alpha
damPres_m1_veg_beta_post  <- hglmm_m1_veg_damPres$sims.list$beta_damPres

damPres_m1_veg_Mod_n <- length(damPres_m1_veg_alpha_post)

damPres_m1_veg_alpha_prior <- rnorm(damPres_m1_veg_Mod_n, 0, sqrt(1/0.04))
damPres_m1_veg_beta_prior  <- rnorm(damPres_m1_veg_Mod_n, 0, sqrt(1/0.04))


# Plot overlap:

plot_prior_post <- function(prior, post, name) {
  
  df <- tibble(
    value = c(prior, post),
    dist  = rep(c("Prior", "Posterior"), each = length(prior))
  )
  
  ggplot(df, aes(x = value, fill = dist)) +
    geom_density(alpha = 0.4) +
    scale_fill_manual(values = c("gray60", "steelblue")) +
    labs(
      title = paste("Prior vs Posterior for", name),
      x = name,
      y = "Density"
    ) +
    theme_minimal(base_size = 14)
}

plot_prior_post(damPres_m1_veg_alpha_prior, damPres_m1_veg_alpha_post, "alpha")
plot_prior_post(damPres_m1_veg_beta_prior, damPres_m1_veg_beta_post, "beta_damPres")

```

----

    PPO indicates that the model is not being driven by posterior specifications

----
----

**Assess effect of dam presence on vegetation area**

```{r,echo=FALSE}
#library(tidyverse)
#library(scales)


#Posterior distribution of damEffect on vegArea:

hist(damPres_m1_veg_beta_post,
     breaks = 50,
     col = "steelblue",
     main = "Posterior of damPresence effect on log(VA + 1)",
     xlab = "beta_damPres")
abline(v = median(damPres_m1_veg_beta_post), col = "red", lwd = 2)
```

**Predicted VegArea with and without Dam Presence**


```{r,echo=FALSE}

m1_n_draws <- length(damPres_m1_veg_post$alpha)

# Random effects averaged out
m1_mu_no_dam <- damPres_m1_veg_post$alpha +
             damPres_m1_veg_post$eta_stream[ , stream] +
             damPres_m1_veg_post$eta_event[ , event] +
             damPres_m1_veg_post$eta_cluster[ , cluster]

m1_mu_yes_dam <- damPres_m1_veg_post$alpha +
              damPres_m1_veg_post$beta_damPres +
              damPres_m1_veg_post$eta_stream[ , stream] +
              damPres_m1_veg_post$eta_event[ , event] +
              damPres_m1_veg_post$eta_cluster[ , cluster]


m1_VA_no_dam  <- exp(m1_mu_no_dam) - 1
m1_VA_yes_dam <- exp(m1_mu_yes_dam) - 1


m1_VA_no_dam_vec  <- as.numeric(m1_VA_no_dam)
m1_VA_yes_dam_vec <- as.numeric(m1_VA_yes_dam)


m1_VA_no_dam_med  <- apply(m1_VA_no_dam, 2, median)
m1_VA_yes_dam_med <- apply(m1_VA_yes_dam, 2, median)

m1_VA_df <- tibble(
  VA = c(m1_VA_no_dam_med, m1_VA_yes_dam_med),
  Dam = factor(rep(c("No Dam", "Dam Present"),
                   each = length(m1_VA_no_dam_med)))
)


ggplot(m1_VA_df, aes(x = Dam, y = VA, fill = Dam)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8) +
  scale_fill_manual(values = c("steelblue", "brown4")) +
  labs(
    title = "Predicted Vegetation Area by Dam Presence (Site Medians)",
    y = "Vegetation Area (m²)",
    x = ""
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")
```


----

----

----

----

----

----

----

**Model 1: VegArea ~ DamPresence**

Effect of damPresence WITH geomorphic covariates

----

```{r,echo=FALSE}
#library(jagsUI)

    sink("HGLMM_m2_Veg_damPresence.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR VEG AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 0.7) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 0.7) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 0.7) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
      
      alpha ~ dnorm(0,0.04)
      beta_damPres ~ dnorm(0,0.04)
      beta_incis ~ dnorm(0,0.04)
      beta_RIPA ~ dnorm(0,0.04)
      beta_UDA ~ dnorm(0,0.04)
    
      ##########################################################################
      
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(0, tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
     
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Amount:
        y[i] ~ dnorm(mu[i], tau)
    
        mu[i] <- alpha
                    + (beta_damPres
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damPres[i]
                    + beta_incis * incis[i] 
                    + beta_RIPA * RIPA[i]
                    + beta_UDA * UDA[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]]
      }
    
    }
    ", fill = TRUE)
    
    sink()
```
    

```{r,echo=FALSE}
hglmm_m2_veg_damPres_data_list <- list(
  N = N,
  y = log(y_vegA + 1),
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damPres = damPres,
  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)),
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hglmm_m2_veg_damPres_inits <- function(){
  list(
    alpha        = rnorm(1, 0, 1),
    sigma        = runif(1, 0.1, 2),

    sigma_beta_stream  = runif(1, 0.1, 2),
    sigma_beta_year    = runif(1, 0.1, 2),
    sigma_beta_event   = runif(1, 0.1, 2),
    sigma_eta_stream   = runif(1, 0.1, 2),
    sigma_eta_event    = runif(1, 0.1, 2),
    sigma_eta_cluster  = runif(1, 0.1, 2)
  )
}

hglmm_m2_veg_damPres_params <- c(
  "mu",
  "alpha",
  "beta_damPres",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_event",
  "sigma_eta_cluster"
)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000

Sys.time()

set.seed(092323)

hglmm_m2_veg_damPres <- jags(data = hglmm_m2_veg_damPres_data_list,
                   inits = hglmm_m2_veg_damPres_inits,
                   parameters.to.save = hglmm_m2_veg_damPres_params,
                   model.file = "HGLMM_m2_Veg_damPresence.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hglmm_m1_veg_damPres) # model with local and watershed dam density as fixed effects
#print(hglmm_m1_veg_damPres$summary[grep("beta_", rownames(hurdle_damPres_SW$summary)), ])

```


----
----

DIAGNOSTICS:

Extract posteriors:
```{r,echo=FALSE}
damPres_m2_veg_post <- hglmm_m2_veg_damPres$sims.list
```

"Chain health" Rhat + ESS:
```{r,echo=FALSE}
# Choose only the parameters you care about
params_m2_veg <- c(
  "alpha",
  "beta_damPres",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "sigma")

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m2_veg_damPres$samples[, params_m2_veg], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m2_veg_damPres$samples[, params_m2_veg])

```

----

    ...

----

**Check traceplots** for critical parameters:

```{r,echo=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

tplot_m2_veg_params <- c("alpha","beta_damPres",
                         "beta_incis","beta_RIPA",
                         "beta_UDA", "sigma", "sigma_beta_stream")


m2_mcmc_veg_sub <- hglmm_m2_veg_damPres$samples[, tplot_m2_veg_params]

coda::traceplot(m2_mcmc_veg_sub)
```

----

    Hm.. not great for beta_damPres... what to do...
    

----

**Posterior predictive plots:**

Model-estimated vegArea vs. observed vegArea

Should be about 1

```{r,echo=FALSE}
par(mfrow = c(1,1),        # 1 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  

# Posterior mean of predicted log(VA + 1)
m2_mu_veg_post <- apply(hglmm_m2_veg_damPres$sims.list$mu, 2, mean)

# Observed log vegetation area
m2_obs_logVA <- log(y_vegA + 1)

# Fit regression on the log scale
m2_fit_log <- lm(m2_obs_logVA ~ m2_mu_veg_post)

m2_slope_log <- coef(m2_fit_log)[2]
m2_R2_log    <- summary(m2_fit_log)$r.squared

# Plot
plot(m2_mu_veg_post,
     m2_obs_logVA,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated log(VA + 1)",
     ylab = "Observed log(VA + 1)")

abline(0, 1, col = "red", lwd = 2)

# Add slope and R² text
text(x = min(m2_mu_veg_post),
     y = max(m2_obs_logVA),
     labels = paste0("Slope = ", round(m2_slope_log, 3)),
     pos = 4)

#text(x = min(m2_mu_veg_post),
 #    y = max(m2_obs_logVA) - 0.3,
  #   labels = paste0("R² = ", round(m2_R2_log, 3)),
   #  pos = 4)

```

----

    Comparison of model-estimated log-transformed vegetation area with observed log-
    transformed vegetation area. "Perfect" model fit would have a slope = 1. A slope
    of 1.09 is reasonable, given model complexity. This indicates that our model
    slightly underestimates the observed variation in log(vegArea+1), but overall
    performs well. Highest disagreement at lower veg areas.

----

**Random effects shrinkage: **

Should be centered around 0(ish)

1. beta_stream; random effect of stream on sw amount (1:6)
2. beta_event; random effect of sample event on sw amount (1:12; 4 sampleEvents per year)

```{r, echo=FALSE}

par(mfrow = c(2, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hglmm_m2_veg_damPres$sims.list$beta_stream)

boxplot(hglmm_m2_veg_damPres$sims.list$beta_event)
```

----

**Posterior mean dam effects:**

damPres -> vegArea (w/geom.covar.)


```{r,echo=FALSE}
m2_coeffparams <- c(#"alpha",
            "beta_damPres",
            "beta_incis",
            "beta_RIPA",
            "beta_UDA",
            "sigma")

m2_coef_df <- hglmm_m2_veg_damPres$summary[m2_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m2_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.2,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```


```{r,echo=FALSE}
summary(hglmm_m2_veg_damPres$sims.list$beta_damPres)

exp(mean(hglmm_m2_veg_damPres$sims.list$beta_damPres))
```

    When you also consider site-level geomorphic conditions, dam presence is associated 
    with an estimated 15-20% increase in vegetation area. More conservative than M1. 
    Still, LOTS of variation, & strongoverlap of 0. Modest positive influence.

----
----

Prior-posterior overlap:

```{r,echo=FALSE}
damPres_m2_veg_alpha_post <- hglmm_m2_veg_damPres$sims.list$alpha
damPres_m2_veg_beta_post  <- hglmm_m2_veg_damPres$sims.list$beta_damPres
damPres_m2_veg_sigmaBetaStream_post <- hglmm_m2_veg_damPres$sims.list$sigma_beta_stream

damPres_m2_veg_Mod_n <- length(damPres_m2_veg_alpha_post)

damPres_m2_veg_alpha_prior <- rnorm(damPres_m2_veg_Mod_n, 0, sqrt(1/0.04))
damPres_m2_veg_beta_prior  <- rnorm(damPres_m2_veg_Mod_n, 0, sqrt(1/0.04))
damPres_m2_veg_sigmaBetaStream_prior <- pmax(rnorm(damPres_m2_veg_Mod_n, 0, 1), 0)


# Plot overlap:

#plot_prior_post <- function(prior, post, name) {
  
#  df <- tibble(
#    value = c(prior, post),
#    dist  = rep(c("Prior", "Posterior"), each = length(prior))
#  )
  
#  ggplot(df, aes(x = value, fill = dist)) +
#    geom_density(alpha = 0.4) +
#    scale_fill_manual(values = c("gray60", "steelblue")) +
#    labs(
#      title = paste("Prior vs Posterior for", name),
#      x = name,
#      y = "Density"
#    ) +
#    theme_minimal(base_size = 14)
#}

plot_prior_post(damPres_m2_veg_alpha_prior, damPres_m2_veg_alpha_post, "alpha")
plot_prior_post(damPres_m2_veg_beta_prior, damPres_m2_veg_beta_post, "beta_damPres")
plot_prior_post(damPres_m2_veg_sigmaBetaStream_prior, damPres_m2_veg_sigmaBetaStream_post,"sigmaBetaStream")

```

----

    PPOs still look pretty good. 

----
----

**Assess effect of dam presence on vegetation area**

```{r,echo=FALSE}
#library(tidyverse)
#library(scales)

par(mfrow = c(1, 1),        # 2 rows, 1 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)  
#Posterior distribution of damEffect on vegArea:

hist(damPres_m2_veg_beta_post,
     breaks = 50,
     col = "steelblue",
     main = "Posterior distribution of damPres effect on log(VA + 1)",
     xlab = "beta_damPres",
     xlim = c(-1,1))
# Add median line
abline(v = median(damPres_m2_veg_beta_post), col = "red", lwd = 2)

```

**Predicted VegArea with and without Dam Presence**


```{r,echo=FALSE}

m2_n_draws <- length(damPres_m2_veg_post$alpha)

# Random effects averaged out
m2_mu_no_dam <- damPres_m2_veg_post$alpha +
             damPres_m2_veg_post$eta_stream[ , stream] +
             damPres_m2_veg_post$eta_event[ , event] +
             damPres_m2_veg_post$eta_cluster[ , cluster] +
             damPres_m2_veg_post$beta_incis +
             damPres_m2_veg_post$beta_RIPA +
             damPres_m2_veg_post$beta_UDA

m2_mu_yes_dam <- damPres_m2_veg_post$alpha +
              damPres_m2_veg_post$beta_damPres +
              damPres_m2_veg_post$eta_stream[ , stream] +
              damPres_m2_veg_post$eta_event[ , event] +
              damPres_m2_veg_post$eta_cluster[ , cluster] +
              damPres_m2_veg_post$beta_incis +
              damPres_m2_veg_post$beta_RIPA +
              damPres_m2_veg_post$beta_UDA


m2_VA_no_dam  <- exp(m2_mu_no_dam) - 1
m2_VA_yes_dam <- exp(m2_mu_yes_dam) - 1


m2_VA_no_dam_vec  <- as.numeric(m2_VA_no_dam)
m2_VA_yes_dam_vec <- as.numeric(m2_VA_yes_dam)


m2_VA_no_dam_med  <- apply(m2_VA_no_dam, 2, median)
m2_VA_yes_dam_med <- apply(m2_VA_yes_dam, 2, median)

m2_VA_df <- tibble(
  VA = c(m2_VA_no_dam_med, m2_VA_yes_dam_med),
  Dam = factor(rep(c("No Dam", "Dam Present"),
                   each = length(m2_VA_no_dam_med)))
)


ggplot(m2_VA_df, aes(x = Dam, y = VA, fill = Dam)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, alpha = 0.8) +
  scale_fill_manual(values = c("steelblue", "brown4")) +
  labs(
    title = "Predicted Vegetation Area by Dam Presence (Site Medians)",
    y = "Vegetation Area (m²)",
    x = ""
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")
```

----

----

----

----


**Partial dependence plots**

Constrain PDP to empirical ranges for VA and target covariates

```{r,echo=FALSE}

names(damPres_m2_veg_post)

# Raw covariates
incis_raw <- 1 - aoi_HurdleData$TO/90
RIPA_raw  <- aoi_HurdleData$riparian_area_m2
UDA_raw   <- aoi_HurdleData$catchment_km2

logSWA_mean <- mean(aoi_HurdleData$logSWA)
logSWA_sd   <- sd(aoi_HurdleData$logSWA)

# Standardization stats
incis_mean_raw <- mean(incis_raw)
incis_sd_raw   <- sd(incis_raw)

RIPA_mean_raw  <- mean(RIPA_raw)
RIPA_sd_raw    <- sd(RIPA_raw)

UDA_mean_raw   <- mean(UDA_raw)
UDA_sd_raw     <- sd(UDA_raw)


##############################3

# Helper Functions to Compute SWA from posterior draws
compute_swa_draws <- function(incis, RIPA, UDA, post, dam_flag) {
  n_draws <- length(damPres_m2_veg_post$alpha)
  swa_mat <- matrix(NA_real_, nrow = length(incis), ncol = n_draws)

  for (j in seq_len(n_draws)) {
    lp <- damPres_m2_veg_post$alpha[j] +
          damPres_m2_veg_post$beta_incis[j] * incis +
          damPres_m2_veg_post$beta_RIPA[j]  * RIPA +
          damPres_m2_veg_post$beta_UDA[j]   * UDA +
          damPres_m2_veg_post$beta_damPres[j] * dam_flag

    swa_mat[, j] <- exp(lp) - 1
  }

  swa_mat
}

summarize_swa <- function(swa_mat) {
  tibble(
    SWA_med = apply(swa_mat, 1, median),
    SWA_lo  = apply(swa_mat, 1, quantile, probs = 0.025),
    SWA_hi  = apply(swa_mat, 1, quantile, probs = 0.975)
  )
}




#########################################

# PDP Builder:

build_pdp <- function(focal_raw, focal_name,
                      incis_raw, RIPA_raw, UDA_raw,
                      post,
                      incis_mean_raw, incis_sd_raw,
                      RIPA_mean_raw, RIPA_sd_raw,
                      UDA_mean_raw, UDA_sd_raw) {

  # Sequence only across empirical range
  focal_seq <- seq(min(focal_raw), max(focal_raw), length.out = 200)

  # Conditional medians of other covariates
  incis_med <- median(incis_raw)
  RIPA_med  <- median(RIPA_raw)
  UDA_med   <- median(UDA_raw)

  # Build grid (200 rows)
  grid <- tibble(
    focal_raw = focal_seq,
    incis_raw = if (focal_name == "incis") focal_seq else incis_med,
    RIPA_raw  = if (focal_name == "RIPA")  focal_seq else RIPA_med,
    UDA_raw   = if (focal_name == "UDA")   focal_seq else UDA_med
  ) %>%
    mutate(
      incis = (incis_raw - incis_mean_raw) / incis_sd_raw,
      RIPA  = (RIPA_raw  - RIPA_mean_raw)  / RIPA_sd_raw,
      UDA   = (UDA_raw   - UDA_mean_raw)   / UDA_sd_raw
    )

  # Compute SWA draws
  swa_nodam <- compute_swa_draws(grid$incis, grid$RIPA, grid$UDA, post, dam_flag = 0)
  swa_dam   <- compute_swa_draws(grid$incis, grid$RIPA, grid$UDA, post, dam_flag = 1)

  # Summaries (400 rows)
  pdp <- bind_rows(
    summarize_swa(swa_nodam) %>% mutate(dam = "No Dam"),
    summarize_swa(swa_dam)   %>% mutate(dam = "Dam Present")
  )

  # Repeat grid twice (400 rows)
  grid2 <- bind_rows(grid, grid)

  # Bind columns safely
  bind_cols(grid2, pdp)
}



#####

m2_pdp_incis <- build_pdp(
  focal_raw = incis_raw,
  focal_name = "incis",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_m2_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)

m2_pdp_RIPA <- build_pdp(
  focal_raw = RIPA_raw,
  focal_name = "RIPA",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_m2_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)

m2_pdp_UDA <- build_pdp(
  focal_raw = UDA_raw,
  focal_name = "UDA",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_m2_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)



##########################################

# Plot Function:

plot_pdp <- function(pdp, xvar, xtitle) {
  ggplot(pdp,
         aes(x = .data[[xvar]], y = SWA_med, color = dam, fill = dam)) +
    geom_ribbon(aes(ymin = SWA_lo, ymax = SWA_hi), alpha = 0.2, color = NA) +
    geom_line(linewidth = 1.2) +   # updated from size = 1.2
    scale_color_manual(values = c("steelblue", "brown4")) +
    scale_fill_manual(values = c("steelblue", "brown4")) +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = paste("Partial Dependence of SWA on", xtitle),
      subtitle = "Predicted VegArea with 95% credible intervals\nConstrained to empirical covariate range",
      x = xtitle,
      y = "Predicted Vegetation Area (m^2)",
      color = "Dam Presence",
      fill  = "Dam Presence"
    ) +
    theme_minimal(base_size = 14)
}



###########################################3

plot_pdp(m2_pdp_incis, "incis_raw", "Channel Incision (raw)")
plot_pdp(m2_pdp_RIPA,  "RIPA_raw",  "Riparian Area (m²)")
plot_pdp(m2_pdp_UDA,   "UDA_raw",   "Contributing Drainage Area (km²)")


```


----

----

----

----

----

----

----

----

----



VegArea ~ Dam Density
----


**Hurdle HGLMM**


**MODEL 3: VegArea ~ damDensity (no Geom.Covar.)**

```{r,echo=FALSE}
#library(jagsUI)

    sink("HGLMM_m3_Veg_damDensity.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 4) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 4) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 4) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Amount:
      alpha ~ dnorm(0,0.04)
      beta_damDen ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR Veg AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(0, tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Presence/absence:
        
        y[i] ~ dnorm(mu[i], tau)
    
        mu[i] <- alpha
                    + (beta_damDen
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damDen[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]]      
      }
    }
    ", fill = TRUE)
    
    sink()
```  
    

```{r,echo=FALSE}
hglmm_damDen_m3_data_list <- list(
  N = N,
  y = log(y_vegA + 1),
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damDen = damDen,
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hglmm_damDen_m3_inits <- function(){
  list(
    alpha        = rnorm(1, 0, 1),
    epsilon      = rnorm(1, 0, 1),

    sigma        = runif(1, 0.1, 2),
    sigma_zeta_stream  = runif(1, 0.1, 2),
    sigma_zeta_year    = runif(1, 0.1, 2),
    sigma_zeta_event   = runif(1, 0.1, 2),
    sigma_iota_stream  = runif(1, 0.1, 2),
    sigma_iota_event   = runif(1, 0.1, 2),
    sigma_iota_cluster = runif(1, 0.1, 2),

    sigma_beta_stream  = runif(1, 0.1, 2),
    sigma_beta_year    = runif(1, 0.1, 2),
    sigma_beta_event   = runif(1, 0.1, 2),
    sigma_eta_stream   = runif(1, 0.1, 2),
    sigma_eta_event    = runif(1, 0.1, 2),
    sigma_eta_cluster  = runif(1, 0.1, 2)
  )
}

hglmm_damDen_m3_params <- c(
  "alpha",
  "mu",
  "beta_damDen",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_event",
  "sigma_eta_cluster"
)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000


Sys.time()

set.seed(092323)

hglmm_m3_damDen_Veg <- jags(data = hglmm_damDen_m3_data_list,
                   inits = hglmm_damDen_m3_inits,
                   parameters.to.save = hglmm_damDen_m3_params,
                   model.file = "HGLMM_m3_Veg_damDensity.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hurdle_damDen_m3_SW) # model with local and watershed dam density as fixed effects
#print(hurdle_damDen_m3_SW$summary[grep("beta_", rownames(hurdle_damDen_m3_SW$summary)), ])
#print(hurdle_damDen_m3_SW$summary[grep("zeta_", rownames(hurdle_damDen_m3_SW$summary)), ])

```


----


DIAGNOSTICS:
----


```{r,echo=FALSE}
# Extract posterior draws

damDen_m3_post <- hglmm_m3_damDen_Veg$sims.list 
```

**Assess "Chain health" Rhat + ESS:**

```{r,echo=FALSE} 
# Choose only the parameters you care about
params_m3 <- c(
  "alpha",
  "beta_damDen",
  "sigma")

# Gelman–Rubin convergence diagnostic
gelman.diag(hglmm_m3_damDen_Veg$samples[, params_m3], multivariate = FALSE)

# Effective sample size
effectiveSize(hglmm_m3_damDen_Veg$samples[, params_m3])

```

    Rhat & Effective Sample Size:
    
    Rhat values look good. All near 1.00. Beta_damDen is slower to converge, but
    still does so nicely.
    
    Effective sample sizes look ok for all parameters (note beta_damDen). 

----

**Check traceplots** for critical parameters:

```{r,echo=FALSE} 

par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


tplot_params_m3 <- c("alpha","beta_damDen","sigma","sigma_beta_stream")


mcmc_sub_m3 <- hglmm_m3_damDen_Veg$samples[, tplot_params_m3]

coda::traceplot(mcmc_sub_m3)
```


    Traceplots:
    
    Traceplots all look sufficiently "grassy," indicating sufficient chain convergence. Though
    some weirdness with beta_damDen towards the end..?


----

Predicted log(SWA) vs observed log(SWA), conditional on SW presence

Goal is 1:1 (ish)

```{r,echo=FALSE} 
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 

# Posterior mean of predicted log(VA + 1)
m3_mu_veg_post <- apply(hglmm_m3_damDen_Veg$sims.list$mu, 2, mean)

# Observed log vegetation area
m3_obs_logVA <- log(y_vegA + 1)

# Fit regression on the log scale
m3_fit_log <- lm(m3_obs_logVA ~ m3_mu_veg_post)

m3_slope_log <- coef(m3_fit_log)[2]
m3_R2_log    <- summary(m3_fit_log)$r.squared

# Plot
plot(m3_mu_veg_post,
     m3_obs_logVA,
     pch = 16,
     col = rgb(0,0,1,0.3),
     xlab = "Model-estimated log(VA + 1)",
     ylab = "Observed log(VA + 1)")

abline(0, 1, col = "red", lwd = 2)

# Add slope and R² text
text(x = min(m3_mu_veg_post),
     y = max(m3_obs_logVA),
     labels = paste0("Slope = ", round(m3_slope_log, 3)),
     pos = 4)

```

----

**Random effects shrinkage: **

Goal is centered around 0(ish)

1. beta_stream; random effect of stream on veg area; n=6
2. beta_event; random effect of sample event on veg area; n=12

```{r, echo=FALSE} 
par(mfrow = c(2, 1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hglmm_m3_damDen_Veg$sims.list$beta_stream)

boxplot(hglmm_m3_damDen_Veg$sims.list$beta_event)
```


----

----

**Posterior mean dam density effect**

damDen --> vegArea (no geom.covar.)

```{r,echo=FALSE}
m3_coeffparams <- c(#"alpha",
            "beta_damDen",
            "sigma")

m3_coef_df <- hglmm_m3_damDen_Veg$summary[m3_coeffparams, ] %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  select(parameter, mean, `2.5%`, `97.5%`)

ggplot(m3_coef_df, aes(y = parameter, x = mean)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(xmin = `2.5%`, xmax = `97.5%`),
    width = 0.2,
    orientation = "y"
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1) +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    x = "Effect Size (log scale)",
    y = ""
  ) +
  theme_bw(base_size = 14)
```


```{r,echo=FALSE}
summary(hglmm_m3_damDen_Veg$sims.list$beta_damDen)

exp(mean(hglmm_m3_damDen_Veg$sims.list$beta_damDen))
```

    Without geomorphic covariates, dam presence corresponds with a ~ 12-14% increase in surface water area. Again, lots of variation, and notable overlap of 0.


----

**Prior-posterior overlap:**

```{r,echo=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 

damDen_alpha_post_m3 <- hglmm_m3_damDen_Veg$sims.list$alpha
damDen_beta_post_m3  <- hglmm_m3_damDen_Veg$sims.list$beta_damDen
damDen_sigBetaStream_post_m3 <- hglmm_m3_damDen_Veg$sims.list$sigma_beta_stream

damDenMod_m3_n <- length(damDen_alpha_post_m3)

damDen_alpha_prior_m3 <- rnorm(damDenMod_m3_n, 0, sqrt(1/0.04))
damDen_beta_prior_m3  <- rnorm(damDenMod_m3_n, 0, sqrt(1/0.04))
damDen_sigBetaStream_prior_m3 <- pmax(rnorm(damDenMod_m3_n, 0, 1), 0)


# Plot overlap:

#plot_prior_post <- function(prior, post, name) {
  
#  df <- tibble(
#    value = c(prior, post),
#    dist  = rep(c("Prior", "Posterior"), each = length(prior))
#  )
  
#  ggplot(df, aes(x = value, fill = dist)) +
#    geom_density(alpha = 0.4) +
#    scale_fill_manual(values = c("gray60", "steelblue")) +
#    labs(
#      title = paste("Prior vs Posterior for", name),
#      x = name,
#      y = "Density"
#    ) +
#    theme_minimal(base_size = 14)
#}

plot_prior_post(damDen_alpha_prior_m3, damDen_alpha_post_m3, "alpha_m3")
plot_prior_post(damDen_beta_prior_m3, damDen_beta_post_m3, "beta_damDen_m3")
plot_prior_post(damDen_sigBetaStream_prior_m3, damDen_sigBetaStream_post_m3, "sigma_beta_stream_m3")
```


    PPOs still look pretty good.

----

**Change in SWA across damDensity by stream and sampleEvent**

```{r,echo=FALSE}

m3_dam_seq <- seq(min(aoi_HurdleData$dam_km),
               max(aoi_HurdleData$dam_km),
               length.out = 50)

# build PDP grid
m3_pdp_stream <- map_dfr(1:n_stream, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m3_post$alpha
  beta_j  <- damDen_m3_post$beta_damDen + damDen_m3_post$beta_stream[, j]
  eta_j   <- damDen_m3_post$eta_stream[, j]

  map_dfr(m3_dam_seq, function(d) {

    mu_draws <- alpha_j + beta_j * d + eta_j
    va_draws <- exp(mu_draws) - 1

    tibble(
      stream = j,
      damDen = d,
      SWA_med = median(va_draws),
      SWA_lo  = quantile(va_draws, 0.025),
      SWA_hi  = quantile(va_draws, 0.975)
    )
  })
})

stream_lookup <- aoi_HurdleData %>%
  distinct(stream, watershed) %>%
  arrange(stream)

m3_pdp_stream <- m3_pdp_stream %>%
  left_join(stream_lookup, by = "stream")



# plot
ggplot(m3_pdp_stream, aes(x = watershed, y = SWA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Partial Dependence of Vegetation Area on Stream and Dam Density",
    x = "Stream",
    y = "Predicted VegArea (m²)",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


###############################################################
################################################################

# build PDP grid
m3_pdp_event <- map_dfr(1:n_event, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m3_post$alpha
  beta_j  <- damDen_m3_post$beta_damDen + damDen_m3_post$beta_event[, j]
  eta_j   <- damDen_m3_post$eta_event[, j]

  map_dfr(m3_dam_seq, function(d) {

    mu_draws <- alpha_j + beta_j * d + eta_j
    va_draws <- exp(mu_draws) - 1

    tibble(
      event = j,
      damDen = d,
      SWA_med = median(va_draws),
      SWA_lo  = quantile(va_draws, 0.025),
      SWA_hi  = quantile(va_draws, 0.975)
    )
  })
})

event_year_lookup <- tibble(
  event = 1:12,
  event_label = c(
    "2019_Aug_1","2019_Sep_1","2019_Aug_2","2019_Sep_2",
    "2021_Aug_1","2021_Sep_1","2021_Aug_2","2021_Sep_2",
    "2023_Aug_1","2023_Sep_1","2023_Aug_2","2023_Sep_2"
  )
)


m3_pdp_event_named <- m3_pdp_event %>%
  left_join(event_year_lookup, by = "event")



# plot
ggplot(m3_pdp_event_named, aes(x = event_label, y = SWA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Partial Dependence of Vegetation Area on Sample Event and Dam Density",
    x = "Sample Event",
    y = "Predicted VegArea (m²)",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```


----

----

----

----

----



**MODEL 4: SW ~  damDensity + Geom.Covar.**

```{r,echo=FALSE}
#library(jagsUI)

    sink("HGLMM_hurdle_damDensity_m4.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR SW PRESENCE (1/2 Normal on SDs):
    
      sigma_zeta_stream  ~ dnorm(0,1) T(0,)
      tau_zeta_stream    <- pow(sigma_zeta_stream, -2)
    
      sigma_zeta_year    ~ dnorm(0,1) T(0,)
      tau_zeta_year      <- pow(sigma_zeta_year, -2)
    
      sigma_zeta_event   ~ dnorm(0,1) T(0,)
      tau_zeta_event     <- pow(sigma_zeta_event, -2)
    
      sigma_iota_stream  ~ dnorm(0,1) T(0,)
      tau_iota_stream    <- pow(sigma_iota_stream, -2)
    
      sigma_iota_event   ~ dnorm(0,1) T(0,)
      tau_iota_event     <- pow(sigma_iota_event, -2)
    
      sigma_iota_cluster ~ dnorm(0,1) T(0,)
      tau_iota_cluster   <- pow(sigma_iota_cluster, -2)
    
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 4) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 4) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 4) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Presence/absence:
      epsilon ~ dnorm(0,0.04)
      zeta_damDen ~ dnorm(0,0.04)
      zeta_incis ~ dnorm(0,0.04)
      zeta_RIPA ~ dnorm(0,0.04)
      zeta_UDA ~ dnorm(0,0.04)
    
      # Amount:
      alpha ~ dnorm(0,0.04)
      beta_damDen ~ dnorm(0,0.04)
      beta_incis ~ dnorm(0,0.04)
      beta_RIPA ~ dnorm(0,0.04)
      beta_UDA ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW PRESENCE:
    
      for(j in 1:n_stream){
        zeta_stream[j] ~ dnorm(0, tau_zeta_stream)
        iota_stream[j] ~ dnorm(0, tau_iota_stream)
      }
    
      for(yr in 1:n_year){
        zeta_year[yr] ~ dnorm(0, tau_zeta_year)
      }
    
      for(k in 1:n_event){
        zeta_event[k] ~ dnorm(0, tau_zeta_event)
        iota_event[k] ~ dnorm(0, tau_iota_event)
      }
    
      for(c in 1:n_cluster){
        iota_cluster[c] ~ dnorm(0, tau_iota_cluster)
      }
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(0, tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # DETECTION PRIORS:
      
      for(k in 1:n_event){
      p_TP[k] ~ dnorm(0.76, pow(0.16,-2)) T(0,1)   # empirically derived recall
      p_FP[k] ~ dnorm(0.01, pow(0.005,-2)) T(0,1)  # empirically derived FPR
      }
      
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Presence/absence:
        z[i] ~ dbern(psi[i])
    
        logit(psi[i]) <- epsilon
                         + (zeta_damDen
                            + zeta_stream[stream[i]]
                            + zeta_year[year[i]]
                            + zeta_event[event[i]]) * damDen[i]
                         + zeta_incis * incis[i]
                         + zeta_RIPA * RIPA[i]
                         + zeta_UDA * UDA[i]
                         + iota_stream[stream[i]]
                         + iota_event[event[i]]
                         + iota_cluster[cluster[i]]
    
        # Detection:
        p_eff[i] <- z[i] * p_TP[event[i]] + (1-z[i]) * p_FP[event[i]]
        y[i]     ~ dbern(p_eff[i])
    
        # Amount:
        y2[i] ~ dnorm(mu[i], tau_eff[i])
    
        mu[i] <- z[i] *
                  (alpha
                    + (beta_damDen
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damDen[i]
                    + beta_incis * incis[i] 
                    + beta_RIPA * RIPA[i]
                    + beta_UDA * UDA[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]])
    
        tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
      }
    
    }
    ", fill = TRUE)
    
    sink()
```   
    

```{r,echo=FALSE}
hurdle_damDen_m4_data_list <- list(
  N = N,
  y = y,
  y2 = y2,
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damDen = damDen,
  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)),
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hurdle_damDen_m4_inits <- function(){
  list(
    alpha        = rnorm(1, 0, 1),
    epsilon      = rnorm(1, 0, 1),

    sigma        = runif(1, 0.1, 2),
    sigma_zeta_stream  = runif(1, 0.1, 2),
    sigma_zeta_year    = runif(1, 0.1, 2),
    sigma_zeta_event   = runif(1, 0.1, 2),
    sigma_iota_stream  = runif(1, 0.1, 2),
    sigma_iota_event   = runif(1, 0.1, 2),
    sigma_iota_cluster = runif(1, 0.1, 2),

    sigma_beta_stream  = runif(1, 0.1, 2),
    sigma_beta_year    = runif(1, 0.1, 2),
    sigma_beta_event   = runif(1, 0.1, 2),
    sigma_eta_stream   = runif(1, 0.1, 2),
    sigma_eta_event    = runif(1, 0.1, 2),
    sigma_eta_cluster  = runif(1, 0.1, 2)
  )
}

hurdle_damDen_m4_params <- c(
  "psi",
  "mu",
  "epsilon",
  "alpha",
  "zeta_damDen",
  "zeta_incis",
  "zeta_RIPA",
  "zeta_UDA",
  "zeta_stream",
  "zeta_year",
  "zeta_event",
  "iota_stream",
  "iota_event",
  "iota_cluster",
  "beta_damDen",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_zeta_stream",
  "sigma_zeta_year",
  "sigma_zeta_event",
  "sigma_iota_stream",
  "sigma_iota_event",
  "sigma_iota_cluster",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_event",
  "sigma_eta_cluster",
  "p_TP",
  "p_FP",
  "z",
  "tau_eff"
)

#nchains <- 3
#nthin <- 5
#niter <- 100000
#nburn <- 25000


Sys.time()

hurdle_damDen_m4_SW <- jags(data = hurdle_damDen_m4_data_list,
                   inits = hurdle_damDen_m4_inits,
                   parameters.to.save = hurdle_damDen_m4_params,
                   model.file = "HGLMM_hurdle_damDensity_m4.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hurdle_damPres_SW) # model with local and watershed dam density as fixed effects
#print(hurdle_damPres_SW$summary[grep("beta_", rownames(hurdle_damPres_SW$summary)), ])
#print(hurdle_damPres_SW$summary[grep("zeta_", rownames(hurdle_damPres_SW$summary)), ])

```


----


DIAGNOSTICS:
----


```{r,echo=FALSE}
# Extract posterior draws

damDen_m4_post <- hurdle_damDen_m4_SW$sims.list 
```

**Assess "Chain health" Rhat + ESS:**

```{r,echo=FALSE} 
# Choose only the parameters you care about
params_m4 <- c(
  "epsilon",
  "alpha",
  "zeta_damDen",
  "beta_damDen",
  "zeta_incis",
  "beta_incis",
  "zeta_RIPA",
  "beta_RIPA",
  "zeta_UDA",
  "beta_UDA",
  "sigma",
  "sigma_zeta_stream",
  "sigma_iota_cluster",
  "p_TP[1]",
  "p_FP[1]")

# Gelman–Rubin convergence diagnostic
gelman.diag(hurdle_damDen_m4_SW$samples[, params_m4], multivariate = FALSE)

# Effective sample size
effectiveSize(hurdle_damDen_m4_SW$samples[, params_m4])

```

    Rhat & Effective Sample Size:
    
    Solid. Rhats near 1.0 (1.03-1.05 is fine); ESS all in the thousands. 
    beta_damDen still the hardest to estimate


----

**Check traceplots** for critical parameters:

```{r,echo=FALSE} 

par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


tplot_params_m4 <- c("alpha","beta_damDen","zeta_damDen", "sigma","sigma_zeta_stream","sigma_beta_stream")


mcmc_sub_m4 <- hurdle_damDen_m4_SW$samples[, tplot_params_m4]

coda::traceplot(mcmc_sub_m4)
```


    Traceplots all look good, again see beta_damDen. 


----

**Posterior predictive plots:**

Predicted Psi vs observed SW presence

```{r,echo=FALSE} 
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

# Posterior mean psi
psi_post_m4 <- apply(hurdle_damDen_m4_SW$sims.list$psi, 2, mean)

# Compare to observed y
plot(psi_post_m4, y, pch=16, col=rgb(0,0,0,0.2))


```

----

Predicted SW vs observed SW

Goal is 1:1 (ish)

```{r,echo=FALSE} 
mu_post_m4 <- apply(hurdle_damDen_m4_SW$sims.list$mu, 2, mean)

plot(mu_post_m4[y==1], y2[y==1], pch=16, col=rgb(0,0,1,0.3))
abline(0,1,col="red")


```

----

**Random effects shrinkage: **

Goal is centered around 0(ish)

1. zeta_stream; random effect of stream on sw presence
2. beta_stream; random effect of stream on sw amount
3. zeta_event; random effect of sample event on sw presence
4. beta_event; random effect of sample event on sw amount

```{r, echo=FALSE} 
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hurdle_damDen_m4_SW$sims.list$zeta_stream)
boxplot(hurdle_damDen_m4_SW$sims.list$beta_stream)
boxplot(hurdle_damDen_m4_SW$sims.list$zeta_event)
boxplot(hurdle_damDen_m4_SW$sims.list$beta_event)
```


----

----


**Posterior mean dam effects:**

damDen --> swPres

```{r,echo=FALSE}
summary(hurdle_damDen_m4_SW$sims.list$zeta_damDen)

exp(mean(hurdle_damDen_m4_SW$sims.list$zeta_damDen))
```

    When we account for geomorphic context, a 1 std dev. increase in dam density corresponds with an estimated 3% increase in the odds of surface water presence
    
    + 1 sd of damDensity ~= + 3.14 dams/km
    
    Effect = NEGLIGIBLE

----

damDen --> swAmount

```{r,echo=FALSE}
summary(hurdle_damDen_m4_SW$sims.list$beta_damDen)

exp(mean(hurdle_damDen_m4_SW$sims.list$beta_damDen))
```

    When we account for geomorhpic context, a 1 std dev. increase in dam density corresponds with a ~ 15% increase in surface water area (conditional on SW presence).
    
    Again - notable overlap of 0 for ALL global parameter estimates, due to massive variation across landscape/time.


----

**Prior-posterior overlap:**

```{r,echo=FALSE}
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8) 

damDen_alpha_m4_post <- hurdle_damDen_m4_SW$sims.list$alpha
damDen_beta_m4_post  <- hurdle_damDen_m4_SW$sims.list$beta_damDen
damDen_zeta_m4_post  <- hurdle_damDen_m4_SW$sims.list$zeta_damDen
damDen_sigBetaStream_m4_post <- hurdle_damDen_m4_SW$sims.list$sigma_beta_stream

damDenMod_m4_n <- length(damDen_alpha_m4_post)

damDen_alpha_m4_prior <- rnorm(damDenMod_m4_n, 0, sqrt(1/0.04))
damDen_beta_m4_prior  <- rnorm(damDenMod_m4_n, 0, sqrt(1/0.04))
damDen_zeta_m4_prior  <- rnorm(damDenMod_m4_n, 0, sqrt(1/0.04))
damDen_sigBetaStream_m4_prior <- pmax(rnorm(damDenMod_m4_n, 0, 1), 0)


# Plot overlap:

#plot_prior_post <- function(prior, post, name) {
  
#  df <- tibble(
#    value = c(prior, post),
#    dist  = rep(c("Prior", "Posterior"), each = length(prior))
#  )
  
#  ggplot(df, aes(x = value, fill = dist)) +
#    geom_density(alpha = 0.4) +
#    scale_fill_manual(values = c("gray60", "steelblue")) +
#    labs(
#      title = paste("Prior vs Posterior for", name),
#      x = name,
#      y = "Density"
#    ) +
#    theme_minimal(base_size = 14)
#}

plot_prior_post(damDen_alpha_m4_prior, damDen_alpha_m4_post, "alpha_m4")
plot_prior_post(damDen_beta_m4_prior, damDen_beta_m4_post, "beta_damDen_m4")
plot_prior_post(damDen_zeta_m4_prior, damDen_zeta_m4_post, "zeta_damDen_m4")
plot_prior_post(damDen_sigBetaStream_m4_prior, damDen_sigBetaStream_m4_post, "sigma_beta_stream_m4")
```

    PPOs still look good pretty good.

----

**Marginal effect** of dam Density on surface water presence, given geomorphic context

**Note** there's a small but negligble positive effect of a 1 SD increase in DamDensity on surface water presence. However, higher dam densities do increase probability of SW presence more notably (empirical range of dam/km is up to +4 standard dev. from mean; 4*0.13 = 0.52 upwards of 52% increase in SWA at highest damDensity) 

```{r,echo=FALSE}
# Build a grid of raw geomorphic covariate values
damDen_m4_geom_grid <- tibble(
  TO_raw = seq(min(aoi_HurdleData$TO),
               max(aoi_HurdleData$TO), length.out = 50),

  incis_raw = 1 - (TO_raw / 90),   # raw incision scale

  RIPA_raw = seq(min(aoi_HurdleData$riparian_area_m2),
                 max(aoi_HurdleData$riparian_area_m2), length.out = 50),

  UDA_raw  = seq(min(aoi_HurdleData$catchment_km2),
                 max(aoi_HurdleData$catchment_km2), length.out = 50)
) %>%
  # Convert raw → standardized (to match model scale)
  mutate(
    incis = as.numeric(scale(incis_raw)),
    RIPA  = as.numeric(scale(RIPA_raw)),
    UDA   = as.numeric(scale(UDA_raw))
  )

# Make dam density sequence
damDen_m4_seq <- seq(min(aoi_HurdleData$dam_km),max(aoi_HurdleData$dam_km),length.out=10)

# Compute surface water across dam density sequence
damDen_m4_compute_psi <- function(damDen_val, geom_grid, damDen_post) {

  eps <- mean(damDen_post$epsilon)
  zd  <- mean(damDen_post$zeta_damDen)
  zi  <- mean(damDen_post$zeta_incis)
  zr  <- mean(damDen_post$zeta_RIPA)
  zu  <- mean(damDen_post$zeta_UDA)

  geom_grid %>%
    mutate(
      psi = plogis(
        eps +
        zd * damDen_val +
        zi * incis +      # standardized incision
        zr * RIPA +       # standardized RIPA
        zu * UDA          # standardized UDA
      ),
      damDen = damDen_val
    )
}


```

```{r,echo=FALSE}
damDen_m4_psi_df <- map_df(damDen_m4_seq,
                        ~ damDen_m4_compute_psi(.x,
                                             damDen_m4_geom_grid,
                                             damDen_m4_post))
```

```{r,echo=FALSE}
ggplot(damDen_m4_psi_df, aes(x = incis_raw, y = psi, color = damDen)) +
  geom_point() +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "Marginal Effect of Dam Density on Surface Water Presence",
    subtitle = "Predicted probability of water presence (psi) across raw incision values",
    x = "Channel Incision (raw scale: 1 - TO/90)",
    y = "Probability of Surface Water",
    color = "Dam Density (dams/km)"
  ) +
  theme_minimal(base_size = 14)

```

```{r,echo=FALSE}
ggplot(damDen_m4_psi_df, aes(x = RIPA_raw, y = psi, color = damDen)) +
  geom_point() +
  scale_color_viridis_c(option="plasma") +
  labs(
    title = "Marginal Effect of Dam Density on Surface Water Presence",
    subtitle = "Predicted probability of water (psi) across raw riparian area (m^2)",
    x = "Riparian Area (m^2)",
    y = "Probability of Surface Water",
    color = "Dam Density (dams/km)"
  ) +
  theme_minimal(base_size = 14)
```

```{r,echo=FALSE}
ggplot(damDen_m4_psi_df, aes(x = UDA_raw, y = psi, color = damDen)) +
  geom_point() +
  scale_color_viridis_c(option="plasma") +
  labs(
    title = "Marginal Effect of Dam Density on Surface Water Presence",
    subtitle = "Predicted probability of water (psi) across raw upstream drainage area (km^2)",
    x = "Upstream Drainage Area (km^2)",
    y = "Probability of Surface Water",
    color = "Dam Density (dams/km)"
  ) +
  theme_minimal(base_size = 14)


```

----

**Partial dependence plots**

Constrain PDP to empirical ranges for SWA and target covariates

```{r,echo=FALSE}
library(tidyverse)
library(scales)

# Raw covariates
incis_m4_raw <- 1 - aoi_HurdleData$TO/90
RIPA_m4_raw  <- aoi_HurdleData$riparian_area_m2
UDA_m4_raw   <- aoi_HurdleData$catchment_km2
#damDen <- aoi_HurdleData$dam_km_std
damDen_m4_raw <- aoi_HurdleData$dam_km
damDen_m4_mean_raw <- mean(damDen_m4_raw)
damDen_m4_sd_raw   <- sd(damDen_m4_raw)

logSWA_m4_mean <- mean(aoi_HurdleData$logSWA)
logSWA_m4_sd   <- sd(aoi_HurdleData$logSWA)

# Standardization stats
incis_m4_mean_raw <- mean(incis_m4_raw)
incis_m4_sd_raw   <- sd(incis_m4_raw)

RIPA_m4_mean_raw  <- mean(RIPA_m4_raw)
RIPA_m4_sd_raw    <- sd(RIPA_m4_raw)

UDA_m4_mean_raw   <- mean(UDA_m4_raw)
UDA_m4_sd_raw     <- sd(UDA_m4_raw)


##############################3

# Helper Functions to Compute SWA from posterior draws
compute_swa_draws_damDen <- function(incis, RIPA, UDA, post, damDen_val) {
  n_draws <- length(post$alpha)
  swa_mat <- matrix(NA_real_, nrow = length(incis), ncol = n_draws)

  for (j in seq_len(n_draws)) {
    lp <- post$alpha[j] +
          post$beta_incis[j] * incis +
          post$beta_RIPA[j]  * RIPA +
          post$beta_UDA[j]   * UDA +
          post$beta_damDen[j] * damDen_val

    swa_mat[, j] <- exp(lp) - 1
  }

  swa_mat
}

summarize_swa <- function(swa_mat) {
  tibble(
    SWA_med = apply(swa_mat, 1, median),
    SWA_lo  = apply(swa_mat, 1, quantile, probs = 0.025),
    SWA_hi  = apply(swa_mat, 1, quantile, probs = 0.975)
  )
}




#########################################

# PDP Builder:

build_pdp_damDen <- function(focal_raw, focal_name,
                      incis_raw, RIPA_raw, UDA_raw,
                      post,
                      incis_mean_raw, incis_sd_raw,
                      RIPA_mean_raw, RIPA_sd_raw,
                      UDA_mean_raw, UDA_sd_raw,
                      damDen_raw) {

  # Sequence only across empirical range
  focal_seq <- seq(min(focal_raw), max(focal_raw), length.out = 200)

  # Conditional medians of other covariates
  incis_med <- median(incis_raw)
  RIPA_med  <- median(RIPA_raw)
  UDA_med   <- median(UDA_raw)
  
  # Empirical dam density range (standardized)
  damDen_seq <- seq(min(damDen_raw), max(damDen_raw), length.out = 10)

  # Build grid (200 rows)
  grid <- tibble(
    focal_raw = focal_seq,
    incis_raw = if (focal_name == "incis") focal_seq else incis_med,
    RIPA_raw  = if (focal_name == "RIPA")  focal_seq else RIPA_med,
    UDA_raw   = if (focal_name == "UDA")   focal_seq else UDA_med
  ) %>%
    mutate(
      incis = (incis_raw - incis_mean_raw) / incis_sd_raw,
      RIPA  = (RIPA_raw  - RIPA_mean_raw)  / RIPA_sd_raw,
      UDA   = (UDA_raw   - UDA_mean_raw)   / UDA_sd_raw
    )
  # Compute PDP curves for each dam density value
  pdp_list <- map(damDen_seq, function(dd) {
    swa_mat <- compute_swa_draws_damDen(grid$incis, grid$RIPA, grid$UDA, post,
                                        (dd - damDen_m4_mean_raw) / damDen_m4_sd_raw)
    summarize_swa(swa_mat) %>% mutate(damDen = dd)
  })

  pdp <- bind_rows(pdp_list)

  # Repeat grid for each damDen value
  grid2 <- bind_rows(replicate(length(damDen_seq), grid, simplify = FALSE))

  bind_cols(grid2, pdp)
}



#####

pdp_incis_m4_damDen <- build_pdp_damDen(
  focal_raw = incis_m4_raw,
  focal_name = "incis",
  incis_m4_raw, RIPA_m4_raw, UDA_m4_raw,
  damDen_m4_post,
  incis_m4_mean_raw, incis_m4_sd_raw,
  RIPA_m4_mean_raw, RIPA_m4_sd_raw,
  UDA_m4_mean_raw, UDA_m4_sd_raw, damDen_m4_raw
)

pdp_RIPA_m4_damDen <- build_pdp_damDen(
  focal_raw = RIPA_m4_raw,
  focal_name = "RIPA",
  incis_m4_raw, RIPA_m4_raw, UDA_m4_raw,
  damDen_m4_post,
  incis_m4_mean_raw, incis_m4_sd_raw,
  RIPA_m4_mean_raw, RIPA_m4_sd_raw,
  UDA_m4_mean_raw, UDA_m4_sd_raw, damDen_m4_raw
)

pdp_UDA_m4_damDen <- build_pdp_damDen(
  focal_raw = UDA_m4_raw,
  focal_name = "UDA",
  incis_raw, RIPA_m4_raw, UDA_m4_raw,
  damDen_m4_post,
  incis_m4_mean_raw, incis_m4_sd_raw,
  RIPA_m4_mean_raw, RIPA_m4_sd_raw,
  UDA_m4_mean_raw, UDA_m4_sd_raw, damDen_m4_raw
)



##########################################3

# Plot Function:

plot_pdp_damDen <- function(pdp, xvar, xtitle) {
  ggplot(pdp,
         aes(x = .data[[xvar]], y = SWA_med,
             color = damDen, fill = damDen)) +
    #geom_ribbon(aes(ymin = SWA_lo, ymax = SWA_hi), alpha = 0.15, color = NA) +
    geom_point() +
    scale_color_viridis_c(option = "plasma") +
    scale_fill_viridis_c(option = "plasma") +
    scale_y_continuous(labels = comma) +
    labs(
      title = paste("Partial Dependence of SWA on dam density and", xtitle),
      subtitle = "Across empirical dam density range",
      x = xtitle,
      y = "Predicted SWA (m^2)",
      color = "Dam Density (dams/km)",
      fill  = "Dam Density (dams/km)"
    ) +
    theme_minimal(base_size = 14)
}



###########################################3


plot_pdp_damDen(pdp_incis_m4_damDen, "incis_raw", "Channel Incision (raw)")
plot_pdp_damDen(pdp_RIPA_m4_damDen,  "RIPA_raw",  "Riparian Area (m²)")
plot_pdp_damDen(pdp_UDA_m4_damDen,   "UDA_raw",   "Contributing Drainage Area (km²)")


```

----

**Change in SWA across damDensity by stream and sampleEvent**

```{r,echo=FALSE}

m4_dam_seq <- seq(min(aoi_HurdleData$dam_km),
               max(aoi_HurdleData$dam_km),
               length.out = 50)

# build PDP grid
m4_pdp_stream <- map_dfr(1:n_stream, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m4_post$alpha
  beta_j  <- damDen_m4_post$beta_damDen + damDen_m4_post$beta_stream[, j]
  eta_j   <- damDen_m4_post$eta_stream[, j]

  map_dfr(m4_dam_seq, function(d) {

    mu_draws <- alpha_j + beta_j * d + eta_j
    swa_draws <- exp(mu_draws) - 1

    tibble(
      stream = j,
      damDen = d,
      SWA_med = median(swa_draws),
      SWA_lo  = quantile(swa_draws, 0.025),
      SWA_hi  = quantile(swa_draws, 0.975)
    )
  })
})

stream_lookup <- aoi_HurdleData %>%
  distinct(stream, watershed) %>%
  arrange(stream)

m4_pdp_stream <- m4_pdp_stream %>%
  left_join(stream_lookup, by = "stream")



# plot
ggplot(m4_pdp_stream, aes(x = watershed, y = SWA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "Partial Dependence of SWA on Stream and Dam Density",
    x = "Stream",
    y = "Predicted SWA (m²)",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


###############################################################
################################################################

# build PDP grid
m4_pdp_event <- map_dfr(1:n_event, function(j) {

  # posterior draws for this stream
  alpha_j <- damDen_m4_post$alpha
  beta_j  <- damDen_m4_post$beta_damDen + damDen_m4_post$beta_event[, j]
  eta_j   <- damDen_m4_post$eta_event[, j]

  map_dfr(m4_dam_seq, function(d) {

    mu_draws <- alpha_j + beta_j * d + eta_j
    swa_draws <- exp(mu_draws) - 1

    tibble(
      event = j,
      damDen = d,
      SWA_med = median(swa_draws),
      SWA_lo  = quantile(swa_draws, 0.025),
      SWA_hi  = quantile(swa_draws, 0.975)
    )
  })
})

event_year_lookup <- tibble(
  event = 1:12,
  event_label = c(
    "2019_Aug_1","2019_Sep_1","2019_Aug_2","2019_Sep_2",
    "2021_Aug_1","2021_Sep_1","2021_Aug_2","2021_Sep_2",
    "2023_Aug_1","2023_Sep_1","2023_Aug_2","2023_Sep_2"
  )
)


m4_pdp_event_named <- m4_pdp_event %>%
  left_join(event_year_lookup, by = "event")



# plot
ggplot(m4_pdp_event_named, aes(x = event_label, y = SWA_med, color = damDen)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "Partial Dependence of SWA on Sample Event and Dam Density",
    x = "Sample Event",
    y = "Predicted SWA (m²)",
    color = "Dam Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```