---
title: "Ch1_Hurdle SWA & VEG ~ BeaverDam"
author: "Matt Webster"
date: "2026-07-14"
output: html_document
editor_options: 
  chunk_output_type: console
---




The purpose of this script is to explore the conditional effects of beaver dams on: 

    - surface water presence, 
    - surface water area, 
    - vegetation presence, 
    - vegetation area, 
    - and NDVI, 


This modeling focuses solely on late summer months; August & September. 

Several models are presented herein. All models have the same random effects structure for stream, year, and sample event. 

    Model 1: Surface water presence & area as a function of beaver dam presence
    
    Model 2: Surface water presence & area as a function of beaver dam density (dams/km^2)
    
    Model 3: Vegetation presence & area as a function of beaver dam presence
    
    Model 4: Vegetation presence & area as a functino of beaver dam density (dams/km^2)
    
    Model 5: NDVI as a function of beaver dam presence
    
    Model 6: NDVI as a function of beaver dam density


Models 1:4 are constructed as **hurdle** models. The goal of the model in this hurdle framework is to 1) estimate the relative influence of dam presence/density on the log-odds (or probability) of surface water presence; & 2) for cases where water is present, estimate the relative influence of dam presence/density on surface water area. 

----

----

----

----

----


```{r,echo=FALSE}
  knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

```{r,echo=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dataRetrieval)
library(knitr)
library(ggplot2)
library(zoo) #for moving window
library(stringr)
library(vioplot)
library(purrr)
library(lavaan)
library(blavaan)
library(ggcorrplot)
library(lme4)
library(broom.mixed)
library(coda)
```
      



USGS Flow Data
----

**Data from USGS LBV Guston Coulee gage, near Saco, MT:**

```{r,echo=FALSE, warning=FALSE}
# LOWER BEAVER CK
# Gage near Guston Coulee
usgsGage_06166000 <- read_waterdata_daily(
  monitoring_location_id = "USGS-06166000",
  parameter_code = "00060",
  time = c("2019-06-01","2023-10-01")
)

###########################################

# Clean USGS
usgs_df <- usgsGage_06166000 %>%
  mutate(
    source = "USGS",
    date   = as.Date(time),
    flow   = value
    ) %>%
  filter(
    (date >= as.Date("2019-06-01") & date <= as.Date("2019-10-01")) |
    (date >= as.Date("2021-06-01") & date <= as.Date("2021-10-01")) |
    (date >= as.Date("2023-06-01") & date <= as.Date("2023-10-01")) ) %>% 
  mutate(
    zFlow = scale(flow)
  ) %>% 
  select(source,
         date,
         zFlow) %>% 
  mutate(surveyPeriod = case_when(between(date, as.Date("2019-06-01"), as.Date("2019-06-14")) ~ 191,
                                between(date, as.Date("2019-06-15"), as.Date("2019-06-30")) ~ 192,
                                between(date, as.Date("2019-07-01"), as.Date("2019-07-14")) ~ 193,
                                between(date, as.Date("2019-07-15"), as.Date("2019-07-31")) ~ 194,
                                between(date, as.Date("2019-08-01"), as.Date("2019-08-14")) ~ 195,
                                between(date, as.Date("2019-08-15"), as.Date("2019-08-31")) ~ 196,
                                between(date, as.Date("2019-09-01"), as.Date("2019-09-14")) ~ 197,
                                between(date, as.Date("2019-09-15"), as.Date("2019-09-30")) ~ 198,
                                between(date, as.Date("2021-06-01"), as.Date("2021-06-14")) ~ 211,
                                between(date, as.Date("2021-06-15"), as.Date("2021-06-30")) ~ 212,
                                between(date, as.Date("2021-07-01"), as.Date("2021-07-14")) ~ 213,
                                between(date, as.Date("2021-07-15"), as.Date("2021-07-31")) ~ 214,
                                between(date, as.Date("2021-08-01"), as.Date("2021-08-14")) ~ 215,
                                between(date, as.Date("2021-08-15"), as.Date("2021-08-31")) ~ 216,
                                between(date, as.Date("2021-09-01"), as.Date("2021-09-14")) ~ 217,
                                between(date, as.Date("2021-09-15"), as.Date("2021-09-30")) ~ 218,
                                between(date, as.Date("2023-06-01"), as.Date("2023-06-14")) ~ 231,
                                between(date, as.Date("2023-06-15"), as.Date("2023-06-30")) ~ 232,
                                between(date, as.Date("2023-07-01"), as.Date("2023-07-14")) ~ 233,
                                between(date, as.Date("2023-07-15"), as.Date("2023-07-31")) ~ 234,
                                between(date, as.Date("2023-08-01"), as.Date("2023-08-14")) ~ 235,
                                between(date, as.Date("2023-08-15"), as.Date("2023-08-31")) ~ 236,
                                between(date, as.Date("2023-09-01"), as.Date("2023-09-14")) ~ 237,
                                between(date, as.Date("2023-09-15"), as.Date("2023-09-30")) ~ 238,
                                TRUE ~ NA_real_)) %>% 
  filter(surveyPeriod > 0)

usgs_df$surveyPeriod <- as.factor(usgs_df$surveyPeriod)


mean_flow_per_survey_period <- usgs_df %>% 
  group_by(surveyPeriod) %>%
  summarise(zFlow_mean = mean(zFlow),
            .groups = "drop")



lateSummerFlow <- usgs_df %>% 
  filter(surveyPeriod != 191,
         surveyPeriod != 192,
         surveyPeriod != 193,
         surveyPeriod != 194,
         surveyPeriod != 211,
         surveyPeriod != 212,
         surveyPeriod != 213,
         surveyPeriod != 214,
         surveyPeriod != 231,
         surveyPeriod != 232,
         surveyPeriod != 233,
         surveyPeriod != 234,
         )
```

**Late Summer Flow Conditions**


**2019:**


Moderate August with small peak in September


```{r,echo=FALSE}

flow19 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2019-06-01") & date <= as.Date("2019-10-01"))
  )


####

ggplot(flow19, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '19 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```


----


**2021:**


Low flow conditions throughout the summer season. Part of 3 year sustained drought conditions (2020, 2021, 2022).


Gage recorded no flow during August & September



```{r,echo=FALSE}

flow21 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2021-06-01") & date <= as.Date("2021-10-01"))
  )


####

ggplot(flow21, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '21 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```  
  
  
----


**2023:**


Moderate to low August & September flows; after early season extreme flooding. 



```{r,echo=FALSE}
flow23 <- lateSummerFlow %>%
  mutate(date = as.Date(date)) %>%
  filter(
    (date >= as.Date("2023-06-01") & date <= as.Date("2023-10-01"))
  )


####

ggplot(flow23, aes(x = date, 
                       y = zFlow,
                       group = lubridate::year(date))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    x = "Date",
    y = "Discharge (cfs)",
    title = "Summer '23 Stream Discharge; USGS 06166000; Lower Beaver",
    subtitle = "z-scaled June through September, 2019, 2021, 2023"
  ) +
  theme_minimal(base_size = 14)

```  

----

GEE-derived data for each stream:
----

```{r, echo=FALSE}

#Read in data frames for each HUC12/Year combo:

# WILLOW:
#2019:
wc_19 <- read.csv("WC_2019_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2019,
         watershed = 'WC')
#2021:
wc_21 <- read.csv("WC_2021_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2021,
         watershed = 'WC')
#2023:
wc_23 <- read.csv("WC_2023_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2023,
         watershed = 'WC')
#AllYears:
WC <- bind_rows(wc_19,wc_21,wc_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LONE TREE:
#2019:
lt_19 <- read.csv("LT_2019_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2019,
         watershed = 'LT')
#2021:
lt_21 <- read.csv("LT_2021_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2021,
         watershed = 'LT')
#2023:
lt_23 <- read.csv("LT_2023_DF_Final_Long_0630Update.csv")%>% 
  mutate(year = 2023,
         watershed = 'LT')

#AllYears:
LT <- bind_rows(lt_19,lt_21,lt_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))

# BRAZIL:
#2019:
brz_19 <- read.csv("BRZ_2019_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2019,
         watershed = 'BRZ')
#2021:
brz_21 <- read.csv("BRZ_2021_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2021,
         watershed = 'BRZ')
#2023:
brz_23 <- read.csv("BRZ_2023_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2023,
         watershed = 'BRZ')

#AllYears:
BRZ <- bind_rows(brz_19,brz_21,brz_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LARB:
#2019:
lrb_19 <- read.csv("LRB_2019_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2019,
         watershed = 'LRB')
#2021:
lrb_21 <- read.csv("LRB_2021_DF_Final_Long_0630Update.csv")%>% 
  mutate(year = 2021,
         watershed = 'LRB')
#2023:
lrb_23 <- read.csv("LRB_2023_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2023,
         watershed = 'LRB')

#AllYears:
LRB <- bind_rows(lrb_19,lrb_21,lrb_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# LOWER BEAVER:
#2019:
lbv_19 <- read.csv("LBV_2019_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2019,
         watershed = 'LBV')
#2021:
lbv_21 <- read.csv("LBV_2021_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2021,
         watershed = 'LBV')
#2023:
lbv_23 <- read.csv("LBV_2023_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2023,
         watershed = 'LBV')

#AllYears:
LBV <- bind_rows(lbv_19,lbv_21,lbv_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
  
# UPPER BEAVER:
#2019:
ubv_19 <- read.csv("UBV_2019_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2019,
         watershed = 'UBV')
#2021:
ubv_21 <- read.csv("UBV_2021_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2021,
         watershed = 'UBV')
#2023:
ubv_23 <- read.csv("UBV_2023_DF_Final_Long_0630Update.csv") %>% 
  mutate(year = 2023,
         watershed = 'UBV')

#AllYears:
UBV <- bind_rows(ubv_19,ubv_21,ubv_23) %>% 
  mutate(surveyPeriod = as.factor(case_when(year == 2019 & t == 1 ~ 191,
                                  year == 2019 & t == 2 ~ 192,
                                  year == 2019 & t == 3 ~ 193,
                                  year == 2019 & t == 4 ~ 194,
                                  year == 2019 & t == 5 ~ 195,
                                  year == 2019 & t == 6 ~ 196,
                                  year == 2019 & t == 7 ~ 197,
                                  year == 2019 & t == 8 ~ 198,
                                  year == 2021 & t == 1 ~ 211,
                                  year == 2021 & t == 2 ~ 212,
                                  year == 2021 & t == 3 ~ 213,
                                  year == 2021 & t == 4 ~ 214,
                                  year == 2021 & t == 5 ~ 215,
                                  year == 2021 & t == 6 ~ 216,
                                  year == 2021 & t == 7 ~ 217,
                                  year == 2021 & t == 8 ~ 218,
                                  year == 2023 & t == 1 ~ 231,
                                  year == 2023 & t == 2 ~ 232,
                                  year == 2023 & t == 3 ~ 233,
                                  year == 2023 & t == 4 ~ 234,
                                  year == 2023 & t == 5 ~ 235,
                                  year == 2023 & t == 6 ~ 236,
                                  year == 2023 & t == 7 ~ 237,
                                  year == 2023 & t == 8 ~ 238)))
```


----

Currently each incoming data frame is of the following structure:

```{r,include=FALSE}
str(brz_19)
```

Where:

    - clusterID_seq = dammed/undammed sample pair ID
    - window_type = whether a sample window is 'dammed' or 'undammed'
    - watershed = watershed identifier (e.g., "BRZ" = Brazil Ck)
    - year = year
    - window_length = length (m) of stream flow line within sample window
    - catchment_km2 = contributing catchment area/upstream drainage area (km^2)
    - riparian_area_m2 = riparian area (m^2) of sample window; 2-m vertical
      buffer from DEM low point (stream flow line)
    - TO = mean topographic openness across window
    - p_window: avg dam detection prob for target window; see CP Ch1
    - f_window: avg dam false positive prob for target window; see CP Ch1
    - beav_count_all: all unique detected dams
    - beav_count_conf: unique confidently detected dams
    - anthro_count: all unique detected anthropogenic dams
    - isAnthro: National Wetland Inventory (NWI) "dyked/impounded"
    - Nhat = estimated dam count within sample window
      - ...= beav_count_all/(p_window(1-f_window)  [McCallum 2005]
    - TO_mean = mean topographic openness across window
    - time_period = 2wk interval 06/01 - 09/30
    - SWA = surface water area (m^2) for window during time (t)
    - t = time (t) {same as time_period}



----

----

----

----



```{r,include=FALSE, fig.width=10}
WC_RelFlow <- WC %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)
```

```{r,include=FALSE, fig.width=10}
LT_RelFlow <- LT %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)

```

```{r,include=FALSE, fig.width=10}
BRZ_RelFlow <- BRZ %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)
```

```{r,include=FALSE, fig.width=10}
LRB_RelFlow <- LRB %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)

```

```{r,include=FALSE, fig.width=10}
LBV_RelFlow <- LBV %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)
```

```{r,include=FALSE, fig.width=10}
UBV_RelFlow <- UBV %>% 
  left_join(mean_flow_per_survey_period,
            by = "surveyPeriod") %>% 
  mutate(dam_km = (Nhat/window_length)*1000)
```


----

----

----

----


```{r,include=FALSE}
# Create GLMM data frame:

aoi_GLM_Data <- WC_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_km) %>% 
  bind_rows(LT_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_km)) %>% 
  bind_rows(BRZ_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_km)) %>% 
  bind_rows(LRB_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_km)) %>% 
  bind_rows(LBV_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_km)) %>% 
  bind_rows(UBV_RelFlow %>% 
  select(clusterID_seq,
         window_type,
         watershed,
         year,
         catchment_km2,
         TO,
         Nhat,
         t,
         surveyPeriod,
         SWA,
         riparian_area_m2,
         dam_km))


aoi_GLM_Data$clusterID_unique <- interaction(
  aoi_GLM_Data$year,
  aoi_GLM_Data$watershed,
  aoi_GLM_Data$clusterID_seq,
  drop = TRUE
)

aoi_GLM_Data$clusterID_num <- as.numeric(aoi_GLM_Data$clusterID_unique)

#aoi_GLM_Data$eventID <- interaction(
#  aoi_GLM_Data$year,
#  aoi_GLM_Data$surveyPeriod,
#  drop = TRUE
#)

#aoi_GLM_Data$eventID_num <- as.numeric(aoi_GLM_Data$eventID)

aoi_GLM_Data$stream <- as.numeric(as.factor(aoi_GLM_Data$watershed))

aoi_GLM_Data$sampleEvent <- as.numeric(aoi_GLM_Data$surveyPeriod)

aoi_GLM_Data$year_id <- as.numeric(as.factor(aoi_GLM_Data$year))

aoi_GLM_Data <- aoi_GLM_Data %>% 
  mutate(damPresence = case_when(Nhat > 0 ~ 1,
                                 Nhat == 0 ~ 0))

str(aoi_GLM_Data)



aoi_GLM_Data_mod <- aoi_GLM_Data %>%
  filter(!is.na(SWA))   # or !is.na(RI) if that’s the response


# Build explicit lookups from the data actually used in the model
stream_lookup <- aoi_GLM_Data_mod %>%
  distinct(stream, watershed) %>%
  arrange(stream)

year_lookup <- aoi_GLM_Data_mod %>%
  distinct(year_id, year) %>%
  arrange(year_id)

uda_lookup <- aoi_GLM_Data_mod %>%
  group_by(watershed) %>%
  summarise(max_catchment = max(catchment_km2, na.rm = TRUE)) %>%
  arrange(max_catchment)



# Start from the full modeled data and remove early summer time periods
# This analysis will look exclusively at late season mesic resource availability.
# Imported data has already been filtered to the target time frame. 
# 08/01-09/30
# t5:t8

aoi_glm_data_mod_refined <- aoi_GLM_Data_mod

# If this is not the case for whatever reason, use the code block below

#aoi_glm_data_mod_refined <- aoi_GLM_Data_mod %>% 
#  filter(!(sampleEvent %in% 1:4) &      # drop 2019 t1-t4
#           !(sampleEvent %in% 9:12) &   # drop 2021 t1-t4
#           !(sampleEvent %in% 17:20))   # drop 2023 t1–t4

event_lookup <- data.frame(
  old = sort(unique(aoi_glm_data_mod_refined$sampleEvent)),
  new = 1:length(unique(aoi_glm_data_mod_refined$sampleEvent))
)

aoi_glm_data_mod_refined <- aoi_glm_data_mod_refined %>%
  left_join(event_lookup, by = c("sampleEvent" = "old")) %>%
  mutate(sampleEvent_new = new) %>%
  select(-new)

event_year_3_4 <- tapply(aoi_glm_data_mod_refined$year_id,
                         aoi_glm_data_mod_refined$sampleEvent_new,
                         unique)
event_year_3_4 <- as.numeric(event_year_3_4)



aoi_glm_data_mod_refined$SWA <- ifelse(aoi_glm_data_mod_refined$SWA < 1,
                                   0,
                                   aoi_glm_data_mod_refined$SWA)
```


```{r,echo=FALSE}

# LOG TRANSFORM SWA TO ADDRESSS RIGHT SKEW

aoi_glm_data_mod_refined$logSWA <- log(aoi_glm_data_mod_refined$SWA+1)

```


Format data for hurdle:
----

```{r,echo=TRUE}
# Create binary indicator of surface water presence, for logistic part of hurdle model:
aoi_HurdleData <- aoi_glm_data_mod_refined
aoi_HurdleData$y <- ifelse(aoi_glm_data_mod_refined$SWA>0,1,0)
aoi_HurdleData$dam_km_std <- scale(aoi_HurdleData$dam_km)

# Full data indices for Bernoulli sub-model
N <- nrow(aoi_HurdleData)
y <- aoi_HurdleData$y
y2 <- aoi_HurdleData$logSWA
stream <- aoi_HurdleData$stream
year <- aoi_HurdleData$year_id
event <- aoi_HurdleData$sampleEvent_new
cluster <- aoi_HurdleData$clusterID_num
damPres <- aoi_HurdleData$damPresence
damDen <- as.numeric(aoi_HurdleData$dam_km_std)
damDen_raw <- aoi_HurdleData$dam_km

n_stream <- length(unique(aoi_HurdleData$stream))
n_year <- length(unique(aoi_HurdleData$year_id))
n_event <- length(unique(aoi_HurdleData$sampleEvent_new))
n_cluster <- length(unique(aoi_HurdleData$clusterID_num))
```

----

**Log-transform SWA to 'normalize' the right skew:**

**Visualize SWA spread:**

---
---

**Raw SWA (m^2):**

```{r,echo=FALSE}
hist(aoi_HurdleData$SWA,
     main = "Distribution of SWA",
     xlab = "log(SWA)")
```

**log(SWA):**

```{r,echo=FALSE}
hist(aoi_HurdleData$logSWA,
     main = "Distribution of log(SWA)",
     xlab = "log(SWA)")
```

----

----

----

**Hurdle HGLMM**
----



    This section explores late summner SWA in relation to beaver dam presence.
    
    Uses hurdle model with nested random slopes for beaver dam effect to allow for stream, year, 
    and sampleEvent to cause deviation from the global mean beaver effect.
    
    Pt1: Surface water presence/absence; logistic regression
    Pt2: Surface water amount; log(Normal) regression
    
    Uses log(SWA) for AOI all years. Adjust priors accordingly...


----

----

**Model 1: SWA ~ DamPresence**

```{r,echo=FALSE}
library(jagsUI)
```
    sink("HGLMM_hurdle_damPresence.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR SW PRESENCE (1/2 Normal on SDs):
    
      sigma_zeta_stream  ~ dnorm(0,1) T(0,)
      tau_zeta_stream    <- pow(sigma_zeta_stream, -2)
    
      sigma_zeta_year    ~ dnorm(0,1) T(0,)
      tau_zeta_year      <- pow(sigma_zeta_year, -2)
    
      sigma_zeta_event   ~ dnorm(0,1) T(0,)
      tau_zeta_event     <- pow(sigma_zeta_event, -2)
    
      sigma_iota_stream  ~ dnorm(0,1) T(0,)
      tau_iota_stream    <- pow(sigma_iota_stream, -2)
    
      sigma_iota_event   ~ dnorm(0,1) T(0,)
      tau_iota_event     <- pow(sigma_iota_event, -2)
    
      sigma_iota_cluster ~ dnorm(0,1) T(0,)
      tau_iota_cluster   <- pow(sigma_iota_cluster, -2)
    
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 0.7) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 0.7) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 0.7) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Presence/absence:
      epsilon ~ dnorm(0,0.04)
      zeta_damPres ~ dnorm(0,0.04)
      zeta_incis ~ dnorm(0,0.04)
      zeta_RIPA ~ dnorm(0,0.04)
      zeta_UDA ~ dnorm(0,0.04)
    
      # Amount:
      alpha ~ dnorm(0,0.04)
      beta_damPres ~ dnorm(0,0.04)
      beta_incis ~ dnorm(0,0.04)
      beta_RIPA ~ dnorm(0,0.04)
      beta_UDA ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW PRESENCE:
    
      for(j in 1:n_stream){
        zeta_stream[j] ~ dnorm(0, tau_zeta_stream)
        iota_stream[j] ~ dnorm(0, tau_iota_stream)
      }
    
      for(yr in 1:n_year){
        zeta_year[yr] ~ dnorm(0, tau_zeta_year)
      }
    
      for(k in 1:n_event){
        zeta_event[k] ~ dnorm(0, tau_zeta_event)
        iota_event[k] ~ dnorm(0, tau_iota_event)
      }
    
      for(c in 1:n_cluster){
        iota_cluster[c] ~ dnorm(0, tau_iota_cluster)
      }
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(0, tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # DETECTION PRIORS:
      
      for(k in 1:n_event){
      p_TP[k] ~ dnorm(0.76, pow(0.16,-2)) T(0,1)   # empirically derived recall
      p_FP[k] ~ dnorm(0.01, pow(0.005,-2)) T(0,1)  # empirically derived FPR
      }
      
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Presence/absence:
        z[i] ~ dbern(psi[i])
    
        logit(psi[i]) <- epsilon
                         + (zeta_damPres
                            + zeta_stream[stream[i]]
                            + zeta_year[year[i]]
                            + zeta_event[event[i]]) * damPres[i]
                         + zeta_incis * incis[i]
                         + zeta_RIPA * RIPA[i]
                         + zeta_UDA * UDA[i]
                         + iota_stream[stream[i]]
                         + iota_event[event[i]]
                         + iota_cluster[cluster[i]]
    
        # Detection:
        p_eff[i] <- z[i] * p_TP[event[i]] + (1-z[i]) * p_FP[event[i]]
        y[i]     ~ dbern(p_eff[i])
    
        # Amount:
        y2[i] ~ dnorm(mu[i], tau_eff[i])
    
        mu[i] <- z[i] *
                  (alpha
                    + (beta_damPres
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damPres[i]
                    + beta_incis * incis[i] 
                    + beta_RIPA * RIPA[i]
                    + beta_UDA * UDA[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]])
    
        tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
      }
    
    }
    ", fill = TRUE)
    
    sink()
    
    

```{r,echo=FALSE}
hurdle_damPres_data_list <- list(
  N = N,
  y = y,
  y2 = y2,
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damPres = damPres,
  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)),
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hurdle_damPres_inits <- function(){
  list(
    alpha        = rnorm(1, 0, 1),
    epsilon      = rnorm(1, 0, 1),

    sigma        = runif(1, 0.1, 2),
    sigma_zeta_stream  = runif(1, 0.1, 2),
    sigma_zeta_year    = runif(1, 0.1, 2),
    sigma_zeta_event   = runif(1, 0.1, 2),
    sigma_iota_stream  = runif(1, 0.1, 2),
    sigma_iota_event   = runif(1, 0.1, 2),
    sigma_iota_cluster = runif(1, 0.1, 2),

    sigma_beta_stream  = runif(1, 0.1, 2),
    sigma_beta_year    = runif(1, 0.1, 2),
    sigma_beta_event   = runif(1, 0.1, 2),
    sigma_eta_stream   = runif(1, 0.1, 2),
    sigma_eta_event    = runif(1, 0.1, 2),
    sigma_eta_cluster  = runif(1, 0.1, 2)
  )
}

hurdle_damPres_params <- c(
  "psi",
  "mu",
  "epsilon",
  "alpha",
  "zeta_damPres",
  "zeta_incis",
  "zeta_RIPA",
  "zeta_UDA",
  "zeta_stream",
  "zeta_year",
  "zeta_event",
  "iota_stream",
  "iota_event",
  "iota_cluster",
  "beta_damPres",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_zeta_stream",
  "sigma_zeta_year",
  "sigma_zeta_event",
  "sigma_iota_stream",
  "sigma_iota_event",
  "sigma_iota_cluster",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_event",
  "sigma_eta_cluster",
  "p_TP",
  "p_FP",
  "z",
  "tau_eff"
)

nchains <- 3
nthin <- 5
niter <- 100000
nburn <- 20000

Sys.time()

hurdle_damPres_SW <- jags(data = hurdle_damPres_data_list,
                   inits = hurdle_damPres_inits,
                   parameters.to.save = hurdle_damPres_params,
                   model.file = "HGLMM_hurdle_damPresence.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hurdle_damPres_SW) # model with local and watershed dam density as fixed effects
#print(hurdle_damPres_SW$summary[grep("beta_", rownames(hurdle_damPres_SW$summary)), ])
#print(hurdle_damPres_SW$summary[grep("zeta_", rownames(hurdle_damPres_SW$summary)), ])

```


----

DIAGNOSTICS:
----

```{r,echo=FALSE}
# Plotting specifications from Rmarkdown.

par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

```


```{r,echo=FALSE}
#Extract posteriors
damPres_post <- hurdle_damPres_SW$sims.list
```

**Assess "MCMC chain health":**

**Rubin-Gelman Rhat** -- assess convergence

**Effective Sample Size** -- assess posterior independence

```{r,echo=FALSE}
# Choose only the parameters you care about
params <- c(
  "epsilon",
  "alpha",
  "zeta_damPres",
  "beta_damPres",
  "zeta_incis",
  "beta_incis",
  "zeta_RIPA",
  "beta_RIPA",
  "zeta_UDA",
  "beta_UDA",
  "sigma",
  "sigma_zeta_stream",
  "sigma_iota_cluster",
  "p_TP[1]",
  "p_FP[1]")

# Gelman–Rubin convergence diagnostic
gelman.diag(hurdle_damPres_SW$samples[, params], multivariate = FALSE)

# Effective sample size
effectiveSize(hurdle_damPres_SW$samples[, params])

```



    Good Rhat values, all around 1.00.
    
    Mostly high effective sample sizes. 
    
    Note: beta_damPres has the most trouble converging; however, Rhat near 1, 
    and moderate ESS for beta_damPres suggest that the parameter is still 
    reliably usable. 


----

**Check traceplots for critical parameters**

```{r,echo=FALSE}
tplot_params <- c("alpha","beta_damPres","zeta_damPres", "sigma","sigma_zeta_stream","sigma_beta_stream")


mcmc_sub <- hurdle_damPres_SW$samples[, tplot_params]

coda::traceplot(mcmc_sub)
```


    Traceplots are all mostly grassy. Again, see beta_damPres. REAL slow to 
    converge; expected, given Rhat and ESS.


----

**Posterior predictive plots:**

Predicted Psi vs observed SW Presence

```{r,echo=FALSE}

par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

# Posterior mean psi
psi_post <- apply(hurdle_damPres_SW$sims.list$psi, 2, mean)

# Compare to observed y
plot(psi_post, y, pch=16, col=rgb(0,0,0,0.2))


```


Predicted SW vs detected SW

Goal is 1:1 (ish)

```{r,echo=FALSE}
mu_post <- apply(hurdle_damPres_SW$sims.list$mu, 2, mean)

plot(mu_post[y==1], y2[y==1], pch=16, col=rgb(0,0,1,0.3))
abline(0,1,col="red")


```

----

**Random effects shrinkage: **

Goal is centered around 0(ish)

1. zeta_stream; random effect of stream on sw presence
2. beta_stream; random effect of stream on sw amount
3. zeta_event; random effect of sample event on sw presence
4. beta_event; random effect of sample event on sw amount

```{r, echo=FALSE, }
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

boxplot(hurdle_damPres_SW$sims.list$zeta_stream,
        main = "zeta_stream")
boxplot(hurdle_damPres_SW$sims.list$beta_stream,
        main = "beta_stream")
boxplot(hurdle_damPres_SW$sims.list$zeta_event,
        main = "zeta_event")
boxplot(hurdle_damPres_SW$sims.list$beta_event,
        main = "beta_event")
```

<!---
Hyperpriors on stream effect: goal sd 0.1:2

```{r,echo=FALSE}

summary(hurdle_damPres_SW$sims.list$sigma_zeta_stream)
summary(hurdle_damPres_SW$sims.list$sigma_beta_stream)

```

Check detection posteriors: goal is near empirical mean recall

```{r,echo=FALSE}
summary(hurdle_damPres_SW$sims.list$p_TP)

summary(hurdle_damPres_SW$sims.list$p_FP)
```

    TP is high and precise; FP is low and precise. This makes sense because these are based on empirical mean() and sd() recall and flase positive rates from the RF classifier (see multiprobability threshod validation).


Check z latent states

```{r,echo=FALSE}
z_post <- apply(hurdle_damPres_SW$sims.list$z, 2, mean)

hist(z_post, main = "Split of posterior z")
```

    Latent z's split mostly evenly

Check tau_eff behavior:

```{r,echo=FALSE}
summary(hurdle_damPres_SW$sims.list$tau_eff)

```

    This shows that the tau_eff trick worked: 
    
    tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
    
    so when z is 1, tau_eff is reasonable. when z is 0, tau_eff is 100 and the prior distribution is basically flat

Check dam effects:
--->


----


**Posterior mean dam effects:**

damPres --> swPres (logistic regression)

```{r,echo=FALSE}
summary(hurdle_damPres_SW$sims.list$zeta_damPres)

exp(mean(hurdle_damPres_SW$sims.list$zeta_damPres))
```

    Dam presence increases the odds of surface‑water presence by about 6%. On the probability scale, this corresponds to only a ~1–2 percentage‑point increase from the baseline probability, indicating that dam presence has a negligible effect on whether water is present at all.
    
    
Log-odds to Probability transformation:

Psi_baseline:

$$1/(1+ e^{-0.68}) = 0.664$$ 
Psi_dam: 

$$1/(1+e^{-0.68 - 0.06}) = 1/(1+e^{-0.74}) = 0.677$$
Dam Effect on Psi:

$$Psi_{dam} - Psi_{baseline} = 0.677 - 0.664 = 0.013$$

**1.3 percetnage point increase on probability scale**

----


damPres --> swAmount (hglmm)

```{r,echo=FALSE}
summary(hurdle_damPres_SW$sims.list$beta_damPres)

exp(mean(hurdle_damPres_SW$sims.list$beta_damPres))
```

    Dam presence increases surface water area by ~ 20-25%.


----



**Prior-posterior overlap:**


```{r,echo=FALSE}
damPres_alpha_post <- hurdle_damPres_SW$sims.list$alpha
damPres_beta_post  <- hurdle_damPres_SW$sims.list$beta_damPres
damPres_zeta_post  <- hurdle_damPres_SW$sims.list$zeta_damPres
damPres_pTP_post <- hurdle_damPres_SW$sims.list$p_TP
damPres_pFP_post <- hurdle_damPres_SW$sims.list$p_FP

damPresMod_n <- length(damPres_alpha_post)

damPres_alpha_prior <- rnorm(damPresMod_n, 0, sqrt(1/0.04))
damPres_beta_prior  <- rnorm(damPresMod_n, 0, sqrt(1/0.04))
damPres_zeta_prior  <- rnorm(damPresMod_n, 0, sqrt(1/0.04))
damPres_pTP_prior <- rnorm(damPresMod_n, 0.76, sqrt(1/0.16))
damPres_pFP_prior <- rnorm(damPresMod_n, 0.01, sqrt(1/0.005))






# Plot overlap:

plot_prior_post <- function(prior, post, name) {
  
  df <- tibble(
    value = c(prior, post),
    dist  = rep(c("Prior", "Posterior"), each = length(prior))
  )
  
  ggplot(df, aes(x = value, fill = dist)) +
    geom_density(alpha = 0.4) +
    scale_fill_manual(values = c("gray60", "steelblue")) +
    labs(
      title = paste("Prior vs Posterior for", name),
      x = name,
      y = "Density"
    ) +
    theme_minimal(base_size = 14)
}

plot_prior_post(damPres_alpha_prior, damPres_alpha_post, "alpha")
plot_prior_post(damPres_beta_prior, damPres_beta_post, "beta")
plot_prior_post(damPres_zeta_prior, damPres_zeta_post, "zeta")
```


----


**Marginal effect** of dam presence on surface water presence, given geomorphic context.

**Note** there's a  TINY positive effect of dam presence on surface water presence. Not really perceptible from these plots. 

```{r,echo=FALSE}
# Build a grid of raw geomorphic covariate values
damPres_geom_grid <- tibble(
  TO_raw = seq(min(aoi_HurdleData$TO),
               max(aoi_HurdleData$TO), length.out = 50),

  incis_raw = 1 - (TO_raw / 90),   # raw incision scale

  RIPA_raw = seq(min(aoi_HurdleData$riparian_area_m2),
                 max(aoi_HurdleData$riparian_area_m2), length.out = 50),

  UDA_raw  = seq(min(aoi_HurdleData$catchment_km2),
                 max(aoi_HurdleData$catchment_km2), length.out = 50)
) %>%
  # Convert raw → standardized (to match model scale)
  mutate(
    incis = as.numeric(scale(incis_raw)),
    RIPA  = as.numeric(scale(RIPA_raw)),
    UDA   = as.numeric(scale(UDA_raw))
  )

damPres_compute_psi <- function(dam_flag, geom_grid, damPres_post) {

  eps <- mean(damPres_post$epsilon)
  zd  <- mean(damPres_post$zeta_damPres)
  zi  <- mean(damPres_post$zeta_incis)
  zr  <- mean(damPres_post$zeta_RIPA)
  zu  <- mean(damPres_post$zeta_UDA)

  geom_grid %>%
    mutate(
      psi = plogis(
        eps +
        zd * dam_flag +
        zi * incis +      # standardized incision
        zr * RIPA +       # standardized RIPA
        zu * UDA          # standardized UDA
      ),
      dam = dam_flag
    )
}


```

```{r,echo=FALSE}
damPres_psi_nodam <- damPres_compute_psi(0, damPres_geom_grid, damPres_post)
damPres_psi_dam   <- damPres_compute_psi(1, damPres_geom_grid, damPres_post)

damPres_psi_df <- bind_rows(damPres_psi_nodam, damPres_psi_dam)
```

```{r,echo=FALSE}
ggplot(damPres_psi_df, aes(x = incis_raw, y = psi, color = factor(dam))) +
  geom_line(size = 1.2) +
  labs(
    title = "Marginal Effect of Dam Presence on Surface Water Presence",
    subtitle = "Predicted probability of water (psi) across raw incision values",
    x = "Channel Incision (raw scale: 1 - TO/90)",
    y = "Probability of Surface Water",
    color = "Dam Presence"
  ) +
  scale_color_manual(values = c("steelblue","brown4"),
                     labels = c("Dam Present", "No Dam")) +
  theme_minimal(base_size = 14)

```

```{r,echo=FALSE}
ggplot(damPres_psi_df, aes(x = RIPA_raw, y = psi, color = factor(dam))) +
  geom_line(size = 1.2) +
  labs(
    title = "Marginal Effect of Dam Presence on Surface Water Presence",
    subtitle = "Predicted psi across raw riparian area (m^2)",
    x = "Riparian Area (m^2)",
    y = "Probability of Surface Water",
    color = "Dam Presence"
  ) +
  scale_color_manual(values = c("steelblue","brown4"),
                     labels = c("Dam Present", "No Dam")) +
  theme_minimal(base_size = 14)


```

```{r,echo=FALSE}
ggplot(damPres_psi_df, aes(x = UDA_raw, y = psi, color = factor(dam))) +
  geom_line(size = 1.2) +
  labs(
    title = "Marginal Effect of Dam Presence on Surface Water Presence",
    subtitle = "Predicted psi across raw contributing drainage area (km^2)",
    x = "Contributing Drainage Area (km^2)",
    y = "Probability of Surface Water",
    color = "Dam Presence"
  ) +
  scale_color_manual(values = c("steelblue","brown4"),
                     labels = c("Dam Present", "No Dam")) +
  theme_minimal(base_size = 14)


```


----


**Partial dependence plots**

Constrain PDP to empirical ranges for SWA and target covariates

```{r,echo=FALSE}
library(tidyverse)
library(scales)

damPres_post <- hurdle_damPres_SW$sims.list

# Raw covariates
incis_raw <- 1 - aoi_HurdleData$TO/90
RIPA_raw  <- aoi_HurdleData$riparian_area_m2
UDA_raw   <- aoi_HurdleData$catchment_km2

logSWA_mean <- mean(aoi_HurdleData$logSWA)
logSWA_sd   <- sd(aoi_HurdleData$logSWA)

# Standardization stats
incis_mean_raw <- mean(incis_raw)
incis_sd_raw   <- sd(incis_raw)

RIPA_mean_raw  <- mean(RIPA_raw)
RIPA_sd_raw    <- sd(RIPA_raw)

UDA_mean_raw   <- mean(UDA_raw)
UDA_sd_raw     <- sd(UDA_raw)


##############################3

# Helper Functions to Compute SWA from posterior draws
compute_swa_draws <- function(incis, RIPA, UDA, post, dam_flag) {
  n_draws <- length(damPres_post$alpha)
  swa_mat <- matrix(NA_real_, nrow = length(incis), ncol = n_draws)

  for (j in seq_len(n_draws)) {
    lp <- damPres_post$alpha[j] +
          damPres_post$beta_incis[j] * incis +
          damPres_post$beta_RIPA[j]  * RIPA +
          damPres_post$beta_UDA[j]   * UDA +
          damPres_post$beta_damPres[j] * dam_flag

    swa_mat[, j] <- exp(lp) - 1
  }

  swa_mat
}

summarize_swa <- function(swa_mat) {
  tibble(
    SWA_med = apply(swa_mat, 1, median),
    SWA_lo  = apply(swa_mat, 1, quantile, probs = 0.025),
    SWA_hi  = apply(swa_mat, 1, quantile, probs = 0.975)
  )
}




#########################################

# PDP Builder:

build_pdp <- function(focal_raw, focal_name,
                      incis_raw, RIPA_raw, UDA_raw,
                      post,
                      incis_mean_raw, incis_sd_raw,
                      RIPA_mean_raw, RIPA_sd_raw,
                      UDA_mean_raw, UDA_sd_raw) {

  # Sequence only across empirical range
  focal_seq <- seq(min(focal_raw), max(focal_raw), length.out = 200)

  # Conditional medians of other covariates
  incis_med <- median(incis_raw)
  RIPA_med  <- median(RIPA_raw)
  UDA_med   <- median(UDA_raw)

  # Build grid (200 rows)
  grid <- tibble(
    focal_raw = focal_seq,
    incis_raw = if (focal_name == "incis") focal_seq else incis_med,
    RIPA_raw  = if (focal_name == "RIPA")  focal_seq else RIPA_med,
    UDA_raw   = if (focal_name == "UDA")   focal_seq else UDA_med
  ) %>%
    mutate(
      incis = (incis_raw - incis_mean_raw) / incis_sd_raw,
      RIPA  = (RIPA_raw  - RIPA_mean_raw)  / RIPA_sd_raw,
      UDA   = (UDA_raw   - UDA_mean_raw)   / UDA_sd_raw
    )

  # Compute SWA draws
  swa_nodam <- compute_swa_draws(grid$incis, grid$RIPA, grid$UDA, post, dam_flag = 0)
  swa_dam   <- compute_swa_draws(grid$incis, grid$RIPA, grid$UDA, post, dam_flag = 1)

  # Summaries (400 rows)
  pdp <- bind_rows(
    summarize_swa(swa_nodam) %>% mutate(dam = "No Dam"),
    summarize_swa(swa_dam)   %>% mutate(dam = "Dam Present")
  )

  # Repeat grid twice (400 rows)
  grid2 <- bind_rows(grid, grid)

  # Bind columns safely
  bind_cols(grid2, pdp)
}



#####

pdp_incis <- build_pdp(
  focal_raw = incis_raw,
  focal_name = "incis",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)

pdp_RIPA <- build_pdp(
  focal_raw = RIPA_raw,
  focal_name = "RIPA",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)

pdp_UDA <- build_pdp(
  focal_raw = UDA_raw,
  focal_name = "UDA",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)



##########################################3

# Plot Function:

plot_pdp <- function(pdp, xvar, xtitle) {
  ggplot(pdp,
         aes(x = .data[[xvar]], y = SWA_med, color = dam, fill = dam)) +
    geom_ribbon(aes(ymin = SWA_lo, ymax = SWA_hi), alpha = 0.2, color = NA) +
    geom_line(size = 1.2) +
    scale_color_manual(values = c("steelblue", "brown4")) +
    scale_fill_manual(values = c("steelblue", "brown4")) +
    scale_y_continuous(labels = comma) +
    labs(
      title = paste("Partial Dependence of SWA on", xtitle),
      subtitle = "Predicted SWA with 95% credible intervals\nConstrained to empirical covariate range",
      x = xtitle,
      y = "Predicted SWA (m²)",
      color = "Dam Presence",
      fill  = "Dam Presence"
    ) +
    theme_minimal(base_size = 14)
}


###########################################3


plot_pdp(pdp_incis, "incis_raw", "Channel Incision (raw)")
plot_pdp(pdp_RIPA,  "RIPA_raw",  "Riparian Area (m²)")
plot_pdp(pdp_UDA,   "UDA_raw",   "Contributing Drainage Area (km²)")


```


----


NEED TO COMPLETE:
Visualize variation in dam effect among streams and seasons:

- heatmap
- line graph



----

----

----

----


Dam Density
----


**Hurdle HGLMM**

```{r,echo=FALSE}
#library(jagsUI)
```
    sink("HGLMM_hurdle_damDensity.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR SW PRESENCE (1/2 Normal on SDs):
    
      sigma_zeta_stream  ~ dnorm(0,1) T(0,)
      tau_zeta_stream    <- pow(sigma_zeta_stream, -2)
    
      sigma_zeta_year    ~ dnorm(0,1) T(0,)
      tau_zeta_year      <- pow(sigma_zeta_year, -2)
    
      sigma_zeta_event   ~ dnorm(0,1) T(0,)
      tau_zeta_event     <- pow(sigma_zeta_event, -2)
    
      sigma_iota_stream  ~ dnorm(0,1) T(0,)
      tau_iota_stream    <- pow(sigma_iota_stream, -2)
    
      sigma_iota_event   ~ dnorm(0,1) T(0,)
      tau_iota_event     <- pow(sigma_iota_event, -2)
    
      sigma_iota_cluster ~ dnorm(0,1) T(0,)
      tau_iota_cluster   <- pow(sigma_iota_cluster, -2)
    
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 0.7) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 0.7) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 0.7) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Presence/absence:
      epsilon ~ dnorm(0,0.04)
      zeta_damDen ~ dnorm(0,0.04)
      zeta_incis ~ dnorm(0,0.04)
      zeta_RIPA ~ dnorm(0,0.04)
      zeta_UDA ~ dnorm(0,0.04)
    
      # Amount:
      alpha ~ dnorm(0,0.04)
      beta_damDen ~ dnorm(0,0.04)
      beta_incis ~ dnorm(0,0.04)
      beta_RIPA ~ dnorm(0,0.04)
      beta_UDA ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW PRESENCE:
    
      for(j in 1:n_stream){
        zeta_stream[j] ~ dnorm(0, tau_zeta_stream)
        iota_stream[j] ~ dnorm(0, tau_iota_stream)
      }
    
      for(yr in 1:n_year){
        zeta_year[yr] ~ dnorm(0, tau_zeta_year)
      }
    
      for(k in 1:n_event){
        zeta_event[k] ~ dnorm(0, tau_zeta_event)
        iota_event[k] ~ dnorm(0, tau_iota_event)
      }
    
      for(c in 1:n_cluster){
        iota_cluster[c] ~ dnorm(0, tau_iota_cluster)
      }
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(0, tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # DETECTION PRIORS:
      
      for(k in 1:n_event){
      p_TP[k] ~ dnorm(0.76, pow(0.16,-2)) T(0,1)   # empirically derived recall
      p_FP[k] ~ dnorm(0.01, pow(0.005,-2)) T(0,1)  # empirically derived FPR
      }
      
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Presence/absence:
        z[i] ~ dbern(psi[i])
    
        logit(psi[i]) <- epsilon
                         + (zeta_damDen
                            + zeta_stream[stream[i]]
                            + zeta_year[year[i]]
                            + zeta_event[event[i]]) * damDen[i]
                         + zeta_incis * incis[i]
                         + zeta_RIPA * RIPA[i]
                         + zeta_UDA * UDA[i]
                         + iota_stream[stream[i]]
                         + iota_event[event[i]]
                         + iota_cluster[cluster[i]]
    
        # Detection:
        p_eff[i] <- z[i] * p_TP[event[i]] + (1-z[i]) * p_FP[event[i]]
        y[i]     ~ dbern(p_eff[i])
    
        # Amount:
        y2[i] ~ dnorm(mu[i], tau_eff[i])
    
        mu[i] <- z[i] *
                  (alpha
                    + (beta_damDen
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damDen[i]
                    + beta_incis * incis[i] 
                    + beta_RIPA * RIPA[i]
                    + beta_UDA * UDA[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]])
    
        tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
      }
    
    }
    ", fill = TRUE)
    
    sink()
   
    

```{r,echo=FALSE}
hurdle_damDen_data_list <- list(
  N = N,
  y = y,
  y2 = y2,
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damDen = damDen,
  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)),
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hurdle_damDen_inits <- function(){
  list(
    alpha        = rnorm(1, 0, 1),
    epsilon      = rnorm(1, 0, 1),

    sigma        = runif(1, 0.1, 2),
    sigma_zeta_stream  = runif(1, 0.1, 2),
    sigma_zeta_year    = runif(1, 0.1, 2),
    sigma_zeta_event   = runif(1, 0.1, 2),
    sigma_iota_stream  = runif(1, 0.1, 2),
    sigma_iota_event   = runif(1, 0.1, 2),
    sigma_iota_cluster = runif(1, 0.1, 2),

    sigma_beta_stream  = runif(1, 0.1, 2),
    sigma_beta_year    = runif(1, 0.1, 2),
    sigma_beta_event   = runif(1, 0.1, 2),
    sigma_eta_stream   = runif(1, 0.1, 2),
    sigma_eta_event    = runif(1, 0.1, 2),
    sigma_eta_cluster  = runif(1, 0.1, 2)
  )
}

hurdle_damDen_params <- c(
  "psi",
  "mu",
  "epsilon",
  "alpha",
  "zeta_damDen",
  "zeta_incis",
  "zeta_RIPA",
  "zeta_UDA",
  "zeta_stream",
  "zeta_year",
  "zeta_event",
  "iota_stream",
  "iota_event",
  "iota_cluster",
  "beta_damDen",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_zeta_stream",
  "sigma_zeta_year",
  "sigma_zeta_event",
  "sigma_iota_stream",
  "sigma_iota_event",
  "sigma_iota_cluster",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_event",
  "sigma_eta_cluster",
  "p_TP",
  "p_FP",
  "z",
  "tau_eff"
)

nchains <- 3
nthin <- 5
niter <- 100000
nburn <- 20000


Sys.time()

hurdle_damDen_SW <- jags(data = hurdle_damDen_data_list,
                   inits = hurdle_damDen_inits,
                   parameters.to.save = hurdle_damDen_params,
                   model.file = "HGLMM_hurdle_damDensity.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hurdle_damPres_SW) # model with local and watershed dam density as fixed effects
#print(hurdle_damPres_SW$summary[grep("beta_", rownames(hurdle_damPres_SW$summary)), ])
#print(hurdle_damPres_SW$summary[grep("zeta_", rownames(hurdle_damPres_SW$summary)), ])

```


----


DIAGNOSTICS:
----


```{r,echo=FALSE}
# Extract posterior draws

damDen_post <- hurdle_damDen_SW$sims.list 
```

**Assess "Chain health" Rhat + ESS:**

```{r,echo=FALSE} 
# Choose only the parameters you care about
params_2 <- c(
  "epsilon",
  "alpha",
  "zeta_damDen",
  "beta_damDen",
  "zeta_incis",
  "beta_incis",
  "zeta_RIPA",
  "beta_RIPA",
  "zeta_UDA",
  "beta_UDA",
  "sigma",
  "sigma_zeta_stream",
  "sigma_iota_cluster",
  "p_TP[1]",
  "p_FP[1]")

# Gelman–Rubin convergence diagnostic
gelman.diag(hurdle_damDen_SW$samples[, params_2], multivariate = FALSE)

# Effective sample size
effectiveSize(hurdle_damDen_SW$samples[, params_2])

```



    Good Rhat values, all around 1.00.
    
    Mostly high effective sample sizes. Note: beta_damDen has the most trouble converging (unsurprising). 


----

**Check traceplots** for critical parameters:

```{r,echo=FALSE} 

par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


tplot_params_2 <- c("alpha","beta_damDen","zeta_damDen", "sigma","sigma_zeta_stream","sigma_beta_stream")


mcmc_sub_2 <- hurdle_damDen_SW$samples[, tplot_params_2]

coda::traceplot(mcmc_sub_2)
```


    Traceplots all look good, except for beta_damDen, which is mostly grassy, but 
    slow to converge - suggesting that its difficult for the model to estimate. 


----

**Posterior predictive plots:**

Predicted Psi vs observed SW presence

```{r,echo=FALSE} 
par(mfrow = c(1,1),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit

# Posterior mean psi
psi_post_2 <- apply(hurdle_damDen_SW$sims.list$psi, 2, mean)

# Compare to observed y
plot(psi_post_2, y, pch=16, col=rgb(0,0,0,0.2))


```

----

Predicted SW vs observed SW

Goal is 1:1 (ish)

```{r,echo=FALSE} 
mu_post_2 <- apply(hurdle_damDen_SW$sims.list$mu, 2, mean)

plot(mu_post_2[y==1], y2[y==1], pch=16, col=rgb(0,0,1,0.3))
abline(0,1,col="red")


```

----

**Random effects shrinkage: **

Goal is centered around 0(ish)

1. zeta_stream; random effect of stream on sw presence
2. beta_stream; random effect of stream on sw amount
3. zeta_event; random effect of sample event on sw presence
4. beta_event; random effect of sample event on sw amount

```{r, echo=FALSE} 
par(mfrow = c(2, 2),        # 2 rows, 2 columns
    mar = c(4, 4, 2, 1),    # smaller margins
    cex = 0.8)              # shrink text/points a bit


boxplot(hurdle_damDen_SW$sims.list$zeta_stream)
boxplot(hurdle_damDen_SW$sims.list$beta_stream)
boxplot(hurdle_damDen_SW$sims.list$zeta_event)
boxplot(hurdle_damDen_SW$sims.list$beta_event)
```


----

----


**Posterior mean dam effects:**

damDen --> swPres

```{r,echo=FALSE}
summary(hurdle_damDen_SW$sims.list$zeta_damDen)

exp(mean(hurdle_damDen_SW$sims.list$zeta_damDen))
```

    1 std dev. increase in dam density increases odds of surface water 
    presence by ~ 0.03. This corresponds to ~0.5 - 1 percentage-point increase 
    on the probability scale. 
    
    1 std dev. of dam density from raw data is about 3.14 dams
    
    Massive overlap of 0. Really no major effect of 1 std dev increase in dam 
    density on surface water presence

----

damDen --> swAmount

```{r,echo=FALSE}
summary(hurdle_damDen_SW$sims.list$beta_damDen)

exp(mean(hurdle_damDen_SW$sims.list$beta_damDen))
```

    Dam presence increases surface water area by ~ 13%.


----

**Prior-posterior overlap:**

```{r,echo=FALSE}
damDen_alpha_post <- hurdle_damDen_SW$sims.list$alpha
damDen_beta_post  <- hurdle_damDen_SW$sims.list$beta_damDen
damDen_zeta_post  <- hurdle_damDen_SW$sims.list$zeta_damDen
damDen_pTP_post <- hurdle_damDen_SW$sims.list$p_TP
damDen_pFP_post <- hurdle_damDen_SW$sims.list$p_FP

damDenMod_n <- length(damDen_alpha_post)

damDen_alpha_prior <- rnorm(damDenMod_n, 0, sqrt(1/0.04))
damDen_beta_prior  <- rnorm(damDenMod_n, 0, sqrt(1/0.04))
damDen_zeta_prior  <- rnorm(damDenMod_n, 0, sqrt(1/0.04))
damDen_pTP_prior <- rnorm(damDenMod_n, 0.76, sqrt(1/0.16))
damDen_pFP_prior <- rnorm(damDenMod_n, 0.01, sqrt(1/0.005))






# Plot overlap:

plot_prior_post <- function(prior, post, name) {
  
  df <- tibble(
    value = c(prior, post),
    dist  = rep(c("Prior", "Posterior"), each = length(prior))
  )
  
  ggplot(df, aes(x = value, fill = dist)) +
    geom_density(alpha = 0.4) +
    scale_fill_manual(values = c("gray60", "steelblue")) +
    labs(
      title = paste("Prior vs Posterior for", name),
      x = name,
      y = "Density"
    ) +
    theme_minimal(base_size = 14)
}

plot_prior_post(damDen_alpha_prior, damPres_alpha_post, "alpha")
plot_prior_post(damDen_beta_prior, damPres_beta_post, "beta")
plot_prior_post(damDen_zeta_prior, damPres_zeta_post, "zeta")
```


----

**Marginal effect** of dam Density on surface water presence, given geomorphic context

**Note** there's a small but negligble positive effect of a 1 stdDev increase in DamDensity on surface water presence. However, higher dam densities do increase probability of SW presence more notably (range is up to +4 standard dev. from mean) 

```{r,echo=FALSE}
# Build a grid of raw geomorphic covariate values
damDen_geom_grid <- tibble(
  TO_raw = seq(min(aoi_HurdleData$TO),
               max(aoi_HurdleData$TO), length.out = 50),

  incis_raw = 1 - (TO_raw / 90),   # raw incision scale

  RIPA_raw = seq(min(aoi_HurdleData$riparian_area_m2),
                 max(aoi_HurdleData$riparian_area_m2), length.out = 50),

  UDA_raw  = seq(min(aoi_HurdleData$catchment_km2),
                 max(aoi_HurdleData$catchment_km2), length.out = 50)
) %>%
  # Convert raw → standardized (to match model scale)
  mutate(
    incis = as.numeric(scale(incis_raw)),
    RIPA  = as.numeric(scale(RIPA_raw)),
    UDA   = as.numeric(scale(UDA_raw))
  )

# Make dam density sequence
damDen_seq <- seq(min(damDen_raw),max(damDen_raw),length.out=10)

# Compute surface water across dam density sequence
damDen_compute_psi <- function(damDen_val, geom_grid, damDen_post) {

  eps <- mean(damDen_post$epsilon)
  zd  <- mean(damDen_post$zeta_damDen)
  zi  <- mean(damDen_post$zeta_incis)
  zr  <- mean(damDen_post$zeta_RIPA)
  zu  <- mean(damDen_post$zeta_UDA)

  geom_grid %>%
    mutate(
      psi = plogis(
        eps +
        zd * damDen_val +
        zi * incis +      # standardized incision
        zr * RIPA +       # standardized RIPA
        zu * UDA          # standardized UDA
      ),
      damDen = damDen_val
    )
}


```

```{r,echo=FALSE}
damDen_psi_df <- map_df(damDen_seq,
                        ~ damDen_compute_psi(.x,
                                             damDen_geom_grid,
                                             damDen_post))
```

```{r,echo=FALSE}
ggplot(damDen_psi_df, aes(x = incis_raw, y = psi, color = damDen)) +
  geom_point() +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "Marginal Effect of Dam Density on Surface Water Presence",
    subtitle = "Predicted probability of water (psi) across raw incision values",
    x = "Channel Incision (raw scale: 1 - TO/90)",
    y = "Probability of Surface Water",
    color = "Dam Density (z-std)"
  ) +
  theme_minimal(base_size = 14)

```

```{r,echo=FALSE}
ggplot(damDen_psi_df, aes(x = RIPA_raw, y = psi, color = damDen)) +
  geom_point() +
  scale_color_viridis_c(option="plasma") +
  labs(
    title = "Marginal Effect of Dam Density on Surface Water Presence",
    subtitle = "Predicted psi across raw riparian area (m^2)",
    x = "Riparian Area (m^2)",
    y = "Probability of Surface Water",
    color = "Dam Density (z-std)"
  ) +
  theme_minimal(base_size = 14)
```

```{r,echo=FALSE}
ggplot(damDen_psi_df, aes(x = UDA_raw, y = psi, color = damDen)) +
  geom_point() +
  scale_color_viridis_c(option="plasma") +
  labs(
    title = "Marginal Effect of Dam Density on Surface Water Presence",
    subtitle = "Predicted psi across raw upstream drainage area (km^2)",
    x = "Upstream Drainage Area (km^2)",
    y = "Probability of Surface Water",
    color = "Dam Density (z-std)"
  ) +
  theme_minimal(base_size = 14)


```

----

**Partial dependence plots**

Constrain PDP to empirical ranges for SWA and target covariates

```{r,echo=FALSE}
library(tidyverse)
library(scales)

damDen_post <- hurdle_damDen_SW$sims.list

# Raw covariates
incis_raw <- 1 - aoi_HurdleData$TO/90
RIPA_raw  <- aoi_HurdleData$riparian_area_m2
UDA_raw   <- aoi_HurdleData$catchment_km2
#damDen <- aoi_HurdleData$dam_km_std
damDen_raw <- aoi_HurdleData$dam_km
damDen_mean_raw <- mean(damDen_raw)
damDen_sd_raw   <- sd(damDen_raw)

logSWA_mean <- mean(aoi_HurdleData$logSWA)
logSWA_sd   <- sd(aoi_HurdleData$logSWA)

# Standardization stats
incis_mean_raw <- mean(incis_raw)
incis_sd_raw   <- sd(incis_raw)

RIPA_mean_raw  <- mean(RIPA_raw)
RIPA_sd_raw    <- sd(RIPA_raw)

UDA_mean_raw   <- mean(UDA_raw)
UDA_sd_raw     <- sd(UDA_raw)


##############################3

# Helper Functions to Compute SWA from posterior draws
compute_swa_draws_damDen <- function(incis, RIPA, UDA, post, damDen_val) {
  n_draws <- length(post$alpha)
  swa_mat <- matrix(NA_real_, nrow = length(incis), ncol = n_draws)

  for (j in seq_len(n_draws)) {
    lp <- post$alpha[j] +
          post$beta_incis[j] * incis +
          post$beta_RIPA[j]  * RIPA +
          post$beta_UDA[j]   * UDA +
          post$beta_damDen[j] * damDen_val

    swa_mat[, j] <- exp(lp) - 1
  }

  swa_mat
}

summarize_swa <- function(swa_mat) {
  tibble(
    SWA_med = apply(swa_mat, 1, median),
    SWA_lo  = apply(swa_mat, 1, quantile, probs = 0.025),
    SWA_hi  = apply(swa_mat, 1, quantile, probs = 0.975)
  )
}




#########################################

# PDP Builder:

build_pdp_damDen <- function(focal_raw, focal_name,
                      incis_raw, RIPA_raw, UDA_raw,
                      post,
                      incis_mean_raw, incis_sd_raw,
                      RIPA_mean_raw, RIPA_sd_raw,
                      UDA_mean_raw, UDA_sd_raw,
                      damDen_raw) {

  # Sequence only across empirical range
  focal_seq <- seq(min(focal_raw), max(focal_raw), length.out = 200)

  # Conditional medians of other covariates
  incis_med <- median(incis_raw)
  RIPA_med  <- median(RIPA_raw)
  UDA_med   <- median(UDA_raw)
  
  # Empirical dam density range (standardized)
  damDen_seq <- seq(min(damDen_raw), max(damDen_raw), length.out = 10)

  # Build grid (200 rows)
  grid <- tibble(
    focal_raw = focal_seq,
    incis_raw = if (focal_name == "incis") focal_seq else incis_med,
    RIPA_raw  = if (focal_name == "RIPA")  focal_seq else RIPA_med,
    UDA_raw   = if (focal_name == "UDA")   focal_seq else UDA_med
  ) %>%
    mutate(
      incis = (incis_raw - incis_mean_raw) / incis_sd_raw,
      RIPA  = (RIPA_raw  - RIPA_mean_raw)  / RIPA_sd_raw,
      UDA   = (UDA_raw   - UDA_mean_raw)   / UDA_sd_raw
    )
  # Compute PDP curves for each dam density value
  pdp_list <- map(damDen_seq, function(dd) {
    swa_mat <- compute_swa_draws_damDen(grid$incis, grid$RIPA, grid$UDA, post,
                                        (dd - damDen_mean_raw) / damDen_sd_raw)
    summarize_swa(swa_mat) %>% mutate(damDen = dd)
  })

  pdp <- bind_rows(pdp_list)

  # Repeat grid for each damDen value
  grid2 <- bind_rows(replicate(length(damDen_seq), grid, simplify = FALSE))

  bind_cols(grid2, pdp)
}



#####

pdp_incis_damDen <- build_pdp_damDen(
  focal_raw = incis_raw,
  focal_name = "incis",
  incis_raw, RIPA_raw, UDA_raw,
  damDen_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw, damDen_raw
)

pdp_RIPA_damDen <- build_pdp_damDen(
  focal_raw = RIPA_raw,
  focal_name = "RIPA",
  incis_raw, RIPA_raw, UDA_raw,
  damDen_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw, damDen_raw
)

pdp_UDA_damDen <- build_pdp_damDen(
  focal_raw = UDA_raw,
  focal_name = "UDA",
  incis_raw, RIPA_raw, UDA_raw,
  damDen_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw, damDen_raw
)



##########################################3

# Plot Function:

plot_pdp_damDen <- function(pdp, xvar, xtitle) {
  ggplot(pdp,
         aes(x = .data[[xvar]], y = SWA_med,
             color = damDen, fill = damDen)) +
    #geom_ribbon(aes(ymin = SWA_lo, ymax = SWA_hi), alpha = 0.15, color = NA) +
    geom_point() +
    scale_color_viridis_c(option = "plasma") +
    scale_fill_viridis_c(option = "plasma") +
    scale_y_continuous(labels = comma) +
    labs(
      title = paste("Partial Dependence of SWA on", xtitle),
      subtitle = "Across empirical dam density range",
      x = xtitle,
      y = "Predicted SWA (m^2)",
      color = "Dam Density (dams/km)",
      fill  = "Dam Density (dams/km)"
    ) +
    theme_minimal(base_size = 14)
}



###########################################3


plot_pdp_damDen(pdp_incis_damDen, "incis_raw", "Channel Incision (raw)")
plot_pdp_damDen(pdp_RIPA_damDen,  "RIPA_raw",  "Riparian Area (m²)")
plot_pdp_damDen(pdp_UDA_damDen,   "UDA_raw",   "Contributing Drainage Area (km²)")


```

 



----
----

----
----

----
----

----
----

----
----

----
----

----
----

----
----

----
----

----
----

----
----

----
----

----
----

----
----

----
----

----
----

----
----

----
----

----
----

----
----

<!---

    This section explores late summner vegetation (veg area (m^2) and NDVI) in 
    relation to beaver dam presence.
    
    Uses hurdle model with nested random slopes for beaver dam effect to allow 
    for stream, year, 
    and sampleEvent to cause deviation from the global mean beaver effect.
    
    Pt1: Surface water presence/absence; logistic regression
    Pt2: Surface water amount; log(Normal) regression
    
    Uses log(SWA) for AOI all years. Adjust priors accordingly...

----
----

**Hurdle HGLMM**
----

```{r,echo=FALSE}
library(jagsUI)
```
    sink("HGLMM_hurdle_damPresence.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR SW PRESENCE (1/2 Normal on SDs):
    
      sigma_zeta_stream  ~ dnorm(0,1) T(0,)
      tau_zeta_stream    <- pow(sigma_zeta_stream, -2)
    
      sigma_zeta_year    ~ dnorm(0,1) T(0,)
      tau_zeta_year      <- pow(sigma_zeta_year, -2)
    
      sigma_zeta_event   ~ dnorm(0,1) T(0,)
      tau_zeta_event     <- pow(sigma_zeta_event, -2)
    
      sigma_iota_stream  ~ dnorm(0,1) T(0,)
      tau_iota_stream    <- pow(sigma_iota_stream, -2)
    
      sigma_iota_event   ~ dnorm(0,1) T(0,)
      tau_iota_event     <- pow(sigma_iota_event, -2)
    
      sigma_iota_cluster ~ dnorm(0,1) T(0,)
      tau_iota_cluster   <- pow(sigma_iota_cluster, -2)
    
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 0.7) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 0.7) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 0.7) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Presence/absence:
      epsilon ~ dnorm(0,0.04)
      zeta_damPres ~ dnorm(0,0.04)
      zeta_incis ~ dnorm(0,0.04)
      zeta_RIPA ~ dnorm(0,0.04)
      zeta_UDA ~ dnorm(0,0.04)
    
      # Amount:
      alpha ~ dnorm(0,0.04)
      beta_damPres ~ dnorm(0,0.04)
      beta_incis ~ dnorm(0,0.04)
      beta_RIPA ~ dnorm(0,0.04)
      beta_UDA ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW PRESENCE:
    
      for(j in 1:n_stream){
        zeta_stream[j] ~ dnorm(0, tau_zeta_stream)
        iota_stream[j] ~ dnorm(0, tau_iota_stream)
      }
    
      for(yr in 1:n_year){
        zeta_year[yr] ~ dnorm(0, tau_zeta_year)
      }
    
      for(k in 1:n_event){
        zeta_event[k] ~ dnorm(0, tau_zeta_event)
        iota_event[k] ~ dnorm(0, tau_iota_event)
      }
    
      for(c in 1:n_cluster){
        iota_cluster[c] ~ dnorm(0, tau_iota_cluster)
      }
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(0, tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # DETECTION PRIORS:
      
      for(k in 1:n_event){
      p_TP[k] ~ dnorm(0.76, pow(0.16,-2)) T(0,1)   # empirically derived recall
      p_FP[k] ~ dnorm(0.01, pow(0.005,-2)) T(0,1)  # empirically derived FPR
      }
      
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Presence/absence:
        z[i] ~ dbern(psi[i])
    
        logit(psi[i]) <- epsilon
                         + (zeta_damPres
                            + zeta_stream[stream[i]]
                            + zeta_year[year[i]]
                            + zeta_event[event[i]]) * damPres[i]
                         + zeta_incis * incis[i]
                         + zeta_RIPA * RIPA[i]
                         + zeta_UDA * UDA[i]
                         + iota_stream[stream[i]]
                         + iota_event[event[i]]
                         + iota_cluster[cluster[i]]
    
        # Detection:
        p_eff[i] <- z[i] * p_TP[event[i]] + (1-z[i]) * p_FP[event[i]]
        y[i]     ~ dbern(p_eff[i])
    
        # Amount:
        y2[i] ~ dnorm(mu[i], tau_eff[i])
    
        mu[i] <- z[i] *
                  (alpha
                    + (beta_damPres
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damPres[i]
                    + beta_incis * incis[i] 
                    + beta_RIPA * RIPA[i]
                    + beta_UDA * UDA[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]])
    
        tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
      }
    
    }
    ", fill = TRUE)
    
    sink()
    
    

```{r,echo=FALSE}
hurdle_damPres_data_list <- list(
  N = N,
  y = y,
  y2 = y2,
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damPres = damPres,
  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)),
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hurdle_damPres_inits <- function(){
  list(
    alpha        = rnorm(1, 0, 1),
    epsilon      = rnorm(1, 0, 1),

    sigma        = runif(1, 0.1, 2),
    sigma_zeta_stream  = runif(1, 0.1, 2),
    sigma_zeta_year    = runif(1, 0.1, 2),
    sigma_zeta_event   = runif(1, 0.1, 2),
    sigma_iota_stream  = runif(1, 0.1, 2),
    sigma_iota_event   = runif(1, 0.1, 2),
    sigma_iota_cluster = runif(1, 0.1, 2),

    sigma_beta_stream  = runif(1, 0.1, 2),
    sigma_beta_year    = runif(1, 0.1, 2),
    sigma_beta_event   = runif(1, 0.1, 2),
    sigma_eta_stream   = runif(1, 0.1, 2),
    sigma_eta_event    = runif(1, 0.1, 2),
    sigma_eta_cluster  = runif(1, 0.1, 2)
  )
}

hurdle_damPres_params <- c(
  "psi",
  "mu",
  "epsilon",
  "alpha",
  "zeta_damPres",
  "zeta_incis",
  "zeta_RIPA",
  "zeta_UDA",
  "zeta_stream",
  "zeta_year",
  "zeta_event",
  "iota_stream",
  "iota_event",
  "iota_cluster",
  "beta_damPres",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_zeta_stream",
  "sigma_zeta_year",
  "sigma_zeta_event",
  "sigma_iota_stream",
  "sigma_iota_event",
  "sigma_iota_cluster",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_event",
  "sigma_eta_cluster",
  "p_TP",
  "p_FP",
  "z",
  "tau_eff"
)

nchains <- 3
nthin <- 5
niter <- 100000
nburn <- 20000

hurdle_damPres_SW <- jags(data = hurdle_damPres_data_list,
                   inits = hurdle_damPres_inits,
                   parameters.to.save = hurdle_damPres_params,
                   model.file = "HGLMM_hurdle_damPresence.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hurdle_damPres_SW) # model with local and watershed dam density as fixed effects
#print(hurdle_damPres_SW$summary[grep("beta_", rownames(hurdle_damPres_SW$summary)), ])
#print(hurdle_damPres_SW$summary[grep("zeta_", rownames(hurdle_damPres_SW$summary)), ])

```


----
----

DIAGNOSTICS:

Extract posteriors:
```{r,echo=FALSE}
damPres_post <- hurdle_damPres_SW$sims.list
```

"Chain health" Rhat + ESS:
```{r,echo=FALSE}
# Choose only the parameters you care about
params <- c(
  "epsilon",
  "alpha",
  "zeta_damPres",
  "beta_damPres",
  "zeta_incis",
  "beta_incis",
  "zeta_RIPA",
  "beta_RIPA",
  "zeta_UDA",
  "beta_UDA",
  "sigma",
  "sigma_zeta_stream",
  "sigma_iota_cluster",
  "p_TP[1]",
  "p_FP[1]")

# Gelman–Rubin convergence diagnostic
gelman.diag(hurdle_damPres_SW$samples[, params], multivariate = FALSE)

# Effective sample size
effectiveSize(hurdle_damPres_SW$samples[, params])

```

----

    Good Rhat values, all around 1.00.
    
    Mostly high effective sample sizes. Note: beta_damPres has the most trouble converging. 


----

**Check traceplots** for critical parameters:

```{r,echo=FALSE}
tplot_params <- c("alpha","beta_damPres","zeta_damPres", "sigma","sigma_zeta_stream","sigma_beta_stream")


mcmc_sub <- hurdle_damPres_SW$samples[, tplot_params]

coda::traceplot(mcmc_sub)
```

----

    Traceplots are all mostly grassy. Again, see beta_damPres. Note: had to updated iterations, burnins, and thin rate to get convergence. 

----

**Posterior predictive plots:**

Predicted Psi vs observed SW Presence

```{r,echo=FALSE}
# Posterior mean psi
psi_post <- apply(hurdle_damPres_SW$sims.list$psi, 2, mean)

# Compare to observed y
plot(psi_post, y, pch=16, col=rgb(0,0,0,0.2))


```

```{r,echo=FALSE}
mu_post <- apply(hurdle_damPres_SW$sims.list$mu, 2, mean)

plot(mu_post[y==1], y2[y==1], pch=16, col=rgb(0,0,1,0.3))
abline(0,1,col="red")


```


Random effects shrinkage: 

goal is centered around 0(ish)

1. zeta_stream; random effect of stream on sw presence
2. beta_stream; random effect of stream on sw amount
3. zeta_event; random effect of sample event on sw presence
4. beta_event; random effect of sample event on sw amount

```{r, echo=FALSE}
boxplot(hurdle_damPres_SW$sims.list$zeta_stream)
boxplot(hurdle_damPres_SW$sims.list$beta_stream)
boxplot(hurdle_damPres_SW$sims.list$zeta_event)
boxplot(hurdle_damPres_SW$sims.list$beta_event)
```

<!---
Hyperpriors on stream effect: goal sd 0.1:2

```{r,echo=FALSE}

summary(hurdle_damPres_SW$sims.list$sigma_zeta_stream)
summary(hurdle_damPres_SW$sims.list$sigma_beta_stream)

```

Check detection posteriors: goal is near empirical mean recall

```{r,echo=FALSE}
summary(hurdle_damPres_SW$sims.list$p_TP)

summary(hurdle_damPres_SW$sims.list$p_FP)
```

    TP is high and precise; FP is low and precise. This makes sense because these are based on empirical mean() and sd() recall and flase positive rates from the RF classifier (see multiprobability threshod validation).


Check z latent states

```{r,echo=FALSE}
z_post <- apply(hurdle_damPres_SW$sims.list$z, 2, mean)

hist(z_post, main = "Split of posterior z")
```

    Latent z's split mostly evenly

Check tau_eff behavior:

```{r,echo=FALSE}
summary(hurdle_damPres_SW$sims.list$tau_eff)

```

    This shows that the tau_eff trick worked: 
    
    tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
    
    so when z is 1, tau_eff is reasonable. when z is 0, tau_eff is 100 and the prior distribution is basically flat

Check dam effects:


Posterior mean dam effects:

damPres -> swPres

```{r,echo=FALSE}
summary(hurdle_damPres_SW$sims.list$zeta_damPres)

exp(mean(hurdle_damPres_SW$sims.list$zeta_damPres))
```

    Dam presence increases probability of surface water presence by ~ 6%.


damPres -> swAmount

```{r,echo=FALSE}
summary(hurdle_damPres_SW$sims.list$beta_damPres)

exp(mean(hurdle_damPres_SW$sims.list$beta_damPres))
```

    Dam presence increases surface water area by ~ 20%.

----
----

Prior-posterior overlap:

```{r,echo=FALSE}
damPres_alpha_post <- hurdle_damPres_SW$sims.list$alpha
damPres_beta_post  <- hurdle_damPres_SW$sims.list$beta_damPres
damPres_zeta_post  <- hurdle_damPres_SW$sims.list$zeta_damPres
damPres_pTP_post <- hurdle_damPres_SW$sims.list$p_TP
damPres_pFP_post <- hurdle_damPres_SW$sims.list$p_FP

damPresMod_n <- length(damPres_alpha_post)

damPres_alpha_prior <- rnorm(damPresMod_n, 0, sqrt(1/0.04))
damPres_beta_prior  <- rnorm(damPresMod_n, 0, sqrt(1/0.04))
damPres_zeta_prior  <- rnorm(damPresMod_n, 0, sqrt(1/0.04))
damPres_pTP_prior <- rnorm(damPresMod_n, 0.76, sqrt(1/0.16))
damPres_pFP_prior <- rnorm(damPresMod_n, 0.01, sqrt(1/0.005))






# Plot overlap:

plot_prior_post <- function(prior, post, name) {
  
  df <- tibble(
    value = c(prior, post),
    dist  = rep(c("Prior", "Posterior"), each = length(prior))
  )
  
  ggplot(df, aes(x = value, fill = dist)) +
    geom_density(alpha = 0.4) +
    scale_fill_manual(values = c("gray60", "steelblue")) +
    labs(
      title = paste("Prior vs Posterior for", name),
      x = name,
      y = "Density"
    ) +
    theme_minimal(base_size = 14)
}

plot_prior_post(damPres_alpha_prior, damPres_alpha_post, "alpha")
plot_prior_post(damPres_beta_prior, damPres_beta_post, "beta")
plot_prior_post(damPres_zeta_prior, damPres_zeta_post, "zeta")
```

----
----

Marginal effect of dam presence on surface water presence, given geomorphic context

**Note** there's a  TINY positive effect of dam presence on surface water presence. Not really perceptible from these plots. 

```{r,echo=FALSE}
# Build a grid of raw geomorphic covariate values
damPres_geom_grid <- tibble(
  TO_raw = seq(min(aoi_HurdleData$TO),
               max(aoi_HurdleData$TO), length.out = 50),

  incis_raw = 1 - (TO_raw / 90),   # raw incision scale

  RIPA_raw = seq(min(aoi_HurdleData$riparian_area_m2),
                 max(aoi_HurdleData$riparian_area_m2), length.out = 50),

  UDA_raw  = seq(min(aoi_HurdleData$catchment_km2),
                 max(aoi_HurdleData$catchment_km2), length.out = 50)
) %>%
  # Convert raw → standardized (to match model scale)
  mutate(
    incis = as.numeric(scale(incis_raw)),
    RIPA  = as.numeric(scale(RIPA_raw)),
    UDA   = as.numeric(scale(UDA_raw))
  )

damPres_compute_psi <- function(dam_flag, geom_grid, damPres_post) {

  eps <- mean(damPres_post$epsilon)
  zd  <- mean(damPres_post$zeta_damPres)
  zi  <- mean(damPres_post$zeta_incis)
  zr  <- mean(damPres_post$zeta_RIPA)
  zu  <- mean(damPres_post$zeta_UDA)

  geom_grid %>%
    mutate(
      psi = plogis(
        eps +
        zd * dam_flag +
        zi * incis +      # standardized incision
        zr * RIPA +       # standardized RIPA
        zu * UDA          # standardized UDA
      ),
      dam = dam_flag
    )
}


```

```{r,echo=FALSE}
damPres_psi_nodam <- damPres_compute_psi(0, damPres_geom_grid, damPres_post)
damPres_psi_dam   <- damPres_compute_psi(1, damPres_geom_grid, damPres_post)

damPres_psi_df <- bind_rows(damPres_psi_nodam, damPres_psi_dam)
```

```{r,echo=FALSE}
ggplot(damPres_psi_df, aes(x = incis_raw, y = psi, color = factor(dam))) +
  geom_line(size = 1.2) +
  labs(
    title = "Marginal Effect of Dam Presence on Surface Water Presence",
    subtitle = "Predicted probability of water (psi) across raw incision values",
    x = "Channel Incision (raw scale: 1 - TO/90)",
    y = "Probability of Surface Water",
    color = "Dam Presence"
  ) +
  scale_color_manual(values = c("brown4", "steelblue"),
                     labels = c("No Dam", "Dam Present")) +
  theme_minimal(base_size = 14)

```

```{r,echo=FALSE}
ggplot(damPres_psi_df, aes(x = RIPA_raw, y = psi, color = factor(dam))) +
  geom_line(size = 1.2) +
  labs(
    title = "Marginal Effect of Dam Presence on Surface Water Presence",
    subtitle = "Predicted psi across raw riparian area (m^2)",
    x = "Riparian Area (m^2)",
    y = "Probability of Surface Water",
    color = "Dam Presence"
  ) +
  scale_color_manual(values = c("brown4", "steelblue"),
                     labels = c("No Dam", "Dam Present")) +
  theme_minimal(base_size = 14)


```

```{r,echo=FALSE}
ggplot(damPres_psi_df, aes(x = UDA_raw, y = psi, color = factor(dam))) +
  geom_line(size = 1.2) +
  labs(
    title = "Marginal Effect of Dam Presence on Surface Water Presence",
    subtitle = "Predicted psi across raw contributing drainage area (km^2)",
    x = "Contributing Drainage Area (km^2)",
    y = "Probability of Surface Water",
    color = "Dam Presence"
  ) +
  scale_color_manual(values = c("brown4", "steelblue"),
                     labels = c("No Dam", "Dam Present")) +
  theme_minimal(base_size = 14)


```


Partial dependence plots

Constrain PDP to empirical ranges for SWA and target covariates

```{r,echo=FALSE}
library(tidyverse)
library(scales)

damPres_post <- hurdle_damPres_SW$sims.list

# Raw covariates
incis_raw <- 1 - aoi_HurdleData$TO/90
RIPA_raw  <- aoi_HurdleData$riparian_area_m2
UDA_raw   <- aoi_HurdleData$catchment_km2

logSWA_mean <- mean(aoi_HurdleData$logSWA)
logSWA_sd   <- sd(aoi_HurdleData$logSWA)

# Standardization stats
incis_mean_raw <- mean(incis_raw)
incis_sd_raw   <- sd(incis_raw)

RIPA_mean_raw  <- mean(RIPA_raw)
RIPA_sd_raw    <- sd(RIPA_raw)

UDA_mean_raw   <- mean(UDA_raw)
UDA_sd_raw     <- sd(UDA_raw)


##############################3

# Helper Functions to Compute SWA from posterior draws
compute_swa_draws <- function(incis, RIPA, UDA, post, dam_flag) {
  n_draws <- length(damPres_post$alpha)
  swa_mat <- matrix(NA_real_, nrow = length(incis), ncol = n_draws)

  for (j in seq_len(n_draws)) {
    lp <- damPres_post$alpha[j] +
          damPres_post$beta_incis[j] * incis +
          damPres_post$beta_RIPA[j]  * RIPA +
          damPres_post$beta_UDA[j]   * UDA +
          damPres_post$beta_damPres[j] * dam_flag

    swa_mat[, j] <- exp(lp) - 1
  }

  swa_mat
}

summarize_swa <- function(swa_mat) {
  tibble(
    SWA_med = apply(swa_mat, 1, median),
    SWA_lo  = apply(swa_mat, 1, quantile, probs = 0.025),
    SWA_hi  = apply(swa_mat, 1, quantile, probs = 0.975)
  )
}




#########################################

# PDP Builder:

build_pdp <- function(focal_raw, focal_name,
                      incis_raw, RIPA_raw, UDA_raw,
                      post,
                      incis_mean_raw, incis_sd_raw,
                      RIPA_mean_raw, RIPA_sd_raw,
                      UDA_mean_raw, UDA_sd_raw) {

  # Sequence only across empirical range
  focal_seq <- seq(min(focal_raw), max(focal_raw), length.out = 200)

  # Conditional medians of other covariates
  incis_med <- median(incis_raw)
  RIPA_med  <- median(RIPA_raw)
  UDA_med   <- median(UDA_raw)

  # Build grid (200 rows)
  grid <- tibble(
    focal_raw = focal_seq,
    incis_raw = if (focal_name == "incis") focal_seq else incis_med,
    RIPA_raw  = if (focal_name == "RIPA")  focal_seq else RIPA_med,
    UDA_raw   = if (focal_name == "UDA")   focal_seq else UDA_med
  ) %>%
    mutate(
      incis = (incis_raw - incis_mean_raw) / incis_sd_raw,
      RIPA  = (RIPA_raw  - RIPA_mean_raw)  / RIPA_sd_raw,
      UDA   = (UDA_raw   - UDA_mean_raw)   / UDA_sd_raw
    )

  # Compute SWA draws
  swa_nodam <- compute_swa_draws(grid$incis, grid$RIPA, grid$UDA, post, dam_flag = 0)
  swa_dam   <- compute_swa_draws(grid$incis, grid$RIPA, grid$UDA, post, dam_flag = 1)

  # Summaries (400 rows)
  pdp <- bind_rows(
    summarize_swa(swa_nodam) %>% mutate(dam = "No Dam"),
    summarize_swa(swa_dam)   %>% mutate(dam = "Dam Present")
  )

  # Repeat grid twice (400 rows)
  grid2 <- bind_rows(grid, grid)

  # Bind columns safely
  bind_cols(grid2, pdp)
}



#####

pdp_incis <- build_pdp(
  focal_raw = incis_raw,
  focal_name = "incis",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)

pdp_RIPA <- build_pdp(
  focal_raw = RIPA_raw,
  focal_name = "RIPA",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)

pdp_UDA <- build_pdp(
  focal_raw = UDA_raw,
  focal_name = "UDA",
  incis_raw, RIPA_raw, UDA_raw,
  damPres_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw
)



##########################################3

# Plot Function:

plot_pdp <- function(pdp, xvar, xtitle) {
  ggplot(pdp,
         aes(x = .data[[xvar]], y = SWA_med, color = dam, fill = dam)) +
    geom_ribbon(aes(ymin = SWA_lo, ymax = SWA_hi), alpha = 0.2, color = NA) +
    geom_line(size = 1.2) +
    scale_color_manual(values = c("steelblue", "brown4")) +
    scale_fill_manual(values = c("steelblue", "brown4")) +
    scale_y_continuous(labels = comma) +
    labs(
      title = paste("Partial Dependence of SWA on", xtitle),
      subtitle = "Median predicted SWA with 95% credible intervals\nConstrained to empirical covariate range",
      x = xtitle,
      y = "Predicted SWA (m²)",
      color = "Dam Presence",
      fill  = "Dam Presence"
    ) +
    theme_minimal(base_size = 14)
}


###########################################3


plot_pdp(pdp_incis, "incis_raw", "Channel Incision (raw)")
plot_pdp(pdp_RIPA,  "RIPA_raw",  "Riparian Area (m²)")
plot_pdp(pdp_UDA,   "UDA_raw",   "Contributing Drainage Area (km²)")


```


----
----

Dam Density
----


**Hurdle HGLMM**
----

```{r,echo=FALSE}
#library(jagsUI)
```
    sink("HGLMM_hurdle_damDensity.jags")
    
    cat("
    model{
      # NEED priors and hyperpriors first so that downstream nodes can be estimated
      
      ##########################################################################
      
      # HYPERPRIORS FOR SW PRESENCE (1/2 Normal on SDs):
    
      sigma_zeta_stream  ~ dnorm(0,1) T(0,)
      tau_zeta_stream    <- pow(sigma_zeta_stream, -2)
    
      sigma_zeta_year    ~ dnorm(0,1) T(0,)
      tau_zeta_year      <- pow(sigma_zeta_year, -2)
    
      sigma_zeta_event   ~ dnorm(0,1) T(0,)
      tau_zeta_event     <- pow(sigma_zeta_event, -2)
    
      sigma_iota_stream  ~ dnorm(0,1) T(0,)
      tau_iota_stream    <- pow(sigma_iota_stream, -2)
    
      sigma_iota_event   ~ dnorm(0,1) T(0,)
      tau_iota_event     <- pow(sigma_iota_event, -2)
    
      sigma_iota_cluster ~ dnorm(0,1) T(0,)
      tau_iota_cluster   <- pow(sigma_iota_cluster, -2)
    
      ##########################################################################
      # HYPERPRIORS FOR SW AMOUNT (1/2 Normal on SDs):
    
      sigma              ~ dnorm(0,1) T(0,)
      tau                <- pow(sigma, -2)
    
      sigma_beta_stream  ~ dnorm(0, 0.7) T(0,)
      tau_beta_stream    <- pow(sigma_beta_stream, -2)
    
      sigma_beta_year    ~ dnorm(0, 0.7) T(0,)
      tau_beta_year      <- pow(sigma_beta_year, -2)
    
      sigma_beta_event   ~ dnorm(0, 0.7) T(0,)
      tau_beta_event     <- pow(sigma_beta_event, -2)
    
      sigma_eta_stream   ~ dnorm(0, 1) T(0,)
      tau_eta_stream     <- pow(sigma_eta_stream, -2)
    
      sigma_eta_event    ~ dnorm(0, 1) T(0,)
      tau_eta_event      <- pow(sigma_eta_event, -2)
    
      sigma_eta_cluster  ~ dnorm(0, 1) T(0,)
      tau_eta_cluster    <- pow(sigma_eta_cluster, -2)
    
      ##########################################################################
      # FIXED EFFECT PRIORS:
    
      # Presence/absence:
      epsilon ~ dnorm(0,0.04)
      zeta_damDen ~ dnorm(0,0.04)
      zeta_incis ~ dnorm(0,0.04)
      zeta_RIPA ~ dnorm(0,0.04)
      zeta_UDA ~ dnorm(0,0.04)
    
      # Amount:
      alpha ~ dnorm(0,0.04)
      beta_damDen ~ dnorm(0,0.04)
      beta_incis ~ dnorm(0,0.04)
      beta_RIPA ~ dnorm(0,0.04)
      beta_UDA ~ dnorm(0,0.04)
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW PRESENCE:
    
      for(j in 1:n_stream){
        zeta_stream[j] ~ dnorm(0, tau_zeta_stream)
        iota_stream[j] ~ dnorm(0, tau_iota_stream)
      }
    
      for(yr in 1:n_year){
        zeta_year[yr] ~ dnorm(0, tau_zeta_year)
      }
    
      for(k in 1:n_event){
        zeta_event[k] ~ dnorm(0, tau_zeta_event)
        iota_event[k] ~ dnorm(0, tau_iota_event)
      }
    
      for(c in 1:n_cluster){
        iota_cluster[c] ~ dnorm(0, tau_iota_cluster)
      }
    
      ##########################################################################
      # RANDOM EFFECT PRIORS FOR SW AMOUNT:
    
      for(j in 1:n_stream){
        beta_stream[j] ~ dnorm(0, tau_beta_stream)
        eta_stream[j]  ~ dnorm(0, tau_eta_stream)
      }
    
      for(yr in 1:n_year){
        beta_year[yr] ~ dnorm(0, tau_beta_year)
      }
    
      for(k in 1:n_event){
        beta_event[k] ~ dnorm(beta_year[event_year[k]], tau_beta_event)
        eta_event[k]  ~ dnorm(0, tau_eta_event)
      }
    
      for(c in 1:n_cluster){
        eta_cluster[c] ~ dnorm(0, tau_eta_cluster)
      }
    
      ##########################################################################
      # DETECTION PRIORS:
      
      for(k in 1:n_event){
      p_TP[k] ~ dnorm(0.76, pow(0.16,-2)) T(0,1)   # empirically derived recall
      p_FP[k] ~ dnorm(0.01, pow(0.005,-2)) T(0,1)  # empirically derived FPR
      }
      
      ##########################################################################
      # MAIN LOOP: PRESENCE, DETECTION, AMOUNT
    
      for(i in 1:N){
    
        # Presence/absence:
        z[i] ~ dbern(psi[i])
    
        logit(psi[i]) <- epsilon
                         + (zeta_damDen
                            + zeta_stream[stream[i]]
                            + zeta_year[year[i]]
                            + zeta_event[event[i]]) * damDen[i]
                         + zeta_incis * incis[i]
                         + zeta_RIPA * RIPA[i]
                         + zeta_UDA * UDA[i]
                         + iota_stream[stream[i]]
                         + iota_event[event[i]]
                         + iota_cluster[cluster[i]]
    
        # Detection:
        p_eff[i] <- z[i] * p_TP[event[i]] + (1-z[i]) * p_FP[event[i]]
        y[i]     ~ dbern(p_eff[i])
    
        # Amount:
        y2[i] ~ dnorm(mu[i], tau_eff[i])
    
        mu[i] <- z[i] *
                  (alpha
                    + (beta_damDen
                      + beta_stream[stream[i]]
                      + beta_year[year[i]]
                      + beta_event[event[i]]) * damDen[i]
                    + beta_incis * incis[i] 
                    + beta_RIPA * RIPA[i]
                    + beta_UDA * UDA[i]
                    + eta_stream[stream[i]]
                    + eta_event[event[i]]
                    + eta_cluster[cluster[i]])
    
        tau_eff[i] <- z[i] * tau + (1 - z[i]) * 100
      }
    
    }
    ", fill = TRUE)
    
    sink()
   
    

```{r,echo=FALSE}
hurdle_damDen_data_list <- list(
  N = N,
  y = y,
  y2 = y2,
  stream = stream,
  year = year,
  event = event,
  cluster = cluster,
  damDen = damDen,
  incis = as.numeric(scale(1-(aoi_HurdleData$TO/90))), # Channel Incision
  RIPA = as.numeric(scale(aoi_HurdleData$riparian_area_m2)),
  UDA = as.numeric(scale(aoi_HurdleData$catchment_km2)),
  n_stream = n_stream,
  n_year = n_year,
  n_event = n_event,
  n_cluster = n_cluster,
  event_year = event_year_3_4
)

hurdle_damDen_inits <- function(){
  list(
    alpha        = rnorm(1, 0, 1),
    epsilon      = rnorm(1, 0, 1),

    sigma        = runif(1, 0.1, 2),
    sigma_zeta_stream  = runif(1, 0.1, 2),
    sigma_zeta_year    = runif(1, 0.1, 2),
    sigma_zeta_event   = runif(1, 0.1, 2),
    sigma_iota_stream  = runif(1, 0.1, 2),
    sigma_iota_event   = runif(1, 0.1, 2),
    sigma_iota_cluster = runif(1, 0.1, 2),

    sigma_beta_stream  = runif(1, 0.1, 2),
    sigma_beta_year    = runif(1, 0.1, 2),
    sigma_beta_event   = runif(1, 0.1, 2),
    sigma_eta_stream   = runif(1, 0.1, 2),
    sigma_eta_event    = runif(1, 0.1, 2),
    sigma_eta_cluster  = runif(1, 0.1, 2)
  )
}

hurdle_damDen_params <- c(
  "psi",
  "mu",
  "epsilon",
  "alpha",
  "zeta_damDen",
  "zeta_incis",
  "zeta_RIPA",
  "zeta_UDA",
  "zeta_stream",
  "zeta_year",
  "zeta_event",
  "iota_stream",
  "iota_event",
  "iota_cluster",
  "beta_damDen",
  "beta_incis",
  "beta_RIPA",
  "beta_UDA",
  "beta_stream",
  "beta_year",
  "beta_event",
  "eta_stream",
  "eta_event",
  "eta_cluster",
  "sigma",
  "sigma_zeta_stream",
  "sigma_zeta_year",
  "sigma_zeta_event",
  "sigma_iota_stream",
  "sigma_iota_event",
  "sigma_iota_cluster",
  "sigma_beta_stream",
  "sigma_beta_year",
  "sigma_beta_event",
  "sigma_eta_stream",
  "sigma_eta_event",
  "sigma_eta_cluster",
  "p_TP",
  "p_FP",
  "z",
  "tau_eff"
)

nchains <- 3
nthin <- 5
niter <- 100000
nburn <- 20000

hurdle_damDen_SW <- jags(data = hurdle_damDen_data_list,
                   inits = hurdle_damDen_inits,
                   parameters.to.save = hurdle_damDen_params,
                   model.file = "HGLMM_hurdle_damDensity.jags",
                   parallel = T,
                   n.chains = nchains,
                   n.thin = nthin,
                   n.iter = niter,
                   n.burnin = nburn)


Sys.time()

#print(hurdle_damPres_SW) # model with local and watershed dam density as fixed effects
#print(hurdle_damPres_SW$summary[grep("beta_", rownames(hurdle_damPres_SW$summary)), ])
#print(hurdle_damPres_SW$summary[grep("zeta_", rownames(hurdle_damPres_SW$summary)), ])

```


----
----

DIAGNOSTICS:

Extract posteriors:
```{r,echo=FALSE}
damDen_post <- hurdle_damDen_SW$sims.list 
```

"Chain health" Rhat + ESS:
```{r,echo=FALSE} 
# Choose only the parameters you care about
params_2 <- c(
  "epsilon",
  "alpha",
  "zeta_damDen",
  "beta_damDen",
  "zeta_incis",
  "beta_incis",
  "zeta_RIPA",
  "beta_RIPA",
  "zeta_UDA",
  "beta_UDA",
  "sigma",
  "sigma_zeta_stream",
  "sigma_iota_cluster",
  "p_TP[1]",
  "p_FP[1]")

# Gelman–Rubin convergence diagnostic
gelman.diag(hurdle_damDen_SW$samples[, params_2], multivariate = FALSE)

# Effective sample size
effectiveSize(hurdle_damDen_SW$samples[, params_2])

```



    Good Rhat values, all around 1.00.
    
    Mostly high effective sample sizes. Note: beta_damPres has the most trouble converging. 




Check traceplots for critical parameters
```{r,echo=FALSE} 
tplot_params_2 <- c("alpha","beta_damDen","zeta_damDen", "sigma","sigma_zeta_stream","sigma_beta_stream")


mcmc_sub_2 <- hurdle_damDen_SW$samples[, tplot_params_2]

coda::traceplot(mcmc_sub_2)
```


    Traceplots all look good, except for beta_damDen, which is mostly grassy, but 
    slow to converge - suggesting that its difficult for the model to estimate. 




Posterior predictive plots:

Predicted SW vs observed SW

Goal is 1:1 (ish)

```{r,echo=FALSE} 
# Posterior mean psi
psi_post_2 <- apply(hurdle_damDen_SW$sims.list$psi, 2, mean)

# Compare to observed y
plot(psi_post_2, y, pch=16, col=rgb(0,0,0,0.2))


```

```{r,echo=FALSE} 
mu_post_2 <- apply(hurdle_damDen_SW$sims.list$mu, 2, mean)

plot(mu_post_2[y==1], y2[y==1], pch=16, col=rgb(0,0,1,0.3))
abline(0,1,col="red")


```


Random effects shrinkage: 

goal is centered around 0(ish)

1. zeta_stream; random effect of stream on sw presence
2. beta_stream; random effect of stream on sw amount
3. zeta_event; random effect of sample event on sw presence
4. beta_event; random effect of sample event on sw amount

```{r, echo=FALSE} 
boxplot(hurdle_damDen_SW$sims.list$zeta_stream)
boxplot(hurdle_damDen_SW$sims.list$beta_stream)
boxplot(hurdle_damDen_SW$sims.list$zeta_event)
boxplot(hurdle_damDen_SW$sims.list$beta_event)
```


----
----

----
----


Posterior mean dam effects:

damDen -> swPres

```{r,echo=FALSE}
summary(hurdle_damDen_SW$sims.list$zeta_damDen)

exp(mean(hurdle_damDen_SW$sims.list$zeta_damDen))
```

    Log-odds ratio = 1.03

    1 std dev. increase in dam density increases log-odds of surface water 
    presence by ~ 3%... so not much. 
    
    1 std dev. of dam density from raw data is about 3.14 dams
    
    Massive overlap of 0. Really no major effect of 1 std dev increase in dam 
    density on surface water presence


damDen -> swAmount

```{r,echo=FALSE}
summary(hurdle_damDen_SW$sims.list$beta_damDen)

exp(mean(hurdle_damDen_SW$sims.list$beta_damDen))
```

    Dam presence increases surface water area by ~ 13%.

----
----

Prior-posterior overlap:

```{r,echo=FALSE}
damDen_alpha_post <- hurdle_damDen_SW$sims.list$alpha
damDen_beta_post  <- hurdle_damDen_SW$sims.list$beta_damDen
damDen_zeta_post  <- hurdle_damDen_SW$sims.list$zeta_damDen
damDen_pTP_post <- hurdle_damDen_SW$sims.list$p_TP
damDen_pFP_post <- hurdle_damDen_SW$sims.list$p_FP

damDenMod_n <- length(damDen_alpha_post)

damDen_alpha_prior <- rnorm(damDenMod_n, 0, sqrt(1/0.04))
damDen_beta_prior  <- rnorm(damDenMod_n, 0, sqrt(1/0.04))
damDen_zeta_prior  <- rnorm(damDenMod_n, 0, sqrt(1/0.04))
damDen_pTP_prior <- rnorm(damDenMod_n, 0.76, sqrt(1/0.16))
damDen_pFP_prior <- rnorm(damDenMod_n, 0.01, sqrt(1/0.005))






# Plot overlap:

plot_prior_post <- function(prior, post, name) {
  
  df <- tibble(
    value = c(prior, post),
    dist  = rep(c("Prior", "Posterior"), each = length(prior))
  )
  
  ggplot(df, aes(x = value, fill = dist)) +
    geom_density(alpha = 0.4) +
    scale_fill_manual(values = c("gray60", "steelblue")) +
    labs(
      title = paste("Prior vs Posterior for", name),
      x = name,
      y = "Density"
    ) +
    theme_minimal(base_size = 14)
}

plot_prior_post(damDen_alpha_prior, damPres_alpha_post, "alpha")
plot_prior_post(damDen_beta_prior, damPres_beta_post, "beta")
plot_prior_post(damDen_zeta_prior, damPres_zeta_post, "zeta")
```

----
----

Marginal effect of dam presence on surface water presence, given geomorphic context

**Note** there's a  TINY positive effect of dam presence on surface water presence. Not really perceptible from these plots. 

```{r,echo=FALSE}
# Build a grid of raw geomorphic covariate values
damDen_geom_grid <- tibble(
  TO_raw = seq(min(aoi_HurdleData$TO),
               max(aoi_HurdleData$TO), length.out = 50),

  incis_raw = 1 - (TO_raw / 90),   # raw incision scale

  RIPA_raw = seq(min(aoi_HurdleData$riparian_area_m2),
                 max(aoi_HurdleData$riparian_area_m2), length.out = 50),

  UDA_raw  = seq(min(aoi_HurdleData$catchment_km2),
                 max(aoi_HurdleData$catchment_km2), length.out = 50)
) %>%
  # Convert raw → standardized (to match model scale)
  mutate(
    incis = as.numeric(scale(incis_raw)),
    RIPA  = as.numeric(scale(RIPA_raw)),
    UDA   = as.numeric(scale(UDA_raw))
  )

# Make dam density sequence
damDen_seq <- seq(min(damDen_raw),max(damDen_raw),length.out=10)

# Compute surface water across dam density sequence
damDen_compute_psi <- function(damDen_val, geom_grid, damDen_post) {

  eps <- mean(damDen_post$epsilon)
  zd  <- mean(damDen_post$zeta_damDen)
  zi  <- mean(damDen_post$zeta_incis)
  zr  <- mean(damDen_post$zeta_RIPA)
  zu  <- mean(damDen_post$zeta_UDA)

  geom_grid %>%
    mutate(
      psi = plogis(
        eps +
        zd * damDen_val +
        zi * incis +      # standardized incision
        zr * RIPA +       # standardized RIPA
        zu * UDA          # standardized UDA
      ),
      damDen = damDen_val
    )
}


```

```{r,echo=FALSE}
damDen_psi_df <- map_df(damDen_seq,
                        ~ damDen_compute_psi(.x,
                                             damDen_geom_grid,
                                             damDen_post))
```

```{r,echo=FALSE}
ggplot(damDen_psi_df, aes(x = incis_raw, y = psi, color = damDen)) +
  geom_point() +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "Marginal Effect of Dam Density on Surface Water Presence",
    subtitle = "Predicted probability of water (psi) across raw incision values",
    x = "Channel Incision (raw scale: 1 - TO/90)",
    y = "Probability of Surface Water",
    color = "Dam Density (z-std)"
  ) +
  theme_minimal(base_size = 14)

```

```{r,echo=FALSE}
ggplot(damDen_psi_df, aes(x = RIPA_raw, y = psi, color = damDen)) +
  geom_point() +
  scale_color_viridis_c(option="plasma") +
  labs(
    title = "Marginal Effect of Dam Density on Surface Water Presence",
    subtitle = "Predicted psi across raw riparian area (m^2)",
    x = "Riparian Area (m^2)",
    y = "Probability of Surface Water",
    color = "Dam Density (z-std)"
  ) +
  theme_minimal(base_size = 14)
```

```{r,echo=FALSE}
ggplot(damDen_psi_df, aes(x = UDA_raw, y = psi, color = damDen)) +
  geom_point() +
  scale_color_viridis_c(option="plasma") +
  labs(
    title = "Marginal Effect of Dam Density on Surface Water Presence",
    subtitle = "Predicted psi across raw upstream drainage area (km^2)",
    x = "Upstream Drainage Area (km^2)",
    y = "Probability of Surface Water",
    color = "Dam Density (z-std)"
  ) +
  theme_minimal(base_size = 14)


```


Partial dependence plots

Constrain PDP to empirical ranges for SWA and target covariates

```{r,echo=FALSE}
library(tidyverse)
library(scales)

damDen_post <- hurdle_damDen_SW$sims.list

# Raw covariates
incis_raw <- 1 - aoi_HurdleData$TO/90
RIPA_raw  <- aoi_HurdleData$riparian_area_m2
UDA_raw   <- aoi_HurdleData$catchment_km2
#damDen <- aoi_HurdleData$dam_km_std
damDen_raw <- aoi_HurdleData$dam_km
damDen_mean_raw <- mean(damDen_raw)
damDen_sd_raw   <- sd(damDen_raw)

logSWA_mean <- mean(aoi_HurdleData$logSWA)
logSWA_sd   <- sd(aoi_HurdleData$logSWA)

# Standardization stats
incis_mean_raw <- mean(incis_raw)
incis_sd_raw   <- sd(incis_raw)

RIPA_mean_raw  <- mean(RIPA_raw)
RIPA_sd_raw    <- sd(RIPA_raw)

UDA_mean_raw   <- mean(UDA_raw)
UDA_sd_raw     <- sd(UDA_raw)


##############################3

# Helper Functions to Compute SWA from posterior draws
compute_swa_draws_damDen <- function(incis, RIPA, UDA, post, damDen_val) {
  n_draws <- length(post$alpha)
  swa_mat <- matrix(NA_real_, nrow = length(incis), ncol = n_draws)

  for (j in seq_len(n_draws)) {
    lp <- post$alpha[j] +
          post$beta_incis[j] * incis +
          post$beta_RIPA[j]  * RIPA +
          post$beta_UDA[j]   * UDA +
          post$beta_damDen[j] * damDen_val

    swa_mat[, j] <- exp(lp) - 1
  }

  swa_mat
}

summarize_swa <- function(swa_mat) {
  tibble(
    SWA_med = apply(swa_mat, 1, median),
    SWA_lo  = apply(swa_mat, 1, quantile, probs = 0.025),
    SWA_hi  = apply(swa_mat, 1, quantile, probs = 0.975)
  )
}




#########################################

# PDP Builder:

build_pdp_damDen <- function(focal_raw, focal_name,
                      incis_raw, RIPA_raw, UDA_raw,
                      post,
                      incis_mean_raw, incis_sd_raw,
                      RIPA_mean_raw, RIPA_sd_raw,
                      UDA_mean_raw, UDA_sd_raw,
                      damDen_raw) {

  # Sequence only across empirical range
  focal_seq <- seq(min(focal_raw), max(focal_raw), length.out = 200)

  # Conditional medians of other covariates
  incis_med <- median(incis_raw)
  RIPA_med  <- median(RIPA_raw)
  UDA_med   <- median(UDA_raw)
  
  # Empirical dam density range (standardized)
  damDen_seq <- seq(min(damDen_raw), max(damDen_raw), length.out = 10)

  # Build grid (200 rows)
  grid <- tibble(
    focal_raw = focal_seq,
    incis_raw = if (focal_name == "incis") focal_seq else incis_med,
    RIPA_raw  = if (focal_name == "RIPA")  focal_seq else RIPA_med,
    UDA_raw   = if (focal_name == "UDA")   focal_seq else UDA_med
  ) %>%
    mutate(
      incis = (incis_raw - incis_mean_raw) / incis_sd_raw,
      RIPA  = (RIPA_raw  - RIPA_mean_raw)  / RIPA_sd_raw,
      UDA   = (UDA_raw   - UDA_mean_raw)   / UDA_sd_raw
    )
  # Compute PDP curves for each dam density value
  pdp_list <- map(damDen_seq, function(dd) {
    swa_mat <- compute_swa_draws_damDen(grid$incis, grid$RIPA, grid$UDA, post,
                                        (dd - damDen_mean_raw) / damDen_sd_raw)
    summarize_swa(swa_mat) %>% mutate(damDen = dd)
  })

  pdp <- bind_rows(pdp_list)

  # Repeat grid for each damDen value
  grid2 <- bind_rows(replicate(length(damDen_seq), grid, simplify = FALSE))

  bind_cols(grid2, pdp)
}



#####

pdp_incis_damDen <- build_pdp_damDen(
  focal_raw = incis_raw,
  focal_name = "incis",
  incis_raw, RIPA_raw, UDA_raw,
  damDen_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw, damDen_raw
)

pdp_RIPA_damDen <- build_pdp_damDen(
  focal_raw = RIPA_raw,
  focal_name = "RIPA",
  incis_raw, RIPA_raw, UDA_raw,
  damDen_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw, damDen_raw
)

pdp_UDA_damDen <- build_pdp_damDen(
  focal_raw = UDA_raw,
  focal_name = "UDA",
  incis_raw, RIPA_raw, UDA_raw,
  damDen_post,
  incis_mean_raw, incis_sd_raw,
  RIPA_mean_raw, RIPA_sd_raw,
  UDA_mean_raw, UDA_sd_raw, damDen_raw
)



##########################################3

# Plot Function:

plot_pdp_damDen <- function(pdp, xvar, xtitle) {
  ggplot(pdp,
         aes(x = .data[[xvar]], y = SWA_med,
             color = damDen, fill = damDen)) +
    #geom_ribbon(aes(ymin = SWA_lo, ymax = SWA_hi), alpha = 0.15, color = NA) +
    geom_point() +
    scale_color_viridis_c(option = "plasma") +
    scale_fill_viridis_c(option = "plasma") +
    scale_y_continuous(labels = comma) +
    labs(
      title = paste("Partial Dependence of SWA on", xtitle),
      subtitle = "Median predicted SWA with 95% credible intervals\nAcross empirical dam density range",
      x = xtitle,
      y = "Predicted SWA (m^2)",
      color = "Dam Density (dams/km)",
      fill  = "Dam Density (dams/km)"
    ) +
    theme_minimal(base_size = 14)
}



###########################################3


plot_pdp_damDen(pdp_incis_damDen, "incis_raw", "Channel Incision (raw)")
plot_pdp_damDen(pdp_RIPA_damDen,  "RIPA_raw",  "Riparian Area (m²)")
plot_pdp_damDen(pdp_UDA_damDen,   "UDA_raw",   "Contributing Drainage Area (km²)")


```
