---
title: "Landscape_Application"
author: "Matt Webster"
date: "2026-08-20"
output: html_document
editor_options: 
  chunk_output_type: console
---

This script is meant to apply model estimates to raw geomorphic/topographic covariate data, in order to create a predictive "heat ribbon" along our study streams that depict absolute change in SWA (or VA) with the addition of *x* dams km^-1.



----


**Data organization**
    

```{r,echo=FALSE,}
suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(ggplot2)
  library(dataRetrieval)
  library(knitr)
  library(ggplot2)
  library(zoo) #for moving window
  library(stringr)
  library(vioplot)
  library(purrr)
  library(lavaan)
  library(blavaan)
  library(ggcorrplot)
  library(lme4)
  library(broom.mixed)
  library(coda)
  library(scales)
  library(loo)
  library(sf)
  library(jsonlite)
  library(magrittr)
  library(slider)
  library(leaflet)

})
```


**Read in shape files and extract TP_ID ("Input_FID"), and geometry attributes:**


    This gives the geometry for the riparian zones of each stream. Recall that riparian
    zones were created using a 2m vertical buffer from stream flow lines. 
    
    The next step will sample geomorphic/topographic covariates within the riparian zone,
    using a moving window approach.
    
    
```{r,echo=FALSE}
# Willow:
wcRiparianGeometries <- suppressMessages(st_read("WC_RIP_GEOMS.shp",quiet=TRUE)) %>% 
  st_make_valid() %>% 
  select(Input_FID,
         geometry) %>% 
  rename(TP_ID = Input_FID) %>% 
  st_transform(4326) %>% 
  filter(TP_ID != 11609) 



wcRipGeom_Area <- st_transform(wcRiparianGeometries, 32613) %>% 
  mutate(riparian_area_m2 = as.numeric(st_area(.)),
         watershed = "WC") %>% 
  arrange(TP_ID)


# Lone Tree:
ltRiparianGeometries <- suppressMessages(st_read("LT_RIP_GEOMS.shp",quiet = TRUE)) %>%
  st_make_valid() %>% 
  select(Input_FID,
         geometry) %>% 
  rename(TP_ID = Input_FID) %>% 
  st_transform(4326)

ltRipGeom_Area <- st_transform(ltRiparianGeometries, 32613) %>% 
  mutate(riparian_area_m2 = as.numeric(st_area(.)),
         watershed = "LT") %>% 
  arrange(TP_ID)

# Brazil:
brzRiparianGeometries <- suppressMessages(st_read("BRZ_RIP_GEOMS.shp",quiet=T)) %>% 
  st_make_valid() %>% 
  select(Input_FID,
         geometry) %>% 
  rename(TP_ID = Input_FID) %>% 
  st_transform(4326) %>%
  filter(TP_ID!=5682)

brzRipGeom_Area <- st_transform(brzRiparianGeometries, 32613) %>% 
  mutate(riparian_area_m2 = as.numeric(st_area(.)),
         watershed = "BRZ") %>% 
  arrange(TP_ID)

# Larb:
lrbRiparianGeometries <- suppressMessages(st_read("LRB_RIP_GEOMS.shp",quiet=T)) %>% 
  st_make_valid() %>% 
  select(Input_FID,
         geometry) %>% 
  rename(TP_ID = Input_FID) %>% 
  st_transform(4326) %>% 
  filter(TP_ID != 4271)

lrbRipGeom_Area <- st_transform(lrbRiparianGeometries, 32613) %>% 
  mutate(riparian_area_m2 = as.numeric(st_area(.)),
         watershed = "LRB") %>% 
  arrange(TP_ID)

# Lower Beaver:
lbvRiparianGeometries <- suppressMessages(st_read("LBV_RIP_GEOMS.shp",quiet=T)) %>% 
  st_make_valid() %>% 
  select(Input_FID,
         geometry) %>%
  rename(TP_ID = Input_FID) %>% 
  st_transform(4326)

lbvRipGeom_Area <- st_transform(lbvRiparianGeometries, 32613) %>% 
  mutate(riparian_area_m2 = as.numeric(st_area(.)),
         watershed = "LBV") %>% 
  arrange(TP_ID)

# Upper Beaver:
ubvRiparianGeometries <- suppressMessages(st_read("ubv_RIP_GEOMS.shp",quiet=T)) %>% 
  st_make_valid() %>% 
  select(Input_FID,
         geometry) %>%
  rename(TP_ID = Input_FID) %>% 
  st_transform(4326)

ubvRipGeom_Area <- st_transform(ubvRiparianGeometries, 32613) %>%
  mutate(riparian_area_m2 = as.numeric(st_area(.)),
         watershed = "UBV") %>% 
  arrange(TP_ID) 


####################################################################################


#### Combine (above) ^^ dataframes into an AOI df:

aoi_ripGEOMs <- bind_rows(brzRipGeom_Area,
                          lbvRipGeom_Area,
                          lrbRipGeom_Area,
                          ltRipGeom_Area,
                          ubvRipGeom_Area,
                          wcRipGeom_Area) %>% 
  group_by(watershed)
```



```{r, echo=FALSE}

aoi_geomCovar <- read.csv("aoi_geomorphicCovariates.csv") %>%
  arrange(TP_ID) %>%
  
  # nearest upstream non-NA
  mutate(
    upstream = zoo::na.locf(catchment_km2, na.rm = FALSE)
  ) %>%
  
  # nearest downstream non-NA
  mutate(
    downstream = zoo::na.locf(catchment_km2, fromLast = TRUE, na.rm = FALSE)
  ) %>%
  
  # fill NA using nearest non-NA neighbors
  mutate(
    catchment_km2_filled = case_when(
      !is.na(catchment_km2) ~ catchment_km2,
      is.na(catchment_km2) & !is.na(upstream) & !is.na(downstream) ~ (upstream + downstream) / 2,
      is.na(catchment_km2) & !is.na(upstream) ~ upstream,
      is.na(catchment_km2) & !is.na(downstream) ~ downstream,
      TRUE ~ NA_real_
    )
  ) %>%
  
  select(-upstream, -downstream)

```


```{r,echo=FALSE}

# Combine df containing geometries with the df containing covariate information

aoi_forPrediction <- aoi_geomCovar %>% 
  left_join(aoi_ripGEOMs,
            by = c("watershed",
                   "TP_ID")) %>%
  mutate(incis = 1-(TO/90))


aoi_forPrediction <- aoi_forPrediction %>% 
  mutate(incis_scaled = as.numeric(scale(incis)),
         ripa_scaled = as.numeric(scale(riparian_area_m2.x)),
         uda_scaled = as.numeric(scale(catchment_km2_filled)))


```


**MOVING WINDOW SAMPLING**

    This section uses a moving 250m long window to sample geomorphic/topographic covariates
    within the riparian zones of each study stream. 250-m was selected because that's what
    was used as the minimum window size in the dam clustering/paired sampling protocol. 
    
    250-m represents the minimum predictable unit. 
    
    USE OVERLAPPING WINDOW; this provides a 'heatribbon' along the stream corridor
    

```{r,echo=FALSE}
####### NON-OVERLAPPING #########################
#df_windows <- aoi_forPrediction %>%
#  arrange(watershed, TP_ID) %>%
#  group_by(watershed) %>%
#  mutate(
#    window_id = ((TP_ID - 1) %/% 25) + 1
#  ) %>%
#  group_by(watershed, window_id) %>%
#  summarise(
#    TP_ID_start = min(TP_ID),
#    TP_ID_end   = max(TP_ID),
#    TO_mean     = mean(TO, na.rm = TRUE),
#    TO_sd       = sd(TO, na.rm = TRUE),
#    RIPA_mean   = mean(riparian_area_m2.x, na.rm = TRUE),
#    RIPA_sd     = sd(riparian_area_m2.x, na.rm = TRUE),
#    UDA_mean    = mean(catchment_km2_filled, na.rm = TRUE),
#    UDA_sd      = sd(catchment_km2_filled, na.rm = TRUE),
#    .groups = "drop"
#  )
#################################################


######### OVERLAPPING WINDOWS ##################

df_windows <- aoi_forPrediction %>%
  arrange(watershed, TP_ID) %>%											#get everything in sequential order
  group_by(watershed) %>%														#create groups (watershed-year)
  group_modify(~ {																	#operate per watershed-year
  
    n <- nrow(.x)																	  #".x" is "watershed-year[i]"; get # of rows
	
    window_id <- seq_len(n - 25 + 1)								#define windows (24 less than max number of rows)

    tibble(
      window_id   = window_id,
      TP_ID_start = window_id,											#start here
      TP_ID_end   = window_id + 24,									 #progress upstream +24 rows = 25 trows total = ~250m (each row is 1 unique sequential TP_ID)

      #Perform sliding window sampling
      incis_mean     = slide_dbl(.x$incis_scaled,   mean, .before = 0, .after = 24)[window_id],
      incis_sd       = slide_dbl(.x$incis_scaled,   sd,   .before = 0, .after = 24)[window_id],

      RIPA_mean   = slide_dbl(.x$ripa_scaled, mean, .before = 0, .after = 24)[window_id],
      RIPA_sd     = slide_dbl(.x$ripa_scaled, sd,   .before = 0, .after = 24)[window_id],

      UDA_mean    = slide_dbl(.x$uda_scaled,  mean, .before = 0, .after = 24)[window_id],
      UDA_sd      = slide_dbl(.x$uda_scaled,  sd,   .before = 0, .after = 24)[window_id]
    )
  }) %>%
  ungroup()



################################################

# NOW you should have a tibble with:

	# watershed
	# year
	# window_id
	# TP_ID_start
	# TP_ID_end
	# incis_mean
	# incis_sd
	# RIPA_mean
	# RIPA_sd
	# UDA_mean
	# UDA_sd

str(df_windows)
	
# THIS is what you will confront with your posterior estimates to get predicted SWA or VA
```


----



**Predict SWA/VA based on geomorphic covariates for window[w].**

    Predict SWA or VA for each stream during late August of 2021;
    This was one of the most water limited periods in our study.

    Predict at damDensity:

    0, 3, 6, 9, 12, 15
    
    0 = "baseline"/ no dam condition
    1 SD increase in damDensity corresponds to ~3.14 dams/km <according to the paired 
    sampling data>
    
    
```{r,echo=FALSE}
# Build 2 PREDICTION FUNCTIONs to predict baseline 1)SWA & 2)VA from TO,RIPA,UDA,watershed, and year, for late August sample event

# Define the fixed effect coefficients as estimated by "model 4" posterior means:
	# alpha
	# beta_damDen
	# beta_incis
	# beta_RIPA
	# beta_UDA
	
coefs_SW_fixed <- c(alpha = 5.776,
					beta_damDen = 0.147,								
					beta_incis = -0.770,
					beta_RIPA = -0.082,
					beta_UDA = 0.538)

coefs_VEG_fixed <- c(alpha = 9.209,
					beta_damDen = 0.119,								
					beta_incis = -0.01,
					beta_RIPA = 0.271,
					beta_UDA = 0.008)

coefs_SW_random_BetaStream <- c(BRZ = 0.118,
							LRB = -0.137,
							LT = -0.022,
							LBV  = 0.028,
							UBV = 0.098,
							WC = -0.086)
					
coefs_SW_random_etaStream <- c(BRZ = 0.247,
							LRB = 0.229,
							LT = -0.637,
							LBV  = -0.520,
							UBV = -0.121,
							WC = 0.858)

coefs_VEG_random_etaStream <- c(BRZ = 0.157,
							LRB = 0.203,
							LT = -0.253,
							LBV  = 0.502,
							UBV = 0.245,
							WC = -0.865)

coefs_SW_random_BetaYear <- 0.025             #mean(beta_year) for 2021

coefs_SW_random_lateAug21_etaEvent<- -0.052   #mean(eta_event) for last 2 wks of Aug, 2021

coefs_SW_random_lateAug21_betaEvent<- -0.015   #mean(beta_event) for last 2 wks of Aug, 2021


coefs_VEG_random_etaYear <- -1.081               #mean(eta_year) for 2021

```



```{r,echo=FALSE}
############################
# PREDICTION FUNCTION
############################

# SURFACE WATER AREA

# year = 2021
# event = late August

predict_SW_window <- function(incis_mean, RIPA_mean, UDA_mean,
                              damDensity,
                              watershed,
                              coefs_SW_fixed,
                              coefs_SW_random_BetaStream,
                              coefs_SW_random_etaStream,
                              coefs_SW_random_BetaYear,
                              coefs_SW_random_lateAug21_etaEvent,
                              coefs_SW_random_lateAug21_betaEvent,
                              link = "log") {
  
  eta <- coefs_SW_fixed["alpha"] +
    coefs_SW_fixed["beta_incis"]     * incis_mean +
    coefs_SW_fixed["beta_RIPA"]      * RIPA_mean +
    coefs_SW_fixed["beta_UDA"]       * UDA_mean +
    (coefs_SW_fixed["beta_damDen"]    * damDensity +
       coefs_SW_random_BetaStream[as.character(watershed)] +
       coefs_SW_random_BetaYear +
       coefs_SW_random_lateAug21_betaEvent) +
    coefs_SW_random_etaStream[as.character(watershed)] +
    coefs_SW_random_lateAug21_etaEvent
	
  SW <- exp(eta) - 1
  
  SW
}
```


----


```{r,echo=FALSE}
# VEGETATION AREA   = log(VA+1)

# year = 2021

predict_VEG_window <- function(incis_mean, RIPA_mean, UDA_mean,
                              damDensity,
                              watershed,
                              coefs_VEG_fixed,
                              coefs_VEG_random_etaStream,
                              coefs_VEG_random_etaYear,
                              link = "log") {
  
  eta <- coefs_VEG_fixed["alpha"] +
    coefs_VEG_fixed["beta_damDen"]    * damDensity +
    coefs_VEG_fixed["beta_incis"]     * incis_mean +
    coefs_VEG_fixed["beta_RIPA"]      * RIPA_mean +
    coefs_VEG_fixed["beta_UDA"]       * UDA_mean +
    coefs_VEG_random_etaStream[as.character(watershed)] +
    coefs_VEG_random_etaYear
	
  VEG <- exp(eta) - 1
  
  VEG
}

```

----

----

----

----

----

----

----

----


```{r,echo=FALSE}
# Translate z-scaled dam density scale from modeled data to (0,3,6,9,12,15):

damDenTlate <- read.csv("damDensityTranslator.csv")
#damDenTlate
```


----

----

----

----


NOTE:
----
    
    The only maps presented below represented the predicted difference in surface
    water or vegetation area bewteen baseline (0 dams km^-1) and 15 dams km^-1. Maps
    can be displayed at other dam densities, but the RMarkdown file is large and un-
    wieldy with multiple leaflet maps in it. GitHub doesn't allow backup of files 
    larger that 100 MB. 
    
    15 dams^km-1 is near the max observed dam density in our sample reaches. Even
    at max observed density, effects are meager.


----

----

Predict Surface Water & Render Map:
----

    Modeled for late August (08/15-08/31) surface water area
    
    Response variable: log(surfaceWaterArea + 1)
    
    Fixed Effects: damDensity + incision + riparianArea + drainageArea
    
    Random Effects: stream[s], year[y], event[e] as random slopes; 
                    stream[s], event[e] as random intercepts
    
    NOTE: for these predictive maps, "year" RE was fixed to the estimated 
          posterior mean for 2021; "event" RE was fixed to the estimated posterior
          mean for the second survey period Aug/15 - Aug/31.



```{r,echo=FALSE}
################################################

# PREDICT SWA and COMPUTE DELTA_SW for each window and dam density

# dam_densities <- c(0,3,6,9,12,15)

df_delta_sw <- df_windows %>%
  rowwise() %>%
  mutate(
    SW_0  = predict_SW_window(incis_mean, RIPA_mean, UDA_mean,
                              damDensity = -0.9396414,   # 0 dam/km
                              watershed = watershed,
                              coefs_SW_fixed = coefs_SW_fixed,
                              coefs_SW_random_BetaStream =  coefs_SW_random_BetaStream,
                              coefs_SW_random_etaStream = coefs_SW_random_etaStream,
                              coefs_SW_random_BetaYear = coefs_SW_random_BetaYear,
                              coefs_SW_random_lateAug21_etaEvent = coefs_SW_random_lateAug21_etaEvent,
                              coefs_SW_random_lateAug21_betaEvent = coefs_SW_random_lateAug21_betaEvent
                              ),
    SW_3  = predict_SW_window(incis_mean, RIPA_mean, UDA_mean,
                              damDensity = 0.00609353,   # ~3 dam/km
                              watershed = watershed,
                              coefs_SW_fixed = coefs_SW_fixed,
                              coefs_SW_random_BetaStream =  coefs_SW_random_BetaStream,
                              coefs_SW_random_etaStream = coefs_SW_random_etaStream,
                              coefs_SW_random_BetaYear = coefs_SW_random_BetaYear,
                              coefs_SW_random_lateAug21_etaEvent = coefs_SW_random_lateAug21_etaEvent,
                              coefs_SW_random_lateAug21_betaEvent = coefs_SW_random_lateAug21_betaEvent
                              ),
    SW_6 = predict_SW_window(incis_mean, RIPA_mean, UDA_mean,
                              damDensity = 0.9383662,   # ~6 dam/km
                              watershed = watershed,
                              coefs_SW_fixed = coefs_SW_fixed,
                              coefs_SW_random_BetaStream =  coefs_SW_random_BetaStream,
                              coefs_SW_random_etaStream = coefs_SW_random_etaStream,
                              coefs_SW_random_BetaYear = coefs_SW_random_BetaYear,
                              coefs_SW_random_lateAug21_etaEvent = coefs_SW_random_lateAug21_etaEvent,
                              coefs_SW_random_lateAug21_betaEvent = coefs_SW_random_lateAug21_betaEvent
                             ),
    SW_9 = predict_SW_window(incis_mean, RIPA_mean, UDA_mean,
                              damDensity = 1.932580,  # ~9 dam/km
                              watershed = watershed,
                              coefs_SW_fixed = coefs_SW_fixed,
                              coefs_SW_random_BetaStream =  coefs_SW_random_BetaStream,
                              coefs_SW_random_etaStream = coefs_SW_random_etaStream,
                              coefs_SW_random_BetaYear = coefs_SW_random_BetaYear,
                              coefs_SW_random_lateAug21_etaEvent = coefs_SW_random_lateAug21_etaEvent,
                              coefs_SW_random_lateAug21_betaEvent = coefs_SW_random_lateAug21_betaEvent
                             ),
	SW_12 = predict_SW_window(incis_mean, RIPA_mean, UDA_mean,
	                          damDensity = 2.865453, #~12 dams/km
                              watershed = watershed,
                              coefs_SW_fixed = coefs_SW_fixed,
                              coefs_SW_random_BetaStream =  coefs_SW_random_BetaStream,
                              coefs_SW_random_etaStream = coefs_SW_random_etaStream,
                              coefs_SW_random_BetaYear = coefs_SW_random_BetaYear,
                              coefs_SW_random_lateAug21_etaEvent = coefs_SW_random_lateAug21_etaEvent,
                              coefs_SW_random_lateAug21_betaEvent = coefs_SW_random_lateAug21_betaEvent
	                          ),
    SW_15 = predict_SW_window(incis_mean, RIPA_mean, UDA_mean,
                              damDensity = 3.881626,   # ~15 dam/km
                              watershed = watershed,
                              coefs_SW_fixed = coefs_SW_fixed,
                              coefs_SW_random_BetaStream =  coefs_SW_random_BetaStream,
                              coefs_SW_random_etaStream = coefs_SW_random_etaStream,
                              coefs_SW_random_BetaYear = coefs_SW_random_BetaYear,
                              coefs_SW_random_lateAug21_etaEvent = coefs_SW_random_lateAug21_etaEvent,
                              coefs_SW_random_lateAug21_betaEvent = coefs_SW_random_lateAug21_betaEvent
                              ),
	delta_SW_3  = SW_3  - SW_0,
	delta_SW_9 = SW_9 - SW_0,
	delta_SW_15 = SW_15 - SW_0
  ) %>%
  ungroup()



colnames(df_delta_sw) <- c("watershed",
                           "window_id",
                           "TP_ID",
                           "TP_ID_end",
                           "incis_mean",
                           "incis_sd",
                           "RIPA_mean",
                           "RIPA_sd",
                           "UDA_mean",
                           "UDA_sd",
                           "SW_0",
                           "SW_3",
                           "SW_6",
                           "SW_9",
                           "SW_12",
                           "SW_15",
                           "delta_SW_3",
                           "delta_SW_9",
                           "delta_SW_15")
```

```{r,echo=FALSE}

df_predSWChange_map <- aoi_forPrediction %>%
  left_join(
    df_delta_sw %>%
      select(watershed, window_id, TP_ID, TP_ID_end,
             delta_SW_3, delta_SW_9, delta_SW_15),
    by = c("watershed","TP_ID")
  ) %>%
  drop_na() %>% 
  ungroup() %>% 
  st_as_sf() %>% 
  st_transform(4326)

  

```


    damDen[0] = baseline; no-dam condition
    
    damDen[k] = dammed condition (3,6,9,12,15) dams km^-1
    
    dams km^-1 = "dams per kilometer"



<!---

**Absolute difference in SWA associated with +3 dams km^-1:**

(Predicted surface water area at dam density [k]) - (predicted baseline surface water area)


```{r,echo=FALSE}
pal_full <- RColorBrewer::brewer.pal(11, "Spectral")

# keep only the darker 5 colors (skip the light yellows)
pal_trimmed <- pal_full[6:11]

palDelt <- colorNumeric(
  palette = pal_full,
  domain  = df_predSWChange_map$delta_SW_3
)
```
```{r,echo=FALSE}

leaflet(df_predSWChange_map) %>%
  addProviderTiles(providers$USGS.USImageryTopo) %>%
  addPolygons(
    data = df_predSWChange_map,
    fillColor = ~palDelt(delta_SW_3),
    fillOpacity = 0.5,
    color = "grey40",
    weight = 0.5,
    popup = ~paste0(
      "watershed: ", watershed,
      "<br>Window: ", window_id,
      "<br>Absolute difference in SWA with + ~3 dams/km: ", round(delta_SW_3, 2)
    )
  ) %>%
  addLegend("bottomright", pal = palDelt, values = df_predSWChange_map$delta_SW_3,
            title = "Absolute difference in SWA w/added 3 dams/km") %>%
  addScaleBar(
    position = "bottomleft",
    options = scaleBarOptions(
      maxWidth = 100,
      metric = TRUE,
      imperial = FALSE,
      updateWhenIdle = TRUE
    )
  )


```


----


**Absolute difference in SWA associated with +9 dams km^-1:**


```{r,echo=FALSE}

palDelt9 <- colorNumeric(
  palette = pal_full,
  domain  = df_predSWChange_map$delta_SW_9
)
```

```{r,echo=FALSE}

leaflet(df_predSWChange_map) %>%
  addProviderTiles(providers$USGS.USImageryTopo) %>%
  addPolygons(
    data = df_predSWChange_map,
    fillColor = ~palDelt9(delta_SW_9),
    fillOpacity = 0.5,
    color = "grey40",
    weight = 0.5,
    popup = ~paste0(
      "watershed: ", watershed,
      "<br>Window: ", window_id,
      "<br>Absolute difference in SWA with + ~9 dams/km: ", round(delta_SW_9, 2)
    )
  ) %>%
  addLegend("bottomright", pal = palDelt9, values = ~delta_SW_9,
            title = "Absolute difference in SWA with +9 dams km^-1") %>%
  addScaleBar(
    position = "bottomleft",
    options = scaleBarOptions(
      maxWidth = 100,
      metric = TRUE,
      imperial = FALSE,
      updateWhenIdle = TRUE
    )
  )
```

--->

----


**Absolute difference in SWA associated +15 dams km^-1:**

```{r,echo=FALSE}
#pal_full <- RColorBrewer::brewer.pal(11, "Spectral")

# keep only the darker 5 colors (skip the light yellows)
#pal_trimmed <- pal_full[6:11]

palDelt15 <- colorNumeric(
  palette = pal_full,
  domain  = df_predSWChange_map$delta_SW_15
)
```
```{r,echo=FALSE}

leaflet(df_predSWChange_map) %>%
  addProviderTiles(providers$USGS.USImageryTopo) %>%
  addPolygons(
    data = df_predSWChange_map,
    fillColor = ~palDelt15(delta_SW_15),
    fillOpacity = 0.5,
    color = "grey40",
    weight = 0.5,
    popup = ~paste0(
      "watershed: ", watershed,
      "<br>Window: ", window_id,
      "<br>Absolute difference in SWA with + ~15 dams/km: ", round(delta_SW_15, 2)
    )
  ) %>%
  addLegend("bottomright", pal = palDelt15, values = ~delta_SW_15,
            title = "Absolute difference in SWA w/added 15 dams/km") %>%
  addScaleBar(
    position = "bottomleft",
    options = scaleBarOptions(
      maxWidth = 100,
      metric = TRUE,
      imperial = FALSE,
      updateWhenIdle = TRUE
    )
  )


```


----

----

----

----

----

----

----

----

----

----

----

Predict Vegetation Area & Render Map:
----

    Modeled for late August (08/15-08/31) vegetation
    
    Response variable: log(vegetationArea + 1)
    
    Fixed Effects: damDensity + incision + riparianArea + drainageArea
    
    Random Effects: stream[s], year[y]
      NOTE: for these predictive maps, "year" RE was fixed to the estimated 
            posterior mean for 2021; no seasonal effect included in veg models
            due to lack of variation during late summer (Aug-Sep)
            

```{r,echo=F}
################################################

# PREDICT VA and COMPUTE DELTA_VA for each window and dam density

# dam_densities <- c(0,3,6,9,12,15)

df_delta_veg <- df_windows %>%
  rowwise() %>%
  mutate(
    VEG_0  = predict_VEG_window(incis_mean, RIPA_mean, UDA_mean,
                              damDensity = -0.9396414,   # 0 dam/km
                              watershed = watershed,
                              coefs_VEG_fixed = coefs_VEG_fixed,
                              coefs_VEG_random_etaStream = coefs_VEG_random_etaStream,
                              coefs_VEG_random_etaYear = coefs_VEG_random_etaYear
                              ),
    VEG_3  = predict_VEG_window(incis_mean, RIPA_mean, UDA_mean,
                              damDensity = 0.00609353,   # ~3 dam/km
                              watershed = watershed,
                              coefs_VEG_fixed = coefs_VEG_fixed,
                              coefs_VEG_random_etaStream = coefs_VEG_random_etaStream,
                              coefs_VEG_random_etaYear = coefs_VEG_random_etaYear
                              ),
    VEG_6 = predict_VEG_window(incis_mean, RIPA_mean, UDA_mean,
                              damDensity = 0.9383662,   # ~6 dam/km
                              watershed = watershed,
                              coefs_VEG_fixed = coefs_VEG_fixed,
                              coefs_VEG_random_etaStream = coefs_VEG_random_etaStream,
                              coefs_VEG_random_etaYear = coefs_VEG_random_etaYear
                             ),
    VEG_9 = predict_VEG_window(incis_mean, RIPA_mean, UDA_mean,
                              damDensity = 1.932580,  # ~9 dam/km
                              watershed = watershed,
                              coefs_VEG_fixed = coefs_VEG_fixed,
                              coefs_VEG_random_etaStream = coefs_VEG_random_etaStream,
                              coefs_VEG_random_etaYear = coefs_VEG_random_etaYear
                             ),
	VEG_12 = predict_VEG_window(incis_mean, RIPA_mean, UDA_mean,
	                          damDensity = 2.865453, #~12 dams/km
                              watershed = watershed,
                              coefs_VEG_fixed = coefs_VEG_fixed,
                              coefs_VEG_random_etaStream = coefs_VEG_random_etaStream,
                              coefs_VEG_random_etaYear = coefs_VEG_random_etaYear
	                          ),
    VEG_15 = predict_VEG_window(incis_mean, RIPA_mean, UDA_mean,
                              damDensity = 3.881626,   # ~15 dam/km
                              watershed = watershed,
                              coefs_VEG_fixed = coefs_VEG_fixed,
                              coefs_VEG_random_etaStream = coefs_VEG_random_etaStream,
                              coefs_VEG_random_etaYear = coefs_VEG_random_etaYear
                              ),
	delta_VEG_3  = VEG_3  - VEG_0,
	delta_VEG_9 = VEG_9 - VEG_0,
	delta_VEG_15 = VEG_15 - VEG_0
  ) %>%
  ungroup()



colnames(df_delta_veg) <- c("watershed",
                           "window_id",
                           "TP_ID",
                           "TP_ID_end",
                           "incis_mean",
                           "incis_sd",
                           "RIPA_mean",
                           "RIPA_sd",
                           "UDA_mean",
                           "UDA_sd",
                           "VEG_0",
                           "VEG_3",
                           "VEG_6",
                           "VEG_9",
                           "VEG_12",
                           "VEG_15",
                           "delta_VEG_3",
                           "delta_VEG_9",
                           "delta_VEG_15")
```

```{r,echo=FALSE}

df_predVEGChange_map <- aoi_forPrediction %>%
  left_join(
    df_delta_veg %>%
      select(watershed, window_id, TP_ID, TP_ID_end,
             delta_VEG_3, delta_VEG_9, delta_VEG_15),
    by = c("watershed","TP_ID")
  ) %>%
  drop_na() %>% 
  ungroup() %>% 
  st_as_sf() %>% 
  st_transform(4326)

  

```

    damDen[0] = baseline; no-dam condition
    
    damDen[k] = dammed condition (3,6,9,12,15) dams km^-1
    
    dams km^-1 = "dams per kilometer"


<!---

**Absolute difference in VEG associated with +3 dams km^-1:**

```{r,echo=FALSE}
pal_full <- RColorBrewer::brewer.pal(11, "Spectral")

# keep only the darker 5 colors (skip the light yellows)
pal_trimmed <- pal_full[6:11]

palDeltVEG_3 <- colorNumeric(
  palette = pal_full,
  domain  = df_predVEGChange_map$delta_VEG_3
)
```
```{r,echo=FALSE}

leaflet(df_predVEGChange_map) %>%
  addProviderTiles(providers$USGS.USImageryTopo) %>%
  addPolygons(
    data = df_predVEGChange_map,
    fillColor = ~palDeltVEG_3(delta_VEG_3),
    fillOpacity = 0.5,
    color = "grey40",
    weight = 0.5,
    popup = ~paste0(
      "watershed: ", watershed,
      "<br>Window: ", window_id,
      "<br>Absolute difference in VA with + ~3 dams/km: ", round(delta_VEG_3, 2)
    )
  ) %>%
  addLegend("bottomright", pal = palDeltVEG_3, values = df_predVEGChange_map$delta_VEG_3,
            title = "Absolute difference in VA with +3 dams km^-1")


```


----


**Absolute difference in VEG associated with +9 dams km^-1:**


```{r,echo=FALSE}

palDeltVEG_9 <- colorNumeric(
  palette = pal_full,
  domain  = df_predVEGChange_map$delta_VEG_9
)
```

```{r,echo=FALSE}

leaflet(df_predVEGChange_map) %>%
  addProviderTiles(providers$USGS.USImageryTopo) %>%
  addPolygons(
    data = df_predVEGChange_map,
    fillColor = ~palDeltVEG_9(delta_VEG_9),
    fillOpacity = 0.5,
    color = "grey40",
    weight = 0.5,
    popup = ~paste0(
      "watershed: ", watershed,
      "<br>Window: ", window_id,
      "<br>Absolute difference in VA with + ~9 dams km^-1: ", round(delta_VEG_9, 2)
    )
  ) %>%
  addLegend("bottomright", pal = palDeltVEG_9, values = ~delta_VEG_9,
            title = "Absolute difference in VA with +9 dams km^-1")
```

--->

----



**Absolute difference in VEG associated with +15 dams km^-1:**

```{r,echo=FALSE}
pal_full <- RColorBrewer::brewer.pal(11, "Spectral")

# keep only the darker 5 colors (skip the light yellows)
pal_trimmed <- pal_full[6:11]

palDeltVEG15 <- colorNumeric(
  palette = pal_full,
  domain  = df_predVEGChange_map$delta_VEG_15
)
```

```{r,echo=FALSE}

leaflet(df_predVEGChange_map) %>%
  addProviderTiles(providers$USGS.USImageryTopo) %>%
  addPolygons(
    data = df_predVEGChange_map,
    fillColor = ~palDeltVEG15(delta_VEG_15),
    fillOpacity = 0.5,
    color = "grey40",
    weight = 0.5,
    popup = ~paste0(
      "watershed: ", watershed,
      "<br>Window: ", window_id,
      "<br>Absolute difference in VA with + ~15 dams km^-1: ", round(delta_VEG_15, 2)
    )
  ) %>%
  addLegend("bottomright", pal = palDeltVEG15, values = ~delta_VEG_15,
            title = "Absolute difference in VA with +15 dams km^-1") %>%
  addScaleBar(
    position = "bottomleft",
    options = scaleBarOptions(
      maxWidth = 100,
      metric = TRUE,
      imperial = FALSE,
      updateWhenIdle = TRUE
    )
  )


```